### 2019-03-15 README.Rmd
initial changes

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below. 
Domproject <- "Title of your paper goes here"
```

# domproject

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to download or install

You can download the compendium as a zip from from this URL: </archive/master.zip>

Or you can install this compendium as an R package, domproject, from GitHub with:
```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("DCPollard94/domproject")
```
### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2019-03-15 analysis/paper/paper.Rmd
initial changes

---
title: "Title Goes Here"
author:
  - Author One
  - Author Two
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  comment = "#>",
  fig.path = "../figures/"
)

# Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
```

# Introduction

Here is a citation [@Marwick2017]

# Background

```{r flowers, fig.cap="Here are some flower measurements"}
library(ggplot2)
ggplot(iris,aes(Sepal.Width,Sepal.Length))+geom_point()
```

We can make a plot of `r nrow(iris)` flower measurements (Fig \@ref(fig:flowers))

```{r a-table}
knitr::kable(head(iris),
             caption="A table of flowers")
```

We can also make a table (Table \@ref(tab:a-table))

# Methods

# Results

```{r get-data, eval = FALSE}
# Note the path that we need to use to access our data files when rendering this document
my_data <- readr::read_csv(here::here('analysis', 'data', 'raw_data', 'my_csv_file.csv'))
```

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
##### pagebreak

# References 
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

##### pagebreak

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
git2r::repository(here::here())
```

### 2019-03-15 README.Rmd
Readme and description changes

@@
 
-```{r, echo = FALSE}
+```{r Title, echo = FALSE}
 knitr::opts_chunk$set(
@@
 # Please put your title here to include it in the file below. 
-Domproject <- "Title of your paper goes here"
+Title <- "Family Fortunes: A Quantitative Analysis of the Early Iron Age Cemeteries of Knossos"
 ```
 
-# domproject
+# knossoscemeteries
 
-This repository contains the data and code for our paper:
+This repository contains the data and code for my paper:
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
+> Pollard, Dominic (2019). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
 
-Our pre-print is online here:
+The pre-print available is online here:
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Pollard, Dominic (2019). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
 
@@
 # install.packages("devtools")
-devtools::install_github("DCPollard94/domproject")
+devtools::install_github("DCPollard94/knossoscemeteries")
 ```

### 2019-03-15 analysis/paper/paper.Rmd
bibliography and templates

@@
 ---
-title: "Title Goes Here"
+title: "Family Fortunes: A Quantitative Analysis of the Early Iron Age Cemeteries at Knossos, Crete"
 author:
-  - Author One
-  - Author Two
+  - Dominic Pollard
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 bibliography: references.bib
-csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
+csl: "../templates/world-archaeology.csl" # Insert path for the bib-style
 abstract: |

### 2019-03-15 analysis/templates/template.Rmd
bibliography and templates

---
title: "My title"
author:
- author 1
- other author
date: '`r Sys.Date()`'
output: word_document
abstract: "My abstract \n"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Heading 1

## Heading 2

### Heading 3

#### Heading 4

##### Heading 5 - this is for the page break with MS Word

In the MS Word document, the following modifications must be made in the 'Modify Style' menu:

+ Set the font color to ‘white’ (rather than ‘Automatic’).
+ Select the smallest font size (8 rather than 11).
+ Select ‘Page break before’ in the ‘Line and Page Breaks’ tab.
+ Set the line spacing to ‘Exactly’ and ‘1 pt’ in the ‘Indents and Spacing’ tab.

After these tweaks, the ‘Heading 5’ style will no longer format a heading of level 5. Instead it will insert a very small and white (and, thus, invisible) line followed by a page break.

Now knit this doc, and edit the styles in the resulting docx file. Then rename as 'template' and use this in the yml:

```
output:
  word_document:
    reference_docx: templates/template1.docx
```

Don't forget line numbers for JAS! Headings are bold, underline, nothing. And Arial Narrow 10pt for tables.

See here for more information: http://rmarkdown.rstudio.com/articles_docx.html

### 2019-03-15 analysis/paper/paper.Rmd
Reference changes
updated references
Beginning writing

--- 2019-03-15 analysis/paper/paper.Rmd
@@
       fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
-bibliography: references.bib
-csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
+      reference_docx: "../templates/articletemplate.docx" # Insert path for the DOCX file
+bibliography: Phd.bib
+csl: "../templates/world-archaeology.csl" # Insert path for the bib-style
 abstract: |
@@
 # Introduction
+It may be felt that, of all ancient sites on Crete, Knossos is the last in need of further archaeological attention.
+The city and ‘palace’ of Knossos, first excavated well over a century ago [@Evans1921; -@Evans1928; -@Evans1930], retain a pre-eminent position in popular and academic accounts the island’s Bronze Age, and few, if any, other Cretan sites have been so thoroughly dug, reconstructed, and contested.
+However, there are reasons why the present study – a quantitative comparison of the two largest Early Iron Age cemeteries at Knossos – is not only justified, but also timely.
 
-Here is a citation [@Marwick2017]
+For one, this is not the Knossos of that illustrous, palatial period, but rather the town as it persisted and evolved during the subsequent Early Iron Age (EIA).
+Research into this once maligned period on which research has grown apace in the last few decades.
+The reframing of the the twelth-seventh centuries as the gestational period of the later classical Greek world [first seen in the works of @Coldstream1977; @Desborough1972; @Snodgrass1971] came first, while more recently scholars have refuted the image of a society bereft of its former palatial glory.
+Certainly, few would any longer defend the appelation or interpretive bagguage of the ‘Dark Ages’ [on the origins of which, see @Kotsonas2016; @Morris1997].Nonetheless, our knowledge of many EIA settlements (both new foundations, and those with BA ancestry like Knossos) remains often underdeveloped in comparison with those of the second millennium BC.
+
+Secondly, and welcome though they are, a number of recent publications on the socio-economic [@Wallace2010a] and settlement [@Nowicki2000] changes which accompanied the BA-EIA transition have directed scholarly attention toward a number of elevated, defensible settlements supposedly founded in response to the turbulent contemporary political climate.
+This has, unfortunately, engendered a certain disregard for continuing occupation at several lowland centres, including Knossos and Phaistos.
+There remains, therefore, a need for studies addressing the social dynamics at these sites, with a long-term view to comparative appraisal of the divergent trajectories of communities across the island in the EIA [@Whitley2011, 667–8].
+
+Finally, a number of authors, drawing on datasets from Crete and the wider Aegean, have in recent years demonstrated the great potential of quantitative, statistical, spatial, and network analyses to elucidate social dynamics, settlement patterning, and systems of inter-regional communication and exchange.
+Such work encourages the use of large (and often pre-existing) datasets, and the sharing of both data and methods to facilitate reproducible, collaborative, and innovative research.
+The cemeteries of Knossos offer an opportunity to introduce such approaches into EIA research, where they have yet to make much of an impact.
+The Fortesta and North Cemetery complexes are both well-published (Brock 1957; Coldstream and Catling 1996) and have been considered variously in relation to their oriental imports (Antoniadis 2012), antiques (Crowe 2016), and religious context (Coldstream 1984a).
+But a lack of quantitative analyses, or systematic comparisons between different cemeteries, has left us with many intuited trends, some clearly exceptional tombs, and a general impression of un-patterned multiplicity in mortuary rituals.
+I suggest that the application of a range of quantitative and statistical methods may provide new insight into the social behaviours underlying the mortuary assemblages.
+
+A recent paper by Kotsonas (2011)  has stressed the importance of quantification in analyses of mortuary material.
+Drawing on data from the EIA Knossian and Eleuthernian cemeteries, he illustrates the clear potential of quantitative methods for elucidating social behaviours.
+Taking the lead from the Kotsonas, I here adopt a broader, more comparative, and more quantitative approach to the Knossian material than has previously been employed.
+With a dataset including information on every find and tomb recorded in the Fortetsa and Knossos North Cemetery publications (Brock 1957; Coldstream and Catling 1996), the following analysis examines both the synchronic and diachronic variations among the tombs of both cemeteries.
+In doing so, I aim to demonstrate the potential for quantitative methods, firstly, to permit more reliable identification, and more nuanced characterisation, of significant patterning in the archaeological material; secondly, to facilitate a more systematic appraisal of the cemeteries’ largest tombs, and set their development in a richer socio-economic context; and, finally, to offer not just generalities, but evidence by which to consider, in tangible and context-specific ways, the motivations and identities of the burial groups themselves.
 
@@
 
-We can make a plot of `r nrow(iris)` flower measurements (Fig \@ref(fig:flowers))
+The end [@Marwick2017].
 

### 2019-03-18 analysis/paper/paper.Rmd
Paper edits

@@
 author:
-  - Dominic Pollard
+  - Dominic Pollard, Institute of Archaeology, University College London
+  - dominic.pollard.16@ucl.ac.uk
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 abstract: |
-  Text of abstract
+  The Early Iron Age cemeteries of Knossos, Crete, represent the richest source of archaeological material at the site for the whole period between the end of the Bronze Age and the emergence of the Archaic *polis*.
+However, despite well-published excavation reports on the two largest burial grounds, the Fortetsa and Knossos North cemeteries, and a number of studies dealing with various artefact classes and individual tombs therein, no substantial comparative, quantitative analysis of these two cemeteries has yet been undertaken.
+This thesis provides such an analysis, examining both the synchronic and diachronic variation among the tombs in these two cemeteries.
+Through the application of a range of quantitative techniques, this study isolates a pattern of increased rates of burial among a few select tombs, mostly founded in the 9th century BC.
+Through comparing the compositions of the different tomb assemblages, it is suggested that these tombs represent certain extended-lineage groups, perhaps akin to the ‘tribes’ or ‘clans’ of later textual sources, and that their exceptional burial numbers and rates reflect social strategies developed in the contested ideological context of the nascent Knossian polis.
+This thesis demonstrates the potential for quantitative analyses to contribute to nuanced, context-specific theories of social practice, and advocates their wider application to the archaeological record of the Cretan Early Iron Age.
 keywords: |
-  keyword 1; keyword 2; keyword 3
-highlights: |
-  These are the highlights. 
+  Early Iron Age; Knossos; Burial; Quantitative;
+
 ---
@@
 For one, this is not the Knossos of that illustrous, palatial period, but rather the town as it persisted and evolved during the subsequent Early Iron Age (EIA).
-Research into this once maligned period on which research has grown apace in the last few decades.
-The reframing of the the twelth-seventh centuries as the gestational period of the later classical Greek world [first seen in the works of @Coldstream1977; @Desborough1972; @Snodgrass1971] came first, while more recently scholars have refuted the image of a society bereft of its former palatial glory.
+Research into this once maligned period has grown apace in the last few decades.
+The reframing of the the twelth-seventh centuries as the gestational period of the later classical Greek world [first seen in the works of @Coldstream1977; @Desborough1972; @Snodgrass1971] came first, while, more recently, scholars have refuted the image of a society bereft of its former palatial glory.
 Certainly, few would any longer defend the appelation or interpretive bagguage of the ‘Dark Ages’ [on the origins of which, see @Kotsonas2016; @Morris1997].Nonetheless, our knowledge of many EIA settlements (both new foundations, and those with BA ancestry like Knossos) remains often underdeveloped in comparison with those of the second millennium BC.
@@
 Secondly, and welcome though they are, a number of recent publications on the socio-economic [@Wallace2010a] and settlement [@Nowicki2000] changes which accompanied the BA-EIA transition have directed scholarly attention toward a number of elevated, defensible settlements supposedly founded in response to the turbulent contemporary political climate.
-This has, unfortunately, engendered a certain disregard for continuing occupation at several lowland centres, including Knossos and Phaistos.
-There remains, therefore, a need for studies addressing the social dynamics at these sites, with a long-term view to comparative appraisal of the divergent trajectories of communities across the island in the EIA [@Whitley2011, 667–8].
+This research agenda is likewise reflected in survey work, which has focussed extensively around the Bay of Mirabello, and the isthmus of Ierapetra.
+This has, unfortunately, engendered a certain disregard for continuing occupation at several lowland centres in the central region of the island, including Knossos and Phaistos.
+There remains, therefore, a need for studies addressing the social dynamics at these sites, with a long-term view to comparative appraisal of the divergent trajectories of communities across the island in the EIA [@Whitley2011, 667–8]
 
@@
 The cemeteries of Knossos offer an opportunity to introduce such approaches into EIA research, where they have yet to make much of an impact.
-The Fortesta and North Cemetery complexes are both well-published (Brock 1957; Coldstream and Catling 1996) and have been considered variously in relation to their oriental imports (Antoniadis 2012), antiques (Crowe 2016), and religious context (Coldstream 1984a).
+The Fortesta and North Cemetery complexes are both well-published [@Brock1957; @Coldstream1996a] and have been considered variously in relation to their oriental imports [@Antoniadis2012], antiques [@Crowe2016], and religious context [@Coldstream1984b].
 But a lack of quantitative analyses, or systematic comparisons between different cemeteries, has left us with many intuited trends, some clearly exceptional tombs, and a general impression of un-patterned multiplicity in mortuary rituals.
@@
 
-A recent paper by Kotsonas (2011)  has stressed the importance of quantification in analyses of mortuary material.
-Drawing on data from the EIA Knossian and Eleuthernian cemeteries, he illustrates the clear potential of quantitative methods for elucidating social behaviours.
-Taking the lead from the Kotsonas, I here adopt a broader, more comparative, and more quantitative approach to the Knossian material than has previously been employed.
-With a dataset including information on every find and tomb recorded in the Fortetsa and Knossos North Cemetery publications (Brock 1957; Coldstream and Catling 1996), the following analysis examines both the synchronic and diachronic variations among the tombs of both cemeteries.
+A recent paper by Kotsonas [-@Kotsonas2011]  has stressed the importance of quantification in analyses of mortuary material.
+Using data drawn from the EIA Knossian and Eleuthernian cemeteries, Kotsonas demonstrates the capacity of even quite basic methods to problematise commonplace assumptions surroudning mortuary behaviour.
+In this spirit, I here adopt a broader, more comparative, and more quantitative approach to the Knossian material than has previously been employed.
+With a dataset including information on every find and tomb recorded in the Fortetsa and Knossos North Cemetery publications [@Brock1957; @Coldstream1996a, the following analysis examines both the synchronic and diachronic variation among the tombs of both cemeteries.
 In doing so, I aim to demonstrate the potential for quantitative methods, firstly, to permit more reliable identification, and more nuanced characterisation, of significant patterning in the archaeological material; secondly, to facilitate a more systematic appraisal of the cemeteries’ largest tombs, and set their development in a richer socio-economic context; and, finally, to offer not just generalities, but evidence by which to consider, in tangible and context-specific ways, the motivations and identities of the burial groups themselves.
 
-# Background
+# Early Iron Age Knossos
+## Settlement Evidence: 'A meagre filling in a very thick sandwich'
+
+Settlement evidence at EIA Knossos is, in a word, insubstantial.
+Speaking of the town’s stratigraphic sequence, Coldstream lamented that “the remains of [this] period are like a meagre filling in a very thick sandwich” between Bronze Age and later Greek and Roman levels [@Coldstream1991, 287].
+Most individual finds of Protogeometric or Geometric date take the form of flimsy foundations, patches of earth floor, wells or simple scatters of pottery, and for a long time it was believed the town contracted, moved, or even dissolved following the collapse of the palace.
+Indeed, Alexiou (1950 CITATION NEEDED) thought it likely that the area comprised several dispersed villages, which later coalesced into a single city, as per Aristotle’s model of synoecism.
+
+However, by the time of Hood and Smyth’s [-@Hood1981] survey-cum-synthesis of prior excavations, a concentration of activity slightly to the west of the old palace presented a good case for continuing, nucleated settlement, something strongly advocated by Coldstream [-@Coldstream1984a; -@Coldstream2000].
+More recently, the findings of the Knossos Urban Landscape Project suggest a PG settlement of up to 40ha, extending from the Kairatos River to the western slope of the Acropolis, and from the northern slope of Gypsades to between the Minoan palace and the Kephala hill [@Kotsonas2011a, 5-8].
+This estimate remains our best guess in the absence of substantial excavatied deposits but, if it is accurate, then EIA Knossos was one of the largest sites of the contemporary Aegean, and we should expect the existence of relatively complex forms of social interaction and organisation within this exceptional community.
+
+## The Early Iron Age Cemeteries
+
+Survey aside, our most abundant evidence for EIA Knossos comes from the mortuary record.
+The transitional period between the LBA and EIA, admittedly, yields few if any securely datable burials.
+In the EIA proper, however, the evidence becomes more plentiful, with multiple burials known from the Kephala ridge [@Coldstream1963, 38; @Hogarth1899, 82-5], the modern suburbs of Ayios Ioannis and Atsalenio [@Boardman1960; @Davaras1968, 133–46], the cave of Mavro Spelio [@Antoniadis2012, 58-9], and the Khaniale Teke site with its well-known tholos, mooted as evidence for the presence immigrant craftsmen from the Near East [@Boardman1967;@Hutchinson1954; cf.
+@Hoffman1997 @Kotsonas2006].
+Taken together, these scattered tombs point to a variegated set of funerary practices, all within about 30 minutes’ walk of each other, each serving some sub-section of the Knossian population.
+But despite their reasonable number when considered collectively, these tombs, so dispersed and often solitary, provide little scope for systematic quantitative analysis.
+
+Fortunately, far larger burial accumulations are to be found in the Fortetsa and the Knossos North cemeteries, which together comprise some 134 tombs, ranging in date from the Subminoan to Late Orienatalising periods – a span of around 400 years.
+Though many were robbed in antiquity, and others destroyed by the digging of later burials, building work, or just the ravages of time, they still represent the most comprehensive and abundant source of evidence for not just mortuary practices, but for social life of any kind at EIA Knossos.
+They are thus the ideal candidates for a quantitative analysis of the kind advanced here.
+
+## The Fortetsa and the Knossos North Cemeteries
+
+The cemetery on the slope facing the eponymous village of Fortetsa is composed of twelve tombs excavated in 1933, and another eight uncovered in 1935 (Brock 1957, xi).
+These tombs cluster in three main groups.
+The largest of these, including tombs IX, VIII, VII, V, II, III, IV, V and P, is arranged on a north-south axis, with the *dromos* of each tomb extending downslope to the west (Tomb P is set on a slightly different orientation, though, as we shall see, this tomb is particularly exceptional).
+Further north lies the cluster of tombs OD, X, XI, LST and BLT.
+The third group lies south of the first, and comprises tombs Θ, Ϙ and F.
+Finally, three tombs excavated in 1933 – L, TFT and Π – which lie close to the main road north of the Acropolis hill [@Brock1957, 1-2] are now considered to belong the Knossos North Cemetery.
+
+To the northeast of the Fortetsa cemetery, a collection of ten badly damaged tombs known, after the year of their excavation, as the Fortetsa 1967 tombs, were probably the southernmost burials of the North Cemetery.
+Unfortunately, all had been looted at the time of excavation, and no in situ deposit was found in any tomb.
+For this reason, unfortunately, they feature little in the present analysis.
+Northwest of these tombs, and just south of the Teke Tholos, the Teke tombs represent the northern reaches of the KNC.
+The Teke tholos itself, and the tombs around it, were almost certainly part of the same complex, with six unpublished tombs excavated by the Archaeological Service during the construction of intervening buildings in the 1980s.
+The central section of the KNC comprises the ‘Medical Faculty’ site, so named because it was the construction of an extension for the university of Crete that prompted major rescue excavations here in 1978.
+Some 89 tombs of Subminoan to Late Orientalising date were found, by far the largest single concentration of EIA tombs in the whole Knossos area [@Coldstream1996a, 53-55].
+
+The 134 tombs  that make up these two cemeteries, some of which were in use for several centuries, suggest an established, relatively stable community for whom tradition formed an important foundation in mortuary beliefs and behaviours [@Coldstream1994, 106-108].
+The tombs fall into three main types, the chamber tomb, shaft grave, and pit cave or tomb; the latter two are largely Subminoam phenomena, with the former by far the most abundant.
+The chamber tomb comprises a descending ramp or passage, the *dromos*, leading to the chamber itself, cut into the bedrock.
+The *dromos* and chamber are separated by the *stomion*, an opening often set a step or two lower than the *dromos*, which is covered by a large stone slab, smaller stacked stones, or a combination of both (Antoniadis 2012, 47–8).
+It is not uncommon for dromoi to have niches cut into them, to house later burials that could not fit in the chamber (e.g.
+T.
+II), or even, in some cases, cinerary urns removed from the chamber so that new interments could be fitted in (e.g.
+T.
+TFT; Brock 1957, 3–4).
+
+The majority of burials in these two cemeteries are cremations, a rite that appears at Knossos in the Subminoan period, in the Tomb 200+ complex (Ts.
+200, 201 and 202 are three chambers of a single pit tomb) in the KNC (Coldstream 1994, 109).
+Inhumation does not completely disappear in the EIA, but by the Orientalising period it appears restricted to child burials, and in those instances where it is attested (Ts.
+L, 45, 98, 112 and F67:5), these burials appear to be among the very earliest in each tomb (Antoniadis 2012, 69).
+This degree of coexistence suggests that the shift between the two rites need not have marked a major ideological break; the preference for communal chamber tombs, for instance, something practiced since the Middle Minoan period at Knossos, remained constant.
+Whether this is linked to a strong sense of historical continuity or tradition, close kinship ties, or simply the need to bury large numbers of people in a small geographical area, we cannot be sure.
+
+## Chronological developments in the cemeteries
+
+The chronological sequence of the cemeteries has received much attention.
+It has been claimed that no tomb crosses the LM IIIC-SM divide (Coldstream 1998, 58; Coldstream 2006, 582), but there is a strong possibility that this is a classificatory artefact (Hallager 2010).
+The earliest tomb appears to be the T.
+200+ complex, with its precocious cremations and imported artefacts.
+There remain ongoing debates about whether the KNC represents a foundation on virgin soil, as instances of BA tombs reused the EIA are known (e.g.
+the Khaniale Teke tholos).
+Cavanagh (1996), for one, could not rule out the possibility of some KNC tombs being of BA date based on his cluster analysis, but other evidence would suggest this is unlikely.
+
+During the use of the cemeteries, correspondences have been highlighted between changing pottery styles, and patterns of tomb construction and abandonment.
+In the Protogeometric period, some 22 tombs then in use in the KNC, and nine at Fortetsa, ceased to receive interments, followed in the succeeding Protogeometric B period by the founding of 22 new tombs (16 and six respectively).
+The PGB was defined by Brock (1957, 143), who hailed it as “the most remarkable phase in Cretan vase-painting”.
+Apparently a Cretan anomaly, this phase sees Attic geometric motifs mixed with experimental patterns possibly deriving from Near Eastern metalwork, as well as iconography drawn seemingly from the Minoan repertoire (Coldstream 1984, 93–4).
+Coldstream has argued that the concurrence of this style and a rash of newly founded tombs points to a reformulation of elite mortuary display, noting how, in this period also, redeposited Minoan larnakes began appearing in and around some tombs, in one case even inspiring the decoration of a PGB vessel (in T.
+107; see Coldstream 1994, 112–3; 1984).
+Whether simply nostalgia (Coldstream 1998, 60), or a more tactical strategy by newly emerging elite groups (1994, 114–5), it seems that the Bronze Age past possessed a significant cultural cachet mobilised by various actors over the lives of these cemeteries.
+
+The final mystery of the chronological sequence at the KNC and Fortetsa is the quite abrupt cessation of new burials around 630 BCE.
+There had been changes in tomb use in both cemeteries in the preceding century, with no new tombs built in Fortetsa, and few in the KNC, though interments in old tombs continued (Cavanagh 1996, 651–3; Brock 1957, 4).
+This curiosity has, along with the equally scanty evidence for the period in the wider settlement, become known as the ‘Archaic Gap’ (Coldstream, Huxley, and Webb 1999).
+The causes of this lacuna, which persists until the recrudescence of archaeological evidence in the Late Archaic period, remain unknown.
+Although various explanations have been forwarded, (Huxley 1994, 126; Coldstream, Huxley, and Webb 1999, 301–2; Kotsonas 2002, 41–44), it seems there are now strong reasons to doubt the settlement was truly abandoned (Whitelaw, pers comm.).
+
+## Heroes and Heirlooms: Object Biographies and Exceptional Burials
+
+A popular approach to the study of Knossos’ EIA tombs has been the investigation of ‘object biography’, a term derived from Appadurai (1986) and Kopytoff (1986).
+Influenced by Mauss’ (1954 [1925]) formulation of ‘the gift’, these authors stressed the mutuable nature of an object’s social meanings, and its capacity for acquiring a ‘biography’, through the transformations of its social existence.
+In Aegean archaeology, these ideas proved attractive because of their resonance with the Homeric motif of gift-giving accompanied by accounts of an object's illustrious heritage (Whitley 2002, 220–1).
+It is hardly suprising that, when direct parallels for such objects have appeared archaeologically – such as the boar’s tusk helmet from Tomb 200+ and that given to Odysseus by Meriones (Il.
+10.260-271) – a particular interpretive lens, melding Homeric and anthropological ideas, has been applied to certain elaborate burials.
+
+Catling’s discussion of the T.
+200+ complex is a notable example.
+Tomb 201 contained the remains of at least two adults, and possibly a child.
+The burial objects included a bronze sword, spearhead, and arrowhead, an iron dirk or knife, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned BA-style boar’s tusk helmet (Catling 1995, 123).
+Catling identifies similarities between this tomb and burials at Tiryns, Kaloriziki, and the famous *hērōön* of Lefkandi.
+All contained imported items at an early date, were of ‘warrior grave’ type, and formed loci for later burials.
+Catling argues that those individuals who dared to travel abroad in this time of more limited seafaring, accruing with precious artefacts and stories of distant lands – ‘heroes returned’ in Catling’s phrase – would have been accorded exceptional treatment in death (Catling 1995, 127–8).
+
+Complementing the theme of heroes abroad has been a similar degree of interest in foreigners at home.
+Two cache of gold jewellery, found in pits flanking the entrance to the Teke Tholos, were argued by Boardman (1967) to be foundation deposits – a known Near Eastern custom – for the burial place of an immigrant craftsman, probably Phoenician, practising his trade at Knossos.
+Kotsonas has revisited this attribution, suggesting that the tholos is more likely the resting place of an elite individual or family with a monopoly over a metal workshop’s output (Kotsonas 2006, 155–9).
+While certainty in such individual cases may be moot, it seems unlikely that traders or craftsmen of non-Cretan heritage never visited or resided at Knossos during the EIA (Hoffman 1997, 176–85; Schreiber 2003, 293-306).
+
+Finally, a recurrent interest has been in the 'presence of the past' in these tombs.
+The practice of hero or ancestor cult, much discussed on the mainland (see Antonaccio 1994; Coldstream 1976), is not common here.
+Instead, we see several Late Minoan larnakes (and a few other vessels) redeposited in tombs and pits in the KNC and Fortetsa from the PGB period onward.
+A thorough analysis of these artefacts has been provided by Crowe (2016), who argues that the larnakes were robbed from tombs in the Knossos area, perhaps with a particular preference for painted specimens where available.
+A particularly evocative case comres from tomb 107, where parts of an elaborate figured LM larnax have been found along with a PGB vessel by the so-called Tree Painter.
+Both feature women in the company of birds and spiralling trees, and it seems certain that the former inspired the design of the latter.
+The PGB vessel also seems to draw on Oriental motifs, with the polos crown worn by the goddess having Near Eastern origins (Coldstream 1984a, 94–101).
+The magpie-like amalgamation of these styles and motifs seem to exemplify the particular cultural openness, fluidity and competition suggested to characterise the PGB period.
+Artefacts such as this hint towards an intimate discourse of (re)discovery; a painter, struck by the strangeness, or perhaps the uncanny familiarity, of centuries-old images, incorporated them into his own work.
+
+## Number Crunching: Quantitative Analyses of the Cemeteries
+
+Generally speaking, the above analyses have tended to rely more on particular finds, single tombs, or observable, but limited, patterns in the data to form their conclusions.
+A few authors have applied more quantative approaches, and these point toward their potential utility.
+Antoniadis (2012, 172-6), in his study of Oriental imports, divided the EIA tombs of Knossos into three groups, those with fewer than 10 pots, fewer than 50 pots, and more than 50 pots, and found that, in each successive case, the proportions of tombs containing imports and imitations increased.
+The implication is that a hierarchy existed, with differing degrees of control over the accumulation of prestigious goods.
+That said, the relationship between such items and tomb ‘wealth’ is examined critically in the present work, and may prove more complex than Antoniadis implies.
+
+Cluster analyses have been employed by both Cavanagh (1996, 653-7) and Antoniadis (2012, 193-7), the most sophisticated statistical techniques so far applied to the present material.
+Cavanagh’s analysis of tomb architecture has been mentioned, while Antoniadis’ more holistic approach corroborates quantitatively some of the intuitive correspondences noted between specific tombs in previous studies.
+Antoniadis does not interrogate his findings as much as he might have, but his highlighting of the relationship between tomb foundation dates, associated finds, and geographical location points towards social processes of self-definition and exclusion among the burial groups that would reward further investigation, something offered by the present study.
 
@@
 
-The end [@Marwick2017].
+The end
 

### 2019-03-18 analysis/paper/paper.Rmd
citations added

@@
 Most individual finds of Protogeometric or Geometric date take the form of flimsy foundations, patches of earth floor, wells or simple scatters of pottery, and for a long time it was believed the town contracted, moved, or even dissolved following the collapse of the palace.
-Indeed, Alexiou (1950 CITATION NEEDED) thought it likely that the area comprised several dispersed villages, which later coalesced into a single city, as per Aristotle’s model of synoecism.
+Indeed, Alexiou [@1950] thought it likely that the area comprised several dispersed villages, which later coalesced into a single city, as per Aristotle’s model of synoecism.
 
@@
 The chamber tomb comprises a descending ramp or passage, the *dromos*, leading to the chamber itself, cut into the bedrock.
-The *dromos* and chamber are separated by the *stomion*, an opening often set a step or two lower than the *dromos*, which is covered by a large stone slab, smaller stacked stones, or a combination of both (Antoniadis 2012, 47–8).
+The *dromos* and chamber are separated by the *stomion*, an opening often set a step or two lower than the *dromos*, which is covered by a large stone slab, smaller stacked stones, or a combination of both [@Antoniadis2012, 47-8].
 It is not uncommon for dromoi to have niches cut into them, to house later burials that could not fit in the chamber (e.g.
@@
 T.
-TFT; Brock 1957, 3–4).
+TFT; @Brock1957, 3-4).
 
 The majority of burials in these two cemeteries are cremations, a rite that appears at Knossos in the Subminoan period, in the Tomb 200+ complex (Ts.
-200, 201 and 202 are three chambers of a single pit tomb) in the KNC (Coldstream 1994, 109).
+200, 201 and 202 are three chambers of a single pit tomb) in the KNC [@Coldstream1994, 109].
 Inhumation does not completely disappear in the EIA, but by the Orientalising period it appears restricted to child burials, and in those instances where it is attested (Ts.
-L, 45, 98, 112 and F67:5), these burials appear to be among the very earliest in each tomb (Antoniadis 2012, 69).
+L, 45, 98, 112 and F67:5), these burials appear to be among the very earliest in each tomb [@Antoniadis2012, 69].
 This degree of coexistence suggests that the shift between the two rites need not have marked a major ideological break; the preference for communal chamber tombs, for instance, something practiced since the Middle Minoan period at Knossos, remained constant.
@@
 The chronological sequence of the cemeteries has received much attention.
-It has been claimed that no tomb crosses the LM IIIC-SM divide (Coldstream 1998, 58; Coldstream 2006, 582), but there is a strong possibility that this is a classificatory artefact (Hallager 2010).
+It has been claimed that no tomb crosses the LM IIIC-SM divide [@Coldstream1998, 58; @Coldstream2006, 582), but there is a strong possibility that this is a classificatory artefact [@Hallager2010].
 The earliest tomb appears to be the T.
@@
 the Khaniale Teke tholos).
-Cavanagh (1996), for one, could not rule out the possibility of some KNC tombs being of BA date based on his cluster analysis, but other evidence would suggest this is unlikely.
+Cavanagh [-@Cavanagh1996], for one, could not rule out the possibility of some KNC tombs being of BA date based on his cluster analysis, but other evidence would suggest this is unlikely.
 
@@
 In the Protogeometric period, some 22 tombs then in use in the KNC, and nine at Fortetsa, ceased to receive interments, followed in the succeeding Protogeometric B period by the founding of 22 new tombs (16 and six respectively).
-The PGB was defined by Brock (1957, 143), who hailed it as “the most remarkable phase in Cretan vase-painting”.
+The PGB was defined by Brock [-@Brock1957], 143], who hailed it as “the most remarkable phase in Cretan vase-painting”.
 Apparently a Cretan anomaly, this phase sees Attic geometric motifs mixed with experimental patterns possibly deriving from Near Eastern metalwork, as well as iconography drawn seemingly from the Minoan repertoire (Coldstream 1984, 93–4).
 Coldstream has argued that the concurrence of this style and a rash of newly founded tombs points to a reformulation of elite mortuary display, noting how, in this period also, redeposited Minoan larnakes began appearing in and around some tombs, in one case even inspiring the decoration of a PGB vessel (in T.
-107; see Coldstream 1994, 112–3; 1984).
-Whether simply nostalgia (Coldstream 1998, 60), or a more tactical strategy by newly emerging elite groups (1994, 114–5), it seems that the Bronze Age past possessed a significant cultural cachet mobilised by various actors over the lives of these cemeteries.
+107; see @Coldstream1984b; -@Coldstream1994, 112–3).
+Whether simply nostalgia [@Coldstream1998, 60], or a more tactical strategy by newly emerging elite groups [@Coldstream1994, 114-5], it seems that the Bronze Age past possessed a significant cultural cachet mobilised by various actors over the lives of these cemeteries.
 
 The final mystery of the chronological sequence at the KNC and Fortetsa is the quite abrupt cessation of new burials around 630 BCE.
-There had been changes in tomb use in both cemeteries in the preceding century, with no new tombs built in Fortetsa, and few in the KNC, though interments in old tombs continued (Cavanagh 1996, 651–3; Brock 1957, 4).
-This curiosity has, along with the equally scanty evidence for the period in the wider settlement, become known as the ‘Archaic Gap’ (Coldstream, Huxley, and Webb 1999).
+There had been changes in tomb use in both cemeteries in the preceding century, with no new tombs built in Fortetsa, and few in the KNC, though interments in old tombs continued [@Brock1957, 4; @Cavanagh1996; 651-3].
+This curiosity has, along with the equally scanty evidence for the period in the wider settlement, become known as the ‘Archaic Gap'.
 The causes of this lacuna, which persists until the recrudescence of archaeological evidence in the Late Archaic period, remain unknown.
-Although various explanations have been forwarded, (Huxley 1994, 126; Coldstream, Huxley, and Webb 1999, 301–2; Kotsonas 2002, 41–44), it seems there are now strong reasons to doubt the settlement was truly abandoned (Whitelaw, pers comm.).
+Although various explanations have been forwarded, [@Coldstream1999, 301-2; @Huxley1994, 126; @Kotsonas2002, 41-4], it seems there are now strong reasons to doubt the settlement was truly abandoned (Whitelaw, pers comm.).
 
@@
 
-A popular approach to the study of Knossos’ EIA tombs has been the investigation of ‘object biography’, a term derived from Appadurai (1986) and Kopytoff (1986).
-Influenced by Mauss’ (1954 [1925]) formulation of ‘the gift’, these authors stressed the mutuable nature of an object’s social meanings, and its capacity for acquiring a ‘biography’, through the transformations of its social existence.
-In Aegean archaeology, these ideas proved attractive because of their resonance with the Homeric motif of gift-giving accompanied by accounts of an object's illustrious heritage (Whitley 2002, 220–1).
-It is hardly suprising that, when direct parallels for such objects have appeared archaeologically – such as the boar’s tusk helmet from Tomb 200+ and that given to Odysseus by Meriones (Il.
-10.260-271) – a particular interpretive lens, melding Homeric and anthropological ideas, has been applied to certain elaborate burials.
+A popular approach to the study of Knossos’ EIA tombs has been the investigation of ‘object biography’, a term derived from Appadurai [-@Appadurai1986] and Kopytoff [-@Kopytoff1986].
+Influenced by Mauss’ [-@Mauss1954] formulation of ‘the gift’, these authors stressed the mutuable nature of an object’s social meanings, and its capacity for acquiring a ‘biography’, through the transformations of its social existence.
+In Aegean archaeology, these ideas proved attractive because of their resonance with the Homeric motif of gift-giving accompanied by accounts of an object's illustrious heritage [@Whitley2002, 220-1].
+It is hardly suprising that, when direct parallels for such objects have appeared archaeologically – such as the boar’s tusk helmet from Tomb 200+ and that given to Odysseus by Meriones (*Iliad* 10.260-271) – a particular interpretive lens, melding Homeric and anthropological ideas, has been applied to certain elaborate burials.
 
@@
 Tomb 201 contained the remains of at least two adults, and possibly a child.
-The burial objects included a bronze sword, spearhead, and arrowhead, an iron dirk or knife, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned BA-style boar’s tusk helmet (Catling 1995, 123).
+The burial objects included a bronze sword, spearhead, and arrowhead, an iron dirk or knife, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned BA-style boar’s tusk helmet [@Catling1995, 123].
 Catling identifies similarities between this tomb and burials at Tiryns, Kaloriziki, and the famous *hērōön* of Lefkandi.
 All contained imported items at an early date, were of ‘warrior grave’ type, and formed loci for later burials.
-Catling argues that those individuals who dared to travel abroad in this time of more limited seafaring, accruing with precious artefacts and stories of distant lands – ‘heroes returned’ in Catling’s phrase – would have been accorded exceptional treatment in death (Catling 1995, 127–8).
+Catling argues that those individuals who dared to travel abroad in this time of more limited seafaring, accruing with precious artefacts and stories of distant lands – ‘heroes returned’ in Catling’s phrase – would have been accorded exceptional treatment in death [@Catling1995], 127–8].
 
 Complementing the theme of heroes abroad has been a similar degree of interest in foreigners at home.
-Two cache of gold jewellery, found in pits flanking the entrance to the Teke Tholos, were argued by Boardman (1967) to be foundation deposits – a known Near Eastern custom – for the burial place of an immigrant craftsman, probably Phoenician, practising his trade at Knossos.
-Kotsonas has revisited this attribution, suggesting that the tholos is more likely the resting place of an elite individual or family with a monopoly over a metal workshop’s output (Kotsonas 2006, 155–9).
-While certainty in such individual cases may be moot, it seems unlikely that traders or craftsmen of non-Cretan heritage never visited or resided at Knossos during the EIA (Hoffman 1997, 176–85; Schreiber 2003, 293-306).
+Two caches of gold jewellery, found in pits flanking the entrance to the Teke Tholos, were argued by Boardman [-@Boardman1967] to be foundation deposits – a known Near Eastern custom – for the burial place of an immigrant craftsman, probably Phoenician, practising his trade at Knossos.
+Kotsonas has revisited this attribution, suggesting that the tholos is more likely the resting place of an elite individual or family with a monopoly over a metal workshop’s output [-@Kotsonas2006, 155–9].
+While certainty in such individual cases may be moot, it seems unlikely that traders or craftsmen of non-Cretan heritage never visited or resided at Knossos during the EIA [@Hoffman1997, 176–85; @Schreiber2003, 293-306].
 
 Finally, a recurrent interest has been in the 'presence of the past' in these tombs.
-The practice of hero or ancestor cult, much discussed on the mainland (see Antonaccio 1994; Coldstream 1976), is not common here.
+The practice of hero or ancestor cult, much discussed on the mainland (see @Antonaccio1994; @Coldstream1976), is not common here.
 Instead, we see several Late Minoan larnakes (and a few other vessels) redeposited in tombs and pits in the KNC and Fortetsa from the PGB period onward.

### 2019-03-18 analysis/paper/paper.Rmd
space

@@
 Instead, we see several Late Minoan larnakes (and a few other vessels) redeposited in tombs and pits in the KNC and Fortetsa from the PGB period onward.
-A thorough analysis of these artefacts has been provided by Crowe (2016), who argues that the larnakes were robbed from tombs in the Knossos area, perhaps with a particular preference for painted specimens where available.
+A thorough analysis of these artefacts has been provided by Crowe [-@Crowe2016], who argues that the larnakes were robbed from tombs in the Knossos area, perhaps with a particular preference for painted specimens where available.
 A particularly evocative case comres from tomb 107, where parts of an elaborate figured LM larnax have been found along with a PGB vessel by the so-called Tree Painter.
 Both feature women in the company of birds and spiralling trees, and it seems certain that the former inspired the design of the latter.
-The PGB vessel also seems to draw on Oriental motifs, with the polos crown worn by the goddess having Near Eastern origins (Coldstream 1984a, 94–101).
+The PGB vessel also seems to draw on Oriental motifs, with the polos crown worn by the goddess having Near Eastern origins [@Coldstream1984b, 94-101].
 The magpie-like amalgamation of these styles and motifs seem to exemplify the particular cultural openness, fluidity and competition suggested to characterise the PGB period.
@@
 A few authors have applied more quantative approaches, and these point toward their potential utility.
-Antoniadis (2012, 172-6), in his study of Oriental imports, divided the EIA tombs of Knossos into three groups, those with fewer than 10 pots, fewer than 50 pots, and more than 50 pots, and found that, in each successive case, the proportions of tombs containing imports and imitations increased.
+Antoniadis [-@Antoniadis2012, 172-6], in his study of Oriental imports, divided the EIA tombs of Knossos into three groups, those with fewer than 10 pots, fewer than 50 pots, and more than 50 pots, and found that, in each successive case, the proportions of tombs containing imports and imitations increased.
 The implication is that a hierarchy existed, with differing degrees of control over the accumulation of prestigious goods.
@@
 
-Cluster analyses have been employed by both Cavanagh (1996, 653-7) and Antoniadis (2012, 193-7), the most sophisticated statistical techniques so far applied to the present material.
+Cluster analyses have been employed by both Cavanagh [-@Coldstream1996, 653-7] and Antoniadis [-@Antoniadis2012, 193-7], the most sophisticated statistical techniques so far applied to the present material.
 Cavanagh’s analysis of tomb architecture has been mentioned, while Antoniadis’ more holistic approach corroborates quantitatively some of the intuitive correspondences noted between specific tombs in previous studies.

### 2019-03-18 analysis/paper/paper.Rmd
mini changes
reconnected
data added and code begun

@@
 
-```{r, setup, echo = FALSE}
-knitr::opts_chunk$set(
+```{r setup}
+knitr::opts_knit$set(
   collapse = TRUE,
@@
 ```{r flowers, fig.cap="Here are some flower measurements"}
+
 library(ggplot2)
-ggplot(iris,aes(Sepal.Width,Sepal.Length))+geom_point()
+
+pots<-read.csv("../data/raw_data/KnossosPots.csv",header=T)
+
+ggplot(pots,aes(pots$ave_date))+geom_histogram()+xlim(1100,600)
 ```

### 2019-03-18 analysis/paper/paper.Rmd
first figs

@@
 
-pots<-read.csv("../data/raw_data/KnossosPots.csv",header=T)
+allpots<-read.csv("../data/raw_data/KnossosPots.csv",header=T)
+pots<-subset(allpots,cemetery=="NC"|cemetery=="FOR"|cemetery=="TK")
+potsnop<-subset(allpots,tomb!="P")
+str(pots)
 
-ggplot(pots,aes(pots$ave_date))+geom_histogram()+xlim(1100,600)
+allpots$tomb_period<- factor(allpots$tomb_period, levels = c("SM", "EPG", "MPG", "PG","LPG","PGB","EG","MG","G","LG",
+                                                       "EO","O","LO",""))
+
+ggplot(pots,aes(pots$ave_date,fill=tomb_period,color=tomb_period))+geom_histogram()+xlim(1100,600)
+ggplot(tombp,aes(tombp$ave_date,fill=period,color=period))+geom_histogram()+xlim(1100,600)
 ```

### 2019-03-19 analysis/paper/paper.Rmd
Writing and figs

--- 2019-03-18 analysis/paper/paper.Rmd
@@
 
-```{r, setup, echo = FALSE}
+```{r setup, include=FALSE}
 knitr::opts_chunk$set(
@@
   message = FALSE,
-  echo = TRUE,
+  echo = FALSE,
+  fig.retina = 2,
   comment = "#>",
@@
 )
-
-# Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
 ```
@@
 
-```{r flowers, fig.cap="Here are some flower measurements"}
+# Methods
+
+## Databases
+
+The cornerstone of the present study is a database (or, rather, three linked databases) based on the excavation reports for the Fortetsa and Knossos North cemeteries [@Brock1957; @Coldstream1996a].
+The first database includes an entry for each individual ceramic vessel, by far the largest portion of all known finds.
+Each entry includes: the tomb from which the vessel came; the vessel shape; whether it was an open or closed vessel; its main dimensions; its ceramic period and corresponding absolute dating measures (see below).
+In addition, all imports, Oriental imitations, and antique (that is, Bronze Age) artefacts were accordingly categorised, using the studies of Antoniadis (2012), Jones (2000), and Crowe (2016) to supplement omissions in the original excavation reports.
+In total, this database comprised just over 4540 individual artefacts.
+
+The second database is similar, but with each entry corresponding to a find other than a ceramic vessel.
+The categories remained the same as in the pottery database, but with the addition of a ‘material’ variable, as of course these finds were not all made of the same substance.
+This database included some 1620 individual artefacts.
+The final database concerns the tombs themselves.
+Each of the 134 tombs was accorded its own entry, with variables including dating (both in relative and absolute terms), surviving dimensions, and the known and estimated number of interments.
+Each entry also included a tally of the objects recorded in the other two databases.
+These tallies were arranged by shape for the pottery, and by material and category (e.g.
+weapon, adornment, seal, funerary paraphernalia) for the other objects.
+Finally, the total number of imports, imitations and antiques in each tomb was calculated.
+
+Taken together, these databases constitute a detailed record of all the finds from the Fortetsa and KNC excavations, and offer the opportunity for carrying out a range of robust quantitative analyses.
+It must be conceded that, with many of the tombs evidently looted, dug into or destroyed over the centuries, this cannot be a complete picture of the tombs’ original contents.
+If we assume, though, that natural or accidental destruction was relatively indiscriminate, and that, even in cases of looting, certain object classes (particularly cinerary urns) were seldom robbed, then the data are by no means rendered useless.
+As Snodgrass (1996) notes, the situation means we should be extremely cautious about making negative arguments – that is, conclusions drawn about the lack of features in certain tombs – but may be somewhat more hopeful in the forming of positive ones.
+
+## Estimating interments
+
+A first important step in preparing the data for analysis was the estimation of the number of individuals buried in each tomb.
+The method chosen for this estimation broadly follows that of Cavanagh [-@Cavanagh1996, 659-660], who took the number of pithoi (the most common urn type) and the number of pithoi, amphorae and kraters (which were in some cases used as urns) as middle and upper estimates of the number of interments, with all securely identified interments (in the form of ashes or osteological material) being the lower limit.
+In the present study, the mean of these three estimates was taken, to provide a single aggregate measure for the purposes of analysis (where no interments were known, this variable was still assigned a value of "1", for all tombs must have contained at least one burial).
+For analysis at the level of individual finds, however, pithoi alone were taken as the best proxy (Figure \@ref(fig:burialestimate)).
+For the bulk of the analyses, any tombs lacking identified remains or a single vessel suitable for use as an urn were not considered, thus excluding the most heavily damaged and looted tombs.
+
+```{r burialestimate, fig.cap="Relationship between known interments, counts of pithoi, and counts of pithoi, amphorae and kraters.
+The straight line represents the mean estimate based on these three values."}
+
 library(ggplot2)
-ggplot(iris,aes(Sepal.Width,Sepal.Length))+geom_point()
+library(RColorBrewer)
+
+tombs<-read.csv("../data/raw_data/KnossosTombs.csv",header=T)
+attach(tombs)
+
+p<-ggplot(tombs)+geom_point(aes(est_burials,all_urns,color="All urns"))+
+  geom_point(aes(est_burials,total_burials,color="Known interments"))+
+    geom_point(aes(est_burials,pithoi,colour="Pithoi"))+
+      geom_line(aes(est_burials,est_burials),alpha=0.4)+
+       xlim(0,60)+ylim(0,60)+theme_minimal()+theme(legend.title=element_blank())+
+        labs(x="Mean Burial Estimate", y="Interment/Vessel Count")+
+          scale_color_brewer(palette = "Set2")
+
+p<-p+coord_fixed()
+plot(p)
+
+cor.test(all_urns,est_burials, method = "pearson")
+
 ```
@@
 ```{r a-table}
+
 knitr::kable(head(iris),
@@
 We can also make a table (Table \@ref(tab:a-table))
-
-# Methods
 

### 2019-03-25 analysis/paper/paper.Rmd
Writing methods

@@
 abstract: |
+  Three principal interests converge in this paper: reorienting the focus of Early Iron Age Cretan archaeology; the application of computational, and quantitative methods to supplement traditional modes of inference; and open-access data and reproducible research.
 The Early Iron Age cemeteries of Knossos, Crete, represent the richest source of archaeological material at the site for the whole period between the end of the Bronze Age and the emergence of the Archaic *polis*.
 However, despite well-published excavation reports on the two largest burial grounds, the Fortetsa and Knossos North cemeteries, and a number of studies dealing with various artefact classes and individual tombs therein, no substantial comparative, quantitative analysis of these two cemeteries has yet been undertaken.
-This thesis provides such an analysis, examining both the synchronic and diachronic variation among the tombs in these two cemeteries.
-Through the application of a range of quantitative techniques, this study isolates a pattern of increased rates of burial among a few select tombs, mostly founded in the 9th century BC.
-Through comparing the compositions of the different tomb assemblages, it is suggested that these tombs represent certain extended-lineage groups, perhaps akin to the ‘tribes’ or ‘clans’ of later textual sources, and that their exceptional burial numbers and rates reflect social strategies developed in the contested ideological context of the nascent Knossian polis.
-This thesis demonstrates the potential for quantitative analyses to contribute to nuanced, context-specific theories of social practice, and advocates their wider application to the archaeological record of the Cretan Early Iron Age.
+This paper provides such an analysis, examining both the synchronic and diachronic variation among the tombs in these two cemeteries.
+Through the application of a range of quantitative techniques, this study isolates a pattern of increased rates of burial among a few select tombs, and further considers their characteristics.
+This paper demonstrates the potential for quantitative analyses to contribute to nuanced, context-specific theories of social practice, and advocates their wider application to the archaeological record of the Cretan Early Iron Age.
+It also serves as a demonstration of maximal research reproducibility, with the full datasets and code underpinning the analysis made freely available to download and use.
 keywords: |
-  Early Iron Age; Knossos; Burial; Quantitative;
+  Early Iron Age; Knossos; Burial; Quantitative; Open-Access; Reproducibility
 
@@
 
-For one, this is not the Knossos of that illustrous, palatial period, but rather the town as it persisted and evolved during the subsequent Early Iron Age (EIA).
+For one, this is not the Knossos of that illustrous, palatial period, but rather the town as it persisted and evolved during the subsequent Early Iron Age (EIA), about which much less is known.
 Research into this once maligned period has grown apace in the last few decades.
-The reframing of the the twelth-seventh centuries as the gestational period of the later classical Greek world [first seen in the works of @Coldstream1977; @Desborough1972; @Snodgrass1971] came first, while, more recently, scholars have refuted the image of a society bereft of its former palatial glory.
+What began as a reframing of the the twelth-seventh centuries as the gestational period of the later classical Greek world [first seen in the works of @Coldstream1977; @Desborough1972; @Snodgrass1971] has culimated in the complete rejection by many scholars of the image of a society bereft of its former palatial glory.
 Certainly, few would any longer defend the appelation or interpretive bagguage of the ‘Dark Ages’ [on the origins of which, see @Kotsonas2016; @Morris1997].Nonetheless, our knowledge of many EIA settlements (both new foundations, and those with BA ancestry like Knossos) remains often underdeveloped in comparison with those of the second millennium BC.
@@
 Secondly, and welcome though they are, a number of recent publications on the socio-economic [@Wallace2010a] and settlement [@Nowicki2000] changes which accompanied the BA-EIA transition have directed scholarly attention toward a number of elevated, defensible settlements supposedly founded in response to the turbulent contemporary political climate.
-This research agenda is likewise reflected in survey work, which has focussed extensively around the Bay of Mirabello, and the isthmus of Ierapetra.
+This research agenda is likewise reflected in survey work, which has focussed extensively around the Bay of Mirabello, and the Isthmus of Ierapetra.
 This has, unfortunately, engendered a certain disregard for continuing occupation at several lowland centres in the central region of the island, including Knossos and Phaistos.
-There remains, therefore, a need for studies addressing the social dynamics at these sites, with a long-term view to comparative appraisal of the divergent trajectories of communities across the island in the EIA [@Whitley2011, 667–8]
+There remains, therefore, a need for studies addressing the social dynamics at these sites, with a long-term view to comparative appraisal of the divergent trajectories of communities across the island in the EIA [@Whitley2011, 667–8].
 
@@
 Such work encourages the use of large (and often pre-existing) datasets, and the sharing of both data and methods to facilitate reproducible, collaborative, and innovative research.
-The cemeteries of Knossos offer an opportunity to introduce such approaches into EIA research, where they have yet to make much of an impact.
-The Fortesta and North Cemetery complexes are both well-published [@Brock1957; @Coldstream1996a] and have been considered variously in relation to their oriental imports [@Antoniadis2012], antiques [@Crowe2016], and religious context [@Coldstream1984b].
-But a lack of quantitative analyses, or systematic comparisons between different cemeteries, has left us with many intuited trends, some clearly exceptional tombs, and a general impression of un-patterned multiplicity in mortuary rituals.
+The cemeteries of Knossos offer an opportunity to introduce such approaches into EIA research, where they have yet to make much of an impact [see @Kotsonas2011].
+The Fortesta and North Cemetery complexes are both well-published [@Brock1957; @Coldstream1996a] and have been considered variously in relation to their "oriental"" imports [@Antoniadis2012], antiques [@Crowe2016], and religious context [@Coldstream1984b].
+But a lack of quantitative analyses, or systematic comparisons between different cemeteries, has left us with many intuited trends, some clearly exceptional tombs, but a general impression of un-patterned multiplicity.
+
 I suggest that the application of a range of quantitative and statistical methods may provide new insight into the social behaviours underlying the mortuary assemblages.
-
-A recent paper by Kotsonas [-@Kotsonas2011]  has stressed the importance of quantification in analyses of mortuary material.
-Using data drawn from the EIA Knossian and Eleuthernian cemeteries, Kotsonas demonstrates the capacity of even quite basic methods to problematise commonplace assumptions surroudning mortuary behaviour.
-In this spirit, I here adopt a broader, more comparative, and more quantitative approach to the Knossian material than has previously been employed.
-With a dataset including information on every find and tomb recorded in the Fortetsa and Knossos North Cemetery publications [@Brock1957; @Coldstream1996a, the following analysis examines both the synchronic and diachronic variation among the tombs of both cemeteries.
+With a dataset including information on every find and tomb recorded in the Fortetsa and Knossos North Cemetery publications [@Brock1957; @Coldstream1996a], the following analysis examines both the synchronic and diachronic variation among the tombs of both cemeteries.
 In doing so, I aim to demonstrate the potential for quantitative methods, firstly, to permit more reliable identification, and more nuanced characterisation, of significant patterning in the archaeological material; secondly, to facilitate a more systematic appraisal of the cemeteries’ largest tombs, and set their development in a richer socio-economic context; and, finally, to offer not just generalities, but evidence by which to consider, in tangible and context-specific ways, the motivations and identities of the burial groups themselves.
@@
 
+## Reproducibility and data-sharing
+
+An ever-growing number of archaeologists have begun to advocate the utility, indeed the necessity, of making available not just the data underpinning their professional output, but the full methodologies by which their analyses were conducted, and in formats which facilitate reproducibilty and further investigation.
+It is felt that, with the opportunities afforded by current technologies (digital databases, statistical software packages, online repositories etc.), there remain ever fewer obstacles - and excuses - preventing the full and free sharing of both data and workflows.
+In Marwick's words [-@Marwick2017, 445], "[t]he technical problems are largely solved; the challenge now is to change the norms of the discipline to make high reproducibility a canonical attribute of high-quality scholarly work".
+
+This is not the place for a detailed investigation into the lack of reproducible research in archaeology (for this, see ...).
+Suffice it to say the problem exists, though, granted, gradual progress is being spearheaded by some.
+This article is, among other things, an attempt to bring these principles of maximal reproducibility to the archaeology of Early Iron Age Crete.
+Within this field, catalogues and excavation reports abound and, though many of these may be forgiven based on their dates of publication, the lack of readily utilisable digital data from a number of recent articles and monographs, indeed, the failure to even present the data such that they could be manually digitsed, is less pardonable.
+Preparation of databases such as those of the present study can (and did) involve prolongued trawling of printed tables (often without consistent formatting, coding or presentation which would enable easy digitisation), cross-referencing artefacts listed inconsistently in different contexts, or else grappling with catalogues which, through their unexplained lack of certain critical information (e.g.
+coordinates in site catalogues), offer little research potential beyond their descriptive face-value.
+
+There are, of course, complications to the sharing of data and methods.
+There are real and valid reservations around issues of copyright, the protection of sites from looting, the possibility of being scooped with one's own data, or simply the significant investment of time demanded by the preparation of open-access research.
+But I would argue that the potential gains outweigh these concerns, and that research on the EIA would benefit greatly from the greater accessability of datasets and workflows.
+
+In this vein, the present article has been composed making use of Ben Marwick's "rrtools", an open-source package for the statistical software R Studio, which provides the author with the tools to create a self-contained "compendium", which can be freely downloaded, and from which anyone can accurately reproduce the entire workflow of the present analysis.
+This analysis was conducted in R, and the full code, as well as the raw and derived datasets, complete bibliography (in BibTex format, for export to any major reference manager), figures, and article as originally submitted to this journal are included in the package.
+It can be downloaded as a repository from GitHub, at the following address: https://github.com/DCPollard94/knossoscemeteries, and the raw datasets have been stored with the online repository Figshare, here:
+
+Though not visible in the present paper, every figure and table herein was produced directly, through written code, from the raw data using R.
+The full code can be viewed and followed by downloading the repository from GitHub, and all the data are made available through an MIT license, which permits unlimited use, ammendation, and dissemination thereof, yet frees the present author from any responsibilty for that use, or for any perceived errors or inaccuracies in the data.
+It is sincerely hoped that others may find interest in, and themselves further explore and expand upon the research presented here, by downloading this compedium.
+
 ## Estimating interments
@@
 
+#The appropriate packages are installed.
+Extrafont allows the use of non-standard fonts in R plots, ggplot2 is a powerful figure and graph plotting tool, and RColorBrewer includes a number of colour pallettes which can be applied to plots.
+The colour scheme in this paper is chosen from those that are suitable for colour-blind readers.
+The font Garmond is installed and stored.
+
+library(extrafont)
 library(ggplot2)
 library(RColorBrewer)
+font_import(paths="../paper/font",pattern="GARA")
+loadfonts(device="win")
+
+#The data from the tomb database are added, and 'attached' so that the variables therein need no longer be referred to in their long format (i.e. 'tombs$total_pottery' becomes simply 'total_pottery')
 
@@
 attach(tombs)
+
+#Figure 1 is plotted, with axis limits set (excluding tomb P, resulting in the warning message delivered) and the style of the plot adjusted (including font and text size, the appearance of the gridlines etc.)
 
@@
       geom_line(aes(est_burials,est_burials),alpha=0.4)+
-       xlim(0,60)+ylim(0,60)+theme_minimal()+theme(legend.title=element_blank())+
+       xlim(0,60)+ylim(0,60)+theme_minimal()+theme(legend.title=element_blank(),
+        text=element_text(family="Garamond", size=12))+
         labs(x="Mean Burial Estimate", y="Interment/Vessel Count")+
           scale_color_brewer(palette = "Set2")
+
+#The coordinates of the plot are adjusted to render the axis as equal, and the corresponding gridlines as squares, and then the figure is plotted.
 
@@
 
-cor.test(all_urns,est_burials, method = "pearson")
+cor_burial_proxy<-cor.test(all_urns,est_burials, method = "pearson")
 

### 2019-03-29 analysis/paper/paper.Rmd
writing edits

@@
 In the present study, the mean of these three estimates was taken, to provide a single aggregate measure for the purposes of analysis (where no interments were known, this variable was still assigned a value of "1", for all tombs must have contained at least one burial).
-For analysis at the level of individual finds, however, pithoi alone were taken as the best proxy (Figure \@ref(fig:burialestimate)).
-For the bulk of the analyses, any tombs lacking identified remains or a single vessel suitable for use as an urn were not considered, thus excluding the most heavily damaged and looted tombs.
 
@@
 
+library(reshape)
+library(data.table)
 library(extrafont)
@@
 library(RColorBrewer)
+library(grid)
+library(dplyr)
 font_import(paths="../paper/font",pattern="GARA")
@@
 tombs<-read.csv("../data/raw_data/KnossosTombs.csv",header=T)
-attach(tombs)
+tombs<-subset(tombs,tombs$cem_condensed=="NC"|tombs$cem_condensed=="FOR")
 
@@
       geom_line(aes(est_burials,est_burials),alpha=0.4)+
-       xlim(0,60)+ylim(0,60)+theme_minimal()+theme(legend.title=element_blank(),
+       xlim(0,60)+ylim(0,60)+theme_minimal()+
+        theme(legend.title=element_blank(),
           text=element_text(family="Garamond", size=12))+
@@
 
-cor_burial_proxy<-cor.test(all_urns,est_burials, method = "pearson")
+cor_burial_proxy<-cor(tombs$all_urns,tombs$est_burials, method = "pearson")
 
@@
 
-The end
+For analysis at the level of individual finds, however, pithoi alone were taken as the best proxy, as they correlated most closely with the mean estimate, producing a pearson correlation coefficient of `r cor_burial_proxy` (Figure \@ref(fig:burialestimate)).
+For the bulk of the analyses, any tombs lacking identified remains or a single vessel suitable for use as an urn were not considered, thus excluding the most heavily damaged and looted tombs.
 
-```{r a-table}
+```{r aorist_all_cems}
 
-knitr::kable(head(iris),
-             caption="A table of flowers")
+aorist_all_tombs <- colSums(main_tombs[157:206])
+aorist_all_tombs<-data.frame(aorist_all_tombs)
+
+colnames(aorist_all_tombs)<-c("aoristic_density")
+dates<-seq(from = 610, to = 1100, by = 10)
+dates<-rev(dates)
+row.names(aorist_all_tombs)<-dates
+aorist_all_tombs<-data.frame(aorist_all_tombs)
+setDT(aorist_all_tombs, keep.rownames = "decade")
+aorist_all_tombs$decade<-as.numeric(aorist_all_tombs$decade)
+
+ggplot(aorist_all_tombs,aes(decade,aoristic_density))+
+  geom_col()+xlim(1100,630)+theme_minimal()+
+        theme(legend.title=element_blank(),
+          text=element_text(family="Garamond", size=12))+
+              labs(x="Date (Years BC)", y="Aoristic Density")+
+                 scale_fill_brewer(palette = "Set2")
+
+```
+
+```{r aorist_uncertainty}
+
+pots<-read.csv("../data/raw_data/KnossosPots.csv",header=T)
+pots<-subset(pots,pots$cemetery=="FOR"|pots$cemetery=="TK"|pots$cemetery=="NC")
+
+aorist_pot_values<-pots[24:73]
+aorist_pot_matrix<-data.matrix(aorist_pot_values)
+aorist_pot_matrix[aorist_pot_matrix == 0] <- NA
+aorist_pot_medians<-apply(aorist_pot_matrix,2,median,na.rm=TRUE)
+
+aorist_pot_medians<-data.frame(aorist_pot_medians)
+colnames(aorist_pot_medians)<-c("aoristic_median")
+row.names(aorist_pot_medians)<-dates
+aorist_pot_medians<-data.frame(aorist_pot_medians)
+setDT(aorist_pot_medians, keep.rownames = "decade")
+aorist_pot_medians$decade<-as.numeric(aorist_pot_medians$decade)
+
+View(aorist_pot_medians)
+
+ggplot(NULL)+geom_col(data=aorist_pot_medians,aes(decade,aoristic_median))+
+    xlim(1100,630)+theme_minimal()+
+      theme(legend.title=element_blank(),
+        text=element_text(family="Garamond", size=12))+
+            labs(x="Date (Years BC)", y="Aoristic Median")+
+               scale_fill_brewer(palette = "Set2")
+
+ggplot(main_tombs,aes(total_finds))+geom_histogram()+xlim(0,200)
+
+aorist_tomb_values<-tombs[c(1,157:206)]
+aorist_tomb_values[aorist_tomb_values == 0] <- NA
+
+aorist_tomb_values<-data.frame(aorist_tomb_values)
+colnames(aorist_tomb_values)[2:51]<-dates
+
+aorist_melt<-melt(aorist_tomb_values)
+aorist_melt$variable<-as.numeric(as.character(aorist_melt$variable))
+
+aorist_melt$period<-ifelse(aorist_melt$variable<=1100 & aorist_melt$variable>=980,"SM",
+		ifelse(aorist_melt$variable<=970 & aorist_melt$variable>=930,"EPG",
+		ifelse(aorist_melt$variable<=920 & aorist_melt$variable>=880,"MPG",
+		ifelse(aorist_melt$variable<=870 & aorist_melt$variable>=850,"LPG",
+		ifelse(aorist_melt$variable<=840 & aorist_melt$variable>=820,"PGB",
+	  ifelse(aorist_melt$variable<=810 & aorist_melt$variable>=800,"EG",
+	  ifelse(aorist_melt$variable<=790 & aorist_melt$variable>=750,"MG",
+	  ifelse(aorist_melt$variable<=740 & aorist_melt$variable>=720,"LG",
+	  ifelse(aorist_melt$variable<=710 & aorist_melt$variable>=680,"EO",
+	  ifelse(aorist_melt$variable<=670 & aorist_melt$variable>=640,"LO",
+	  ifelse(aorist_melt$variable<=630 & aorist_melt$variable>=600,"A",""
+		)))))))))))
+
+decade_ranges<-seq(from = 600, to = 1100, by = 10)
+decade_ranges<-rev(decade_ranges)
+decade_ranges<-as.factor(decade_ranges)
+aorist_melt$variable<-as.factor(aorist_melt$variable)
+aorist_melt$variable<-factor(aorist_melt$variable, levels=(decade_ranges))
+aorist_melt$period<-factor(aorist_melt$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
+
+ggplot(aorist_melt,aes(variable,value, fill=period))+geom_boxplot(varwidth=TRUE)+theme_bw()+
+      theme(legend.title=element_blank(),
+        text=element_text(family="Garamond", size=12),
+        axis.text.x = element_text(angle = 70, hjust = 1),
+        panel.grid.major.x = element_blank())+
+            labs(x="Date (Years BC)", y="Aoristic Sums per Tomb")+
+               scale_fill_brewer(palette = "Set3")+
+                scale_y_continuous(expand = c(0, 0),limits=c(0,20))
+
 ```
@@
 We can also make a table (Table \@ref(tab:a-table))
+
+```{r aoristknc}
+KNC<-subset(tombs,tombs$cem_condensed=="NC")
+
+aorist_KNC <- colSums(KNC[157:206])
+aorist_KNC<-data.frame(aorist_KNC)
+
+colnames(aorist_KNC)<-c("aoristic_density")
+row.names(aorist_KNC)<-dates
+aorist_KNC<-data.frame(aorist_KNC)
+setDT(aorist_KNC, keep.rownames = "decade")
+aorist_KNC$decade<-as.numeric(aorist_KNC$decade)
+
+FOR<-subset(tombs,tombs$cem_condensed=="FOR")
+
+aorist_FOR <-colSums(FOR[157:206])
+aorist_FOR<-data.frame(aorist_FOR)
+
+colnames(aorist_FOR)<-c("aoristic_density")
+row.names(aorist_FOR)<-dates
+aorist_FOR<-data.frame(aorist_FOR)
+setDT(aorist_FOR, keep.rownames = "decade")
+aorist_FOR$decade<-as.numeric(aorist_FOR$decade)
+
+ggplot(NULL)+geom_col(data=aorist_all_tombs,aes(decade,aoristic_density,fill="All tombs"))+
+  geom_col(data=aorist_KNC,aes(decade,aoristic_density,fill="KNC"))+
+  geom_col(data=aorist_FOR,aes(decade,aoristic_density, fill="Fortetsa"))+
+      xlim(1100,630)+theme_minimal()+
+        theme(legend.title=element_blank(),
+          text=element_text(family="Garamond", size=12))+
+              labs(x="Date (Years BC)", y="Aoristic Density")+
+                 scale_fill_brewer(palette = "Set2")
+
+```
+
+```{r aorist_burials}
+
+pithoi<-subset(pots,pots$shape=="Pithos")
+all_urns<-subset(pots,pots$shape=="Pithos"|pots$shape=="Amphora"|pots$shape=="Krater")
+
+aorist_pithoi<-colSums(pithoi[24:73])
+aorist_pithoi<-data.frame(aorist_pithoi)
+
+colnames(aorist_pithoi)<-c("aoristic_density")
+row.names(aorist_pithoi)<-dates
+aorist_pithoi<-data.frame(aorist_pithoi)
+setDT(aorist_pithoi, keep.rownames = "decade")
+aorist_pithoi$decade<-as.numeric(aorist_pithoi$decade)
+
+aorist_all_urns<-colSums(all_urns[24:73])
+aorist_all_urns<-data.frame(aorist_all_urns)
+
+colnames(aorist_all_urns)<-c("aoristic_density")
+row.names(aorist_all_urns)<-dates
+aorist_all_urns<-data.frame(aorist_all_urns)
+setDT(aorist_all_urns, keep.rownames = "decade")
+aorist_all_urns$decade<-as.numeric(aorist_all_urns$decade)
+
+ggplot(NULL)+geom_col(data=aorist_all_urns,aes(decade,aoristic_density,fill="All urns"))+
+  geom_col(data=aorist_pithoi,aes(decade,aoristic_density,fill="Pithoi"))+
+      xlim(1100,630)+theme_minimal()+
+        theme(legend.title=element_blank(),
+          text=element_text(family="Garamond", size=12))+
+              labs(x="Date (Years BC)", y="Aoristic Density")+
+                 scale_fill_brewer(palette = "Set2")
+
+```
+
+```{r aorist_tomb_adjust}
+
+no_tombs<-colSums(main_tombs[157:206]!=0)
+no_tombs<-data.frame(no_tombs)
+colnames(no_tombs)<-c("no_tombs")
+row.names(no_tombs)<-dates
+no_tombs<-data.frame(no_tombs)
+setDT(no_tombs, keep.rownames = "decade")
+no_tombs$decade<-as.numeric(no_tombs$decade)
+
+aorist_all_tombs1<-aorist_all_tombs
+aorist_by_tomb<-merge(aorist_all_tombs1,no_tombs)
+
+aorist_by_tomb$aorist_by_tomb<-aorist_by_tomb$aoristic_density/aorist_by_tomb$no_tombs
+
+ggplot(aorist_by_tomb,aes(decade,aorist_by_tomb))+geom_col()+xlim(1100,600)
+
+#Now for the 9th century tombs
+
+tombs_9th<-subset(tombs,tombs$century_built=="9th")
+no_tombs_9th<-colSums(tombs_9th[157:206]!=0)
+
+no_tombs_9th<-data.frame(no_tombs_9th)
+colnames(no_tombs_9th)<-c("no_tombs")
+row.names(no_tombs_9th)<-dates
+no_tombs_9th<-data.frame(no_tombs_9th)
+setDT(no_tombs_9th, keep.rownames = "decade")
+no_tombs_9th$decade<-as.numeric(no_tombs_9th$decade)
+
+aorist_tombs_9th<-colSums(tombs_9th[157:206])
+aorist_tombs_9th<-data.frame(aorist_tombs_9th)
+colnames(aorist_tombs_9th)<-c("aoristic_density")
+row.names(aorist_tombs_9th)<-dates
+aorist_tombs_9th<-data.frame(aorist_tombs_9th)
+setDT(aorist_tombs_9th, keep.rownames = "decade")
+aorist_tombs_9th$decade<-as.numeric(aorist_tombs_9th$decade)
+
+aorist_by_tomb_9th<-merge(aorist_tombs_9th,no_tombs_9th)
+aorist_by_tomb_9th$aorist_by_tomb<-aorist_by_tomb_9th$aoristic_density/aorist_by_tomb_9th$no_tombs
+
+ggplot()+geom_col(data=aorist_by_tomb_9th,aes(decade,aorist_by_tomb,fill="9th Century Tombs"))+
+  geom_col(data=aorist_by_tomb,aes(decade,aorist_by_tomb, fill="All Tombs"),alpha=0.7)+
+    xlim(1100,600)+theme_minimal()+
+        theme(legend.title=element_blank(),
+          text=element_text(family="Garamond", size=12))+
+              labs(x="Date (Years BC)", y="Aoristic Density/Number of Tombs Represented")+
+                 scale_fill_brewer(palette = "Set2")
+
+names(main_tombs)
+tombs_in_use<-colSums(main_tombs[5:14]!="")
+tombs_in_use<-data.frame(tombs_in_use)
+colnames(tombs_in_use)<-c("no_tombs")
+tombs_in_use<-data.frame(tombs_in_use)
+setDT(tombs_in_use, keep.rownames = "period")
+tombs_in_use$period<-factor(tombs_in_use$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
+
+ggplot(no_tombs,aes(decade,no_tombs, fill="Tombs in Use"))+geom_col()+theme_classic()+
+     theme(legend.title=element_blank(),
+          text=element_text(family="Garamond", size=12))+
+            xlim(1100,600)+
+              labs(x="Date (Years BC)", y="Number of Tombs in Use")+
+                 scale_fill_brewer(palette = "Set2")
+
+ggplot(tombs_in_use,aes(period,no_tombs, fill="Tombs in Use"))+geom_col()+theme_classic()+
+        theme(legend.title=element_blank(),
+          text=element_text(family="Garamond", size=12))+
+              labs(x="Period", y="Number of Tombs in Use")+
+                 scale_fill_brewer(palette = "Set2")
+
+ggplot(tombs,aes(start_period, fill="New Tombs"))+geom_bar()+theme_classic()+
+        theme(legend.title=element_blank(),
+          text=element_text(family="Garamond", size=12))+
+              labs(x="Period", y="Number of New Tombs")+
+                 scale_fill_brewer(palette = "Set2")
+
+tombs$start_period<-factor(tombs$start_period,levels=c("SM", "EPG", "MPG","PG", "LPG", "PGB", "EG", "MG","G", "LG", "EO","O", "LO"))
+
+```
+
+```{r rates of burials}
+
+tombs_ranked<-main_tombs[with(main_tombs, order(-main_tombs$total_pottery)), ]
+tombs_top_12<-head(tombs_ranked,12)
+
+no_tombs_top<-colSums(tombs_top_12[157:206]!=0)
+
+no_tombs_top<-data.frame(no_tombs_top)
+colnames(no_tombs_top)<-c("no_tombs")
+row.names(no_tombs_top)<-dates
+no_tombs_top<-data.frame(no_tombs_top)
+setDT(no_tombs_top, keep.rownames = "decade")
+no_tombs_top$decade<-as.numeric(no_tombs_top$decade)
+
+aorist_tombs_top<-colSums(tombs_top_12[157:206])
+aorist_tombs_top<-data.frame(aorist_tombs_top)
+colnames(aorist_tombs_top)<-c("aoristic_density")
+row.names(aorist_tombs_top)<-dates
+aorist_tombs_top<-data.frame(aorist_tombs_top)
+setDT(aorist_tombs_top, keep.rownames = "decade")
+aorist_tombs_top$decade<-as.numeric(aorist_tombs_top$decade)
+
+aorist_by_tomb_top<-merge(aorist_tombs_top,no_tombs_top)
+aorist_by_tomb_top$aorist_by_tomb<-aorist_by_tomb_top$aoristic_density/aorist_by_tomb_top$no_tombs
+
+ggplot()+geom_col(data=aorist_by_tomb_top,aes(decade,aorist_by_tomb,fill="12 Largest Tombs"))+
+  geom_col(data=aorist_by_tomb,aes(decade,aorist_by_tomb, fill="All Tombs"),alpha=0.7)+
+    xlim(1100,600)+theme_minimal()+
+        theme(legend.title=element_blank(),
+          text=element_text(family="Garamond", size=12))+
+              labs(x="Date (Years BC)", y="Aoristic Density/Number of Tombs Represented")+
+                 scale_fill_brewer(palette = "Set2")
+
+```
 
@@
 my_data <- readr::read_csv(here::here('analysis', 'data', 'raw_data', 'my_csv_file.csv'))
+
+knitr::kable(head(iris),
+             caption="A table of flowers")
 ```

### 2019-03-29 analysis/paper/paper.Rmd
gini estimates

@@
 
+library(ineq)
+library(ggrepel)
 library(reshape)
@@
                 scale_y_continuous(expand = c(0, 0),limits=c(0,20))
+
+                   #geom_label_repel(data=subset(aorist_melt,aorist_melt$value>8 & aorist_melt$variable=="710"),
+                      #aes(label=tomb), show.legend = FALSE,ylim=c(0,40))
+
+tombs_gini<-matrix(0, ncol = 0, nrow = 1)
+tombs_gini<-data.frame(tombs_gini)
+for(i in colnames(aorist_tomb_values[2:51])) {
+  tombs_gini[[paste0(i)]] <- ineq(aorist_tomb_vals_only[[i]],type="Gini")
+}
+
+View(tombs_gini)
+
+tombs_gini<-t(tombs_gini)
+tombs_gini<-data.frame(tombs_gini)
+
+colnames(tombs_gini)<-c("gini_coefficient")
+row.names(tombs_gini)<-dates
+tombs_gini<-data.frame(tombs_gini)
+setDT(tombs_gini, keep.rownames = "decade")
+tombs_gini$decade<-as.numeric(tombs_gini$decade)
+
+ggplot(tombs_gini,aes(decade,gini_coefficient))+geom_line()+
+      xlim(1100,600)+ylim(0.3,0.8)+theme_minimal()+
+        theme(legend.title=element_blank(),
+          text=element_text(family="Garamond", size=12))+
+              labs(x="Date (Years BC)", y="Gini Coefficient")+
+                 scale_fill_brewer(palette = "Set2")
+
+rects <- data.frame(xstart = c(1100,970,920,875,840,810,790,745,70,670,630), 
+                    xend = c(970,920,875,840,810,790,745,710,670,630,600), 
+                    col = c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
+rects$col<-factor(rects$col,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
+
+ggplot() + 
+  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf, fill = col), alpha = 0.4) +
+     geom_line(data = tombs_gini, aes(decade,gini_coefficient))+xlim(1100,600)+ylim(0.5,0.8)+
+      theme_bw()+theme(legend.title=element_blank(),
+          text=element_text(family="Garamond", size=12))+
+              labs(x="Date (Years BC)", y="Gini Coefficient")+
+                 scale_fill_brewer(palette = "Set3")+ scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
+                    scale_y_continuous(expand = c(0, 0),limits=c(0.5,0.8))
+
+plot(Lc(aorist_tomb_values$`800`))
 

### 2019-03-29 analysis/paper/paper.Rmd
image edits

@@
 
-aorist_all_tombs <- colSums(main_tombs[157:206])
+aorist_all_tombs <- colSums(tombs[157:206])
 aorist_all_tombs<-data.frame(aorist_all_tombs)
@@
 
-```{r aorist_uncertainty}
+```{r aorist_uncertainty, fig.dim = c(7, 4)}
 
@@
 
-ggplot(main_tombs,aes(total_finds))+geom_histogram()+xlim(0,200)
+ggplot(tombs,aes(total_finds))+geom_histogram()+xlim(0,200)
 
@@
 for(i in colnames(aorist_tomb_values[2:51])) {
-  tombs_gini[[paste0(i)]] <- ineq(aorist_tomb_vals_only[[i]],type="Gini")
+  tombs_gini[[paste0(i)]] <- ineq(aorist_tomb_values[[i]],type="Gini")
 }
@@
   geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf, fill = col), alpha = 0.4) +
-     geom_line(data = tombs_gini, aes(decade,gini_coefficient))+xlim(1100,600)+ylim(0.5,0.8)+
+     geom_line(data = tombs_gini, aes(decade,gini_coefficient))+
       theme_bw()+theme(legend.title=element_blank(),
@@
               labs(x="Date (Years BC)", y="Gini Coefficient")+
-                 scale_fill_brewer(palette = "Set3")+ scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
-                    scale_y_continuous(expand = c(0, 0),limits=c(0.5,0.8))
+                 scale_fill_brewer(palette = "Set3")+ coord_cartesian(
+                    xlim = c(1100, 600), ylim = c(0.4,0.8))+ scale_x_reverse(expand = c(0, 0),limits=c(1100,600))
 
@@
 
-no_tombs<-colSums(main_tombs[157:206]!=0)
+no_tombs<-colSums(tombs[157:206]!=0)
 no_tombs<-data.frame(no_tombs)
@@
 
-names(main_tombs)
-tombs_in_use<-colSums(main_tombs[5:14]!="")
+names(tombs)
+tombs_in_use<-colSums(tombs[5:14]!="")
 tombs_in_use<-data.frame(tombs_in_use)
@@
 
-tombs_ranked<-main_tombs[with(main_tombs, order(-main_tombs$total_pottery)), ]
+tombs_ranked<-tombs[with(tombs, order(-tombs$total_pottery)), ]
 tombs_top_12<-head(tombs_ranked,12)

### 2019-04-01 analysis/paper/paper.Rmd
Writing methods and analysis

@@
   comment = "#>",
-  fig.path = "../figures/"
+  fig.path = "../figures/",
+  dev="png"
 )
@@
 A first important step in preparing the data for analysis was the estimation of the number of individuals buried in each tomb.
-The method chosen for this estimation broadly follows that of Cavanagh [-@Cavanagh1996, 659-660], who took the number of pithoi (the most common urn type) and the number of pithoi, amphorae and kraters (which were in some cases used as urns) as middle and upper estimates of the number of interments, with all securely identified interments (in the form of ashes or osteological material) being the lower limit.
+The method chosen for this estimation broadly follows that of Cavanagh [-@Cavanagh1996, 659-660], who took the number of pithoi (the most common urn type) and the number of pithoi, amphorae and kraters (which were in some cases used as urns) as middle and upper estimates of the number of interments, with all securely identified interments (in the form of ashes or osteological material) being the lower limit (Figure \@ref(fig:burialestimate)).
 In the present study, the mean of these three estimates was taken, to provide a single aggregate measure for the purposes of analysis (where no interments were known, this variable was still assigned a value of "1", for all tombs must have contained at least one burial).
@@
 
+library(ggpubr)
+library(cowplot)
+library(gridExtra)
 library(ineq)
@@
 
-For analysis at the level of individual finds, however, pithoi alone were taken as the best proxy, as they correlated most closely with the mean estimate, producing a pearson correlation coefficient of `r cor_burial_proxy` (Figure \@ref(fig:burialestimate)).
-For the bulk of the analyses, any tombs lacking identified remains or a single vessel suitable for use as an urn were not considered, thus excluding the most heavily damaged and looted tombs.
+For most diachronic analyses, however, pithoi alone were taken as the best proxy for burial numbers, as they correlated most closely with the mean estimate, (Pearson's *r* = `r cor_burial_proxy`).
+In most cases, and often simply by virute of the methods employed, any tombs lacking identified remains or a single vessel suitable for use as an urn were not considered, thus excluding what are presumably among the most heavily damaged and looted tombs.
 
-```{r aorist_all_cems}
+##Aoristic approaches to dating
+
+A central concern of the present study was how to deal with the temporal uncertainty implicit within the assigning of individual vessels to stylistically-defined periods.
+A vessel dated to the Protogeometric B period could, on our best estimates of absolute chronology, date to any point between 840 and 810 BC.
+This situation results in major challenges to the visualisation and analysis of such data through mean which normally rely on event- and not range-based data inputs.
+A simple, if unsatisfying, solution is to use the midpoint of the range asigned to each vessel (i.e.
+825 for our PGB example).
+This method is used in a couple of instances in the present paper when the course-grained sorting of the data (into 50 year periods) renders the innacuracies of such a method less significant.
+
+```{r allpithoi, fig.cap="Counts of pithoi across both cemeteries through time.
+Vessel dates have here been based on the midpoint of the absolute date range accompanying their respective ceramic periods."}
+
+pots<-read.csv("../data/raw_data/KnossosPots.csv",header=T)
+pots<-subset(pots,pots$cemetery=="FOR"|pots$cemetery=="TK"|pots$cemetery=="NC")
+pithoi<-subset(pots,pots$shape=="Pithos")
+all_urns<-subset(pots,pots$shape=="Pithos"|pots$shape=="Amphora"|pots$shape=="Krater")
+
+ggplot(pots, aes(ave_date,fill="Pots"))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+      text=element_text(family="Garamond", size=12))+
+          labs(x="Date (Years BC)", y="Total Vessel Counts")+
+              scale_fill_brewer(palette = "Set3")+ 
+                theme(legend.position="none")+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,600))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))
+
+```
+
+But the artificial clustering this technique engenders around select absolute dates proves misleading when, for instance, the total number of pithoi deposited by date are plotted via a histogram (Figure \@ref(fig:allpithoi)); the apparent cessation of burial activity for 30 years in the 8th century, followed by the most intense episode of deposition yet seen, is simply a fiction generated by the processing of the data.
+
+A alternative way of dealing with temporal uncertainty, and the method employed extensively in this paper, is the use of so-called "aoristic analysis".
+Here, rather than assign a pot a single date, the entire timespan under investigation is divided into equal chunks (in this case, decades), and probability values summing to 1 are then assigned for each vessel, per chunk of time, based on its accordant date range.
+So, our PGB vessel, dateable between 840 and 810 BC would register a value of 0.33 for each of the three decades within which we it could theoretically have been produced.
+These values for individual vessels can can then simply be summed to produce tomb- or cemetery-level signatures.
+The greatest contribution of this method is that it incorporates temporal uncertainty into the size of the resulting values, that is, "events with tight temporal definition contribute more to the total probability over their range than do loosely defined events" (JOHNSON 2003 p2).
+
+```{r aoristallcems, fig.cap="Aoristic sums of all vessels across both cemeteries. This represents a probabilistically weighted visualisation of depositional activity through time."}
 
@@
 
-ggplot(aorist_all_tombs,aes(decade,aoristic_density))+
-  geom_col()+xlim(1100,630)+theme_minimal()+
-        theme(legend.title=element_blank(),
+ggplot(aorist_all_tombs,aes(decade,aoristic_density, fill="Aoristic Sum"))+
+  geom_col()+theme_bw()+theme(legend.title=element_blank(),
       text=element_text(family="Garamond", size=12))+
-              labs(x="Date (Years BC)", y="Aoristic Density")+
-                 scale_fill_brewer(palette = "Set2")
+          labs(x="Date (Years BC)", y="Aoristic Sum")+
+              scale_fill_brewer(palette = "Set3")+ 
+                theme(legend.position="none")+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,225))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))
 
@@
 
-```{r aorist_uncertainty, fig.dim = c(7, 4)}
+As seen in Figure \@ref(fig:aoristallcems), the artifical peaks and troughs of the simple averaging method are avoided, and different patterns become visible.
+Though these methods present their own limitations, which will be explored in the following sections, they are considered a more robust and productive approach to the temporal uncertainty of ceramic dating and, as such, are used for much of the present analysis.
 
-pots<-read.csv("../data/raw_data/KnossosPots.csv",header=T)
-pots<-subset(pots,pots$cemetery=="FOR"|pots$cemetery=="TK"|pots$cemetery=="NC")
+##Theory and method
 
-aorist_pot_values<-pots[24:73]
-aorist_pot_matrix<-data.matrix(aorist_pot_values)
-aorist_pot_matrix[aorist_pot_matrix == 0] <- NA
-aorist_pot_medians<-apply(aorist_pot_matrix,2,median,na.rm=TRUE)
+Before proceeding to the main discussion, it should be stressed that, though I hope to demonstrate the power of quantitative and statistical approaches to the mortuary record, data-points cannot replace description, and theory must not be absent in the use of such methods.
+Quantifications and statistical analyses should, ideally, be employed as tools for the development and exploration of theoretically-informed models of social practice.
+Throughout the following analysis, I aim to link the observed patterning in the material back to possible forms of human behaviour, which in turn guides the focus of subsequent analyses, and so on.
+It is hoped that the value of such an approach will become clear.
 
-aorist_pot_medians<-data.frame(aorist_pot_medians)
-colnames(aorist_pot_medians)<-c("aoristic_median")
-row.names(aorist_pot_medians)<-dates
-aorist_pot_medians<-data.frame(aorist_pot_medians)
-setDT(aorist_pot_medians, keep.rownames = "decade")
-aorist_pot_medians$decade<-as.numeric(aorist_pot_medians$decade)
+#Analysis
 
-View(aorist_pot_medians)
+##Introduction
+The great advantage of the large, quantitative dataset compiled for the present analysis is that it permits investigation of several nested levels of potentially significant patterning.
+From individual burials, to collective tombs, to the cemeteries as a whole, and between cemeteries in the landscape, we have the opportunity to study and compare both the chronological and contemporary variation across these various levels.
+The following analysis is divided broadly into these two distinct, though complementary, perspectives.
+I consider, firstly, how temporal variations in the material may relate to broader social dynamics and, secondly, how the form and content of the tombs and their assemblages can enrich the schematic picture derived from the chronological patterns.
+Taken together, these two approaches enable us to develop a nuanced account of the changing social landscape of EIA Knossos, drawing together issues of group identity, external relations, and the communication of wealth and status through funerary practice.
 
-ggplot(NULL)+geom_col(data=aorist_pot_medians,aes(decade,aoristic_median))+
-    xlim(1100,630)+theme_minimal()+
-      theme(legend.title=element_blank(),
-        text=element_text(family="Garamond", size=12))+
-            labs(x="Date (Years BC)", y="Aoristic Median")+
-               scale_fill_brewer(palette = "Set2")
+##Temporal variation in the Knossian Cemeteries
 
-ggplot(tombs,aes(total_finds))+geom_histogram()+xlim(0,200)
+An obvious place to begin is with the chronology of the burials themselves.
+The rate and number of burials being made through time provides the background against which to consider all other developments.
 
-aorist_tomb_values<-tombs[c(1,157:206)]
-aorist_tomb_values[aorist_tomb_values == 0] <- NA
+```{r aoristbycem, fig.cap="Aorist sums of pithoi (left) as a proxy for cremation burials, and all vessels (right).
+The sums are colour coded to show the values of the two cemeteries, and the total values.", fig.dim=c(8,4)}
+KNC<-subset(tombs,tombs$cem_condensed=="NC")
 
-aorist_tomb_values<-data.frame(aorist_tomb_values)
-colnames(aorist_tomb_values)[2:51]<-dates
+aorist_KNC_pithoi <- colSums(KNC[207:256])
+aorist_KNC_pithoi<-data.frame(aorist_KNC_pithoi)
 
-aorist_melt<-melt(aorist_tomb_values)
-aorist_melt$variable<-as.numeric(as.character(aorist_melt$variable))
+colnames(aorist_KNC_pithoi)<-c("aoristic_density")
+row.names(aorist_KNC_pithoi)<-dates
+aorist_KNC_pithoi<-data.frame(aorist_KNC_pithoi)
+setDT(aorist_KNC_pithoi, keep.rownames = "decade")
+aorist_KNC_pithoi$decade<-as.numeric(aorist_KNC_pithoi$decade)
 
-aorist_melt$period<-ifelse(aorist_melt$variable<=1100 & aorist_melt$variable>=980,"SM",
-		ifelse(aorist_melt$variable<=970 & aorist_melt$variable>=930,"EPG",
-		ifelse(aorist_melt$variable<=920 & aorist_melt$variable>=880,"MPG",
-		ifelse(aorist_melt$variable<=870 & aorist_melt$variable>=850,"LPG",
-		ifelse(aorist_melt$variable<=840 & aorist_melt$variable>=820,"PGB",
-	  ifelse(aorist_melt$variable<=810 & aorist_melt$variable>=800,"EG",
-	  ifelse(aorist_melt$variable<=790 & aorist_melt$variable>=750,"MG",
-	  ifelse(aorist_melt$variable<=740 & aorist_melt$variable>=720,"LG",
-	  ifelse(aorist_melt$variable<=710 & aorist_melt$variable>=680,"EO",
-	  ifelse(aorist_melt$variable<=670 & aorist_melt$variable>=640,"LO",
-	  ifelse(aorist_melt$variable<=630 & aorist_melt$variable>=600,"A",""
-		)))))))))))
+FOR<-subset(tombs,tombs$cem_condensed=="FOR")
 
-decade_ranges<-seq(from = 600, to = 1100, by = 10)
-decade_ranges<-rev(decade_ranges)
-decade_ranges<-as.factor(decade_ranges)
-aorist_melt$variable<-as.factor(aorist_melt$variable)
-aorist_melt$variable<-factor(aorist_melt$variable, levels=(decade_ranges))
-aorist_melt$period<-factor(aorist_melt$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
+aorist_FOR_pithoi <-colSums(FOR[207:256])
+aorist_FOR_pithoi<-data.frame(aorist_FOR_pithoi)
 
-ggplot(aorist_melt,aes(variable,value, fill=period))+geom_boxplot(varwidth=TRUE)+theme_bw()+
-      theme(legend.title=element_blank(),
-        text=element_text(family="Garamond", size=12),
-        axis.text.x = element_text(angle = 70, hjust = 1),
-        panel.grid.major.x = element_blank())+
-            labs(x="Date (Years BC)", y="Aoristic Sums per Tomb")+
-               scale_fill_brewer(palette = "Set3")+
-                scale_y_continuous(expand = c(0, 0),limits=c(0,20))
+colnames(aorist_FOR_pithoi)<-c("aoristic_density")
+row.names(aorist_FOR_pithoi)<-dates
+aorist_FOR_pithoi<-data.frame(aorist_FOR_pithoi)
+setDT(aorist_FOR_pithoi, keep.rownames = "decade")
+aorist_FOR_pithoi$decade<-as.numeric(aorist_FOR_pithoi$decade)
 
-                   #geom_label_repel(data=subset(aorist_melt,aorist_melt$value>8 & aorist_melt$variable=="710"),
-                      #aes(label=tomb), show.legend = FALSE,ylim=c(0,40))
+aorist_all_pithoi <- colSums(tombs[207:256])
+aorist_all_pithoi<-data.frame(aorist_all_pithoi)
 
-tombs_gini<-matrix(0, ncol = 0, nrow = 1)
-tombs_gini<-data.frame(tombs_gini)
-for(i in colnames(aorist_tomb_values[2:51])) {
-  tombs_gini[[paste0(i)]] <- ineq(aorist_tomb_values[[i]],type="Gini")
-}
+colnames(aorist_all_pithoi)<-c("aoristic_density")
 
-View(tombs_gini)
+row.names(aorist_all_pithoi)<-dates
+aorist_all_pithoi<-data.frame(aorist_all_pithoi)
+setDT(aorist_all_pithoi, keep.rownames = "decade")
+aorist_all_pithoi$decade<-as.numeric(aorist_all_pithoi$decade)
 
-tombs_gini<-t(tombs_gini)
-tombs_gini<-data.frame(tombs_gini)
-
-colnames(tombs_gini)<-c("gini_coefficient")
-row.names(tombs_gini)<-dates
-tombs_gini<-data.frame(tombs_gini)
-setDT(tombs_gini, keep.rownames = "decade")
-tombs_gini$decade<-as.numeric(tombs_gini$decade)
-
-ggplot(tombs_gini,aes(decade,gini_coefficient))+geom_line()+
-      xlim(1100,600)+ylim(0.3,0.8)+theme_minimal()+
-        theme(legend.title=element_blank(),
-          text=element_text(family="Garamond", size=12))+
-              labs(x="Date (Years BC)", y="Gini Coefficient")+
-                 scale_fill_brewer(palette = "Set2")
-
-rects <- data.frame(xstart = c(1100,970,920,875,840,810,790,745,70,670,630), 
-                    xend = c(970,920,875,840,810,790,745,710,670,630,600), 
-                    col = c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
-rects$col<-factor(rects$col,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
-
-ggplot() + 
-  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf, fill = col), alpha = 0.4) +
-     geom_line(data = tombs_gini, aes(decade,gini_coefficient))+
+aorist_pithoi_plot<-ggplot(NULL)+
+  geom_col(data=aorist_all_pithoi,aes(decade,aoristic_density,fill=" All tombs  "))+
+    geom_col(data=aorist_KNC_pithoi,aes(decade,aoristic_density,fill=" KNC  "))+
+      geom_col(data=aorist_FOR_pithoi,aes(decade,aoristic_density, fill=" Fortetsa  "))+
         theme_bw()+theme(legend.title=element_blank(),
           text=element_text(family="Garamond", size=12))+
-              labs(x="Date (Years BC)", y="Gini Coefficient")+
-                 scale_fill_brewer(palette = "Set3")+ coord_cartesian(
-                    xlim = c(1100, 600), ylim = c(0.4,0.8))+ scale_x_reverse(expand = c(0, 0),limits=c(1100,600))
+            labs(x="Date (Years BC)", y="Aoristic Sum (Pithoi)")+
+              scale_fill_brewer(palette = "Set2")+ 
+                  scale_y_continuous(expand=c(0,0),limits=c(0,45))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
 
-plot(Lc(aorist_tomb_values$`800`))
+aorist_KNC <- colSums(KNC[157:206])
+aorist_KNC<-data.frame(aorist_KNC)
+
+colnames(aorist_KNC)<-c("aoristic_density")
+row.names(aorist_KNC)<-dates
+aorist_KNC<-data.frame(aorist_KNC)
+setDT(aorist_KNC, keep.rownames = "decade")
+aorist_KNC$decade<-as.numeric(aorist_KNC$decade)
+
+FOR<-subset(tombs,tombs$cem_condensed=="FOR")
+
+aorist_FOR <-colSums(FOR[157:206])
+aorist_FOR<-data.frame(aorist_FOR)
+
+colnames(aorist_FOR)<-c("aoristic_density")
+row.names(aorist_FOR)<-dates
+aorist_FOR<-data.frame(aorist_FOR)
+setDT(aorist_FOR, keep.rownames = "decade")
+aorist_FOR$decade<-as.numeric(aorist_FOR$decade)
+
+aorist_pots_plot<-ggplot(NULL)+
+  geom_col(data=aorist_all_tombs,aes(decade,aoristic_density,fill=" All tombs  "))+
+    geom_col(data=aorist_KNC,aes(decade,aoristic_density,fill=" KNC  "))+
+      geom_col(data=aorist_FOR,aes(decade,aoristic_density, fill=" Fortetsa  "))+
+        theme_bw()+theme(legend.title=element_blank(),
+          text=element_text(family="Garamond", size=12))+
+            labs(x="Date (Years BC)", y="Aoristic Sum (All Vessels)")+
+              scale_fill_brewer(palette = "Set2")+ 
+                  scale_y_continuous(expand=c(0,0),limits=c(0,225))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+ggarrange(aorist_pithoi_plot,aorist_pots_plot,ncol=2,common.legend=TRUE, legend="bottom")
 
@@
 
-We can also make a table (Table \@ref(tab:a-table))
+Figure \@ref(fig:aoristbycem) depicts the aoristic sums of pithoi (as a proxy for burials) and of all vessels, colour-coded by cemetery.
+Immediately apparent are the broad correspondences between both cemeteries.
+Despite differences in scale and timing, we can see the Fortetsa and Knossos North Cemeteries underwent similar chronological developments in the volume of burials and, to a lesser degree, grave goods.
+In both cemeteries, the rate of burial began rising around the end of the 9th century, plateaued slightly in the early 8th, and then rose again to a peak in the late 8th and early 7th centuries.
+For associated vessels, the developments are more volatile, but again there is clear growth beginning in the later 10th century - reaching an especially marked peak in the KNC by the end of the 9th - then a slight decline and levelling into the 8th century, before a final early 7th century apogee.
+
+The near total absence of pithoi from the 11th and early 10th centuries reflects the delayed onset of cremation burial, with inhumations representing a significant proportion of earlier interments.
+Nonetheless, the clear inference would appear to be that, at both cemeteries, the late 8th and early 7th centuries witnessed an unprecedented level of burial activity, something which makes the sudden and total drop of chamber-tomb burials around 630 BC even more remarkable.
 
@@
 
+```{r aorist_uncertainty, fig.dim = c(7, 4), dpi=300}
+
+aorist_pot_values<-pots[24:73]
+aorist_pot_matrix<-data.matrix(aorist_pot_values)
+aorist_pot_matrix[aorist_pot_matrix == 0] <- NA
+aorist_pot_medians<-apply(aorist_pot_matrix,2,median,na.rm=TRUE)
+
+aorist_pot_medians<-data.frame(aorist_pot_medians)
+colnames(aorist_pot_medians)<-c("aoristic_median")
+row.names(aorist_pot_medians)<-dates
+aorist_pot_medians<-data.frame(aorist_pot_medians)
+setDT(aorist_pot_medians, keep.rownames = "decade")
+aorist_pot_medians$decade<-as.numeric(aorist_pot_medians$decade)
+
+ggplot(NULL)+geom_col(data=aorist_pot_medians,aes(decade,aoristic_median))+
+    xlim(1100,630)+theme_minimal()+
+      theme(legend.title=element_blank(),
+        text=element_text(family="Garamond", size=12))+
+            labs(x="Date (Years BC)", y="Aoristic Median")+
+               scale_fill_brewer(palette = "Set2")
+
+```
+
+```{r aoristmedianadjust}
+
+aorist_pot_medians$aorist_sums<-aorist_all_tombs$aoristic_density
+aorist_pot_medians$aorist_median_adjust<-aorist_pot_medians$aorist_sums/aorist_pot_medians$aoristic_median
+View(aorist_pot_medians)
+
+ggplot(NULL)+geom_col(data=aorist_pot_medians,aes(decade,aorist_median_adjust))+
+    xlim(1100,630)+theme_minimal()+
+      theme(legend.title=element_blank(),
+        text=element_text(family="Garamond", size=12))+
+            labs(x="Date (Years BC)", y="Aoristic Median")+
+               scale_fill_brewer(palette = "Set2")
+
+```
+
+```{r }
+
+ggplot(tombs,aes(total_finds))+geom_histogram()+xlim(0,200)
+
+aorist_tomb_values<-tombs[c(1,157:206)]
+aorist_tomb_values[aorist_tomb_values == 0] <- NA
+
+aorist_tomb_values<-data.frame(aorist_tomb_values)
+colnames(aorist_tomb_values)[2:51]<-dates
+
+aorist_melt<-melt(aorist_tomb_values)
+aorist_melt$variable<-as.numeric(as.character(aorist_melt$variable))
+
+aorist_melt$period<-ifelse(aorist_melt$variable<=1100 & aorist_melt$variable>=980,"SM",
+		ifelse(aorist_melt$variable<=970 & aorist_melt$variable>=930,"EPG",
+		ifelse(aorist_melt$variable<=920 & aorist_melt$variable>=880,"MPG",
+		ifelse(aorist_melt$variable<=870 & aorist_melt$variable>=850,"LPG",
+		ifelse(aorist_melt$variable<=840 & aorist_melt$variable>=820,"PGB",
+	  ifelse(aorist_melt$variable<=810 & aorist_melt$variable>=800,"EG",
+	  ifelse(aorist_melt$variable<=790 & aorist_melt$variable>=750,"MG",
+	  ifelse(aorist_melt$variable<=740 & aorist_melt$variable>=720,"LG",
+	  ifelse(aorist_melt$variable<=710 & aorist_melt$variable>=680,"EO",
+	  ifelse(aorist_melt$variable<=670 & aorist_melt$variable>=640,"LO",
+	  ifelse(aorist_melt$variable<=630 & aorist_melt$variable>=600,"A",""
+		)))))))))))
+
+decade_ranges<-seq(from = 600, to = 1100, by = 10)
+decade_ranges<-rev(decade_ranges)
+decade_ranges<-as.factor(decade_ranges)
+aorist_melt$variable<-as.factor(aorist_melt$variable)
+aorist_melt$variable<-factor(aorist_melt$variable, levels=(decade_ranges))
+aorist_melt$period<-factor(aorist_melt$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
+
+ggplot(aorist_melt,aes(variable,value, fill=period))+geom_boxplot()+theme_bw()+
+      theme(legend.title=element_blank(),
+        text=element_text(family="Garamond", size=12),
+        axis.text.x = element_text(angle = 70, hjust = 1),
+        panel.grid.major.x = element_blank())+
+            labs(x="Date (Years BC)", y="Aoristic Sums per Tomb")+
+               scale_fill_brewer(palette = "Set3")+
+                scale_y_continuous(expand = c(0, 0),limits=c(0,20))
+
+                   #geom_label_repel(data=subset(aorist_melt,aorist_melt$value>8 & aorist_melt$variable=="710"),
+                      #aes(label=tomb), show.legend = FALSE,ylim=c(0,40))
+
+```
+
+```{r }
+tombs_gini<-matrix(0, ncol = 0, nrow = 1)
+tombs_gini<-data.frame(tombs_gini)
+for(i in colnames(aorist_tomb_values[2:51])) {
+  tombs_gini[[paste0(i)]] <- ineq(aorist_tomb_values[[i]],type="Gini")
+}
+
+View(tombs_gini)
+
+tombs_gini<-t(tombs_gini)
+tombs_gini<-data.frame(tombs_gini)
+
+colnames(tombs_gini)<-c("gini_coefficient")
+row.names(tombs_gini)<-dates
+tombs_gini<-data.frame(tombs_gini)
+setDT(tombs_gini, keep.rownames = "decade")
+tombs_gini$decade<-as.numeric(tombs_gini$decade)
+
+ggplot(tombs_gini,aes(decade,gini_coefficient))+geom_line()+
+      xlim(1100,600)+ylim(0.3,0.8)+theme_minimal()+
+        theme(legend.title=element_blank(),
+          text=element_text(family="Garamond", size=12))+
+              labs(x="Date (Years BC)", y="Gini Coefficient")+
+                 scale_fill_brewer(palette = "Set2")
+
+rects <- data.frame(xstart = c(1100,970,920,875,840,810,790,745,70,670,630), 
+                    xend = c(970,920,875,840,810,790,745,710,670,630,600), 
+                    col = c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
+rects$col<-factor(rects$col,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
+
+ggplot() + 
+  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf, fill = col), alpha = 0.4) +
+     geom_line(data = tombs_gini, aes(decade,gini_coefficient),size=1)+
+      theme_bw()+theme(legend.title=element_blank(),
+          text=element_text(family="Garamond", size=12))+
+              labs(x="Date (Years BC)", y="Gini Coefficient")+
+                 scale_fill_brewer(palette = "Set3")+ coord_cartesian(
+                    xlim = c(1100, 600), ylim = c(0.5,0.8))+ scale_x_reverse(expand = c(0, 0),limits=c(1100,600))
+
+plot(Lc(aorist_tomb_values$`800`))
+
+```
+
+We can also make a table (Table \@ref(tab:a-table))
+
 ```{r aorist_burials}
-
-pithoi<-subset(pots,pots$shape=="Pithos")
-all_urns<-subset(pots,pots$shape=="Pithos"|pots$shape=="Amphora"|pots$shape=="Krater")
 

### 2019-04-03 analysis/paper/paper.Rmd
Writing up

@@
 
-colnames(aorist_all_tombs)<-c("aoristic_density")
+colnames(aorist_all_tombs)<-c("aoristic_sum")
 dates<-seq(from = 610, to = 1100, by = 10)
@@
 
-ggplot(aorist_all_tombs,aes(decade,aoristic_density, fill="Aoristic Sum"))+
+ggplot(aorist_all_tombs,aes(decade,aoristic_sum, fill="Aoristic Sum"))+
   geom_col()+theme_bw()+theme(legend.title=element_blank(),
@@
 
-colnames(aorist_KNC_pithoi)<-c("aoristic_density")
+colnames(aorist_KNC_pithoi)<-c("aoristic_sum")
 row.names(aorist_KNC_pithoi)<-dates
@@
 
-colnames(aorist_FOR_pithoi)<-c("aoristic_density")
+colnames(aorist_FOR_pithoi)<-c("aoristic_sum")
 row.names(aorist_FOR_pithoi)<-dates
@@
 
-colnames(aorist_all_pithoi)<-c("aoristic_density")
+colnames(aorist_all_pithoi)<-c("aoristic_sum")
 
@@
 aorist_pithoi_plot<-ggplot(NULL)+
-  geom_col(data=aorist_all_pithoi,aes(decade,aoristic_density,fill=" All tombs  "))+
-    geom_col(data=aorist_KNC_pithoi,aes(decade,aoristic_density,fill=" KNC  "))+
-      geom_col(data=aorist_FOR_pithoi,aes(decade,aoristic_density, fill=" Fortetsa  "))+
+  geom_col(data=aorist_all_pithoi,aes(decade,aoristic_sum,fill=" All tombs  "))+
+    geom_col(data=aorist_KNC_pithoi,aes(decade,aoristic_sum,fill=" KNC  "))+
+      geom_col(data=aorist_FOR_pithoi,aes(decade,aoristic_sum, fill=" Fortetsa  "))+
         theme_bw()+theme(legend.title=element_blank(),
@@
 
-colnames(aorist_KNC)<-c("aoristic_density")
+colnames(aorist_KNC)<-c("aoristic_sum")
 row.names(aorist_KNC)<-dates
@@
 
-colnames(aorist_FOR)<-c("aoristic_density")
+colnames(aorist_FOR)<-c("aoristic_sum")
 row.names(aorist_FOR)<-dates
@@
 aorist_pots_plot<-ggplot(NULL)+
-  geom_col(data=aorist_all_tombs,aes(decade,aoristic_density,fill=" All tombs  "))+
-    geom_col(data=aorist_KNC,aes(decade,aoristic_density,fill=" KNC  "))+
-      geom_col(data=aorist_FOR,aes(decade,aoristic_density, fill=" Fortetsa  "))+
+  geom_col(data=aorist_all_tombs,aes(decade,aoristic_sum,fill=" All tombs  "))+
+    geom_col(data=aorist_KNC,aes(decade,aoristic_sum,fill=" KNC  "))+
+      geom_col(data=aorist_FOR,aes(decade,aoristic_sum, fill=" Fortetsa  "))+
         theme_bw()+theme(legend.title=element_blank(),
@@
 The near total absence of pithoi from the 11th and early 10th centuries reflects the delayed onset of cremation burial, with inhumations representing a significant proportion of earlier interments.
-Nonetheless, the clear inference would appear to be that, at both cemeteries, the late 8th and early 7th centuries witnessed an unprecedented level of burial activity, something which makes the sudden and total drop of chamber-tomb burials around 630 BC even more remarkable.
+Nonetheless, the clear inference would appear to be that, at both cemeteries, the late 8th and early 7th centuries witnessed an unprecedented level of burial activity, something which makes the sudden and total decline of chamber-tomb burials around 630 BC even more notable.
 
-```{r aoristknc}
-KNC<-subset(tombs,tombs$cem_condensed=="NC")
+```{r tombconstruction}
 
-aorist_KNC <- colSums(KNC[157:206])
-aorist_KNC<-data.frame(aorist_KNC)
+tomb_dates_aorist<-read.csv("../data/derived_data/tomb_dates_aorist.csv",header=T)
+tomb_dates_aorist<-subset(tomb_dates_aorist,tomb_dates_aorist$cem_condensed=="NC"|
+                            tomb_dates_aorist$cem_condensed=="FOR")
 
-colnames(aorist_KNC)<-c("aoristic_density")
-row.names(aorist_KNC)<-dates
-aorist_KNC<-data.frame(aorist_KNC)
-setDT(aorist_KNC, keep.rownames = "decade")
-aorist_KNC$decade<-as.numeric(aorist_KNC$decade)
+aorist_tomb_dates<-colSums(tomb_dates_aorist[16:65])
+aorist_tomb_dates<-data.frame(aorist_tomb_dates)
 
-FOR<-subset(tombs,tombs$cem_condensed=="FOR")
+colnames(aorist_tomb_dates)<-c("aoristic_sum")
+row.names(aorist_tomb_dates)<-dates
+aorist_tomb_dates<-data.frame(aorist_tomb_dates)
+setDT(aorist_tomb_dates, keep.rownames = "decade")
+aorist_tomb_dates$decade<-as.numeric(aorist_tomb_dates$decade)
 
-aorist_FOR <-colSums(FOR[157:206])
-aorist_FOR<-data.frame(aorist_FOR)
+aorist_tomb_values<-tomb_dates_aorist[16:65]
+aorist_tomb_matrix<-data.matrix(aorist_tomb_values)
+aorist_tomb_matrix[aorist_tomb_matrix == 0] <- NA
+aorist_tomb_dates$medians<-apply(aorist_pot_matrix,2,median,na.rm=TRUE)
 
-colnames(aorist_FOR)<-c("aoristic_density")
-row.names(aorist_FOR)<-dates
-aorist_FOR<-data.frame(aorist_FOR)
-setDT(aorist_FOR, keep.rownames = "decade")
-aorist_FOR$decade<-as.numeric(aorist_FOR$decade)
+aorist_tomb_dates$median_adjust<-aorist_tomb_dates$aoristic_sum/aorist_tomb_dates$medians
 
-ggplot(NULL)+geom_col(data=aorist_all_tombs,aes(decade,aoristic_density,fill="All tombs"))+
-  geom_col(data=aorist_KNC,aes(decade,aoristic_density,fill="KNC"))+
-  geom_col(data=aorist_FOR,aes(decade,aoristic_density, fill="Fortetsa"))+
-      xlim(1100,630)+theme_minimal()+
-        theme(legend.title=element_blank(),
+ggplot(aorist_tomb_dates,aes(decade,median_adjust))+geom_col()+
+  theme_bw()+theme(legend.title=element_blank(),
           text=element_text(family="Garamond", size=12))+
-              labs(x="Date (Years BC)", y="Aoristic Density")+
-                 scale_fill_brewer(palette = "Set2")
+            labs(x="Date (Years BC)", y="Aoristic Sum (All Vessels)")+
+              scale_fill_brewer(palette = "Set2")+ 
+                  scale_y_continuous(expand=c(0,0),limits=c(0,65))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+aorist_tomb_periods <-colSums(tomb_dates_aorist[5:14])
+aorist_tomb_periods<-data.frame(aorist_tomb_periods)
+
+colnames(aorist_tomb_periods)<-c("aoristic_sum")
+row.names(aorist_tomb_periods)<-c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO")
+aorist_tomb_periods<-data.frame(aorist_tomb_periods)
+setDT(aorist_tomb_periods, keep.rownames = "period")
+aorist_tomb_periods$period<-factor(aorist_tomb_periods$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
+
+ggplot(aorist_tomb_periods,aes(period,aoristic_sum, fill="Period"))+geom_col()+
+  theme_bw()+theme(legend.title=element_blank(),
+          text=element_text(family="Garamond", size=12))+
+            labs(x="Period", y="Aoristic Sum")+
+              scale_fill_brewer(palette = "Set2")+ 
+                 scale_y_continuous(expand=c(0,0),limits=c(0,30))+
+                    theme(legend.position="none")
 
@@
 
-aorist_pot_medians$aorist_sums<-aorist_all_tombs$aoristic_density
-aorist_pot_medians$aorist_median_adjust<-aorist_pot_medians$aorist_sums/aorist_pot_medians$aoristic_median
-View(aorist_pot_medians)
+aorist_pot_medians$aoristic_sums<-aorist_all_tombs$aoristic_sum
+aorist_pot_medians$aorist_median_adjust<-aorist_pot_medians$aoristic_sums/aorist_pot_medians$aoristic_median
 
@@
 
-colnames(aorist_pithoi)<-c("aoristic_density")
+colnames(aorist_pithoi)<-c("aoristic_sum")
 row.names(aorist_pithoi)<-dates
@@
 
-colnames(aorist_all_urns)<-c("aoristic_density")
+colnames(aorist_all_urns)<-c("aoristic_sum")
 row.names(aorist_all_urns)<-dates
@@
 
-ggplot(NULL)+geom_col(data=aorist_all_urns,aes(decade,aoristic_density,fill="All urns"))+
-  geom_col(data=aorist_pithoi,aes(decade,aoristic_density,fill="Pithoi"))+
+ggplot(NULL)+geom_col(data=aorist_all_urns,aes(decade,aoristic_sum,fill="All urns"))+
+  geom_col(data=aorist_pithoi,aes(decade,aoristic_sum,fill="Pithoi"))+
       xlim(1100,630)+theme_minimal()+
@@
 
-aorist_by_tomb$aorist_by_tomb<-aorist_by_tomb$aoristic_density/aorist_by_tomb$no_tombs
+aorist_by_tomb$aorist_by_tomb<-aorist_by_tomb$aoristic_sum/aorist_by_tomb$no_tombs
 
@@
 aorist_tombs_9th<-data.frame(aorist_tombs_9th)
-colnames(aorist_tombs_9th)<-c("aoristic_density")
+colnames(aorist_tombs_9th)<-c("aoristic_sum")
 row.names(aorist_tombs_9th)<-dates
@@
 aorist_by_tomb_9th<-merge(aorist_tombs_9th,no_tombs_9th)
-aorist_by_tomb_9th$aorist_by_tomb<-aorist_by_tomb_9th$aoristic_density/aorist_by_tomb_9th$no_tombs
+aorist_by_tomb_9th$aorist_by_tomb<-aorist_by_tomb_9th$aoristic_sum/aorist_by_tomb_9th$no_tombs
 
@@
 
-tombs$start_period<-factor(tombs$start_period,levels=c("SM", "EPG", "MPG","PG", "LPG", "PGB", "EG", "MG","G", "LG", "EO","O", "LO"))
-
 ```
@@
 aorist_tombs_top<-data.frame(aorist_tombs_top)
-colnames(aorist_tombs_top)<-c("aoristic_density")
+colnames(aorist_tombs_top)<-c("aoristic_sum")
 row.names(aorist_tombs_top)<-dates
@@
 aorist_by_tomb_top<-merge(aorist_tombs_top,no_tombs_top)
-aorist_by_tomb_top$aorist_by_tomb<-aorist_by_tomb_top$aoristic_density/aorist_by_tomb_top$no_tombs
+aorist_by_tomb_top$aorist_by_tomb<-aorist_by_tomb_top$aoristic_sum/aorist_by_tomb_top$no_tombs
 

### 2019-04-11 analysis/paper/paper.Rmd
Figure creation and writing

@@
 )
+library(ggpubr)
+library(cowplot)
+library(gridExtra)
+library(ineq)
+library(ggrepel)
+library(reshape)
+library(data.table)
+library(extrafont)
+library(ggplot2)
+library(RColorBrewer)
+library(grid)
+library(dplyr)
+font_import(paths="../paper/font",pattern="GARA")
+loadfonts(device="win")
+
 ```
@@
 
-```{r burialestimate, fig.cap="Relationship between known interments, counts of pithoi, and counts of pithoi, amphorae and kraters.
+```{r burialestimate, fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="Relationship between known interments, counts of pithoi, and counts of pithoi, amphorae and kraters.
 The straight line represents the mean estimate based on these three values."}
@@
 The font Garmond is installed and stored.
-
-library(ggpubr)
-library(cowplot)
-library(gridExtra)
-library(ineq)
-library(ggrepel)
-library(reshape)
-library(data.table)
-library(extrafont)
-library(ggplot2)
-library(RColorBrewer)
-library(grid)
-library(dplyr)
-font_import(paths="../paper/font",pattern="GARA")
-loadfonts(device="win")
-
-#The data from the tomb database are added, and 'attached' so that the variables therein need no longer be referred to in their long format (i.e. 'tombs$total_pottery' becomes simply 'total_pottery')
 
@@
 colnames(aorist_all_tombs)<-c("aoristic_sum")
-dates<-seq(from = 610, to = 1100, by = 10)
-dates<-rev(dates)
+dates<-seq(from = 1100, to = 610, by = -10)
 row.names(aorist_all_tombs)<-dates
@@
 aorist_tomb_matrix[aorist_tomb_matrix == 0] <- NA
-aorist_tomb_dates$medians<-apply(aorist_pot_matrix,2,median,na.rm=TRUE)
+aorist_tomb_dates$medians<-apply(aorist_tomb_matrix,2,median,na.rm=TRUE)
 

### 2019-05-13 analysis/paper/paper.Rmd
Writing analysis part 1

@@
       reference_docx: "../templates/articletemplate.docx" # Insert path for the DOCX file
-bibliography: Phd.bib
+bibliography: Lib.bib
 csl: "../templates/world-archaeology.csl" # Insert path for the bib-style
@@
 )
+
+#The appropriate packages are installed.
+Extrafont allows the use of non-standard fonts in R plots, ggplot2 is a powerful figure and graph plotting tool, and RColorBrewer includes a number of colour pallettes which can be applied to plots.
+The colour scheme in this paper is chosen from those that are suitable for colour-blind readers.
+The font Garmond is installed and stored.
 library(ggpubr)
@@
 library(dplyr)
+library(plyr)
 font_import(paths="../paper/font",pattern="GARA")
@@
 
-#The appropriate packages are installed.
-Extrafont allows the use of non-standard fonts in R plots, ggplot2 is a powerful figure and graph plotting tool, and RColorBrewer includes a number of colour pallettes which can be applied to plots.
-The colour scheme in this paper is chosen from those that are suitable for colour-blind readers.
-The font Garmond is installed and stored.
+#The tombs database is imported and, for the purposes of this article, the entire database is subsetted to include only those tombs from the Knossos North and Fortetsa cemeteries (excluding the Tekke Tombs).
 
-tombs<-read.csv("../data/raw_data/KnossosTombs.csv",header=T)
-tombs<-subset(tombs,tombs$cem_condensed=="NC"|tombs$cem_condensed=="FOR")
+tombs<-read.csv("../data/raw_data/KnossosTombs.csv",header=TRUE)
+tombs<-subset(tombs,tombs$cem_condensed=="KNC"|tombs$cem_condensed=="FOR")
 
-#Figure 1 is plotted, with axis limits set (excluding tomb P, resulting in the warning message delivered) and the style of the plot adjusted (including font and text size, the appearance of the gridlines etc.)
+#Figure 1 is plotted, with axis limits set (excluding tomb P, resulting in the warning message delivered) and the style of the plot adjusted (including font and text size, the appearance of the gridlines, selecting a colour palette etc.)
 
@@
 
-#The coordinates of the plot are adjusted to render the axis as equal, and the corresponding gridlines as squares, and then the figure is plotted.
+#The coordinates of the plot are adjusted to render the axes equal, and the corresponding gridlines as squares, and then the figure is plotted.
 
@@
 
-cor_burial_proxy<-cor(tombs$all_urns,tombs$est_burials, method = "pearson")
-
 ```
 
-For most diachronic analyses, however, pithoi alone were taken as the best proxy for burial numbers, as they correlated most closely with the mean estimate, (Pearson's *r* = `r cor_burial_proxy`).
-In most cases, and often simply by virute of the methods employed, any tombs lacking identified remains or a single vessel suitable for use as an urn were not considered, thus excluding what are presumably among the most heavily damaged and looted tombs.
+For most diachronic analyses, however, pithoi alone were taken as the best proxy for burial numbers.
+In most cases, and often simply by virtue of the methods employed, any tombs lacking identified remains or a single vessel suitable for use as an urn were not considered, thus excluding what are presumably among the most heavily damaged and looted tombs.
 
@@
 This situation results in major challenges to the visualisation and analysis of such data through mean which normally rely on event- and not range-based data inputs.
-A simple, if unsatisfying, solution is to use the midpoint of the range asigned to each vessel (i.e.
+A simple, if unsatisfying, solution is to use the midpoint of the range assigned to each vessel (i.e.
 825 for our PGB example).
-This method is used in a couple of instances in the present paper when the course-grained sorting of the data (into 50 year periods) renders the innacuracies of such a method less significant.
+This method is used in a couple of instances in the present paper when the course-grained sorting of the data (into 50 year periods) renders the inaccuracies of such a method less significant.
 
-```{r allpithoi, fig.cap="Counts of pithoi across both cemeteries through time.
+```{r allpithoi,fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="Counts of pithoi across both cemeteries through time.
 Vessel dates have here been based on the midpoint of the absolute date range accompanying their respective ceramic periods."}
 
-pots<-read.csv("../data/raw_data/KnossosPots.csv",header=T)
+#The database of all vessels is imported and, again only those tombs from the KNC and Fortetsa cemetery are selected.
+A new object "pithoi" is made by subsetting only those vessels labelled "Pithos", and a similar object "all_urns" by subsetting all pithoi, amphorae and kraters.
+
+pots<-read.csv("../data/raw_data/KnossosPots.csv",header=TRUE)
 pots<-subset(pots,pots$cemetery=="FOR"|pots$cemetery=="TK"|pots$cemetery=="NC")
@@
 all_urns<-subset(pots,pots$shape=="Pithos"|pots$shape=="Amphora"|pots$shape=="Krater")
+
+#The figure is plotted by creating a histogram of all vessels based on their average dates ("ave_date"), which is simply the midpoint of the absolute chronological range associated with their respective ceramic phase.
+Similar stylistic adjustments are made (font, labels, colour palette, removing a legend, setting the x and y scales, and expanding the borders of the plot slightly to prevent the "600" label being cut off, which would occur without the final line of code.)
 
@@
                   scale_y_continuous(expand=c(0,0),limits=c(0,600))+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))
+                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
 
@@
 
-A alternative way of dealing with temporal uncertainty, and the method employed extensively in this paper, is the use of so-called "aoristic analysis".
-Here, rather than assign a pot a single date, the entire timespan under investigation is divided into equal chunks (in this case, decades), and probability values summing to 1 are then assigned for each vessel, per chunk of time, based on its accordant date range.
-So, our PGB vessel, dateable between 840 and 810 BC would register a value of 0.33 for each of the three decades within which we it could theoretically have been produced.
+A alternative way of dealing with temporal uncertainty, and the method employed extensively in this paper, is the use of so-called "aoristic analysis" [for original use in criminology, see @Ratcliffe2000; for archaeological applications, see @Crema2010; @Crema2012; @Johnson2004a].
+Here, rather than assign a pot a single date, the entire time span under investigation is divided into equal chunks (in this case, decades), and probability values summing to 1 are then assigned for each vessel, per chunk of time, based on its accordant date range.
+So, our PGB vessel, datable between 840 and 810 BC would register a value of 0.33 for each of the three decades within which we it could theoretically have been produced.
 These values for individual vessels can can then simply be summed to produce tomb- or cemetery-level signatures.
-The greatest contribution of this method is that it incorporates temporal uncertainty into the size of the resulting values, that is, "events with tight temporal definition contribute more to the total probability over their range than do loosely defined events" (JOHNSON 2003 p2).
+The major contribution of this method (though it too creates problems; see @Crema2010, 1123-1124) is that it incorporates temporal uncertainty into the size of the resulting values, that is, "events with tight temporal definition contribute more to the total probability over their range than do loosely defined events" [@Johnson2004a, 450].
 
-```{r aoristallcems, fig.cap="Aoristic sums of all vessels across both cemeteries. This represents a probabilistically weighted visualisation of depositional activity through time."}
+```{r aoristallcems, fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="Aoristic sums of all vessels across both cemeteries.
+This represents a probabilistically weighted visualisation of depositional activity through time."}
 
-aorist_all_tombs <- colSums(tombs[157:206])
+#The aoristic values for every tomb are summed by decade, and then this object is converted to a data frame.
+
+aorist_all_tombs<-colSums(tombs[158:207])
 aorist_all_tombs<-data.frame(aorist_all_tombs)
+
+#The column of values is labelled "aoristic_sum", an vector named "dates" is created with values between 1100 and 610 (in that order) at intervals of 10, and then this used to rename the rows of the data frame.
+Finally, these "row names" are converted into a functioning row in their own right within the data frame (this makes them plottable as a variable) and then are converted into numeric variables, allowing the plotting of them along a continuous scale.
 
@@
 row.names(aorist_all_tombs)<-dates
-aorist_all_tombs<-data.frame(aorist_all_tombs)
 setDT(aorist_all_tombs, keep.rownames = "decade")
 aorist_all_tombs$decade<-as.numeric(aorist_all_tombs$decade)
+
+#The plot is produced, following the same stylistic adjustments of the previous plot.
 
@@
                   scale_y_continuous(expand=c(0,0),limits=c(0,225))+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))
+                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
 
@@
 
-As seen in Figure \@ref(fig:aoristallcems), the artifical peaks and troughs of the simple averaging method are avoided, and different patterns become visible.
+As seen in Figure \@ref(fig:aoristallcems), the artificial peaks and troughs of the simple averaging method are avoided, and different patterns become visible.
 Though these methods present their own limitations, which will be explored in the following sections, they are considered a more robust and productive approach to the temporal uncertainty of ceramic dating and, as such, are used for much of the present analysis.
@@
 ##Introduction
-The great advantage of the large, quantitative dataset compiled for the present analysis is that it permits investigation of several nested levels of potentially significant patterning.
-From individual burials, to collective tombs, to the cemeteries as a whole, and between cemeteries in the landscape, we have the opportunity to study and compare both the chronological and contemporary variation across these various levels.
+The great advantage of the large, quantitative data set compiled for the present analysis is that it permits investigation of several nested levels of potentially significant patterning in the material.
+From individual burials, to collective tombs, to the cemeteries as a whole, and between cemeteries in the landscape, we have the opportunity to study and compare both the synchronic and diachronic variation across these various levels.
 The following analysis is divided broadly into these two distinct, though complementary, perspectives.
@@
 The sums are colour coded to show the values of the two cemeteries, and the total values.", fig.dim=c(8,4)}
-KNC<-subset(tombs,tombs$cem_condensed=="NC")
+KNC<-subset(tombs,tombs$cem_condensed=="KNC")
 
-aorist_KNC_pithoi <- colSums(KNC[207:256])
+aorist_KNC_pithoi <- colSums(KNC[208:257])
 aorist_KNC_pithoi<-data.frame(aorist_KNC_pithoi)
@@
 
-aorist_FOR_pithoi <-colSums(FOR[207:256])
+aorist_FOR_pithoi <-colSums(FOR[208:257])
 aorist_FOR_pithoi<-data.frame(aorist_FOR_pithoi)
@@
 
-aorist_all_pithoi <- colSums(tombs[207:256])
+aorist_all_pithoi <- colSums(tombs[208:257])
 aorist_all_pithoi<-data.frame(aorist_all_pithoi)
@@
 
-aorist_KNC <- colSums(KNC[157:206])
+aorist_KNC <- colSums(KNC[158:207])
 aorist_KNC<-data.frame(aorist_KNC)
@@
 
-aorist_FOR <-colSums(FOR[157:206])
+aorist_FOR <-colSums(FOR[158:207])
 aorist_FOR<-data.frame(aorist_FOR)
@@
 
-Figure \@ref(fig:aoristbycem) depicts the aoristic sums of pithoi (as a proxy for burials) and of all vessels, colour-coded by cemetery.
+Figure \@ref(fig:aoristbycem) presents the aoristic sums of pithoi (as a proxy for burials) and of all vessels, respectively.
 Immediately apparent are the broad correspondences between both cemeteries.
-Despite differences in scale and timing, we can see the Fortetsa and Knossos North Cemeteries underwent similar chronological developments in the volume of burials and, to a lesser degree, grave goods.
-In both cemeteries, the rate of burial began rising around the end of the 9th century, plateaued slightly in the early 8th, and then rose again to a peak in the late 8th and early 7th centuries.
-For associated vessels, the developments are more volatile, but again there is clear growth beginning in the later 10th century - reaching an especially marked peak in the KNC by the end of the 9th - then a slight decline and levelling into the 8th century, before a final early 7th century apogee.
+Despite differences in scale and timing, we can see the Fortetsa and Knossos North Cemeteries experieced similar chronological developments in the volume of burials and, to a slightly lesser degree, grave goods.
+In both cemeteries, the rate of cremation burial began rising around the end of the 9th century, plateaued slightly in the early 8th, and then rose again to a peak in the late 8th and early 7th centuries.
+Across all vessels, the developments are more volatile, but again there was growth beginning in the later 10th century - reaching an especially marked peak in the KNC by the end of the 9th - then a slight decline and levelling into the 8th century, before a final early 7th century flourit of depositional activity.
 
@@
 Nonetheless, the clear inference would appear to be that, at both cemeteries, the late 8th and early 7th centuries witnessed an unprecedented level of burial activity, something which makes the sudden and total decline of chamber-tomb burials around 630 BC even more notable.
+We might assume that these developments would be mirrored in the construction of new tombs, but this proves not to be the case.
 
-```{r tombconstruction}
+```{r tombconstruction, fig.dim=c(8,4), dpi=300, fig.cap="Visulations of tomb construction through time at both cemeteries.
+On the left are counts based on the earliest dated pottery in each tomb, while on the left are aoristic sums likewise based on the earliest ceramic-based assignations.
+In both cases, tombs lacking any securely dateable pottery (i.e.
+N/As) are excluded."}
 
-tomb_dates_aorist<-read.csv("../data/derived_data/tomb_dates_aorist.csv",header=T)
-tomb_dates_aorist<-subset(tomb_dates_aorist,tomb_dates_aorist$cem_condensed=="NC"|
-                            tomb_dates_aorist$cem_condensed=="FOR")
+tomb_dates_aorist<-read.csv("../data/derived_data/tomb_dates_aorist.csv",header=TRUE)
+tomb_dates_aorist_KNC<-subset(tomb_dates_aorist,tomb_dates_aorist$cem_condensed=="NC")
+tomb_dates_aorist_FOR<-subset(tomb_dates_aorist,tomb_dates_aorist$cem_condensed=="FOR")
 
-aorist_tomb_dates<-colSums(tomb_dates_aorist[16:65])
-aorist_tomb_dates<-data.frame(aorist_tomb_dates)
+aorist_tomb_periods<-colSums(tomb_dates_aorist_KNC[5:14])
+aorist_tomb_periods<-data.frame(aorist_tomb_periods)
+aorist_tomb_periods$FOR<-colSums(tomb_dates_aorist_FOR[5:14])
+colnames(aorist_tomb_periods)<-c("KNC","FOR")
+setDT(aorist_tomb_periods, keep.rownames = "period")
+aorist_tomb_periods$period<-factor(aorist_tomb_periods$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
+aorist_tomb_periods_melt<-melt(aorist_tomb_periods)
+colnames(aorist_tomb_periods_melt)<-c("period","cemetery","aoristic_sum")
 
-colnames(aorist_tomb_dates)<-c("aoristic_sum")
-row.names(aorist_tomb_dates)<-dates
-aorist_tomb_dates<-data.frame(aorist_tomb_dates)
-setDT(aorist_tomb_dates, keep.rownames = "decade")
-aorist_tomb_dates$decade<-as.numeric(aorist_tomb_dates$decade)
+tombs$start_period<-factor(tombs$start_period,levels=c("SM", "EPG", "MPG", "PG","LPG", "PGB", "EG", "MG","G", "LG", "EO","O","LO"))
 
-aorist_tomb_values<-tomb_dates_aorist[16:65]
-aorist_tomb_matrix<-data.matrix(aorist_tomb_values)
-aorist_tomb_matrix[aorist_tomb_matrix == 0] <- NA
-aorist_tomb_dates$medians<-apply(aorist_tomb_matrix,2,median,na.rm=TRUE)
+tombs$cem_condensed<-factor(tombs$cem_condensed,levels=c("KNC","FOR"))
 
-aorist_tomb_dates$median_adjust<-aorist_tomb_dates$aoristic_sum/aorist_tomb_dates$medians
-
-ggplot(aorist_tomb_dates,aes(decade,median_adjust))+geom_col()+
+new_tomb_counts_plot<-ggplot(data=subset(tombs,!is.na(start_period)),aes(start_period, fill=cem_condensed))+geom_bar(position="identity")+theme_classic()+
         theme_bw()+theme(legend.title=element_blank(),
           text=element_text(family="Garamond", size=12))+
-            labs(x="Date (Years BC)", y="Aoristic Sum (All Vessels)")+
-              scale_fill_brewer(palette = "Set2")+ 
-                  scale_y_continuous(expand=c(0,0),limits=c(0,65))+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+            labs(x="Period", y="Count of New Tombs")+
+              scale_fill_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+ 
+                 scale_y_continuous(expand=c(0,0),limits=c(0,30))
 
-aorist_tomb_periods <-colSums(tomb_dates_aorist[5:14])
-aorist_tomb_periods<-data.frame(aorist_tomb_periods)
-
-colnames(aorist_tomb_periods)<-c("aoristic_sum")
-row.names(aorist_tomb_periods)<-c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO")
-aorist_tomb_periods<-data.frame(aorist_tomb_periods)
-setDT(aorist_tomb_periods, keep.rownames = "period")
-aorist_tomb_periods$period<-factor(aorist_tomb_periods$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
-
-ggplot(aorist_tomb_periods,aes(period,aoristic_sum, fill="Period"))+geom_col()+
+new_tomb_aorist_plot<-ggplot(aorist_tomb_periods_melt,aes(period,aoristic_sum, fill=cemetery))+geom_col(position="identity")+
   theme_bw()+theme(legend.title=element_blank(),
           text=element_text(family="Garamond", size=12))+
-            labs(x="Period", y="Aoristic Sum")+
-              scale_fill_brewer(palette = "Set2")+ 
-                 scale_y_continuous(expand=c(0,0),limits=c(0,30))+
-                    theme(legend.position="none")
+            labs(x="Period", y="Aoristic Sum of New Tombs")+
+              scale_fill_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+ 
+                 scale_y_continuous(expand=c(0,0),limits=c(0,30))
+
+ggarrange(new_tomb_counts_plot,new_tomb_aorist_plot,ncol=2,common.legend=TRUE, legend="bottom")
 
@@
 
-```{r aorist_uncertainty, fig.dim = c(7, 4), dpi=300}
+Figure \@ref(fig:tombconstruction) presents two ways of dating the construction of the tombs.
+On the left, counts are presented, based on the ceramic phase considered most likely to correspond to the tomb's first use, including the general assignations of "PG", "G" and "O" when no greater specificity is possible.
+On the right, aoristic sums are presented, and essentially represent the spreading of the uncertainty associated with those general assignations across the sub-periods they encompass.
+In both cases, the trend appears clear.
+The North Cemetery began with a rash of building during the, admittedly prolongued, SM period, followed by a protracted decline in rates of consruction, offset by small resurgences in the late 9th and 8th centuries.
+It should also be noted, too, that many of the later tombs are not chamber tombs, but a mix of pit, larnax, and pithos burials.
+The Fortetsa, meanwhile, saw next to no new chamber tomb construction from the 8th century onwards.
+In both cases, then, the rates of tomb building run counter the late 8th-early 7th century spate of burial activity.
+These countervalent trends present a number of interpretive problems.
+The abatement of chamber tomb construction has been noted before (Brock 1957, 4; Cavanagh 1996, 652–3) and has been somewhat implicitly read as evidence of restructuring of the funerary rite, if not a decline in the fortunes of the wider community.
+But this image of a waigning investment in chamber burials sits uneasily with the apparent growth in mortuary activity suggested by both counts of artefacts, and estimates of the number of tombs in use in each period.
 
-aorist_pot_values<-pots[24:73]
-aorist_pot_matrix<-data.matrix(aorist_pot_values)
-aorist_pot_matrix[aorist_pot_matrix == 0] <- NA
-aorist_pot_medians<-apply(aorist_pot_matrix,2,median,na.rm=TRUE)
+```{r tombsinuse, fig.dim=c(7,4), dpi=300, fig.cap="Estimates of the number of tombs in use during each of the ceramic phases attested in both cemetaries."}
 
-aorist_pot_medians<-data.frame(aorist_pot_medians)
-colnames(aorist_pot_medians)<-c("aoristic_median")
-row.names(aorist_pot_medians)<-dates
-aorist_pot_medians<-data.frame(aorist_pot_medians)
-setDT(aorist_pot_medians, keep.rownames = "decade")
-aorist_pot_medians$decade<-as.numeric(aorist_pot_medians$decade)
+KNC_tombs_in_use<-colSums(KNC[6:15]!="")
+KNC_tombs_in_use<-data.frame(KNC_tombs_in_use)
+colnames(KNC_tombs_in_use)<-c("no_tombs")
+setDT(KNC_tombs_in_use, keep.rownames = "period")
+KNC_tombs_in_use$period<-factor(KNC_tombs_in_use$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
+KNC_tombs_in_use$cemetery<-"KNC"
 
-ggplot(NULL)+geom_col(data=aorist_pot_medians,aes(decade,aoristic_median))+
-    xlim(1100,630)+theme_minimal()+
-      theme(legend.title=element_blank(),
+FOR_tombs_in_use<-colSums(FOR[6:15]!="")
+FOR_tombs_in_use<-data.frame(FOR_tombs_in_use)
+colnames(FOR_tombs_in_use)<-c("no_tombs")
+setDT(FOR_tombs_in_use, keep.rownames = "period")
+FOR_tombs_in_use$period<-factor(FOR_tombs_in_use$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
+FOR_tombs_in_use$cemetery<-"FOR"
+
+tombs_in_use_cem<-rbind(KNC_tombs_in_use,FOR_tombs_in_use)
+tombs_in_use_cem$cemetery<-factor(tombs_in_use_cem$cemetery,levels=c("KNC","FOR"))
+
+ggplot(tombs_in_use_cem,aes(period,no_tombs, fill=cemetery))+geom_col(position="identity")+theme_classic()+
+       theme_bw()+theme(legend.title=element_blank(),
           text=element_text(family="Garamond", size=12))+
-            labs(x="Date (Years BC)", y="Aoristic Median")+
-               scale_fill_brewer(palette = "Set2")
-
+            labs(x="Period", y="Tombs in Use")+
+              scale_fill_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+ 
+                 scale_y_continuous(expand=c(0,0),limits=c(0,50))
 ```
 
-```{r aoristmedianadjust}
+```{r }
+tombs_11th<-subset(tombs,tombs$century_built=="11th")
+tombs_10th<-subset(tombs,tombs$century_built=="10th")
+tombs_9th<-subset(tombs,tombs$century_built=="9th")
+tombs_8th<-subset(tombs,tombs$century_built=="8th")
+tombs_7th<-subset(tombs,tombs$century_built=="7th")
 
-aorist_pot_medians$aoristic_sums<-aorist_all_tombs$aoristic_sum
-aorist_pot_medians$aorist_median_adjust<-aorist_pot_medians$aoristic_sums/aorist_pot_medians$aoristic_median
+tombs_in_use_11th<-colSums(tombs_11th[6:15]!="")
+tombs_in_use_11th<-data.frame(tombs_in_use_11th)
+colnames(tombs_in_use_11th)<-c("no_tombs")
+setDT(tombs_in_use_11th, keep.rownames = "period")
+tombs_in_use_11th$period<-factor(tombs_in_use_11th$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
+tombs_in_use_11th$tomb_century<-"11th"
 
-ggplot(NULL)+geom_col(data=aorist_pot_medians,aes(decade,aorist_median_adjust))+
-    xlim(1100,630)+theme_minimal()+
-      theme(legend.title=element_blank(),
+tombs_in_use_10th<-colSums(tombs_10th[6:15]!="")
+tombs_in_use_10th<-data.frame(tombs_in_use_10th)
+colnames(tombs_in_use_10th)<-c("no_tombs")
+setDT(tombs_in_use_10th, keep.rownames = "period")
+tombs_in_use_10th$period<-factor(tombs_in_use_10th$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
+tombs_in_use_10th$tomb_century<-"10th"
+
+tombs_in_use_9th<-colSums(tombs_9th[6:15]!="")
+tombs_in_use_9th<-data.frame(tombs_in_use_9th)
+colnames(tombs_in_use_9th)<-c("no_tombs")
+setDT(tombs_in_use_9th, keep.rownames = "period")
+tombs_in_use_9th$period<-factor(tombs_in_use_9th$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
+tombs_in_use_9th$tomb_century<-"9th"
+
+tombs_in_use_8th<-colSums(tombs_8th[6:15]!="")
+tombs_in_use_8th<-data.frame(tombs_in_use_8th)
+colnames(tombs_in_use_8th)<-c("no_tombs")
+setDT(tombs_in_use_8th, keep.rownames = "period")
+tombs_in_use_8th$period<-factor(tombs_in_use_8th$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
+tombs_in_use_8th$tomb_century<-"8th"
+
+tombs_in_use_7th<-colSums(tombs_7th[6:15]!="")
+tombs_in_use_7th<-data.frame(tombs_in_use_7th)
+colnames(tombs_in_use_7th)<-c("no_tombs")
+setDT(tombs_in_use_7th, keep.rownames = "period")
+tombs_in_use_7th$period<-factor(tombs_in_use_7th$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
+tombs_in_use_7th$tomb_century<-"7th"
+
+tomb_use<-rbind(tombs_in_use_11th,tombs_in_use_10th,tombs_in_use_9th,tombs_in_use_8th,tombs_in_use_7th)
+tomb_use$tomb_century<-factor(tomb_use$tomb_century,levels=c("11th","10th","9th","8th","7th"))
+
+ggplot(data=tomb_use,aes(period,no_tombs))+geom_col(alpha=0.2)+
+            geom_col(data=tomb_use,aes(period,no_tombs,fill=tomb_century),position="dodge")+
+              theme_bw()+theme(legend.title=element_blank(),
                text=element_text(family="Garamond", size=12))+
-            labs(x="Date (Years BC)", y="Aoristic Median")+
-               scale_fill_brewer(palette = "Set2")
+                 labs(x="Period", y="Estimated No. of Tombs in Use")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0),limits=c(0,60))
 
@@
 
-aorist_tomb_values<-tombs[c(1,157:206)]
+aorist_tomb_values<-tombs[c(1,158:207)]
 aorist_tomb_values[aorist_tomb_values == 0] <- NA
@@
 
-no_tombs<-colSums(tombs[157:206]!=0)
+no_tombs<-colSums(tombs[158:207]!=0)
 no_tombs<-data.frame(no_tombs)
@@
 tombs_9th<-subset(tombs,tombs$century_built=="9th")
-no_tombs_9th<-colSums(tombs_9th[157:206]!=0)
+no_tombs_9th<-colSums(tombs_9th[158:207]!=0)
 
+View(tombs_9th)
 no_tombs_9th<-data.frame(no_tombs_9th)
@@
 
-aorist_tombs_9th<-colSums(tombs_9th[157:206])
+aorist_tombs_9th<-colSums(tombs_9th[158:207])
 aorist_tombs_9th<-data.frame(aorist_tombs_9th)
@@
 names(tombs)
-tombs_in_use<-colSums(tombs[5:14]!="")
+tombs_in_use<-colSums(tombs[6:15]!="")
 tombs_in_use<-data.frame(tombs_in_use)
@@
 
-ggplot(no_tombs,aes(decade,no_tombs, fill="Tombs in Use"))+geom_col()+theme_classic()+
+ggplot(no_tombs_9th,aes(decade,no_tombs, fill="Tombs in Use"))+geom_col()+theme_classic()+
      theme(legend.title=element_blank(),
@@
 
-ggplot(tombs,aes(start_period, fill="New Tombs"))+geom_bar()+theme_classic()+
+ggplot(tombs,aes(start_period, fill=tomb_type_condensed))+geom_bar()+theme_classic()+
         theme(legend.title=element_blank(),
@@
 
-no_tombs_top<-colSums(tombs_top_12[157:206]!=0)
+View(tombs_top_12)
+
+no_tombs_top<-colSums(tombs_top_12[158:207]!=0)
 
@@
 
-aorist_tombs_top<-colSums(tombs_top_12[157:206])
+aorist_tombs_top<-colSums(tombs_top_12[158:207])
 aorist_tombs_top<-data.frame(aorist_tombs_top)
@@
                  scale_fill_brewer(palette = "Set2")
+
+```
+
+```{r top_tombs, fig.dim = c(7, 5), dpi=300, fig.align='center',}
+
+cols<-c(1,158:207)
+aorist_all_tomb_names<-tombs_top_12[,cols]
+aorist_all_tomb_names<-data.frame(aorist_all_tomb_names)
+colnames(aorist_all_tomb_names)<-c("tomb",dates)
+
+aorist_all_tomb_melt<-melt(aorist_all_tomb_names)
+str(aorist_all_tomb_melt)
+aorist_all_tomb_melt$variable<-as.numeric(as.character(aorist_all_tomb_melt$variable))
+
+top_12<-tombs_top_12$tomb
+aorist_all_tomb_melt$tomb<-factor(aorist_all_tomb_melt$tomb,levels=top_12)
+
+ggplot(aorist_all_tomb_melt,aes(variable, value))+geom_col()+theme(legend.position = "none")+facet_wrap(~tomb)+
+  geom_col()+theme_bw()+theme(legend.title=element_blank(),
+      text=element_text(family="Garamond", size=12))+
+          labs(x="Date (Years BC)", y="Aoristic Sum")+
+              scale_fill_brewer(palette = "Set3")+ 
+                theme(legend.position="none")+
+                  coord_cartesian(ylim=c(0,30),xlim=c(1100,610))+scale_y_continuous(expand=c(0,0))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1100,610))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+```
+
+```{r cumfreq}
 

### 2019-05-14 analysis/paper/paper.Rmd
Continuing writing analysis

@@
 library(plyr)
+library(ggforce)
 font_import(paths="../paper/font",pattern="GARA")
@@
 pots<-read.csv("../data/raw_data/KnossosPots.csv",header=TRUE)
-pots<-subset(pots,pots$cemetery=="FOR"|pots$cemetery=="TK"|pots$cemetery=="NC")
+pots<-subset(pots,pots$cem_condensed=="FOR"|pots$cem_condensed=="NC")
 pithoi<-subset(pots,pots$shape=="Pithos")
@@
 Figure \@ref(fig:tombconstruction) presents two ways of dating the construction of the tombs.
-On the left, counts are presented, based on the ceramic phase considered most likely to correspond to the tomb's first use, including the general assignations of "PG", "G" and "O" when no greater specificity is possible.
+On the left, counts are presented, based on the ceramic phase considered most likely to correspond to the tomb's first use, including the broad labels "PG", "G" and "O" where no greater specificity is possible.
 On the right, aoristic sums are presented, and essentially represent the spreading of the uncertainty associated with those general assignations across the sub-periods they encompass.
@@
 The North Cemetery began with a rash of building during the, admittedly prolongued, SM period, followed by a protracted decline in rates of consruction, offset by small resurgences in the late 9th and 8th centuries.
-It should also be noted, too, that many of the later tombs are not chamber tombs, but a mix of pit, larnax, and pithos burials.
+It should be noted, too, that many of the later "tombs" are not chamber tombs, but a mix of pit, larnax, and pithos burials.
 The Fortetsa, meanwhile, saw next to no new chamber tomb construction from the 8th century onwards.
@@
 
-ggplot(tombs_in_use_cem,aes(period,no_tombs, fill=cemetery))+geom_col(position="identity")+theme_classic()+
+ggplot(tombs_in_use_cem,aes(period,no_tombs, fill=cemetery))+geom_col(position="dodge")+theme_classic()+
        theme_bw()+theme(legend.title=element_blank(),
@@
               scale_fill_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+ 
-                 scale_y_continuous(expand=c(0,0),limits=c(0,50))
+                 scale_y_continuous(expand=c(0,0),limits=c(0,45))
 ```
 
-```{r }
+Figure \@ref(fig:tombsinuse) presents the estimated number of tombs in use throughout the EIA at both cemeteries, that is, the number of tombs which yielded pottery of each period.
+At the Forteta, this number remained rlatively stable from the MPG to LG periods before declining (and bear in mind, there are only around 20 tombs belonging to this cemetery throughout the period), while at the KNC, the numbers of tombs in use rose to a preak in the late 8th century before declining.
+The cessation of burials there certainly appears more sudden and marked than at Fortetsa, though, again, it was always a larger cemetery.
+But with the data available, we can interrogate these trends further.
+
+```{r tombsinusebybuilddate, fig.dim=c(7,4), dpi=300, fig.cap="Estimates of the number of tombs in use through time across both cemeteries.
+The grey background represents the total number of tombs in use in each period, while the smaller coloured bars correspond to tombs built in each of the five centuries of EIA activity at the cemeteries."}
 tombs_11th<-subset(tombs,tombs$century_built=="11th")
@@
 
+Figure \@ref(fig:tombsinusebybuilddate) presents the number of tombs in use in each period, divided by the centuries in which those tombs were constructed, all overlaid on the total number of tombs in use across both cemeteries.
+Perhaps unsurprisingly, a wave-like pattern can be observed with each century marked by the construction of new tombs which then steadily go out of use, as new ones are built.
+This would seem to accord with Cavanagh's suggestion that "[t]he length of time the tombs last seems [...] to reflect the duration of the social group which had access to the sepulchre", and perhaps even that this access was lineally inehrited, leading to frequent fission and, thus, limited trans-generational use.
+However, something else is going here, which can only be appreciated by considering which tombs became the focus of the intense period of depositional activity in the late 8th and erly 7th centuries.
+
+```{r potsbydateandtombdate,  fig.dim=c(8,4), dpi=300, fig.cap="Counts of vessels from across all tombs in both cemeteries, arranged, on the left, by the century to which they are assigned based on their ceramic period and, on the right, by the centuries in which the tombs they come from were built."}
+
+pots$century<-factor(pots$century,levels=c("11th","10th","9th","8th","7th"))
+pots$tomb_century<-factor(pots$tomb_century,levels=c("11th","10th","9th","8th","7th"))
+
+date_by_vessel<-ggplot(data=subset(pots,pots$century!=""),aes(century,fill="Date"))+geom_bar()+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Century (Date of Vessel)", y="Vessel Count")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0),limits=c(0,2800))
+
+date_by_tomb<-ggplot(pots,aes(tomb_century,fill="Date"))+geom_bar()+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Century (Date of Tomb)", y="")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0),limits=c(0,2800))
+
+ggarrange(date_by_vessel,date_by_tomb,ncol=2)
+
+```
+
+Figure \@ref(fig:potsbydateandtombdate) shows, firstly, the number of vessels dating to each century across both cemeteries, based on their assigned ceramic period, essentially a simplified version of the trend illustrated in Figure \@ref(fig:aoristallcems).
+But, on the right, Figure \@ref(fig:potsbydateandtombdate) shows those pots divided into the centuries to which the tombs they were found in date.
+So as can be seen, though the late 8th and early 7th centuries represented the apex of burial activity in across both cemeteries, it was not the tombs built in those centuries, but rather those built in the centuries prior, which received the lion's share of the burials and associated ceramics.
+This is all the more notable when, returning to Figure \@ref(fig:tombsinusebybuilddate), we see that the number of 9th and 10th century tombs in use by the late 8th and early 7th centuries were declining.
+
 ```{r }
+
+pots10th10th<-sum(pithoi$tomb_century=="10th" & pithoi$century=="10th",na.rm=TRUE)
+pots10th9th<-sum(pithoi$tomb_century=="10th" & pithoi$century=="9th",na.rm=TRUE)
+pots10th8th<-sum(pithoi$tomb_century=="10th" & pithoi$century=="8th",na.rm=TRUE)
+pots10th7th<-sum(pithoi$tomb_century=="10th" & pithoi$century=="7th",na.rm=TRUE)
+
+notombs10th10th<-nrow(unique(subset(pithoi,tomb_century=="10th" & century=="10th",select=tomb)))
+notombs10th9th<-nrow(unique(subset(pithoi,tomb_century=="10th" & century=="9th",select=tomb)))
+notombs10th8th<-nrow(unique(subset(pithoi,tomb_century=="10th" & century=="8th",select=tomb)))
+notombs10th7th<-nrow(unique(subset(pithoi,tomb_century=="10th" & century=="7th",select=tomb)))
+
+pots9th9th<-sum(pithoi$tomb_century=="9th" & pithoi$century=="9th",na.rm=TRUE)
+pots9th8th<-sum(pithoi$tomb_century=="9th" & pithoi$century=="8th",na.rm=TRUE)
+pots9th7th<-sum(pithoi$tomb_century=="9th" & pithoi$century=="7th",na.rm=TRUE)
+
+notombs9th9th<-nrow(unique(subset(pithoi,tomb_century=="9th" & century=="9th",select=tomb)))
+notombs9th8th<-nrow(unique(subset(pithoi,tomb_century=="9th" & century=="8th",select=tomb)))
+notombs9th7th<-nrow(unique(subset(pithoi,tomb_century=="9th" & century=="7th",select=tomb)))
+
+pots8th8th<-sum(pithoi$tomb_century=="8th" & pithoi$century=="8th",na.rm=TRUE)
+pots8th7th<-sum(pithoi$tomb_century=="8th" & pithoi$century=="7th",na.rm=TRUE)
+
+notombs8th8th<-nrow(unique(subset(pithoi,tomb_century=="8th" & century=="8th",select=tomb)))
+notombs8th7th<-nrow(unique(subset(pithoi,tomb_century=="8th" & century=="7th",select=tomb)))
+
+pots7th7th<-sum(pithoi$tomb_century=="7th" & pithoi$century=="7th",na.rm=TRUE)
+
+notombs7th7th<-nrow(unique(subset(pithoi,tomb_century=="7th" & century=="7th",select=tomb)))
+
+```
+
+|**Tombs**          | **Date of pithoi (burials)**| **No. of pithoi (burials)**| **Number of tombs represented** | **Burials per tomb**                   |
+|-------------------|--------------------------|-------------------------|-----------------------------|--------------------------------------------------|
+|*10th century tombs*|                         |                         |                             |                                                  |
+|                   |10th century              | `r pots10th10th`        | `r notombs10th10th`         | `r round(pots10th10th/notombs10th10th,digits=2)` |
+|                   |9th century               | `r pots10th9th`         | `r notombs10th9th`          | `r round(pots10th9th/notombs10th9th,digits=2)`   |
+|                   |8th century               | `r pots10th8th`         | `r notombs10th8th`          | `r round(pots10th8th/notombs10th8th,digits=2)`   |
+|                   |7th century               | `r pots10th7th`         | `r notombs10th7th`          | `r round(pots10th7th/notombs10th7th,digits=2)`   |
+|*9th century tombs*|                          |                         |                             |                                                  |
+|                   |9th century               | `r pots9th9th`          | `r notombs9th9th`           | `r round(pots9th9th/notombs9th9th,digits=2)`     |
+|                   |8th century               | `r pots9th8th`          | `r notombs9th8th`           | `r round(pots9th8th/notombs9th8th,digits=2)`     |
+|                   |7th century               | `r pots9th7th`          | `r notombs9th7th`           | `r round(pots9th7th/notombs9th7th,digits=2)`     |
+|*8th century tombs*|                          |                         |                             |                                                  |
+|                   |8th century               | `r pots8th8th`          | `r notombs8th8th`           | `r round(pots8th8th/notombs8th8th,digits=2)`     |
+|                   |7th century               | `r pots8th7th`          | `r notombs8th7th`           | `r round(pots8th7th/notombs8th7th,digits=2)`     |
+|*7th century tombs*|                          |                         |                             |                                                  |
+|                   |7th century               | `r pots7th7th`          | `r notombs7th7th`           | `r round(pots7th7th/notombs7th7th,digits=2)`     |
+
+Table: Table 1 The number of pithoi (as a proxy for burials) deposited in tombs dating to the 10th, 9th, 8th and 7th centuries, through each of the centuries they were in use, as well as the number of tombs from which these burials came, allowing calculation of the changing rate of burial through time.
+
+The significance of these findings can be seen in Table 1.
+Taking the number of pithoi present in each tomb as a proxy for burials, we can estimate the changing rates of burial through the centuries.
+And while for most of the cemeteries' lives the burial rate remained around three per tomb per century, in the 8th and 7th centuries tombs founded in the 9th century saw a rapid rise in the rate of burial, peaking at around 12 per tomb per century.
+Though these estimates are no doubt very rough, the marked nature of this divergence between the 9th century tombs and the rest suggests it is not simply an artefact of imperfect data.
+
+```{r boxplot,  fig.dim=c(9,5), dpi=300,}
 
@@
 aorist_melt$period<-factor(aorist_melt$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
+aorist_melt$variable<-as.numeric(as.character(aorist_melt$variable))
 
-ggplot(aorist_melt,aes(variable,value, fill=period))+geom_boxplot()+theme_bw()+
+ggplot(aorist_melt,aes(variable,value,fill=period))+geom_boxplot(aes(group = cut_width(variable, 10)))+theme_bw()+
       theme(legend.title=element_blank(),
@@
                scale_fill_brewer(palette = "Set3")+
-                scale_y_continuous(expand = c(0, 0),limits=c(0,20))
+                scale_x_reverse()+
+                  scale_y_continuous()+
+                  facet_zoom(x=variable<900& variable>640,y=value<20)
 
-                   #geom_label_repel(data=subset(aorist_melt,aorist_melt$value>8 & aorist_melt$variable=="710"),
-                      #aes(label=tomb), show.legend = FALSE,ylim=c(0,40))
+ggplot(aorist_melt,aes(variable,value,fill=period))+geom_boxplot(aes(group = cut_width(variable, 10)))+theme_bw()+
+      theme(legend.title=element_blank(),
+        text=element_text(family="Garamond", size=12),
+        axis.text.x = element_text(angle = 70, hjust = 1),
+        panel.grid.major.x = element_blank())+
+            labs(x="Date (Years BC)", y="Aoristic Sums per Tomb")+
+               scale_fill_brewer(palette = "Set3")+
+                scale_x_reverse()+
+                  scale_y_continuous(limits=c(0,20))+geom_label(data=subset(aorist_melt,value>19),aes(label=tomb))
 
@@
 
-```{r }
+```{r dpi=300}
 tombs_gini<-matrix(0, ncol = 0, nrow = 1)
@@
 }
-
-View(tombs_gini)
 
@@
   geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf, fill = col), alpha = 0.4) +
-     geom_line(data = tombs_gini, aes(decade,gini_coefficient),size=1)+
+     geom_line(data = tombs_gini, aes(decade,gini_coefficient),size=0.5)+
       theme_bw()+theme(legend.title=element_blank(),
@@
 
-We can also make a table (Table \@ref(tab:a-table))
-
 ```{r aorist_burials}
 
-aorist_pithoi<-colSums(pithoi[24:73])
+aorist_pithoi<-colSums(pithoi[25:74])
 aorist_pithoi<-data.frame(aorist_pithoi)
@@
 
-aorist_all_urns<-colSums(all_urns[24:73])
+aorist_all_urns<-colSums(all_urns[25:74])
 aorist_all_urns<-data.frame(aorist_all_urns)

### 2019-05-15 analysis/paper/paper.Rmd
Cut downs and edits

@@
 
-```{r boxplot,  fig.dim=c(9,5), dpi=300,}
+Also apparent from Table 1, though, are the markedly low rates of burial which edominated at both cemeteries.
+Two to four burials per tomb per century amounts to a very selective burial rite, even considering the restricted portion of the population probably represented by these cemeteries [@Cavanagh1996, 664].
+It seems unlikely, then, that the rising rates of burial noted above are, for instance, the direct result of population growth; the increase seems too rapid, and is concentrated only in certain (odler) tombs, not to mention that, based on KULP’s survey data, the major population increase at EIA Knossos should be dated a century or more before the late 8th century peak in burial, in the Protogeometric period (Whitelaw, per comm.).
 
-ggplot(tombs,aes(total_finds))+geom_histogram()+xlim(0,200)
+We might assume that the tombs which became the focus of intense burial activity in the late 8th and early 7th centuries would be those which yielded the most burials and grave goods when excavated.
+Certainly, the largest 12 tombs are all chamber tombs, with 10 dating to the 9th century, and two to the 10th.
+But looking more closely at the aoristic sums of their assemblages through time, not all of them appear to have exhibited the noted intensification of burial activity into the Orientalising period.
+
+```{r top_tombs, fig.dim = c(7, 5), dpi=300, fig.align='center',fig.cap="Aoristic sums for the 12 largest tombs (in terms of grave goods), arranged in decreasing order of total recovered artefacts.
+Note, for the sake of better illustrating the relative patterns, rather than absolute values, the scales on each plot are distinct."}
+
+cols<-c(1,158:207)
+aorist_all_tomb_names<-tombs_top_20[,cols]
+aorist_all_tomb_names<-data.frame(aorist_all_tomb_names)
+colnames(aorist_all_tomb_names)<-c("tomb",dates)
+
+aorist_all_tomb_melt<-melt(aorist_all_tomb_names)
+aorist_all_tomb_melt$variable<-as.numeric(as.character(aorist_all_tomb_melt$variable))
+
+top_20<-tombs_top_20$tomb
+aorist_all_tomb_melt$tomb<-factor(aorist_all_tomb_melt$tomb,levels=top_20)
+
+ggplot(aorist_all_tomb_melt,aes(variable, value))+geom_col()+theme(legend.position = "none")+facet_wrap(~tomb,scales="free_y")+
+  geom_col()+theme_bw()+theme(legend.title=element_blank(),
+      text=element_text(family="Garamond", size=12))+
+          labs(x="Date (Years BC)", y="Aoristic Sums")+
+              scale_fill_brewer(palette = "Set3")+ 
+                theme(legend.position="none")+
+                  coord_cartesian(xlim=c(1100,610))+scale_y_continuous()+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1100,610))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+```
+
+From Figure \@ref(fig:top_tombs), we can see that, though seven of the 12 did exhibit late, intense episodes of deposition, tombs X, G, Q, 104 and 283 evince a different pattern, with their respective peaks coming in the PGB and EG periods, around a century earlier.
+We have seen already in Figure \@ref(fig:aoristbycem) that, particularly in the KNC, the end of the 9th century was another period of increased burial activity
+
+```{r boxplot,  fig.dim=c(7,4), dpi=300,}
+
+tombs_top_20<-head(tombs_ranked,20)
+top20<-as.vector(tombs_top_20$tomb)
+pots_top_20$tomb<-factor(pots_top_20$tomb,levels=top20)
+
+pots_top_20<-pots[is.element(pots$tomb, top20),]
+pots_top_20_9th_7th<-subset(pots_top_20,century=="9th"|century=="8th"|century=="7th")
+ggplot(pots_top_20_9th_7th,aes(tomb,fill=tomb_century))+geom_bar()+facet_wrap(~century)
+
+tombs$group<-ifelse(tombs$tomb=="X"|tombs$tomb=="G"|tombs$tomb=="Q"|tombs$tomb=="104"|tombs$tomb=="283","A",
+                   ifelse(tombs$tomb=="P"|tombs$tomb=="292"|tombs$tomb=="75"|tombs$tomb=="107"|tombs$tomb=="285"|tombs$tomb=="II"|tombs$tomb=="218","B","OTHER"))
+names(tombs)
+ggplot(tombs,aes(antique,fill=group))+geom_histogram()
 
@@
 
-ggplot(aorist_melt,aes(variable,value,fill=period))+geom_boxplot(aes(group = cut_width(variable, 10)))+theme_bw()+
+ggplot(aorist_melt,aes(variable,value,fill=period,outlier.size=0.3))+geom_boxplot(aes(group = cut_width(variable, 10)))+theme_bw()+
       theme(legend.title=element_blank(),
@@
                scale_fill_brewer(palette = "Set3")+
-                scale_x_reverse()+
-                  scale_y_continuous()+
-                  facet_zoom(x=variable<900& variable>640,y=value<20)
-
-ggplot(aorist_melt,aes(variable,value,fill=period))+geom_boxplot(aes(group = cut_width(variable, 10)))+theme_bw()+
-      theme(legend.title=element_blank(),
-        text=element_text(family="Garamond", size=12),
-        axis.text.x = element_text(angle = 70, hjust = 1),
-        panel.grid.major.x = element_blank())+
-            labs(x="Date (Years BC)", y="Aoristic Sums per Tomb")+
-               scale_fill_brewer(palette = "Set3")+
-                scale_x_reverse()+
-                  scale_y_continuous(limits=c(0,20))+geom_label(data=subset(aorist_melt,value>19),aes(label=tomb))
-
+                scale_x_reverse(expand=c(0,0))+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,20))
+names(tombs)
+tombs$century_built<-factor(tombs$century_built,levels=c("11th","10th","9th","8th","7th"))
+ggplot(tombs,aes(start_period,dromos_length))+geom_boxplot()
+ggplot(data=subset(pots,pots$import=="Y"),aes(period))+geom_histogram()+xlim(1100,630)
 ```
@@
 ```{r dpi=300}
+aorist_tomb_values<-tombs[c(1,158:207)]
+aorist_tomb_values[aorist_tomb_values == 0] <- NA
+
+aorist_tomb_values<-data.frame(aorist_tomb_values)
+colnames(aorist_tomb_values)[2:51]<-dates
+
 tombs_gini<-matrix(0, ncol = 0, nrow = 1)
@@
 
-ggplot(tombs_gini,aes(decade,gini_coefficient))+geom_line()+
-      xlim(1100,600)+ylim(0.3,0.8)+theme_minimal()+
-        theme(legend.title=element_blank(),
-          text=element_text(family="Garamond", size=12))+
-              labs(x="Date (Years BC)", y="Gini Coefficient")+
-                 scale_fill_brewer(palette = "Set2")
-
 rects <- data.frame(xstart = c(1100,970,920,875,840,810,790,745,70,670,630), 
@@
                     xlim = c(1100, 600), ylim = c(0.5,0.8))+ scale_x_reverse(expand = c(0, 0),limits=c(1100,600))
-
-plot(Lc(aorist_tomb_values$`800`))
-
-```
-
-```{r aorist_burials}
-
-aorist_pithoi<-colSums(pithoi[25:74])
-aorist_pithoi<-data.frame(aorist_pithoi)
-
-colnames(aorist_pithoi)<-c("aoristic_sum")
-row.names(aorist_pithoi)<-dates
-aorist_pithoi<-data.frame(aorist_pithoi)
-setDT(aorist_pithoi, keep.rownames = "decade")
-aorist_pithoi$decade<-as.numeric(aorist_pithoi$decade)
-
-aorist_all_urns<-colSums(all_urns[25:74])
-aorist_all_urns<-data.frame(aorist_all_urns)
-
-colnames(aorist_all_urns)<-c("aoristic_sum")
-row.names(aorist_all_urns)<-dates
-aorist_all_urns<-data.frame(aorist_all_urns)
-setDT(aorist_all_urns, keep.rownames = "decade")
-aorist_all_urns$decade<-as.numeric(aorist_all_urns$decade)
-
-ggplot(NULL)+geom_col(data=aorist_all_urns,aes(decade,aoristic_sum,fill="All urns"))+
-  geom_col(data=aorist_pithoi,aes(decade,aoristic_sum,fill="Pithoi"))+
-      xlim(1100,630)+theme_minimal()+
-        theme(legend.title=element_blank(),
-          text=element_text(family="Garamond", size=12))+
-              labs(x="Date (Years BC)", y="Aoristic Density")+
-                 scale_fill_brewer(palette = "Set2")
-
-```
-
-```{r aorist_tomb_adjust}
-
-no_tombs<-colSums(tombs[158:207]!=0)
-no_tombs<-data.frame(no_tombs)
-colnames(no_tombs)<-c("no_tombs")
-row.names(no_tombs)<-dates
-no_tombs<-data.frame(no_tombs)
-setDT(no_tombs, keep.rownames = "decade")
-no_tombs$decade<-as.numeric(no_tombs$decade)
-
-aorist_all_tombs1<-aorist_all_tombs
-aorist_by_tomb<-merge(aorist_all_tombs1,no_tombs)
-
-aorist_by_tomb$aorist_by_tomb<-aorist_by_tomb$aoristic_sum/aorist_by_tomb$no_tombs
-
-ggplot(aorist_by_tomb,aes(decade,aorist_by_tomb))+geom_col()+xlim(1100,600)
-
-#Now for the 9th century tombs
-
-tombs_9th<-subset(tombs,tombs$century_built=="9th")
-no_tombs_9th<-colSums(tombs_9th[158:207]!=0)
-
-View(tombs_9th)
-no_tombs_9th<-data.frame(no_tombs_9th)
-colnames(no_tombs_9th)<-c("no_tombs")
-row.names(no_tombs_9th)<-dates
-no_tombs_9th<-data.frame(no_tombs_9th)
-setDT(no_tombs_9th, keep.rownames = "decade")
-no_tombs_9th$decade<-as.numeric(no_tombs_9th$decade)
-
-aorist_tombs_9th<-colSums(tombs_9th[158:207])
-aorist_tombs_9th<-data.frame(aorist_tombs_9th)
-colnames(aorist_tombs_9th)<-c("aoristic_sum")
-row.names(aorist_tombs_9th)<-dates
-aorist_tombs_9th<-data.frame(aorist_tombs_9th)
-setDT(aorist_tombs_9th, keep.rownames = "decade")
-aorist_tombs_9th$decade<-as.numeric(aorist_tombs_9th$decade)
-
-aorist_by_tomb_9th<-merge(aorist_tombs_9th,no_tombs_9th)
-aorist_by_tomb_9th$aorist_by_tomb<-aorist_by_tomb_9th$aoristic_sum/aorist_by_tomb_9th$no_tombs
-
-ggplot()+geom_col(data=aorist_by_tomb_9th,aes(decade,aorist_by_tomb,fill="9th Century Tombs"))+
-  geom_col(data=aorist_by_tomb,aes(decade,aorist_by_tomb, fill="All Tombs"),alpha=0.7)+
-    xlim(1100,600)+theme_minimal()+
-        theme(legend.title=element_blank(),
-          text=element_text(family="Garamond", size=12))+
-              labs(x="Date (Years BC)", y="Aoristic Density/Number of Tombs Represented")+
-                 scale_fill_brewer(palette = "Set2")
-
-names(tombs)
-tombs_in_use<-colSums(tombs[6:15]!="")
-tombs_in_use<-data.frame(tombs_in_use)
-colnames(tombs_in_use)<-c("no_tombs")
-tombs_in_use<-data.frame(tombs_in_use)
-setDT(tombs_in_use, keep.rownames = "period")
-tombs_in_use$period<-factor(tombs_in_use$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
-
-ggplot(no_tombs_9th,aes(decade,no_tombs, fill="Tombs in Use"))+geom_col()+theme_classic()+
-     theme(legend.title=element_blank(),
-          text=element_text(family="Garamond", size=12))+
-            xlim(1100,600)+
-              labs(x="Date (Years BC)", y="Number of Tombs in Use")+
-                 scale_fill_brewer(palette = "Set2")
-
-ggplot(tombs_in_use,aes(period,no_tombs, fill="Tombs in Use"))+geom_col()+theme_classic()+
-        theme(legend.title=element_blank(),
-          text=element_text(family="Garamond", size=12))+
-              labs(x="Period", y="Number of Tombs in Use")+
-                 scale_fill_brewer(palette = "Set2")
-
-ggplot(tombs,aes(start_period, fill=tomb_type_condensed))+geom_bar()+theme_classic()+
-        theme(legend.title=element_blank(),
-          text=element_text(family="Garamond", size=12))+
-              labs(x="Period", y="Number of New Tombs")+
-                 scale_fill_brewer(palette = "Set2")
-
-```
-
-```{r rates of burials}
-
-tombs_ranked<-tombs[with(tombs, order(-tombs$total_pottery)), ]
-tombs_top_12<-head(tombs_ranked,12)
-
-View(tombs_top_12)
-
-no_tombs_top<-colSums(tombs_top_12[158:207]!=0)
-
-no_tombs_top<-data.frame(no_tombs_top)
-colnames(no_tombs_top)<-c("no_tombs")
-row.names(no_tombs_top)<-dates
-no_tombs_top<-data.frame(no_tombs_top)
-setDT(no_tombs_top, keep.rownames = "decade")
-no_tombs_top$decade<-as.numeric(no_tombs_top$decade)
-
-aorist_tombs_top<-colSums(tombs_top_12[158:207])
-aorist_tombs_top<-data.frame(aorist_tombs_top)
-colnames(aorist_tombs_top)<-c("aoristic_sum")
-row.names(aorist_tombs_top)<-dates
-aorist_tombs_top<-data.frame(aorist_tombs_top)
-setDT(aorist_tombs_top, keep.rownames = "decade")
-aorist_tombs_top$decade<-as.numeric(aorist_tombs_top$decade)
-
-aorist_by_tomb_top<-merge(aorist_tombs_top,no_tombs_top)
-aorist_by_tomb_top$aorist_by_tomb<-aorist_by_tomb_top$aoristic_sum/aorist_by_tomb_top$no_tombs
-
-ggplot()+geom_col(data=aorist_by_tomb_top,aes(decade,aorist_by_tomb,fill="12 Largest Tombs"))+
-  geom_col(data=aorist_by_tomb,aes(decade,aorist_by_tomb, fill="All Tombs"),alpha=0.7)+
-    xlim(1100,600)+theme_minimal()+
-        theme(legend.title=element_blank(),
-          text=element_text(family="Garamond", size=12))+
-              labs(x="Date (Years BC)", y="Aoristic Density/Number of Tombs Represented")+
-                 scale_fill_brewer(palette = "Set2")
-
-```
-
-```{r top_tombs, fig.dim = c(7, 5), dpi=300, fig.align='center',}
-
-cols<-c(1,158:207)
-aorist_all_tomb_names<-tombs_top_12[,cols]
-aorist_all_tomb_names<-data.frame(aorist_all_tomb_names)
-colnames(aorist_all_tomb_names)<-c("tomb",dates)
-
-aorist_all_tomb_melt<-melt(aorist_all_tomb_names)
-str(aorist_all_tomb_melt)
-aorist_all_tomb_melt$variable<-as.numeric(as.character(aorist_all_tomb_melt$variable))
-
-top_12<-tombs_top_12$tomb
-aorist_all_tomb_melt$tomb<-factor(aorist_all_tomb_melt$tomb,levels=top_12)
-
-ggplot(aorist_all_tomb_melt,aes(variable, value))+geom_col()+theme(legend.position = "none")+facet_wrap(~tomb)+
-  geom_col()+theme_bw()+theme(legend.title=element_blank(),
-      text=element_text(family="Garamond", size=12))+
-          labs(x="Date (Years BC)", y="Aoristic Sum")+
-              scale_fill_brewer(palette = "Set3")+ 
-                theme(legend.position="none")+
-                  coord_cartesian(ylim=c(0,30),xlim=c(1100,610))+scale_y_continuous(expand=c(0,0))+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1100,610))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
 

### 2019-05-17 analysis/paper/paper.Rmd
Analysis first draft

@@
 
-```{r }
+```{r table1}
 
@@
 
-```{r top_tombs, fig.dim = c(7, 5), dpi=300, fig.align='center',fig.cap="Aoristic sums for the 12 largest tombs (in terms of grave goods), arranged in decreasing order of total recovered artefacts.
+```{r toptombs, fig.dim = c(7, 5), dpi=300, fig.align='center',fig.cap="Aoristic sums for the 12 largest tombs (in terms of grave goods), arranged in decreasing order of total recovered artefacts.
 Note, for the sake of better illustrating the relative patterns, rather than absolute values, the scales on each plot are distinct."}
 
+tombs_ranked<-tombs[with(tombs, order(-tombs$total_pottery)), ]
+tombs_top_12<-head(tombs_ranked,12)
+top12<-as.vector(tombs_top_12$tomb)
+pots_top_12<-pots[is.element(pots$tomb, top12),]
+pots_top_12$tomb<-factor(pots_top_12$tomb,levels=top12)
+
 cols<-c(1,158:207)
-aorist_all_tomb_names<-tombs_top_20[,cols]
+aorist_all_tomb_names<-tombs_top_12[,cols]
 aorist_all_tomb_names<-data.frame(aorist_all_tomb_names)
@@
 
-top_20<-tombs_top_20$tomb
-aorist_all_tomb_melt$tomb<-factor(aorist_all_tomb_melt$tomb,levels=top_20)
+top_12<-tombs_top_12$tomb
+aorist_all_tomb_melt$tomb<-factor(aorist_all_tomb_melt$tomb,levels=top_12)
 
@@
 
-```
-
-From Figure \@ref(fig:top_tombs), we can see that, though seven of the 12 did exhibit late, intense episodes of deposition, tombs X, G, Q, 104 and 283 evince a different pattern, with their respective peaks coming in the PGB and EG periods, around a century earlier.
-We have seen already in Figure \@ref(fig:aoristbycem) that, particularly in the KNC, the end of the 9th century was another period of increased burial activity
-
-```{r boxplot,  fig.dim=c(7,4), dpi=300,}
-
-tombs_top_20<-head(tombs_ranked,20)
-top20<-as.vector(tombs_top_20$tomb)
-pots_top_20$tomb<-factor(pots_top_20$tomb,levels=top20)
-
-pots_top_20<-pots[is.element(pots$tomb, top20),]
-pots_top_20_9th_7th<-subset(pots_top_20,century=="9th"|century=="8th"|century=="7th")
-ggplot(pots_top_20_9th_7th,aes(tomb,fill=tomb_century))+geom_bar()+facet_wrap(~century)
-
-tombs$group<-ifelse(tombs$tomb=="X"|tombs$tomb=="G"|tombs$tomb=="Q"|tombs$tomb=="104"|tombs$tomb=="283","A",
-                   ifelse(tombs$tomb=="P"|tombs$tomb=="292"|tombs$tomb=="75"|tombs$tomb=="107"|tombs$tomb=="285"|tombs$tomb=="II"|tombs$tomb=="218","B","OTHER"))
-names(tombs)
-ggplot(tombs,aes(antique,fill=group))+geom_histogram()
+tombs$group<-ifelse(tombs$tomb=="P"|tombs$tomb=="292"|tombs$tomb=="75"|tombs$tomb=="107"|tombs$tomb=="285"|tombs$tomb=="II"|tombs$tomb=="218","A","B")
+total_pottery_by_groups<-aggregate(tombs$total_pottery, by=list(tombs$group), FUN=sum)
+calc1<-(total_pottery_by_groups[1,2]/sum(total_pottery_by_groups$x))*100
+
+```
+
+In Figure \@ref(fig:toptombs), we see that, seven of the 12 did exhibit late, intense episodes of deposition.
+Tombs X, G, Q, 104 and 283 do, however, evince a different pattern, with their respective peaks coming in the PGB and EG periods, around a century earlier.
+We have seen already in Figure \@ref(fig:aoristbycem) that, particularly in the KNC, the end of the 9th century was another period of increased burial activity, and these tombs shall be considered again shortly.
+But the other seven do appear to underpin the late 8th to early 7th century spate of burials, and between them accrount for `r round(calc1,digits=2)`% of the total pottery assemblage recovered from the `r nrow(tombs)` tombs considered in the present analysis.
+The question arises then, as to whether these tombs were in any way exceptional at the time of their construction, and it is to this possibility which we now turn.
+
+##Tomb characteristics and the composition of the burial assemblages
+
+The indices discussed so far – the distribution of of artefacts between the tombs, and the construction and use of the tombs through time – only inform us of the temporal or aggregate features of the cemeteries; they tell us little of the tombs as they were built or first used.
+Synchronic comparisons are harder to make with the available data, but the dimensions of the tombs offer some scope for doing so.
+
+```{r tombdimensions, fig.dim = c(7.5, 4), dpi=300, fig.align='center', fig.cap="Scatterplots "}
+
+tombareas<-ggplot(subset(tombs,tombs$est_burials>1),aes(chamber_area,est_burials,color=cem_condensed))+geom_point()+
+  theme_bw()+theme(legend.title=element_blank(),
+        text=element_text(family="Garamond", size=12))+
+            labs(x="Chamber Area (m2)", y="Estimated Total Burials")+
+               scale_color_brewer(palette = "Set2")+
+                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+dromoslengths<-ggplot(subset(tombs,tombs$est_burials>1),aes(dromos_length,est_burials,color=cem_condensed))+geom_point()+
+  theme_bw()+theme(legend.title=element_blank(),
+        text=element_text(family="Garamond", size=12))+
+            labs(x="Dromos Length (m)", y="")+
+               scale_color_brewer(palette = "Set2")+
+                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+ggarrange(tombareas,dromoslengths,ncol=2,common.legend = TRUE,legend="bottom")
+
+```
+
+Looking at , however, it seems that no clear relationship existed between the physical size of a tomb and its buried population at EIA Knossos.
+It is true that over half of all tombs do not exceed 2.6m2 in area, while nine of the ten tombs with the most burials do surpass this value.
+But many of the very largest tombs are decidedly middling when it comes to burial numbers.
+Things are even less clear for the dromos, the most visible part of the tomb, and a plausible arena for demonstrations of scale and grandeur during the burial itself.
+Admittedly, the very largest two tombs by burial numbers also had exceptionally long dromoi, but the next eight tombs with the largest dromoi do not rank highly when it comes to number of interments.
+There are difficulties here in that the dromoi especially were common victims of encroachment by later tombs, and our estimates of their original length are tentative in many cases.
+
+```{r dimensions boxplots}
+
+tombs$century_built<-factor(tombs$century_built,levels=c("11th","10th","9th","8th","7th"))
+dromwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_width,fill=century_built))+geom_boxplot()+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Century of Tomb Construction", y="Dromos Width")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0.1))
+dromlength<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_length,fill=century_built))+geom_boxplot()+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Century of Tomb Construction", y="Dromos Length")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0.5))
+stomwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,stomion_width,fill=century_built))+geom_boxplot()+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Century of Tomb Construction", y="Stomion Width")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0.1))
+chamarea<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,chamber_area,fill=century_built))+geom_boxplot()+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Century of Tomb Construction", y="Chamber Area")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0.5))
+
+ggarrange(dromwidth,dromlength,stomwidth,chamarea,nrow=2,ncol=2)
+
+```
+
+Looking at ...
+we can see that, through the centuries, the dimensions of the tombs varied, sometimes consistently, as with the trend toward narrower **dromoi**.
+Tombs of the 9th century do not particularly stand out here, though it is worth noting that this period did witness the peak in the size of the actual chambers of tombs, as well as being the period in which these sizes varied most.
+But, then again, even the largest chambers are not exactly cavernous, and the extra labour expended on the construction of the larger chambers would have been limited.
+The dromos is a plausible area for conspicuous demonstrations of scale but, with their frequently poor preservation, we cannot be confident in assessing this.
+It appears at any rate that they were gradually declining in size, albeit with some notable outliers.
+
+What these figures do suggest, though, is that if tomb size carried a social meaning at the time of construction, this proved irrelevant in the selection of certain tombs for extensive later reuse.
+That is, the tomb’s later popularity seems not have been foreseen.
+Many large tombs had niches cut in their dromoi, or the stomion extended, to fit in later burial urns.
+Sometimes, urns from within the tomb itself were placed in such niches to make space for new interments in the chamber itself.
+There is an impression of individuals vying to be included in the chamber itself.
+That extra space had to be made, and competition was fierce for inclusion into the sepulchral group, supports the idea that the tombs’ original builders little anticipated their eventual popularity and so did not scale their tombs accordingly.
+
+Though it may not be immediately apparent from the dimensions of the tombs, however, the late 9th and early 8th centuries are consistently interpreted as periods of artistic development, increased overseas trade, and renewed vigour in the construction of tombs and the visibility of the burial rite.
+The PGB style, first identified by Brock, purportedly reflects a uniquely Cretan (perhaps even Knossian) melding of Attic and Near Eastern forms and motifs, while the appearance of Bronze Age 'antiques' in the tombs from this period onwards is felt to reflect renewed, even competitive interest in the Minoan past, as a tool for mortuary display or even lineage affirmation.
+We can
+
+```{r}
+
+imports<-subset(pots,pots$import!="")
+greekimports<-subset(imports,imports$import_region=="Greece")
+NEimports<-subset(imports,imports$import_region=="Near East")
+
+import_aorist<-colSums(imports[27:76])
+import_aorist<-data.frame(import_aorist)
+colnames(import_aorist)<-c("aoristic_sum")
+row.names(import_aorist)<-dates
+setDT(import_aorist, keep.rownames = "decade")
+import_aorist$decade<-as.numeric(import_aorist$decade)
+
+greekimport_aorist<-colSums(greekimports[27:76])
+greekimport_aorist<-data.frame(greekimport_aorist)
+colnames(greekimport_aorist)<-c("aoristic_sum")
+row.names(greekimport_aorist)<-dates
+setDT(greekimport_aorist, keep.rownames = "decade")
+greekimport_aorist$decade<-as.numeric(greekimport_aorist$decade)
+
+NEimport_aorist<-colSums(NEimports[27:76])
+NEimport_aorist<-data.frame(NEimport_aorist)
+colnames(NEimport_aorist)<-c("aoristic_sum")
+row.names(NEimport_aorist)<-dates
+setDT(NEimport_aorist, keep.rownames = "decade")
+NEimport_aorist$decade<-as.numeric(NEimport_aorist$decade)
+
+greekimport_aorist$origin<-"Greece"
+NEimport_aorist$origin<-"Near East"
+allimport_aorist<-rbind(greekimport_aorist,NEimport_aorist)
+View(allimport_aorist)
+
+ggplot()+geom_col(data=import_aorist,aes(decade,aoristic_sum),alpha=0.2)+
+          geom_col(data=allimport_aorist,aes(decade,aoristic_sum, fill=origin),position="dodge")+
+            theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                labs(x="Date (Years BC)", y="Aoristic Sum")+
+                  scale_fill_brewer(palette = "Set2")+
+                   scale_y_continuous(expand=c(0,0),limits=c(0,20))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+```
+
+The rise in imported items is clearly reflected in the tomb assemblages, with a succession of increased Greek and Near Eastern imports respectively in the late 9th and early 8th centuries.
+Yet whilst confirming some of the prior inferences about the LPG-EG periods, the deposition of these imported wares can also cast light on the original pattern drawn out in the present analysis: the late surge in burial activity among some select 9th and 10th century tombs.
+
+```{r importstombcenturies, fig.dim=c(8,4),dpi=300}
+
+imports<-subset(pots,import!="")
+
+import_aorist<-colSums(imports[27:76])
+import_aorist<-data.frame(import_aorist)
+colnames(import_aorist)<-c("aoristic_sum")
+row.names(import_aorist)<-dates
+setDT(import_aorist, keep.rownames = "decade")
+import_aorist$decade<-as.numeric(import_aorist$decade)
+
+imports_11th<-subset(imports,tomb_century=="11th")
+imports_10th<-subset(imports,tomb_century=="10th")
+imports_9th<-subset(imports,tomb_century=="9th")
+imports_8th<-subset(imports,tomb_century=="8th")
+imports_7th<-subset(imports,tomb_century=="7th")
+
+import_11th_aorist<-colSums(imports_11th[27:76])
+import_11th_aorist<-data.frame(import_11th_aorist)
+colnames(import_11th_aorist)<-c("aoristic_sum")
+row.names(import_11th_aorist)<-dates
+setDT(import_11th_aorist, keep.rownames = "decade")
+import_11th_aorist$decade<-as.numeric(import_11th_aorist$decade)
+
+import_11th_aorist$century<-"11th"
+
+import_10th_aorist<-colSums(imports_10th[27:76])
+import_10th_aorist<-data.frame(import_10th_aorist)
+colnames(import_10th_aorist)<-c("aoristic_sum")
+row.names(import_10th_aorist)<-dates
+setDT(import_10th_aorist, keep.rownames = "decade")
+import_10th_aorist$decade<-as.numeric(import_10th_aorist$decade)
+
+import_10th_aorist$century<-"10th"
+
+import_9th_aorist<-colSums(imports_9th[27:76])
+import_9th_aorist<-data.frame(import_9th_aorist)
+colnames(import_9th_aorist)<-c("aoristic_sum")
+row.names(import_9th_aorist)<-dates
+setDT(import_9th_aorist, keep.rownames = "decade")
+import_9th_aorist$decade<-as.numeric(import_9th_aorist$decade)
+
+import_9th_aorist$century<-"9th"
+
+import_8th_aorist<-colSums(imports_8th[27:76])
+import_8th_aorist<-data.frame(import_8th_aorist)
+colnames(import_8th_aorist)<-c("aoristic_sum")
+row.names(import_8th_aorist)<-dates
+setDT(import_8th_aorist, keep.rownames = "decade")
+import_8th_aorist$decade<-as.numeric(import_8th_aorist$decade)
+
+import_8th_aorist$century<-"8th"
+
+import_7th_aorist<-colSums(imports_7th[27:76])
+import_7th_aorist<-data.frame(import_7th_aorist)
+colnames(import_7th_aorist)<-c("aoristic_sum")
+row.names(import_7th_aorist)<-dates
+setDT(import_7th_aorist, keep.rownames = "decade")
+import_7th_aorist$decade<-as.numeric(import_7th_aorist$decade)
+
+import_7th_aorist$century<-"7th"
+
+all_import_aorist<-rbind(import_11th_aorist,import_10th_aorist,import_9th_aorist,import_8th_aorist,import_7th_aorist)
+all_import_aorist$century<-factor(all_import_aorist$century,levels=c("11th","10th","9th","8th","7th"))
+
+ggplot()+ geom_col(data=subset(all_import_aorist,century!="7th" & century!="8th"),aes(decade,aoristic_sum,fill=century),position="dodge")+
+  scale_x_reverse(limits=c(1000,630))+scale_fill_brewer(palette="Set2")+
+  theme_bw()+theme(legend.position="bottom",text=element_text(family="Garamond", size=12))+
+            labs(x="Date (Years BC)", y="Aoristic Sum")+
+  guides(fill=guide_legend(title="Tomb Century"))+
+              scale_fill_brewer(palette = "Set2")+ 
+                  scale_y_continuous(expand=c(0,0),limits=c(0,12))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+```
+
+Leaving aside the later tombs for a moment, we can see in Figure ...
+that the tombs already built by the end of the PGB period, particularly those built in the 9th and, to a lesser extent, 10th centuries, were major beneficiaries of the influx of imported items during this time, but that, in the following centuries, in line with the rest of the cemetery, the number of imported items deposited within them declined.
+Imported ceramics are used here as something of a proxy for items of status value, a shorthand that would ideally be bolstered by other artefacts (such as precious vessels, weapons, jewellery etc.) which, owing to their less developed chronologies, must unfortunately be left out of such diachronic analyses.
+But the beginnings of my final argument are seen here in Figure ..., namely that the late surge in burial activity seen across both cemeteries in the late 8th and early 7th centuries would appear to be of a different kind to that of a century earlier.
+
+```{r}
 
@@
                   scale_y_continuous(expand=c(0,0),limits=c(0,20))
-names(tombs)
-tombs$century_built<-factor(tombs$century_built,levels=c("11th","10th","9th","8th","7th"))
-ggplot(tombs,aes(start_period,dromos_length))+geom_boxplot()
-ggplot(data=subset(pots,pots$import=="Y"),aes(period))+geom_histogram()+xlim(1100,630)
-```
-
-```{r dpi=300}
+
+tombP<-subset(tombs,tombs$tomb=="P")
+tombP$X650
+```
+
+Figure ...
+is a boxplot for the aoristic sums of each decade across both cemeteries, colour-coded by ceramic phase.
+Note that, for clarity, tomb G in the EG period, and tomb P in the late EO-LO period are excluded, with values ranging from 29-34, and 36-44 respectively.
+The important observation is that the PGB and EG exhibit the largest interquartile ranges (IQRs) of any period, and each possess only a single outlier (beyond 1.5 times the IQR).
+The Orientalising period, on the other hand, exhibit much reduced IQRs, and a proliferation of outliers far and above the range of the other tombs.
+In other words, while both the PGB-EG and EO-LO periods witnessed an expanding volume and range of burial activity, in the earlier period this was shared somewhat more equally among the total burying population.
+One speaks of increasing plurality, the other of divergence.
+
+```{r gini}
+
 aorist_tomb_values<-tombs[c(1,158:207)]
@@
                  scale_fill_brewer(palette = "Set3")+ coord_cartesian(
-                    xlim = c(1100, 600), ylim = c(0.5,0.8))+ scale_x_reverse(expand = c(0, 0),limits=c(1100,600))
-
-```
-
-```{r cumfreq}
-
-```
-
-# Results
-
-```{r get-data, eval = FALSE}
-# Note the path that we need to use to access our data files when rendering this document
-my_data <- readr::read_csv(here::here('analysis', 'data', 'raw_data', 'my_csv_file.csv'))
-
-knitr::kable(head(iris),
-             caption="A table of flowers")
-```
+                    xlim = c(1100, 600), ylim = c(0.5,0.8))+ scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
+                       scale_y_continuous(expand=c(0,0))
+```
+
+This possibility is furthered somewhat by considering the gini coefficient of the distribution of pottery between the tombs.
+Using the aoristic sums as a probablistically-weighted estimate of the volume of material deposited in each tomb in each decade, use of the gini coefficient can give an (albeit rough) estimate of how unevenly distributed that material was.
+And what we see is that, though values were always relatively high (i.e.
+unequal) within the cemeteries, the major spike in the gini coefficient seen in the Orientalising period contrasts markedly with the modest rise and plateau of the PGB-EG.
+The gini coefficient is most commonly used in the assessment of income inequality, but I want to stress here that it is not "wealth" in any meaningful sense that we are talking about here.
+Indeed, if anything, the late flourit of burial activity here discussed may be a case of quantity, and not quality.
+
+```{r}
+pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A","B")
+
+potcompKNC<-ggplot(subset(pots,century!="" & cem_condensed=="NC"),aes(century,fill=type))+geom_bar(position="fill")+
+  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title = element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(title="KNC",x="", y="Proportion of Total Assemblage")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" Amphora  "," Jug  "," Oil vessel  ", " Pithos"," Drinking vessel  "," Lid  "," Other"))+ 
+                    scale_y_continuous(expand=c(0,0.0))
+potcompFOR<-ggplot(subset(pots,century!="" & cem_condensed=="FOR"),aes(century,fill=type))+geom_bar(position="fill")+
+  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(title="Fortetsa",x="Century BC", y="")+
+                  scale_fill_brewer(palette = "Set2", labels=c(" Amphora  "," Jug  "," Oil vessel  ", " Pithos"," Drinking vessel  "," Lid  "," Other"))+ 
+                    scale_y_continuous(expand=c(0,0.0))
+
+potcomptop7<-ggplot(subset(pots,century!="" & century!="11th"& group=="A"),aes(century,fill=type))+geom_bar(position="fill")+
+  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(title="P, 292, 75, 107, 285, II & 218",x="", y="")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" Amphora  "," Jug  "," Oil vessel  ", " Pithos"," Drinking vessel  "," Lid  "," Other"))+ 
+                    scale_y_continuous(expand=c(0,0.0))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+ggarrange(potcompKNC,potcompFOR,potcomptop7,ncol=3,common.legend =TRUE,legend="bottom")
+```
+
+Figure ...
+shows the composition of the total pottery assemblage from each cemetery, and from the seven tombs drawn out earlier, through the centuries.
+In all cases, there appears to be the steady crystallisation of a basic 'burial kit', comprising a pithos and lid, an oil vessel, a jug, and a drinking vessel.
+The proportion of grave goods comprising 'other' vessels - pyxides, askoi, trays, stands, dinoi, kernoi, kraters, lekanai and more - declines to become only a minor part of the standardised repertoire of the 7th century.
+The seven tombs that underpinned the dramatic increase in burial activity in the later 8th century did not break with this mould, and their assemblages seem as conventional as any other during the last century of use in the cemeteries.
+
+It has no doubt been apparent that the main focus of this article has been on pottery and not other artefacts of metal, stone, glass, and organic materials.
+This is largely because the chronologies for these artefacts are insufficiently developed to allow for the kind of diachronic analysis made thus far.
+But I do want to briefly comment on one aspect of the distribution of non-ceramic artefacts that may make more sense in the light of the foregoing discussion.
+There is, perhaps unsurprisingly, a relatively strong (Spearman's rank correlation of `r spearmanobjectspots`) correlation between the number of ceramic and non-ceramic finds across the tombs, and, certainly, many of the most unique and precious objects were recovered from among the larger tombs.
+But, the monopoly those tombs would appear to have over such objects, perhaps tempting one initially to speak of 'elite' or 'high-status' tombs, evaporates when the number of those objects is divided by the number of burials represented.
+
+```{r }
+tombs$group<-factor(tombs$group,levels=c("A","B"))
+total_object<-ggplot(subset(tombs,est_burials>1),aes(total_object, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Total Non-Ceramic Finds", y="Number of Tombs")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+total_metal<-ggplot(subset(tombs,est_burials>1),aes(weapon, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Total Metal Artefacts", y="Number of Tombs")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+total_import<-ggplot(subset(tombs,est_burials>1),aes(total_import, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Total Imports", y="Number of Tombs")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+
+objectsbyburial<-ggplot(subset(tombs,est_burials>1),aes(total_object/est_burials, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Total Non-Ceramics/Burials", y="Number of Tombs")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+metalbyburial<-ggplot(subset(tombs,est_burials>1),aes(weapon/est_burials, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Total Metal Artefacts/Burials", y="Number of Tombs")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+
+importsbyburial<-ggplot(subset(tombs,est_burials>1),aes(total_import/est_burials, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Total Imports/Burials", y="Number of Tombs")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+
+ggarrange(total_import,total_weapon,total_object,importsbyburial,weaponbyburial,objectsbyburial,ncol=3,nrow=2,common.legend = TRUE,legend="bottom")
+```
+
+Figure ...
+thus supports the view which has emerged from the preceeding analysis, that the major increase in burial activity at both Fortetsa and the KNC was not a phenomenon linked in any straightforward way to simple principles of 'elite' or 'wealthy' burial.
+To summarise, towards the end of 8th century BC, a select group of tombs, all of which had been constructed back in the 9th or 10th, became the foci of a major increase in burials, something which did not occur among the tombs built around the same time.
+We have seen how this surge in acivity occurred against the backdrop of a declining number of tombs dating back to the 10th and 9th centuries, and I have suggested that the rates of change are too rapid, and unevenly distributed, to be explicable through population growth alone.
+We have seen how little about the physical dimensions of these tombs presaged their later popularity, and that, while a similar, if smaller, increase in burial activity occurred around a century earlier, in the PGB-EG periods, this contrasts in the comparative diversity of its associated assemblages, and their distribution across the cemetries.
+The later burials were marked by great uniformity in the associated grave goods, and this apparently limited and consistent repertoire may explain why many of the most used tombs prove exceptional only in the absolute quantity of artefacts, and not the accoutrements of each individual burial.
 

### 2019-05-20 analysis/paper/paper.Rmd
First draft analysis

@@
 
-```{r dimensions boxplots}
+```{r dimensions boxplots, fig.dim=c(8,4),dpi=300}
 
@@
 
-```{r}
+```{r imports, fig.dim=c(6,4),dpi=300}
 
@@
 
-```{r}
+```{r bigbox, fig.dim=c(8,4),dpi=300}
 
@@
 
-```{r gini}
+```{r gini, fig.dim=c(7,4),dpi=300}
 
@@
 
-```{r}
+```{r comps, fig.dim=c(8,4),dpi=300}
 pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A","B")
@@
 ggarrange(potcompKNC,potcompFOR,potcomptop7,ncol=3,common.legend =TRUE,legend="bottom")
+
+spearmanobjectspots<-cor(tombs$total_object,tombs$total_pottery,method="spearman")
 ```
@@
 
-```{r }
+```{r hists, fig.dim=c(8,4),dpi=300}
 tombs$group<-factor(tombs$group,levels=c("A","B"))
@@
 
-ggarrange(total_import,total_weapon,total_object,importsbyburial,weaponbyburial,objectsbyburial,ncol=3,nrow=2,common.legend = TRUE,legend="bottom")
+ggarrange(total_import,total_metal,total_object,importsbyburial,metalbyburial,objectsbyburial,ncol=3,nrow=2,common.legend = TRUE,legend="bottom")
 ```

### 2019-05-20 analysis/paper/paper.Rmd
Figs complete and csl edited

@@
 title: "Family Fortunes: A Quantitative Analysis of the Early Iron Age Cemeteries at Knossos, Crete"
+subtitle: "Early Iron Age Cemeteries at Knossos"
 author:
@@
 csl: "../templates/world-archaeology.csl" # Insert path for the bib-style
+
 abstract: |
-  Three principal interests converge in this paper: reorienting the focus of Early Iron Age Cretan archaeology; the application of computational, and quantitative methods to supplement traditional modes of inference; and open-access data and reproducible research.
-The Early Iron Age cemeteries of Knossos, Crete, represent the richest source of archaeological material at the site for the whole period between the end of the Bronze Age and the emergence of the Archaic *polis*.
-However, despite well-published excavation reports on the two largest burial grounds, the Fortetsa and Knossos North cemeteries, and a number of studies dealing with various artefact classes and individual tombs therein, no substantial comparative, quantitative analysis of these two cemeteries has yet been undertaken.
-This paper provides such an analysis, examining both the synchronic and diachronic variation among the tombs in these two cemeteries.
-Through the application of a range of quantitative techniques, this study isolates a pattern of increased rates of burial among a few select tombs, and further considers their characteristics.
-This paper demonstrates the potential for quantitative analyses to contribute to nuanced, context-specific theories of social practice, and advocates their wider application to the archaeological record of the Cretan Early Iron Age.
+  The Early Iron Age cemeteries of Knossos, Crete, represent the richest source of archaeological material at the site between the end of the Bronze Age and the emergence of the Archaic *polis*.
+However, despite well-published excavation reports on the two largest burial grounds and a number of studies dealing with various artefacts and individual tombs therein, no substantial comparative, quantitative analysis has yet been undertaken.
+This paper does just that, and in so doing identifies, and attempts to explain, a pattern of increased burial rates among a few select tombs.
+This paper demonstrates the potential for quantitative analyses to contribute towards nuanced, context-specific theories of social practice, and advocates their wider application to the archaeological record of the Cretan Early Iron Age.
 It also serves as a demonstration of maximal research reproducibility, with the full datasets and code underpinning the analysis made freely available to download and use.
@@
 ---
-
-<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
 
@@
 library(ggpubr)
-library(cowplot)
-library(gridExtra)
 library(ineq)
-library(ggrepel)
 library(reshape)
@@
 # Introduction
-It may be felt that, of all ancient sites on Crete, Knossos is the last in need of further archaeological attention.
+Three principal interests converge in this paper: reorienting the focus of Early Iron Age Cretan archaeology; the application of computational, and quantitative methods to supplement traditional modes of inference; and open-access data and reproducible research.It may be felt that, of all ancient sites on Crete, Knossos is the last in need of further archaeological attention.
 The city and ‘palace’ of Knossos, first excavated well over a century ago [@Evans1921; -@Evans1928; -@Evans1930], retain a pre-eminent position in popular and academic accounts the island’s Bronze Age, and few, if any, other Cretan sites have been so thoroughly dug, reconstructed, and contested.
@@
       geom_line(aes(est_burials,est_burials),alpha=0.4)+
-       xlim(0,60)+ylim(0,60)+theme_minimal()+
+       scale_x_continuous(expand=c(0,0),limits=c(0,55))+
+        scale_y_continuous(expand=c(0,0),limits=c(0,55))+theme_bw()+
          theme(legend.title=element_blank(),
@@
       text=element_text(family="Garamond", size=12))+
-          labs(x="Date (Years BC)", y="Total Vessel Counts")+
+          labs(x="Date (Years BC)", y="Vessel Count")+
               scale_fill_brewer(palette = "Set3")+ 
@@
 
-```{r tombconstruction, fig.dim=c(8,4), dpi=300, fig.cap="Visulations of tomb construction through time at both cemeteries.
+```{r tombconstruction, fig.dim=c(7.5,4), dpi=300, fig.cap="Visulations of tomb construction through time at both cemeteries.
 On the left are counts based on the earliest dated pottery in each tomb, while on the left are aoristic sums likewise based on the earliest ceramic-based assignations.
@@
             geom_col(data=tomb_use,aes(period,no_tombs,fill=tomb_century),position="dodge")+
-              theme_bw()+theme(legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
+              theme_bw()+theme(text=element_text(family="Garamond", size=12))+
                  labs(x="Period", y="Estimated No. of Tombs in Use")+
-                  scale_fill_brewer(palette = "Set2")+ 
+                  scale_fill_brewer(name="Century 
+Built",palette = "Set2")+ 
                     scale_y_continuous(expand=c(0,0),limits=c(0,60))
@@
 
-```{r potsbydateandtombdate,  fig.dim=c(8,4), dpi=300, fig.cap="Counts of vessels from across all tombs in both cemeteries, arranged, on the left, by the century to which they are assigned based on their ceramic period and, on the right, by the centuries in which the tombs they come from were built."}
+```{r potsbydateandtombdate,  fig.dim=c(7,4), dpi=300, fig.cap="Counts of vessels from across all tombs in both cemeteries, arranged, on the left, by the century to which they are assigned based on their ceramic period and, on the right, by the centuries in which the tombs they come from were built."}
 
@@
                text=element_text(family="Garamond", size=12))+
-                 labs(x="Century (Date of Vessel)", y="Vessel Count")+
+                 labs(x="Century BC (Date of Vessel)", y="Vessel Count")+
                   scale_fill_brewer(palette = "Set2")+ 
@@
 
-date_by_tomb<-ggplot(pots,aes(tomb_century,fill="Date"))+geom_bar()+
+date_by_tomb<-ggplot(data=subset(pots,pots$century!=""),aes(tomb_century,fill="Date"))+geom_bar()+
   theme_bw()+theme(legend.position="none",
                text=element_text(family="Garamond", size=12))+
-                 labs(x="Century (Date of Tomb)", y="")+
+                 labs(x="Century BC (Date of Tomb)", y="Vessel Count")+
                   scale_fill_brewer(palette = "Set2")+ 
@@
 ```
+
+##### pagebreak
 
@@
 
-```{r tombdimensions, fig.dim = c(7.5, 4), dpi=300, fig.align='center', fig.cap="Scatterplots "}
+```{r tombdimensions, fig.dim = c(7.5, 4), dpi=300, fig.align='center', fig.cap="Scatterplots comparing tomb dimensions (chamber area and *dromos* length) to estimated total burials by tomb."}
 
@@
         text=element_text(family="Garamond", size=12))+
-            labs(x="Chamber Area (m2)", y="Estimated Total Burials")+
-               scale_color_brewer(palette = "Set2")+
+            labs(x=bquote('Chamber Area ('*m^2*')'), y="Estimated Total Burials")+
+               scale_color_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+
                 scale_x_continuous(expand=c(0,0),limits=c(0,10))+
@@
         text=element_text(family="Garamond", size=12))+
-            labs(x="Dromos Length (m)", y="")+
-               scale_color_brewer(palette = "Set2")+
+            labs(x="Dromos Length (m)", y="Estimated Total Burials")+
+               scale_color_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+
                 scale_x_continuous(expand=c(0,0),limits=c(0,10))+
@@
 
-```{r dimensions boxplots, fig.dim=c(8,4),dpi=300}
+```{r dimensionsboxplots, fig.dim=c(6,5),dpi=300, fig.align='center',fig.cap="Boxplots comparing various tomb dimensions as they vary by the century of tomb construction."}
 
@@
                text=element_text(family="Garamond", size=12))+
-                 labs(x="Century of Tomb Construction", y="Dromos Width")+
+                 labs(x="Century of Tomb Construction", y="Dromos Width (m)")+
                   scale_fill_brewer(palette = "Set2")+ 
@@
                text=element_text(family="Garamond", size=12))+
-                 labs(x="Century of Tomb Construction", y="Dromos Length")+
+                 labs(x="Century of Tomb Construction", y="Dromos Length (m)")+
                   scale_fill_brewer(palette = "Set2")+ 
@@
                text=element_text(family="Garamond", size=12))+
-                 labs(x="Century of Tomb Construction", y="Stomion Width")+
+                 labs(x="Century of Tomb Construction", y="Stomion Width (m)")+
                   scale_fill_brewer(palette = "Set2")+ 
@@
                text=element_text(family="Garamond", size=12))+
-                 labs(x="Century of Tomb Construction", y="Chamber Area")+
+                 labs(x="Century of Tomb Construction", y=bquote('Chamber Area ('*m^2*')'))+
                   scale_fill_brewer(palette = "Set2")+ 
@@
 
+```{r tombscompthroughtime, fig.dim=c(7,4),dpi=300,fig.align='center',fig.cap="Counts of vessels belonging to the 9th, 8th and 7th centuries, found in tombs dating to the 10th or 9th century which yielded upwards of 50 vessels when excavated.
+The seven large tombs with late peaks in burial activity identified earlier are shown in green."}
+
+pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A","B")
+tombs9th10th<-subset(tombs,century_built=="9th" & total_pottery>50|century_built=="10th" & total_pottery>50)
+pots9th10th<-subset(pots,tomb_century=="9th"|tomb_century=="10th")
+
+tombs9th10th_ranked<-tombs9th10th[with(tombs9th10th, order(-tombs9th10th$total_pottery)), ]
+tombs9th10th_ranked<-as.vector(tombs9th10th_ranked$tomb)
+pots9th10th$tomb<-factor(pots9th10th$tomb,levels=tombs9th10th_ranked)
+pots9th10th<-subset(pots9th10th,century=="9th" & tomb!="NA"|century=="8th" & tomb!="NA"|century=="7th" & tomb!="NA")
+pots9th10th$century<-mapvalues(pots9th10th$century, from = c("9th","8th","7th"), to = c("9th Century", "8th Century","7th Century"))
+
+ggplot(pots9th10th,aes(tomb,fill=group))+geom_bar()+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="Garamond", size=12),
+               axis.text.x = element_text(angle = 90,hjust=1,vjust=0.4))+
+                 labs(x="Tomb", y="Vessel Count")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0))+facet_grid(~century,scales="free_x",space="free")
+```
+
 Though it may not be immediately apparent from the dimensions of the tombs, however, the late 9th and early 8th centuries are consistently interpreted as periods of artistic development, increased overseas trade, and renewed vigour in the construction of tombs and the visibility of the burial rite.
@@
 
-```{r imports, fig.dim=c(6,4),dpi=300}
+```{r imports, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="Aoristic sums of imported ceramics through time across both cemeteries, colour-coded by the broad regional origin of the vessels."}
 
@@
 allimport_aorist<-rbind(greekimport_aorist,NEimport_aorist)
-View(allimport_aorist)
 
@@
           geom_col(data=allimport_aorist,aes(decade,aoristic_sum, fill=origin),position="dodge")+
-            theme_bw()+theme(legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
+            theme_bw()+theme(text=element_text(family="Garamond", size=12))+
                 labs(x="Date (Years BC)", y="Aoristic Sum")+
-                  scale_fill_brewer(palette = "Set2")+
+                  scale_fill_brewer(name="Origin",palette = "Set2")+
                    scale_y_continuous(expand=c(0,0),limits=c(0,20))+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))
 
@@
 
-```{r importstombcenturies, fig.dim=c(8,4),dpi=300}
+```{r importstombcenturies, fig.dim=c(8,4),dpi=300, fig.align='center',fig.cap="Aoristic sums of imported vessels through time, from tombs built in the 11th through 9th centuries."}
 
@@
             labs(x="Date (Years BC)", y="Aoristic Sum")+
-  guides(fill=guide_legend(title="Tomb Century"))+
+  guides(fill=guide_legend(title="Century of Tomb Construction"))+
               scale_fill_brewer(palette = "Set2")+ 
@@
 
-```{r bigbox, fig.dim=c(8,4),dpi=300}
+```{r bigbox, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="Boxplot for the aoristic sums of all vessels across both cemeteries through time, colour-coded by ceramic period.
+Boxes represent the interquartile range, horizontal lines the median value, whiskers values within 1.5 times the interquartile range, and points outliers beyond this."}
 
@@
 
-ggplot(aorist_melt,aes(variable,value,fill=period,outlier.size=0.3))+geom_boxplot(aes(group = cut_width(variable, 10)))+theme_bw()+
-      theme(legend.title=element_blank(),
-        text=element_text(family="Garamond", size=12),
-        axis.text.x = element_text(angle = 70, hjust = 1),
-        panel.grid.major.x = element_blank())+
+ggplot(aorist_melt,aes(variable,value,fill=period))+geom_boxplot(aes(group = cut_width(variable, 10)),outlier.size=0.8)+theme_bw()+
+      theme(text=element_text(family="Garamond", size=12))+
             labs(x="Date (Years BC)", y="Aoristic Sums per Tomb")+
-               scale_fill_brewer(palette = "Set3")+
+               scale_fill_brewer(name="Period",palette = "Set3")+
                 scale_x_reverse(expand=c(0,0))+
@@
 
-tombP<-subset(tombs,tombs$tomb=="P")
-tombP$X650
 ```
@@
 
-```{r gini, fig.dim=c(7,4),dpi=300}
+```{r gini, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="Gini coefficient values through time based on the aoristic sums for all vessels across both cemeteries.
+Background colours correspond to ceramic periods."}
 
@@
      geom_line(data = tombs_gini, aes(decade,gini_coefficient),size=0.5)+
-      theme_bw()+theme(legend.title=element_blank(),
-          text=element_text(family="Garamond", size=12))+
+      theme_bw()+theme(text=element_text(family="Garamond", size=12))+
               labs(x="Date (Years BC)", y="Gini Coefficient")+
-                 scale_fill_brewer(palette = "Set3")+ coord_cartesian(
+                 scale_fill_brewer(name="Period",palette = "Set3")+ coord_cartesian(
                     xlim = c(1100, 600), ylim = c(0.5,0.8))+ scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
@@
 
-```{r comps, fig.dim=c(8,4),dpi=300}
+```{r comps, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="Plots showing the composition of the total pottery assemblage for each century across both cemeteries, and among the 7 notable tombs drawn out earlier in the analysis."}
 pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A","B")
@@
                text=element_text(family="Garamond", size=12))+
-                 labs(title="P, 292, 75, 107, 285, II & 218",x="", y="")+
+                 labs(title="7 Tombs",x="", y="")+
                   scale_fill_brewer(palette = "Set2",labels=c(" Amphora  "," Jug  "," Oil vessel  ", " Pithos"," Drinking vessel  "," Lid  "," Other"))+ 
-                    scale_y_continuous(expand=c(0,0.0))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+                    scale_y_continuous(expand=c(0,0.0))
 
@@
 
-```{r hists, fig.dim=c(8,4),dpi=300}
+```{r hists, fig.dim=c(6,4),dpi=300, fig.align='center',fig.cap="Histograms relating, first, to counts of non-ceramic artefacts, metal objects, and imported items across all tombs in both cemeteries and, secondly, to those counts divided by the estimated number of burials in each tomb."}
 tombs$group<-factor(tombs$group,levels=c("A","B"))
-total_object<-ggplot(subset(tombs,est_burials>1),aes(total_object, fill=group))+geom_histogram()+
+total_object<-ggplot(subset(tombs,est_burials>1 & total_object>0),aes(total_object, fill=group))+geom_histogram()+
   theme_bw()+theme(legend.title=element_blank(),
@@
                     scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-total_metal<-ggplot(subset(tombs,est_burials>1),aes(weapon, fill=group))+geom_histogram()+
+total_metal<-ggplot(subset(tombs,est_burials>1 & weapon>0),aes(weapon, fill=group))+geom_histogram()+
   theme_bw()+theme(legend.title=element_blank(),
@@
                     scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-total_import<-ggplot(subset(tombs,est_burials>1),aes(total_import, fill=group))+geom_histogram()+
+total_import<-ggplot(subset(tombs,est_burials>1 & total_import>0),aes(total_import, fill=group))+geom_histogram()+
   theme_bw()+theme(legend.title=element_blank(),
@@
 
-objectsbyburial<-ggplot(subset(tombs,est_burials>1),aes(total_object/est_burials, fill=group))+geom_histogram()+
+objectsbyburial<-ggplot(subset(tombs,est_burials>1 & total_object>0),aes(total_object/est_burials, fill=group))+geom_histogram()+
   theme_bw()+theme(legend.title=element_blank(),
@@
                     scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-metalbyburial<-ggplot(subset(tombs,est_burials>1),aes(weapon/est_burials, fill=group))+geom_histogram()+
+metalbyburial<-ggplot(subset(tombs,est_burials>1 & weapon>0),aes(weapon/est_burials, fill=group))+geom_histogram()+
   theme_bw()+theme(legend.title=element_blank(),
@@
 
-importsbyburial<-ggplot(subset(tombs,est_burials>1),aes(total_import/est_burials, fill=group))+geom_histogram()+
+importsbyburial<-ggplot(subset(tombs,est_burials>1 & total_import>0),aes(total_import/est_burials, fill=group))+geom_histogram()+
   theme_bw()+theme(legend.title=element_blank(),

### 2019-05-21 analysis/paper/paper.Rmd
Writing and image importing

@@
 library(ggforce)
+library(png)
 font_import(paths="../paper/font",pattern="GARA")
@@
 # Introduction
-Three principal interests converge in this paper: reorienting the focus of Early Iron Age Cretan archaeology; the application of computational, and quantitative methods to supplement traditional modes of inference; and open-access data and reproducible research.It may be felt that, of all ancient sites on Crete, Knossos is the last in need of further archaeological attention.
-The city and ‘palace’ of Knossos, first excavated well over a century ago [@Evans1921; -@Evans1928; -@Evans1930], retain a pre-eminent position in popular and academic accounts the island’s Bronze Age, and few, if any, other Cretan sites have been so thoroughly dug, reconstructed, and contested.
-However, there are reasons why the present study – a quantitative comparison of the two largest Early Iron Age cemeteries at Knossos – is not only justified, but also timely.
-
-For one, this is not the Knossos of that illustrous, palatial period, but rather the town as it persisted and evolved during the subsequent Early Iron Age (EIA), about which much less is known.
-Research into this once maligned period has grown apace in the last few decades.
-What began as a reframing of the the twelth-seventh centuries as the gestational period of the later classical Greek world [first seen in the works of @Coldstream1977; @Desborough1972; @Snodgrass1971] has culimated in the complete rejection by many scholars of the image of a society bereft of its former palatial glory.
-Certainly, few would any longer defend the appelation or interpretive bagguage of the ‘Dark Ages’ [on the origins of which, see @Kotsonas2016; @Morris1997].Nonetheless, our knowledge of many EIA settlements (both new foundations, and those with BA ancestry like Knossos) remains often underdeveloped in comparison with those of the second millennium BC.
-
-Secondly, and welcome though they are, a number of recent publications on the socio-economic [@Wallace2010a] and settlement [@Nowicki2000] changes which accompanied the BA-EIA transition have directed scholarly attention toward a number of elevated, defensible settlements supposedly founded in response to the turbulent contemporary political climate.
+Three principal interests converge in this paper: a reorientation of the focus of Early Iron Age Cretan archaeology; the application of computational, and quantitative methods to supplement traditional modes of archaeological inference; and open-access data and reproducible research.
+To begin with the first of these, it may be felt that, of all ancient sites on Crete, Knossos is the last in need of further archaeological attention.
+The city and palace of Knossos, first excavated well over a century ago [@Evans1921; -@Evans1928; -@Evans1930], retain a pre-eminent position in popular and academic accounts the island’s Bronze Age, and few, if any, other Cretan sites have been so thoroughly dug, reconstructed, and contested.
+
+For one, though, this is not the Knossos of that illustrous, palatial period, but rather the town as it persisted and evolved during the subsequent Early Iron Age (EIA).
+Research into this once maligned period has grown apace in the last few decades, beginning with the pioneering works of Coldstream [-@Coldstream1977], Desborough [-@Desborough1972], and Snodgrass [-@Snodgrass1971] and culminating in the rejection by many scholars of the image of a society bereft of its former palatial glory.
+Certainly, few would any longer defend the appelation or interpretive bagguage of the ‘Dark Ages’ [on the origins of which, see @Kotsonas2016; @Morris1997].Nonetheless, our knowledge of many EIA settlements remains often underdeveloped in comparison with those of the second millennium BC.
+
+But more importantly, and welcome though they are, a number of recent publications on the socio-economic [@Wallace2010a] and settlement [@Nowicki2000] changes which accompanied the BA-EIA transition have directed scholarly attention toward a number of elevated, defensible settlements supposedly founded in response to the dangers of this turbulent period.
 This research agenda is likewise reflected in survey work, which has focussed extensively around the Bay of Mirabello, and the Isthmus of Ierapetra.
-This has, unfortunately, engendered a certain disregard for continuing occupation at several lowland centres in the central region of the island, including Knossos and Phaistos.
-There remains, therefore, a need for studies addressing the social dynamics at these sites, with a long-term view to comparative appraisal of the divergent trajectories of communities across the island in the EIA [@Whitley2011, 667–8].
-
-Finally, a number of authors, drawing on datasets from Crete and the wider Aegean, have in recent years demonstrated the great potential of quantitative, statistical, spatial, and network analyses to elucidate social dynamics, settlement patterning, and systems of inter-regional communication and exchange.
-Such work encourages the use of large (and often pre-existing) datasets, and the sharing of both data and methods to facilitate reproducible, collaborative, and innovative research.
+This has, unfortunately, engendered a certain disregard for settlement characterised by quite different topographies, especially in the central region of the island.
+There remains, therefore, a need for studies addressing the social dynamics at such sites (Knossos, Phaistos, Grivila etc.), with a long-term view to comparative appraisal of the divergent trajectories of communities across the island in the EIA [@Kotsonas2011b; @Whitley2011, 667–8].
+
+As for the second and third of my stated interests, several authors have recently demonstrated the great potential of quantitative, statistical, spatial, and network analyses in elucidating social dynamics, settlement patterning, and systems of inter-regional communication and exchange across Crete and the wider Aegean.
+Such work encourages the use of large (often pre-existing) datasets, and the sharing of both data and methods to facilitate reproducible, collaborative, and innovative research [@Bevan2013; @Knappett2011a].
 The cemeteries of Knossos offer an opportunity to introduce such approaches into EIA research, where they have yet to make much of an impact [see @Kotsonas2011].
-The Fortesta and North Cemetery complexes are both well-published [@Brock1957; @Coldstream1996a] and have been considered variously in relation to their "oriental"" imports [@Antoniadis2012], antiques [@Crowe2016], and religious context [@Coldstream1984b].
+The Fortesta and North Cemetery complexes are both well-published [@Brock1957; @Coldstream1996a] and have been considered variously in relation to their 'oriental' imports [@Antoniadis2012], antique artefacts [@Crowe2016], and religious context [@Coldstream1984b].
 But a lack of quantitative analyses, or systematic comparisons between different cemeteries, has left us with many intuited trends, some clearly exceptional tombs, but a general impression of un-patterned multiplicity.
@@
 With a dataset including information on every find and tomb recorded in the Fortetsa and Knossos North Cemetery publications [@Brock1957; @Coldstream1996a], the following analysis examines both the synchronic and diachronic variation among the tombs of both cemeteries.
-In doing so, I aim to demonstrate the potential for quantitative methods, firstly, to permit more reliable identification, and more nuanced characterisation, of significant patterning in the archaeological material; secondly, to facilitate a more systematic appraisal of the cemeteries’ largest tombs, and set their development in a richer socio-economic context; and, finally, to offer not just generalities, but evidence by which to consider, in tangible and context-specific ways, the motivations and identities of the burial groups themselves.
+In doing so, I aim to demonstrate the potential for quantitative methods, firstly, to permit identification, and nuanced characterisation, of significant patterning in the archaeological material; secondly, to facilitate a more systematic appraisal of the cemeteries’ largest tombs and their distinctive histories; and, finally, to offer not just generalities, but tangible and context-specific considerations of the motivations and identities behind the burial groups themselves, with a view to eventually comparing such results with mortuary assemblages across the island.
 
@@
 
-However, by the time of Hood and Smyth’s [-@Hood1981] survey-cum-synthesis of prior excavations, a concentration of activity slightly to the west of the old palace presented a good case for continuing, nucleated settlement, something strongly advocated by Coldstream [-@Coldstream1984a; -@Coldstream2000].
-More recently, the findings of the Knossos Urban Landscape Project suggest a PG settlement of up to 40ha, extending from the Kairatos River to the western slope of the Acropolis, and from the northern slope of Gypsades to between the Minoan palace and the Kephala hill [@Kotsonas2011a, 5-8].
-This estimate remains our best guess in the absence of substantial excavatied deposits but, if it is accurate, then EIA Knossos was one of the largest sites of the contemporary Aegean, and we should expect the existence of relatively complex forms of social interaction and organisation within this exceptional community.
+However, a concentration of activity slightly to the west of the old palace presented a good case for continuing, nucleated settlement [@Hood1981], something strongly advocated by Coldstream [-@Coldstream1984a; -@Coldstream2000], while, more recently, findings of the Knossos Urban Landscape Project suggest a PG settlement of up to 40ha [@Kotsonas2011a, 5-8].
+This estimate remains our best guess in the absence of substantial excavated deposits but, if it is accurate, then EIA Knossos was one of the largest sites of the contemporary Aegean.
 
 ## The Early Iron Age Cemeteries
-
-Survey aside, our most abundant evidence for EIA Knossos comes from the mortuary record.
-The transitional period between the LBA and EIA, admittedly, yields few if any securely datable burials.
-In the EIA proper, however, the evidence becomes more plentiful, with multiple burials known from the Kephala ridge [@Coldstream1963, 38; @Hogarth1899, 82-5], the modern suburbs of Ayios Ioannis and Atsalenio [@Boardman1960; @Davaras1968, 133–46], the cave of Mavro Spelio [@Antoniadis2012, 58-9], and the Khaniale Teke site with its well-known tholos, mooted as evidence for the presence immigrant craftsmen from the Near East [@Boardman1967;@Hutchinson1954; cf.
+```{r map, fig.cap="Map of the Knossos area showing the locations of the various tomb clusters and cemeteries, with EIA pottery distributions from the KULP marked in orange"}
+knitr::include_graphics("../figures/Knossos Tombs Map.png")
+```
+
+Survey aside, our most abundant evidence for EIA Knossos comes from the mortuary record (see Figure \@ref(fig:map).
+The transitional period between the LBA and EIA yields few if any securely datable interments, but later the evidence becomes more plentiful, with multiple burials known from the Kephala ridge [@Coldstream1963, 38; @Hogarth1899, 82-5], the modern suburbs of Ayios Ioannis and Atsalenio [@Boardman1960; @Davaras1968, 133–46], the cave of Mavro Spelio [@Antoniadis2012, 58-9], and the Khaniale Teke site with its well-known tholos, mooted as evidence for the presence immigrant craftsmen from the Near East [@Boardman1967;@Hutchinson1954; cf.
 @Hoffman1997 @Kotsonas2006].
-Taken together, these scattered tombs point to a variegated set of funerary practices, all within about 30 minutes’ walk of each other, each serving some sub-section of the Knossian population.
-But despite their reasonable number when considered collectively, these tombs, so dispersed and often solitary, provide little scope for systematic quantitative analysis.
-
-Fortunately, far larger burial accumulations are to be found in the Fortetsa and the Knossos North cemeteries, which together comprise some 134 tombs, ranging in date from the Subminoan to Late Orienatalising periods – a span of around 400 years.
-Though many were robbed in antiquity, and others destroyed by the digging of later burials, building work, or just the ravages of time, they still represent the most comprehensive and abundant source of evidence for not just mortuary practices, but for social life of any kind at EIA Knossos.
+Taken together, these scattered tombs point to a richly variegated set of funerary practices but, despite their reasonable number, are so dispersed and often isolated that they provide little scope for systematic quantitative analysis.
+
+Fortunately, far larger burial accumulations are to be found in the Fortetsa and the Knossos North cemeteries, which together comprise some 134 tombs, ranging in date from the Subminoan to Late Orienatalising periods.
+Though many were robbed in antiquity, and others destroyed by the digging of later burials, building work, or just the ravages of time, they still represent the most abundant source of evidence for not just mortuary practices, but for social life of any kind at EIA Knossos.
 They are thus the ideal candidates for a quantitative analysis of the kind advanced here.
@@
 These tombs cluster in three main groups.
-The largest of these, including tombs IX, VIII, VII, V, II, III, IV, V and P, is arranged on a north-south axis, with the *dromos* of each tomb extending downslope to the west (Tomb P is set on a slightly different orientation, though, as we shall see, this tomb is particularly exceptional).
+The largest of these, including tombs IX, VIII, VII, V, II, III, IV, V and P, is arranged on a north-south axis, with the *dromos* of each tomb extending downslope to the west.
 Further north lies the cluster of tombs OD, X, XI, LST and BLT.
 The third group lies south of the first, and comprises tombs Θ, Ϙ and F.
-Finally, three tombs excavated in 1933 – L, TFT and Π – which lie close to the main road north of the Acropolis hill [@Brock1957, 1-2] are now considered to belong the Knossos North Cemetery.
+Finally, three tombs excavated in 1933 – L, TFT and Π – which lie close to the main road north of the Acropolis hill [@Brock1957, 1-2], are now considered to belong the Knossos North Cemetery.
 
 To the northeast of the Fortetsa cemetery, a collection of ten badly damaged tombs known, after the year of their excavation, as the Fortetsa 1967 tombs, were probably the southernmost burials of the North Cemetery.
-Unfortunately, all had been looted at the time of excavation, and no in situ deposit was found in any tomb.
-For this reason, unfortunately, they feature little in the present analysis.
+Unfortunately, all had been looted at the time of excavation, and no *in situ* deposit was found in any tomb.
+They are thus excluded from present analysis.
 Northwest of these tombs, and just south of the Teke Tholos, the Teke tombs represent the northern reaches of the KNC.
-The Teke tholos itself, and the tombs around it, were almost certainly part of the same complex, with six unpublished tombs excavated by the Archaeological Service during the construction of intervening buildings in the 1980s.
-The central section of the KNC comprises the ‘Medical Faculty’ site, so named because it was the construction of an extension for the university of Crete that prompted major rescue excavations here in 1978.
+The central section of the KNC comprises the ‘Medical Faculty’ site, named for the extension of the University of Crete that prompted major rescue excavations here in 1978.
 Some 89 tombs of Subminoan to Late Orientalising date were found, by far the largest single concentration of EIA tombs in the whole Knossos area [@Coldstream1996a, 53-55].
 
-The 134 tombs  that make up these two cemeteries, some of which were in use for several centuries, suggest an established, relatively stable community for whom tradition formed an important foundation in mortuary beliefs and behaviours [@Coldstream1994, 106-108].
-The tombs fall into three main types, the chamber tomb, shaft grave, and pit cave or tomb; the latter two are largely Subminoam phenomena, with the former by far the most abundant.
-The chamber tomb comprises a descending ramp or passage, the *dromos*, leading to the chamber itself, cut into the bedrock.
-The *dromos* and chamber are separated by the *stomion*, an opening often set a step or two lower than the *dromos*, which is covered by a large stone slab, smaller stacked stones, or a combination of both [@Antoniadis2012, 47-8].
-It is not uncommon for dromoi to have niches cut into them, to house later burials that could not fit in the chamber (e.g.
-T.
-II), or even, in some cases, cinerary urns removed from the chamber so that new interments could be fitted in (e.g.
-T.
-TFT; @Brock1957, 3-4).
-
-The majority of burials in these two cemeteries are cremations, a rite that appears at Knossos in the Subminoan period, in the Tomb 200+ complex (Ts.
+The tombs themselves fall into three main types, the chamber tomb, shaft grave, and pit cave or tomb; the latter two are largely Subminoam phenomena, with the former by far the most abundant.
+It comprises a descending ramp or passage, the *dromos*, leading to the chamber itself, cut into the *kouskouras* bedrock.
+The *dromos* and chamber are separated by the *stomion*, an opening often set a step or two lower than the *dromos*, which is covered by a large stone slab, smaller stacked stones, or combination of both [@Antoniadis2012, 47-8].
+The majority of burials are cremations, a rite that appears at Knossos in the Subminoan period, in the Tomb 200+ complex (Ts.
 200, 201 and 202 are three chambers of a single pit tomb) in the KNC [@Coldstream1994, 109].
@@
 L, 45, 98, 112 and F67:5), these burials appear to be among the very earliest in each tomb [@Antoniadis2012, 69].
-This degree of coexistence suggests that the shift between the two rites need not have marked a major ideological break; the preference for communal chamber tombs, for instance, something practiced since the Middle Minoan period at Knossos, remained constant.
-Whether this is linked to a strong sense of historical continuity or tradition, close kinship ties, or simply the need to bury large numbers of people in a small geographical area, we cannot be sure.
 
@@
 
-The chronological sequence of the cemeteries has received much attention.
-It has been claimed that no tomb crosses the LM IIIC-SM divide [@Coldstream1998, 58; @Coldstream2006, 582), but there is a strong possibility that this is a classificatory artefact [@Hallager2010].
-The earliest tomb appears to be the T.
-200+ complex, with its precocious cremations and imported artefacts.
-There remain ongoing debates about whether the KNC represents a foundation on virgin soil, as instances of BA tombs reused the EIA are known (e.g.
-the Khaniale Teke tholos).
-Cavanagh [-@Cavanagh1996], for one, could not rule out the possibility of some KNC tombs being of BA date based on his cluster analysis, but other evidence would suggest this is unlikely.
+The initial foundation of the cemeteries has received much attention.
+It has been claimed that no tomb crosses the LM IIIC-SM divide [@Coldstream1998, 58; @Coldstream2006, 582), though there is a possibility that this is a classificatory artefact [@Hallager2010].
+Instances of BA tombs reused the EIA are known (e.g.
+the Khaniale Teke tholos), and Cavanagh [-@Cavanagh1996], based on a cluster analysis of tomb dimensions, could not reject the possibility of some KNC tombs being of BA date, though other evidence would suggest this is unlikely (Whitelaw, pers.
+comm.).
 
 During the use of the cemeteries, correspondences have been highlighted between changing pottery styles, and patterns of tomb construction and abandonment.
-In the Protogeometric period, some 22 tombs then in use in the KNC, and nine at Fortetsa, ceased to receive interments, followed in the succeeding Protogeometric B period by the founding of 22 new tombs (16 and six respectively).
-The PGB was defined by Brock [-@Brock1957], 143], who hailed it as “the most remarkable phase in Cretan vase-painting”.
-Apparently a Cretan anomaly, this phase sees Attic geometric motifs mixed with experimental patterns possibly deriving from Near Eastern metalwork, as well as iconography drawn seemingly from the Minoan repertoire (Coldstream 1984, 93–4).
-Coldstream has argued that the concurrence of this style and a rash of newly founded tombs points to a reformulation of elite mortuary display, noting how, in this period also, redeposited Minoan larnakes began appearing in and around some tombs, in one case even inspiring the decoration of a PGB vessel (in T.
+The Protogeometric B style, first defined by Brock [-@Brock1957], 143], who hailed it as “the most remarkable phase in Cretan vase-painting”, is apparently a Cretan anomaly, and evinces a melding of Attic geometric motifs, experimental patterns possibly deriving from Near Eastern metalwork, and iconography drawn seemingly from the Minoan repertoire (Coldstream 1984, 93–4).
+Coldstream has argued that the concurrence of this style with a rash of newly founded tombs points to a reformulation of elite mortuary display, noting how, in this period also, redeposited Minoan larnakes began appearing in and around some tombs, in one case even inspiring the decoration of a PGB vessel (in T.
 107; see @Coldstream1984b; -@Coldstream1994, 112–3).
@@
 
-The final mystery of the chronological sequence at the KNC and Fortetsa is the quite abrupt cessation of new burials around 630 BCE.
-There had been changes in tomb use in both cemeteries in the preceding century, with no new tombs built in Fortetsa, and few in the KNC, though interments in old tombs continued [@Brock1957, 4; @Cavanagh1996; 651-3].
+The final mystery of the chronological sequence at the KNC and Fortetsa is the quite abrupt cessation of burials around 630 BCE.
+There had been changes in tomb use in both cemeteries in the preceding century, with few new tombs built, though interments in old tombs continued [@Brock1957, 4; @Cavanagh1996; 651-3].
 This curiosity has, along with the equally scanty evidence for the period in the wider settlement, become known as the ‘Archaic Gap'.
 The causes of this lacuna, which persists until the recrudescence of archaeological evidence in the Late Archaic period, remain unknown.
-Although various explanations have been forwarded, [@Coldstream1999, 301-2; @Huxley1994, 126; @Kotsonas2002, 41-4], it seems there are now strong reasons to doubt the settlement was truly abandoned (Whitelaw, pers comm.).
+Although various explanations have been forwarded, [@Coldstream1999, 301-2; @Huxley1994, 126; @Kotsonas2002, 41-4], there are now strong reasons to doubt the settlement was abandoned (Whitelaw, pers comm.).
 
@@
 Influenced by Mauss’ [-@Mauss1954] formulation of ‘the gift’, these authors stressed the mutuable nature of an object’s social meanings, and its capacity for acquiring a ‘biography’, through the transformations of its social existence.
-In Aegean archaeology, these ideas proved attractive because of their resonance with the Homeric motif of gift-giving accompanied by accounts of an object's illustrious heritage [@Whitley2002, 220-1].
-It is hardly suprising that, when direct parallels for such objects have appeared archaeologically – such as the boar’s tusk helmet from Tomb 200+ and that given to Odysseus by Meriones (*Iliad* 10.260-271) – a particular interpretive lens, melding Homeric and anthropological ideas, has been applied to certain elaborate burials.
-
-Catling’s discussion of the T.
-200+ complex is a notable example.
+These ideas proved attractive with their evocation of the Homeric motif of gift-giving [@Whitley2002, 220-1], and the remarkable archaeological parallels for such objects – such as the boar’s tusk helmet from Tomb 200+ and that given to Odysseus by Meriones (*Iliad* 10.260-271).
+Thus, a particular interpretive lens, melding Homeric and anthropological ideas, has been applied to certain elaborate burials.
+
+Catling’s discussion of the Tomb 200+ complex is a notable example.
 Tomb 201 contained the remains of at least two adults, and possibly a child.
-The burial objects included a bronze sword, spearhead, and arrowhead, an iron dirk or knife, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned BA-style boar’s tusk helmet [@Catling1995, 123].
+Objects included a bronze sword, spearhead, and arrowhead, an iron dirk or knife, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned helmet [@Catling1995, 123].
 Catling identifies similarities between this tomb and burials at Tiryns, Kaloriziki, and the famous *hērōön* of Lefkandi.
 All contained imported items at an early date, were of ‘warrior grave’ type, and formed loci for later burials.
-Catling argues that those individuals who dared to travel abroad in this time of more limited seafaring, accruing with precious artefacts and stories of distant lands – ‘heroes returned’ in Catling’s phrase – would have been accorded exceptional treatment in death [@Catling1995], 127–8].
-
-Complementing the theme of heroes abroad has been a similar degree of interest in foreigners at home.
-Two caches of gold jewellery, found in pits flanking the entrance to the Teke Tholos, were argued by Boardman [-@Boardman1967] to be foundation deposits – a known Near Eastern custom – for the burial place of an immigrant craftsman, probably Phoenician, practising his trade at Knossos.
+Catling argues that those individuals who dared to travel abroad in this time of more limited seafaring, accruing precious artefacts and stories of distant lands, would have been accorded exceptional treatment in death [@Catling1995, 127–8].
+
+Complementing the theme of heroes abroad has been a similar interest in foreigners at home.
+Two caches of gold jewellery, found in pits flanking the entrance to the Teke Tholos, were argued by Boardman [-@Boardman1967] to be foundation deposits – a Near Eastern custom – for the burial place of an immigrant craftsman, probably Phoenician in ancestry.
 Kotsonas has revisited this attribution, suggesting that the tholos is more likely the resting place of an elite individual or family with a monopoly over a metal workshop’s output [-@Kotsonas2006, 155–9].
@@
 
-Finally, a recurrent interest has been in the 'presence of the past' in these tombs.
-The practice of hero or ancestor cult, much discussed on the mainland (see @Antonaccio1994; @Coldstream1976), is not common here.
-Instead, we see several Late Minoan larnakes (and a few other vessels) redeposited in tombs and pits in the KNC and Fortetsa from the PGB period onward.
-A thorough analysis of these artefacts has been provided by Crowe [-@Crowe2016], who argues that the larnakes were robbed from tombs in the Knossos area, perhaps with a particular preference for painted specimens where available.
-A particularly evocative case comres from tomb 107, where parts of an elaborate figured LM larnax have been found along with a PGB vessel by the so-called Tree Painter.
-Both feature women in the company of birds and spiralling trees, and it seems certain that the former inspired the design of the latter.
-The PGB vessel also seems to draw on Oriental motifs, with the polos crown worn by the goddess having Near Eastern origins [@Coldstream1984b, 94-101].
-The magpie-like amalgamation of these styles and motifs seem to exemplify the particular cultural openness, fluidity and competition suggested to characterise the PGB period.
-Artefacts such as this hint towards an intimate discourse of (re)discovery; a painter, struck by the strangeness, or perhaps the uncanny familiarity, of centuries-old images, incorporated them into his own work.
-
 ## Number Crunching: Quantitative Analyses of the Cemeteries
 
-Generally speaking, the above analyses have tended to rely more on particular finds, single tombs, or observable, but limited, patterns in the data to form their conclusions.
-A few authors have applied more quantative approaches, and these point toward their potential utility.
-Antoniadis [-@Antoniadis2012, 172-6], in his study of Oriental imports, divided the EIA tombs of Knossos into three groups, those with fewer than 10 pots, fewer than 50 pots, and more than 50 pots, and found that, in each successive case, the proportions of tombs containing imports and imitations increased.
-The implication is that a hierarchy existed, with differing degrees of control over the accumulation of prestigious goods.
-That said, the relationship between such items and tomb ‘wealth’ is examined critically in the present work, and may prove more complex than Antoniadis implies.
-
+Generally speaking, the above analyses have tended to rely more on particular finds, tombs, or observable, but limited, patterns in the data.
+Only a few authors have applied more quantative approaches, but these point toward their potential utility.
 Cluster analyses have been employed by both Cavanagh [-@Coldstream1996, 653-7] and Antoniadis [-@Antoniadis2012, 193-7], the most sophisticated statistical techniques so far applied to the present material.
 Cavanagh’s analysis of tomb architecture has been mentioned, while Antoniadis’ more holistic approach corroborates quantitatively some of the intuitive correspondences noted between specific tombs in previous studies.
-Antoniadis does not interrogate his findings as much as he might have, but his highlighting of the relationship between tomb foundation dates, associated finds, and geographical location points towards social processes of self-definition and exclusion among the burial groups that would reward further investigation, something offered by the present study.
+Antoniadis [-@Antoniadis2012, 172-6] also divided the EIA tombs of Knossos into three groups, those with fewer than 10 pots, fewer than 50, and more than 50, and found that, in each successive case, the proportions of tombs containing 'oriental' imports and imitations increased, the implication being that a hierarchy existed, with unequal access to such prestigious goods.
+Yet even here the available data could be further interrogated, such as with the relationship between imported items and so-called tomb ‘wealth’.
+The present study seeks to extend and nuance a number of observations that have been made in previous studies, as well as revealing patterns thus far unacknowledged, through use of a more systematic approach to quantitative analysis.
 
@@
 The cornerstone of the present study is a database (or, rather, three linked databases) based on the excavation reports for the Fortetsa and Knossos North cemeteries [@Brock1957; @Coldstream1996a].
-The first database includes an entry for each individual ceramic vessel, by far the largest portion of all known finds.
+The first includes an entry for each individual ceramic vessel, by far the largest portion of all known finds.
 Each entry includes: the tomb from which the vessel came; the vessel shape; whether it was an open or closed vessel; its main dimensions; its ceramic period and corresponding absolute dating measures (see below).
 In addition, all imports, Oriental imitations, and antique (that is, Bronze Age) artefacts were accordingly categorised, using the studies of Antoniadis (2012), Jones (2000), and Crowe (2016) to supplement omissions in the original excavation reports.
-In total, this database comprised just over 4540 individual artefacts.
+In total, this database comprises just over 4540 individual artefacts.
 
 The second database is similar, but with each entry corresponding to a find other than a ceramic vessel.
-The categories remained the same as in the pottery database, but with the addition of a ‘material’ variable, as of course these finds were not all made of the same substance.
+A ‘material’ variable was also added, as of course these finds were not all made of the same substance.
 This database included some 1620 individual artefacts.
+
 The final database concerns the tombs themselves.
-Each of the 134 tombs was accorded its own entry, with variables including dating (both in relative and absolute terms), surviving dimensions, and the known and estimated number of interments.
-Each entry also included a tally of the objects recorded in the other two databases.
+Each tomb was accorded its own entry, with variables including dating (both in relative and absolute terms), surviving dimensions, and the known and estimated number of interments, as well as a tally of the objects recorded in the other two databases.
 These tallies were arranged by shape for the pottery, and by material and category (e.g.
@@
 This is not the place for a detailed investigation into the lack of reproducible research in archaeology (for this, see ...).
-Suffice it to say the problem exists, though, granted, gradual progress is being spearheaded by some.
-This article is, among other things, an attempt to bring these principles of maximal reproducibility to the archaeology of Early Iron Age Crete.
+Bu this article is, among other things, an attempt to bring these principles of maximal reproducibility to the archaeology of Early Iron Age Crete.
 Within this field, catalogues and excavation reports abound and, though many of these may be forgiven based on their dates of publication, the lack of readily utilisable digital data from a number of recent articles and monographs, indeed, the failure to even present the data such that they could be manually digitsed, is less pardonable.
-Preparation of databases such as those of the present study can (and did) involve prolongued trawling of printed tables (often without consistent formatting, coding or presentation which would enable easy digitisation), cross-referencing artefacts listed inconsistently in different contexts, or else grappling with catalogues which, through their unexplained lack of certain critical information (e.g.
-coordinates in site catalogues), offer little research potential beyond their descriptive face-value.
+Preparation of databases such as those of the present study can (and did) involve prolongued trawling of printed tables (often without consistent formatting, coding or presentation which would enable easy digitisation), cross-referencing artefacts listed inconsistently in different contexts, or else grappling with catalogues which, through their unexplained lack of critical contextual information, offer little research potential beyond their descriptive face-value.
 
@@
 In this vein, the present article has been composed making use of Ben Marwick's "rrtools", an open-source package for the statistical software R Studio, which provides the author with the tools to create a self-contained "compendium", which can be freely downloaded, and from which anyone can accurately reproduce the entire workflow of the present analysis.
-This analysis was conducted in R, and the full code, as well as the raw and derived datasets, complete bibliography (in BibTex format, for export to any major reference manager), figures, and article as originally submitted to this journal are included in the package.
+This analysis was conducted in R, and the full code, as well as the raw and derived datasets, complete bibliography (in BibTex format, for export to any major reference manager), figures, and article as originally written are included in the package.
 It can be downloaded as a repository from GitHub, at the following address: https://github.com/DCPollard94/knossoscemeteries, and the raw datasets have been stored with the online repository Figshare, here:
 
-Though not visible in the present paper, every figure and table herein was produced directly, through written code, from the raw data using R.
+Though not visible in the present paper, every chart and table herein was produced directly, through written code, from the raw data using R.
 The full code can be viewed and followed by downloading the repository from GitHub, and all the data are made available through an MIT license, which permits unlimited use, ammendation, and dissemination thereof, yet frees the present author from any responsibilty for that use, or for any perceived errors or inaccuracies in the data.
-It is sincerely hoped that others may find interest in, and themselves further explore and expand upon the research presented here, by downloading this compedium.
+It is sincerely hoped that others may find interest in, further explore, and expand upon the research presented here by downloading this compedium.
 
@@
 A first important step in preparing the data for analysis was the estimation of the number of individuals buried in each tomb.
-The method chosen for this estimation broadly follows that of Cavanagh [-@Cavanagh1996, 659-660], who took the number of pithoi (the most common urn type) and the number of pithoi, amphorae and kraters (which were in some cases used as urns) as middle and upper estimates of the number of interments, with all securely identified interments (in the form of ashes or osteological material) being the lower limit (Figure \@ref(fig:burialestimate)).
+The method chosen broadly follows that of Cavanagh [-@Cavanagh1996, 659-660], who took the number of pithoi (the most common urn type) and the number of pithoi, amphorae and kraters (which were in some cases used as urns) as middle and upper estimates of the number of interments, with all securely identified interments (in the form of ashes or osteological material) being the lower limit (Figure \@ref(fig:burialestimate)).
 In the present study, the mean of these three estimates was taken, to provide a single aggregate measure for the purposes of analysis (where no interments were known, this variable was still assigned a value of "1", for all tombs must have contained at least one burial).
@@
 
-But the artificial clustering this technique engenders around select absolute dates proves misleading when, for instance, the total number of pithoi deposited by date are plotted via a histogram (Figure \@ref(fig:allpithoi)); the apparent cessation of burial activity for 30 years in the 8th century, followed by the most intense episode of deposition yet seen, is simply a fiction generated by the processing of the data.
+But the artificial clustering this technique engenders around select absolute dates proves misleading when, for instance, the total number of pithoi deposited by date are plotted via a histogram (Figure \@ref(fig:allpithoi)); the apparent cessation of burial activity for 30 years in the 8th century is simply a fiction generated by the processing of the data.
 
@@
 Here, rather than assign a pot a single date, the entire time span under investigation is divided into equal chunks (in this case, decades), and probability values summing to 1 are then assigned for each vessel, per chunk of time, based on its accordant date range.
-So, our PGB vessel, datable between 840 and 810 BC would register a value of 0.33 for each of the three decades within which we it could theoretically have been produced.
+So, our PGB vessel, datable between 840 and 810 BC would register a value of 0.33 for each of the three decades within which it was likely produced.
 These values for individual vessels can can then simply be summed to produce tomb- or cemetery-level signatures.
-The major contribution of this method (though it too creates problems; see @Crema2010, 1123-1124) is that it incorporates temporal uncertainty into the size of the resulting values, that is, "events with tight temporal definition contribute more to the total probability over their range than do loosely defined events" [@Johnson2004a, 450].
+The major contribution of this method is that it incorporates temporal uncertainty into the size of the resulting values, that is, "events with tight temporal definition contribute more to the total probability over their range than do loosely defined events" [@Johnson2004a, 450].
 
@@
 As seen in Figure \@ref(fig:aoristallcems), the artificial peaks and troughs of the simple averaging method are avoided, and different patterns become visible.
-Though these methods present their own limitations, which will be explored in the following sections, they are considered a more robust and productive approach to the temporal uncertainty of ceramic dating and, as such, are used for much of the present analysis.
-
-##Theory and method
-
-Before proceeding to the main discussion, it should be stressed that, though I hope to demonstrate the power of quantitative and statistical approaches to the mortuary record, data-points cannot replace description, and theory must not be absent in the use of such methods.
-Quantifications and statistical analyses should, ideally, be employed as tools for the development and exploration of theoretically-informed models of social practice.
-Throughout the following analysis, I aim to link the observed patterning in the material back to possible forms of human behaviour, which in turn guides the focus of subsequent analyses, and so on.
-It is hoped that the value of such an approach will become clear.
+Though these methods present their own limitations (see @Crema2010, 1123-1124), they are considered a more robust and productive approach to the temporal uncertainty of ceramic dating and, as such, are used for much of the present analysis.
 
@@
 From individual burials, to collective tombs, to the cemeteries as a whole, and between cemeteries in the landscape, we have the opportunity to study and compare both the synchronic and diachronic variation across these various levels.
-The following analysis is divided broadly into these two distinct, though complementary, perspectives.
-I consider, firstly, how temporal variations in the material may relate to broader social dynamics and, secondly, how the form and content of the tombs and their assemblages can enrich the schematic picture derived from the chronological patterns.
-Taken together, these two approaches enable us to develop a nuanced account of the changing social landscape of EIA Knossos, drawing together issues of group identity, external relations, and the communication of wealth and status through funerary practice.
-
-##Temporal variation in the Knossian Cemeteries
+In so doing, I first investigate the broad temporal trends in tomb construction, burial, and pottery deposition across both cemeteries.
+Through this, a number of notable tombs are isolated, which I then examine in an attempt to understand what sets them apart from the rest.
+Tomb dimensions, rates of burials, and changes in the composition of assemblages through time are all considered, to produce an account which speaks to the changing social landscape of EIA Knossos, drawing together issues of group identity, external relations, and the communication of wealth and status through funerary practice.
+
+##Temporal developments in the Knossian Cemeteries
 
@@
 ggarrange(aorist_pithoi_plot,aorist_pots_plot,ncol=2,common.legend=TRUE, legend="bottom")
-
 ```
@@
 Figure \@ref(fig:aoristbycem) presents the aoristic sums of pithoi (as a proxy for burials) and of all vessels, respectively.
-Immediately apparent are the broad correspondences between both cemeteries.
-Despite differences in scale and timing, we can see the Fortetsa and Knossos North Cemeteries experieced similar chronological developments in the volume of burials and, to a slightly lesser degree, grave goods.
+Despite differences in scale and timing, it is immediately apparent that the Fortetsa and Knossos North Cemeteries experieced similar chronological developments in the volume of burials and, to a slightly lesser degree, grave goods.
 In both cemeteries, the rate of cremation burial began rising around the end of the 9th century, plateaued slightly in the early 8th, and then rose again to a peak in the late 8th and early 7th centuries.
@@
 Figure \@ref(fig:tombconstruction) presents two ways of dating the construction of the tombs.
-On the left, counts are presented, based on the ceramic phase considered most likely to correspond to the tomb's first use, including the broad labels "PG", "G" and "O" where no greater specificity is possible.
-On the right, aoristic sums are presented, and essentially represent the spreading of the uncertainty associated with those general assignations across the sub-periods they encompass.
+On the left, counts are based on the ceramic phase considered most likely to correspond to the tomb's first use, including the broad labels "PG", "G" and "O" where no greater specificity is possible.
+On the right, aoristic sums are presented, with the uncertainty associated with those general assignations spread across the sub-periods they encompass.
 In both cases, the trend appears clear.
@@
 The Fortetsa, meanwhile, saw next to no new chamber tomb construction from the 8th century onwards.
-In both cases, then, the rates of tomb building run counter the late 8th-early 7th century spate of burial activity.
-These countervalent trends present a number of interpretive problems.
 The abatement of chamber tomb construction has been noted before (Brock 1957, 4; Cavanagh 1996, 652–3) and has been somewhat implicitly read as evidence of restructuring of the funerary rite, if not a decline in the fortunes of the wider community.
-But this image of a waigning investment in chamber burials sits uneasily with the apparent growth in mortuary activity suggested by both counts of artefacts, and estimates of the number of tombs in use in each period.
+But this image of a waigning investment in chamber burials sits uneasily with the observable growth in mortuary activity in the late 8th century.
 
@@
 
-Figure \@ref(fig:tombsinuse) presents the estimated number of tombs in use throughout the EIA at both cemeteries, that is, the number of tombs which yielded pottery of each period.
-At the Forteta, this number remained rlatively stable from the MPG to LG periods before declining (and bear in mind, there are only around 20 tombs belonging to this cemetery throughout the period), while at the KNC, the numbers of tombs in use rose to a preak in the late 8th century before declining.
+Such countervalent trends are observable too in the number of tombs receiving in each period.
+Figure \@ref(fig:tombsinuse) presents the estimated number of tombs in use throughout the EIA at both cemeteries.
+At the Forteta, this number remained relatively stable from the MPG to LG periods before declining (and bear in mind, there were only around 20 tombs belonging to this cemetery throughout the period), while at the KNC, the numbers of tombs in use rose to a preak in the late 8th century before declining.
 The cessation of burials there certainly appears more sudden and marked than at Fortetsa, though, again, it was always a larger cemetery.
-But with the data available, we can interrogate these trends further.
+But this is not the whole story.
 
@@
 
-Figure \@ref(fig:tombsinusebybuilddate) presents the number of tombs in use in each period, divided by the centuries in which those tombs were constructed, all overlaid on the total number of tombs in use across both cemeteries.
-Perhaps unsurprisingly, a wave-like pattern can be observed with each century marked by the construction of new tombs which then steadily go out of use, as new ones are built.
-This would seem to accord with Cavanagh's suggestion that "[t]he length of time the tombs last seems [...] to reflect the duration of the social group which had access to the sepulchre", and perhaps even that this access was lineally inehrited, leading to frequent fission and, thus, limited trans-generational use.
-However, something else is going here, which can only be appreciated by considering which tombs became the focus of the intense period of depositional activity in the late 8th and erly 7th centuries.
+Figure \@ref(fig:tombsinusebybuilddate) shows the number of tombs in use in each period, divided into the centuries in which those tombs were constructed, all overlaid on the total number of tombs in use across both cemeteries.
+A wave-like pattern can be observed with each century marked by the construction of new tombs which then steadily go out of use, as new ones are built.
+This would seem to accord with Cavanagh's suggestion that "[t]he length of time the tombs last seems [...] to reflect the duration of the social group which had access to the sepulchre", and that, if such access was lineally inehrited, frequent fission would result in limited trans-generational use.
+We might expect, then, that it was the 8th and 7th century tombs coming into use that spurred the dramatic rise in burial rates previoulsy noted in this period.
+But this is not the case.
 
@@
 
-Figure \@ref(fig:potsbydateandtombdate) shows, firstly, the number of vessels dating to each century across both cemeteries, based on their assigned ceramic period, essentially a simplified version of the trend illustrated in Figure \@ref(fig:aoristallcems).
-But, on the right, Figure \@ref(fig:potsbydateandtombdate) shows those pots divided into the centuries to which the tombs they were found in date.
-So as can be seen, though the late 8th and early 7th centuries represented the apex of burial activity in across both cemeteries, it was not the tombs built in those centuries, but rather those built in the centuries prior, which received the lion's share of the burials and associated ceramics.
-This is all the more notable when, returning to Figure \@ref(fig:tombsinusebybuilddate), we see that the number of 9th and 10th century tombs in use by the late 8th and early 7th centuries were declining.
+Figure \@ref(fig:potsbydateandtombdate) shows, firstly, the number of vessels dating to each century across both cemeteries, based on their assigned ceramic period.
+But on the right, those pots are divided into the centuries to which the tombs they were found in originally date.
+So as can be seen, though the late 8th and early 7th centuries represented the apex of burial activity across both cemeteries, it was not the tombs built in those centuries which received the lion's share of the burials and associated ceramics.
+This is all the more notable when, returning to Figure \@ref(fig:tombsinusebybuilddate), we see that the number of 9th and 10th century tombs in use by the late 8th and early 7th centuries was declining.
 
@@
 The significance of these findings can be seen in Table 1.
-Taking the number of pithoi present in each tomb as a proxy for burials, we can estimate the changing rates of burial through the centuries.
-And while for most of the cemeteries' lives the burial rate remained around three per tomb per century, in the 8th and 7th centuries tombs founded in the 9th century saw a rapid rise in the rate of burial, peaking at around 12 per tomb per century.
-Though these estimates are no doubt very rough, the marked nature of this divergence between the 9th century tombs and the rest suggests it is not simply an artefact of imperfect data.
-
-Also apparent from Table 1, though, are the markedly low rates of burial which edominated at both cemeteries.
+Using pithoi as a proxy for burials, we can estimate the changing rates of burial through the centuries based on how many tombs yielded how many pithoi of each period.
+And while for the most part burial rates remained around three per tomb per century, in the 8th and 7th centuries, tombs founded in the 9th century saw a rapid rise in the rate of burial, peaking at over 11 per tomb per century.
+Though these estimates are no doubt very rough, the marked nature of this divergence suggests it is not simply an artefact of imperfect data.
+
+Also apparent from Table 1, though, are the markedly low rates of burial which predominated at both cemeteries.
 Two to four burials per tomb per century amounts to a very selective burial rite, even considering the restricted portion of the population probably represented by these cemeteries [@Cavanagh1996, 664].
-It seems unlikely, then, that the rising rates of burial noted above are, for instance, the direct result of population growth; the increase seems too rapid, and is concentrated only in certain (odler) tombs, not to mention that, based on KULP’s survey data, the major population increase at EIA Knossos should be dated a century or more before the late 8th century peak in burial, in the Protogeometric period (Whitelaw, per comm.).
-
-We might assume that the tombs which became the focus of intense burial activity in the late 8th and early 7th centuries would be those which yielded the most burials and grave goods when excavated.
-Certainly, the largest 12 tombs are all chamber tombs, with 10 dating to the 9th century, and two to the 10th.
-But looking more closely at the aoristic sums of their assemblages through time, not all of them appear to have exhibited the noted intensification of burial activity into the Orientalising period.
+It seems unlikely, then, that the rising rates of burial noted above are, for instance, the direct result of population growth; the increase seems too rapid, and is concentrated only in certain (older) tombs, not to mention that, based on KULP’s survey data, the major population increase at EIA Knossos should be dated a century or more before the late 8th century peak in burial, in the Protogeometric period (Whitelaw, per comm.).
+
+##Isolating the phenomenon
+
+We are yet to draw out *which tombs exactly* are responsible for the observible patterning so far discussed.
+We might assume that the tombs which became the focus of intense burial activity in the late 8th and early 7th centuries would be those which yielded the most burials and grave goods when excavated and, happily, this supposition proves true.
 
@@
 
-In Figure \@ref(fig:toptombs), we see that, seven of the 12 did exhibit late, intense episodes of deposition.
-Tombs X, G, Q, 104 and 283 do, however, evince a different pattern, with their respective peaks coming in the PGB and EG periods, around a century earlier.
-We have seen already in Figure \@ref(fig:aoristbycem) that, particularly in the KNC, the end of the 9th century was another period of increased burial activity, and these tombs shall be considered again shortly.
-But the other seven do appear to underpin the late 8th to early 7th century spate of burials, and between them accrount for `r round(calc1,digits=2)`% of the total pottery assemblage recovered from the `r nrow(tombs)` tombs considered in the present analysis.
-The question arises then, as to whether these tombs were in any way exceptional at the time of their construction, and it is to this possibility which we now turn.
-
-##Tomb characteristics and the composition of the burial assemblages
+The 12 tombs with the most associated vessels are all chamber tombs, with 10 dating to the 9th century, and two to the 10th.
+And, in Figure \@ref(fig:toptombs), we see that seven of the 12 did exhibit late, intense episodes of deposition.
+Tombs P, 292, 75, 107, 285, II, 218, and 104 between them accrount for `r round(calc1,digits=2)`% of the entire pottery assemblage recovered from the `r nrow(tombs)` tombs considered here.
+All but Tomb 285 date to the 9th century, and together they underpinned the late spate of burials here considered.
+Interestingly, Tombs X, G, Q, 104 and 283, however, evince a different pattern, with their respective peaks coming in the PGB and EG periods, around a century earlier.
+We have seen already in Figure \@ref(fig:aoristbycem) that, particularly in the KNC, the end of the 9th century was another period of increased burial activity.
+What, then, set these two groups of tombs apart?
+The recipients of the greatest numbers of burials and grave goods in both cemeteries, they nonetheless seem to have achieved that distinction via different histories.
+By looking more closely at the periods in which they were built, and in which they were most popular, I hope to shed light on the broader social contexts in which these tombs were flourished.
+
+##Not all tombs are created equal
 
@@
 Synchronic comparisons are harder to make with the available data, but the dimensions of the tombs offer some scope for doing so.
+Perhaps the tombs which went on to receive the most burials were notably large or grand when first built, reflecting the status of the associated burying group.
 
@@
 
-Looking at , however, it seems that no clear relationship existed between the physical size of a tomb and its buried population at EIA Knossos.
-It is true that over half of all tombs do not exceed 2.6m2 in area, while nine of the ten tombs with the most burials do surpass this value.
+Looking at Figure \@ref(fig:tombdimensions), however, it seems that no clear relationship existed between the physical size of a tomb and its buried population at EIA Knossos.
+It is true that over half of all tombs do not exceed 2.6 sq m in area, while nine of the ten tombs with the most burials do surpass this value.
 But many of the very largest tombs are decidedly middling when it comes to burial numbers.
-Things are even less clear for the dromos, the most visible part of the tomb, and a plausible arena for demonstrations of scale and grandeur during the burial itself.
-Admittedly, the very largest two tombs by burial numbers also had exceptionally long dromoi, but the next eight tombs with the largest dromoi do not rank highly when it comes to number of interments.
-There are difficulties here in that the dromoi especially were common victims of encroachment by later tombs, and our estimates of their original length are tentative in many cases.
+Things are even less clear for the *dromos*, the most visible part of the tomb, and a plausible arena for demonstrations of scale and grandeur during the burial itself.
+The very largest two tombs by burial numbers also had exceptionally long *dromoi*, but the next eight tombs with the largest *dromoi* do not rank highly when it comes to interments.
+There are difficulties here in that the *dromoi* especially were common victims of encroachment by later tombs, and our estimates of their original length are tentative in many cases.
 
@@
 
-Looking at ...
-we can see that, through the centuries, the dimensions of the tombs varied, sometimes consistently, as with the trend toward narrower **dromoi**.
-Tombs of the 9th century do not particularly stand out here, though it is worth noting that this period did witness the peak in the size of the actual chambers of tombs, as well as being the period in which these sizes varied most.
-But, then again, even the largest chambers are not exactly cavernous, and the extra labour expended on the construction of the larger chambers would have been limited.
-The dromos is a plausible area for conspicuous demonstrations of scale but, with their frequently poor preservation, we cannot be confident in assessing this.
+Looking at Figure \@ref(fig:dimensionsboxplots), though, we can see that, through the centuries, the dimensions of the tombs varied and sometimes consistently, as with the trend toward narrower *dromoi*.
+The 9th-century tombs, including all but one of those seven noted above, do not particularly stand out.
+However, it does seem that chamber were, on average, larger in this period as well as varying more in size between tombs.
+Then again, even the largest chambers are not exactly cavernous, and the extra labour expended on the construction of the larger chambers would have been limited.
+The *dromos* is a plausible area for conspicuous demonstrations of scale but, with their frequently poor preservation, we cannot be confident in assessing this.
 It appears at any rate that they were gradually declining in size, albeit with some notable outliers.
 
-What these figures do suggest, though, is that if tomb size carried a social meaning at the time of construction, this proved irrelevant in the selection of certain tombs for extensive later reuse.
-That is, the tomb’s later popularity seems not have been foreseen.
-Many large tombs had niches cut in their dromoi, or the stomion extended, to fit in later burial urns.
-Sometimes, urns from within the tomb itself were placed in such niches to make space for new interments in the chamber itself.
-There is an impression of individuals vying to be included in the chamber itself.
-That extra space had to be made, and competition was fierce for inclusion into the sepulchral group, supports the idea that the tombs’ original builders little anticipated their eventual popularity and so did not scale their tombs accordingly.
+So perhaps, at their time of building, such aspects of tomb construction carried social meaning but, even if they did, these Figures suggest this proved irrelevant to their later reuse.
+Several tombs had niches cut into their *dromoi*, or the *stomion* extended, to fit in later burial urns.
+Sometimes, urns from within the tomb were placed in such niches to make space for new interments in the chamber itself.
+There is an impression of individuals vying to be included in the chamber itself, of competition was for inclusion into the sepulchral group, supporting the idea that the tombs’ original builders little anticipated their eventual popularity and so did not scale their tombs accordingly.
 
@@
 
+##Imports, interments, and indices of wealth
+
 Though it may not be immediately apparent from the dimensions of the tombs, however, the late 9th and early 8th centuries are consistently interpreted as periods of artistic development, increased overseas trade, and renewed vigour in the construction of tombs and the visibility of the burial rite.
-The PGB style, first identified by Brock, purportedly reflects a uniquely Cretan (perhaps even Knossian) melding of Attic and Near Eastern forms and motifs, while the appearance of Bronze Age 'antiques' in the tombs from this period onwards is felt to reflect renewed, even competitive interest in the Minoan past, as a tool for mortuary display or even lineage affirmation.
-We can
+The PGB style reflects a uniquely Cretan blend of Attic and Near Eastern forms, while the appearance of Bronze Age 'antiques' in the tombs from this period onwards is felt to reflect renewed, even competitive interest in the Minoan past, as a tool for mortuary display or even lineage affirmation.
 
@@
 
-The rise in imported items is clearly reflected in the tomb assemblages, with a succession of increased Greek and Near Eastern imports respectively in the late 9th and early 8th centuries.
-Yet whilst confirming some of the prior inferences about the LPG-EG periods, the deposition of these imported wares can also cast light on the original pattern drawn out in the present analysis: the late surge in burial activity among some select 9th and 10th century tombs.
+The rise in imported items is clearly reflected in the tomb assemblages, with a succession of increased Greek and Near Eastern imports respectively in the late 9th and early 8th centuries (see Figure \@ref(fig:imports)).
+Yet beyond conforming to our prior inferences about the LPG-EG periods, the deposition of these imported wares may also cast light on the original pattern drawn out in the present analysis: the late surge in burial activity among some select 9th and 10th century tombs.
 
@@
 
-Leaving aside the later tombs for a moment, we can see in Figure ...
-that the tombs already built by the end of the PGB period, particularly those built in the 9th and, to a lesser extent, 10th centuries, were major beneficiaries of the influx of imported items during this time, but that, in the following centuries, in line with the rest of the cemetery, the number of imported items deposited within them declined.
-Imported ceramics are used here as something of a proxy for items of status value, a shorthand that would ideally be bolstered by other artefacts (such as precious vessels, weapons, jewellery etc.) which, owing to their less developed chronologies, must unfortunately be left out of such diachronic analyses.
-But the beginnings of my final argument are seen here in Figure ..., namely that the late surge in burial activity seen across both cemeteries in the late 8th and early 7th centuries would appear to be of a different kind to that of a century earlier.
+In Figure \@ref(fig:importstombcenturies), we see that the tombs already built by the end of the PGB period, particularly those built in the 9th and, to a lesser extent, 10th centuries, were major beneficiaries of the influx of imported items during this time, but that, in the following centuries, in line with the rest of the cemetery, the number of imported items deposited within them declined.
+Imported ceramics are used here as something of a proxy for items of status value, a shorthand that would ideally be bolstered by other artefacts (such as precious vessels, weapons, jewellery etc.) which, owing to their less developed chronologies, must unfortunately be left out of such diachronic analyses for now.
+But the beginnings of my final argument are seen here too, namely, that the late surge in burial activity seen across both cemeteries in the late 8th and early 7th centuries would appear to be of a different kind to that of a century earlier.
 

### 2019-05-23 analysis/paper/paper.Rmd
Discussion and Conclusion

@@
 
-Looking at Figure \@ref(fig:dimensionsboxplots), though, we can see that, through the centuries, the dimensions of the tombs varied and sometimes consistently, as with the trend toward narrower *dromoi*.
+Looking at Figure \@ref(fig:dimensionsboxplots), though, we can see that, through the centuries, the dimensions of the tombs varied, and sometimes consistently, as with the trend toward narrower *dromoi*.
 The 9th-century tombs, including all but one of those seven noted above, do not particularly stand out.
-However, it does seem that chamber were, on average, larger in this period as well as varying more in size between tombs.
+It does seem that chambers were, on average, larger in this period as well as varying more in size than in any other period.
 Then again, even the largest chambers are not exactly cavernous, and the extra labour expended on the construction of the larger chambers would have been limited.
@@
 
+imitations<-subset(pots,pots$imitation!="")
+imitation_aorist<-colSums(imitations[27:76])
+imitation_aorist<-data.frame(imitation_aorist)
+colnames(imitation_aorist)<-c("aoristic_sum")
+row.names(imitation_aorist)<-dates
+setDT(imitation_aorist, keep.rownames = "decade")
+imitation_aorist$decade<-as.numeric(imitation_aorist$decade)
+
 ggplot()+geom_col(data=import_aorist,aes(decade,aoristic_sum),alpha=0.2)+
           geom_col(data=allimport_aorist,aes(decade,aoristic_sum, fill=origin),position="dodge")+
+            geom_col(data=imitation_aorist,aes(decade,aoristic_sum))+
             theme_bw()+theme(text=element_text(family="Garamond", size=12))+
@@
 
-The rise in imported items is clearly reflected in the tomb assemblages, with a succession of increased Greek and Near Eastern imports respectively in the late 9th and early 8th centuries (see Figure \@ref(fig:imports)).
+The rise in imported items is clearly seen in the tomb assemblages, with a succession of increased Greek and Near Eastern imports respectively in the late 9th and early 8th centuries (see Figure \@ref(fig:imports)).
 Yet beyond conforming to our prior inferences about the LPG-EG periods, the deposition of these imported wares may also cast light on the original pattern drawn out in the present analysis: the late surge in burial activity among some select 9th and 10th century tombs.
@@
 Imported ceramics are used here as something of a proxy for items of status value, a shorthand that would ideally be bolstered by other artefacts (such as precious vessels, weapons, jewellery etc.) which, owing to their less developed chronologies, must unfortunately be left out of such diachronic analyses for now.
-But the beginnings of my final argument are seen here too, namely, that the late surge in burial activity seen across both cemeteries in the late 8th and early 7th centuries would appear to be of a different kind to that of a century earlier.
+But, nonetheless, this pattern hints towards a possibility that shall now be explored, that the surge in burial activity seen across both cemeteries in the late 8th and early 7th centuries would appear to be of a different kind to that of a century earlier.
 
@@
 
-Figure ...
-is a boxplot for the aoristic sums of each decade across both cemeteries, colour-coded by ceramic phase.
-Note that, for clarity, tomb G in the EG period, and tomb P in the late EO-LO period are excluded, with values ranging from 29-34, and 36-44 respectively.
+Figure \@ref(fig:bigbox) is a boxplot for the aoristic sums of each decade across both cemeteries, colour-coded by ceramic phase.
+Note that, for visual clarity, the outliers tomb G in the EG period, and tomb P in the late EO-LO period are excluded, with values ranging from 29-34, and 36-44 respectively.
 The important observation is that the PGB and EG exhibit the largest interquartile ranges (IQRs) of any period, and each possess only a single outlier (beyond 1.5 times the IQR).
-The Orientalising period, on the other hand, exhibit much reduced IQRs, and a proliferation of outliers far and above the range of the other tombs.
-In other words, while both the PGB-EG and EO-LO periods witnessed an expanding volume and range of burial activity, in the earlier period this was shared somewhat more equally among the total burying population.
+The Orientalising period, on the other hand, exhibits much reduced IQRs, and a proliferation of outliers far and above the range of the other tombs.
+In other words, while both the PGB-EG and EO-LO periods witnessed an expanding volume and range of burial activity, in the earlier period this was distributed somewhat more evenly across the total burying population.
 One speaks of increasing plurality, the other of divergence.
@@
 
-This possibility is furthered somewhat by considering the gini coefficient of the distribution of pottery between the tombs.
-Using the aoristic sums as a probablistically-weighted estimate of the volume of material deposited in each tomb in each decade, use of the gini coefficient can give an (albeit rough) estimate of how unevenly distributed that material was.
+This possibility is supported further by considering the gini coefficient of the distribution of pottery between the tombs.
+Using the aoristic sums as a probablistically-weighted estimate of the volume of material deposited in each tomb in each decade, use of the gini coefficient can give an (albeit rough) estimate of how unequally distributed that material was.
 And what we see is that, though values were always relatively high (i.e.
@@
 
-Figure ...
-shows the composition of the total pottery assemblage from each cemetery, and from the seven tombs drawn out earlier, through the centuries.
+Figure \@ref(fig:comps) shows the composition of the total pottery assemblage from each cemetery, and from the seven tombs drawn out earlier, through the centuries.
 In all cases, there appears to be the steady crystallisation of a basic 'burial kit', comprising a pithos and lid, an oil vessel, a jug, and a drinking vessel.
@@
 But I do want to briefly comment on one aspect of the distribution of non-ceramic artefacts that may make more sense in the light of the foregoing discussion.
-There is, perhaps unsurprisingly, a relatively strong (Spearman's rank correlation of `r spearmanobjectspots`) correlation between the number of ceramic and non-ceramic finds across the tombs, and, certainly, many of the most unique and precious objects were recovered from among the larger tombs.
+There is, perhaps unsurprisingly, a relatively strong (Spearman's ρ=`r spearmanobjectspots`) correlation between the number of ceramic and non-ceramic finds across the tombs, and, certainly, many of the most unique and precious objects were recovered from among the larger tombs.
 But, the monopoly those tombs would appear to have over such objects, perhaps tempting one initially to speak of 'elite' or 'high-status' tombs, evaporates when the number of those objects is divided by the number of burials represented.
@@
 
-Figure ...
-thus supports the view which has emerged from the preceeding analysis, that the major increase in burial activity at both Fortetsa and the KNC was not a phenomenon linked in any straightforward way to simple principles of 'elite' or 'wealthy' burial.
-To summarise, towards the end of 8th century BC, a select group of tombs, all of which had been constructed back in the 9th or 10th, became the foci of a major increase in burials, something which did not occur among the tombs built around the same time.
-We have seen how this surge in acivity occurred against the backdrop of a declining number of tombs dating back to the 10th and 9th centuries, and I have suggested that the rates of change are too rapid, and unevenly distributed, to be explicable through population growth alone.
-We have seen how little about the physical dimensions of these tombs presaged their later popularity, and that, while a similar, if smaller, increase in burial activity occurred around a century earlier, in the PGB-EG periods, this contrasts in the comparative diversity of its associated assemblages, and their distribution across the cemetries.
+Figure \@ref(fig:hists) thus supports the view which has emerged from the foregoing analysis, that the major increase in burial activity at both Fortetsa and the KNC was not a phenomenon linked in any straightforward way to simple principles of 'elite' or 'wealthy' burial.
+To summarise, towards the end of 8th century BC, a select group of tombs, all of which had been constructed back in the 9th or 10th, became the foci of a major increase in burials, something which did not occur among the tombs built around that time.
+We have seen how this surge in acivity occurred against the backdrop of a declining number of older tombs in use, and I have suggested that the rates of change are too rapid, and unevenly distributed, to be explicable through population growth alone.
+We have seen how little about the physical dimensions or early use of these tombs presaged their later popularity, and that, while a similar, if smaller, increase in burial activity occurred around a century earlier, in the PGB-EG periods, this contrasted in the comparative diversity of its associated assemblages, and their distribution across the tombs.
 The later burials were marked by great uniformity in the associated grave goods, and this apparently limited and consistent repertoire may explain why many of the most used tombs prove exceptional only in the absolute quantity of artefacts, and not the accoutrements of each individual burial.
@@
 
+The preceding analysis has revealed a number of potentially significant patterns in the use of the Fortetsa and Knossos North Cemeteries, in particular the exceptional trajectories of a select group of 10th and 9th century tombs.
+It only remains to consider how we whould characterise the burying groups responsible for these observed trends.
+The traditional view is that such groups were immediate families (Brock 1957, 41; Boardman 1967, 63; Coldstream & Huxley 1999; 291).
+The relatively short lifespan of many tombs (often only two or three generations) has been taken as evidence for direct, lineal inheritance, with the tomb going out of use as families moved, individuals founded new tombs, or the line came to an end.
+The idea of lineal groups is supported by the, admittedly circumstantial, evidence of men, women and children being found within single tombs, and one (contentious) instance of an apparently hereditary mandibular deformity in three non-contemporary skulls in the same tomb (Musgrave 1996).
+
+The model of lineal inheritance has been recently challenged by Kotsonas (2011, 131), however, who considers it ‘probably the greatest factoid in the archaeology of Crete of the given period’.
+He argues that the rise in interments during the 8th century, noted also in the present study, is too great to be explained by population growth, as would be necessary for the possibility of direct lineal inheritance to be preserved.
+Certainly, it seems illogical that population increases would lead to the consolidation of burials within only extant tombs, rather than increasing tomb construction.
+The trend instead must reflect choices made by the burying groups themselves, which Kotsonas feels may have comprised various elements, including perhaps age- and sex-grades.
+But it seems probable that such corporate groups, if they existed intersected with, and existed alongside, real or fictive kinship networks.
+
+Yet it is hard to be conclusive, as our current understanding of the social organisation of the early Cretan *poleis* is patchy at best.
+Perlman and Gagarin's [-@Gagarin2016] recent study of the Cretan law codes reminds us, importantly, if lamentably, that there is an often underappreciated degree of diversity apparent from the Archaic inscriptions in the offices, institutions, and social groups of the early city-states.
+In the traditional view, political authority was concentrated in the hands of the *kosmoi*, aristocratic magistrates selected from the leading, hereditary *startoi*, or ‘clans’, networks of extended lineages controlling the ancestral estates on which worked the serfs and/or slaves who made up the lower rungs of the social hierarchy (Willetts 1982, 240).
+But despite elements of this account probably being accurate, others remain equivocal, such as the nature of the *pyla*, or 'tribe', and its relation to the *startos* (which may itself be a military grouping), the distinction between an *astos* and a *poliatas*, both clearly some kind of formulation of the rights-bearing civic actor, and the ancestry and spread of the *agelai*, apparently an adolescent age-grade so far only attested at Eltynia [@Gagarin2016, 36; 78-80].
+Indeed, many terms which might illuminate social divisions and groupings within the Archaic Cretan populace are only attested at one or two sites, and even then sometimes appear to differ in their form or function.
+Frustrating though this can be, it has also seen the image of Cretan austerity and uniformity begin to be challenged, or at least nuanced, by an appreciation of the regional variety and idiosyncracy apparent from the laws.
+
+It wouldn't be helpful at present, then, to draw any simple parallel between the Knossian burial groups and, say, powerful *startoi*, a term which is not even attested outside of Gortyn in this period (ibid., 1-2).
+But archaeological and textual evidence suggests that, in general terms, such supra-familial kinship groups were undergoing a transition during this period, in ways reflected at Knossos, and thus supporting the attribution of the chamber tombs to such groups.
+Elsewhere on Crete, changes around this time include a shift to unmarked, individual burials, and more austere grave goods, as well as an apparent reorientation of sanctuary deposits away from extra-mural to suburban spaces, and a rise in utilitarian, mass-produced votives (Perlman & Gagarin 2016, 34-6).
+At Azoria, major urban remodelling in the 7th century saw the earlier, agglutinative house forms of neighbouring Kavousi Vronda and Kastro give way to larger, more formally planned dwellings with little evidence of renovation or remodelling, while larger, apprently civic, buildings were constructed to accommodate major public gatherings.
+At Knossos, a small number of pithos burials in pits do appear in the KNC and Fortetsa volumes (Brock 1957, 98; Coldstream & Catling 1996, 162), and recent re-excavation of the Teke plot identified numerous small pit burials around three chamber tombs (Evely 2012), raising the possibility that many more such burials filled the interstices between larger chamber tombs elsewhere, but have gone unrecognised.
+The insecure dates of these pits, however, makes it unclear whether they represent a genuine Archaic trend.
+
+These varied lines of evidence may all suggest that with the emergence of the archaic *poleis* on Crete, there was a concomitant restructuring of wider social groups, institutions, and behaviours.
+Perlman and Gagarin (Gagarin & Perlman 2016, 32-7) argue that, at least by the Classical period, supra-household kinship units such as the *startos* or *pyla*, whatever their exact formulations, had become mostly ceremonial, political entities, and had little influence over the social organisation of the Cretan city states.
+Was this transformation linked to an emerging emphasis on the citizen, the state, and ideologies of relative egality, within which communal gatherings, intra-mural ritual, and individual burial replaced ostentatious, kin-centred acts of gathering, dedication, and burial?
+It is plausible, though as Perlman and Gagarin (32-37) concede, the decline of ostentatious burial could just as well represent a voluntary abandonment by groups assured of their economic and political influence, finding the symbolic expression of kin-relations in death no longer a necessary expense.
+
+These issues, naturally, recall the conclusions of Morris (1987, 171-211), who argued 30 years ago that the emergence of the *polis* across the Aegean led to a tension between elite preferences for conspicuous display, and a downplaying of social difference in death, as promoted by the ‘middling’ ideology that the *polis* engendered and promulgated.
+In many areas, this resulted in increasing restriction of elaborate tombs, despite a ‘rise in aristocratic family consciousness’ that saw certain wealthy tombs buck the trend, until ‘such overt groupings were subsumed within the larger “citizen cemeteries”’ (ibid., 185).
+In his well-known model of the *agathoi* and *kakoi*, Morris argued that, at various times, the relevant spheres of kinship demonstrated in burial could widen and constrict, with more and less distantly related lineages drawn into association through changing social strategies, until the pre-eminence of kinship relations in the structuring of political power declined.
+
+Returning to Knossos, while aspects of the noted expansion in the burying population would seem readily explicable within a model like Morris', certain observations sit more uneasily.
+Firstly, the burials at issue here are in many ways *not* ostentatious in comparison to the norm.
+We have seen that the 'richest' tombs in terms of finds are not so 'rich' after all, when one takes account of the number of burials.
+They are not phyiscally larger, and do not exhibit monopolies over any of a range of potentially 'valuable' offerings.
+Likewise, Morris claims that periods of expansion in the rite of burial will correspond to greater variety in the associated assemblages, yet, at Knossos, we have seen that the range of grave goods in fact diminished and crystallised during the late 8th and early 7th centuries.
+Perhaps the expansion was driven by a widening of access to the burial rite by the so-called *agathoi*, with the standardised burial assemblages coming to reflect their shared affinal relations.
+Or else, the widening of access to these tombs actually heralds the breakdown in traditional kin-based stewardship thereof, perhaps with new, non-lineal corporate groups becoming  the custodians of the tombs, with the limited funereal repertoire thus reflecting the levelling ideology of the emergent state.
+The exclusive focus of the expansion in a very limited number of older tombs, though, would seem to argue in favour of the former.
+What seems certain is that this was a period of transition wherein, whatever their identity, certain actors invested heavily in a final flurry of chamber tomb burials, with an emphasis on tradition, an established repertoire of items, rites, and venues for one's passage to the afterlife.
+
+And this image ties into the broader picture developing of the island at this time.
+It is telling that the peak for imported items in the tombs came in the 9th century; the 'Orientalising' as a time period feels increasingly misapplied to Crete, when, at Knossos at least, it witnessed far lower rates of oriental imports, and only limited adoption of the figural motifs which characterise its appearance elsewhere.
+The observed rise in imitations of Near Eastern vessels in the late 8th and early 7th century further accords with our growing awareness of the selective nature of archaic and classical Crete's external relations.
+Erickson has, contrary to many, argued that *symposia* were practised on Crete, but that their practitioners simply favoured simpler vessel forms.
+The local, selective adoption of, and responses to, wider social and material trends across the contemporary Aegean is perhaps becoming the hallmark of archaic Crete, and finds further justification in the trends identified in the EIA cemeteries at Knossos.
+
 # Conclusion
+
+It is important that the present study was a comparative investigation of both the Fortetsa and Knossos North Cemetery.
+It is notable how similar the observed trends are across both these cemeteries, adding significant credence to the inferences drawn.
+That both underwent similar developments during the EIA suggests not only that they catered to similar sub-sections of the Knossian population, but that some of the socio-economic and political changes inferred and discussed were real, and community-wide.
+We should be cautious in speculating as to whether, for instance, particular families, lineages, groups, or classes made persistent use of these different necropoleis, but such possiblities are not inconceivable.
+Further work should seek to extend these comparisons, both into trends unexplored in the present study (such as, for instance, the curious reuse of a number of 11th century tombs in the late 8th and early 7th centuries), and out to the wider funerary landscape of Knossos.
+I hope to make just such a study but, for the time being, the datasets I am making available alongside this article include the other published EIA tombs in the area, so that others may also explore their possible significance.
+
+There were undeniable limitations to the present study, also.
+For one, non-ceramic artefacts were underrepresented, mostly because they lack accurate dating in the manner of pottery and future work could focus on dating these artefacts more closely, permitting their future inclusion into temporal analyses.
+Issues of gender were sadly also little considered, though this was largely due to limitations in the data themselves.
+There have also been a number of tombs discovered in the last 20 years that fill in the spaces between the Teke plot and the Medical Faculty Site, and ideally these unpublished tombs should be included in further analyses as they, along with the unconsidered Teke tholos and associated tombs, are almost surely part of a single contiguous cemetery (Blackman 1999, 113; Catling 1983, 51; Evely 2012; French 1991, 68–9).
+Finally, a preliminary study was recently taken by the author of select boxes marked as 'unpublished' from the KNC deposits kept at the British School at Athens at Knossos.
+Despite going unmentioned in the publication, this material was, at least in the boxes seen, not inconsiderable and so, regretably, future study will be necessary to determine whether it threatens the integrity of quantitative analyses such as this.
+Suffice to say that, at present, it is another pinch of salt with which to take the findings.
+
+These limitations notwithstanding, the foregoing analysis demonstrates the value of quantitative approaches to the archaeological record of the period.
+Such methods allow us not only to identify types of variability in the archaeological material, but also assess the scale of that variability.
+With large datasets, such as those used in the present study, we are able to locate and visualise covariance and significant correlations between different variables, plot changes in the data through time and, ultimately, draw on our findings to inform discussions of social processes.
+Though the construction of databases can be time consuming, once made they offer the opportunity to subdivide, compare, and plot the data in numerous ways quickly and easily, facilitating investigation of the material from multiple angles.
+Such methods are also more independently verifiable than conclusions drawn from more qualitative analyses, for two main reasons.
+Firstly, though explanations for any patterning in the data still require interpretation, the existence of such patterns may be more easily and reliably assessed via the use of quantitative and statistical tests of correlation and covariance.
+And secondly, by making datasets available for use by others, as with the data and code for the present study, one’s conclusions can be independently tested, and alternative interpretations can be forwarded, by other scholars.
+It is hoped that the present study demonstrates the wealth of insights yet to be uncovered by the application of more systematic, quantitative methods.
+The databases produced for the above analysis contain a large of amount of unutilised data that could facilitate a range of future investigations, and all are welcome to it.
+With more such analyses, we could begin to a construct a fuller account of the regional diversity in mortuary, and wider social, practices that characterised the communities of EIA and archaic Crete.
 

### 2019-05-24 analysis/paper/JMAArticle.Rmd
Further edits and new Figures

--- 2019-05-23 analysis/paper/paper.Rmd
@@
 
+tombs<-read.csv("../data/raw_data/KnossosTombs.csv",header=TRUE)
+tombs<-subset(tombs,tombs$cem_condensed=="KNC"|tombs$cem_condensed=="FOR")
+
 ```
@@
 ## The Early Iron Age Cemeteries
-```{r map, fig.cap="Map of the Knossos area showing the locations of the various tomb clusters and cemeteries, with EIA pottery distributions from the KULP marked in orange"}
-knitr::include_graphics("../figures/Knossos Tombs Map.png")
-```
 
@@
 
-```{r burialestimate, fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="Relationship between known interments, counts of pithoi, and counts of pithoi, amphorae and kraters.
-The straight line represents the mean estimate based on these three values."}
-
-#The tombs database is imported and, for the purposes of this article, the entire database is subsetted to include only those tombs from the Knossos North and Fortetsa cemeteries (excluding the Tekke Tombs).
-
-tombs<-read.csv("../data/raw_data/KnossosTombs.csv",header=TRUE)
-tombs<-subset(tombs,tombs$cem_condensed=="KNC"|tombs$cem_condensed=="FOR")
-
-#Figure 1 is plotted, with axis limits set (excluding tomb P, resulting in the warning message delivered) and the style of the plot adjusted (including font and text size, the appearance of the gridlines, selecting a colour palette etc.)
-
-p<-ggplot(tombs)+geom_point(aes(est_burials,all_urns,color="All urns"))+
-  geom_point(aes(est_burials,total_burials,color="Known interments"))+
-    geom_point(aes(est_burials,pithoi,colour="Pithoi"))+
-      geom_line(aes(est_burials,est_burials),alpha=0.4)+
-       scale_x_continuous(expand=c(0,0),limits=c(0,55))+
-        scale_y_continuous(expand=c(0,0),limits=c(0,55))+theme_bw()+
-         theme(legend.title=element_blank(),
-          text=element_text(family="Garamond", size=12))+
-              labs(x="Mean Burial Estimate", y="Interment/Vessel Count")+
-                 scale_color_brewer(palette = "Set2")
-
-#The coordinates of the plot are adjusted to render the axes equal, and the corresponding gridlines as squares, and then the figure is plotted.
-
-p<-p+coord_fixed()
-plot(p)
-
-```
+Insert Figure \@ref(fig:burialestimate) around here
 
@@
 
-```{r allpithoi,fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="Counts of pithoi across both cemeteries through time.
-Vessel dates have here been based on the midpoint of the absolute date range accompanying their respective ceramic periods."}
-
-#The database of all vessels is imported and, again only those tombs from the KNC and Fortetsa cemetery are selected.
-A new object "pithoi" is made by subsetting only those vessels labelled "Pithos", and a similar object "all_urns" by subsetting all pithoi, amphorae and kraters.
-
-pots<-read.csv("../data/raw_data/KnossosPots.csv",header=TRUE)
-pots<-subset(pots,pots$cem_condensed=="FOR"|pots$cem_condensed=="NC")
-pithoi<-subset(pots,pots$shape=="Pithos")
-all_urns<-subset(pots,pots$shape=="Pithos"|pots$shape=="Amphora"|pots$shape=="Krater")
-
-#The figure is plotted by creating a histogram of all vessels based on their average dates ("ave_date"), which is simply the midpoint of the absolute chronological range associated with their respective ceramic phase.
-Similar stylistic adjustments are made (font, labels, colour palette, removing a legend, setting the x and y scales, and expanding the borders of the plot slightly to prevent the "600" label being cut off, which would occur without the final line of code.)
-
-ggplot(pots, aes(ave_date,fill="Pots"))+geom_histogram()+
-  theme_bw()+theme(legend.title=element_blank(),
-      text=element_text(family="Garamond", size=12))+
-          labs(x="Date (Years BC)", y="Vessel Count")+
-              scale_fill_brewer(palette = "Set3")+ 
-                theme(legend.position="none")+
-                  scale_y_continuous(expand=c(0,0),limits=c(0,600))+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-```
+Insert Figure \@ref(fig:allpithoi) about here
 
@@
 
-```{r aoristallcems, fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="Aoristic sums of all vessels across both cemeteries.
-This represents a probabilistically weighted visualisation of depositional activity through time."}
-
-#The aoristic values for every tomb are summed by decade, and then this object is converted to a data frame.
-
-aorist_all_tombs<-colSums(tombs[158:207])
-aorist_all_tombs<-data.frame(aorist_all_tombs)
-
-#The column of values is labelled "aoristic_sum", an vector named "dates" is created with values between 1100 and 610 (in that order) at intervals of 10, and then this used to rename the rows of the data frame.
-Finally, these "row names" are converted into a functioning row in their own right within the data frame (this makes them plottable as a variable) and then are converted into numeric variables, allowing the plotting of them along a continuous scale.
-
-colnames(aorist_all_tombs)<-c("aoristic_sum")
-dates<-seq(from = 1100, to = 610, by = -10)
-row.names(aorist_all_tombs)<-dates
-setDT(aorist_all_tombs, keep.rownames = "decade")
-aorist_all_tombs$decade<-as.numeric(aorist_all_tombs$decade)
-
-#The plot is produced, following the same stylistic adjustments of the previous plot.
-
-ggplot(aorist_all_tombs,aes(decade,aoristic_sum, fill="Aoristic Sum"))+
-  geom_col()+theme_bw()+theme(legend.title=element_blank(),
-      text=element_text(family="Garamond", size=12))+
-          labs(x="Date (Years BC)", y="Aoristic Sum")+
-              scale_fill_brewer(palette = "Set3")+ 
-                theme(legend.position="none")+
-                  scale_y_continuous(expand=c(0,0),limits=c(0,225))+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-```
+Insert Figure \@ref(fig:aoristallcems) around here
 
@@
 
-```{r aoristbycem, fig.cap="Aorist sums of pithoi (left) as a proxy for cremation burials, and all vessels (right).
-The sums are colour coded to show the values of the two cemeteries, and the total values.", fig.dim=c(8,4)}
-KNC<-subset(tombs,tombs$cem_condensed=="KNC")
-
-aorist_KNC_pithoi <- colSums(KNC[208:257])
-aorist_KNC_pithoi<-data.frame(aorist_KNC_pithoi)
-
-colnames(aorist_KNC_pithoi)<-c("aoristic_sum")
-row.names(aorist_KNC_pithoi)<-dates
-aorist_KNC_pithoi<-data.frame(aorist_KNC_pithoi)
-setDT(aorist_KNC_pithoi, keep.rownames = "decade")
-aorist_KNC_pithoi$decade<-as.numeric(aorist_KNC_pithoi$decade)
-
-FOR<-subset(tombs,tombs$cem_condensed=="FOR")
-
-aorist_FOR_pithoi <-colSums(FOR[208:257])
-aorist_FOR_pithoi<-data.frame(aorist_FOR_pithoi)
-
-colnames(aorist_FOR_pithoi)<-c("aoristic_sum")
-row.names(aorist_FOR_pithoi)<-dates
-aorist_FOR_pithoi<-data.frame(aorist_FOR_pithoi)
-setDT(aorist_FOR_pithoi, keep.rownames = "decade")
-aorist_FOR_pithoi$decade<-as.numeric(aorist_FOR_pithoi$decade)
-
-aorist_all_pithoi <- colSums(tombs[208:257])
-aorist_all_pithoi<-data.frame(aorist_all_pithoi)
-
-colnames(aorist_all_pithoi)<-c("aoristic_sum")
-
-row.names(aorist_all_pithoi)<-dates
-aorist_all_pithoi<-data.frame(aorist_all_pithoi)
-setDT(aorist_all_pithoi, keep.rownames = "decade")
-aorist_all_pithoi$decade<-as.numeric(aorist_all_pithoi$decade)
-
-aorist_pithoi_plot<-ggplot(NULL)+
-  geom_col(data=aorist_all_pithoi,aes(decade,aoristic_sum,fill=" All tombs  "))+
-    geom_col(data=aorist_KNC_pithoi,aes(decade,aoristic_sum,fill=" KNC  "))+
-      geom_col(data=aorist_FOR_pithoi,aes(decade,aoristic_sum, fill=" Fortetsa  "))+
-        theme_bw()+theme(legend.title=element_blank(),
-          text=element_text(family="Garamond", size=12))+
-            labs(x="Date (Years BC)", y="Aoristic Sum (Pithoi)")+
-              scale_fill_brewer(palette = "Set2")+ 
-                  scale_y_continuous(expand=c(0,0),limits=c(0,45))+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-aorist_KNC <- colSums(KNC[158:207])
-aorist_KNC<-data.frame(aorist_KNC)
-
-colnames(aorist_KNC)<-c("aoristic_sum")
-row.names(aorist_KNC)<-dates
-aorist_KNC<-data.frame(aorist_KNC)
-setDT(aorist_KNC, keep.rownames = "decade")
-aorist_KNC$decade<-as.numeric(aorist_KNC$decade)
-
-FOR<-subset(tombs,tombs$cem_condensed=="FOR")
-
-aorist_FOR <-colSums(FOR[158:207])
-aorist_FOR<-data.frame(aorist_FOR)
-
-colnames(aorist_FOR)<-c("aoristic_sum")
-row.names(aorist_FOR)<-dates
-aorist_FOR<-data.frame(aorist_FOR)
-setDT(aorist_FOR, keep.rownames = "decade")
-aorist_FOR$decade<-as.numeric(aorist_FOR$decade)
-
-aorist_pots_plot<-ggplot(NULL)+
-  geom_col(data=aorist_all_tombs,aes(decade,aoristic_sum,fill=" All tombs  "))+
-    geom_col(data=aorist_KNC,aes(decade,aoristic_sum,fill=" KNC  "))+
-      geom_col(data=aorist_FOR,aes(decade,aoristic_sum, fill=" Fortetsa  "))+
-        theme_bw()+theme(legend.title=element_blank(),
-          text=element_text(family="Garamond", size=12))+
-            labs(x="Date (Years BC)", y="Aoristic Sum (All Vessels)")+
-              scale_fill_brewer(palette = "Set2")+ 
-                  scale_y_continuous(expand=c(0,0),limits=c(0,225))+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-ggarrange(aorist_pithoi_plot,aorist_pots_plot,ncol=2,common.legend=TRUE, legend="bottom")
-```
+Insert Figure \@ref(fig:aoristbycem) around here
 
@@
 
-```{r tombconstruction, fig.dim=c(7.5,4), dpi=300, fig.cap="Visulations of tomb construction through time at both cemeteries.
-On the left are counts based on the earliest dated pottery in each tomb, while on the left are aoristic sums likewise based on the earliest ceramic-based assignations.
-In both cases, tombs lacking any securely dateable pottery (i.e.
-N/As) are excluded."}
-
-tomb_dates_aorist<-read.csv("../data/derived_data/tomb_dates_aorist.csv",header=TRUE)
-tomb_dates_aorist_KNC<-subset(tomb_dates_aorist,tomb_dates_aorist$cem_condensed=="NC")
-tomb_dates_aorist_FOR<-subset(tomb_dates_aorist,tomb_dates_aorist$cem_condensed=="FOR")
-
-aorist_tomb_periods<-colSums(tomb_dates_aorist_KNC[5:14])
-aorist_tomb_periods<-data.frame(aorist_tomb_periods)
-aorist_tomb_periods$FOR<-colSums(tomb_dates_aorist_FOR[5:14])
-colnames(aorist_tomb_periods)<-c("KNC","FOR")
-setDT(aorist_tomb_periods, keep.rownames = "period")
-aorist_tomb_periods$period<-factor(aorist_tomb_periods$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
-aorist_tomb_periods_melt<-melt(aorist_tomb_periods)
-colnames(aorist_tomb_periods_melt)<-c("period","cemetery","aoristic_sum")
-
-tombs$start_period<-factor(tombs$start_period,levels=c("SM", "EPG", "MPG", "PG","LPG", "PGB", "EG", "MG","G", "LG", "EO","O","LO"))
-
-tombs$cem_condensed<-factor(tombs$cem_condensed,levels=c("KNC","FOR"))
-
-new_tomb_counts_plot<-ggplot(data=subset(tombs,!is.na(start_period)),aes(start_period, fill=cem_condensed))+geom_bar(position="identity")+theme_classic()+
-        theme_bw()+theme(legend.title=element_blank(),
-          text=element_text(family="Garamond", size=12))+
-            labs(x="Period", y="Count of New Tombs")+
-              scale_fill_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+ 
-                 scale_y_continuous(expand=c(0,0),limits=c(0,30))
-
-new_tomb_aorist_plot<-ggplot(aorist_tomb_periods_melt,aes(period,aoristic_sum, fill=cemetery))+geom_col(position="identity")+
-  theme_bw()+theme(legend.title=element_blank(),
-          text=element_text(family="Garamond", size=12))+
-            labs(x="Period", y="Aoristic Sum of New Tombs")+
-              scale_fill_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+ 
-                 scale_y_continuous(expand=c(0,0),limits=c(0,30))
-
-ggarrange(new_tomb_counts_plot,new_tomb_aorist_plot,ncol=2,common.legend=TRUE, legend="bottom")
-
-```
+Insert Figure \@ref(fig:tombconstruction) around here
 
@@
 
+Insert Figure \@ref(fig:tombsinuse) around here
+
+Such countervalent trends are observable too in the number of tombs receiving in each period.
+Figure \@ref(fig:tombsinuse) presents the estimated number of tombs in use throughout the EIA at both cemeteries.
+At the Forteta, this number remained relatively stable from the MPG to LG periods before declining (and bear in mind, there were only around 20 tombs belonging to this cemetery throughout the period), while at the KNC, the numbers of tombs in use rose to a preak in the late 8th century before declining.
+The cessation of burials there certainly appears more sudden and marked than at Fortetsa, though, again, it was always a larger cemetery.
+But this is not the whole story.
+
+Insert Figure \@ref(fig:tombsinusebybuilddate) around here
+
+Figure \@ref(fig:tombsinusebybuilddate) shows the number of tombs in use in each period, divided into the centuries in which those tombs were constructed, all overlaid on the total number of tombs in use across both cemeteries.
+A wave-like pattern can be observed with each century marked by the construction of new tombs which then steadily go out of use, as new ones are built.
+This would seem to accord with Cavanagh's suggestion that "[t]he length of time the tombs last seems [...] to reflect the duration of the social group which had access to the sepulchre", and that, if such access was lineally inehrited, frequent fission would result in limited trans-generational use.
+We might expect, then, that it was the 8th and 7th century tombs coming into use that spurred the dramatic rise in burial rates previoulsy noted in this period.
+But this is not the case.
+
+Insert Figure \@ref(fig:potsbydateandtombdate) around here
+
+Figure \@ref(fig:potsbydateandtombdate) shows, firstly, the number of vessels dating to each century across both cemeteries, based on their assigned ceramic period.
+But on the right, those pots are divided into the centuries to which the tombs they were found in originally date.
+So as can be seen, though the late 8th and early 7th centuries represented the apex of burial activity across both cemeteries, it was not the tombs built in those centuries which received the lion's share of the burials and associated ceramics.
+This is all the more notable when, returning to Figure \@ref(fig:tombsinusebybuilddate), we see that the number of 9th and 10th century tombs in use by the late 8th and early 7th centuries was declining.
+
+Insert Table 1 around here
+
+The significance of these findings can be seen in Table 1.
+Using pithoi as a proxy for burials, we can estimate the changing rates of burial through the centuries based on how many tombs yielded how many pithoi of each period.
+And while for the most part burial rates remained around three per tomb per century, in the 8th and 7th centuries, tombs founded in the 9th century saw a rapid rise in the rate of burial, peaking at over 11 per tomb per century.
+Though these estimates are no doubt very rough, the marked nature of this divergence suggests it is not simply an artefact of imperfect data.
+
+Also apparent from Table 1, though, are the markedly low rates of burial which predominated at both cemeteries.
+Two to four burials per tomb per century amounts to a very selective burial rite, even considering the restricted portion of the population probably represented by these cemeteries [@Cavanagh1996, 664].
+It seems unlikely, then, that the rising rates of burial noted above are, for instance, the direct result of population growth; the increase seems too rapid, and is concentrated only in certain (older) tombs, not to mention that, based on KULP’s survey data, the major population increase at EIA Knossos should be dated a century or more before the late 8th century peak in burial, in the Protogeometric period (Whitelaw, per comm.).
+
+##Isolating the phenomenon
+
+We are yet to draw out *which tombs exactly* are responsible for the observible patterning so far discussed.
+We might assume that the tombs which became the focus of intense burial activity in the late 8th and early 7th centuries would be those which yielded the most burials and grave goods when excavated and, happily, this supposition proves true.
+
+Insert Figure \@ref(fig:toptombs) around here
+
+```{r calc}
+
+tombs$group<-ifelse(tombs$tomb=="P"|tombs$tomb=="292"|tombs$tomb=="75"|tombs$tomb=="107"|tombs$tomb=="285"|tombs$tomb=="II"|tombs$tomb=="218","A","B")
+total_pottery_by_groups<-aggregate(tombs$total_pottery, by=list(tombs$group), FUN=sum)
+calc1<-(total_pottery_by_groups[1,2]/sum(total_pottery_by_groups$x))*100
+
+```
+The 12 tombs with the most associated vessels are all chamber tombs, with 10 dating to the 9th century, and two to the 10th.
+And, in Figure \@ref(fig:toptombs), we see that seven of the 12 did exhibit late, intense episodes of deposition.
+Tombs P, 292, 75, 107, 285, II, 218, and 104 between them accrount for `r round(calc1,digits=2)`% of the entire pottery assemblage recovered from the `r nrow(tombs)` tombs considered here.
+All but Tomb 285 date to the 9th century, and together they underpinned the late spate of burials here considered.
+Interestingly, Tombs X, G, Q, 104 and 283, however, evince a different pattern, with their respective peaks coming in the PGB and EG periods, around a century earlier.
+We have seen already in Figure \@ref(fig:aoristbycem) that, particularly in the KNC, the end of the 9th century was another period of increased burial activity.
+What, then, set these two groups of tombs apart?
+The recipients of the greatest numbers of burials and grave goods in both cemeteries, they nonetheless seem to have achieved that distinction via different histories.
+By looking more closely at the periods in which they were built, and in which they were most popular, I hope to shed light on the broader social contexts in which these tombs were flourished.
+
+##Not all tombs are created equal
+
+The indices discussed so far – the distribution of of artefacts between the tombs, and the construction and use of the tombs through time – only inform us of the temporal or aggregate features of the cemeteries; they tell us little of the tombs as they were built or first used.
+Synchronic comparisons are harder to make with the available data, but the dimensions of the tombs offer some scope for doing so.
+Perhaps the tombs which went on to receive the most burials were notably large or grand when first built, reflecting the status of the associated burying group.
+
+Insert Figure \@ref(fig:tombdimensions) around here
+
+Looking at Figure \@ref(fig:tombdimensions), however, it seems that no clear relationship existed between the physical size of a tomb and its buried population at EIA Knossos.
+It is true that over half of all tombs do not exceed 2.6 sq m in area, while nine of the ten tombs with the most burials do surpass this value.
+But many of the very largest tombs are decidedly middling when it comes to burial numbers.
+Things are even less clear for the *dromos*, the most visible part of the tomb, and a plausible arena for demonstrations of scale and grandeur during the burial itself.
+The very largest two tombs by burial numbers also had exceptionally long *dromoi*, but the next eight tombs with the largest *dromoi* do not rank highly when it comes to interments.
+There are difficulties here in that the *dromoi* especially were common victims of encroachment by later tombs, and our estimates of their original length are tentative in many cases.
+
+Insert Figure \@ref(fig:dimensionsboxplots) around here
+
+Looking at Figure \@ref(fig:dimensionsboxplots), though, we can see that, through the centuries, the dimensions of the tombs varied, and sometimes consistently, as with the trend toward narrower *dromoi*.
+The 9th-century tombs, including all but one of those seven noted above, do not particularly stand out.
+It does seem that chambers were, on average, larger in this period as well as varying more in size than in any other period.
+Then again, even the largest chambers are not exactly cavernous, and the extra labour expended on the construction of the larger chambers would have been limited.
+The *dromos* is a plausible area for conspicuous demonstrations of scale but, with their frequently poor preservation, we cannot be confident in assessing this.
+It appears at any rate that they were gradually declining in size, albeit with some notable outliers.
+
+So perhaps, at their time of building, such aspects of tomb construction carried social meaning but, even if they did, these Figures suggest this proved irrelevant to their later reuse.
+Several tombs had niches cut into their *dromoi*, or the *stomion* extended, to fit in later burial urns.
+Sometimes, urns from within the tomb were placed in such niches to make space for new interments in the chamber itself.
+There is an impression of individuals vying to be included in the chamber itself, of competition was for inclusion into the sepulchral group, supporting the idea that the tombs’ original builders little anticipated their eventual popularity and so did not scale their tombs accordingly.
+
+Insert Figure \@ref(fig:tombscompthroughtime) around here
+
+##Imports, interments, and indices of wealth
+
+Though it may not be immediately apparent from the dimensions of the tombs, however, the late 9th and early 8th centuries are consistently interpreted as periods of artistic development, increased overseas trade, and renewed vigour in the construction of tombs and the visibility of the burial rite.
+The PGB style reflects a uniquely Cretan blend of Attic and Near Eastern forms, while the appearance of Bronze Age 'antiques' in the tombs from this period onwards is felt to reflect renewed, even competitive interest in the Minoan past, as a tool for mortuary display or even lineage affirmation.
+
+Insert Figure \@ref(fig:imports) around here
+
+The rise in imported items is clearly seen in the tomb assemblages, with a succession of increased Greek and Near Eastern imports respectively in the late 9th and early 8th centuries (see Figure \@ref(fig:imports)).
+Yet beyond conforming to our prior inferences about the LPG-EG periods, the deposition of these imported wares may also cast light on the original pattern drawn out in the present analysis: the late surge in burial activity among some select 9th and 10th century tombs.
+
+Insert Figure \@ref(fig:importstombcenturies) around here
+
+In Figure \@ref(fig:importstombcenturies), we see that the tombs already built by the end of the PGB period, particularly those built in the 9th and, to a lesser extent, 10th centuries, were major beneficiaries of the influx of imported items during this time, but that, in the following centuries, in line with the rest of the cemetery, the number of imported items deposited within them declined.
+Imported ceramics are used here as something of a proxy for items of status value, a shorthand that would ideally be bolstered by other artefacts (such as precious vessels, weapons, jewellery etc.) which, owing to their less developed chronologies, must unfortunately be left out of such diachronic analyses for now.
+But, nonetheless, this pattern hints towards a possibility that shall now be explored, that the surge in burial activity seen across both cemeteries in the late 8th and early 7th centuries would appear to be of a different kind to that of a century earlier.
+
+Inset Figure \@ref(fig:bigbox)
+
+Figure \@ref(fig:bigbox) is a boxplot for the aoristic sums of each decade across both cemeteries, colour-coded by ceramic phase.
+Note that, for visual clarity, the outliers tomb G in the EG period, and tomb P in the late EO-LO period are excluded, with values ranging from 29-34, and 36-44 respectively.
+The important observation is that the PGB and EG exhibit the largest interquartile ranges (IQRs) of any period, and each possess only a single outlier (beyond 1.5 times the IQR).
+The Orientalising period, on the other hand, exhibits much reduced IQRs, and a proliferation of outliers far and above the range of the other tombs.
+In other words, while both the PGB-EG and EO-LO periods witnessed an expanding volume and range of burial activity, in the earlier period this was distributed somewhat more evenly across the total burying population.
+One speaks of increasing plurality, the other of divergence.
+
+Insert Figure \@ref(fig:gini) around here
+
+This possibility is supported further by considering the gini coefficient of the distribution of pottery between the tombs.
+Using the aoristic sums as a probablistically-weighted estimate of the volume of material deposited in each tomb in each decade, use of the gini coefficient can give an (albeit rough) estimate of how unequally distributed that material was.
+And what we see is that, though values were always relatively high (i.e.
+unequal) within the cemeteries, the major spike in the gini coefficient seen in the Orientalising period contrasts markedly with the modest rise and plateau of the PGB-EG.
+The gini coefficient is most commonly used in the assessment of income inequality, but I want to stress here that it is not "wealth" in any meaningful sense that we are talking about here.
+Indeed, if anything, the late flourit of burial activity here discussed may be a case of quantity, and not quality.
+
+Insert Figure \@ref(fig:comps) around here
+
+Figure \@ref(fig:comps) shows the composition of the total pottery assemblage from each cemetery, and from the seven tombs drawn out earlier, through the centuries.
+In all cases, there appears to be the steady crystallisation of a basic 'burial kit', comprising a pithos and lid, an oil vessel, a jug, and a drinking vessel.
+The proportion of grave goods comprising 'other' vessels - pyxides, askoi, trays, stands, dinoi, kernoi, kraters, lekanai and more - declines to become only a minor part of the standardised repertoire of the 7th century.
+The seven tombs that underpinned the dramatic increase in burial activity in the later 8th century did not break with this mould, and their assemblages seem as conventional as any other during the last century of use in the cemeteries.
+
+It has no doubt been apparent that the main focus of this article has been on pottery and not other artefacts of metal, stone, glass, and organic materials.
+This is largely because the chronologies for these artefacts are insufficiently developed to allow for the kind of diachronic analysis made thus far.
+But I do want to briefly comment on one aspect of the distribution of non-ceramic artefacts that may make more sense in the light of the foregoing discussion.
+There is, perhaps unsurprisingly, a relatively strong (Spearman's ρ=`r spearmanobjectspots`) correlation between the number of ceramic and non-ceramic finds across the tombs, and, certainly, many of the most unique and precious objects were recovered from among the larger tombs.
+But, the monopoly those tombs would appear to have over such objects, perhaps tempting one initially to speak of 'elite' or 'high-status' tombs, evaporates when the number of those objects is divided by the number of burials represented.
+
+Insert Figure \@ref(fig:hists) around here
+
+Figure \@ref(fig:hists) thus supports the view which has emerged from the foregoing analysis, that the major increase in burial activity at both Fortetsa and the KNC was not a phenomenon linked in any straightforward way to simple principles of 'elite' or 'wealthy' burial.
+To summarise, towards the end of 8th century BC, a select group of tombs, all of which had been constructed back in the 9th or 10th, became the foci of a major increase in burials, something which did not occur among the tombs built around that time.
+We have seen how this surge in acivity occurred against the backdrop of a declining number of older tombs in use, and I have suggested that the rates of change are too rapid, and unevenly distributed, to be explicable through population growth alone.
+We have seen how little about the physical dimensions or early use of these tombs presaged their later popularity, and that, while a similar, if smaller, increase in burial activity occurred around a century earlier, in the PGB-EG periods, this contrasted in the comparative diversity of its associated assemblages, and their distribution across the tombs.
+The later burials were marked by great uniformity in the associated grave goods, and this apparently limited and consistent repertoire may explain why many of the most used tombs prove exceptional only in the absolute quantity of artefacts, and not the accoutrements of each individual burial.
+
+# Discussion
+
+The preceding analysis has revealed a number of potentially significant patterns in the use of the Fortetsa and Knossos North Cemeteries, in particular the exceptional trajectories of a select group of 10th and 9th century tombs.
+It only remains to consider how we whould characterise the burying groups responsible for these observed trends.
+The traditional view is that such groups were immediate families (Brock 1957, 41; Boardman 1967, 63; Coldstream & Huxley 1999; 291).
+The relatively short lifespan of many tombs (often only two or three generations) has been taken as evidence for direct, lineal inheritance, with the tomb going out of use as families moved, individuals founded new tombs, or the line came to an end.
+The idea of lineal groups is supported by the, admittedly circumstantial, evidence of men, women and children being found within single tombs, and one (contentious) instance of an apparently hereditary mandibular deformity in three non-contemporary skulls in the same tomb (Musgrave 1996).
+
+The model of lineal inheritance has been recently challenged by Kotsonas (2011, 131), however, who considers it ‘probably the greatest factoid in the archaeology of Crete of the given period’.
+He argues that the rise in interments during the 8th century, noted also in the present study, is too great to be explained by population growth, as would be necessary for the possibility of direct lineal inheritance to be preserved.
+Certainly, it seems illogical that population increases would lead to the consolidation of burials within only extant tombs, rather than increasing tomb construction.
+The trend instead must reflect choices made by the burying groups themselves, which Kotsonas feels may have comprised various elements, including perhaps age- and sex-grades.
+But it seems probable that such corporate groups, if they existed intersected with, and existed alongside, real or fictive kinship networks.
+
+Yet it is hard to be conclusive, as our current understanding of the social organisation of the early Cretan *poleis* is patchy at best.
+Perlman and Gagarin's [-@Gagarin2016] recent study of the Cretan law codes reminds us, importantly, if lamentably, that there is an often underappreciated degree of diversity apparent from the Archaic inscriptions in the offices, institutions, and social groups of the early city-states.
+In the traditional view, political authority was concentrated in the hands of the *kosmoi*, aristocratic magistrates selected from the leading, hereditary *startoi*, or ‘clans’, networks of extended lineages controlling the ancestral estates on which worked the serfs and/or slaves who made up the lower rungs of the social hierarchy (Willetts 1982, 240).
+But despite elements of this account probably being accurate, others remain equivocal, such as the nature of the *pyla*, or 'tribe', and its relation to the *startos* (which may itself be a military grouping), the distinction between an *astos* and a *poliatas*, both clearly some kind of formulation of the rights-bearing civic actor, and the ancestry and spread of the *agelai*, apparently an adolescent age-grade so far only attested at Eltynia [@Gagarin2016, 36; 78-80].
+Indeed, many terms which might illuminate social divisions and groupings within the Archaic Cretan populace are only attested at one or two sites, and even then sometimes appear to differ in their form or function.
+Frustrating though this can be, it has also seen the image of Cretan austerity and uniformity begin to be challenged, or at least nuanced, by an appreciation of the regional variety and idiosyncracy apparent from the laws.
+
+It wouldn't be helpful at present, then, to draw any simple parallel between the Knossian burial groups and, say, powerful *startoi*, a term which is not even attested outside of Gortyn in this period (ibid., 1-2).
+But archaeological and textual evidence suggests that, in general terms, such supra-familial kinship groups were undergoing a transition during this period, in ways reflected at Knossos, and thus supporting the attribution of the chamber tombs to such groups.
+Elsewhere on Crete, changes around this time include a shift to unmarked, individual burials, and more austere grave goods, as well as an apparent reorientation of sanctuary deposits away from extra-mural to suburban spaces, and a rise in utilitarian, mass-produced votives (Perlman & Gagarin 2016, 34-6).
+At Azoria, major urban remodelling in the 7th century saw the earlier, agglutinative house forms of neighbouring Kavousi Vronda and Kastro give way to larger, more formally planned dwellings with little evidence of renovation or remodelling, while larger, apprently civic, buildings were constructed to accommodate major public gatherings.
+At Knossos, a small number of pithos burials in pits do appear in the KNC and Fortetsa volumes (Brock 1957, 98; Coldstream & Catling 1996, 162), and recent re-excavation of the Teke plot identified numerous small pit burials around three chamber tombs (Evely 2012), raising the possibility that many more such burials filled the interstices between larger chamber tombs elsewhere, but have gone unrecognised.
+The insecure dates of these pits, however, makes it unclear whether they represent a genuine Archaic trend.
+
+These varied lines of evidence may all suggest that with the emergence of the archaic *poleis* on Crete, there was a concomitant restructuring of wider social groups, institutions, and behaviours.
+Perlman and Gagarin (Gagarin & Perlman 2016, 32-7) argue that, at least by the Classical period, supra-household kinship units such as the *startos* or *pyla*, whatever their exact formulations, had become mostly ceremonial, political entities, and had little influence over the social organisation of the Cretan city states.
+Was this transformation linked to an emerging emphasis on the citizen, the state, and ideologies of relative egality, within which communal gatherings, intra-mural ritual, and individual burial replaced ostentatious, kin-centred acts of gathering, dedication, and burial?
+It is plausible, though as Perlman and Gagarin (32-37) concede, the decline of ostentatious burial could just as well represent a voluntary abandonment by groups assured of their economic and political influence, finding the symbolic expression of kin-relations in death no longer a necessary expense.
+
+These issues, naturally, recall the conclusions of Morris (1987, 171-211), who argued 30 years ago that the emergence of the *polis* across the Aegean led to a tension between elite preferences for conspicuous display, and a downplaying of social difference in death, as promoted by the ‘middling’ ideology that the *polis* engendered and promulgated.
+In many areas, this resulted in increasing restriction of elaborate tombs, despite a ‘rise in aristocratic family consciousness’ that saw certain wealthy tombs buck the trend, until ‘such overt groupings were subsumed within the larger “citizen cemeteries”’ (ibid., 185).
+In his well-known model of the *agathoi* and *kakoi*, Morris argued that, at various times, the relevant spheres of kinship demonstrated in burial could widen and constrict, with more and less distantly related lineages drawn into association through changing social strategies, until the pre-eminence of kinship relations in the structuring of political power declined.
+
+Returning to Knossos, while aspects of the noted expansion in the burying population would seem readily explicable within a model like Morris', certain observations sit more uneasily.
+Firstly, the burials at issue here are in many ways *not* ostentatious in comparison to the norm.
+We have seen that the 'richest' tombs in terms of finds are not so 'rich' after all, when one takes account of the number of burials.
+They are not phyiscally larger, and do not exhibit monopolies over any of a range of potentially 'valuable' offerings.
+Likewise, Morris claims that periods of expansion in the rite of burial will correspond to greater variety in the associated assemblages, yet, at Knossos, we have seen that the range of grave goods in fact diminished and crystallised during the late 8th and early 7th centuries.
+Perhaps the expansion was driven by a widening of access to the burial rite by the so-called *agathoi*, with the standardised burial assemblages coming to reflect their shared affinal relations.
+Or else, the widening of access to these tombs actually heralds the breakdown in traditional kin-based stewardship thereof, perhaps with new, non-lineal corporate groups becoming  the custodians of the tombs, with the limited funereal repertoire thus reflecting the levelling ideology of the emergent state.
+The exclusive focus of the expansion in a very limited number of older tombs, though, would seem to argue in favour of the former.
+What seems certain is that this was a period of transition wherein, whatever their identity, certain actors invested heavily in a final flurry of chamber tomb burials, with an emphasis on tradition, an established repertoire of items, rites, and venues for one's passage to the afterlife.
+
+And this image ties into the broader picture developing of the island at this time.
+It is telling that the peak for imported items in the tombs came in the 9th century; the 'Orientalising' as a time period feels increasingly misapplied to Crete, when, at Knossos at least, it witnessed far lower rates of oriental imports, and only limited adoption of the figural motifs which characterise its appearance elsewhere.
+The observed rise in imitations of Near Eastern vessels in the late 8th and early 7th century further accords with our growing awareness of the selective nature of archaic and classical Crete's external relations.
+Erickson has, contrary to many, argued that *symposia* were practised on Crete, but that their practitioners simply favoured simpler vessel forms.
+The local, selective adoption of, and responses to, wider social and material trends across the contemporary Aegean is perhaps becoming the hallmark of archaic Crete, and finds further justification in the trends identified in the EIA cemeteries at Knossos.
+
+# Conclusion
+
+It is important that the present study was a comparative investigation of both the Fortetsa and Knossos North Cemetery.
+It is notable how similar the observed trends are across both these cemeteries, adding significant credence to the inferences drawn.
+That both underwent similar developments during the EIA suggests not only that they catered to similar sub-sections of the Knossian population, but that some of the socio-economic and political changes inferred and discussed were real, and community-wide.
+We should be cautious in speculating as to whether, for instance, particular families, lineages, groups, or classes made persistent use of these different necropoleis, but such possiblities are not inconceivable.
+Further work should seek to extend these comparisons, both into trends unexplored in the present study (such as, for instance, the curious reuse of a number of 11th century tombs in the late 8th and early 7th centuries), and out to the wider funerary landscape of Knossos.
+I hope to make just such a study but, for the time being, the datasets I am making available alongside this article include the other published EIA tombs in the area, so that others may also explore their possible significance.
+
+There were undeniable limitations to the present study, also.
+For one, non-ceramic artefacts were underrepresented, mostly because they lack accurate dating in the manner of pottery and future work could focus on dating these artefacts more closely, permitting their future inclusion into temporal analyses.
+Issues of gender were sadly also little considered, though this was largely due to limitations in the data themselves.
+There have also been a number of tombs discovered in the last 20 years that fill in the spaces between the Teke plot and the Medical Faculty Site, and ideally these unpublished tombs should be included in further analyses as they, along with the unconsidered Teke tholos and associated tombs, are almost surely part of a single contiguous cemetery (Blackman 1999, 113; Catling 1983, 51; Evely 2012; French 1991, 68–9).
+Finally, a preliminary study was recently taken by the author of select boxes marked as 'unpublished' from the KNC deposits kept at the British School at Athens at Knossos.
+Despite going unmentioned in the publication, this material was, at least in the boxes seen, not inconsiderable and so, regretably, future study will be necessary to determine whether it threatens the integrity of quantitative analyses such as this.
+Suffice to say that, at present, it is another pinch of salt with which to take the findings.
+
+These limitations notwithstanding, the foregoing analysis demonstrates the value of quantitative approaches to the archaeological record of the period.
+Such methods allow us not only to identify types of variability in the archaeological material, but also assess the scale of that variability.
+With large datasets, such as those used in the present study, we are able to locate and visualise covariance and significant correlations between different variables, plot changes in the data through time and, ultimately, draw on our findings to inform discussions of social processes.
+Though the construction of databases can be time consuming, once made they offer the opportunity to subdivide, compare, and plot the data in numerous ways quickly and easily, facilitating investigation of the material from multiple angles.
+Such methods are also more independently verifiable than conclusions drawn from more qualitative analyses, for two main reasons.
+Firstly, though explanations for any patterning in the data still require interpretation, the existence of such patterns may be more easily and reliably assessed via the use of quantitative and statistical tests of correlation and covariance.
+And secondly, by making datasets available for use by others, as with the data and code for the present study, one’s conclusions can be independently tested, and alternative interpretations can be forwarded, by other scholars.
+It is hoped that the present study demonstrates the wealth of insights yet to be uncovered by the application of more systematic, quantitative methods.
+The databases produced for the above analysis contain a large of amount of unutilised data that could facilitate a range of future investigations, and all are welcome to it.
+With more such analyses, we could begin to a construct a fuller account of the regional diversity in mortuary, and wider social, practices that characterised the communities of EIA and archaic Crete.
+
+# Acknowledgements
+
+<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
+##### pagebreak
+
+# References 
+<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
+<div id="refs"></div>
+
+##### pagebreak
+
+# Captions
+
+##### pagebreak
+
+# Figures
+
+```{r map, fig.cap="Map of the Knossos area showing the locations of the various tomb clusters and cemeteries, with EIA pottery distributions from the KULP marked in orange"}
+knitr::include_graphics("../figures/Knossos Tombs Map.png")
+```
+
+```{r burialestimate, fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="Relationship between known interments, counts of pithoi, and counts of pithoi, amphorae and kraters.
+The straight line represents the mean estimate based on these three values."}
+
+#The tombs database is imported and, for the purposes of this article, the entire database is subsetted to include only those tombs from the Knossos North and Fortetsa cemeteries (excluding the Tekke Tombs).
+
+tombs<-read.csv("../data/raw_data/KnossosTombs.csv",header=TRUE)
+tombs<-subset(tombs,tombs$cem_condensed=="KNC"|tombs$cem_condensed=="FOR")
+
+#Figure 1 is plotted, with axis limits set (excluding tomb P, resulting in the warning message delivered) and the style of the plot adjusted (including font and text size, the appearance of the gridlines, selecting a colour palette etc.)
+
+p<-ggplot(tombs)+geom_point(aes(est_burials,all_urns,color="All urns"))+
+  geom_point(aes(est_burials,total_burials,color="Known interments"))+
+    geom_point(aes(est_burials,pithoi,colour="Pithoi"))+
+      geom_line(aes(est_burials,est_burials),alpha=0.4)+
+       scale_x_continuous(expand=c(0,0),limits=c(0,55))+
+        scale_y_continuous(expand=c(0,0),limits=c(0,55))+theme_bw()+
+         theme(legend.title=element_blank(),
+          text=element_text(family="Garamond", size=12))+
+              labs(x="Mean Burial Estimate", y="Interment/Vessel Count")+
+                 scale_color_brewer(palette = "Set2")
+
+#The coordinates of the plot are adjusted to render the axes equal, and the corresponding gridlines as squares, and then the figure is plotted.
+
+p<-p+coord_fixed()
+plot(p)
+
+```
+
+```{r allpithoi,fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="Counts of pithoi across both cemeteries through time.
+Vessel dates have here been based on the midpoint of the absolute date range accompanying their respective ceramic periods."}
+
+#The database of all vessels is imported and, again only those tombs from the KNC and Fortetsa cemetery are selected.
+A new object "pithoi" is made by subsetting only those vessels labelled "Pithos", and a similar object "all_urns" by subsetting all pithoi, amphorae and kraters.
+
+pots<-read.csv("../data/raw_data/KnossosPots.csv",header=TRUE)
+pots<-subset(pots,pots$cem_condensed=="FOR"|pots$cem_condensed=="NC")
+pithoi<-subset(pots,pots$shape=="Pithos")
+all_urns<-subset(pots,pots$shape=="Pithos"|pots$shape=="Amphora"|pots$shape=="Krater")
+
+#The figure is plotted by creating a histogram of all vessels based on their average dates ("ave_date"), which is simply the midpoint of the absolute chronological range associated with their respective ceramic phase.
+Similar stylistic adjustments are made (font, labels, colour palette, removing a legend, setting the x and y scales, and expanding the borders of the plot slightly to prevent the "600" label being cut off, which would occur without the final line of code.)
+
+ggplot(pots, aes(ave_date,fill="Pots"))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+      text=element_text(family="Garamond", size=12))+
+          labs(x="Date (Years BC)", y="Vessel Count")+
+              scale_fill_brewer(palette = "Set3")+ 
+                theme(legend.position="none")+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,600))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+```
+
+```{r aoristallcems, fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="Aoristic sums of all vessels across both cemeteries.
+This represents a probabilistically weighted visualisation of depositional activity through time."}
+
+#The aoristic values for every tomb are summed by decade, and then this object is converted to a data frame.
+
+aorist_all_tombs<-colSums(tombs[158:207])
+aorist_all_tombs<-data.frame(aorist_all_tombs)
+
+#The column of values is labelled "aoristic_sum", an vector named "dates" is created with values between 1100 and 610 (in that order) at intervals of 10, and then this used to rename the rows of the data frame.
+Finally, these "row names" are converted into a functioning row in their own right within the data frame (this makes them plottable as a variable) and then are converted into numeric variables, allowing the plotting of them along a continuous scale.
+
+colnames(aorist_all_tombs)<-c("aoristic_sum")
+dates<-seq(from = 1100, to = 610, by = -10)
+row.names(aorist_all_tombs)<-dates
+setDT(aorist_all_tombs, keep.rownames = "decade")
+aorist_all_tombs$decade<-as.numeric(aorist_all_tombs$decade)
+
+#The plot is produced, following the same stylistic adjustments of the previous plot.
+
+ggplot(aorist_all_tombs,aes(decade,aoristic_sum, fill="Aoristic Sum"))+
+  geom_col()+theme_bw()+theme(legend.title=element_blank(),
+      text=element_text(family="Garamond", size=12))+
+          labs(x="Date (Years BC)", y="Aoristic Sum")+
+              scale_fill_brewer(palette = "Set3")+ 
+                theme(legend.position="none")+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,225))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+```
+
+```{r aoristbycem, fig.cap="Aorist sums of pithoi (left) as a proxy for cremation burials, and all vessels (right).
+The sums are colour coded to show the values of the two cemeteries, and the total values.", fig.dim=c(8,4)}
+KNC<-subset(tombs,tombs$cem_condensed=="KNC")
+
+aorist_KNC_pithoi <- colSums(KNC[208:257])
+aorist_KNC_pithoi<-data.frame(aorist_KNC_pithoi)
+
+colnames(aorist_KNC_pithoi)<-c("aoristic_sum")
+row.names(aorist_KNC_pithoi)<-dates
+aorist_KNC_pithoi<-data.frame(aorist_KNC_pithoi)
+setDT(aorist_KNC_pithoi, keep.rownames = "decade")
+aorist_KNC_pithoi$decade<-as.numeric(aorist_KNC_pithoi$decade)
+
+FOR<-subset(tombs,tombs$cem_condensed=="FOR")
+
+aorist_FOR_pithoi <-colSums(FOR[208:257])
+aorist_FOR_pithoi<-data.frame(aorist_FOR_pithoi)
+
+colnames(aorist_FOR_pithoi)<-c("aoristic_sum")
+row.names(aorist_FOR_pithoi)<-dates
+aorist_FOR_pithoi<-data.frame(aorist_FOR_pithoi)
+setDT(aorist_FOR_pithoi, keep.rownames = "decade")
+aorist_FOR_pithoi$decade<-as.numeric(aorist_FOR_pithoi$decade)
+
+aorist_all_pithoi <- colSums(tombs[208:257])
+aorist_all_pithoi<-data.frame(aorist_all_pithoi)
+
+colnames(aorist_all_pithoi)<-c("aoristic_sum")
+
+row.names(aorist_all_pithoi)<-dates
+aorist_all_pithoi<-data.frame(aorist_all_pithoi)
+setDT(aorist_all_pithoi, keep.rownames = "decade")
+aorist_all_pithoi$decade<-as.numeric(aorist_all_pithoi$decade)
+
+aorist_pithoi_plot<-ggplot(NULL)+
+  geom_col(data=aorist_all_pithoi,aes(decade,aoristic_sum,fill=" All tombs  "))+
+    geom_col(data=aorist_KNC_pithoi,aes(decade,aoristic_sum,fill=" KNC  "))+
+      geom_col(data=aorist_FOR_pithoi,aes(decade,aoristic_sum, fill=" Fortetsa  "))+
+        theme_bw()+theme(legend.title=element_blank(),
+          text=element_text(family="Garamond", size=12))+
+            labs(x="Date (Years BC)", y="Aoristic Sum (Pithoi)")+
+              scale_fill_brewer(palette = "Set2")+ 
+                  scale_y_continuous(expand=c(0,0),limits=c(0,45))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+aorist_KNC <- colSums(KNC[158:207])
+aorist_KNC<-data.frame(aorist_KNC)
+
+colnames(aorist_KNC)<-c("aoristic_sum")
+row.names(aorist_KNC)<-dates
+aorist_KNC<-data.frame(aorist_KNC)
+setDT(aorist_KNC, keep.rownames = "decade")
+aorist_KNC$decade<-as.numeric(aorist_KNC$decade)
+
+FOR<-subset(tombs,tombs$cem_condensed=="FOR")
+
+aorist_FOR <-colSums(FOR[158:207])
+aorist_FOR<-data.frame(aorist_FOR)
+
+colnames(aorist_FOR)<-c("aoristic_sum")
+row.names(aorist_FOR)<-dates
+aorist_FOR<-data.frame(aorist_FOR)
+setDT(aorist_FOR, keep.rownames = "decade")
+aorist_FOR$decade<-as.numeric(aorist_FOR$decade)
+
+aorist_pots_plot<-ggplot(NULL)+
+  geom_col(data=aorist_all_tombs,aes(decade,aoristic_sum,fill=" All tombs  "))+
+    geom_col(data=aorist_KNC,aes(decade,aoristic_sum,fill=" KNC  "))+
+      geom_col(data=aorist_FOR,aes(decade,aoristic_sum, fill=" Fortetsa  "))+
+        theme_bw()+theme(legend.title=element_blank(),
+          text=element_text(family="Garamond", size=12))+
+            labs(x="Date (Years BC)", y="Aoristic Sum (All Vessels)")+
+              scale_fill_brewer(palette = "Set2")+ 
+                  scale_y_continuous(expand=c(0,0),limits=c(0,225))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+ggarrange(aorist_pithoi_plot,aorist_pots_plot,ncol=2,common.legend=TRUE, legend="bottom")
+```
+
+```{r tombconstruction, fig.dim=c(7.5,4), dpi=300, fig.cap="Visulations of tomb construction through time at both cemeteries.
+On the left are counts based on the earliest dated pottery in each tomb, while on the left are aoristic sums likewise based on the earliest ceramic-based assignations.
+In both cases, tombs lacking any securely dateable pottery (i.e.
+N/As) are excluded."}
+
+tomb_dates_aorist<-read.csv("../data/derived_data/tomb_dates_aorist.csv",header=TRUE)
+tomb_dates_aorist_KNC<-subset(tomb_dates_aorist,tomb_dates_aorist$cem_condensed=="NC")
+tomb_dates_aorist_FOR<-subset(tomb_dates_aorist,tomb_dates_aorist$cem_condensed=="FOR")
+
+aorist_tomb_periods<-colSums(tomb_dates_aorist_KNC[5:14])
+aorist_tomb_periods<-data.frame(aorist_tomb_periods)
+aorist_tomb_periods$FOR<-colSums(tomb_dates_aorist_FOR[5:14])
+colnames(aorist_tomb_periods)<-c("KNC","FOR")
+setDT(aorist_tomb_periods, keep.rownames = "period")
+aorist_tomb_periods$period<-factor(aorist_tomb_periods$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
+aorist_tomb_periods_melt<-melt(aorist_tomb_periods)
+colnames(aorist_tomb_periods_melt)<-c("period","cemetery","aoristic_sum")
+
+tombs$start_period<-factor(tombs$start_period,levels=c("SM", "EPG", "MPG", "PG","LPG", "PGB", "EG", "MG","G", "LG", "EO","O","LO"))
+
+tombs$cem_condensed<-factor(tombs$cem_condensed,levels=c("KNC","FOR"))
+
+new_tomb_counts_plot<-ggplot(data=subset(tombs,!is.na(start_period)),aes(start_period, fill=cem_condensed))+geom_bar(position="identity")+theme_classic()+
+        theme_bw()+theme(legend.title=element_blank(),
+          text=element_text(family="Garamond", size=12))+
+            labs(x="Period", y="Count of New Tombs")+
+              scale_fill_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+ 
+                 scale_y_continuous(expand=c(0,0),limits=c(0,30))
+
+new_tomb_aorist_plot<-ggplot(aorist_tomb_periods_melt,aes(period,aoristic_sum, fill=cemetery))+geom_col(position="identity")+
+  theme_bw()+theme(legend.title=element_blank(),
+          text=element_text(family="Garamond", size=12))+
+            labs(x="Period", y="Aoristic Sum of New Tombs")+
+              scale_fill_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+ 
+                 scale_y_continuous(expand=c(0,0),limits=c(0,30))
+
+ggarrange(new_tomb_counts_plot,new_tomb_aorist_plot,ncol=2,common.legend=TRUE, legend="bottom")
+
+```
+
 ```{r tombsinuse, fig.dim=c(7,4), dpi=300, fig.cap="Estimates of the number of tombs in use during each of the ceramic phases attested in both cemetaries."}
@@
 ```
-
-Such countervalent trends are observable too in the number of tombs receiving in each period.
-Figure \@ref(fig:tombsinuse) presents the estimated number of tombs in use throughout the EIA at both cemeteries.
-At the Forteta, this number remained relatively stable from the MPG to LG periods before declining (and bear in mind, there were only around 20 tombs belonging to this cemetery throughout the period), while at the KNC, the numbers of tombs in use rose to a preak in the late 8th century before declining.
-The cessation of burials there certainly appears more sudden and marked than at Fortetsa, though, again, it was always a larger cemetery.
-But this is not the whole story.
 
@@
 
-Figure \@ref(fig:tombsinusebybuilddate) shows the number of tombs in use in each period, divided into the centuries in which those tombs were constructed, all overlaid on the total number of tombs in use across both cemeteries.
-A wave-like pattern can be observed with each century marked by the construction of new tombs which then steadily go out of use, as new ones are built.
-This would seem to accord with Cavanagh's suggestion that "[t]he length of time the tombs last seems [...] to reflect the duration of the social group which had access to the sepulchre", and that, if such access was lineally inehrited, frequent fission would result in limited trans-generational use.
-We might expect, then, that it was the 8th and 7th century tombs coming into use that spurred the dramatic rise in burial rates previoulsy noted in this period.
-But this is not the case.
-
 ```{r potsbydateandtombdate,  fig.dim=c(7,4), dpi=300, fig.cap="Counts of vessels from across all tombs in both cemeteries, arranged, on the left, by the century to which they are assigned based on their ceramic period and, on the right, by the centuries in which the tombs they come from were built."}
@@
 
-Figure \@ref(fig:potsbydateandtombdate) shows, firstly, the number of vessels dating to each century across both cemeteries, based on their assigned ceramic period.
-But on the right, those pots are divided into the centuries to which the tombs they were found in originally date.
-So as can be seen, though the late 8th and early 7th centuries represented the apex of burial activity across both cemeteries, it was not the tombs built in those centuries which received the lion's share of the burials and associated ceramics.
-This is all the more notable when, returning to Figure \@ref(fig:tombsinusebybuilddate), we see that the number of 9th and 10th century tombs in use by the late 8th and early 7th centuries was declining.
+```{r toptombs, fig.dim = c(7, 5), dpi=300, fig.align='center',fig.cap="Aoristic sums for the 12 largest tombs (in terms of grave goods), arranged in decreasing order of total recovered artefacts.
+Note, for the sake of better illustrating the relative patterns, rather than absolute values, the scales on each plot are distinct."}
+
+tombs_ranked<-tombs[with(tombs, order(-tombs$total_pottery)), ]
+tombs_top_12<-head(tombs_ranked,12)
+top12<-as.vector(tombs_top_12$tomb)
+pots_top_12<-pots[is.element(pots$tomb, top12),]
+pots_top_12$tomb<-factor(pots_top_12$tomb,levels=top12)
+
+cols<-c(1,158:207)
+aorist_all_tomb_names<-tombs_top_12[,cols]
+aorist_all_tomb_names<-data.frame(aorist_all_tomb_names)
+colnames(aorist_all_tomb_names)<-c("tomb",dates)
+
+aorist_all_tomb_melt<-melt(aorist_all_tomb_names)
+aorist_all_tomb_melt$variable<-as.numeric(as.character(aorist_all_tomb_melt$variable))
+
+top_12<-tombs_top_12$tomb
+aorist_all_tomb_melt$tomb<-factor(aorist_all_tomb_melt$tomb,levels=top_12)
+
+ggplot(aorist_all_tomb_melt,aes(variable, value))+geom_col()+theme(legend.position = "none")+facet_wrap(~tomb,scales="free_y")+
+  geom_col()+theme_bw()+theme(legend.title=element_blank(),
+      text=element_text(family="Garamond", size=12))+
+          labs(x="Date (Years BC)", y="Aoristic Sums")+
+              scale_fill_brewer(palette = "Set3")+ 
+                theme(legend.position="none")+
+                  coord_cartesian(xlim=c(1100,610))+scale_y_continuous()+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1100,610))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+```
+
+```{r tombdimensions, fig.dim = c(7.5, 4), dpi=300, fig.align='center', fig.cap="Scatterplots comparing tomb dimensions (chamber area and *dromos* length) to estimated total burials by tomb."}
+
+tombareas<-ggplot(subset(tombs,tombs$est_burials>1),aes(chamber_area,est_burials,color=cem_condensed))+geom_point()+
+  theme_bw()+theme(legend.title=element_blank(),
+        text=element_text(family="Garamond", size=12))+
+            labs(x=bquote('Chamber Area ('*m^2*')'), y="Estimated Total Burials")+
+               scale_color_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+
+                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+dromoslengths<-ggplot(subset(tombs,tombs$est_burials>1),aes(dromos_length,est_burials,color=cem_condensed))+geom_point()+
+  theme_bw()+theme(legend.title=element_blank(),
+        text=element_text(family="Garamond", size=12))+
+            labs(x="Dromos Length (m)", y="Estimated Total Burials")+
+               scale_color_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+
+                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+ggarrange(tombareas,dromoslengths,ncol=2,common.legend = TRUE,legend="bottom")
+
+```
+
+```{r dimensionsboxplots, fig.dim=c(6,5),dpi=300, fig.align='center',fig.cap="Boxplots comparing various tomb dimensions as they vary by the century of tomb construction."}
+
+tombs$century_built<-factor(tombs$century_built,levels=c("11th","10th","9th","8th","7th"))
+dromwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_width,fill=century_built))+geom_boxplot()+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Century of Tomb Construction", y="Dromos Width (m)")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0.1))
+dromlength<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_length,fill=century_built))+geom_boxplot()+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Century of Tomb Construction", y="Dromos Length (m)")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0.5))
+stomwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,stomion_width,fill=century_built))+geom_boxplot()+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Century of Tomb Construction", y="Stomion Width (m)")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0.1))
+chamarea<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,chamber_area,fill=century_built))+geom_boxplot()+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Century of Tomb Construction", y=bquote('Chamber Area ('*m^2*')'))+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0.5))
+
+ggarrange(dromwidth,dromlength,stomwidth,chamarea,nrow=2,ncol=2)
+
+```
+
+```{r tombscompthroughtime, fig.dim=c(7,4),dpi=300,fig.align='center',fig.cap="Counts of vessels belonging to the 9th, 8th and 7th centuries, found in tombs dating to the 10th or 9th century which yielded upwards of 50 vessels when excavated.
+The seven large tombs with late peaks in burial activity identified earlier are shown in green."}
+
+pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A","B")
+tombs9th10th<-subset(tombs,century_built=="9th" & total_pottery>50|century_built=="10th" & total_pottery>50)
+pots9th10th<-subset(pots,tomb_century=="9th"|tomb_century=="10th")
+
+tombs9th10th_ranked<-tombs9th10th[with(tombs9th10th, order(-tombs9th10th$total_pottery)), ]
+tombs9th10th_ranked<-as.vector(tombs9th10th_ranked$tomb)
+pots9th10th$tomb<-factor(pots9th10th$tomb,levels=tombs9th10th_ranked)
+pots9th10th<-subset(pots9th10th,century=="9th" & tomb!="NA"|century=="8th" & tomb!="NA"|century=="7th" & tomb!="NA")
+pots9th10th$century<-mapvalues(pots9th10th$century, from = c("9th","8th","7th"), to = c("9th Century", "8th Century","7th Century"))
+
+ggplot(pots9th10th,aes(tomb,fill=group))+geom_bar()+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="Garamond", size=12),
+               axis.text.x = element_text(angle = 90,hjust=1,vjust=0.4))+
+                 labs(x="Tomb", y="Vessel Count")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0))+facet_grid(~century,scales="free_x",space="free")
+```
+
+```{r imports, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="Aoristic sums of imported ceramics through time across both cemeteries, colour-coded by the broad regional origin of the vessels."}
+
+imports<-subset(pots,pots$import!="")
+greekimports<-subset(imports,imports$import_region=="Greece")
+NEimports<-subset(imports,imports$import_region=="Near East")
+
+import_aorist<-colSums(imports[27:76])
+import_aorist<-data.frame(import_aorist)
+colnames(import_aorist)<-c("aoristic_sum")
+row.names(import_aorist)<-dates
+setDT(import_aorist, keep.rownames = "decade")
+import_aorist$decade<-as.numeric(import_aorist$decade)
+
+greekimport_aorist<-colSums(greekimports[27:76])
+greekimport_aorist<-data.frame(greekimport_aorist)
+colnames(greekimport_aorist)<-c("aoristic_sum")
+row.names(greekimport_aorist)<-dates
+setDT(greekimport_aorist, keep.rownames = "decade")
+greekimport_aorist$decade<-as.numeric(greekimport_aorist$decade)
+
+NEimport_aorist<-colSums(NEimports[27:76])
+NEimport_aorist<-data.frame(NEimport_aorist)
+colnames(NEimport_aorist)<-c("aoristic_sum")
+row.names(NEimport_aorist)<-dates
+setDT(NEimport_aorist, keep.rownames = "decade")
+NEimport_aorist$decade<-as.numeric(NEimport_aorist$decade)
+
+greekimport_aorist$origin<-"Greece"
+NEimport_aorist$origin<-"Near East"
+allimport_aorist<-rbind(greekimport_aorist,NEimport_aorist)
+
+imitations<-subset(pots,pots$imitation!="")
+imitation_aorist<-colSums(imitations[27:76])
+imitation_aorist<-data.frame(imitation_aorist)
+colnames(imitation_aorist)<-c("aoristic_sum")
+row.names(imitation_aorist)<-dates
+setDT(imitation_aorist, keep.rownames = "decade")
+imitation_aorist$decade<-as.numeric(imitation_aorist$decade)
+
+ggplot()+geom_col(data=import_aorist,aes(decade,aoristic_sum),alpha=0.2)+
+          geom_col(data=allimport_aorist,aes(decade,aoristic_sum, fill=origin),position="dodge")+
+            geom_col(data=imitation_aorist,aes(decade,aoristic_sum))+
+            theme_bw()+theme(text=element_text(family="Garamond", size=12))+
+                labs(x="Date (Years BC)", y="Aoristic Sum")+
+                  scale_fill_brewer(name="Origin",palette = "Set2")+
+                   scale_y_continuous(expand=c(0,0),limits=c(0,20))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))
+
+```
+
+```{r importstombcenturies, fig.dim=c(8,4),dpi=300, fig.align='center',fig.cap="Aoristic sums of imported vessels through time, from tombs built in the 11th through 9th centuries."}
+
+imports<-subset(pots,import!="")
+
+import_aorist<-colSums(imports[27:76])
+import_aorist<-data.frame(import_aorist)
+colnames(import_aorist)<-c("aoristic_sum")
+row.names(import_aorist)<-dates
+setDT(import_aorist, keep.rownames = "decade")
+import_aorist$decade<-as.numeric(import_aorist$decade)
+
+imports_11th<-subset(imports,tomb_century=="11th")
+imports_10th<-subset(imports,tomb_century=="10th")
+imports_9th<-subset(imports,tomb_century=="9th")
+imports_8th<-subset(imports,tomb_century=="8th")
+imports_7th<-subset(imports,tomb_century=="7th")
+
+import_11th_aorist<-colSums(imports_11th[27:76])
+import_11th_aorist<-data.frame(import_11th_aorist)
+colnames(import_11th_aorist)<-c("aoristic_sum")
+row.names(import_11th_aorist)<-dates
+setDT(import_11th_aorist, keep.rownames = "decade")
+import_11th_aorist$decade<-as.numeric(import_11th_aorist$decade)
+
+import_11th_aorist$century<-"11th"
+
+import_10th_aorist<-colSums(imports_10th[27:76])
+import_10th_aorist<-data.frame(import_10th_aorist)
+colnames(import_10th_aorist)<-c("aoristic_sum")
+row.names(import_10th_aorist)<-dates
+setDT(import_10th_aorist, keep.rownames = "decade")
+import_10th_aorist$decade<-as.numeric(import_10th_aorist$decade)
+
+import_10th_aorist$century<-"10th"
+
+import_9th_aorist<-colSums(imports_9th[27:76])
+import_9th_aorist<-data.frame(import_9th_aorist)
+colnames(import_9th_aorist)<-c("aoristic_sum")
+row.names(import_9th_aorist)<-dates
+setDT(import_9th_aorist, keep.rownames = "decade")
+import_9th_aorist$decade<-as.numeric(import_9th_aorist$decade)
+
+import_9th_aorist$century<-"9th"
+
+import_8th_aorist<-colSums(imports_8th[27:76])
+import_8th_aorist<-data.frame(import_8th_aorist)
+colnames(import_8th_aorist)<-c("aoristic_sum")
+row.names(import_8th_aorist)<-dates
+setDT(import_8th_aorist, keep.rownames = "decade")
+import_8th_aorist$decade<-as.numeric(import_8th_aorist$decade)
+
+import_8th_aorist$century<-"8th"
+
+import_7th_aorist<-colSums(imports_7th[27:76])
+import_7th_aorist<-data.frame(import_7th_aorist)
+colnames(import_7th_aorist)<-c("aoristic_sum")
+row.names(import_7th_aorist)<-dates
+setDT(import_7th_aorist, keep.rownames = "decade")
+import_7th_aorist$decade<-as.numeric(import_7th_aorist$decade)
+
+import_7th_aorist$century<-"7th"
+
+all_import_aorist<-rbind(import_11th_aorist,import_10th_aorist,import_9th_aorist,import_8th_aorist,import_7th_aorist)
+all_import_aorist$century<-factor(all_import_aorist$century,levels=c("11th","10th","9th","8th","7th"))
+
+ggplot()+ geom_col(data=subset(all_import_aorist,century!="7th" & century!="8th"),aes(decade,aoristic_sum,fill=century),position="dodge")+
+  scale_x_reverse(limits=c(1000,630))+scale_fill_brewer(palette="Set2")+
+  theme_bw()+theme(legend.position="bottom",text=element_text(family="Garamond", size=12))+
+            labs(x="Date (Years BC)", y="Aoristic Sum")+
+  guides(fill=guide_legend(title="Century of Tomb Construction"))+
+              scale_fill_brewer(palette = "Set2")+ 
+                  scale_y_continuous(expand=c(0,0),limits=c(0,12))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+```
+
+```{r bigbox, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="Boxplot for the aoristic sums of all vessels across both cemeteries through time, colour-coded by ceramic period.
+Boxes represent the interquartile range, horizontal lines the median value, whiskers values within 1.5 times the interquartile range, and points outliers beyond this."}
+
+aorist_tomb_values<-tombs[c(1,158:207)]
+aorist_tomb_values[aorist_tomb_values == 0] <- NA
+
+aorist_tomb_values<-data.frame(aorist_tomb_values)
+colnames(aorist_tomb_values)[2:51]<-dates
+
+aorist_melt<-melt(aorist_tomb_values)
+aorist_melt$variable<-as.numeric(as.character(aorist_melt$variable))
+
+aorist_melt$period<-ifelse(aorist_melt$variable<=1100 & aorist_melt$variable>=980,"SM",
+		ifelse(aorist_melt$variable<=970 & aorist_melt$variable>=930,"EPG",
+		ifelse(aorist_melt$variable<=920 & aorist_melt$variable>=880,"MPG",
+		ifelse(aorist_melt$variable<=870 & aorist_melt$variable>=850,"LPG",
+		ifelse(aorist_melt$variable<=840 & aorist_melt$variable>=820,"PGB",
+	  ifelse(aorist_melt$variable<=810 & aorist_melt$variable>=800,"EG",
+	  ifelse(aorist_melt$variable<=790 & aorist_melt$variable>=750,"MG",
+	  ifelse(aorist_melt$variable<=740 & aorist_melt$variable>=720,"LG",
+	  ifelse(aorist_melt$variable<=710 & aorist_melt$variable>=680,"EO",
+	  ifelse(aorist_melt$variable<=670 & aorist_melt$variable>=640,"LO",
+	  ifelse(aorist_melt$variable<=630 & aorist_melt$variable>=600,"A",""
+		)))))))))))
+
+decade_ranges<-seq(from = 600, to = 1100, by = 10)
+decade_ranges<-rev(decade_ranges)
+decade_ranges<-as.factor(decade_ranges)
+aorist_melt$variable<-as.factor(aorist_melt$variable)
+aorist_melt$variable<-factor(aorist_melt$variable, levels=(decade_ranges))
+aorist_melt$period<-factor(aorist_melt$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
+aorist_melt$variable<-as.numeric(as.character(aorist_melt$variable))
+
+ggplot(aorist_melt,aes(variable,value,fill=period))+geom_boxplot(aes(group = cut_width(variable, 10)),outlier.size=0.8)+theme_bw()+
+      theme(text=element_text(family="Garamond", size=12))+
+            labs(x="Date (Years BC)", y="Aoristic Sums per Tomb")+
+               scale_fill_brewer(name="Period",palette = "Set3")+
+                scale_x_reverse(expand=c(0,0))+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,20))
+
+```
+
+```{r gini, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="Gini coefficient values through time based on the aoristic sums for all vessels across both cemeteries.
+Background colours correspond to ceramic periods."}
+
+aorist_tomb_values<-tombs[c(1,158:207)]
+aorist_tomb_values[aorist_tomb_values == 0] <- NA
+
+aorist_tomb_values<-data.frame(aorist_tomb_values)
+colnames(aorist_tomb_values)[2:51]<-dates
+
+tombs_gini<-matrix(0, ncol = 0, nrow = 1)
+tombs_gini<-data.frame(tombs_gini)
+for(i in colnames(aorist_tomb_values[2:51])) {
+  tombs_gini[[paste0(i)]] <- ineq(aorist_tomb_values[[i]],type="Gini")
+}
+
+tombs_gini<-t(tombs_gini)
+tombs_gini<-data.frame(tombs_gini)
+
+colnames(tombs_gini)<-c("gini_coefficient")
+row.names(tombs_gini)<-dates
+tombs_gini<-data.frame(tombs_gini)
+setDT(tombs_gini, keep.rownames = "decade")
+tombs_gini$decade<-as.numeric(tombs_gini$decade)
+
+rects <- data.frame(xstart = c(1100,970,920,875,840,810,790,745,70,670,630), 
+                    xend = c(970,920,875,840,810,790,745,710,670,630,600), 
+                    col = c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
+rects$col<-factor(rects$col,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
+
+ggplot() + 
+  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf, fill = col), alpha = 0.4) +
+     geom_line(data = tombs_gini, aes(decade,gini_coefficient),size=0.5)+
+      theme_bw()+theme(text=element_text(family="Garamond", size=12))+
+              labs(x="Date (Years BC)", y="Gini Coefficient")+
+                 scale_fill_brewer(name="Period",palette = "Set3")+ coord_cartesian(
+                    xlim = c(1100, 600), ylim = c(0.5,0.8))+ scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
+                       scale_y_continuous(expand=c(0,0))
+```
+
+```{r comps, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="Plots showing the composition of the total pottery assemblage for each century across both cemeteries, and among the 7 notable tombs drawn out earlier in the analysis."}
+pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A","B")
+
+potcompKNC<-ggplot(subset(pots,century!="" & cem_condensed=="NC"),aes(century,fill=type))+geom_bar(position="fill")+
+  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title = element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(title="KNC",x="", y="Proportion of Total Assemblage")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" Amphora  "," Jug  "," Oil vessel  ", " Pithos"," Drinking vessel  "," Lid  "," Other"))+ 
+                    scale_y_continuous(expand=c(0,0.0))
+potcompFOR<-ggplot(subset(pots,century!="" & cem_condensed=="FOR"),aes(century,fill=type))+geom_bar(position="fill")+
+  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(title="Fortetsa",x="Century BC", y="")+
+                  scale_fill_brewer(palette = "Set2", labels=c(" Amphora  "," Jug  "," Oil vessel  ", " Pithos"," Drinking vessel  "," Lid  "," Other"))+ 
+                    scale_y_continuous(expand=c(0,0.0))
+
+potcomptop7<-ggplot(subset(pots,century!="" & century!="11th"& group=="A"),aes(century,fill=type))+geom_bar(position="fill")+
+  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(title="7 Tombs",x="", y="")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" Amphora  "," Jug  "," Oil vessel  ", " Pithos"," Drinking vessel  "," Lid  "," Other"))+ 
+                    scale_y_continuous(expand=c(0,0.0))
+
+ggarrange(potcompKNC,potcompFOR,potcomptop7,ncol=3,common.legend =TRUE,legend="bottom")
+
+spearmanobjectspots<-cor(tombs$total_object,tombs$total_pottery,method="spearman")
+```
+
+```{r hists, fig.dim=c(6,4),dpi=300, fig.align='center',fig.cap="Histograms relating, first, to counts of non-ceramic artefacts, metal objects, and imported items across all tombs in both cemeteries and, secondly, to those counts divided by the estimated number of burials in each tomb."}
+tombs$group<-factor(tombs$group,levels=c("A","B"))
+total_object<-ggplot(subset(tombs,est_burials>1 & total_object>0),aes(total_object, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Total Non-Ceramic Finds", y="Number of Tombs")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+total_metal<-ggplot(subset(tombs,est_burials>1 & weapon>0),aes(weapon, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Total Metal Artefacts", y="Number of Tombs")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+total_import<-ggplot(subset(tombs,est_burials>1 & total_import>0),aes(total_import, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Total Imports", y="Number of Tombs")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+
+objectsbyburial<-ggplot(subset(tombs,est_burials>1 & total_object>0),aes(total_object/est_burials, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Total Non-Ceramics/Burials", y="Number of Tombs")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+metalbyburial<-ggplot(subset(tombs,est_burials>1 & weapon>0),aes(weapon/est_burials, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Total Metal Artefacts/Burials", y="Number of Tombs")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+
+importsbyburial<-ggplot(subset(tombs,est_burials>1 & total_import>0),aes(total_import/est_burials, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Total Imports/Burials", y="Number of Tombs")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+
+ggarrange(total_import,total_metal,total_object,importsbyburial,metalbyburial,objectsbyburial,ncol=3,nrow=2,common.legend = TRUE,legend="bottom")
+```
+
+##### pagebreak
+
+# Tables
 
@@
 
-The significance of these findings can be seen in Table 1.
-Using pithoi as a proxy for burials, we can estimate the changing rates of burial through the centuries based on how many tombs yielded how many pithoi of each period.
-And while for the most part burial rates remained around three per tomb per century, in the 8th and 7th centuries, tombs founded in the 9th century saw a rapid rise in the rate of burial, peaking at over 11 per tomb per century.
-Though these estimates are no doubt very rough, the marked nature of this divergence suggests it is not simply an artefact of imperfect data.
-
-Also apparent from Table 1, though, are the markedly low rates of burial which predominated at both cemeteries.
-Two to four burials per tomb per century amounts to a very selective burial rite, even considering the restricted portion of the population probably represented by these cemeteries [@Cavanagh1996, 664].
-It seems unlikely, then, that the rising rates of burial noted above are, for instance, the direct result of population growth; the increase seems too rapid, and is concentrated only in certain (older) tombs, not to mention that, based on KULP’s survey data, the major population increase at EIA Knossos should be dated a century or more before the late 8th century peak in burial, in the Protogeometric period (Whitelaw, per comm.).
-
-##Isolating the phenomenon
-
-We are yet to draw out *which tombs exactly* are responsible for the observible patterning so far discussed.
-We might assume that the tombs which became the focus of intense burial activity in the late 8th and early 7th centuries would be those which yielded the most burials and grave goods when excavated and, happily, this supposition proves true.
-
-```{r toptombs, fig.dim = c(7, 5), dpi=300, fig.align='center',fig.cap="Aoristic sums for the 12 largest tombs (in terms of grave goods), arranged in decreasing order of total recovered artefacts.
-Note, for the sake of better illustrating the relative patterns, rather than absolute values, the scales on each plot are distinct."}
-
-tombs_ranked<-tombs[with(tombs, order(-tombs$total_pottery)), ]
-tombs_top_12<-head(tombs_ranked,12)
-top12<-as.vector(tombs_top_12$tomb)
-pots_top_12<-pots[is.element(pots$tomb, top12),]
-pots_top_12$tomb<-factor(pots_top_12$tomb,levels=top12)
-
-cols<-c(1,158:207)
-aorist_all_tomb_names<-tombs_top_12[,cols]
-aorist_all_tomb_names<-data.frame(aorist_all_tomb_names)
-colnames(aorist_all_tomb_names)<-c("tomb",dates)
-
-aorist_all_tomb_melt<-melt(aorist_all_tomb_names)
-aorist_all_tomb_melt$variable<-as.numeric(as.character(aorist_all_tomb_melt$variable))
-
-top_12<-tombs_top_12$tomb
-aorist_all_tomb_melt$tomb<-factor(aorist_all_tomb_melt$tomb,levels=top_12)
-
-ggplot(aorist_all_tomb_melt,aes(variable, value))+geom_col()+theme(legend.position = "none")+facet_wrap(~tomb,scales="free_y")+
-  geom_col()+theme_bw()+theme(legend.title=element_blank(),
-      text=element_text(family="Garamond", size=12))+
-          labs(x="Date (Years BC)", y="Aoristic Sums")+
-              scale_fill_brewer(palette = "Set3")+ 
-                theme(legend.position="none")+
-                  coord_cartesian(xlim=c(1100,610))+scale_y_continuous()+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1100,610))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-tombs$group<-ifelse(tombs$tomb=="P"|tombs$tomb=="292"|tombs$tomb=="75"|tombs$tomb=="107"|tombs$tomb=="285"|tombs$tomb=="II"|tombs$tomb=="218","A","B")
-total_pottery_by_groups<-aggregate(tombs$total_pottery, by=list(tombs$group), FUN=sum)
-calc1<-(total_pottery_by_groups[1,2]/sum(total_pottery_by_groups$x))*100
-
-```
-
-The 12 tombs with the most associated vessels are all chamber tombs, with 10 dating to the 9th century, and two to the 10th.
-And, in Figure \@ref(fig:toptombs), we see that seven of the 12 did exhibit late, intense episodes of deposition.
-Tombs P, 292, 75, 107, 285, II, 218, and 104 between them accrount for `r round(calc1,digits=2)`% of the entire pottery assemblage recovered from the `r nrow(tombs)` tombs considered here.
-All but Tomb 285 date to the 9th century, and together they underpinned the late spate of burials here considered.
-Interestingly, Tombs X, G, Q, 104 and 283, however, evince a different pattern, with their respective peaks coming in the PGB and EG periods, around a century earlier.
-We have seen already in Figure \@ref(fig:aoristbycem) that, particularly in the KNC, the end of the 9th century was another period of increased burial activity.
-What, then, set these two groups of tombs apart?
-The recipients of the greatest numbers of burials and grave goods in both cemeteries, they nonetheless seem to have achieved that distinction via different histories.
-By looking more closely at the periods in which they were built, and in which they were most popular, I hope to shed light on the broader social contexts in which these tombs were flourished.
-
-##Not all tombs are created equal
-
-The indices discussed so far – the distribution of of artefacts between the tombs, and the construction and use of the tombs through time – only inform us of the temporal or aggregate features of the cemeteries; they tell us little of the tombs as they were built or first used.
-Synchronic comparisons are harder to make with the available data, but the dimensions of the tombs offer some scope for doing so.
-Perhaps the tombs which went on to receive the most burials were notably large or grand when first built, reflecting the status of the associated burying group.
-
-```{r tombdimensions, fig.dim = c(7.5, 4), dpi=300, fig.align='center', fig.cap="Scatterplots comparing tomb dimensions (chamber area and *dromos* length) to estimated total burials by tomb."}
-
-tombareas<-ggplot(subset(tombs,tombs$est_burials>1),aes(chamber_area,est_burials,color=cem_condensed))+geom_point()+
-  theme_bw()+theme(legend.title=element_blank(),
-        text=element_text(family="Garamond", size=12))+
-            labs(x=bquote('Chamber Area ('*m^2*')'), y="Estimated Total Burials")+
-               scale_color_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+
-                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
-                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-dromoslengths<-ggplot(subset(tombs,tombs$est_burials>1),aes(dromos_length,est_burials,color=cem_condensed))+geom_point()+
-  theme_bw()+theme(legend.title=element_blank(),
-        text=element_text(family="Garamond", size=12))+
-            labs(x="Dromos Length (m)", y="Estimated Total Burials")+
-               scale_color_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+
-                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
-                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-ggarrange(tombareas,dromoslengths,ncol=2,common.legend = TRUE,legend="bottom")
-
-```
-
-Looking at Figure \@ref(fig:tombdimensions), however, it seems that no clear relationship existed between the physical size of a tomb and its buried population at EIA Knossos.
-It is true that over half of all tombs do not exceed 2.6 sq m in area, while nine of the ten tombs with the most burials do surpass this value.
-But many of the very largest tombs are decidedly middling when it comes to burial numbers.
-Things are even less clear for the *dromos*, the most visible part of the tomb, and a plausible arena for demonstrations of scale and grandeur during the burial itself.
-The very largest two tombs by burial numbers also had exceptionally long *dromoi*, but the next eight tombs with the largest *dromoi* do not rank highly when it comes to interments.
-There are difficulties here in that the *dromoi* especially were common victims of encroachment by later tombs, and our estimates of their original length are tentative in many cases.
-
-```{r dimensionsboxplots, fig.dim=c(6,5),dpi=300, fig.align='center',fig.cap="Boxplots comparing various tomb dimensions as they vary by the century of tomb construction."}
-
-tombs$century_built<-factor(tombs$century_built,levels=c("11th","10th","9th","8th","7th"))
-dromwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_width,fill=century_built))+geom_boxplot()+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Century of Tomb Construction", y="Dromos Width (m)")+
-                  scale_fill_brewer(palette = "Set2")+ 
-                    scale_y_continuous(expand=c(0,0.1))
-dromlength<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_length,fill=century_built))+geom_boxplot()+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Century of Tomb Construction", y="Dromos Length (m)")+
-                  scale_fill_brewer(palette = "Set2")+ 
-                    scale_y_continuous(expand=c(0,0.5))
-stomwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,stomion_width,fill=century_built))+geom_boxplot()+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Century of Tomb Construction", y="Stomion Width (m)")+
-                  scale_fill_brewer(palette = "Set2")+ 
-                    scale_y_continuous(expand=c(0,0.1))
-chamarea<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,chamber_area,fill=century_built))+geom_boxplot()+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Century of Tomb Construction", y=bquote('Chamber Area ('*m^2*')'))+
-                  scale_fill_brewer(palette = "Set2")+ 
-                    scale_y_continuous(expand=c(0,0.5))
-
-ggarrange(dromwidth,dromlength,stomwidth,chamarea,nrow=2,ncol=2)
-
-```
-
-Looking at Figure \@ref(fig:dimensionsboxplots), though, we can see that, through the centuries, the dimensions of the tombs varied, and sometimes consistently, as with the trend toward narrower *dromoi*.
-The 9th-century tombs, including all but one of those seven noted above, do not particularly stand out.
-It does seem that chambers were, on average, larger in this period as well as varying more in size than in any other period.
-Then again, even the largest chambers are not exactly cavernous, and the extra labour expended on the construction of the larger chambers would have been limited.
-The *dromos* is a plausible area for conspicuous demonstrations of scale but, with their frequently poor preservation, we cannot be confident in assessing this.
-It appears at any rate that they were gradually declining in size, albeit with some notable outliers.
-
-So perhaps, at their time of building, such aspects of tomb construction carried social meaning but, even if they did, these Figures suggest this proved irrelevant to their later reuse.
-Several tombs had niches cut into their *dromoi*, or the *stomion* extended, to fit in later burial urns.
-Sometimes, urns from within the tomb were placed in such niches to make space for new interments in the chamber itself.
-There is an impression of individuals vying to be included in the chamber itself, of competition was for inclusion into the sepulchral group, supporting the idea that the tombs’ original builders little anticipated their eventual popularity and so did not scale their tombs accordingly.
-
-```{r tombscompthroughtime, fig.dim=c(7,4),dpi=300,fig.align='center',fig.cap="Counts of vessels belonging to the 9th, 8th and 7th centuries, found in tombs dating to the 10th or 9th century which yielded upwards of 50 vessels when excavated.
-The seven large tombs with late peaks in burial activity identified earlier are shown in green."}
-
-pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A","B")
-tombs9th10th<-subset(tombs,century_built=="9th" & total_pottery>50|century_built=="10th" & total_pottery>50)
-pots9th10th<-subset(pots,tomb_century=="9th"|tomb_century=="10th")
-
-tombs9th10th_ranked<-tombs9th10th[with(tombs9th10th, order(-tombs9th10th$total_pottery)), ]
-tombs9th10th_ranked<-as.vector(tombs9th10th_ranked$tomb)
-pots9th10th$tomb<-factor(pots9th10th$tomb,levels=tombs9th10th_ranked)
-pots9th10th<-subset(pots9th10th,century=="9th" & tomb!="NA"|century=="8th" & tomb!="NA"|century=="7th" & tomb!="NA")
-pots9th10th$century<-mapvalues(pots9th10th$century, from = c("9th","8th","7th"), to = c("9th Century", "8th Century","7th Century"))
-
-ggplot(pots9th10th,aes(tomb,fill=group))+geom_bar()+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="Garamond", size=12),
-               axis.text.x = element_text(angle = 90,hjust=1,vjust=0.4))+
-                 labs(x="Tomb", y="Vessel Count")+
-                  scale_fill_brewer(palette = "Set2")+ 
-                    scale_y_continuous(expand=c(0,0))+facet_grid(~century,scales="free_x",space="free")
-```
-
-##Imports, interments, and indices of wealth
-
-Though it may not be immediately apparent from the dimensions of the tombs, however, the late 9th and early 8th centuries are consistently interpreted as periods of artistic development, increased overseas trade, and renewed vigour in the construction of tombs and the visibility of the burial rite.
-The PGB style reflects a uniquely Cretan blend of Attic and Near Eastern forms, while the appearance of Bronze Age 'antiques' in the tombs from this period onwards is felt to reflect renewed, even competitive interest in the Minoan past, as a tool for mortuary display or even lineage affirmation.
-
-```{r imports, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="Aoristic sums of imported ceramics through time across both cemeteries, colour-coded by the broad regional origin of the vessels."}
-
-imports<-subset(pots,pots$import!="")
-greekimports<-subset(imports,imports$import_region=="Greece")
-NEimports<-subset(imports,imports$import_region=="Near East")
-
-import_aorist<-colSums(imports[27:76])
-import_aorist<-data.frame(import_aorist)
-colnames(import_aorist)<-c("aoristic_sum")
-row.names(import_aorist)<-dates
-setDT(import_aorist, keep.rownames = "decade")
-import_aorist$decade<-as.numeric(import_aorist$decade)
-
-greekimport_aorist<-colSums(greekimports[27:76])
-greekimport_aorist<-data.frame(greekimport_aorist)
-colnames(greekimport_aorist)<-c("aoristic_sum")
-row.names(greekimport_aorist)<-dates
-setDT(greekimport_aorist, keep.rownames = "decade")
-greekimport_aorist$decade<-as.numeric(greekimport_aorist$decade)
-
-NEimport_aorist<-colSums(NEimports[27:76])
-NEimport_aorist<-data.frame(NEimport_aorist)
-colnames(NEimport_aorist)<-c("aoristic_sum")
-row.names(NEimport_aorist)<-dates
-setDT(NEimport_aorist, keep.rownames = "decade")
-NEimport_aorist$decade<-as.numeric(NEimport_aorist$decade)
-
-greekimport_aorist$origin<-"Greece"
-NEimport_aorist$origin<-"Near East"
-allimport_aorist<-rbind(greekimport_aorist,NEimport_aorist)
-
-imitations<-subset(pots,pots$imitation!="")
-imitation_aorist<-colSums(imitations[27:76])
-imitation_aorist<-data.frame(imitation_aorist)
-colnames(imitation_aorist)<-c("aoristic_sum")
-row.names(imitation_aorist)<-dates
-setDT(imitation_aorist, keep.rownames = "decade")
-imitation_aorist$decade<-as.numeric(imitation_aorist$decade)
-
-ggplot()+geom_col(data=import_aorist,aes(decade,aoristic_sum),alpha=0.2)+
-          geom_col(data=allimport_aorist,aes(decade,aoristic_sum, fill=origin),position="dodge")+
-            geom_col(data=imitation_aorist,aes(decade,aoristic_sum))+
-            theme_bw()+theme(text=element_text(family="Garamond", size=12))+
-                labs(x="Date (Years BC)", y="Aoristic Sum")+
-                  scale_fill_brewer(name="Origin",palette = "Set2")+
-                   scale_y_continuous(expand=c(0,0),limits=c(0,20))+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))
-
-```
-
-The rise in imported items is clearly seen in the tomb assemblages, with a succession of increased Greek and Near Eastern imports respectively in the late 9th and early 8th centuries (see Figure \@ref(fig:imports)).
-Yet beyond conforming to our prior inferences about the LPG-EG periods, the deposition of these imported wares may also cast light on the original pattern drawn out in the present analysis: the late surge in burial activity among some select 9th and 10th century tombs.
-
-```{r importstombcenturies, fig.dim=c(8,4),dpi=300, fig.align='center',fig.cap="Aoristic sums of imported vessels through time, from tombs built in the 11th through 9th centuries."}
-
-imports<-subset(pots,import!="")
-
-import_aorist<-colSums(imports[27:76])
-import_aorist<-data.frame(import_aorist)
-colnames(import_aorist)<-c("aoristic_sum")
-row.names(import_aorist)<-dates
-setDT(import_aorist, keep.rownames = "decade")
-import_aorist$decade<-as.numeric(import_aorist$decade)
-
-imports_11th<-subset(imports,tomb_century=="11th")
-imports_10th<-subset(imports,tomb_century=="10th")
-imports_9th<-subset(imports,tomb_century=="9th")
-imports_8th<-subset(imports,tomb_century=="8th")
-imports_7th<-subset(imports,tomb_century=="7th")
-
-import_11th_aorist<-colSums(imports_11th[27:76])
-import_11th_aorist<-data.frame(import_11th_aorist)
-colnames(import_11th_aorist)<-c("aoristic_sum")
-row.names(import_11th_aorist)<-dates
-setDT(import_11th_aorist, keep.rownames = "decade")
-import_11th_aorist$decade<-as.numeric(import_11th_aorist$decade)
-
-import_11th_aorist$century<-"11th"
-
-import_10th_aorist<-colSums(imports_10th[27:76])
-import_10th_aorist<-data.frame(import_10th_aorist)
-colnames(import_10th_aorist)<-c("aoristic_sum")
-row.names(import_10th_aorist)<-dates
-setDT(import_10th_aorist, keep.rownames = "decade")
-import_10th_aorist$decade<-as.numeric(import_10th_aorist$decade)
-
-import_10th_aorist$century<-"10th"
-
-import_9th_aorist<-colSums(imports_9th[27:76])
-import_9th_aorist<-data.frame(import_9th_aorist)
-colnames(import_9th_aorist)<-c("aoristic_sum")
-row.names(import_9th_aorist)<-dates
-setDT(import_9th_aorist, keep.rownames = "decade")
-import_9th_aorist$decade<-as.numeric(import_9th_aorist$decade)
-
-import_9th_aorist$century<-"9th"
-
-import_8th_aorist<-colSums(imports_8th[27:76])
-import_8th_aorist<-data.frame(import_8th_aorist)
-colnames(import_8th_aorist)<-c("aoristic_sum")
-row.names(import_8th_aorist)<-dates
-setDT(import_8th_aorist, keep.rownames = "decade")
-import_8th_aorist$decade<-as.numeric(import_8th_aorist$decade)
-
-import_8th_aorist$century<-"8th"
-
-import_7th_aorist<-colSums(imports_7th[27:76])
-import_7th_aorist<-data.frame(import_7th_aorist)
-colnames(import_7th_aorist)<-c("aoristic_sum")
-row.names(import_7th_aorist)<-dates
-setDT(import_7th_aorist, keep.rownames = "decade")
-import_7th_aorist$decade<-as.numeric(import_7th_aorist$decade)
-
-import_7th_aorist$century<-"7th"
-
-all_import_aorist<-rbind(import_11th_aorist,import_10th_aorist,import_9th_aorist,import_8th_aorist,import_7th_aorist)
-all_import_aorist$century<-factor(all_import_aorist$century,levels=c("11th","10th","9th","8th","7th"))
-
-ggplot()+ geom_col(data=subset(all_import_aorist,century!="7th" & century!="8th"),aes(decade,aoristic_sum,fill=century),position="dodge")+
-  scale_x_reverse(limits=c(1000,630))+scale_fill_brewer(palette="Set2")+
-  theme_bw()+theme(legend.position="bottom",text=element_text(family="Garamond", size=12))+
-            labs(x="Date (Years BC)", y="Aoristic Sum")+
-  guides(fill=guide_legend(title="Century of Tomb Construction"))+
-              scale_fill_brewer(palette = "Set2")+ 
-                  scale_y_continuous(expand=c(0,0),limits=c(0,12))+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-```
-
-In Figure \@ref(fig:importstombcenturies), we see that the tombs already built by the end of the PGB period, particularly those built in the 9th and, to a lesser extent, 10th centuries, were major beneficiaries of the influx of imported items during this time, but that, in the following centuries, in line with the rest of the cemetery, the number of imported items deposited within them declined.
-Imported ceramics are used here as something of a proxy for items of status value, a shorthand that would ideally be bolstered by other artefacts (such as precious vessels, weapons, jewellery etc.) which, owing to their less developed chronologies, must unfortunately be left out of such diachronic analyses for now.
-But, nonetheless, this pattern hints towards a possibility that shall now be explored, that the surge in burial activity seen across both cemeteries in the late 8th and early 7th centuries would appear to be of a different kind to that of a century earlier.
-
-```{r bigbox, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="Boxplot for the aoristic sums of all vessels across both cemeteries through time, colour-coded by ceramic period.
-Boxes represent the interquartile range, horizontal lines the median value, whiskers values within 1.5 times the interquartile range, and points outliers beyond this."}
-
-aorist_tomb_values<-tombs[c(1,158:207)]
-aorist_tomb_values[aorist_tomb_values == 0] <- NA
-
-aorist_tomb_values<-data.frame(aorist_tomb_values)
-colnames(aorist_tomb_values)[2:51]<-dates
-
-aorist_melt<-melt(aorist_tomb_values)
-aorist_melt$variable<-as.numeric(as.character(aorist_melt$variable))
-
-aorist_melt$period<-ifelse(aorist_melt$variable<=1100 & aorist_melt$variable>=980,"SM",
-		ifelse(aorist_melt$variable<=970 & aorist_melt$variable>=930,"EPG",
-		ifelse(aorist_melt$variable<=920 & aorist_melt$variable>=880,"MPG",
-		ifelse(aorist_melt$variable<=870 & aorist_melt$variable>=850,"LPG",
-		ifelse(aorist_melt$variable<=840 & aorist_melt$variable>=820,"PGB",
-	  ifelse(aorist_melt$variable<=810 & aorist_melt$variable>=800,"EG",
-	  ifelse(aorist_melt$variable<=790 & aorist_melt$variable>=750,"MG",
-	  ifelse(aorist_melt$variable<=740 & aorist_melt$variable>=720,"LG",
-	  ifelse(aorist_melt$variable<=710 & aorist_melt$variable>=680,"EO",
-	  ifelse(aorist_melt$variable<=670 & aorist_melt$variable>=640,"LO",
-	  ifelse(aorist_melt$variable<=630 & aorist_melt$variable>=600,"A",""
-		)))))))))))
-
-decade_ranges<-seq(from = 600, to = 1100, by = 10)
-decade_ranges<-rev(decade_ranges)
-decade_ranges<-as.factor(decade_ranges)
-aorist_melt$variable<-as.factor(aorist_melt$variable)
-aorist_melt$variable<-factor(aorist_melt$variable, levels=(decade_ranges))
-aorist_melt$period<-factor(aorist_melt$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
-aorist_melt$variable<-as.numeric(as.character(aorist_melt$variable))
-
-ggplot(aorist_melt,aes(variable,value,fill=period))+geom_boxplot(aes(group = cut_width(variable, 10)),outlier.size=0.8)+theme_bw()+
-      theme(text=element_text(family="Garamond", size=12))+
-            labs(x="Date (Years BC)", y="Aoristic Sums per Tomb")+
-               scale_fill_brewer(name="Period",palette = "Set3")+
-                scale_x_reverse(expand=c(0,0))+
-                  scale_y_continuous(expand=c(0,0),limits=c(0,20))
-
-```
-
-Figure \@ref(fig:bigbox) is a boxplot for the aoristic sums of each decade across both cemeteries, colour-coded by ceramic phase.
-Note that, for visual clarity, the outliers tomb G in the EG period, and tomb P in the late EO-LO period are excluded, with values ranging from 29-34, and 36-44 respectively.
-The important observation is that the PGB and EG exhibit the largest interquartile ranges (IQRs) of any period, and each possess only a single outlier (beyond 1.5 times the IQR).
-The Orientalising period, on the other hand, exhibits much reduced IQRs, and a proliferation of outliers far and above the range of the other tombs.
-In other words, while both the PGB-EG and EO-LO periods witnessed an expanding volume and range of burial activity, in the earlier period this was distributed somewhat more evenly across the total burying population.
-One speaks of increasing plurality, the other of divergence.
-
-```{r gini, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="Gini coefficient values through time based on the aoristic sums for all vessels across both cemeteries.
-Background colours correspond to ceramic periods."}
-
-aorist_tomb_values<-tombs[c(1,158:207)]
-aorist_tomb_values[aorist_tomb_values == 0] <- NA
-
-aorist_tomb_values<-data.frame(aorist_tomb_values)
-colnames(aorist_tomb_values)[2:51]<-dates
-
-tombs_gini<-matrix(0, ncol = 0, nrow = 1)
-tombs_gini<-data.frame(tombs_gini)
-for(i in colnames(aorist_tomb_values[2:51])) {
-  tombs_gini[[paste0(i)]] <- ineq(aorist_tomb_values[[i]],type="Gini")
-}
-
-tombs_gini<-t(tombs_gini)
-tombs_gini<-data.frame(tombs_gini)
-
-colnames(tombs_gini)<-c("gini_coefficient")
-row.names(tombs_gini)<-dates
-tombs_gini<-data.frame(tombs_gini)
-setDT(tombs_gini, keep.rownames = "decade")
-tombs_gini$decade<-as.numeric(tombs_gini$decade)
-
-rects <- data.frame(xstart = c(1100,970,920,875,840,810,790,745,70,670,630), 
-                    xend = c(970,920,875,840,810,790,745,710,670,630,600), 
-                    col = c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
-rects$col<-factor(rects$col,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
-
-ggplot() + 
-  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf, fill = col), alpha = 0.4) +
-     geom_line(data = tombs_gini, aes(decade,gini_coefficient),size=0.5)+
-      theme_bw()+theme(text=element_text(family="Garamond", size=12))+
-              labs(x="Date (Years BC)", y="Gini Coefficient")+
-                 scale_fill_brewer(name="Period",palette = "Set3")+ coord_cartesian(
-                    xlim = c(1100, 600), ylim = c(0.5,0.8))+ scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
-                       scale_y_continuous(expand=c(0,0))
-```
-
-This possibility is supported further by considering the gini coefficient of the distribution of pottery between the tombs.
-Using the aoristic sums as a probablistically-weighted estimate of the volume of material deposited in each tomb in each decade, use of the gini coefficient can give an (albeit rough) estimate of how unequally distributed that material was.
-And what we see is that, though values were always relatively high (i.e.
-unequal) within the cemeteries, the major spike in the gini coefficient seen in the Orientalising period contrasts markedly with the modest rise and plateau of the PGB-EG.
-The gini coefficient is most commonly used in the assessment of income inequality, but I want to stress here that it is not "wealth" in any meaningful sense that we are talking about here.
-Indeed, if anything, the late flourit of burial activity here discussed may be a case of quantity, and not quality.
-
-```{r comps, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="Plots showing the composition of the total pottery assemblage for each century across both cemeteries, and among the 7 notable tombs drawn out earlier in the analysis."}
-pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A","B")
-
-potcompKNC<-ggplot(subset(pots,century!="" & cem_condensed=="NC"),aes(century,fill=type))+geom_bar(position="fill")+
-  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title = element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(title="KNC",x="", y="Proportion of Total Assemblage")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" Amphora  "," Jug  "," Oil vessel  ", " Pithos"," Drinking vessel  "," Lid  "," Other"))+ 
-                    scale_y_continuous(expand=c(0,0.0))
-potcompFOR<-ggplot(subset(pots,century!="" & cem_condensed=="FOR"),aes(century,fill=type))+geom_bar(position="fill")+
-  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(title="Fortetsa",x="Century BC", y="")+
-                  scale_fill_brewer(palette = "Set2", labels=c(" Amphora  "," Jug  "," Oil vessel  ", " Pithos"," Drinking vessel  "," Lid  "," Other"))+ 
-                    scale_y_continuous(expand=c(0,0.0))
-
-potcomptop7<-ggplot(subset(pots,century!="" & century!="11th"& group=="A"),aes(century,fill=type))+geom_bar(position="fill")+
-  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(title="7 Tombs",x="", y="")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" Amphora  "," Jug  "," Oil vessel  ", " Pithos"," Drinking vessel  "," Lid  "," Other"))+ 
-                    scale_y_continuous(expand=c(0,0.0))
-
-ggarrange(potcompKNC,potcompFOR,potcomptop7,ncol=3,common.legend =TRUE,legend="bottom")
-
-spearmanobjectspots<-cor(tombs$total_object,tombs$total_pottery,method="spearman")
-```
-
-Figure \@ref(fig:comps) shows the composition of the total pottery assemblage from each cemetery, and from the seven tombs drawn out earlier, through the centuries.
-In all cases, there appears to be the steady crystallisation of a basic 'burial kit', comprising a pithos and lid, an oil vessel, a jug, and a drinking vessel.
-The proportion of grave goods comprising 'other' vessels - pyxides, askoi, trays, stands, dinoi, kernoi, kraters, lekanai and more - declines to become only a minor part of the standardised repertoire of the 7th century.
-The seven tombs that underpinned the dramatic increase in burial activity in the later 8th century did not break with this mould, and their assemblages seem as conventional as any other during the last century of use in the cemeteries.
-
-It has no doubt been apparent that the main focus of this article has been on pottery and not other artefacts of metal, stone, glass, and organic materials.
-This is largely because the chronologies for these artefacts are insufficiently developed to allow for the kind of diachronic analysis made thus far.
-But I do want to briefly comment on one aspect of the distribution of non-ceramic artefacts that may make more sense in the light of the foregoing discussion.
-There is, perhaps unsurprisingly, a relatively strong (Spearman's ρ=`r spearmanobjectspots`) correlation between the number of ceramic and non-ceramic finds across the tombs, and, certainly, many of the most unique and precious objects were recovered from among the larger tombs.
-But, the monopoly those tombs would appear to have over such objects, perhaps tempting one initially to speak of 'elite' or 'high-status' tombs, evaporates when the number of those objects is divided by the number of burials represented.
-
-```{r hists, fig.dim=c(6,4),dpi=300, fig.align='center',fig.cap="Histograms relating, first, to counts of non-ceramic artefacts, metal objects, and imported items across all tombs in both cemeteries and, secondly, to those counts divided by the estimated number of burials in each tomb."}
-tombs$group<-factor(tombs$group,levels=c("A","B"))
-total_object<-ggplot(subset(tombs,est_burials>1 & total_object>0),aes(total_object, fill=group))+geom_histogram()+
-  theme_bw()+theme(legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Non-Ceramic Finds", y="Number of Tombs")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
-                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-total_metal<-ggplot(subset(tombs,est_burials>1 & weapon>0),aes(weapon, fill=group))+geom_histogram()+
-  theme_bw()+theme(legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Metal Artefacts", y="Number of Tombs")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
-                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-total_import<-ggplot(subset(tombs,est_burials>1 & total_import>0),aes(total_import, fill=group))+geom_histogram()+
-  theme_bw()+theme(legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Imports", y="Number of Tombs")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
-                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-
-objectsbyburial<-ggplot(subset(tombs,est_burials>1 & total_object>0),aes(total_object/est_burials, fill=group))+geom_histogram()+
-  theme_bw()+theme(legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Non-Ceramics/Burials", y="Number of Tombs")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
-                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-metalbyburial<-ggplot(subset(tombs,est_burials>1 & weapon>0),aes(weapon/est_burials, fill=group))+geom_histogram()+
-  theme_bw()+theme(legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Metal Artefacts/Burials", y="Number of Tombs")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
-                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-
-importsbyburial<-ggplot(subset(tombs,est_burials>1 & total_import>0),aes(total_import/est_burials, fill=group))+geom_histogram()+
-  theme_bw()+theme(legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Imports/Burials", y="Number of Tombs")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
-                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-
-ggarrange(total_import,total_metal,total_object,importsbyburial,metalbyburial,objectsbyburial,ncol=3,nrow=2,common.legend = TRUE,legend="bottom")
-```
-
-Figure \@ref(fig:hists) thus supports the view which has emerged from the foregoing analysis, that the major increase in burial activity at both Fortetsa and the KNC was not a phenomenon linked in any straightforward way to simple principles of 'elite' or 'wealthy' burial.
-To summarise, towards the end of 8th century BC, a select group of tombs, all of which had been constructed back in the 9th or 10th, became the foci of a major increase in burials, something which did not occur among the tombs built around that time.
-We have seen how this surge in acivity occurred against the backdrop of a declining number of older tombs in use, and I have suggested that the rates of change are too rapid, and unevenly distributed, to be explicable through population growth alone.
-We have seen how little about the physical dimensions or early use of these tombs presaged their later popularity, and that, while a similar, if smaller, increase in burial activity occurred around a century earlier, in the PGB-EG periods, this contrasted in the comparative diversity of its associated assemblages, and their distribution across the tombs.
-The later burials were marked by great uniformity in the associated grave goods, and this apparently limited and consistent repertoire may explain why many of the most used tombs prove exceptional only in the absolute quantity of artefacts, and not the accoutrements of each individual burial.
-
-# Discussion
-
-The preceding analysis has revealed a number of potentially significant patterns in the use of the Fortetsa and Knossos North Cemeteries, in particular the exceptional trajectories of a select group of 10th and 9th century tombs.
-It only remains to consider how we whould characterise the burying groups responsible for these observed trends.
-The traditional view is that such groups were immediate families (Brock 1957, 41; Boardman 1967, 63; Coldstream & Huxley 1999; 291).
-The relatively short lifespan of many tombs (often only two or three generations) has been taken as evidence for direct, lineal inheritance, with the tomb going out of use as families moved, individuals founded new tombs, or the line came to an end.
-The idea of lineal groups is supported by the, admittedly circumstantial, evidence of men, women and children being found within single tombs, and one (contentious) instance of an apparently hereditary mandibular deformity in three non-contemporary skulls in the same tomb (Musgrave 1996).
-
-The model of lineal inheritance has been recently challenged by Kotsonas (2011, 131), however, who considers it ‘probably the greatest factoid in the archaeology of Crete of the given period’.
-He argues that the rise in interments during the 8th century, noted also in the present study, is too great to be explained by population growth, as would be necessary for the possibility of direct lineal inheritance to be preserved.
-Certainly, it seems illogical that population increases would lead to the consolidation of burials within only extant tombs, rather than increasing tomb construction.
-The trend instead must reflect choices made by the burying groups themselves, which Kotsonas feels may have comprised various elements, including perhaps age- and sex-grades.
-But it seems probable that such corporate groups, if they existed intersected with, and existed alongside, real or fictive kinship networks.
-
-Yet it is hard to be conclusive, as our current understanding of the social organisation of the early Cretan *poleis* is patchy at best.
-Perlman and Gagarin's [-@Gagarin2016] recent study of the Cretan law codes reminds us, importantly, if lamentably, that there is an often underappreciated degree of diversity apparent from the Archaic inscriptions in the offices, institutions, and social groups of the early city-states.
-In the traditional view, political authority was concentrated in the hands of the *kosmoi*, aristocratic magistrates selected from the leading, hereditary *startoi*, or ‘clans’, networks of extended lineages controlling the ancestral estates on which worked the serfs and/or slaves who made up the lower rungs of the social hierarchy (Willetts 1982, 240).
-But despite elements of this account probably being accurate, others remain equivocal, such as the nature of the *pyla*, or 'tribe', and its relation to the *startos* (which may itself be a military grouping), the distinction between an *astos* and a *poliatas*, both clearly some kind of formulation of the rights-bearing civic actor, and the ancestry and spread of the *agelai*, apparently an adolescent age-grade so far only attested at Eltynia [@Gagarin2016, 36; 78-80].
-Indeed, many terms which might illuminate social divisions and groupings within the Archaic Cretan populace are only attested at one or two sites, and even then sometimes appear to differ in their form or function.
-Frustrating though this can be, it has also seen the image of Cretan austerity and uniformity begin to be challenged, or at least nuanced, by an appreciation of the regional variety and idiosyncracy apparent from the laws.
-
-It wouldn't be helpful at present, then, to draw any simple parallel between the Knossian burial groups and, say, powerful *startoi*, a term which is not even attested outside of Gortyn in this period (ibid., 1-2).
-But archaeological and textual evidence suggests that, in general terms, such supra-familial kinship groups were undergoing a transition during this period, in ways reflected at Knossos, and thus supporting the attribution of the chamber tombs to such groups.
-Elsewhere on Crete, changes around this time include a shift to unmarked, individual burials, and more austere grave goods, as well as an apparent reorientation of sanctuary deposits away from extra-mural to suburban spaces, and a rise in utilitarian, mass-produced votives (Perlman & Gagarin 2016, 34-6).
-At Azoria, major urban remodelling in the 7th century saw the earlier, agglutinative house forms of neighbouring Kavousi Vronda and Kastro give way to larger, more formally planned dwellings with little evidence of renovation or remodelling, while larger, apprently civic, buildings were constructed to accommodate major public gatherings.
-At Knossos, a small number of pithos burials in pits do appear in the KNC and Fortetsa volumes (Brock 1957, 98; Coldstream & Catling 1996, 162), and recent re-excavation of the Teke plot identified numerous small pit burials around three chamber tombs (Evely 2012), raising the possibility that many more such burials filled the interstices between larger chamber tombs elsewhere, but have gone unrecognised.
-The insecure dates of these pits, however, makes it unclear whether they represent a genuine Archaic trend.
-
-These varied lines of evidence may all suggest that with the emergence of the archaic *poleis* on Crete, there was a concomitant restructuring of wider social groups, institutions, and behaviours.
-Perlman and Gagarin (Gagarin & Perlman 2016, 32-7) argue that, at least by the Classical period, supra-household kinship units such as the *startos* or *pyla*, whatever their exact formulations, had become mostly ceremonial, political entities, and had little influence over the social organisation of the Cretan city states.
-Was this transformation linked to an emerging emphasis on the citizen, the state, and ideologies of relative egality, within which communal gatherings, intra-mural ritual, and individual burial replaced ostentatious, kin-centred acts of gathering, dedication, and burial?
-It is plausible, though as Perlman and Gagarin (32-37) concede, the decline of ostentatious burial could just as well represent a voluntary abandonment by groups assured of their economic and political influence, finding the symbolic expression of kin-relations in death no longer a necessary expense.
-
-These issues, naturally, recall the conclusions of Morris (1987, 171-211), who argued 30 years ago that the emergence of the *polis* across the Aegean led to a tension between elite preferences for conspicuous display, and a downplaying of social difference in death, as promoted by the ‘middling’ ideology that the *polis* engendered and promulgated.
-In many areas, this resulted in increasing restriction of elaborate tombs, despite a ‘rise in aristocratic family consciousness’ that saw certain wealthy tombs buck the trend, until ‘such overt groupings were subsumed within the larger “citizen cemeteries”’ (ibid., 185).
-In his well-known model of the *agathoi* and *kakoi*, Morris argued that, at various times, the relevant spheres of kinship demonstrated in burial could widen and constrict, with more and less distantly related lineages drawn into association through changing social strategies, until the pre-eminence of kinship relations in the structuring of political power declined.
-
-Returning to Knossos, while aspects of the noted expansion in the burying population would seem readily explicable within a model like Morris', certain observations sit more uneasily.
-Firstly, the burials at issue here are in many ways *not* ostentatious in comparison to the norm.
-We have seen that the 'richest' tombs in terms of finds are not so 'rich' after all, when one takes account of the number of burials.
-They are not phyiscally larger, and do not exhibit monopolies over any of a range of potentially 'valuable' offerings.
-Likewise, Morris claims that periods of expansion in the rite of burial will correspond to greater variety in the associated assemblages, yet, at Knossos, we have seen that the range of grave goods in fact diminished and crystallised during the late 8th and early 7th centuries.
-Perhaps the expansion was driven by a widening of access to the burial rite by the so-called *agathoi*, with the standardised burial assemblages coming to reflect their shared affinal relations.
-Or else, the widening of access to these tombs actually heralds the breakdown in traditional kin-based stewardship thereof, perhaps with new, non-lineal corporate groups becoming  the custodians of the tombs, with the limited funereal repertoire thus reflecting the levelling ideology of the emergent state.
-The exclusive focus of the expansion in a very limited number of older tombs, though, would seem to argue in favour of the former.
-What seems certain is that this was a period of transition wherein, whatever their identity, certain actors invested heavily in a final flurry of chamber tomb burials, with an emphasis on tradition, an established repertoire of items, rites, and venues for one's passage to the afterlife.
-
-And this image ties into the broader picture developing of the island at this time.
-It is telling that the peak for imported items in the tombs came in the 9th century; the 'Orientalising' as a time period feels increasingly misapplied to Crete, when, at Knossos at least, it witnessed far lower rates of oriental imports, and only limited adoption of the figural motifs which characterise its appearance elsewhere.
-The observed rise in imitations of Near Eastern vessels in the late 8th and early 7th century further accords with our growing awareness of the selective nature of archaic and classical Crete's external relations.
-Erickson has, contrary to many, argued that *symposia* were practised on Crete, but that their practitioners simply favoured simpler vessel forms.
-The local, selective adoption of, and responses to, wider social and material trends across the contemporary Aegean is perhaps becoming the hallmark of archaic Crete, and finds further justification in the trends identified in the EIA cemeteries at Knossos.
-
-# Conclusion
-
-It is important that the present study was a comparative investigation of both the Fortetsa and Knossos North Cemetery.
-It is notable how similar the observed trends are across both these cemeteries, adding significant credence to the inferences drawn.
-That both underwent similar developments during the EIA suggests not only that they catered to similar sub-sections of the Knossian population, but that some of the socio-economic and political changes inferred and discussed were real, and community-wide.
-We should be cautious in speculating as to whether, for instance, particular families, lineages, groups, or classes made persistent use of these different necropoleis, but such possiblities are not inconceivable.
-Further work should seek to extend these comparisons, both into trends unexplored in the present study (such as, for instance, the curious reuse of a number of 11th century tombs in the late 8th and early 7th centuries), and out to the wider funerary landscape of Knossos.
-I hope to make just such a study but, for the time being, the datasets I am making available alongside this article include the other published EIA tombs in the area, so that others may also explore their possible significance.
-
-There were undeniable limitations to the present study, also.
-For one, non-ceramic artefacts were underrepresented, mostly because they lack accurate dating in the manner of pottery and future work could focus on dating these artefacts more closely, permitting their future inclusion into temporal analyses.
-Issues of gender were sadly also little considered, though this was largely due to limitations in the data themselves.
-There have also been a number of tombs discovered in the last 20 years that fill in the spaces between the Teke plot and the Medical Faculty Site, and ideally these unpublished tombs should be included in further analyses as they, along with the unconsidered Teke tholos and associated tombs, are almost surely part of a single contiguous cemetery (Blackman 1999, 113; Catling 1983, 51; Evely 2012; French 1991, 68–9).
-Finally, a preliminary study was recently taken by the author of select boxes marked as 'unpublished' from the KNC deposits kept at the British School at Athens at Knossos.
-Despite going unmentioned in the publication, this material was, at least in the boxes seen, not inconsiderable and so, regretably, future study will be necessary to determine whether it threatens the integrity of quantitative analyses such as this.
-Suffice to say that, at present, it is another pinch of salt with which to take the findings.
-
-These limitations notwithstanding, the foregoing analysis demonstrates the value of quantitative approaches to the archaeological record of the period.
-Such methods allow us not only to identify types of variability in the archaeological material, but also assess the scale of that variability.
-With large datasets, such as those used in the present study, we are able to locate and visualise covariance and significant correlations between different variables, plot changes in the data through time and, ultimately, draw on our findings to inform discussions of social processes.
-Though the construction of databases can be time consuming, once made they offer the opportunity to subdivide, compare, and plot the data in numerous ways quickly and easily, facilitating investigation of the material from multiple angles.
-Such methods are also more independently verifiable than conclusions drawn from more qualitative analyses, for two main reasons.
-Firstly, though explanations for any patterning in the data still require interpretation, the existence of such patterns may be more easily and reliably assessed via the use of quantitative and statistical tests of correlation and covariance.
-And secondly, by making datasets available for use by others, as with the data and code for the present study, one’s conclusions can be independently tested, and alternative interpretations can be forwarded, by other scholars.
-It is hoped that the present study demonstrates the wealth of insights yet to be uncovered by the application of more systematic, quantitative methods.
-The databases produced for the above analysis contain a large of amount of unutilised data that could facilitate a range of future investigations, and all are welcome to it.
-With more such analyses, we could begin to a construct a fuller account of the regional diversity in mortuary, and wider social, practices that characterised the communities of EIA and archaic Crete.
-
-# Acknowledgements
-
-<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
-##### pagebreak
-
-# References 
-<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
-<div id="refs"></div>
-
-##### pagebreak
-
 ### Colophon

### 2019-05-24 analysis/paper/paper2.Rmd
Further edits and new Figures

--- 2019-05-23 analysis/paper/paper.Rmd
@@
   fig.path = "../figures/",
-  dev="png"
+  dev="jpg"
 )
@@
 To begin with the first of these, it may be felt that, of all ancient sites on Crete, Knossos is the last in need of further archaeological attention.
-The city and palace of Knossos, first excavated well over a century ago [@Evans1921; -@Evans1928; -@Evans1930], retain a pre-eminent position in popular and academic accounts the island’s Bronze Age, and few, if any, other Cretan sites have been so thoroughly dug, reconstructed, and contested.
+The city and palace of Knossos, first excavated well over a century ago [@Evans1921; -@Evans1928; -@Evans1930], retain a pre-eminent position in popular and academic accounts the island’s Bronze Age (BA), and few, if any, other Cretan sites have been so thoroughly dug, reconstructed, and contested.
 
@@
 Research into this once maligned period has grown apace in the last few decades, beginning with the pioneering works of Coldstream [-@Coldstream1977], Desborough [-@Desborough1972], and Snodgrass [-@Snodgrass1971] and culminating in the rejection by many scholars of the image of a society bereft of its former palatial glory.
-Certainly, few would any longer defend the appelation or interpretive bagguage of the ‘Dark Ages’ [on the origins of which, see @Kotsonas2016; @Morris1997].Nonetheless, our knowledge of many EIA settlements remains often underdeveloped in comparison with those of the second millennium BC.
+Certainly, few would any longer defend the appelation or interpretive bagguage of the ‘Dark Ages’ [on the origins of which, see @Kotsonas2016; @Morris1997].
+Nonetheless, our knowledge of many EIA settlements remains often underdeveloped in comparison with those of the second millennium BC.
 
 But more importantly, and welcome though they are, a number of recent publications on the socio-economic [@Wallace2010a] and settlement [@Nowicki2000] changes which accompanied the BA-EIA transition have directed scholarly attention toward a number of elevated, defensible settlements supposedly founded in response to the dangers of this turbulent period.
-This research agenda is likewise reflected in survey work, which has focussed extensively around the Bay of Mirabello, and the Isthmus of Ierapetra.
+This is reflected too in survey work, which has focussed extensively around the Bay of Mirabello, and the Isthmus of Ierapetra, where such sites are numerous.
 This has, unfortunately, engendered a certain disregard for settlement characterised by quite different topographies, especially in the central region of the island.
-There remains, therefore, a need for studies addressing the social dynamics at such sites (Knossos, Phaistos, Grivila etc.), with a long-term view to comparative appraisal of the divergent trajectories of communities across the island in the EIA [@Kotsonas2011b; @Whitley2011, 667–8].
-
-As for the second and third of my stated interests, several authors have recently demonstrated the great potential of quantitative, statistical, spatial, and network analyses in elucidating social dynamics, settlement patterning, and systems of inter-regional communication and exchange across Crete and the wider Aegean.
+There remains, therefore, a need for studies addressing the social dynamics at such sites (Knossos, Phaistos, Grivila, Eleutherna etc.), with a long-term view to comparative appraisal of the divergent trajectories of communities across the island in the EIA [@Kotsonas2011b; @Whitley2011, 667–8].
+
+As to the second and third of my stated interests, several authors have recently demonstrated the great potential of quantitative, statistical, spatial, and network analyses for elucidating social dynamics, settlement patterning, and systems of inter-regional communication across Crete and the wider Aegean.
 Such work encourages the use of large (often pre-existing) datasets, and the sharing of both data and methods to facilitate reproducible, collaborative, and innovative research [@Bevan2013; @Knappett2011a].
@@
 ```{r map, fig.cap="Map of the Knossos area showing the locations of the various tomb clusters and cemeteries, with EIA pottery distributions from the KULP marked in orange"}
-knitr::include_graphics("../figures/Knossos Tombs Map.png")
+knitr::include_graphics("../figures/Knossos Tombs Map.jpg")
 ```
@@
 Survey aside, our most abundant evidence for EIA Knossos comes from the mortuary record (see Figure \@ref(fig:map).
-The transitional period between the LBA and EIA yields few if any securely datable interments, but later the evidence becomes more plentiful, with multiple burials known from the Kephala ridge [@Coldstream1963, 38; @Hogarth1899, 82-5], the modern suburbs of Ayios Ioannis and Atsalenio [@Boardman1960; @Davaras1968, 133–46], the cave of Mavro Spelio [@Antoniadis2012, 58-9], and the Khaniale Teke site with its well-known tholos, mooted as evidence for the presence immigrant craftsmen from the Near East [@Boardman1967;@Hutchinson1954; cf.
+The transitional period between the Late Bronze Age (LBA) and EIA yields few if any securely datable interments, but later evidence is more plentiful, with multiple burials known from the Kephala ridge [@Coldstream1963, 38; @Hogarth1899, 82-5], the modern suburbs of Ayios Ioannis and Atsalenio [@Boardman1960; @Davaras1968, 133–46], the cave of Mavro Spelio [@Antoniadis2012, 58-9], and the Khaniale Teke site with its well-known tholos [@Boardman1967;@Hutchinson1954; cf.
 @Hoffman1997 @Kotsonas2006].
@@
 
-Fortunately, far larger burial accumulations are to be found in the Fortetsa and the Knossos North cemeteries, which together comprise some 134 tombs, ranging in date from the Subminoan to Late Orienatalising periods.
+Fortunately, far larger burial accumulations are to be found in the Fortetsa and the Knossos North cemeteries, which together comprise some 134 tombs, ranging in date from the Subminoan (SM) to Late Orienatalising (LO) periods.
 Though many were robbed in antiquity, and others destroyed by the digging of later burials, building work, or just the ravages of time, they still represent the most abundant source of evidence for not just mortuary practices, but for social life of any kind at EIA Knossos.
-They are thus the ideal candidates for a quantitative analysis of the kind advanced here.
 
@@
 These tombs cluster in three main groups.
-The largest of these, including tombs IX, VIII, VII, V, II, III, IV, V and P, is arranged on a north-south axis, with the *dromos* of each tomb extending downslope to the west.
-Further north lies the cluster of tombs OD, X, XI, LST and BLT.
-The third group lies south of the first, and comprises tombs Θ, Ϙ and F.
-Finally, three tombs excavated in 1933 – L, TFT and Π – which lie close to the main road north of the Acropolis hill [@Brock1957, 1-2], are now considered to belong the Knossos North Cemetery.
-
-To the northeast of the Fortetsa cemetery, a collection of ten badly damaged tombs known, after the year of their excavation, as the Fortetsa 1967 tombs, were probably the southernmost burials of the North Cemetery.
+The largest of these, comprising nine tombs, is arranged on a north-south axis, with the *dromos* of each tomb extending downslope to the west.
+To the north lies a cluster of five tombs and, to the south, another of three.
+Finally, three tombs excavated in 1933 – L, TFT and Π – which lie close to the main road north of the Acropolis hill [@Brock1957, 1-2], are now considered to belong the Knossos North Cemetery (KNC).
+
+To the northeast of the Fortetsa cemetery, a collection of ten badly damaged tombs known, after the year of their excavation, as the Fortetsa 1967 tombs, were probably the southernmost burials of the KNC.
 Unfortunately, all had been looted at the time of excavation, and no *in situ* deposit was found in any tomb.
 They are thus excluded from present analysis.
-Northwest of these tombs, and just south of the Teke Tholos, the Teke tombs represent the northern reaches of the KNC.
+Northwest of these, and just south of the Teke Tholos, the Teke tombs represent the northern reaches of the KNC.
 The central section of the KNC comprises the ‘Medical Faculty’ site, named for the extension of the University of Crete that prompted major rescue excavations here in 1978.
-Some 89 tombs of Subminoan to Late Orientalising date were found, by far the largest single concentration of EIA tombs in the whole Knossos area [@Coldstream1996a, 53-55].
+Some 89 tombs of Subminoan to Late Orientalising date were found, by far the largest single concentration in the whole Knossos area [@Coldstream1996a, 53-55].
 
@@
 The initial foundation of the cemeteries has received much attention.
-It has been claimed that no tomb crosses the LM IIIC-SM divide [@Coldstream1998, 58; @Coldstream2006, 582), though there is a possibility that this is a classificatory artefact [@Hallager2010].
+It has been claimed that no tomb crosses the Late Minoan (LM) IIIC-SM divide [@Coldstream1998, 58; @Coldstream2006, 582), though there is a possibility that this is a classificatory artefact [@Hallager2010].
 Instances of BA tombs reused the EIA are known (e.g.
@@
 
-During the use of the cemeteries, correspondences have been highlighted between changing pottery styles, and patterns of tomb construction and abandonment.
-The Protogeometric B style, first defined by Brock [-@Brock1957], 143], who hailed it as “the most remarkable phase in Cretan vase-painting”, is apparently a Cretan anomaly, and evinces a melding of Attic geometric motifs, experimental patterns possibly deriving from Near Eastern metalwork, and iconography drawn seemingly from the Minoan repertoire (Coldstream 1984, 93–4).
+During the use of the cemeteries, correspondences have been highlighted between changing pottery styles and episodes of tomb construction and abandonment.
+The Protogeometric B (PGB) style, first defined by Brock [-@Brock1957], 143], who hailed it as “the most remarkable phase in Cretan vase-painting”, is apparently a Cretan anomaly, and evinces a melding of Attic geometric motifs, experimental patterns possibly deriving from Near Eastern metalwork, and iconography drawn seemingly from the Minoan repertoire (Coldstream 1984, 93–4).
 Coldstream has argued that the concurrence of this style with a rash of newly founded tombs points to a reformulation of elite mortuary display, noting how, in this period also, redeposited Minoan larnakes began appearing in and around some tombs, in one case even inspiring the decoration of a PGB vessel (in T.
@@
 
-The final mystery of the chronological sequence at the KNC and Fortetsa is the quite abrupt cessation of burials around 630 BCE.
+The final mystery of the chronological sequence at the KNC and Fortetsa is the quite abrupt cessation of burials around 630 BC.
 There had been changes in tomb use in both cemeteries in the preceding century, with few new tombs built, though interments in old tombs continued [@Brock1957, 4; @Cavanagh1996; 651-3].
-This curiosity has, along with the equally scanty evidence for the period in the wider settlement, become known as the ‘Archaic Gap'.
+This curiosity has, along with the equally scanty evidence for the subsequent period in the wider settlement, become known as the ‘Archaic Gap'.
 The causes of this lacuna, which persists until the recrudescence of archaeological evidence in the Late Archaic period, remain unknown.
@@
 A popular approach to the study of Knossos’ EIA tombs has been the investigation of ‘object biography’, a term derived from Appadurai [-@Appadurai1986] and Kopytoff [-@Kopytoff1986].
-Influenced by Mauss’ [-@Mauss1954] formulation of ‘the gift’, these authors stressed the mutuable nature of an object’s social meanings, and its capacity for acquiring a ‘biography’, through the transformations of its social existence.
-These ideas proved attractive with their evocation of the Homeric motif of gift-giving [@Whitley2002, 220-1], and the remarkable archaeological parallels for such objects – such as the boar’s tusk helmet from Tomb 200+ and that given to Odysseus by Meriones (*Iliad* 10.260-271).
-Thus, a particular interpretive lens, melding Homeric and anthropological ideas, has been applied to certain elaborate burials.
+Influenced by Mauss’ [-@Mauss1954] formulation of ‘the gift’, they stressed the mutuable nature of an object’s value, and its capacity for acquiring a ‘biography’ through the transformations of its social existence.
+These ideas, fertilised by the Homeric motif of gift-giving [@Whitley2002, 220-1], and the remarkable archaeological parallels for such objects – such as the boar’s tusk helmet from Tomb 200+ and that given to Odysseus by Meriones (*Iliad* 10.260-271) - have been brought to bear on certain elaborate burials of the EIA.
 
 Catling’s discussion of the Tomb 200+ complex is a notable example.
-Tomb 201 contained the remains of at least two adults, and possibly a child.
-Objects included a bronze sword, spearhead, and arrowhead, an iron dirk or knife, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned helmet [@Catling1995, 123].
+Tomb 201 contained the remains of at least two adults, and possibly a child, a bronze sword, spearhead, and arrowhead, an iron dirk or knife, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned helmet [@Catling1995, 123].
 Catling identifies similarities between this tomb and burials at Tiryns, Kaloriziki, and the famous *hērōön* of Lefkandi.
-All contained imported items at an early date, were of ‘warrior grave’ type, and formed loci for later burials.
+All contained imported items, were of ‘warrior grave’ type, and formed loci for later burials.
 Catling argues that those individuals who dared to travel abroad in this time of more limited seafaring, accruing precious artefacts and stories of distant lands, would have been accorded exceptional treatment in death [@Catling1995, 127–8].
@@
 Complementing the theme of heroes abroad has been a similar interest in foreigners at home.
-Two caches of gold jewellery, found in pits flanking the entrance to the Teke Tholos, were argued by Boardman [-@Boardman1967] to be foundation deposits – a Near Eastern custom – for the burial place of an immigrant craftsman, probably Phoenician in ancestry.
-Kotsonas has revisited this attribution, suggesting that the tholos is more likely the resting place of an elite individual or family with a monopoly over a metal workshop’s output [-@Kotsonas2006, 155–9].
-While certainty in such individual cases may be moot, it seems unlikely that traders or craftsmen of non-Cretan heritage never visited or resided at Knossos during the EIA [@Hoffman1997, 176–85; @Schreiber2003, 293-306].
+Two caches of gold jewellery, found in pits flanking the entrance to the Teke Tholos, were argued by Boardman [-@Boardman1967] to be foundation deposits – a Near Eastern custom – for the burial place of an immigrant Phoenician craftsman.
+Kotsonas has revisited this attribution, suggesting that the tholos is more likely that of an elite individual or family with a monopoly over a metal workshop’s output [-@Kotsonas2006, 155–9].
+While certainty in such individual cases may be moot, it seems unlikely that foreign traders or craftsmen never visited or resided at Knossos during the EIA [@Hoffman1997, 176–85; @Schreiber2003, 293-306].
 
@@
 
-Generally speaking, the above analyses have tended to rely more on particular finds, tombs, or observable, but limited, patterns in the data.
-Only a few authors have applied more quantative approaches, but these point toward their potential utility.
-Cluster analyses have been employed by both Cavanagh [-@Coldstream1996, 653-7] and Antoniadis [-@Antoniadis2012, 193-7], the most sophisticated statistical techniques so far applied to the present material.
+Generally speaking, the above analyses have tended to rely more on particular finds, tombs, or observable, but limited, patterns in the data; only a few authors have pursued more quantative approaches.
+Cluster analyses have been employed by both Cavanagh [-@Coldstream1996, 653-7] and Antoniadis [-@Antoniadis2012, 193-7], the most sophisticated statistical techniques so far applied to the material.
 Cavanagh’s analysis of tomb architecture has been mentioned, while Antoniadis’ more holistic approach corroborates quantitatively some of the intuitive correspondences noted between specific tombs in previous studies.
 Antoniadis [-@Antoniadis2012, 172-6] also divided the EIA tombs of Knossos into three groups, those with fewer than 10 pots, fewer than 50, and more than 50, and found that, in each successive case, the proportions of tombs containing 'oriental' imports and imitations increased, the implication being that a hierarchy existed, with unequal access to such prestigious goods.
-Yet even here the available data could be further interrogated, such as with the relationship between imported items and so-called tomb ‘wealth’.
-The present study seeks to extend and nuance a number of observations that have been made in previous studies, as well as revealing patterns thus far unacknowledged, through use of a more systematic approach to quantitative analysis.
+Yet even here the available data could be further interrogated, such as with the number of imported items *per burial*, something Antoniadis does not explore.
+The present paper seeks to extend and nuance a number of observations that have been made in previous studies, as well as revealing patterns thus far unacknowledged, through use of a more systematic approach to quantitative analysis.
 
@@
 The first includes an entry for each individual ceramic vessel, by far the largest portion of all known finds.
-Each entry includes: the tomb from which the vessel came; the vessel shape; whether it was an open or closed vessel; its main dimensions; its ceramic period and corresponding absolute dating measures (see below).
-In addition, all imports, Oriental imitations, and antique (that is, Bronze Age) artefacts were accordingly categorised, using the studies of Antoniadis (2012), Jones (2000), and Crowe (2016) to supplement omissions in the original excavation reports.
+Each entry includes, among other things, the tomb, the vessel shape, its surviving dimensions, its ceramic period and corresponding absolute dating measures.
+In addition, all imports, Oriental imitations, and antique (that is, Bronze Age) artefacts were accordingly categorised, using the studies of Antoniadis (2012), Jones (2000), and Crowe (2016).
 In total, this database comprises just over 4540 individual artefacts.
@@
 The second database is similar, but with each entry corresponding to a find other than a ceramic vessel.
-A ‘material’ variable was also added, as of course these finds were not all made of the same substance.
+A ‘material’ variable was added, as of course these finds were not all made of the same substance.
 This database included some 1620 individual artefacts.
@@
 Each tomb was accorded its own entry, with variables including dating (both in relative and absolute terms), surviving dimensions, and the known and estimated number of interments, as well as a tally of the objects recorded in the other two databases.
-These tallies were arranged by shape for the pottery, and by material and category (e.g.
-weapon, adornment, seal, funerary paraphernalia) for the other objects.
 Finally, the total number of imports, imitations and antiques in each tomb was calculated.
 
-Taken together, these databases constitute a detailed record of all the finds from the Fortetsa and KNC excavations, and offer the opportunity for carrying out a range of robust quantitative analyses.
-It must be conceded that, with many of the tombs evidently looted, dug into or destroyed over the centuries, this cannot be a complete picture of the tombs’ original contents.
+Taken together, these databases constitute a detailed record of all the finds from the Fortetsa and KNC publications, and offer the opportunity for carrying out a range of quantitative analyses.
+It must be conceded that, with many of the tombs evidently looted or damaged over the centuries, this cannot be a complete picture of the tombs’ original contents.
 If we assume, though, that natural or accidental destruction was relatively indiscriminate, and that, even in cases of looting, certain object classes (particularly cinerary urns) were seldom robbed, then the data are by no means rendered useless.
-As Snodgrass (1996) notes, the situation means we should be extremely cautious about making negative arguments – that is, conclusions drawn about the lack of features in certain tombs – but may be somewhat more hopeful in the forming of positive ones.
+As Snodgrass [-@Snodgrass1996] notes, the situation means we should be extremely cautious about making negative arguments – that is, conclusions drawn about the lack of features in certain tombs – but may be somewhat more hopeful in the forming of positive ones.
 
@@
 
-An ever-growing number of archaeologists have begun to advocate the utility, indeed the necessity, of making available not just the data underpinning their professional output, but the full methodologies by which their analyses were conducted, and in formats which facilitate reproducibilty and further investigation.
+An ever-growing number of archaeologists have begun to advocate the utility, indeed the necessity, of making available not just the data underpinning our academic output, but the full methodologies by which our analyses are conducted, in formats facilitating reproducibilty and further investigation.
 It is felt that, with the opportunities afforded by current technologies (digital databases, statistical software packages, online repositories etc.), there remain ever fewer obstacles - and excuses - preventing the full and free sharing of both data and workflows.
@@
 
-This is not the place for a detailed investigation into the lack of reproducible research in archaeology (for this, see ...).
-Bu this article is, among other things, an attempt to bring these principles of maximal reproducibility to the archaeology of Early Iron Age Crete.
-Within this field, catalogues and excavation reports abound and, though many of these may be forgiven based on their dates of publication, the lack of readily utilisable digital data from a number of recent articles and monographs, indeed, the failure to even present the data such that they could be manually digitsed, is less pardonable.
-Preparation of databases such as those of the present study can (and did) involve prolongued trawling of printed tables (often without consistent formatting, coding or presentation which would enable easy digitisation), cross-referencing artefacts listed inconsistently in different contexts, or else grappling with catalogues which, through their unexplained lack of critical contextual information, offer little research potential beyond their descriptive face-value.
+This is not the place for a detailed investigation into the lack of reproducible research in archaeology.
+Bu this article is, among other things, an attempt to bring these principles of maximal reproducibility to the archaeology of EIA Crete.
+Within this field, catalogues and excavation reports abound and, though many of these may be forgiven based on their dates of publication, the lack of readily utilisable digital data from a number of recent articles and monographs, indeed, the failure to even present data such that they could be manually digitsed, is less pardonable.
 
@@
 
-In this vein, the present article has been composed making use of Ben Marwick's "rrtools", an open-source package for the statistical software R Studio, which provides the author with the tools to create a self-contained "compendium", which can be freely downloaded, and from which anyone can accurately reproduce the entire workflow of the present analysis.
-This analysis was conducted in R, and the full code, as well as the raw and derived datasets, complete bibliography (in BibTex format, for export to any major reference manager), figures, and article as originally written are included in the package.
-It can be downloaded as a repository from GitHub, at the following address: https://github.com/DCPollard94/knossoscemeteries, and the raw datasets have been stored with the online repository Figshare, here:
-
+In this vein, the present article has been composed making use of Ben Marwick's "rrtools" [see @Marwick2017], an open-source package for the statistical software R Studio, which provides the author with the tools to create a self-contained "compendium", which can be freely downloaded, and from which anyone can accurately reproduce the entire workflow of the present analysis.
 Though not visible in the present paper, every chart and table herein was produced directly, through written code, from the raw data using R.
-The full code can be viewed and followed by downloading the repository from GitHub, and all the data are made available through an MIT license, which permits unlimited use, ammendation, and dissemination thereof, yet frees the present author from any responsibilty for that use, or for any perceived errors or inaccuracies in the data.
-It is sincerely hoped that others may find interest in, further explore, and expand upon the research presented here by downloading this compedium.
+This code, as well as the raw and derived datasets, complete bibliography (in BibTex format, for export to any major reference manager), figures, and article as originally written are included in the compendium.
+It can be downloaded as a repository from GitHub, at the following address: https://github.com/DCPollard94/knossoscemeteries, and the raw datasets have been stored with the online repository Figshare, here: (NB a URL will be included pending acceptane of the article).
+The data are made available through an MIT license, which permits unlimited use, ammendation, and dissemination thereof, yet frees the present author from any responsibilty for that use, or for any perceived errors or inaccuracies in the data.
+It is sincerely hoped that others may further explore and expand upon the research presented here by downloading this compendium.
 
@@
 A first important step in preparing the data for analysis was the estimation of the number of individuals buried in each tomb.
-The method chosen broadly follows that of Cavanagh [-@Cavanagh1996, 659-660], who took the number of pithoi (the most common urn type) and the number of pithoi, amphorae and kraters (which were in some cases used as urns) as middle and upper estimates of the number of interments, with all securely identified interments (in the form of ashes or osteological material) being the lower limit (Figure \@ref(fig:burialestimate)).
+The method chosen broadly follows that of Cavanagh [-@Cavanagh1996, 659-660], who took the number of pithoi (the most common urn type) and the number of pithoi, amphorae and kraters (which were sometimes used as urns) as middle and upper estimates of the number of interments, with all securely identified interments (in the form of ashes or osteological material) being the lower limit (Figure \@ref(fig:burialestimate)).
 In the present study, the mean of these three estimates was taken, to provide a single aggregate measure for the purposes of analysis (where no interments were known, this variable was still assigned a value of "1", for all tombs must have contained at least one burial).
@@
 For most diachronic analyses, however, pithoi alone were taken as the best proxy for burial numbers.
-In most cases, and often simply by virtue of the methods employed, any tombs lacking identified remains or a single vessel suitable for use as an urn were not considered, thus excluding what are presumably among the most heavily damaged and looted tombs.
+In most cases, any tombs lacking identified remains or a single vessel suitable for use as an urn were not considered, thus excluding what are presumably among the most heavily damaged and looted tombs.
 
@@
 
-A central concern of the present study was how to deal with the temporal uncertainty implicit within the assigning of individual vessels to stylistically-defined periods.
-A vessel dated to the Protogeometric B period could, on our best estimates of absolute chronology, date to any point between 840 and 810 BC.
-This situation results in major challenges to the visualisation and analysis of such data through mean which normally rely on event- and not range-based data inputs.
-A simple, if unsatisfying, solution is to use the midpoint of the range assigned to each vessel (i.e.
+A central concern of the present study was how to deal with the temporal uncertainty implicit in the assigning of individual vessels to stylistically-defined periods.
+A vessel dated to the PGB period could, on our best estimates, date anywhere between 840 and 810 BC.
+This situation results in major challenges to the visualisation and analysis of such data through means which normally rely on value- and not range-based data inputs.
+A simple, if coarse, solution is to use the midpoint of the range assigned to each vessel (i.e.
 825 for our PGB example).
-This method is used in a couple of instances in the present paper when the course-grained sorting of the data (into 50 year periods) renders the inaccuracies of such a method less significant.
+This is used in a couple of instances in the present paper when the division of vessels into centuries renders the inaccuracies of such a method less significant.
 
@@
 
-But the artificial clustering this technique engenders around select absolute dates proves misleading when, for instance, the total number of pithoi deposited by date are plotted via a histogram (Figure \@ref(fig:allpithoi)); the apparent cessation of burial activity for 30 years in the 8th century is simply a fiction generated by the processing of the data.
+But the artificial clustering this technique engenders around select absolute dates can prove misleading, as in Figure \@ref(fig:allpithoi), a simple plot of vessel counts based on these average dates; the apparent cessation of depositions for 30 years in the 8th century is simply a fiction generated by the processing of the data.
 
@@
 Here, rather than assign a pot a single date, the entire time span under investigation is divided into equal chunks (in this case, decades), and probability values summing to 1 are then assigned for each vessel, per chunk of time, based on its accordant date range.
-So, our PGB vessel, datable between 840 and 810 BC would register a value of 0.33 for each of the three decades within which it was likely produced.
+So, our PGB vessel, datable between 840 and 810 BC, would register a value of 0.33 for each of those three decades.
 These values for individual vessels can can then simply be summed to produce tomb- or cemetery-level signatures.
@@
 ##Introduction
-The great advantage of the large, quantitative data set compiled for the present analysis is that it permits investigation of several nested levels of potentially significant patterning in the material.
+The great advantage of the large, quantitative data set employed here is that it permits investigation of several nested levels of potentially significant patterning.
 From individual burials, to collective tombs, to the cemeteries as a whole, and between cemeteries in the landscape, we have the opportunity to study and compare both the synchronic and diachronic variation across these various levels.
-In so doing, I first investigate the broad temporal trends in tomb construction, burial, and pottery deposition across both cemeteries.
-Through this, a number of notable tombs are isolated, which I then examine in an attempt to understand what sets them apart from the rest.
+In so doing, I first investigate the broad temporal trends in tomb construction, burial, and pottery deposition across both cemeteries, isolating a notable trend among a select group of  tombs.
+I then examine these more closely, in an attempt to understand what sets them apart from the rest.
 Tomb dimensions, rates of burials, and changes in the composition of assemblages through time are all considered, to produce an account which speaks to the changing social landscape of EIA Knossos, drawing together issues of group identity, external relations, and the communication of wealth and status through funerary practice.
@@
 An obvious place to begin is with the chronology of the burials themselves.
-The rate and number of burials being made through time provides the background against which to consider all other developments.
+The rate and number of interments being made through time provides the background against which to consider all other developments.
 
@@
 In both cemeteries, the rate of cremation burial began rising around the end of the 9th century, plateaued slightly in the early 8th, and then rose again to a peak in the late 8th and early 7th centuries.
-Across all vessels, the developments are more volatile, but again there was growth beginning in the later 10th century - reaching an especially marked peak in the KNC by the end of the 9th - then a slight decline and levelling into the 8th century, before a final early 7th century flourit of depositional activity.
-
-The near total absence of pithoi from the 11th and early 10th centuries reflects the delayed onset of cremation burial, with inhumations representing a significant proportion of earlier interments.
-Nonetheless, the clear inference would appear to be that, at both cemeteries, the late 8th and early 7th centuries witnessed an unprecedented level of burial activity, something which makes the sudden and total decline of chamber-tomb burials around 630 BC even more notable.
+Across all vessels, the developments are more volatile, but again there was growth beginning in the later 10th century - reaching an especially marked peak in the KNC by the end of the 9th - then a slight decline and levelling into the 8th century, before a final flourit of depositional activity.
 We might assume that these developments would be mirrored in the construction of new tombs, but this proves not to be the case.
 
-```{r tombconstruction, fig.dim=c(7.5,4), dpi=300, fig.cap="Visulations of tomb construction through time at both cemeteries.
+```{r tombconstruction, fig.dim=c(7.5,4), dpi=300, fig.cap="Tomb construction through time at both cemeteries.
 On the left are counts based on the earliest dated pottery in each tomb, while on the left are aoristic sums likewise based on the earliest ceramic-based assignations.
-In both cases, tombs lacking any securely dateable pottery (i.e.
-N/As) are excluded."}
+In both cases, tombs lacking any securely dateable pottery are excluded."}
 
@@
 Figure \@ref(fig:tombconstruction) presents two ways of dating the construction of the tombs.
-On the left, counts are based on the ceramic phase considered most likely to correspond to the tomb's first use, including the broad labels "PG", "G" and "O" where no greater specificity is possible.
+On the left, counts are based on the ceramic phase considered most likely to correspond to the tomb's first use, including "PG", "G" and "O" where no greater specificity was possible.
 On the right, aoristic sums are presented, with the uncertainty associated with those general assignations spread across the sub-periods they encompass.
 In both cases, the trend appears clear.
-The North Cemetery began with a rash of building during the, admittedly prolongued, SM period, followed by a protracted decline in rates of consruction, offset by small resurgences in the late 9th and 8th centuries.
+The KNC began with a large number of new tombs during the, admittedly prolongued, SM period, followed by a protracted decline in rates of consruction, offset by small resurgences in the late 9th and 8th centuries.
 It should be noted, too, that many of the later "tombs" are not chamber tombs, but a mix of pit, larnax, and pithos burials.
 The Fortetsa, meanwhile, saw next to no new chamber tomb construction from the 8th century onwards.
-The abatement of chamber tomb construction has been noted before (Brock 1957, 4; Cavanagh 1996, 652–3) and has been somewhat implicitly read as evidence of restructuring of the funerary rite, if not a decline in the fortunes of the wider community.
-But this image of a waigning investment in chamber burials sits uneasily with the observable growth in mortuary activity in the late 8th century.
+The abatement of chamber tomb construction has been noted before (Brock 1957, 4; Cavanagh 1996, 652–3) and somewhat implicitly read as evidence of restructuring of the funerary rite, if not a decline in the fortunes of the wider community.
+But this image of a waigning investment in chamber tombs sits uneasily with the observable growth in mortuary activity in the late 8th century.
 
@@
 
-Such countervalent trends are observable too in the number of tombs receiving in each period.
+These countervalent trends are observable too in the number of tombs receiving depositions in each period.
 Figure \@ref(fig:tombsinuse) presents the estimated number of tombs in use throughout the EIA at both cemeteries.
-At the Forteta, this number remained relatively stable from the MPG to LG periods before declining (and bear in mind, there were only around 20 tombs belonging to this cemetery throughout the period), while at the KNC, the numbers of tombs in use rose to a preak in the late 8th century before declining.
-The cessation of burials there certainly appears more sudden and marked than at Fortetsa, though, again, it was always a larger cemetery.
+At the Forteta, this number remained relatively stable from the MPG to LG periods before declining (and bear in mind, there were only around 20 tombs belonging to this cemetery throughout the period), while at the KNC, the number of tombs in use rose to a preak in the late 8th century before declining.
 But this is not the whole story.
@@
 This would seem to accord with Cavanagh's suggestion that "[t]he length of time the tombs last seems [...] to reflect the duration of the social group which had access to the sepulchre", and that, if such access was lineally inehrited, frequent fission would result in limited trans-generational use.
-We might expect, then, that it was the 8th and 7th century tombs coming into use that spurred the dramatic rise in burial rates previoulsy noted in this period.
+We might expect, then, that it was the 8th and 7th century tombs coming into use that spurred the dramatic rise in burial rates previoulsy noted.
 But this is not the case.
@@
 So as can be seen, though the late 8th and early 7th centuries represented the apex of burial activity across both cemeteries, it was not the tombs built in those centuries which received the lion's share of the burials and associated ceramics.
-This is all the more notable when, returning to Figure \@ref(fig:tombsinusebybuilddate), we see that the number of 9th and 10th century tombs in use by the late 8th and early 7th centuries was declining.
+This is all the more notable when, returning to Figure \@ref(fig:tombsinusebybuilddate), we see that the number of 9th and 10th century tombs in use by this time was declining.
 
@@
 The significance of these findings can be seen in Table 1.
-Using pithoi as a proxy for burials, we can estimate the changing rates of burial through the centuries based on how many tombs yielded how many pithoi of each period.
-And while for the most part burial rates remained around three per tomb per century, in the 8th and 7th centuries, tombs founded in the 9th century saw a rapid rise in the rate of burial, peaking at over 11 per tomb per century.
+Using pithoi as a proxy for burials, we can estimate the changing rates of burial through time based on how many tombs yielded how many pithoi during each centuring And while for the most part burial rates remained around three per tomb per century, in the 8th and 7th centuries, tombs founded in the 9th century saw a rapid rise in the rate of burial, peaking at over 11 per tomb per century.
 Though these estimates are no doubt very rough, the marked nature of this divergence suggests it is not simply an artefact of imperfect data.
@@
 
-We are yet to draw out *which tombs exactly* are responsible for the observible patterning so far discussed.
+We are yet to draw out which older tombs precisely are responsible for the observible patterning so far discussed.
 We might assume that the tombs which became the focus of intense burial activity in the late 8th and early 7th centuries would be those which yielded the most burials and grave goods when excavated and, happily, this supposition proves true.
@@
 And, in Figure \@ref(fig:toptombs), we see that seven of the 12 did exhibit late, intense episodes of deposition.
-Tombs P, 292, 75, 107, 285, II, 218, and 104 between them accrount for `r round(calc1,digits=2)`% of the entire pottery assemblage recovered from the `r nrow(tombs)` tombs considered here.
-All but Tomb 285 date to the 9th century, and together they underpinned the late spate of burials here considered.
+In fact, Tombs P, 292, 75, 107, 285, II and 218 between them account for `r round(calc1,digits=2)`% of the entire pottery assemblage recovered from the `r nrow(tombs)` tombs considered here.
+All but Tomb 285 date to the 9th century, and together they underpinned the discussed late spate of burials.
 Interestingly, Tombs X, G, Q, 104 and 283, however, evince a different pattern, with their respective peaks coming in the PGB and EG periods, around a century earlier.
 We have seen already in Figure \@ref(fig:aoristbycem) that, particularly in the KNC, the end of the 9th century was another period of increased burial activity.
-What, then, set these two groups of tombs apart?
-The recipients of the greatest numbers of burials and grave goods in both cemeteries, they nonetheless seem to have achieved that distinction via different histories.
-By looking more closely at the periods in which they were built, and in which they were most popular, I hope to shed light on the broader social contexts in which these tombs were flourished.
+Though recipients of the greatest numbers of burials and grave goods in both cemeteries, they nonetheless seem to have achieved that distinction via different routes.
+By looking more closely at the periods in which they were built, and in which they were most popular, I hope to shed light on the broader social contexts in which these tombs flourished.
 
@@
 Perhaps the tombs which went on to receive the most burials were notably large or grand when first built, reflecting the status of the associated burying group.
+
+```{r dimensionsboxplots, fig.dim=c(6,5),dpi=300, fig.align='center',fig.cap="Boxplots comparing various tomb dimensions as they vary by the century of tomb construction."}
+
+tombs$century_built<-factor(tombs$century_built,levels=c("11th","10th","9th","8th","7th"))
+dromwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_width,fill=century_built))+geom_boxplot()+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Century of Tomb Construction", y="Dromos Width (m)")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0.1))
+dromlength<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_length,fill=century_built))+geom_boxplot()+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Century of Tomb Construction", y="Dromos Length (m)")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0.5))
+stomwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,stomion_width,fill=century_built))+geom_boxplot()+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Century of Tomb Construction", y="Stomion Width (m)")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0.1))
+chamarea<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,chamber_area,fill=century_built))+geom_boxplot()+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Century of Tomb Construction", y="Chamber Area (sq m)")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0.5))
+
+ggarrange(dromwidth,dromlength,stomwidth,chamarea,nrow=2,ncol=2)
+
+```
+
+Looking at Figure \@ref(fig:dimensionsboxplots), we see that, through the centuries, the dimensions of the tombs did vary, and sometimes consistently, as with the trend toward narrower *dromoi*.
+The 9th-century tombs, though, which include all but two of the 12 discussed above, do not particularly stand out.
+It does seem that chambers were, on average, larger in this period as well as varying more in size than at any other time.
+Then again, even the largest chambers are not exactly cavernous, and the extra labour expended on construction would have been limited.
+The *dromos* is a plausible area for conspicuous demonstrations of scale but, with their frequently poor preservation, we cannot be confident in assessing this.
+It appears at any rate that they were gradually declining in size, albeit with some notable outliers.
 
@@
 
-Looking at Figure \@ref(fig:tombdimensions), however, it seems that no clear relationship existed between the physical size of a tomb and its buried population at EIA Knossos.
+Furthermore, it seems that no clear relationship existed between the physical size of a tomb and its buried population at EIA Knossos (see Figure \@ref(fig:tombdimensions)).
 It is true that over half of all tombs do not exceed 2.6 sq m in area, while nine of the ten tombs with the most burials do surpass this value.
 But many of the very largest tombs are decidedly middling when it comes to burial numbers.
-Things are even less clear for the *dromos*, the most visible part of the tomb, and a plausible arena for demonstrations of scale and grandeur during the burial itself.
-The very largest two tombs by burial numbers also had exceptionally long *dromoi*, but the next eight tombs with the largest *dromoi* do not rank highly when it comes to interments.
-There are difficulties here in that the *dromoi* especially were common victims of encroachment by later tombs, and our estimates of their original length are tentative in many cases.
-
-```{r dimensionsboxplots, fig.dim=c(6,5),dpi=300, fig.align='center',fig.cap="Boxplots comparing various tomb dimensions as they vary by the century of tomb construction."}
-
-tombs$century_built<-factor(tombs$century_built,levels=c("11th","10th","9th","8th","7th"))
-dromwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_width,fill=century_built))+geom_boxplot()+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Century of Tomb Construction", y="Dromos Width (m)")+
-                  scale_fill_brewer(palette = "Set2")+ 
-                    scale_y_continuous(expand=c(0,0.1))
-dromlength<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_length,fill=century_built))+geom_boxplot()+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Century of Tomb Construction", y="Dromos Length (m)")+
-                  scale_fill_brewer(palette = "Set2")+ 
-                    scale_y_continuous(expand=c(0,0.5))
-stomwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,stomion_width,fill=century_built))+geom_boxplot()+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Century of Tomb Construction", y="Stomion Width (m)")+
-                  scale_fill_brewer(palette = "Set2")+ 
-                    scale_y_continuous(expand=c(0,0.1))
-chamarea<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,chamber_area,fill=century_built))+geom_boxplot()+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Century of Tomb Construction", y=bquote('Chamber Area ('*m^2*')'))+
-                  scale_fill_brewer(palette = "Set2")+ 
-                    scale_y_continuous(expand=c(0,0.5))
-
-ggarrange(dromwidth,dromlength,stomwidth,chamarea,nrow=2,ncol=2)
-
-```
-
-Looking at Figure \@ref(fig:dimensionsboxplots), though, we can see that, through the centuries, the dimensions of the tombs varied, and sometimes consistently, as with the trend toward narrower *dromoi*.
-The 9th-century tombs, including all but one of those seven noted above, do not particularly stand out.
-It does seem that chambers were, on average, larger in this period as well as varying more in size than in any other period.
-Then again, even the largest chambers are not exactly cavernous, and the extra labour expended on the construction of the larger chambers would have been limited.
-The *dromos* is a plausible area for conspicuous demonstrations of scale but, with their frequently poor preservation, we cannot be confident in assessing this.
-It appears at any rate that they were gradually declining in size, albeit with some notable outliers.
-
-So perhaps, at their time of building, such aspects of tomb construction carried social meaning but, even if they did, these Figures suggest this proved irrelevant to their later reuse.
+Things are even less clear for the *dromos*; the largest two tombs by burial numbers also had exceptionally long *dromoi*, but the next eight tombs with the longest *dromoi* do not rank highly on interments.
+
+It seems dubious, then, whether the scale of the tombs when built carried overt social meanings but, even if it did, this clearly proved irrelevant to their later reuse.
 Several tombs had niches cut into their *dromoi*, or the *stomion* extended, to fit in later burial urns.
-Sometimes, urns from within the tomb were placed in such niches to make space for new interments in the chamber itself.
-There is an impression of individuals vying to be included in the chamber itself, of competition was for inclusion into the sepulchral group, supporting the idea that the tombs’ original builders little anticipated their eventual popularity and so did not scale their tombs accordingly.
+Sometimes, urns from within the chamber were relocated to such niches to make space for new interments.
+There is an impression of competition for inclusion into the sepulchral group, supporting the idea that the tombs’ original builders little anticipated their eventual popularity and so did not scale their tombs accordingly.
+
+##Imports, interments, and indices of wealth
+
+The dimensions of the tombs may have shed little light on the early use of what would become the most used tombs in both cemeteries, but we can still investigate the changing scale and composition of the assemblages of these tombs through time.
 
@@
 
-pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A","B")
+pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A",
+       ifelse(pots$tomb=="X"|pots$tomb=="G"|pots$tomb=="104"|pots$tomb=="Q"|pots$tomb=="283","B","C"))
 tombs9th10th<-subset(tombs,century_built=="9th" & total_pottery>50|century_built=="10th" & total_pottery>50)
@@
 
-##Imports, interments, and indices of wealth
-
-Though it may not be immediately apparent from the dimensions of the tombs, however, the late 9th and early 8th centuries are consistently interpreted as periods of artistic development, increased overseas trade, and renewed vigour in the construction of tombs and the visibility of the burial rite.
-The PGB style reflects a uniquely Cretan blend of Attic and Near Eastern forms, while the appearance of Bronze Age 'antiques' in the tombs from this period onwards is felt to reflect renewed, even competitive interest in the Minoan past, as a tool for mortuary display or even lineage affirmation.
+And in Figure \@ref(fig:tombscompthroughtime) we see that the tombs of Group A were relatively unremarkable in the scale of their depositions during the 9th century, only diverging significantly from the rest in the subsequent two hundred years.
+The Group B tombs, meanwhile, had their respective peaks in the 9th or 8th centuries, with a near total decline afterwards.
+As a working hypothesis, let us suppose that these two distinct trajectories correspond to changing socio-politcal and economic contexts, made manifest by the burying groups in their mortuary depositions.
+How far might this hypothesis take us in explicating these patterns, and how visible might it be?
+
+The late 9th and early 8th centuries are consistently interpreted as periods of artistic development, increased overseas trade, and renewed vigour in the construction of tombs and the visibility of the burial rite.
+The PGB style reflects a uniquely Cretan blend of Attic and Near Eastern forms, while the appearance of BA 'antiques' in the tombs from this period onwards is felt to reflect renewed, even competitive interest in the Minoan past, as a tool for mortuary display or even lineage affirmation.
 
@@
 
-imitations<-subset(pots,pots$imitation!="")
-imitation_aorist<-colSums(imitations[27:76])
-imitation_aorist<-data.frame(imitation_aorist)
-colnames(imitation_aorist)<-c("aoristic_sum")
-row.names(imitation_aorist)<-dates
-setDT(imitation_aorist, keep.rownames = "decade")
-imitation_aorist$decade<-as.numeric(imitation_aorist$decade)
-
 ggplot()+geom_col(data=import_aorist,aes(decade,aoristic_sum),alpha=0.2)+
           geom_col(data=allimport_aorist,aes(decade,aoristic_sum, fill=origin),position="dodge")+
-            geom_col(data=imitation_aorist,aes(decade,aoristic_sum))+
             theme_bw()+theme(text=element_text(family="Garamond", size=12))+
@@
 The rise in imported items is clearly seen in the tomb assemblages, with a succession of increased Greek and Near Eastern imports respectively in the late 9th and early 8th centuries (see Figure \@ref(fig:imports)).
-Yet beyond conforming to our prior inferences about the LPG-EG periods, the deposition of these imported wares may also cast light on the original pattern drawn out in the present analysis: the late surge in burial activity among some select 9th and 10th century tombs.
-
-```{r importstombcenturies, fig.dim=c(8,4),dpi=300, fig.align='center',fig.cap="Aoristic sums of imported vessels through time, from tombs built in the 11th through 9th centuries."}
+This much has been relatively well understood for some time, at least at the KNC, but we can further examine how these imports were distributed among the various groups of tombs.
+
+```{r importsgroups, fig.dim=c(8,4),dpi=300, fig.align='center',fig.cap="Aoristic sums of imported vessels through time, from tombs built in the 11th through 9th centuries."}
+
+pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A",
+                   ifelse(pots$tomb=="X"|pots$tomb=="G"|pots$tomb=="104"|pots$tomb=="Q"|pots$tomb=="283","B","C"))
 
 imports<-subset(pots,import!="")
-
-import_aorist<-colSums(imports[27:76])
-import_aorist<-data.frame(import_aorist)
-colnames(import_aorist)<-c("aoristic_sum")
-row.names(import_aorist)<-dates
-setDT(import_aorist, keep.rownames = "decade")
-import_aorist$decade<-as.numeric(import_aorist$decade)
-
-imports_11th<-subset(imports,tomb_century=="11th")
-imports_10th<-subset(imports,tomb_century=="10th")
-imports_9th<-subset(imports,tomb_century=="9th")
-imports_8th<-subset(imports,tomb_century=="8th")
-imports_7th<-subset(imports,tomb_century=="7th")
-
-import_11th_aorist<-colSums(imports_11th[27:76])
-import_11th_aorist<-data.frame(import_11th_aorist)
-colnames(import_11th_aorist)<-c("aoristic_sum")
-row.names(import_11th_aorist)<-dates
-setDT(import_11th_aorist, keep.rownames = "decade")
-import_11th_aorist$decade<-as.numeric(import_11th_aorist$decade)
-
-import_11th_aorist$century<-"11th"
-
-import_10th_aorist<-colSums(imports_10th[27:76])
-import_10th_aorist<-data.frame(import_10th_aorist)
-colnames(import_10th_aorist)<-c("aoristic_sum")
-row.names(import_10th_aorist)<-dates
-setDT(import_10th_aorist, keep.rownames = "decade")
-import_10th_aorist$decade<-as.numeric(import_10th_aorist$decade)
-
-import_10th_aorist$century<-"10th"
-
-import_9th_aorist<-colSums(imports_9th[27:76])
-import_9th_aorist<-data.frame(import_9th_aorist)
-colnames(import_9th_aorist)<-c("aoristic_sum")
-row.names(import_9th_aorist)<-dates
-setDT(import_9th_aorist, keep.rownames = "decade")
-import_9th_aorist$decade<-as.numeric(import_9th_aorist$decade)
-
-import_9th_aorist$century<-"9th"
-
-import_8th_aorist<-colSums(imports_8th[27:76])
-import_8th_aorist<-data.frame(import_8th_aorist)
-colnames(import_8th_aorist)<-c("aoristic_sum")
-row.names(import_8th_aorist)<-dates
-setDT(import_8th_aorist, keep.rownames = "decade")
-import_8th_aorist$decade<-as.numeric(import_8th_aorist$decade)
-
-import_8th_aorist$century<-"8th"
-
-import_7th_aorist<-colSums(imports_7th[27:76])
-import_7th_aorist<-data.frame(import_7th_aorist)
-colnames(import_7th_aorist)<-c("aoristic_sum")
-row.names(import_7th_aorist)<-dates
-setDT(import_7th_aorist, keep.rownames = "decade")
-import_7th_aorist$decade<-as.numeric(import_7th_aorist$decade)
-
-import_7th_aorist$century<-"7th"
-
-all_import_aorist<-rbind(import_11th_aorist,import_10th_aorist,import_9th_aorist,import_8th_aorist,import_7th_aorist)
-all_import_aorist$century<-factor(all_import_aorist$century,levels=c("11th","10th","9th","8th","7th"))
-
-ggplot()+ geom_col(data=subset(all_import_aorist,century!="7th" & century!="8th"),aes(decade,aoristic_sum,fill=century),position="dodge")+
-  scale_x_reverse(limits=c(1000,630))+scale_fill_brewer(palette="Set2")+
+imports_A<-subset(imports,group=="A")
+imports_B<-subset(imports,group=="B")
+imports_C<-subset(imports,group=="C")
+
+import_A_aorist<-colSums(imports_A[27:76])
+import_A_aorist<-data.frame(import_A_aorist)
+colnames(import_A_aorist)<-c("aoristic_sum")
+row.names(import_A_aorist)<-dates
+setDT(import_A_aorist, keep.rownames = "decade")
+import_A_aorist$decade<-as.numeric(import_A_aorist$decade)
+
+import_A_aorist$group<-"A"
+
+import_B_aorist<-colSums(imports_B[27:76])
+import_B_aorist<-data.frame(import_B_aorist)
+colnames(import_B_aorist)<-c("aoristic_sum")
+row.names(import_B_aorist)<-dates
+setDT(import_B_aorist, keep.rownames = "decade")
+import_B_aorist$decade<-as.numeric(import_B_aorist$decade)
+
+import_B_aorist$group<-"B"
+
+import_C_aorist<-colSums(imports_C[27:76])
+import_C_aorist<-data.frame(import_C_aorist)
+colnames(import_C_aorist)<-c("aoristic_sum")
+row.names(import_C_aorist)<-dates
+setDT(import_C_aorist, keep.rownames = "decade")
+import_C_aorist$decade<-as.numeric(import_C_aorist$decade)
+
+import_C_aorist$group<-"C"
+
+aorist_import_groups<-rbind(import_A_aorist,import_B_aorist,import_C_aorist)
+
+ggplot(aorist_import_groups,aes(decade,aoristic_sum,fill=group))+ geom_col(position="dodge")+
   theme_bw()+theme(legend.position="bottom",text=element_text(family="Garamond", size=12))+
   labs(x="Date (Years BC)", y="Aoristic Sum")+
-  guides(fill=guide_legend(title="Century of Tomb Construction"))+
+  guides(fill=guide_legend(title="Group"))+
   scale_fill_brewer(palette = "Set2")+ 
-                  scale_y_continuous(expand=c(0,0),limits=c(0,12))+
+  scale_y_continuous(expand=c(0,0),limits=c(0,10))+
   scale_x_reverse(expand = c(0, 0),limits=c(1000,600))+
@@
 
-In Figure \@ref(fig:importstombcenturies), we see that the tombs already built by the end of the PGB period, particularly those built in the 9th and, to a lesser extent, 10th centuries, were major beneficiaries of the influx of imported items during this time, but that, in the following centuries, in line with the rest of the cemetery, the number of imported items deposited within them declined.
+In Figure \@ref(fig:importstombcenturies), not only is it apparent that a range of tombs were beneficiaries of the increase in imports in the LPG-EG periods, but also that there was potentially significant patterning in where those artefacts ended up.
+The Group A and B tombs closely trace the trends for Greek and Near Eastern imports respectively (and, indeed, tombs Q and G contained only Greek imports), while across the other tombs, there was also no doubt variegation not observable here.
 Imported ceramics are used here as something of a proxy for items of status value, a shorthand that would ideally be bolstered by other artefacts (such as precious vessels, weapons, jewellery etc.) which, owing to their less developed chronologies, must unfortunately be left out of such diachronic analyses for now.
-But, nonetheless, this pattern hints towards a possibility that shall now be explored, that the surge in burial activity seen across both cemeteries in the late 8th and early 7th centuries would appear to be of a different kind to that of a century earlier.
+But let us suggest, for the time being, that these trends are representative of a larger truth, namely that the late 9th and early 8th centuries saw an unmatched degree in the variety of depositions, which went unmatched in later periods.
 
@@
 
-Figure \@ref(fig:bigbox) is a boxplot for the aoristic sums of each decade across both cemeteries, colour-coded by ceramic phase.
+To test such an, admittedly tentative, proposition, Figure \@ref(fig:bigbox) presents a boxplot of the aoristic sums of each decade across both cemeteries, colour-coded by ceramic phase.
 Note that, for visual clarity, the outliers tomb G in the EG period, and tomb P in the late EO-LO period are excluded, with values ranging from 29-34, and 36-44 respectively.
@@
 The Orientalising period, on the other hand, exhibits much reduced IQRs, and a proliferation of outliers far and above the range of the other tombs.
-In other words, while both the PGB-EG and EO-LO periods witnessed an expanding volume and range of burial activity, in the earlier period this was distributed somewhat more evenly across the total burying population.
+In other words, whilst both the PGB-EG and EO-LO periods witnessed an expanding volume and range of burial activity, in the earlier period this was distributed more evenly across the total burying population.
 One speaks of increasing plurality, the other of divergence.
@@
 
-This possibility is supported further by considering the gini coefficient of the distribution of pottery between the tombs.
+This possibility is supported further by considering the gini coefficient of the distribution of pottery between the tombs (see Figure \@ref(fig:gini).
 Using the aoristic sums as a probablistically-weighted estimate of the volume of material deposited in each tomb in each decade, use of the gini coefficient can give an (albeit rough) estimate of how unequally distributed that material was.
-And what we see is that, though values were always relatively high (i.e.
-unequal) within the cemeteries, the major spike in the gini coefficient seen in the Orientalising period contrasts markedly with the modest rise and plateau of the PGB-EG.
+And, though values were always relatively high (i.e.
+unequal) within the cemeteries, the major spike in the gini coefficient seen in the Orientalising period does contrast markedly with the modest rise and plateau of the PGB-EG.
 The gini coefficient is most commonly used in the assessment of income inequality, but I want to stress here that it is not "wealth" in any meaningful sense that we are talking about here.
@@
                text=element_text(family="Garamond", size=12))+
-                 labs(title="7 Tombs",x="", y="")+
+                 labs(title="Group A",x="", y="")+
                   scale_fill_brewer(palette = "Set2",labels=c(" Amphora  "," Jug  "," Oil vessel  ", " Pithos"," Drinking vessel  "," Lid  "," Other"))+ 
@@
 
-Figure \@ref(fig:comps) shows the composition of the total pottery assemblage from each cemetery, and from the seven tombs drawn out earlier, through the centuries.
+Figure \@ref(fig:comps) shows the composition of the total pottery assemblage from each cemetery, and of the Group A tombs, through the centuries.
 In all cases, there appears to be the steady crystallisation of a basic 'burial kit', comprising a pithos and lid, an oil vessel, a jug, and a drinking vessel.
@@
 ```{r hists, fig.dim=c(6,4),dpi=300, fig.align='center',fig.cap="Histograms relating, first, to counts of non-ceramic artefacts, metal objects, and imported items across all tombs in both cemeteries and, secondly, to those counts divided by the estimated number of burials in each tomb."}
-tombs$group<-factor(tombs$group,levels=c("A","B"))
+tombs$group<-factor(tombs$group,levels=c("A","B",))
 total_object<-ggplot(subset(tombs,est_burials>1 & total_object>0),aes(total_object, fill=group))+geom_histogram()+
@@
 The trend instead must reflect choices made by the burying groups themselves, which Kotsonas feels may have comprised various elements, including perhaps age- and sex-grades.
-But it seems probable that such corporate groups, if they existed intersected with, and existed alongside, real or fictive kinship networks.
+But it seems probable that such corporate groups, if they existed, intersected with, and operated alongside, real or fictive kinship networks.
 
@@
 Elsewhere on Crete, changes around this time include a shift to unmarked, individual burials, and more austere grave goods, as well as an apparent reorientation of sanctuary deposits away from extra-mural to suburban spaces, and a rise in utilitarian, mass-produced votives (Perlman & Gagarin 2016, 34-6).
-At Azoria, major urban remodelling in the 7th century saw the earlier, agglutinative house forms of neighbouring Kavousi Vronda and Kastro give way to larger, more formally planned dwellings with little evidence of renovation or remodelling, while larger, apprently civic, buildings were constructed to accommodate major public gatherings.
 At Knossos, a small number of pithos burials in pits do appear in the KNC and Fortetsa volumes (Brock 1957, 98; Coldstream & Catling 1996, 162), and recent re-excavation of the Teke plot identified numerous small pit burials around three chamber tombs (Evely 2012), raising the possibility that many more such burials filled the interstices between larger chamber tombs elsewhere, but have gone unrecognised.
 The insecure dates of these pits, however, makes it unclear whether they represent a genuine Archaic trend.
+At Azoria, major urban remodelling in the 7th century saw the earlier, agglutinative house forms of neighbouring Kavousi Vronda and Kastro give way to larger, more formally planned dwellings with little evidence of renovation or remodelling, while larger, apprently civic, buildings were constructed to accommodate major public gatherings.
 

### 2019-05-24 R/Samples.R

@@ -1,0 +1,381 @@
+```{r aorist_uncertainty, fig.dim = c(7, 4), dpi=300}
+
+aorist_pot_values<-pots[25:74]
+aorist_pot_matrix<-data.matrix(aorist_pot_values)
+aorist_pot_matrix[aorist_pot_matrix == 0] <- NA
+aorist_pot_medians<-apply(aorist_pot_matrix,2,median,na.rm=TRUE)
+
+aorist_pot_medians<-data.frame(aorist_pot_medians)
+colnames(aorist_pot_medians)<-c("aoristic_median")
+row.names(aorist_pot_medians)<-dates
+aorist_pot_medians<-data.frame(aorist_pot_medians)
+setDT(aorist_pot_medians, keep.rownames = "decade")
+aorist_pot_medians$decade<-as.numeric(aorist_pot_medians$decade)
+
+
+ggplot(NULL)+geom_col(data=aorist_pot_medians,aes(decade,aoristic_median))+
+  xlim(1100,630)+ylim(0,0.3)+theme_minimal()+
+  theme(legend.title=element_blank(),
+        text=element_text(family="Garamond", size=12))+
+  labs(x="Date (Years BC)", y="Aoristic Median")+
+  scale_fill_brewer(palette = "Set2")
+
+```
+```{r aoristmedianadjust}
+
+aorist_pot_medians$aoristic_sums<-aorist_all_tombs$aoristic_sum
+aorist_pot_medians$aorist_median_adjust<-aorist_pot_medians$aoristic_sums/aorist_pot_medians$aoristic_median
+
+
+ggplot(NULL)+geom_col(data=aorist_pot_medians,aes(decade,aorist_median_adjust))+
+  xlim(1100,630)+theme_minimal()+
+  theme(legend.title=element_blank(),
+        text=element_text(family="Garamond", size=12))+
+  labs(x="Date (Years BC)", y="Aoristic Median")+
+  scale_fill_brewer(palette = "Set2")
+
+```
+
+```{r boxplot,  fig.dim=c(9,5), dpi=300,}
+
+ggplot(tombs,aes(total_finds))+geom_histogram()+xlim(0,200)
+
+aorist_tomb_values<-tombs[c(1,158:207)]
+aorist_tomb_values[aorist_tomb_values == 0] <- NA
+
+aorist_tomb_values<-data.frame(aorist_tomb_values)
+colnames(aorist_tomb_values)[2:51]<-dates
+
[…]

### 2019-05-24 R/Samples2.R

@@ -1,0 +1,71 @@
+tombs_ranked<-tombs[with(tombs, order(-tombs$total_pottery)), ]
+tombs_top_20<-head(tombs_ranked,20)
+top20<-as.vector(tombs_top_20$tomb)
+pots_top_20<-pots[is.element(pots$tomb, top20),]
+pots_top_20$tomb<-factor(pots_top_20$tomb,levels=top20)
+
+cols<-c(1,158:207)
+aorist_all_tomb_names<-tombs_top_20[,cols]
+aorist_all_tomb_names<-data.frame(aorist_all_tomb_names)
+colnames(aorist_all_tomb_names)<-c("tomb",dates)
+
+aorist_all_tomb_melt<-melt(aorist_all_tomb_names)
+aorist_all_tomb_melt$variable<-as.numeric(as.character(aorist_all_tomb_melt$variable))
+
+
+top_20<-tombs_top_20$tomb
+aorist_all_tomb_melt$tomb<-factor(aorist_all_tomb_melt$tomb,levels=top_20)
+
+
+ggplot(aorist_all_tomb_melt,aes(variable, value))+geom_col()+theme(legend.position = "none")+facet_wrap(~tomb,scales="free_y")+
+  geom_col()+theme_bw()+theme(legend.title=element_blank(),
+                              text=element_text(family="Garamond", size=20))+
+  labs(x="Date (Years BC)", y="Aoristic Sums")+
+  scale_fill_brewer(palette = "Set3")+
+  theme(legend.position="none")+
+  coord_cartesian(xlim=c(1100,610))+scale_y_continuous()+
+  scale_x_reverse(expand = c(0, 0),limits=c(1100,610))+
+  theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+View(tombs_9th_aorist)
+tombs_9th_aorist<-rowSums(tombs_9th[,178:187])
+tombs_9th_aorist<-data.frame(tombs_9th_aorist)
+row.names(tombs_9th_aorist)<-tombs_9th$tomb
+tombs_9th$aorist<-tombs_9th_aorist$tombs_9th_aorist
+
+ggplot(subset(tombs_9th,tombs_9th$est_burials>1),aes(chamber_area,aorist,color=cem_condensed))+geom_point()+
+  theme_bw()+theme(legend.title=element_blank(),text=element_text(family="Garamond", size=12))+
[…]

### 2019-05-30 analysis/paper/paper2.Rmd
Final writing and first edits

@@
       reference_docx: "../templates/articletemplate.docx" # Insert path for the DOCX file
-bibliography: Lib.bib
+bibliography: references.bib
 csl: "../templates/world-archaeology.csl" # Insert path for the bib-style
 
-abstract: |
-  The Early Iron Age cemeteries of Knossos, Crete, represent the richest source of archaeological material at the site between the end of the Bronze Age and the emergence of the Archaic *polis*.
-However, despite well-published excavation reports on the two largest burial grounds and a number of studies dealing with various artefacts and individual tombs therein, no substantial comparative, quantitative analysis has yet been undertaken.
-This paper does just that, and in so doing identifies, and attempts to explain, a pattern of increased burial rates among a few select tombs.
-This paper demonstrates the potential for quantitative analyses to contribute towards nuanced, context-specific theories of social practice, and advocates their wider application to the archaeological record of the Cretan Early Iron Age.
-It also serves as a demonstration of maximal research reproducibility, with the full datasets and code underpinning the analysis made freely available to download and use.
-keywords: |
+abstract: 
+   "The Early Iron Age cemeteries of Knossos, Crete, represent the richest source of archaeological material at the site for the whole period between the end of the Bronze Age and the emergence of the Archaic *polis*.
+However, despite well-published excavation reports on the two largest burial grounds, the Fortetsa and Knossos North cemeteries, and a number of studies dealing with various artefact classes and individual tombs therein, no substantial comparative, quantitative analysis of these two cemeteries has yet been undertaken.
+This paper provides such an analysis, examining both the synchronic and diachronic variation among the tombs in these two cemeteries.
+Through the application of a range of quantitative techniques, this study isolates a pattern of increased rates of burial among a few select tombs, and goes on to consider the broader social contexts within which these tombs were built and used, as well as discussing the possible nature of the burying groups.
+This paper demonstrates the potential for quantitative analyses to contribute to nuanced, context-specific theories of social practice, and advocates their wider application to the archaeological record of the Cretan Early Iron Age.
+It also serves as a demonstration of maximal research reproducibility, with the full datasets and code underpinning the analysis made freely available to download and use"
+
+---
+# Keywords: 
 Early Iron Age; Knossos; Burial; Quantitative; Open-Access; Reproducibility
-
----
 
@@
   fig.path = "../figures/",
-  dev="jpg"
+  dev="png"
 )
@@
 
+##### pagebreak
+
 # Introduction
@@
 For one, though, this is not the Knossos of that illustrous, palatial period, but rather the town as it persisted and evolved during the subsequent Early Iron Age (EIA).
-Research into this once maligned period has grown apace in the last few decades, beginning with the pioneering works of Coldstream [-@Coldstream1977], Desborough [-@Desborough1972], and Snodgrass [-@Snodgrass1971] and culminating in the rejection by many scholars of the image of a society bereft of its former palatial glory.
-Certainly, few would any longer defend the appelation or interpretive bagguage of the ‘Dark Ages’ [on the origins of which, see @Kotsonas2016; @Morris1997].
+Research into this once maligned period has grown apace in the last few decades, beginning with the pioneering works of Coldstream [-@Coldstream1977], Desborough [-@Desborough1972], and Snodgrass [-@Snodgrass1971a] and culminating in the rejection by most scholars of the appelation and interpretive bagguage of the so-called ‘Dark Ages’ [on the origins of which, see @Kotsonas2016; @Morris1997].
 Nonetheless, our knowledge of many EIA settlements remains often underdeveloped in comparison with those of the second millennium BC.
 
-But more importantly, and welcome though they are, a number of recent publications on the socio-economic [@Wallace2010a] and settlement [@Nowicki2000] changes which accompanied the BA-EIA transition have directed scholarly attention toward a number of elevated, defensible settlements supposedly founded in response to the dangers of this turbulent period.
-This is reflected too in survey work, which has focussed extensively around the Bay of Mirabello, and the Isthmus of Ierapetra, where such sites are numerous.
-This has, unfortunately, engendered a certain disregard for settlement characterised by quite different topographies, especially in the central region of the island.
-There remains, therefore, a need for studies addressing the social dynamics at such sites (Knossos, Phaistos, Grivila, Eleutherna etc.), with a long-term view to comparative appraisal of the divergent trajectories of communities across the island in the EIA [@Kotsonas2011b; @Whitley2011, 667–8].
-
-As to the second and third of my stated interests, several authors have recently demonstrated the great potential of quantitative, statistical, spatial, and network analyses for elucidating social dynamics, settlement patterning, and systems of inter-regional communication across Crete and the wider Aegean.
-Such work encourages the use of large (often pre-existing) datasets, and the sharing of both data and methods to facilitate reproducible, collaborative, and innovative research [@Bevan2013; @Knappett2011a].
-The cemeteries of Knossos offer an opportunity to introduce such approaches into EIA research, where they have yet to make much of an impact [see @Kotsonas2011].
-The Fortesta and North Cemetery complexes are both well-published [@Brock1957; @Coldstream1996a] and have been considered variously in relation to their 'oriental' imports [@Antoniadis2012], antique artefacts [@Crowe2016], and religious context [@Coldstream1984b].
-But a lack of quantitative analyses, or systematic comparisons between different cemeteries, has left us with many intuited trends, some clearly exceptional tombs, but a general impression of un-patterned multiplicity.
-
-I suggest that the application of a range of quantitative and statistical methods may provide new insight into the social behaviours underlying the mortuary assemblages.
-With a dataset including information on every find and tomb recorded in the Fortetsa and Knossos North Cemetery publications [@Brock1957; @Coldstream1996a], the following analysis examines both the synchronic and diachronic variation among the tombs of both cemeteries.
+Secondly, and welcome though they are, a number of recent publications on the socio-economic [@Wallace2010a] and settlement [@Nowicki2000] changes which accompanied the BA-EIA transition have directed scholarly attention toward a number of elevated, defensible settlements supposedly founded in response to the dangers of this turbulent period.
+This is reflected in survey work, which has focussed extensively around the Bay of Mirabello, where such sites are numerous [@Haggis2005; @Hayden2005; @Watrous2012].
+This has, unfortunately, engendered a certain disregard for settlement characterised by different topographies, especially in the centre of the island.
+There remains, therefore, a need for studies addressing the social dynamics at such sites (Knossos, Phaistos, Grivila, Eleutherna etc.), with a long-term view to comparative appraisal of the divergent trajectories of communities across Crete in the EIA [@Kotsonas2011b; @Whitley2011, 667–8].
+
+As to the second and third of my stated interests, several authors have recently demonstrated the great potential of quantitative, statistical, spatial, and network analyses to elucidate social dynamics, settlement patterning, and systems of inter-regional communication across Crete and the wider Aegean.
+Such work encourages the use of large (often pre-existing) datasets, and the sharing of both data and methods to facilitate reproducible, collaborative research [@Bevan2013; @Knappett2011a; @Knappett2011].
+The cemeteries of Knossos offer an opportunity to introduce such approaches into EIA research on Crete, where they have yet to make much of an impact [see @Kotsonas2011].
+The Fortesta and North Cemetery complexes are both well-published [@Brock1957; @Coldstream1996a] and have been studied with regard to their 'oriental' imports [@Antoniadis2012], antique artefacts [@Crowe2016], and religious context [@Coldstream1984b].
+But a lack of quantitative analyses, or systematic comparisons between the cemeteries, has left us with many intuited trends, some exceptional tombs, but a general impression of un-patterned multiplicity.
+
+With a dataset including information on every find and tomb recorded in the Fortetsa and Knossos North Cemetery publications [@Brock1957; @Coldstream1996a], the following analysis seeks to address this present lack.
 In doing so, I aim to demonstrate the potential for quantitative methods, firstly, to permit identification, and nuanced characterisation, of significant patterning in the archaeological material; secondly, to facilitate a more systematic appraisal of the cemeteries’ largest tombs and their distinctive histories; and, finally, to offer not just generalities, but tangible and context-specific considerations of the motivations and identities behind the burial groups themselves, with a view to eventually comparing such results with mortuary assemblages across the island.
@@
 Speaking of the town’s stratigraphic sequence, Coldstream lamented that “the remains of [this] period are like a meagre filling in a very thick sandwich” between Bronze Age and later Greek and Roman levels [@Coldstream1991, 287].
-Most individual finds of Protogeometric or Geometric date take the form of flimsy foundations, patches of earth floor, wells or simple scatters of pottery, and for a long time it was believed the town contracted, moved, or even dissolved following the collapse of the palace.
-Indeed, Alexiou [@1950] thought it likely that the area comprised several dispersed villages, which later coalesced into a single city, as per Aristotle’s model of synoecism.
+Most individual finds of Protogeometric or Geometric date take the form of flimsy foundations, patches of earth floor, wells or simple scatters of pottery, and for a time it was argued that the area in the EIA comprised several dispersed villages, which later coalesced into a single city, as per Aristotle’s model of synoecism [@1950].
 
@@
 ## The Early Iron Age Cemeteries
-```{r map, fig.cap="Map of the Knossos area showing the locations of the various tomb clusters and cemeteries, with EIA pottery distributions from the KULP marked in orange"}
+
+```{r map, fig.cap="goes around here"}
 knitr::include_graphics("../figures/Knossos Tombs Map.jpg")
@@
 Survey aside, our most abundant evidence for EIA Knossos comes from the mortuary record (see Figure \@ref(fig:map).
-The transitional period between the Late Bronze Age (LBA) and EIA yields few if any securely datable interments, but later evidence is more plentiful, with multiple burials known from the Kephala ridge [@Coldstream1963, 38; @Hogarth1899, 82-5], the modern suburbs of Ayios Ioannis and Atsalenio [@Boardman1960; @Davaras1968, 133–46], the cave of Mavro Spelio [@Antoniadis2012, 58-9], and the Khaniale Teke site with its well-known tholos [@Boardman1967;@Hutchinson1954; cf.
+The transitional period between the Late Bronze Age (LBA) and EIA yields few if any securely datable interments, but later evidence is more plentiful, with multiple burials known from the Kephala ridge [@Coldstream1963, 38; @Hogarth1899, 82-5], the modern suburbs of Ayios Ioannis and Atsalenio [@Boardman1960a; @Davaras1968, 133–46], the cave of Mavro Spelio [@Antoniadis2012, 58-9], and the Khaniale Teke site with its well-known tholos [@Boardman1967;@Hutchinson1954; cf.
 @Hoffman1997 @Kotsonas2006].
@@
 
-## The Fortetsa and the Knossos North Cemeteries
-
-The cemetery on the slope facing the eponymous village of Fortetsa is composed of twelve tombs excavated in 1933, and another eight uncovered in 1935 (Brock 1957, xi).
+## The Fortetsa and Knossos North Cemeteries
+
+The cemetery on the slope facing the eponymous village of Fortetsa is composed of twelve tombs excavated in 1933, and another eight uncovered in 1935 (Brock 1957: xi).
 These tombs cluster in three main groups.
@@
 To the north lies a cluster of five tombs and, to the south, another of three.
-Finally, three tombs excavated in 1933 – L, TFT and Π – which lie close to the main road north of the Acropolis hill [@Brock1957, 1-2], are now considered to belong the Knossos North Cemetery (KNC).
+Finally, three tombs excavated in 1933 – L, TFT and Π – which lie close to the main road north of the Acropolis hill [@Brock1957: 1-2], are now considered to belong the Knossos North Cemetery (KNC).
 
@@
 Northwest of these, and just south of the Teke Tholos, the Teke tombs represent the northern reaches of the KNC.
-The central section of the KNC comprises the ‘Medical Faculty’ site, named for the extension of the University of Crete that prompted major rescue excavations here in 1978.
-Some 89 tombs of Subminoan to Late Orientalising date were found, by far the largest single concentration in the whole Knossos area [@Coldstream1996a, 53-55].
+The central section of the KNC comprises the ‘Medical Faculty’ site, named for the extension of the University of Crete that prompted major rescue excavations here in 1978 [@Coldstream1996a: 53-55].
 
@@
 It comprises a descending ramp or passage, the *dromos*, leading to the chamber itself, cut into the *kouskouras* bedrock.
-The *dromos* and chamber are separated by the *stomion*, an opening often set a step or two lower than the *dromos*, which is covered by a large stone slab, smaller stacked stones, or combination of both [@Antoniadis2012, 47-8].
-The majority of burials are cremations, a rite that appears at Knossos in the Subminoan period, in the Tomb 200+ complex (Ts.
-200, 201 and 202 are three chambers of a single pit tomb) in the KNC [@Coldstream1994, 109].
-Inhumation does not completely disappear in the EIA, but by the Orientalising period it appears restricted to child burials, and in those instances where it is attested (Ts.
-L, 45, 98, 112 and F67:5), these burials appear to be among the very earliest in each tomb [@Antoniadis2012, 69].
-
-## Chronological developments in the cemeteries
+The *dromos* and chamber are separated by the *stomion*, an opening often set a step or two lower than the *dromos*, which is covered by a large stone slab, smaller stacked stones, or combination of both [@Antoniadis2012: 47-8].
+The majority of burials are cremations, a rite that appears at Knossos in the Subminoan period, in the Tomb 200+ complex in the KNC [@Coldstream1994: 109].
+Inhumation does not completely disappear in the EIA, but by the Orientalising period it appears restricted to child burials and [@Antoniadis2012: 69].
+
+## Chronological Developments in the Cemeteries
 
 The initial foundation of the cemeteries has received much attention.
-It has been claimed that no tomb crosses the Late Minoan (LM) IIIC-SM divide [@Coldstream1998, 58; @Coldstream2006, 582), though there is a possibility that this is a classificatory artefact [@Hallager2010].
-Instances of BA tombs reused the EIA are known (e.g.
+It has been claimed that no tomb crosses the Late Minoan (LM) IIIC-SM divide [@Coldstream1998, 58; @Coldstream2006, 582], though there is a possibility that this is a classificatory artefact [@Hallager2010].
+Instances of BA tombs reused in the EIA are known (e.g.
 the Khaniale Teke tholos), and Cavanagh [-@Cavanagh1996], based on a cluster analysis of tomb dimensions, could not reject the possibility of some KNC tombs being of BA date, though other evidence would suggest this is unlikely (Whitelaw, pers.
@@
 During the use of the cemeteries, correspondences have been highlighted between changing pottery styles and episodes of tomb construction and abandonment.
-The Protogeometric B (PGB) style, first defined by Brock [-@Brock1957], 143], who hailed it as “the most remarkable phase in Cretan vase-painting”, is apparently a Cretan anomaly, and evinces a melding of Attic geometric motifs, experimental patterns possibly deriving from Near Eastern metalwork, and iconography drawn seemingly from the Minoan repertoire (Coldstream 1984, 93–4).
-Coldstream has argued that the concurrence of this style with a rash of newly founded tombs points to a reformulation of elite mortuary display, noting how, in this period also, redeposited Minoan larnakes began appearing in and around some tombs, in one case even inspiring the decoration of a PGB vessel (in T.
-107; see @Coldstream1984b; -@Coldstream1994, 112–3).
-Whether simply nostalgia [@Coldstream1998, 60], or a more tactical strategy by newly emerging elite groups [@Coldstream1994, 114-5], it seems that the Bronze Age past possessed a significant cultural cachet mobilised by various actors over the lives of these cemeteries.
+The Protogeometric B (PGB) style, first defined by Brock [-@Brock1957, 143], who hailed it as “the most remarkable phase in Cretan vase-painting”, is apparently a Cretan anomaly, evincing a melding of Attic geometric motifs, patterns possibly deriving from Near Eastern metalwork, and iconography drawn from the Minoan repertoire [@Coldstream1984b, 93-94].
+Coldstream has argued that the concurrence of this style with a rash of newly founded tombs points to a reformulation of elite mortuary display, noting how, in this period also, redeposited Minoan larnakes began appearing in and around some tombs, in one case even inspiring the decoration of a PGB vessel [in T.
+107; see @Coldstream1984b; -@Coldstream1994, 112–13].
+Whether simply nostalgia [@Coldstream1998, 60], or a more tactical strategy by newly emerging elite groups [@Coldstream1994, 114-15], it seems that the Bronze Age past possessed a significant cultural cachet mobilised by various EIA burying groups.
 
@@
 All contained imported items, were of ‘warrior grave’ type, and formed loci for later burials.
-Catling argues that those individuals who dared to travel abroad in this time of more limited seafaring, accruing precious artefacts and stories of distant lands, would have been accorded exceptional treatment in death [@Catling1995, 127–8].
+Catling argues that those individuals who dared to travel abroad in this time of more limited seafaring, accruing exotica and stories from distant lands, would have been accorded exceptional treatment in death [@Catling1995, 127–8].
 
 Complementing the theme of heroes abroad has been a similar interest in foreigners at home.
-Two caches of gold jewellery, found in pits flanking the entrance to the Teke Tholos, were argued by Boardman [-@Boardman1967] to be foundation deposits – a Near Eastern custom – for the burial place of an immigrant Phoenician craftsman.
-Kotsonas has revisited this attribution, suggesting that the tholos is more likely that of an elite individual or family with a monopoly over a metal workshop’s output [-@Kotsonas2006, 155–9].
+Two caches of gold jewellery, found in pits flanking the entrance to the Teke Tholos, were argued by Boardman [-@Boardman1967] to be foundation deposits – a Near Eastern custom – for the tomb of an immigrant Phoenician craftsman.
+Kotsonas has revisited this attribution, deeming the tholos more probably that of an elite individual or family with a monopoly over a metal workshop’s output [-@Kotsonas2006, 155–9].
 While certainty in such individual cases may be moot, it seems unlikely that foreign traders or craftsmen never visited or resided at Knossos during the EIA [@Hoffman1997, 176–85; @Schreiber2003, 293-306].
@@
 Cluster analyses have been employed by both Cavanagh [-@Coldstream1996, 653-7] and Antoniadis [-@Antoniadis2012, 193-7], the most sophisticated statistical techniques so far applied to the material.
-Cavanagh’s analysis of tomb architecture has been mentioned, while Antoniadis’ more holistic approach corroborates quantitatively some of the intuitive correspondences noted between specific tombs in previous studies.
-Antoniadis [-@Antoniadis2012, 172-6] also divided the EIA tombs of Knossos into three groups, those with fewer than 10 pots, fewer than 50, and more than 50, and found that, in each successive case, the proportions of tombs containing 'oriental' imports and imitations increased, the implication being that a hierarchy existed, with unequal access to such prestigious goods.
+Antoniadis [-@Antoniadis2012, 172-6] also divided the EIA tombs of Knossos into three groups, those with fewer than 10 pots, fewer than 50, and more than 50, and found that, in each successive case, the proportions of tombs containing 'oriental' imports and imitations increased, suggesting unequal access to such prestigious goods.
 Yet even here the available data could be further interrogated, such as with the number of imported items *per burial*, something Antoniadis does not explore.
-The present paper seeks to extend and nuance a number of observations that have been made in previous studies, as well as revealing patterns thus far unacknowledged, through use of a more systematic approach to quantitative analysis.
+The present paper thus seeks to extend and nuance a number of observations that have been made in previous studies, as well as revealing patterns thus far unacknowledged, through use of a more systematic approach to quantitative analysis.
 
@@
 The cornerstone of the present study is a database (or, rather, three linked databases) based on the excavation reports for the Fortetsa and Knossos North cemeteries [@Brock1957; @Coldstream1996a].
-The first includes an entry for each individual ceramic vessel, by far the largest portion of all known finds.
+The first includes an entry for each individual ceramic vessel.
 Each entry includes, among other things, the tomb, the vessel shape, its surviving dimensions, its ceramic period and corresponding absolute dating measures.
-In addition, all imports, Oriental imitations, and antique (that is, Bronze Age) artefacts were accordingly categorised, using the studies of Antoniadis (2012), Jones (2000), and Crowe (2016).
+In addition, all imports, 'oriental' imitations, and antique (that is, Bronze Age) artefacts were accordingly categorised, using the studies of Antoniadis (2012), Jones [-@Jones2000], and Crowe (2016).
 In total, this database comprises just over 4540 individual artefacts.
@@
 
-## Reproducibility and data-sharing
-
-An ever-growing number of archaeologists have begun to advocate the utility, indeed the necessity, of making available not just the data underpinning our academic output, but the full methodologies by which our analyses are conducted, in formats facilitating reproducibilty and further investigation.
-It is felt that, with the opportunities afforded by current technologies (digital databases, statistical software packages, online repositories etc.), there remain ever fewer obstacles - and excuses - preventing the full and free sharing of both data and workflows.
+## Reproducibility and Data-sharing
+
+Evermore archaeologists have begun to advocate the utility, indeed the necessity, of making available not just the data underpinning our academic output, but the full methodologies by which our analyses are conducted, in formats facilitating reproducibilty.
+It is felt that, with the opportunities afforded by current technologies (digital databases, statistical software packages, online repositories etc.), there remain ever fewer obstacles to the full and free sharing of both data and workflows [see @article].
 In Marwick's words [-@Marwick2017, 445], "[t]he technical problems are largely solved; the challenge now is to change the norms of the discipline to make high reproducibility a canonical attribute of high-quality scholarly work".
 
-This is not the place for a detailed investigation into the lack of reproducible research in archaeology.
-Bu this article is, among other things, an attempt to bring these principles of maximal reproducibility to the archaeology of EIA Crete.
+This article is, among other things, an attempt to bring these principles of maximal reproducibility to the archaeology of EIA Crete.
 Within this field, catalogues and excavation reports abound and, though many of these may be forgiven based on their dates of publication, the lack of readily utilisable digital data from a number of recent articles and monographs, indeed, the failure to even present data such that they could be manually digitsed, is less pardonable.
@@
 There are, of course, complications to the sharing of data and methods.
-There are real and valid reservations around issues of copyright, the protection of sites from looting, the possibility of being scooped with one's own data, or simply the significant investment of time demanded by the preparation of open-access research.
+There are valid reservations around issues of copyright, the protection of sites from looting, the possibility of being scooped with one's own data, or simply the significant time investment of preparing open-access research.
 But I would argue that the potential gains outweigh these concerns, and that research on the EIA would benefit greatly from the greater accessability of datasets and workflows.
 
-In this vein, the present article has been composed making use of Ben Marwick's "rrtools" [see @Marwick2017], an open-source package for the statistical software R Studio, which provides the author with the tools to create a self-contained "compendium", which can be freely downloaded, and from which anyone can accurately reproduce the entire workflow of the present analysis.
+In this vein, the present article has been composed making use of Ben Marwick's "rrtools" [see @Marwick2017; @inbook], an open-source package for the statistical software R Studio, which provides the author with the tools to create a self-contained "compendium", which can be freely downloaded, and from which anyone can accurately reproduce the entire workflow of the present analysis.
 Though not visible in the present paper, every chart and table herein was produced directly, through written code, from the raw data using R.
 This code, as well as the raw and derived datasets, complete bibliography (in BibTex format, for export to any major reference manager), figures, and article as originally written are included in the compendium.
-It can be downloaded as a repository from GitHub, at the following address: https://github.com/DCPollard94/knossoscemeteries, and the raw datasets have been stored with the online repository Figshare, here: (NB a URL will be included pending acceptane of the article).
-The data are made available through an MIT license, which permits unlimited use, ammendation, and dissemination thereof, yet frees the present author from any responsibilty for that use, or for any perceived errors or inaccuracies in the data.
+It can be downloaded as a repository from GitHub, at the following address: https://github.com/DCPollard94/knossoscemeteries, and the raw datasets have been stored with the online repository Figshare, here: (**NB a URL will be included pending acceptance of the article**).
+The data are made available through an MIT license, which permits unlimited use, ammendation, and dissemination thereof, yet frees the present author from any responsibilty for that use, or perceived errors or inaccuracies in the data.
 It is sincerely hoped that others may further explore and expand upon the research presented here by downloading this compendium.
 
-## Estimating interments
+## Estimating Interments
 
@@
 The method chosen broadly follows that of Cavanagh [-@Cavanagh1996, 659-660], who took the number of pithoi (the most common urn type) and the number of pithoi, amphorae and kraters (which were sometimes used as urns) as middle and upper estimates of the number of interments, with all securely identified interments (in the form of ashes or osteological material) being the lower limit (Figure \@ref(fig:burialestimate)).
-In the present study, the mean of these three estimates was taken, to provide a single aggregate measure for the purposes of analysis (where no interments were known, this variable was still assigned a value of "1", for all tombs must have contained at least one burial).
-
-```{r burialestimate, fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="Relationship between known interments, counts of pithoi, and counts of pithoi, amphorae and kraters.
-The straight line represents the mean estimate based on these three values."}
+In the present study, the mean of these three estimates was taken, to provide a single aggregate measure for the purposes of analysis.
+
+```{r burialestimate, fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="goes around here"}
 
@@
 
-##Aoristic approaches to dating
+##Aoristic Approaches to Dating
 
@@
 A vessel dated to the PGB period could, on our best estimates, date anywhere between 840 and 810 BC.
-This situation results in major challenges to the visualisation and analysis of such data through means which normally rely on value- and not range-based data inputs.
+This situation results in major challenges to the visualisation and analysis of such data through means which normally require values and not ranges.
 A simple, if coarse, solution is to use the midpoint of the range assigned to each vessel (i.e.
@@
 
-```{r allpithoi,fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="Counts of pithoi across both cemeteries through time.
-Vessel dates have here been based on the midpoint of the absolute date range accompanying their respective ceramic periods."}
+```{r allpithoi,fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="goes around here"}
 
@@
 So, our PGB vessel, datable between 840 and 810 BC, would register a value of 0.33 for each of those three decades.
-These values for individual vessels can can then simply be summed to produce tomb- or cemetery-level signatures.
+These values for individual vessels can then simply be summed to produce tomb- or cemetery-level signatures.
 The major contribution of this method is that it incorporates temporal uncertainty into the size of the resulting values, that is, "events with tight temporal definition contribute more to the total probability over their range than do loosely defined events" [@Johnson2004a, 450].
 
-```{r aoristallcems, fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="Aoristic sums of all vessels across both cemeteries.
-This represents a probabilistically weighted visualisation of depositional activity through time."}
+```{r aoristallcems, fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="goes around here"}
 
@@
 I then examine these more closely, in an attempt to understand what sets them apart from the rest.
-Tomb dimensions, rates of burials, and changes in the composition of assemblages through time are all considered, to produce an account which speaks to the changing social landscape of EIA Knossos, drawing together issues of group identity, external relations, and the communication of wealth and status through funerary practice.
-
-##Temporal developments in the Knossian Cemeteries
+Tomb dimensions, rates of burial, and changes in the composition of assemblages through time are all considered, to produce an account which speaks to the changing social landscape of EIA Knossos, drawing together issues of group identity, external relations, and the communication of wealth and status through funerary practice.
+
+##Temporal Developments in the Knossian Cemeteries
 
@@
 
-```{r aoristbycem, fig.cap="Aorist sums of pithoi (left) as a proxy for cremation burials, and all vessels (right).
-The sums are colour coded to show the values of the two cemeteries, and the total values.", fig.dim=c(8,4)}
+```{r aoristbycem, dpi=300, fig.align="center", fig.cap="goesaround here.", fig.dim=c(8,4)}
 KNC<-subset(tombs,tombs$cem_condensed=="KNC")
@@
 Figure \@ref(fig:aoristbycem) presents the aoristic sums of pithoi (as a proxy for burials) and of all vessels, respectively.
-Despite differences in scale and timing, it is immediately apparent that the Fortetsa and Knossos North Cemeteries experieced similar chronological developments in the volume of burials and, to a slightly lesser degree, grave goods.
+Despite differences in scale and timing, it is immediately apparent that Fortetsa and the KNC experieced similar chronological developments in the volume of burials and, to a slightly lesser degree, grave goods.
 In both cemeteries, the rate of cremation burial began rising around the end of the 9th century, plateaued slightly in the early 8th, and then rose again to a peak in the late 8th and early 7th centuries.
-Across all vessels, the developments are more volatile, but again there was growth beginning in the later 10th century - reaching an especially marked peak in the KNC by the end of the 9th - then a slight decline and levelling into the 8th century, before a final flourit of depositional activity.
+Across all vessels, the developments are more volatile, but largely similar, especially regarding the late spike in burial activity.
 We might assume that these developments would be mirrored in the construction of new tombs, but this proves not to be the case.
 
-```{r tombconstruction, fig.dim=c(7.5,4), dpi=300, fig.cap="Tomb construction through time at both cemeteries.
-On the left are counts based on the earliest dated pottery in each tomb, while on the left are aoristic sums likewise based on the earliest ceramic-based assignations.
-In both cases, tombs lacking any securely dateable pottery are excluded."}
+```{r tombconstruction, fig.dim=c(7.5,4), dpi=300, fig.cap="goes around here"}
 
@@
 On the left, counts are based on the ceramic phase considered most likely to correspond to the tomb's first use, including "PG", "G" and "O" where no greater specificity was possible.
-On the right, aoristic sums are presented, with the uncertainty associated with those general assignations spread across the sub-periods they encompass.
+On the right, aoristic sums are presented, spreading the uncertainty associated with those general assignations.
 In both cases, the trend appears clear.
-The KNC began with a large number of new tombs during the, admittedly prolongued, SM period, followed by a protracted decline in rates of consruction, offset by small resurgences in the late 9th and 8th centuries.
+The KNC began with a large number of new tombs built during the, admittedly prolongued, SM period, followed by a protracted decline in rates of consruction, offset by small resurgences in the late 9th and 8th centuries.
 It should be noted, too, that many of the later "tombs" are not chamber tombs, but a mix of pit, larnax, and pithos burials.
@@
 
-```{r tombsinuse, fig.dim=c(7,4), dpi=300, fig.cap="Estimates of the number of tombs in use during each of the ceramic phases attested in both cemetaries."}
+```{r tombsinuse, fig.dim=c(7,4), dpi=300, fig.cap="goes around here"}
 
@@
 
-These countervalent trends are observable too in the number of tombs receiving depositions in each period.
-Figure \@ref(fig:tombsinuse) presents the estimated number of tombs in use throughout the EIA at both cemeteries.
-At the Forteta, this number remained relatively stable from the MPG to LG periods before declining (and bear in mind, there were only around 20 tombs belonging to this cemetery throughout the period), while at the KNC, the number of tombs in use rose to a preak in the late 8th century before declining.
+These countervalent trends are observable too in the number of tombs receiving depositions in each period (Figure \@ref(fig:tombsinuse)).
+At Forteta, this number remained relatively stable from the MPG to LG periods before declining (and bear in mind, there were only around 20 tombs belonging to this cemetery throughout the period), while at the KNC, the number of tombs in use rose to a preak in the late 8th century before declining.
 But this is not the whole story.
 
-```{r tombsinusebybuilddate, fig.dim=c(7,4), dpi=300, fig.cap="Estimates of the number of tombs in use through time across both cemeteries.
-The grey background represents the total number of tombs in use in each period, while the smaller coloured bars correspond to tombs built in each of the five centuries of EIA activity at the cemeteries."}
+```{r tombsinusebybuilddate, fig.dim=c(7,4), dpi=300, fig.cap="goes around here"}
 tombs_11th<-subset(tombs,tombs$century_built=="11th")
@@
 Figure \@ref(fig:tombsinusebybuilddate) shows the number of tombs in use in each period, divided into the centuries in which those tombs were constructed, all overlaid on the total number of tombs in use across both cemeteries.
-A wave-like pattern can be observed with each century marked by the construction of new tombs which then steadily go out of use, as new ones are built.
+A wave-like pattern can be observed, with each cohort of tombs steadily going out of use as new ones were built.
 This would seem to accord with Cavanagh's suggestion that "[t]he length of time the tombs last seems [...] to reflect the duration of the social group which had access to the sepulchre", and that, if such access was lineally inehrited, frequent fission would result in limited trans-generational use.
-We might expect, then, that it was the 8th and 7th century tombs coming into use that spurred the dramatic rise in burial rates previoulsy noted.
+We might expect, then, that it was the 8th century tombs coming into use that spurred the dramatic rise in burial rates previoulsy noted.
 But this is not the case.
 
-```{r potsbydateandtombdate,  fig.dim=c(7,4), dpi=300, fig.cap="Counts of vessels from across all tombs in both cemeteries, arranged, on the left, by the century to which they are assigned based on their ceramic period and, on the right, by the centuries in which the tombs they come from were built."}
+```{r potsbydateandtombdate,  fig.dim=c(7,4), dpi=300, fig.cap="goes around here"}
 
@@
 
-Figure \@ref(fig:potsbydateandtombdate) shows, firstly, the number of vessels dating to each century across both cemeteries, based on their assigned ceramic period.
-But on the right, those pots are divided into the centuries to which the tombs they were found in originally date.
-So as can be seen, though the late 8th and early 7th centuries represented the apex of burial activity across both cemeteries, it was not the tombs built in those centuries which received the lion's share of the burials and associated ceramics.
+In Figure \@ref(fig:potsbydateandtombdate), all vessels are assigned, firstly, to the century in which they were produced and, secondly, to the century in which the tomb they were found in was built.
+So, though the late 8th and early 7th centuries represented the apex of burial activity across both cemeteries, it was not the tombs built in those centuries which received the lion's share of the burials and associated ceramics.
 This is all the more notable when, returning to Figure \@ref(fig:tombsinusebybuilddate), we see that the number of 9th and 10th century tombs in use by this time was declining.
@@
 
+*Table 1 goes around here*
+
+The significance of these findings can be seen in Table 1.
+Using pithoi as a proxy for burials, we can estimate the changing rates of burial through time.
+And while for the most these remained around three per tomb per century, in the 8th and 7th centuries, tombs founded in the 9th century saw a rapid rise in the rate of burial, peaking at over 11 per tomb per century.
+Though these estimates are no doubt very rough, the marked nature of this divergence suggests it is not simply an artefact of imperfect data.
+
+But Table 1 also demonstrates the markedly low rates of burial which predominated at both cemeteries; two to four burials per tomb per century amounts to a very selective burial rite.
+It seems unlikely, then, that the rising rates of burial noted above are, for instance, the direct result of population growth; the increase is too rapid, and appears concentrated only in certain (older) tombs, not to mention that, based on KULP’s survey data, the major population increase at EIA Knossos probably occurred back in the Protogeometric period (Whitelaw, per comm.).
+
+##Isolating the Phenomenon
+
+We are yet to draw out which older tombs precisely are responsible for the observible patterning so far discussed.
+We might assume the tombs which became the focus of intense later burial activity would be those which yielded the most burials and grave goods when excavated and, happily, this supposition proves true.
+
+```{r toptombs, fig.dim = c(7, 5), dpi=300, fig.align='center',fig.cap="goes around here"}
+
+tombs_ranked<-tombs[with(tombs, order(-tombs$total_pottery)), ]
+tombs_top_12<-head(tombs_ranked,12)
+top12<-as.vector(tombs_top_12$tomb)
+pots_top_12<-pots[is.element(pots$tomb, top12),]
+pots_top_12$tomb<-factor(pots_top_12$tomb,levels=top12)
+
+cols<-c(1,158:207)
+aorist_all_tomb_names<-tombs_top_12[,cols]
+aorist_all_tomb_names<-data.frame(aorist_all_tomb_names)
+colnames(aorist_all_tomb_names)<-c("tomb",dates)
+
+aorist_all_tomb_melt<-melt(aorist_all_tomb_names)
+aorist_all_tomb_melt$variable<-as.numeric(as.character(aorist_all_tomb_melt$variable))
+
+top_12<-tombs_top_12$tomb
+aorist_all_tomb_melt$tomb<-factor(aorist_all_tomb_melt$tomb,levels=top_12)
+
+ggplot(aorist_all_tomb_melt,aes(variable, value))+geom_col()+theme(legend.position = "none")+facet_wrap(~tomb,scales="free_y")+
+  geom_col()+theme_bw()+theme(legend.title=element_blank(),
+      text=element_text(family="Garamond", size=12))+
+          labs(x="Date (Years BC)", y="Aoristic Sums")+
+              scale_fill_brewer(palette = "Set3")+ 
+                theme(legend.position="none")+
+                  coord_cartesian(xlim=c(1100,610))+scale_y_continuous()+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1100,610))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+tombs$group<-ifelse(tombs$tomb=="P"|tombs$tomb=="292"|tombs$tomb=="75"|tombs$tomb=="107"|tombs$tomb=="285"|tombs$tomb=="II"|tombs$tomb=="218","A","B")
+total_pottery_by_groups<-aggregate(tombs$total_pottery, by=list(tombs$group), FUN=sum)
+calc1<-(total_pottery_by_groups[1,2]/sum(total_pottery_by_groups$x))*100
+
+```
+
+The 12 tombs with the most associated vessels are all chamber tombs, 10 dating to the 9th century and two to the 10th, and seven of which did exhibit late, intense episodes of deposition (Figure \@ref(fig:toptombs)).
+In fact, Tombs P, 292, 75, 107, 285, II and 218 (henceforth 'Group A') between them account for `r round(calc1,digits=2)`% of the entire ceramic assemblage from both cemeteries.
+All but Tomb 285 date to the 9th century, and together they underpinned the discussed late spate of burials.
+Interestingly, Tombs X, G, Q, 104 and 283 ('Group B'), however, evince a different pattern, with their respective peaks coming around a century earlier, which, as seen in Figure \@ref(fig:aoristbycem) was another period of increased burial activity.
+So despite receiving the greatest quantities of burials and grave goods, these tombs nonetheless seem to have achieved that distinction via different routes.
+By looking more closely at the periods in which they were built, and in which they were most popular, I believe the differences between these tombs go a long way to explaining the late rise in burial activity here discussed.
+
+##Not All Tombs are Created Equal
+
+The indices employed so far – the distribution of artefacts between tombs, and the construction and use of the tombs through time – inform us of the temporal or aggregate features of the cemeterie, but they tell us little of the tombs as they were built or first used.
+In attempting more synchronic comparisons, we shall first consider the physical dimensions of the tombs as built - a potentially rich context for the negotiation of broader issues of social power, status and group membership.
+
+```{r dimensionsboxplots, fig.dim=c(6,5),dpi=300, fig.align='center',fig.cap="goes around here"}
+
+tombs$century_built<-factor(tombs$century_built,levels=c("11th","10th","9th","8th","7th"))
+dromwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_width,fill=century_built))+geom_boxplot()+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Century of Tomb Construction", y="Dromos Width (m)")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0.1))
+dromlength<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_length,fill=century_built))+geom_boxplot()+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Century of Tomb Construction", y="Dromos Length (m)")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0.5))
+stomwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,stomion_width,fill=century_built))+geom_boxplot()+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Century of Tomb Construction", y="Stomion Width (m)")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0.1))
+chamarea<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,chamber_area,fill=century_built))+geom_boxplot()+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Century of Tomb Construction", y="Chamber Area (sq m)")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0.5))
+
+ggarrange(dromwidth,dromlength,stomwidth,chamarea,nrow=2,ncol=2)
+
+```
+
+In Figure \@ref(fig:dimensionsboxplots), we see that, through the centuries, tomb dimensions did vary, and sometimes consistently, as with the trend toward narrower *dromoi*.
+The 9th-century tombs, though, which include all but two of the 12 discussed above, do not particularly stand out.
+It does seem that chambers were, on average, larger in this period as well as exhibiting greater variety than at any other time.
+Then again, even the largest chambers are not exactly cavernous, and the extra labour expended on construction would have been limited.
+The *dromos* is a plausible area for conspicuous demonstrations of scale but, with their frequently poor preservation, we cannot be confident in assessing this.
+It appears at any rate that they were gradually declining in size, albeit with some notable outliers.
+
+```{r tombdimensions, fig.dim = c(7.5, 4), dpi=300, fig.align='center', fig.cap="goes around here"}
+
+tombareas<-ggplot(subset(tombs,tombs$est_burials>1),aes(chamber_area,est_burials,color=cem_condensed))+geom_point()+
+  theme_bw()+theme(legend.title=element_blank(),
+        text=element_text(family="Garamond", size=12))+
+            labs(x=bquote('Chamber Area ('*m^2*')'), y="Estimated Total Burials")+
+               scale_color_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+
+                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+dromoslengths<-ggplot(subset(tombs,tombs$est_burials>1),aes(dromos_length,est_burials,color=cem_condensed))+geom_point()+
+  theme_bw()+theme(legend.title=element_blank(),
+        text=element_text(family="Garamond", size=12))+
+            labs(x="Dromos Length (m)", y="Estimated Total Burials")+
+               scale_color_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+
+                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+ggarrange(tombareas,dromoslengths,ncol=2,common.legend = TRUE,legend="bottom")
+
+```
+
+Furthermore, it seems that no clear relationship existed between the physical size of a tomb and its buried population at EIA Knossos (see Figure \@ref(fig:tombdimensions)).
+True, over half of all tombs do not exceed 2.6 sq m in area, while nine of the ten tombs with the most burials do surpass this value.
+But many of the very largest tombs are decidedly middling when it comes to burial numbers.
+Things are even less clear for the *dromos*; the largest two tombs by burial numbers also had exceptionally long *dromoi*, but the next eight tombs with the longest *dromoi* do not rank highly on interments.
+
+It seems dubious, then, whether the scale of the tombs when built carried overt social meanings but, even if it did, this didn't prove a determining factor in their later reuse.
+Several tombs had niches cut into their *dromoi*, or the *stomion* extended, to fit in later burial urns.
+Sometimes, urns from within the chamber were relocated to such niches to make space for new interments.
+There is an impression of competition for inclusion into the sepulchral group, further implying the tombs’ original builders little anticipated their eventual popularity.
+
+##Imports, Interments, and Indices of Wealth
+
+Dimensions aside, we can still consider the early use of these tombs through comparisons between the scale and composition of their respective assemblages through time.
+
+```{r tombscompthroughtime, fig.dim=c(7,4),dpi=300,fig.align='center',fig.cap="goes around here"}
+
+pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A",
+       ifelse(pots$tomb=="X"|pots$tomb=="G"|pots$tomb=="104"|pots$tomb=="Q"|pots$tomb=="283","B","C"))
+tombs9th10th<-subset(tombs,century_built=="9th" & total_pottery>50|century_built=="10th" & total_pottery>50)
+pots9th10th<-subset(pots,tomb_century=="9th"|tomb_century=="10th")
+
+tombs9th10th_ranked<-tombs9th10th[with(tombs9th10th, order(-tombs9th10th$total_pottery)), ]
+tombs9th10th_ranked<-as.vector(tombs9th10th_ranked$tomb)
+pots9th10th$tomb<-factor(pots9th10th$tomb,levels=tombs9th10th_ranked)
+pots9th10th<-subset(pots9th10th,century=="9th" & tomb!="NA"|century=="8th" & tomb!="NA"|century=="7th" & tomb!="NA")
+pots9th10th$century<-mapvalues(pots9th10th$century, from = c("9th","8th","7th"), to = c("9th Century", "8th Century","7th Century"))
+
+ggplot(pots9th10th,aes(tomb,fill=group))+geom_bar()+
+  theme_bw()+theme(legend.position="bottom",text=element_text(family="Garamond", size=12),
+               axis.text.x = element_text(angle = 90,hjust=1,vjust=0.4))+
+                 guides(fill=guide_legend(title="Group"))+labs(x="Tomb", y="Vessel Count")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0))+facet_grid(~century,scales="free_x",space="free")
+```
+
+And in Figure \@ref(fig:tombscompthroughtime) we see that the tombs of Group A were relatively unremarkable in the scale of their depositions during the 9th century, only diverging significantly from the rest in the subsequent two hundred years.
+Group B, meanwhile, had their respective peaks in the 9th or 8th centuries, with a near total decline afterwards.
+As a working hypothesis, let us suppose that these two distinct trajectories correspond to changing socio-politcal and economic contexts, made manifest by the burying groups in their mortuary depositions.
+How far might this hypothesis take us in explicating these patterns, and how visible might it be?
+
+The late 9th and early 8th centuries are consistently interpreted as periods of artistic development, increased overseas trade, and renewed vigour in the construction of tombs and the visibility of the burial rite.
+The PGB style reflects a uniquely Cretan blend of Attic and Near Eastern forms, while the appearance of BA 'antiques' in the tombs from this period onwards is felt to reflect rejuvinated, even competitive interest in the Minoan past.
+
+```{r imports, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="goes around here"}
+
+imports<-subset(pots,pots$import!="")
+greekimports<-subset(imports,imports$import_region=="Greece")
+NEimports<-subset(imports,imports$import_region=="Near East")
+
+import_aorist<-colSums(imports[27:76])
+import_aorist<-data.frame(import_aorist)
+colnames(import_aorist)<-c("aoristic_sum")
+row.names(import_aorist)<-dates
+setDT(import_aorist, keep.rownames = "decade")
+import_aorist$decade<-as.numeric(import_aorist$decade)
+
+greekimport_aorist<-colSums(greekimports[27:76])
+greekimport_aorist<-data.frame(greekimport_aorist)
+colnames(greekimport_aorist)<-c("aoristic_sum")
+row.names(greekimport_aorist)<-dates
+setDT(greekimport_aorist, keep.rownames = "decade")
+greekimport_aorist$decade<-as.numeric(greekimport_aorist$decade)
+
+NEimport_aorist<-colSums(NEimports[27:76])
+NEimport_aorist<-data.frame(NEimport_aorist)
+colnames(NEimport_aorist)<-c("aoristic_sum")
+row.names(NEimport_aorist)<-dates
+setDT(NEimport_aorist, keep.rownames = "decade")
+NEimport_aorist$decade<-as.numeric(NEimport_aorist$decade)
+
+greekimport_aorist$origin<-"Greece"
+NEimport_aorist$origin<-"Near East"
+allimport_aorist<-rbind(greekimport_aorist,NEimport_aorist)
+
+ggplot()+geom_col(data=import_aorist,aes(decade,aoristic_sum),alpha=0.2)+
+          geom_col(data=allimport_aorist,aes(decade,aoristic_sum, fill=origin),position="dodge")+
+            theme_bw()+theme(text=element_text(family="Garamond", size=12))+
+                labs(x="Date (Years BC)", y="Aoristic Sum")+
+                  scale_fill_brewer(name="Origin",palette = "Set2")+
+                   scale_y_continuous(expand=c(0,0),limits=c(0,20))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))
+
+```
+
+The rise in imported items is clearly seen in the tomb assemblages, with a succession of increased Greek and Near Eastern imports respectively in the late 9th and early 8th centuries (Figure \@ref(fig:imports)).
+This much has been relatively well understood for some time, at least at the KNC, but we can further examine how these imports were distributed among the various groups of tombs.
+
+```{r importsgroups, fig.dim=c(8,4),dpi=300, fig.align='center',fig.cap="Insert Figure 15 here"}
+
+pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A",
+                   ifelse(pots$tomb=="X"|pots$tomb=="G"|pots$tomb=="104"|pots$tomb=="Q"|pots$tomb=="283","B","C"))
+
+imports<-subset(pots,import!="")
+imports_A<-subset(imports,group=="A")
+imports_B<-subset(imports,group=="B")
+imports_C<-subset(imports,group=="C")
+
+import_A_aorist<-colSums(imports_A[27:76])
+import_A_aorist<-data.frame(import_A_aorist)
+colnames(import_A_aorist)<-c("aoristic_sum")
+row.names(import_A_aorist)<-dates
+setDT(import_A_aorist, keep.rownames = "decade")
+import_A_aorist$decade<-as.numeric(import_A_aorist$decade)
+
+import_A_aorist$group<-"A"
+
+import_B_aorist<-colSums(imports_B[27:76])
+import_B_aorist<-data.frame(import_B_aorist)
+colnames(import_B_aorist)<-c("aoristic_sum")
+row.names(import_B_aorist)<-dates
+setDT(import_B_aorist, keep.rownames = "decade")
+import_B_aorist$decade<-as.numeric(import_B_aorist$decade)
+
+import_B_aorist$group<-"B"
+
+import_C_aorist<-colSums(imports_C[27:76])
+import_C_aorist<-data.frame(import_C_aorist)
+colnames(import_C_aorist)<-c("aoristic_sum")
+row.names(import_C_aorist)<-dates
+setDT(import_C_aorist, keep.rownames = "decade")
+import_C_aorist$decade<-as.numeric(import_C_aorist$decade)
+
+import_C_aorist$group<-"C"
+
+aorist_import_groups<-rbind(import_A_aorist,import_B_aorist,import_C_aorist)
+
+ggplot(aorist_import_groups,aes(decade,aoristic_sum,fill=group))+ geom_col(position="dodge")+
+  theme_bw()+theme(legend.position="bottom",text=element_text(family="Garamond", size=12))+
+  labs(x="Date (Years BC)", y="Aoristic Sum")+
+  guides(fill=guide_legend(title="Group"))+
+  scale_fill_brewer(palette = "Set2")+ 
+  scale_y_continuous(expand=c(0,0),limits=c(0,12))+
+  scale_x_reverse(expand = c(0, 0),limits=c(1000,600))+
+  theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+```
+
+In Figure \@ref(fig:importsgroups), not only is it apparent that a range of tombs were beneficiaries of the increase in imports in the LPG-EG periods, but that there was also potentially significant patterning in where those artefacts ended up.
+The Group A and B tombs closely trace the trends for Greek and Near Eastern imports respectively (and, indeed, tombs Q and G contained only Greek imports), while across all other tombs ('Group C'), there was also no doubt variegation not observable here.
+Imported ceramics are used here as something of a proxy for items of status value, a shorthand that would ideally be bolstered by other artefacts (precious vessels, weapons, jewellery etc.) which, owing to their less-developed chronologies, must unfortunately be excluded.
+But let us suggest, for the time being, that these trends are representative of a larger truth, namely that the late 9th and early 8th centuries witnessed a peak in the variety of depositions, which went unmatched in later periods.
+
+```{r bigbox, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="goes around here"}
+
+aorist_tomb_values<-tombs[c(1,158:207)]
+aorist_tomb_values[aorist_tomb_values == 0] <- NA
+
+aorist_tomb_values<-data.frame(aorist_tomb_values)
+colnames(aorist_tomb_values)[2:51]<-dates
+
+aorist_melt<-melt(aorist_tomb_values)
+aorist_melt$variable<-as.numeric(as.character(aorist_melt$variable))
+
+aorist_melt$period<-ifelse(aorist_melt$variable<=1100 & aorist_melt$variable>=980,"SM",
+		ifelse(aorist_melt$variable<=970 & aorist_melt$variable>=930,"EPG",
+		ifelse(aorist_melt$variable<=920 & aorist_melt$variable>=880,"MPG",
+		ifelse(aorist_melt$variable<=870 & aorist_melt$variable>=850,"LPG",
+		ifelse(aorist_melt$variable<=840 & aorist_melt$variable>=820,"PGB",
+	  ifelse(aorist_melt$variable<=810 & aorist_melt$variable>=800,"EG",
+	  ifelse(aorist_melt$variable<=790 & aorist_melt$variable>=750,"MG",
+	  ifelse(aorist_melt$variable<=740 & aorist_melt$variable>=720,"LG",
+	  ifelse(aorist_melt$variable<=710 & aorist_melt$variable>=680,"EO",
+	  ifelse(aorist_melt$variable<=670 & aorist_melt$variable>=640,"LO",
+	  ifelse(aorist_melt$variable<=630 & aorist_melt$variable>=600,"A",""
+		)))))))))))
+
+decade_ranges<-seq(from = 600, to = 1100, by = 10)
+decade_ranges<-rev(decade_ranges)
+decade_ranges<-as.factor(decade_ranges)
+aorist_melt$variable<-as.factor(aorist_melt$variable)
+aorist_melt$variable<-factor(aorist_melt$variable, levels=(decade_ranges))
+aorist_melt$period<-factor(aorist_melt$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
+aorist_melt$variable<-as.numeric(as.character(aorist_melt$variable))
+
+ggplot(aorist_melt,aes(variable,value,fill=period))+geom_boxplot(aes(group = cut_width(variable, 10)),outlier.size=0.8)+theme_bw()+
+      theme(text=element_text(family="Garamond", size=12))+
+            labs(x="Date (Years BC)", y="Aoristic Sums per Tomb")+
+               scale_fill_brewer(name="Period",palette = "Set3")+
+                scale_x_reverse(expand=c(0,0))+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,20))
+
+```
+
+To test this, admittedly tentative, proposition, Figure \@ref(fig:bigbox) presents a boxplot of the total aoristic sums across both cemeteries, colour-coded by ceramic phase.
+Note that, for visual clarity, the outliers tomb G in the EG period, and tomb P in the late EO-LO period are excluded, with values ranging from 29-34, and 36-44 respectively.
+The important observation is that the PGB and EG exhibit the largest interquartile ranges (IQRs) of any period, and each possess only a single outlier (beyond 1.5 times the IQR).
+The Orientalising period, on the other hand, exhibits much reduced IQRs, and a proliferation of outliers far and above the range of the other tombs.
+In other words, whilst both the PGB-EG and EO-LO periods witnessed an expanding volume and range of burial activity, in the earlier period this was distributed more evenly across the total burying population.
+One speaks of increasing plurality, the other of divergence.
+
+```{r gini, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="goes around here"}
+
+aorist_tomb_values<-tombs[c(1,158:207)]
+aorist_tomb_values[aorist_tomb_values == 0] <- NA
+
+aorist_tomb_values<-data.frame(aorist_tomb_values)
+colnames(aorist_tomb_values)[2:51]<-dates
+
+tombs_gini<-matrix(0, ncol = 0, nrow = 1)
+tombs_gini<-data.frame(tombs_gini)
+for(i in colnames(aorist_tomb_values[2:51])) {
+  tombs_gini[[paste0(i)]] <- ineq(aorist_tomb_values[[i]],type="Gini")
+}
+
+tombs_gini<-t(tombs_gini)
+tombs_gini<-data.frame(tombs_gini)
+
+colnames(tombs_gini)<-c("gini_coefficient")
+row.names(tombs_gini)<-dates
+tombs_gini<-data.frame(tombs_gini)
+setDT(tombs_gini, keep.rownames = "decade")
+tombs_gini$decade<-as.numeric(tombs_gini$decade)
+
+rects <- data.frame(xstart = c(1100,970,920,875,840,810,790,745,70,670,630), 
+                    xend = c(970,920,875,840,810,790,745,710,670,630,600), 
+                    col = c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
+rects$col<-factor(rects$col,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
+
+ggplot() + 
+  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf, fill = col), alpha = 0.4) +
+     geom_line(data = tombs_gini, aes(decade,gini_coefficient),size=0.5)+
+      theme_bw()+theme(text=element_text(family="Garamond", size=12))+
+              labs(x="Date (Years BC)", y="Gini Coefficient")+
+                 scale_fill_brewer(name="Period",palette = "Set3")+ coord_cartesian(
+                    xlim = c(1100, 600), ylim = c(0.5,0.8))+ scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
+                       scale_y_continuous(expand=c(0,0))
+```
+
+This possibility is supported further by considering the gini coefficient of the distribution of pottery between the tombs (see Figure \@ref(fig:gini).
+Using the aoristic sums as a probablistically-weighted estimate of the volume of material deposited in each tomb in each decade, use of the gini coefficient can give an (albeit rough) estimate of how unequally distributed that material was.
+And, though values were always relatively high (i.e.
+unequal) within the cemeteries, the major spike in the gini coefficient seen in the Orientalising period does contrast markedly with the modest rise and plateau of the PGB-EG.
+The gini coefficient is most commonly used in the assessment of income inequality, but I want to stress here that it is not "wealth" in any meaningful sense that we are talking about here.
+Indeed, if anything, the late flourit of burial activity here discussed may be a case of quantity, and not quality.
+
+```{r comps, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="goes around here"}
+
+pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A","B")
+
+potcompKNC<-ggplot(subset(pots,century!="" & cem_condensed=="NC"),aes(century,fill=type))+geom_bar(position="fill")+
+  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title = element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(title="KNC",x="", y="Proportion of Total Assemblage")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" Amphora  "," Jug  "," Oil vessel  ", " Pithos"," Drinking vessel  "," Lid  "," Other"))+ 
+                    scale_y_continuous(expand=c(0,0.0))
+potcompFOR<-ggplot(subset(pots,century!="" & cem_condensed=="FOR"),aes(century,fill=type))+geom_bar(position="fill")+
+  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(title="Fortetsa",x="Century BC", y="")+
+                  scale_fill_brewer(palette = "Set2", labels=c(" Amphora  "," Jug  "," Oil vessel  ", " Pithos"," Drinking vessel  "," Lid  "," Other"))+ 
+                    scale_y_continuous(expand=c(0,0.0))
+
+potcomptop7<-ggplot(subset(pots,century!="" & century!="11th"& group=="A"),aes(century,fill=type))+geom_bar(position="fill")+
+  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(title="Group A",x="", y="")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" Amphora  "," Jug  "," Oil vessel  ", " Pithos"," Drinking vessel  "," Lid  "," Other"))+ 
+                    scale_y_continuous(expand=c(0,0.0))
+
+ggarrange(potcompKNC,potcompFOR,potcomptop7,ncol=3,common.legend =TRUE,legend="bottom")
+
+spearmanobjectspots<-cor(tombs$total_object,tombs$total_pottery,method="spearman")
+```
+
+Figure \@ref(fig:comps) shows the composition of the total pottery assemblage from each cemetery, and of the Group A tombs, through the centuries.
+In all cases, there appears to be the steady crystallisation of a basic 'burial kit', comprising a pithos and lid, an oil vessel, a jug, and a drinking vessel.
+The proportion of grave goods comprising 'other' vessels - pyxides, askoi, trays, stands, dinoi, kernoi, kraters, lekanai and more - declines to become only a minor part of the standardised repertoire of the 7th century.
+The tombs of Group A did not break with this mould, and their assemblages seem as conventional as any other during the last century of use in the cemeteries.
+
+Plainly, the main focus of this article has been on ceramics and not other artefacts of metal, stone, glass, and organic materials, due to their less well-developed chronologies, but even the raw distribution of such artefacts may make more sense in the light of the foregoing discussion.
+There is, perhaps unsurprisingly, a relatively strong (Spearman's ρ=`r spearmanobjectspots`) correlation between the number of ceramic and non-ceramic finds across the tombs, and, certainly, many of the most unique and precious objects were recovered from among the larger tombs.
+But the monopoly those tombs would appear to have over such objects, perhaps tempting one initially to speak of 'elite' or 'high-status' tombs, evaporates when the number of those objects is divided by the number of burials represented.
+This fact has seldom, if ever, been appreciated in prior discussions of the cemeteries.
+
+```{r hists, fig.dim=c(6,4),dpi=300, fig.align='center',fig.cap="goes around here"}
+tombs$group<-revalue(tombs$group,c("A"="A","B"="B","C"="B"))
+total_object<-ggplot(subset(tombs,est_burials>1 & total_object>0),aes(total_object, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Total Non-Ceramic Finds", y="Number of Tombs")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" Group A  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+total_metal<-ggplot(subset(tombs,est_burials>1 & weapon>0),aes(weapon, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Total Metal Artefacts", y="Number of Tombs")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" Group A  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+total_import<-ggplot(subset(tombs,est_burials>1 & total_import>0),aes(total_import, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Total Imports", y="Number of Tombs")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" Group A  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+
+objectsbyburial<-ggplot(subset(tombs,est_burials>1 & total_object>0),aes(total_object/est_burials, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Total Non-Ceramics/Burials", y="Number of Tombs")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" Group A  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+metalbyburial<-ggplot(subset(tombs,est_burials>1 & weapon>0),aes(weapon/est_burials, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Total Metal Artefacts/Burials", y="Number of Tombs")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" Group A  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+
+importsbyburial<-ggplot(subset(tombs,est_burials>1 & total_import>0),aes(total_import/est_burials, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="Garamond", size=12))+
+                 labs(x="Total Imports/Burials", y="Number of Tombs")+
+                  scale_fill_brewer(palette = "Set2",labels=c(" Group A  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+
+ggarrange(total_import,total_metal,total_object,importsbyburial,metalbyburial,objectsbyburial,ncol=3,nrow=2,common.legend = TRUE,legend="bottom")
+```
+
+Figure \@ref(fig:hists) thus supports the view which has emerged from the foregoing analysis, that the major increase in burial activity at both Fortetsa and the KNC was not a phenomenon linked in any straightforward way to simple principles of 'elite' or 'wealthy' burial.
+To summarise, towards the end of 8th century BC, a select group of tombs, all of which had been constructed back in the 9th or 10th, became the foci of a major increase in burials, something which did not occur among the tombs built around that time.
+We have seen how this surge in acivity occurred against the backdrop of a declining number of older tombs in use, and I have suggested that the rates of change are too rapid, and unevenly distributed, to be explicable through population growth alone.
+We have seen how little about the physical dimensions or early use of these tombs presaged their later popularity, and that, while a similar, if smaller, increase in burial activity occurred around a century earlier, in the PGB-EG periods, this contrasted in the comparative diversity of its associated assemblages, and their distribution across the tombs.
+The later burials were marked by great uniformity in the associated grave goods, and this apparently limited and consistent repertoire may explain why many of the most-used tombs prove exceptional only in their absolute quantities of artefacts, and not the accoutrements of each individual burial.
+
+# Discussion
+
+The preceding analysis has revealed a number of potentially significant patterns in the use of the Fortetsa and Knossos North Cemeteries; it only remains to consider how we might characterise the burying groups responsible behind them.
+The traditional view is that such groups were immediate families (Brock 1957, 41; @Boardman1967, 63; @Coldstream1996, 291).
+The relatively short lifespan of many tombs (often only two or three generations) has been taken as evidence for direct, lineal inheritance, with the tomb going out of use as families moved, individuals founded new tombs, or the line came to an end.
+The idea of lineal groups is supported by the, admittedly circumstantial, evidence of men, women and children being found within single tombs, and one (contentious) instance of an apparently hereditary mandibular deformity in three non-contemporary skulls in the same tomb [@Musgrave1996].
+
+The model of lineal inheritance has been recently challenged by Kotsonas [-@Kotsonas2011, 131], however, who considers it ‘probably the greatest factoid in the archaeology of Crete of the given period’.
+He argues that the rise in interments during the 8th century is too great to be explained by population growth, as would be necessary for the possibility of direct lineal inheritance to be preserved.
+Certainly, it seems illogical that population increases would lead to the consolidation of burials within only extant tombs.
+The trend instead must reflect reconfigurations of the burying groups themselves, which Kotsonas feels may have, also or instead, comprised age- and sex-grades.
+
+Yet it is hard to be conclusive, as our current understanding of the social organisation of the early Cretan *poleis* remains patchy.
+Perlman and Gagarin's [-@Gagarin2016] recent study of the Cretan law codes reminds us, importantly, if lamentably, that there is an often underappreciated degree of diversity apparent from the Archaic inscriptions in the offices, institutions, and social groups of the early city-states.
+In the traditional view, political authority was concentrated in the hands of the *kosmoi*, aristocratic magistrates selected from the leading, hereditary *startoi*, or ‘clans’, networks of extended lineages controlling the ancestral estates on which worked attendent serfs and/or slaves (Willetts 1982, 240).
+But despite elements of this account probably being accurate, others remain equivocal, such as the nature of the *pyla*, or 'tribe', and its relation to the *startos* (which may itself be a military grouping), the distinction between an *astos* and a *poliatas*, both clearly formulations of the rights-bearing civic actor, and the ancestry and spread of the *agelai*, seemingly an adolescent age-grade so far only attested at Eltynia [@Gagarin2016, 36; 78-80].
+Indeed, many terms which might illuminate social divisions and groupings within the Archaic Cretan populace are only attested at one or two sites, and even then sometimes appear to differ in their form or function.
+
+It wouldn't be helpful at present, then, to draw any simple parallel between the Knossian burial groups and, say, powerful *startoi*, a term which is not even attested outside of Gortyn in this period (ibid., 1-2).
+But archaeological and textual evidence suggests that, in general terms, such supra-familial kinship groups were undergoing a transition during this period, suggesting some such attribution for the burying groups at Knossos is not unreasonable, given the noted changes there also.
+Elsewhere on Crete, we see a shift to unmarked, individual burials, and more austere grave goods, as well as an apparent reorientation of sanctuary deposits away from extra-mural to suburban spaces, and a rise in utilitarian, mass-produced votives (ibid., 34-6).
+At Knossos, a small number of pithos burials in pits do appear in the KNC and Fortetsa volumes (Brock 1957, 98; Coldstream and Catling 1996, 162), and recent re-excavation of the Teke plot identified numerous small pit burials around three chamber tombs [@Evely2012], raising the possibility that others exist, but have gone unrecognised in prior excavations.
+At Azoria, major urban remodelling in the 7th century saw the earlier, agglutinative house forms of neighbouring Kavousi Vronda and Kastro give way to larger, more formally planned dwellings with little evidence of renovation or remodelling, while larger, apprently civic, buildings were constructed to accommodate major public gatherings [@Haggis2011; @Haggis2011b; @Haggis2014].
+
+Together, these may be indications that with the emergence of the archaic *poleis* on Crete, there was a concomitant restructuring of wider social groups, institutions, and behaviours.
+Was this transformation linked to an emerging emphasis on the citizen, the state, and ideologies of relative egality, within which communal gatherings, intra-mural ritual, and individual graves replaced ostentatious, kin-centred acts of gathering, dedication, and burial?
+It is plausible, though as Perlman and Gagarin (32-37) concede, the decline of ostentatious burial could just as well represent a voluntary abandonment by groups assured of their economic and political influence, finding the symbolic expression of kin-relations in death no longer a necessary expense.
+
+These issues, naturally, recall the conclusions of Morris [-@Morris1987, 171-211], who argued 30 years ago that the emergence of the *polis* across the Aegean led to a tension between elite preferences for conspicuous display, and a downplaying of social difference in death, as promoted by the ‘middling’ ideology that the *polis* engendered and promulgated.
+In his well-known model of the *agathoi* and *kakoi*, Morris argued that, at various times, the relevant spheres of kinship demonstrated in burial could widen and constrict, with more and less distantly related lineages drawn into association through changing social strategies, until the pre-eminence of kinship relations in the structuring of political power declined, along with its manifestiations in mortuary practice.
+
+At Knossos, while some aspects of the noted expansion in the burying population seem explicable within a model like Morris', others sit more uneasily.
+Firstly, the burials at issue here are in many ways *not* ostentatious in comparison to the norm.
+We have seen that the 'richest' tombs in terms of finds are far less so when one accounts for the number of burials represented.
+Likewise, Morris claims that periods of expansion in the rite of burial correspond to greater variety in the associated assemblages, yet, at Knossos, we have seen that the range of grave goods in fact diminished and crystallised during the late 8th and early 7th centuries.
+But whether this reflected some standard set of accoutrements intentionally shared by the hypothetical *agathoi* of Knossos, or else nascent, egalitarian strictures over elaborate burial is hard to judge.
+The PGB period seems a far better candidate for such an expansion, with its greater variety in chamber dimensions, relatively broad distribution of (local and imported) grave goods, not to mention the innovative decoration and reuse of BA artefacts noted elsewhere.
+One thing is clear; the social context which produced these signatures contrasted markedly with that of a century later, with its ever greater volume of interments in an ever-smaller cohort of tombs, its more uniform assemblages, and its decline of imported items.
+
+It is tempting, though potentially perilous, to link the diminishing range of burial goods to Archaic Crete's purported austerity, its (quite literally) spartan lifestyles.
+But, as a more nuanced picture emerges of the island's relationship with the wider Mediterranean, subtler characterisations are possible.
+It is telling that the peak for imported items in the tombs came in the 9th century; the 'Orientalising' as a time period feels increasingly misapplied to Crete, when, at Knossos at least, it witnessed far lower rates of 'oriental' imports, and only limited adoption of the figural motifs which characterise its appearance elsewhere.
+This decline in imports was, however, counterpointed by a rise in imitations of Near Eastern vessels [@Antoniadis2012, 170], something which further accords with our growing awareness of the selective nature of Archaic and Classical Crete's external relations.
+Erickson [@Erickson2010] has, likewise, argued that *symposia* were practised on Crete, something often disputed, but that their practitioners merely favoured simpler vessel forms than the figurally-decorated wares popular on the mainland.
+Local, selective adoption of, and responses to, wider social and material trends across the contemporary Aegean is perhaps becoming the hallmark of Archaic Crete, and finds further endorsement in the trends identified in the EIA cemeteries at Knossos.
+What seems certain is that this was a period of transition wherein, whatever their exact identity, certain actors invested heavily in a final flurry of burials in chamber tombs, with an emphasis on tradition, an established repertoire of items, rites, and venues for one's passage to the afterlife, before changing social forces effected their abandonment.
+
+# Conclusion
+
+It is important that the present study was a comparative investigation of both the Fortetsa and Knossos North cemeteries.
+It is notable how similar the observed trends are at both locations.
+That both underwent similar developments during the EIA suggests, not only that they catered to similar sub-sections of the Knossian population, but that some of the socio-economic and political changes inferred and discussed were real, and community-wide.
+Further work should seek to extend these comparisons, both into trends unexplored in the present study, and out to the wider funerary landscape of Knossos.
+I hope to produced such a study but, for the time being, the datasets I am making available alongside this article include the other published EIA tombs in the area, so that others may also explore their possible significance.
+
+There were undeniable limitations to the present work.
+For one, non-ceramic artefacts were underrepresented due to difficulties with dating, and future work could seek to address this.
+Issues of gender were sadly little considered, though this was largely due to limitations in the data themselves.
+There have also been a number of tombs discovered in the last 20 years that fill in the spaces between the Teke plot and the Medical Faculty Site, and ideally these unpublished tombs should be included in further analyses as they, along with the unconsidered Teke tholos and associated tombs, are almost surely part of a single contiguous cemetery [@Blackman1999, 113; @Catling1983, 51; @French1991, 68-69].
+Finally, a preliminary study was recently taken by the author of select boxes marked as 'unpublished' from the KNC deposits kept at the British School at Athens at Knossos.
+Despite going unmentioned in the publication, this material was not inconsiderable and so, regretably, future study will be necessary to determine whether it threatens the integrity of quantitative analyses such as this.
+Suffice to say that, at present, it is another pinch of salt with which to take the present findings.
+
+These limitations notwithstanding, the foregoing analysis demonstrates the value of quantitative approaches to the archaeological record of the period.
+With large datasets, we are able to locate and visualise covariance and significant correlations between different variables, plot changes in the data through time and, ultimately, draw on our findings to inform discussions of social processes.
+Though the construction of databases can be time consuming, once made they offer the opportunity to subdivide, compare, and plot the data in numerous ways quickly and easily, facilitating investigation of the material from multiple angles.
+By making datasets available for use by others, as with the data and code for the present study, one’s conclusions can also be independently tested, and alternative interpretations can be voiced.
+The databases produced for the above analysis contain a large of amount of unutilised data that could facilitate a range of future investigations, and all are welcome to it.
+With more such analyses, we could begin to a construct a fuller account of the regional diversity in mortuary, and wider social, practices that characterised the communities of EIA and archaic Crete.
+
 ##### pagebreak
+# Acknowledgements
+
+I would like to thank Borja Legarra Herrero and Corinna Riva for their help and advice from the earliest stages of the present research, as well as the BSA at Knossos for permitting me to examine some of the material from the Medical Faculty excavations.
+I am indebted most of all to Professor Todd Whitelaw, for his unstinting guidance, feedback and facilitation of my ongoing research.
+
+# About the Author
+
+Dominic Pollard is a PhD candidate at University College London.
+His research focusses on the Early Iron Age of Crete, with particular interest in the use of computational methods to supplement traditional modes of archaeological inference.
+
+##### pagebreak
+# References 
+<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
+<div id="refs"></div>
+
+##### pagebreak
+# Captions
+
+*Figure 1 Map of the Knossos area, showing the locations of the various EIA tomb clusters and cemeteries, with EIA pottery distributions from the Knossos Urban Landscape Project marked in orange, giving an impression of the contemporary urban extent (underlying image credit: T.
+Whitelaw).*
+
+*Figure 2 Relationship between known interments, counts of pithoi, and counts of pithoi, amphorae and kraters across both cemeteries.
+The straight line represents the mean estimate based on these three values.*
+
+*Figure 3 Counts of pithoi across both cemeteries through time. Vessel dates have here been based on the midpoint (or mean) of the absolute date range accompanying their respective ceramic periods.*
+
+*Figure 4 Aoristic sums of all vessels across both cemeteries. This represents a probabilistically weighted visualisation of depositional activity through time.*
+
+*Figure 5 Aoristic sums of pithoi (left), as a proxy for cremation burials, and of all vessels (right). The sums are colour coded to show the values of the two cemeteries, and the total values*
+
+*Figure 6 Tomb construction through time at both cemeteries.
+On the left are counts based on the earliest dated pottery in each tomb, while on the left are aoristic sums likewise based on the earliest ceramic-based assignations.
+In both cases, tombs lacking any securely dateable pottery are excluded.*
+
+*Figure 7 Estimates of the number of tombs in use during each of the ceramic phases attested in both cemetaries, based on counts of tombs yielding vessels of each period.*
+
+*Figure 8 Estimates of the number of tombs in use through time across both cemeteries.
+The grey background represents the total number of tombs in use, while the smaller coloured bars correspond to tombs built in each of the five centuries of EIA activity at the cemeteries.*
+
+*Figure 9 Counts of vessels from all tombs in both cemeteries, arranged, on the left, by the century to which they are assigned based on their ceramic period and, on the right, by the centuries in which the tombs they come from were built.*
+
+*Figure 10 Aoristic sums for the 12 tombs with the largest ceramic assemblages, arranged in decreasing order.
+Note, for the sake of better illustrating the relative temporal patterns, rather than absolute values, the scales on each plot are distinct.*
+
+*Figure 11 Boxplots comparing various tomb dimensions, divided, in each case, by the century of tomb construction.*
+
+*Figure 12 Scatterplots comparing tomb dimensions (chamber area and *dromos* length) to the estimated total number of burials by tomb.*
+
+*Figure 13 Counts of vessels dating to the 9th, 8th and 7th centuries respectively, found in tombs built in the 10th or 9th century which yielded upwards of 50 vessels when excavated.
+The seven tombs with late peaks in burial activity (Group A) are shown in green, the remaining five of the largest 12 (Group B) in orange, and all other tombs (Group C) in blue.*
+
+*Figure 14 Aoristic sums of imported ceramics through time across both cemeteries, colour-coded by the broad regional origin of the vessels, and overlain on the grand totals in pale grey.*
+
+*Figure 15 Aoristic sums of imported vessels through time. The sums are colour coded as per Figure 13 (thoug here Group C represents all other tombs, not just those with 50+ vessels).*
+
+*Figure 16 Boxplot showing the aoristic sums of all vessels across both cemeteries through time, colour-coded by ceramic period.
+Boxes represent the interquartile range, horizontal lines the median value, whiskers values within 1.5 times the interquartile range, and points outliers beyond this.
+The Y axis has been limited to 20 for ease of interpretation.*
+
+*Figure 17 Gini coefficient values through time based on the aoristic sums for all vessels across both cemeteries. Background colours correspond to ceramic periods.*
+
+*Figure 18 The composition of the total pottery assemblage for each century across both cemeteries, and among the 7 notable tombs (Group A) drawn out earlier in the analysis.*
+
+*Figure 19 Histograms corresponding, first, to counts of non-ceramic artefacts, metal objects, and imported items by tomb across both cemeteries and, secondly, to those counts divided by the estimated number of burials in each tomb.*
+
+##### pagebreak
+# Figures
+
+##### pagebreak
+# Tables
 
@@
 
-Table: Table 1 The number of pithoi (as a proxy for burials) deposited in tombs dating to the 10th, 9th, 8th and 7th centuries, through each of the centuries they were in use, as well as the number of tombs from which these burials came, allowing calculation of the changing rate of burial through time.
-
-The significance of these findings can be seen in Table 1.
-Using pithoi as a proxy for burials, we can estimate the changing rates of burial through time based on how many tombs yielded how many pithoi during each centuring And while for the most part burial rates remained around three per tomb per century, in the 8th and 7th centuries, tombs founded in the 9th century saw a rapid rise in the rate of burial, peaking at over 11 per tomb per century.
-Though these estimates are no doubt very rough, the marked nature of this divergence suggests it is not simply an artefact of imperfect data.
-
-Also apparent from Table 1, though, are the markedly low rates of burial which predominated at both cemeteries.
-Two to four burials per tomb per century amounts to a very selective burial rite, even considering the restricted portion of the population probably represented by these cemeteries [@Cavanagh1996, 664].
-It seems unlikely, then, that the rising rates of burial noted above are, for instance, the direct result of population growth; the increase seems too rapid, and is concentrated only in certain (older) tombs, not to mention that, based on KULP’s survey data, the major population increase at EIA Knossos should be dated a century or more before the late 8th century peak in burial, in the Protogeometric period (Whitelaw, per comm.).
-
-##Isolating the phenomenon
-
-We are yet to draw out which older tombs precisely are responsible for the observible patterning so far discussed.
-We might assume that the tombs which became the focus of intense burial activity in the late 8th and early 7th centuries would be those which yielded the most burials and grave goods when excavated and, happily, this supposition proves true.
-
-```{r toptombs, fig.dim = c(7, 5), dpi=300, fig.align='center',fig.cap="Aoristic sums for the 12 largest tombs (in terms of grave goods), arranged in decreasing order of total recovered artefacts.
-Note, for the sake of better illustrating the relative patterns, rather than absolute values, the scales on each plot are distinct."}
-
-tombs_ranked<-tombs[with(tombs, order(-tombs$total_pottery)), ]
-tombs_top_12<-head(tombs_ranked,12)
-top12<-as.vector(tombs_top_12$tomb)
-pots_top_12<-pots[is.element(pots$tomb, top12),]
-pots_top_12$tomb<-factor(pots_top_12$tomb,levels=top12)
-
-cols<-c(1,158:207)
-aorist_all_tomb_names<-tombs_top_12[,cols]
-aorist_all_tomb_names<-data.frame(aorist_all_tomb_names)
-colnames(aorist_all_tomb_names)<-c("tomb",dates)
-
-aorist_all_tomb_melt<-melt(aorist_all_tomb_names)
-aorist_all_tomb_melt$variable<-as.numeric(as.character(aorist_all_tomb_melt$variable))
-
-top_12<-tombs_top_12$tomb
-aorist_all_tomb_melt$tomb<-factor(aorist_all_tomb_melt$tomb,levels=top_12)
-
-ggplot(aorist_all_tomb_melt,aes(variable, value))+geom_col()+theme(legend.position = "none")+facet_wrap(~tomb,scales="free_y")+
-  geom_col()+theme_bw()+theme(legend.title=element_blank(),
-      text=element_text(family="Garamond", size=12))+
-          labs(x="Date (Years BC)", y="Aoristic Sums")+
-              scale_fill_brewer(palette = "Set3")+ 
-                theme(legend.position="none")+
-                  coord_cartesian(xlim=c(1100,610))+scale_y_continuous()+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1100,610))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-tombs$group<-ifelse(tombs$tomb=="P"|tombs$tomb=="292"|tombs$tomb=="75"|tombs$tomb=="107"|tombs$tomb=="285"|tombs$tomb=="II"|tombs$tomb=="218","A","B")
-total_pottery_by_groups<-aggregate(tombs$total_pottery, by=list(tombs$group), FUN=sum)
-calc1<-(total_pottery_by_groups[1,2]/sum(total_pottery_by_groups$x))*100
-
-```
-
-The 12 tombs with the most associated vessels are all chamber tombs, with 10 dating to the 9th century, and two to the 10th.
-And, in Figure \@ref(fig:toptombs), we see that seven of the 12 did exhibit late, intense episodes of deposition.
-In fact, Tombs P, 292, 75, 107, 285, II and 218 between them account for `r round(calc1,digits=2)`% of the entire pottery assemblage recovered from the `r nrow(tombs)` tombs considered here.
-All but Tomb 285 date to the 9th century, and together they underpinned the discussed late spate of burials.
-Interestingly, Tombs X, G, Q, 104 and 283, however, evince a different pattern, with their respective peaks coming in the PGB and EG periods, around a century earlier.
-We have seen already in Figure \@ref(fig:aoristbycem) that, particularly in the KNC, the end of the 9th century was another period of increased burial activity.
-Though recipients of the greatest numbers of burials and grave goods in both cemeteries, they nonetheless seem to have achieved that distinction via different routes.
-By looking more closely at the periods in which they were built, and in which they were most popular, I hope to shed light on the broader social contexts in which these tombs flourished.
-
-##Not all tombs are created equal
-
-The indices discussed so far – the distribution of of artefacts between the tombs, and the construction and use of the tombs through time – only inform us of the temporal or aggregate features of the cemeteries; they tell us little of the tombs as they were built or first used.
-Synchronic comparisons are harder to make with the available data, but the dimensions of the tombs offer some scope for doing so.
-Perhaps the tombs which went on to receive the most burials were notably large or grand when first built, reflecting the status of the associated burying group.
-
-```{r dimensionsboxplots, fig.dim=c(6,5),dpi=300, fig.align='center',fig.cap="Boxplots comparing various tomb dimensions as they vary by the century of tomb construction."}
-
-tombs$century_built<-factor(tombs$century_built,levels=c("11th","10th","9th","8th","7th"))
-dromwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_width,fill=century_built))+geom_boxplot()+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Century of Tomb Construction", y="Dromos Width (m)")+
-                  scale_fill_brewer(palette = "Set2")+ 
-                    scale_y_continuous(expand=c(0,0.1))
-dromlength<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_length,fill=century_built))+geom_boxplot()+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Century of Tomb Construction", y="Dromos Length (m)")+
-                  scale_fill_brewer(palette = "Set2")+ 
-                    scale_y_continuous(expand=c(0,0.5))
-stomwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,stomion_width,fill=century_built))+geom_boxplot()+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Century of Tomb Construction", y="Stomion Width (m)")+
-                  scale_fill_brewer(palette = "Set2")+ 
-                    scale_y_continuous(expand=c(0,0.1))
-chamarea<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,chamber_area,fill=century_built))+geom_boxplot()+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Century of Tomb Construction", y="Chamber Area (sq m)")+
-                  scale_fill_brewer(palette = "Set2")+ 
-                    scale_y_continuous(expand=c(0,0.5))
-
-ggarrange(dromwidth,dromlength,stomwidth,chamarea,nrow=2,ncol=2)
-
-```
-
-Looking at Figure \@ref(fig:dimensionsboxplots), we see that, through the centuries, the dimensions of the tombs did vary, and sometimes consistently, as with the trend toward narrower *dromoi*.
-The 9th-century tombs, though, which include all but two of the 12 discussed above, do not particularly stand out.
-It does seem that chambers were, on average, larger in this period as well as varying more in size than at any other time.
-Then again, even the largest chambers are not exactly cavernous, and the extra labour expended on construction would have been limited.
-The *dromos* is a plausible area for conspicuous demonstrations of scale but, with their frequently poor preservation, we cannot be confident in assessing this.
-It appears at any rate that they were gradually declining in size, albeit with some notable outliers.
-
-```{r tombdimensions, fig.dim = c(7.5, 4), dpi=300, fig.align='center', fig.cap="Scatterplots comparing tomb dimensions (chamber area and *dromos* length) to estimated total burials by tomb."}
-
-tombareas<-ggplot(subset(tombs,tombs$est_burials>1),aes(chamber_area,est_burials,color=cem_condensed))+geom_point()+
-  theme_bw()+theme(legend.title=element_blank(),
-        text=element_text(family="Garamond", size=12))+
-            labs(x=bquote('Chamber Area ('*m^2*')'), y="Estimated Total Burials")+
-               scale_color_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+
-                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
-                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-dromoslengths<-ggplot(subset(tombs,tombs$est_burials>1),aes(dromos_length,est_burials,color=cem_condensed))+geom_point()+
-  theme_bw()+theme(legend.title=element_blank(),
-        text=element_text(family="Garamond", size=12))+
-            labs(x="Dromos Length (m)", y="Estimated Total Burials")+
-               scale_color_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+
-                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
-                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-ggarrange(tombareas,dromoslengths,ncol=2,common.legend = TRUE,legend="bottom")
-
-```
-
-Furthermore, it seems that no clear relationship existed between the physical size of a tomb and its buried population at EIA Knossos (see Figure \@ref(fig:tombdimensions)).
-It is true that over half of all tombs do not exceed 2.6 sq m in area, while nine of the ten tombs with the most burials do surpass this value.
-But many of the very largest tombs are decidedly middling when it comes to burial numbers.
-Things are even less clear for the *dromos*; the largest two tombs by burial numbers also had exceptionally long *dromoi*, but the next eight tombs with the longest *dromoi* do not rank highly on interments.
-
-It seems dubious, then, whether the scale of the tombs when built carried overt social meanings but, even if it did, this clearly proved irrelevant to their later reuse.
-Several tombs had niches cut into their *dromoi*, or the *stomion* extended, to fit in later burial urns.
-Sometimes, urns from within the chamber were relocated to such niches to make space for new interments.
-There is an impression of competition for inclusion into the sepulchral group, supporting the idea that the tombs’ original builders little anticipated their eventual popularity and so did not scale their tombs accordingly.
-
-##Imports, interments, and indices of wealth
-
-The dimensions of the tombs may have shed little light on the early use of what would become the most used tombs in both cemeteries, but we can still investigate the changing scale and composition of the assemblages of these tombs through time.
-
-```{r tombscompthroughtime, fig.dim=c(7,4),dpi=300,fig.align='center',fig.cap="Counts of vessels belonging to the 9th, 8th and 7th centuries, found in tombs dating to the 10th or 9th century which yielded upwards of 50 vessels when excavated.
-The seven large tombs with late peaks in burial activity identified earlier are shown in green."}
-
-pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A",
-       ifelse(pots$tomb=="X"|pots$tomb=="G"|pots$tomb=="104"|pots$tomb=="Q"|pots$tomb=="283","B","C"))
-tombs9th10th<-subset(tombs,century_built=="9th" & total_pottery>50|century_built=="10th" & total_pottery>50)
-pots9th10th<-subset(pots,tomb_century=="9th"|tomb_century=="10th")
-
-tombs9th10th_ranked<-tombs9th10th[with(tombs9th10th, order(-tombs9th10th$total_pottery)), ]
-tombs9th10th_ranked<-as.vector(tombs9th10th_ranked$tomb)
-pots9th10th$tomb<-factor(pots9th10th$tomb,levels=tombs9th10th_ranked)
-pots9th10th<-subset(pots9th10th,century=="9th" & tomb!="NA"|century=="8th" & tomb!="NA"|century=="7th" & tomb!="NA")
-pots9th10th$century<-mapvalues(pots9th10th$century, from = c("9th","8th","7th"), to = c("9th Century", "8th Century","7th Century"))
-
-ggplot(pots9th10th,aes(tomb,fill=group))+geom_bar()+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="Garamond", size=12),
-               axis.text.x = element_text(angle = 90,hjust=1,vjust=0.4))+
-                 labs(x="Tomb", y="Vessel Count")+
-                  scale_fill_brewer(palette = "Set2")+ 
-                    scale_y_continuous(expand=c(0,0))+facet_grid(~century,scales="free_x",space="free")
-```
-
-And in Figure \@ref(fig:tombscompthroughtime) we see that the tombs of Group A were relatively unremarkable in the scale of their depositions during the 9th century, only diverging significantly from the rest in the subsequent two hundred years.
-The Group B tombs, meanwhile, had their respective peaks in the 9th or 8th centuries, with a near total decline afterwards.
-As a working hypothesis, let us suppose that these two distinct trajectories correspond to changing socio-politcal and economic contexts, made manifest by the burying groups in their mortuary depositions.
-How far might this hypothesis take us in explicating these patterns, and how visible might it be?
-
-The late 9th and early 8th centuries are consistently interpreted as periods of artistic development, increased overseas trade, and renewed vigour in the construction of tombs and the visibility of the burial rite.
-The PGB style reflects a uniquely Cretan blend of Attic and Near Eastern forms, while the appearance of BA 'antiques' in the tombs from this period onwards is felt to reflect renewed, even competitive interest in the Minoan past, as a tool for mortuary display or even lineage affirmation.
-
-```{r imports, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="Aoristic sums of imported ceramics through time across both cemeteries, colour-coded by the broad regional origin of the vessels."}
-
-imports<-subset(pots,pots$import!="")
-greekimports<-subset(imports,imports$import_region=="Greece")
-NEimports<-subset(imports,imports$import_region=="Near East")
-
-import_aorist<-colSums(imports[27:76])
-import_aorist<-data.frame(import_aorist)
-colnames(import_aorist)<-c("aoristic_sum")
-row.names(import_aorist)<-dates
-setDT(import_aorist, keep.rownames = "decade")
-import_aorist$decade<-as.numeric(import_aorist$decade)
-
-greekimport_aorist<-colSums(greekimports[27:76])
-greekimport_aorist<-data.frame(greekimport_aorist)
-colnames(greekimport_aorist)<-c("aoristic_sum")
-row.names(greekimport_aorist)<-dates
-setDT(greekimport_aorist, keep.rownames = "decade")
-greekimport_aorist$decade<-as.numeric(greekimport_aorist$decade)
-
-NEimport_aorist<-colSums(NEimports[27:76])
-NEimport_aorist<-data.frame(NEimport_aorist)
-colnames(NEimport_aorist)<-c("aoristic_sum")
-row.names(NEimport_aorist)<-dates
-setDT(NEimport_aorist, keep.rownames = "decade")
-NEimport_aorist$decade<-as.numeric(NEimport_aorist$decade)
-
-greekimport_aorist$origin<-"Greece"
-NEimport_aorist$origin<-"Near East"
-allimport_aorist<-rbind(greekimport_aorist,NEimport_aorist)
-
-ggplot()+geom_col(data=import_aorist,aes(decade,aoristic_sum),alpha=0.2)+
-          geom_col(data=allimport_aorist,aes(decade,aoristic_sum, fill=origin),position="dodge")+
-            theme_bw()+theme(text=element_text(family="Garamond", size=12))+
-                labs(x="Date (Years BC)", y="Aoristic Sum")+
-                  scale_fill_brewer(name="Origin",palette = "Set2")+
-                   scale_y_continuous(expand=c(0,0),limits=c(0,20))+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))
-
-```
-
-The rise in imported items is clearly seen in the tomb assemblages, with a succession of increased Greek and Near Eastern imports respectively in the late 9th and early 8th centuries (see Figure \@ref(fig:imports)).
-This much has been relatively well understood for some time, at least at the KNC, but we can further examine how these imports were distributed among the various groups of tombs.
-
-```{r importsgroups, fig.dim=c(8,4),dpi=300, fig.align='center',fig.cap="Aoristic sums of imported vessels through time, from tombs built in the 11th through 9th centuries."}
-
-pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A",
-                   ifelse(pots$tomb=="X"|pots$tomb=="G"|pots$tomb=="104"|pots$tomb=="Q"|pots$tomb=="283","B","C"))
-
-imports<-subset(pots,import!="")
-imports_A<-subset(imports,group=="A")
-imports_B<-subset(imports,group=="B")
-imports_C<-subset(imports,group=="C")
-
-import_A_aorist<-colSums(imports_A[27:76])
-import_A_aorist<-data.frame(import_A_aorist)
-colnames(import_A_aorist)<-c("aoristic_sum")
-row.names(import_A_aorist)<-dates
-setDT(import_A_aorist, keep.rownames = "decade")
-import_A_aorist$decade<-as.numeric(import_A_aorist$decade)
-
-import_A_aorist$group<-"A"
-
-import_B_aorist<-colSums(imports_B[27:76])
-import_B_aorist<-data.frame(import_B_aorist)
-colnames(import_B_aorist)<-c("aoristic_sum")
-row.names(import_B_aorist)<-dates
-setDT(import_B_aorist, keep.rownames = "decade")
-import_B_aorist$decade<-as.numeric(import_B_aorist$decade)
-
-import_B_aorist$group<-"B"
-
-import_C_aorist<-colSums(imports_C[27:76])
-import_C_aorist<-data.frame(import_C_aorist)
-colnames(import_C_aorist)<-c("aoristic_sum")
-row.names(import_C_aorist)<-dates
-setDT(import_C_aorist, keep.rownames = "decade")
-import_C_aorist$decade<-as.numeric(import_C_aorist$decade)
-
-import_C_aorist$group<-"C"
-
-aorist_import_groups<-rbind(import_A_aorist,import_B_aorist,import_C_aorist)
-
-ggplot(aorist_import_groups,aes(decade,aoristic_sum,fill=group))+ geom_col(position="dodge")+
-  theme_bw()+theme(legend.position="bottom",text=element_text(family="Garamond", size=12))+
-  labs(x="Date (Years BC)", y="Aoristic Sum")+
-  guides(fill=guide_legend(title="Group"))+
-  scale_fill_brewer(palette = "Set2")+ 
-  scale_y_continuous(expand=c(0,0),limits=c(0,10))+
-  scale_x_reverse(expand = c(0, 0),limits=c(1000,600))+
-  theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-```
-
-In Figure \@ref(fig:importstombcenturies), not only is it apparent that a range of tombs were beneficiaries of the increase in imports in the LPG-EG periods, but also that there was potentially significant patterning in where those artefacts ended up.
-The Group A and B tombs closely trace the trends for Greek and Near Eastern imports respectively (and, indeed, tombs Q and G contained only Greek imports), while across the other tombs, there was also no doubt variegation not observable here.
-Imported ceramics are used here as something of a proxy for items of status value, a shorthand that would ideally be bolstered by other artefacts (such as precious vessels, weapons, jewellery etc.) which, owing to their less developed chronologies, must unfortunately be left out of such diachronic analyses for now.
-But let us suggest, for the time being, that these trends are representative of a larger truth, namely that the late 9th and early 8th centuries saw an unmatched degree in the variety of depositions, which went unmatched in later periods.
-
-```{r bigbox, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="Boxplot for the aoristic sums of all vessels across both cemeteries through time, colour-coded by ceramic period.
-Boxes represent the interquartile range, horizontal lines the median value, whiskers values within 1.5 times the interquartile range, and points outliers beyond this."}
-
-aorist_tomb_values<-tombs[c(1,158:207)]
-aorist_tomb_values[aorist_tomb_values == 0] <- NA
-
-aorist_tomb_values<-data.frame(aorist_tomb_values)
-colnames(aorist_tomb_values)[2:51]<-dates
-
-aorist_melt<-melt(aorist_tomb_values)
-aorist_melt$variable<-as.numeric(as.character(aorist_melt$variable))
-
-aorist_melt$period<-ifelse(aorist_melt$variable<=1100 & aorist_melt$variable>=980,"SM",
-		ifelse(aorist_melt$variable<=970 & aorist_melt$variable>=930,"EPG",
-		ifelse(aorist_melt$variable<=920 & aorist_melt$variable>=880,"MPG",
-		ifelse(aorist_melt$variable<=870 & aorist_melt$variable>=850,"LPG",
-		ifelse(aorist_melt$variable<=840 & aorist_melt$variable>=820,"PGB",
-	  ifelse(aorist_melt$variable<=810 & aorist_melt$variable>=800,"EG",
-	  ifelse(aorist_melt$variable<=790 & aorist_melt$variable>=750,"MG",
-	  ifelse(aorist_melt$variable<=740 & aorist_melt$variable>=720,"LG",
-	  ifelse(aorist_melt$variable<=710 & aorist_melt$variable>=680,"EO",
-	  ifelse(aorist_melt$variable<=670 & aorist_melt$variable>=640,"LO",
-	  ifelse(aorist_melt$variable<=630 & aorist_melt$variable>=600,"A",""
-		)))))))))))
-
-decade_ranges<-seq(from = 600, to = 1100, by = 10)
-decade_ranges<-rev(decade_ranges)
-decade_ranges<-as.factor(decade_ranges)
-aorist_melt$variable<-as.factor(aorist_melt$variable)
-aorist_melt$variable<-factor(aorist_melt$variable, levels=(decade_ranges))
-aorist_melt$period<-factor(aorist_melt$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
-aorist_melt$variable<-as.numeric(as.character(aorist_melt$variable))
-
-ggplot(aorist_melt,aes(variable,value,fill=period))+geom_boxplot(aes(group = cut_width(variable, 10)),outlier.size=0.8)+theme_bw()+
-      theme(text=element_text(family="Garamond", size=12))+
-            labs(x="Date (Years BC)", y="Aoristic Sums per Tomb")+
-               scale_fill_brewer(name="Period",palette = "Set3")+
-                scale_x_reverse(expand=c(0,0))+
-                  scale_y_continuous(expand=c(0,0),limits=c(0,20))
-
-```
-
-To test such an, admittedly tentative, proposition, Figure \@ref(fig:bigbox) presents a boxplot of the aoristic sums of each decade across both cemeteries, colour-coded by ceramic phase.
-Note that, for visual clarity, the outliers tomb G in the EG period, and tomb P in the late EO-LO period are excluded, with values ranging from 29-34, and 36-44 respectively.
-The important observation is that the PGB and EG exhibit the largest interquartile ranges (IQRs) of any period, and each possess only a single outlier (beyond 1.5 times the IQR).
-The Orientalising period, on the other hand, exhibits much reduced IQRs, and a proliferation of outliers far and above the range of the other tombs.
-In other words, whilst both the PGB-EG and EO-LO periods witnessed an expanding volume and range of burial activity, in the earlier period this was distributed more evenly across the total burying population.
-One speaks of increasing plurality, the other of divergence.
-
-```{r gini, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="Gini coefficient values through time based on the aoristic sums for all vessels across both cemeteries.
-Background colours correspond to ceramic periods."}
-
-aorist_tomb_values<-tombs[c(1,158:207)]
-aorist_tomb_values[aorist_tomb_values == 0] <- NA
-
-aorist_tomb_values<-data.frame(aorist_tomb_values)
-colnames(aorist_tomb_values)[2:51]<-dates
-
-tombs_gini<-matrix(0, ncol = 0, nrow = 1)
-tombs_gini<-data.frame(tombs_gini)
-for(i in colnames(aorist_tomb_values[2:51])) {
-  tombs_gini[[paste0(i)]] <- ineq(aorist_tomb_values[[i]],type="Gini")
-}
-
-tombs_gini<-t(tombs_gini)
-tombs_gini<-data.frame(tombs_gini)
-
-colnames(tombs_gini)<-c("gini_coefficient")
-row.names(tombs_gini)<-dates
-tombs_gini<-data.frame(tombs_gini)
-setDT(tombs_gini, keep.rownames = "decade")
-tombs_gini$decade<-as.numeric(tombs_gini$decade)
-
-rects <- data.frame(xstart = c(1100,970,920,875,840,810,790,745,70,670,630), 
-                    xend = c(970,920,875,840,810,790,745,710,670,630,600), 
-                    col = c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
-rects$col<-factor(rects$col,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
-
-ggplot() + 
-  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf, fill = col), alpha = 0.4) +
-     geom_line(data = tombs_gini, aes(decade,gini_coefficient),size=0.5)+
-      theme_bw()+theme(text=element_text(family="Garamond", size=12))+
-              labs(x="Date (Years BC)", y="Gini Coefficient")+
-                 scale_fill_brewer(name="Period",palette = "Set3")+ coord_cartesian(
-                    xlim = c(1100, 600), ylim = c(0.5,0.8))+ scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
-                       scale_y_continuous(expand=c(0,0))
-```
-
-This possibility is supported further by considering the gini coefficient of the distribution of pottery between the tombs (see Figure \@ref(fig:gini).
-Using the aoristic sums as a probablistically-weighted estimate of the volume of material deposited in each tomb in each decade, use of the gini coefficient can give an (albeit rough) estimate of how unequally distributed that material was.
-And, though values were always relatively high (i.e.
-unequal) within the cemeteries, the major spike in the gini coefficient seen in the Orientalising period does contrast markedly with the modest rise and plateau of the PGB-EG.
-The gini coefficient is most commonly used in the assessment of income inequality, but I want to stress here that it is not "wealth" in any meaningful sense that we are talking about here.
-Indeed, if anything, the late flourit of burial activity here discussed may be a case of quantity, and not quality.
-
-```{r comps, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="Plots showing the composition of the total pottery assemblage for each century across both cemeteries, and among the 7 notable tombs drawn out earlier in the analysis."}
-pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A","B")
-
-potcompKNC<-ggplot(subset(pots,century!="" & cem_condensed=="NC"),aes(century,fill=type))+geom_bar(position="fill")+
-  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title = element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(title="KNC",x="", y="Proportion of Total Assemblage")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" Amphora  "," Jug  "," Oil vessel  ", " Pithos"," Drinking vessel  "," Lid  "," Other"))+ 
-                    scale_y_continuous(expand=c(0,0.0))
-potcompFOR<-ggplot(subset(pots,century!="" & cem_condensed=="FOR"),aes(century,fill=type))+geom_bar(position="fill")+
-  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(title="Fortetsa",x="Century BC", y="")+
-                  scale_fill_brewer(palette = "Set2", labels=c(" Amphora  "," Jug  "," Oil vessel  ", " Pithos"," Drinking vessel  "," Lid  "," Other"))+ 
-                    scale_y_continuous(expand=c(0,0.0))
-
-potcomptop7<-ggplot(subset(pots,century!="" & century!="11th"& group=="A"),aes(century,fill=type))+geom_bar(position="fill")+
-  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(title="Group A",x="", y="")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" Amphora  "," Jug  "," Oil vessel  ", " Pithos"," Drinking vessel  "," Lid  "," Other"))+ 
-                    scale_y_continuous(expand=c(0,0.0))
-
-ggarrange(potcompKNC,potcompFOR,potcomptop7,ncol=3,common.legend =TRUE,legend="bottom")
-
-spearmanobjectspots<-cor(tombs$total_object,tombs$total_pottery,method="spearman")
-```
-
-Figure \@ref(fig:comps) shows the composition of the total pottery assemblage from each cemetery, and of the Group A tombs, through the centuries.
-In all cases, there appears to be the steady crystallisation of a basic 'burial kit', comprising a pithos and lid, an oil vessel, a jug, and a drinking vessel.
-The proportion of grave goods comprising 'other' vessels - pyxides, askoi, trays, stands, dinoi, kernoi, kraters, lekanai and more - declines to become only a minor part of the standardised repertoire of the 7th century.
-The seven tombs that underpinned the dramatic increase in burial activity in the later 8th century did not break with this mould, and their assemblages seem as conventional as any other during the last century of use in the cemeteries.
-
-It has no doubt been apparent that the main focus of this article has been on pottery and not other artefacts of metal, stone, glass, and organic materials.
-This is largely because the chronologies for these artefacts are insufficiently developed to allow for the kind of diachronic analysis made thus far.
-But I do want to briefly comment on one aspect of the distribution of non-ceramic artefacts that may make more sense in the light of the foregoing discussion.
-There is, perhaps unsurprisingly, a relatively strong (Spearman's ρ=`r spearmanobjectspots`) correlation between the number of ceramic and non-ceramic finds across the tombs, and, certainly, many of the most unique and precious objects were recovered from among the larger tombs.
-But, the monopoly those tombs would appear to have over such objects, perhaps tempting one initially to speak of 'elite' or 'high-status' tombs, evaporates when the number of those objects is divided by the number of burials represented.
-
-```{r hists, fig.dim=c(6,4),dpi=300, fig.align='center',fig.cap="Histograms relating, first, to counts of non-ceramic artefacts, metal objects, and imported items across all tombs in both cemeteries and, secondly, to those counts divided by the estimated number of burials in each tomb."}
-tombs$group<-factor(tombs$group,levels=c("A","B",))
-total_object<-ggplot(subset(tombs,est_burials>1 & total_object>0),aes(total_object, fill=group))+geom_histogram()+
-  theme_bw()+theme(legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Non-Ceramic Finds", y="Number of Tombs")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
-                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-total_metal<-ggplot(subset(tombs,est_burials>1 & weapon>0),aes(weapon, fill=group))+geom_histogram()+
-  theme_bw()+theme(legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Metal Artefacts", y="Number of Tombs")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
-                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-total_import<-ggplot(subset(tombs,est_burials>1 & total_import>0),aes(total_import, fill=group))+geom_histogram()+
-  theme_bw()+theme(legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Imports", y="Number of Tombs")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
-                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-
-objectsbyburial<-ggplot(subset(tombs,est_burials>1 & total_object>0),aes(total_object/est_burials, fill=group))+geom_histogram()+
-  theme_bw()+theme(legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Non-Ceramics/Burials", y="Number of Tombs")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
-                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-metalbyburial<-ggplot(subset(tombs,est_burials>1 & weapon>0),aes(weapon/est_burials, fill=group))+geom_histogram()+
-  theme_bw()+theme(legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Metal Artefacts/Burials", y="Number of Tombs")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
-                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-
-importsbyburial<-ggplot(subset(tombs,est_burials>1 & total_import>0),aes(total_import/est_burials, fill=group))+geom_histogram()+
-  theme_bw()+theme(legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Imports/Burials", y="Number of Tombs")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" 7 Tombs  ", " Other Tombs"))+
-                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-
-ggarrange(total_import,total_metal,total_object,importsbyburial,metalbyburial,objectsbyburial,ncol=3,nrow=2,common.legend = TRUE,legend="bottom")
-```
-
-Figure \@ref(fig:hists) thus supports the view which has emerged from the foregoing analysis, that the major increase in burial activity at both Fortetsa and the KNC was not a phenomenon linked in any straightforward way to simple principles of 'elite' or 'wealthy' burial.
-To summarise, towards the end of 8th century BC, a select group of tombs, all of which had been constructed back in the 9th or 10th, became the foci of a major increase in burials, something which did not occur among the tombs built around that time.
-We have seen how this surge in acivity occurred against the backdrop of a declining number of older tombs in use, and I have suggested that the rates of change are too rapid, and unevenly distributed, to be explicable through population growth alone.
-We have seen how little about the physical dimensions or early use of these tombs presaged their later popularity, and that, while a similar, if smaller, increase in burial activity occurred around a century earlier, in the PGB-EG periods, this contrasted in the comparative diversity of its associated assemblages, and their distribution across the tombs.
-The later burials were marked by great uniformity in the associated grave goods, and this apparently limited and consistent repertoire may explain why many of the most used tombs prove exceptional only in the absolute quantity of artefacts, and not the accoutrements of each individual burial.
-
-# Discussion
-
-The preceding analysis has revealed a number of potentially significant patterns in the use of the Fortetsa and Knossos North Cemeteries, in particular the exceptional trajectories of a select group of 10th and 9th century tombs.
-It only remains to consider how we whould characterise the burying groups responsible for these observed trends.
-The traditional view is that such groups were immediate families (Brock 1957, 41; Boardman 1967, 63; Coldstream & Huxley 1999; 291).
-The relatively short lifespan of many tombs (often only two or three generations) has been taken as evidence for direct, lineal inheritance, with the tomb going out of use as families moved, individuals founded new tombs, or the line came to an end.
-The idea of lineal groups is supported by the, admittedly circumstantial, evidence of men, women and children being found within single tombs, and one (contentious) instance of an apparently hereditary mandibular deformity in three non-contemporary skulls in the same tomb (Musgrave 1996).
-
-The model of lineal inheritance has been recently challenged by Kotsonas (2011, 131), however, who considers it ‘probably the greatest factoid in the archaeology of Crete of the given period’.
-He argues that the rise in interments during the 8th century, noted also in the present study, is too great to be explained by population growth, as would be necessary for the possibility of direct lineal inheritance to be preserved.
-Certainly, it seems illogical that population increases would lead to the consolidation of burials within only extant tombs, rather than increasing tomb construction.
-The trend instead must reflect choices made by the burying groups themselves, which Kotsonas feels may have comprised various elements, including perhaps age- and sex-grades.
-But it seems probable that such corporate groups, if they existed, intersected with, and operated alongside, real or fictive kinship networks.
-
-Yet it is hard to be conclusive, as our current understanding of the social organisation of the early Cretan *poleis* is patchy at best.
-Perlman and Gagarin's [-@Gagarin2016] recent study of the Cretan law codes reminds us, importantly, if lamentably, that there is an often underappreciated degree of diversity apparent from the Archaic inscriptions in the offices, institutions, and social groups of the early city-states.
-In the traditional view, political authority was concentrated in the hands of the *kosmoi*, aristocratic magistrates selected from the leading, hereditary *startoi*, or ‘clans’, networks of extended lineages controlling the ancestral estates on which worked the serfs and/or slaves who made up the lower rungs of the social hierarchy (Willetts 1982, 240).
-But despite elements of this account probably being accurate, others remain equivocal, such as the nature of the *pyla*, or 'tribe', and its relation to the *startos* (which may itself be a military grouping), the distinction between an *astos* and a *poliatas*, both clearly some kind of formulation of the rights-bearing civic actor, and the ancestry and spread of the *agelai*, apparently an adolescent age-grade so far only attested at Eltynia [@Gagarin2016, 36; 78-80].
-Indeed, many terms which might illuminate social divisions and groupings within the Archaic Cretan populace are only attested at one or two sites, and even then sometimes appear to differ in their form or function.
-Frustrating though this can be, it has also seen the image of Cretan austerity and uniformity begin to be challenged, or at least nuanced, by an appreciation of the regional variety and idiosyncracy apparent from the laws.
-
-It wouldn't be helpful at present, then, to draw any simple parallel between the Knossian burial groups and, say, powerful *startoi*, a term which is not even attested outside of Gortyn in this period (ibid., 1-2).
-But archaeological and textual evidence suggests that, in general terms, such supra-familial kinship groups were undergoing a transition during this period, in ways reflected at Knossos, and thus supporting the attribution of the chamber tombs to such groups.
-Elsewhere on Crete, changes around this time include a shift to unmarked, individual burials, and more austere grave goods, as well as an apparent reorientation of sanctuary deposits away from extra-mural to suburban spaces, and a rise in utilitarian, mass-produced votives (Perlman & Gagarin 2016, 34-6).
-At Knossos, a small number of pithos burials in pits do appear in the KNC and Fortetsa volumes (Brock 1957, 98; Coldstream & Catling 1996, 162), and recent re-excavation of the Teke plot identified numerous small pit burials around three chamber tombs (Evely 2012), raising the possibility that many more such burials filled the interstices between larger chamber tombs elsewhere, but have gone unrecognised.
-The insecure dates of these pits, however, makes it unclear whether they represent a genuine Archaic trend.
-At Azoria, major urban remodelling in the 7th century saw the earlier, agglutinative house forms of neighbouring Kavousi Vronda and Kastro give way to larger, more formally planned dwellings with little evidence of renovation or remodelling, while larger, apprently civic, buildings were constructed to accommodate major public gatherings.
-
-These varied lines of evidence may all suggest that with the emergence of the archaic *poleis* on Crete, there was a concomitant restructuring of wider social groups, institutions, and behaviours.
-Perlman and Gagarin (Gagarin & Perlman 2016, 32-7) argue that, at least by the Classical period, supra-household kinship units such as the *startos* or *pyla*, whatever their exact formulations, had become mostly ceremonial, political entities, and had little influence over the social organisation of the Cretan city states.
-Was this transformation linked to an emerging emphasis on the citizen, the state, and ideologies of relative egality, within which communal gatherings, intra-mural ritual, and individual burial replaced ostentatious, kin-centred acts of gathering, dedication, and burial?
-It is plausible, though as Perlman and Gagarin (32-37) concede, the decline of ostentatious burial could just as well represent a voluntary abandonment by groups assured of their economic and political influence, finding the symbolic expression of kin-relations in death no longer a necessary expense.
-
-These issues, naturally, recall the conclusions of Morris (1987, 171-211), who argued 30 years ago that the emergence of the *polis* across the Aegean led to a tension between elite preferences for conspicuous display, and a downplaying of social difference in death, as promoted by the ‘middling’ ideology that the *polis* engendered and promulgated.
-In many areas, this resulted in increasing restriction of elaborate tombs, despite a ‘rise in aristocratic family consciousness’ that saw certain wealthy tombs buck the trend, until ‘such overt groupings were subsumed within the larger “citizen cemeteries”’ (ibid., 185).
-In his well-known model of the *agathoi* and *kakoi*, Morris argued that, at various times, the relevant spheres of kinship demonstrated in burial could widen and constrict, with more and less distantly related lineages drawn into association through changing social strategies, until the pre-eminence of kinship relations in the structuring of political power declined.
-
-Returning to Knossos, while aspects of the noted expansion in the burying population would seem readily explicable within a model like Morris', certain observations sit more uneasily.
-Firstly, the burials at issue here are in many ways *not* ostentatious in comparison to the norm.
-We have seen that the 'richest' tombs in terms of finds are not so 'rich' after all, when one takes account of the number of burials.
-They are not phyiscally larger, and do not exhibit monopolies over any of a range of potentially 'valuable' offerings.
-Likewise, Morris claims that periods of expansion in the rite of burial will correspond to greater variety in the associated assemblages, yet, at Knossos, we have seen that the range of grave goods in fact diminished and crystallised during the late 8th and early 7th centuries.
-Perhaps the expansion was driven by a widening of access to the burial rite by the so-called *agathoi*, with the standardised burial assemblages coming to reflect their shared affinal relations.
-Or else, the widening of access to these tombs actually heralds the breakdown in traditional kin-based stewardship thereof, perhaps with new, non-lineal corporate groups becoming  the custodians of the tombs, with the limited funereal repertoire thus reflecting the levelling ideology of the emergent state.
-The exclusive focus of the expansion in a very limited number of older tombs, though, would seem to argue in favour of the former.
-What seems certain is that this was a period of transition wherein, whatever their identity, certain actors invested heavily in a final flurry of chamber tomb burials, with an emphasis on tradition, an established repertoire of items, rites, and venues for one's passage to the afterlife.
-
-And this image ties into the broader picture developing of the island at this time.
-It is telling that the peak for imported items in the tombs came in the 9th century; the 'Orientalising' as a time period feels increasingly misapplied to Crete, when, at Knossos at least, it witnessed far lower rates of oriental imports, and only limited adoption of the figural motifs which characterise its appearance elsewhere.
-The observed rise in imitations of Near Eastern vessels in the late 8th and early 7th century further accords with our growing awareness of the selective nature of archaic and classical Crete's external relations.
-Erickson has, contrary to many, argued that *symposia* were practised on Crete, but that their practitioners simply favoured simpler vessel forms.
-The local, selective adoption of, and responses to, wider social and material trends across the contemporary Aegean is perhaps becoming the hallmark of archaic Crete, and finds further justification in the trends identified in the EIA cemeteries at Knossos.
-
-# Conclusion
-
-It is important that the present study was a comparative investigation of both the Fortetsa and Knossos North Cemetery.
-It is notable how similar the observed trends are across both these cemeteries, adding significant credence to the inferences drawn.
-That both underwent similar developments during the EIA suggests not only that they catered to similar sub-sections of the Knossian population, but that some of the socio-economic and political changes inferred and discussed were real, and community-wide.
-We should be cautious in speculating as to whether, for instance, particular families, lineages, groups, or classes made persistent use of these different necropoleis, but such possiblities are not inconceivable.
-Further work should seek to extend these comparisons, both into trends unexplored in the present study (such as, for instance, the curious reuse of a number of 11th century tombs in the late 8th and early 7th centuries), and out to the wider funerary landscape of Knossos.
-I hope to make just such a study but, for the time being, the datasets I am making available alongside this article include the other published EIA tombs in the area, so that others may also explore their possible significance.
-
-There were undeniable limitations to the present study, also.
-For one, non-ceramic artefacts were underrepresented, mostly because they lack accurate dating in the manner of pottery and future work could focus on dating these artefacts more closely, permitting their future inclusion into temporal analyses.
-Issues of gender were sadly also little considered, though this was largely due to limitations in the data themselves.
-There have also been a number of tombs discovered in the last 20 years that fill in the spaces between the Teke plot and the Medical Faculty Site, and ideally these unpublished tombs should be included in further analyses as they, along with the unconsidered Teke tholos and associated tombs, are almost surely part of a single contiguous cemetery (Blackman 1999, 113; Catling 1983, 51; Evely 2012; French 1991, 68–9).
-Finally, a preliminary study was recently taken by the author of select boxes marked as 'unpublished' from the KNC deposits kept at the British School at Athens at Knossos.
-Despite going unmentioned in the publication, this material was, at least in the boxes seen, not inconsiderable and so, regretably, future study will be necessary to determine whether it threatens the integrity of quantitative analyses such as this.
-Suffice to say that, at present, it is another pinch of salt with which to take the findings.
-
-These limitations notwithstanding, the foregoing analysis demonstrates the value of quantitative approaches to the archaeological record of the period.
-Such methods allow us not only to identify types of variability in the archaeological material, but also assess the scale of that variability.
-With large datasets, such as those used in the present study, we are able to locate and visualise covariance and significant correlations between different variables, plot changes in the data through time and, ultimately, draw on our findings to inform discussions of social processes.
-Though the construction of databases can be time consuming, once made they offer the opportunity to subdivide, compare, and plot the data in numerous ways quickly and easily, facilitating investigation of the material from multiple angles.
-Such methods are also more independently verifiable than conclusions drawn from more qualitative analyses, for two main reasons.
-Firstly, though explanations for any patterning in the data still require interpretation, the existence of such patterns may be more easily and reliably assessed via the use of quantitative and statistical tests of correlation and covariance.
-And secondly, by making datasets available for use by others, as with the data and code for the present study, one’s conclusions can be independently tested, and alternative interpretations can be forwarded, by other scholars.
-It is hoped that the present study demonstrates the wealth of insights yet to be uncovered by the application of more systematic, quantitative methods.
-The databases produced for the above analysis contain a large of amount of unutilised data that could facilitate a range of future investigations, and all are welcome to it.
-With more such analyses, we could begin to a construct a fuller account of the regional diversity in mortuary, and wider social, practices that characterised the communities of EIA and archaic Crete.
-
-# Acknowledgements
-
-<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
-##### pagebreak
-
-# References 
-<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
-<div id="refs"></div>
-
-##### pagebreak
+*Table 1 The number of pithoi (as a proxy for burials) deposited in tombs dating to the 10th, 9th, 8th and 7th centuries, through each of the centuries they were in use, as well as the number of tombs from which these burials came, allowing calculation of the changing rate of burial through time.*
+
+#####pagebreak
 

### 2019-05-30 R/Samples.R

@@ -379,3 +379,74 @@ pots$polychrome <- ifelse(grepl("olychrome",pots$additional), "Polychrome", "Oth
 count(groupB$polychrome)
 122/811*100
 66/561*100
+
+pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A",
+                   ifelse(pots$tomb=="X"|pots$tomb=="G"|pots$tomb=="104"|pots$tomb=="Q"|pots$tomb=="283","B","C"))
+
+imports<-subset(pots,import!="")
+imports_A<-subset(imports,group=="A")
+imports_B<-subset(imports,group=="B")
+imports_C<-subset(imports,group=="C")
+
+import_A_aorist<-colSums(imports_A[27:76])
+import_A_aorist<-data.frame(import_A_aorist)
+colnames(import_A_aorist)<-c("aoristic_sum")
+row.names(import_A_aorist)<-dates
+setDT(import_A_aorist, keep.rownames = "decade")
+import_A_aorist$decade<-as.numeric(import_A_aorist$decade)
+
+import_A_aorist$group<-"A"
+import_A_aorist$aoristic_sum<-import_A_aorist$aoristic_sum/7
+
+import_B_aorist<-colSums(imports_B[27:76])
+import_B_aorist<-data.frame(import_B_aorist)
+colnames(import_B_aorist)<-c("aoristic_sum")
+row.names(import_B_aorist)<-dates
+setDT(import_B_aorist, keep.rownames = "decade")
+import_B_aorist$decade<-as.numeric(import_B_aorist$decade)
+
+import_B_aorist$group<-"B"
+import_B_aorist$aoristic_sum<-import_B_aorist$aoristic_sum/5
+
+import_C_aorist<-colSums(imports_C[27:76])
+import_C_aorist<-data.frame(import_C_aorist)
+colnames(import_C_aorist)<-c("aoristic_sum")
+row.names(import_C_aorist)<-dates
+setDT(import_C_aorist, keep.rownames = "decade")
+import_C_aorist$decade<-as.numeric(import_C_aorist$decade)
+
+import_C_aorist$group<-"C"
+import_C_aorist$aoristic_sum<-import_C_aorist$aoristic_sum/35
+
+imports_C$tomb
+
[…]

### 2019-06-03 analysis/paper/paper2.Rmd
Re-writes

@@
 # Introduction
-Three principal interests converge in this paper: a reorientation of the focus of Early Iron Age Cretan archaeology; the application of computational, and quantitative methods to supplement traditional modes of archaeological inference; and open-access data and reproducible research.
-To begin with the first of these, it may be felt that, of all ancient sites on Crete, Knossos is the last in need of further archaeological attention.
+Three principal interests converge in this paper: a reorientation of the focus of Early Iron Age Cretan archaeology; the application of computational, and quantitative methods to supplement traditional modes of archaeological inference; and open-access, reproducible research.
+To begin with the first, it may be felt that, of all ancient sites on Crete, Knossos is the last in need of further archaeological attention.
 The city and palace of Knossos, first excavated well over a century ago [@Evans1921; -@Evans1928; -@Evans1930], retain a pre-eminent position in popular and academic accounts the island’s Bronze Age (BA), and few, if any, other Cretan sites have been so thoroughly dug, reconstructed, and contested.
@@
 For one, though, this is not the Knossos of that illustrous, palatial period, but rather the town as it persisted and evolved during the subsequent Early Iron Age (EIA).
-Research into this once maligned period has grown apace in the last few decades, beginning with the pioneering works of Coldstream [-@Coldstream1977], Desborough [-@Desborough1972], and Snodgrass [-@Snodgrass1971a] and culminating in the rejection by most scholars of the appelation and interpretive bagguage of the so-called ‘Dark Ages’ [on the origins of which, see @Kotsonas2016; @Morris1997].
+Research into this once maligned period has grown apace in recent decades, beginning with the pioneering works of Coldstream [-@Coldstream1977], Desborough [-@Desborough1972], and Snodgrass [-@Snodgrass1971a] and culminating in the rejection by most scholars of the appelation and interpretive bagguage of the so-called ‘Dark Ages’ [see @Kotsonas2016; @Morris1997].
 Nonetheless, our knowledge of many EIA settlements remains often underdeveloped in comparison with those of the second millennium BC.
 
-Secondly, and welcome though they are, a number of recent publications on the socio-economic [@Wallace2010a] and settlement [@Nowicki2000] changes which accompanied the BA-EIA transition have directed scholarly attention toward a number of elevated, defensible settlements supposedly founded in response to the dangers of this turbulent period.
+Secondly, and welcome though they are, a number of recent publications on the socio-economic [@Wallace2010a] and settlement [@Nowicki2000] changes which accompanied the BA-EIA transition have directed scholarly attention toward a pattern of elevated, defensible settlements supposedly founded in response to the dangers of this turbulent period.
 This is reflected in survey work, which has focussed extensively around the Bay of Mirabello, where such sites are numerous [@Haggis2005; @Hayden2005; @Watrous2012].
 This has, unfortunately, engendered a certain disregard for settlement characterised by different topographies, especially in the centre of the island.
-There remains, therefore, a need for studies addressing the social dynamics at such sites (Knossos, Phaistos, Grivila, Eleutherna etc.), with a long-term view to comparative appraisal of the divergent trajectories of communities across Crete in the EIA [@Kotsonas2011b; @Whitley2011, 667–8].
+There remains, therefore, a need for studies addressing the social dynamics at such sites (Knossos, Phaistos, Grivila, Eleutherna etc.), with a long-term view to comparative appraisal of the divergent trajectories of communities across Crete in the EIA [@Kotsonas2011b; @Whitley2011, 667–68].
 
@@
 The cemeteries of Knossos offer an opportunity to introduce such approaches into EIA research on Crete, where they have yet to make much of an impact [see @Kotsonas2011].
-The Fortesta and North Cemetery complexes are both well-published [@Brock1957; @Coldstream1996a] and have been studied with regard to their 'oriental' imports [@Antoniadis2012], antique artefacts [@Crowe2016], and religious context [@Coldstream1984b].
-But a lack of quantitative analyses, or systematic comparisons between the cemeteries, has left us with many intuited trends, some exceptional tombs, but a general impression of un-patterned multiplicity.
-
-With a dataset including information on every find and tomb recorded in the Fortetsa and Knossos North Cemetery publications [@Brock1957; @Coldstream1996a], the following analysis seeks to address this present lack.
-In doing so, I aim to demonstrate the potential for quantitative methods, firstly, to permit identification, and nuanced characterisation, of significant patterning in the archaeological material; secondly, to facilitate a more systematic appraisal of the cemeteries’ largest tombs and their distinctive histories; and, finally, to offer not just generalities, but tangible and context-specific considerations of the motivations and identities behind the burial groups themselves, with a view to eventually comparing such results with mortuary assemblages across the island.
+The Fortesta and North Cemetery complexes are both well-published [@Brock1957; @Coldstream1996a] and studied, but a lack of systematic, quantitative comparisons between the cemeteries has left us with many intuited trends, some exceptional tombs, yet a general impression of un-patterned multiplicity.
+
+With a dataset including information on every find and tomb recorded in the Fortetsa and Knossos North Cemetery publications [@Brock1957; @Coldstream1996a], the following analysis addresses this present lack.
+In doing so, I aim to demonstrate the potential for quantitative methods, firstly, to permit identification, and nuanced characterisation, of significant patterning in the archaeological material; secondly, to facilitate a more systematic appraisal of the cemeteries’ largest tombs and their distinctive histories; and, finally, to offer tangible and context-specific evidence as to the behaviours and identities of the burying groups, with a view to eventually comparing such results with mortuary assemblages across Crete.
 
@@
 Settlement evidence at EIA Knossos is, in a word, insubstantial.
-Speaking of the town’s stratigraphic sequence, Coldstream lamented that “the remains of [this] period are like a meagre filling in a very thick sandwich” between Bronze Age and later Greek and Roman levels [@Coldstream1991, 287].
-Most individual finds of Protogeometric or Geometric date take the form of flimsy foundations, patches of earth floor, wells or simple scatters of pottery, and for a time it was argued that the area in the EIA comprised several dispersed villages, which later coalesced into a single city, as per Aristotle’s model of synoecism [@1950].
-
-However, a concentration of activity slightly to the west of the old palace presented a good case for continuing, nucleated settlement [@Hood1981], something strongly advocated by Coldstream [-@Coldstream1984a; -@Coldstream2000], while, more recently, findings of the Knossos Urban Landscape Project suggest a PG settlement of up to 40ha [@Kotsonas2011a, 5-8].
-This estimate remains our best guess in the absence of substantial excavated deposits but, if it is accurate, then EIA Knossos was one of the largest sites of the contemporary Aegean.
+Speaking of the town’s stratigraphic sequence, Coldstream lamented that “the remains of [this] period are like a meagre filling in a very thick sandwich” between Bronze Age and later Greek and Roman levels [@Coldstream1991: 287].
+Most individual finds of Protogeometric (PG) or Geometric date take the form of flimsy foundations, patches of earth floor, wells or pottery scatters, and for a time it was argued that the area in the EIA comprised several dispersed villages, which later coalesced into a single city, as per Aristotle’s model of synoecism [@1950].
+
+However, a concentration of activity slightly to the west of the palace presented a good case for continuing, nucleated settlement [@Hood1981], something strongly advocated by Coldstream [-@Coldstream1984a; -@Coldstream2000], and, more recently, findings of the Knossos Urban Landscape Project suggest a PG settlement of up to 40ha [@Kotsonas2011a: 5-8].
+This estimate remains our best guess in the absence of substantial excavated deposits but, if accurate, suggests that EIA Knossos was one of the largest sites of the contemporary Aegean.
 
@@
 
-Survey aside, our most abundant evidence for EIA Knossos comes from the mortuary record (see Figure \@ref(fig:map).
-The transitional period between the Late Bronze Age (LBA) and EIA yields few if any securely datable interments, but later evidence is more plentiful, with multiple burials known from the Kephala ridge [@Coldstream1963, 38; @Hogarth1899, 82-5], the modern suburbs of Ayios Ioannis and Atsalenio [@Boardman1960a; @Davaras1968, 133–46], the cave of Mavro Spelio [@Antoniadis2012, 58-9], and the Khaniale Teke site with its well-known tholos [@Boardman1967;@Hutchinson1954; cf.
+Survey aside, our most abundant evidence comes from the mortuary record (see Figure \@ref(fig:map)).
+The transitional period between the Late Bronze Age (LBA) and EIA yields few if any securely datable interments, but later evidence is more plentiful, with burials known from the Kephala ridge [@Coldstream1963: 38; @Hogarth1899: 82-5], the modern suburbs of Ayios Ioannis and Atsalenio [@Boardman1960a; @Davaras1968: 133–46], the cave of Mavro Spelio [@Antoniadis2012: 58-9], and the Khaniale Teke site with its well-known tholos [@Boardman1967;@Hutchinson1954; cf.
 @Hoffman1997 @Kotsonas2006].
-Taken together, these scattered tombs point to a richly variegated set of funerary practices but, despite their reasonable number, are so dispersed and often isolated that they provide little scope for systematic quantitative analysis.
+Taken together, these scattered tombs point to a richly variegated set of funerary practices but, despite their reasonable number, are so dispersed that they provide little scope for systematic quantitative analysis.
 
 Fortunately, far larger burial accumulations are to be found in the Fortetsa and the Knossos North cemeteries, which together comprise some 134 tombs, ranging in date from the Subminoan (SM) to Late Orienatalising (LO) periods.
-Though many were robbed in antiquity, and others destroyed by the digging of later burials, building work, or just the ravages of time, they still represent the most abundant source of evidence for not just mortuary practices, but for social life of any kind at EIA Knossos.
+Though many were robbed in antiquity, and others destroyed by the digging of later burials, building work, or just the ravages of time, they still represent the most abundant source of evidence for not just mortuary, but social practice of any kind at EIA Knossos.
 
@@
 
-The cemetery on the slope facing the eponymous village of Fortetsa is composed of twelve tombs excavated in 1933, and another eight uncovered in 1935 (Brock 1957: xi).
-These tombs cluster in three main groups.
-The largest of these, comprising nine tombs, is arranged on a north-south axis, with the *dromos* of each tomb extending downslope to the west.
+The cemetery on the slope facing the eponymous village of Fortetsa comprises twelve tombs excavated in 1933, and another eight uncovered in 1935 (Brock 1957: xi).
+These cluster in three main groups.
+The largest of these contains nine tombs, arranged on a north-south axis, with the *dromos* of each extending downslope to the west.
 To the north lies a cluster of five tombs and, to the south, another of three.
@@
 
-To the northeast of the Fortetsa cemetery, a collection of ten badly damaged tombs known, after the year of their excavation, as the Fortetsa 1967 tombs, were probably the southernmost burials of the KNC.
-Unfortunately, all had been looted at the time of excavation, and no *in situ* deposit was found in any tomb.
-They are thus excluded from present analysis.
+To the northeast of the Fortetsa cemetery, a collection of ten badly damaged tombs known as the Fortetsa 1967 tombs were probably the southernmost burials of the KNC.
+Unfortunately, all had been thoroughly looted at the time of excavation, and are thus excluded from present analysis.
 Northwest of these, and just south of the Teke Tholos, the Teke tombs represent the northern reaches of the KNC.
-The central section of the KNC comprises the ‘Medical Faculty’ site, named for the extension of the University of Crete that prompted major rescue excavations here in 1978 [@Coldstream1996a: 53-55].
+The central section of the KNC comprises the ‘Medical Faculty’ site, named for the extension of the University of Crete that prompted major rescue excavations there in 1978 [@Coldstream1996a: 53-55].
 
@@
 It comprises a descending ramp or passage, the *dromos*, leading to the chamber itself, cut into the *kouskouras* bedrock.
-The *dromos* and chamber are separated by the *stomion*, an opening often set a step or two lower than the *dromos*, which is covered by a large stone slab, smaller stacked stones, or combination of both [@Antoniadis2012: 47-8].
+The *dromos* and chamber are separated by the *stomion*, an opening often set a step or two lower than the *dromos*, which is covered by a large stone slab, smaller stacked stones, or combination of both [@Antoniadis2012: 47-48].
 The majority of burials are cremations, a rite that appears at Knossos in the Subminoan period, in the Tomb 200+ complex in the KNC [@Coldstream1994: 109].
-Inhumation does not completely disappear in the EIA, but by the Orientalising period it appears restricted to child burials and [@Antoniadis2012: 69].
+Inhumation does not completely disappear in the EIA, but by the Orientalising period it appears restricted solely to child burials [@Antoniadis2012: 69].
 
@@
 
-The initial foundation of the cemeteries has received much attention.
-It has been claimed that no tomb crosses the Late Minoan (LM) IIIC-SM divide [@Coldstream1998, 58; @Coldstream2006, 582], though there is a possibility that this is a classificatory artefact [@Hallager2010].
+The foundation of the cemeteries has received much attention.
+It has been claimed that no tomb crosses the Late Minoan IIIC-SM divide [@Coldstream1998: 58; @Coldstream2006: 582], though there is a possibility that this is a classificatory artefact [@Hallager2010].
 Instances of BA tombs reused in the EIA are known (e.g.
@@
 During the use of the cemeteries, correspondences have been highlighted between changing pottery styles and episodes of tomb construction and abandonment.
-The Protogeometric B (PGB) style, first defined by Brock [-@Brock1957, 143], who hailed it as “the most remarkable phase in Cretan vase-painting”, is apparently a Cretan anomaly, evincing a melding of Attic geometric motifs, patterns possibly deriving from Near Eastern metalwork, and iconography drawn from the Minoan repertoire [@Coldstream1984b, 93-94].
+The Protogeometric B (PGB) style, first defined by Brock [-@Brock1957: 143], who hailed it as “the most remarkable phase in Cretan vase-painting”, is apparently a Cretan anomaly, evincing a melding of Attic geometric motifs, patterns possibly deriving from Near Eastern metalwork, and iconography drawn from the Minoan repertoire [@Coldstream1984b: 93-94].
 Coldstream has argued that the concurrence of this style with a rash of newly founded tombs points to a reformulation of elite mortuary display, noting how, in this period also, redeposited Minoan larnakes began appearing in and around some tombs, in one case even inspiring the decoration of a PGB vessel [in T.
-107; see @Coldstream1984b; -@Coldstream1994, 112–13].
-Whether simply nostalgia [@Coldstream1998, 60], or a more tactical strategy by newly emerging elite groups [@Coldstream1994, 114-15], it seems that the Bronze Age past possessed a significant cultural cachet mobilised by various EIA burying groups.
-
-The final mystery of the chronological sequence at the KNC and Fortetsa is the quite abrupt cessation of burials around 630 BC.
-There had been changes in tomb use in both cemeteries in the preceding century, with few new tombs built, though interments in old tombs continued [@Brock1957, 4; @Cavanagh1996; 651-3].
-This curiosity has, along with the equally scanty evidence for the subsequent period in the wider settlement, become known as the ‘Archaic Gap'.
+107; see @Coldstream1984b; -@Coldstream1994: 112–13].
+Whether simply nostalgia [@Coldstream1998: 60], or a more tactical strategy by newly emerging elite groups [@Coldstream1994: 114-15], the Bronze Age past clearly possessed a significant and enduring cultural cachet for the EIA Knossians.
+
+As contentious as their founding is the abrupt abandonment of both cemeteries around 630 BC.
+Admittedly, few new tombs had been built in the preceding century, though interments in old tombs had certainly continued [@Brock1957: 4; @Cavanagh1996: 651-53].
+The ensuing period, accompanied by equally scanty settlement evidence, has become known as the ‘Archaic Gap'.
 The causes of this lacuna, which persists until the recrudescence of archaeological evidence in the Late Archaic period, remain unknown.
-Although various explanations have been forwarded, [@Coldstream1999, 301-2; @Huxley1994, 126; @Kotsonas2002, 41-4], there are now strong reasons to doubt the settlement was abandoned (Whitelaw, pers comm.).
+Although various explanations have been forwarded, [@Coldstream1999: 301-302; @Huxley1994: 126; @Kotsonas2002: 41-44], there are now strong reasons to doubt the settlement was abandoned (Whitelaw, pers comm.).
 
@@
 
-A popular approach to the study of Knossos’ EIA tombs has been the investigation of ‘object biography’, a term derived from Appadurai [-@Appadurai1986] and Kopytoff [-@Kopytoff1986].
-Influenced by Mauss’ [-@Mauss1954] formulation of ‘the gift’, they stressed the mutuable nature of an object’s value, and its capacity for acquiring a ‘biography’ through the transformations of its social existence.
-These ideas, fertilised by the Homeric motif of gift-giving [@Whitley2002, 220-1], and the remarkable archaeological parallels for such objects – such as the boar’s tusk helmet from Tomb 200+ and that given to Odysseus by Meriones (*Iliad* 10.260-271) - have been brought to bear on certain elaborate burials of the EIA.
-
-Catling’s discussion of the Tomb 200+ complex is a notable example.
-Tomb 201 contained the remains of at least two adults, and possibly a child, a bronze sword, spearhead, and arrowhead, an iron dirk or knife, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned helmet [@Catling1995, 123].
-Catling identifies similarities between this tomb and burials at Tiryns, Kaloriziki, and the famous *hērōön* of Lefkandi.
-All contained imported items, were of ‘warrior grave’ type, and formed loci for later burials.
-Catling argues that those individuals who dared to travel abroad in this time of more limited seafaring, accruing exotica and stories from distant lands, would have been accorded exceptional treatment in death [@Catling1995, 127–8].
+A popular approach to the study of Knossos’ EIA tombs has been the investigation of ‘object biography’ [@Appadurai1986; @Kopytoff1986l; @Mauss1954], an anthropological lens emphasising the mutuable nature of an object’s value, the transformations latent within its social existence.
+Fertilised by the Homeric motif of gift-giving [@Whitley2002: 220-21], and the remarkable archaeological parallels for such items – such as the boar’s tusk helmet from Tomb 200+ and that given to Odysseus by Meriones (*Iliad* 10.260-271) - such approaches have been brought to bear on certain elaborate burials of the EIA.
+
+For instance, Tomb 201 in the KNC contained a bronze sword, spearhead, and arrowhead, an iron dirk or knife, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned helmet [@Catling1995: 123].
+Catling identifies similarities between this tomb and burials at Tiryns, Kaloriziki, and the famous *hērōön* of Lefkandi; all contained imported items, were of ‘warrior grave’ type, and formed loci for later burials.
+He has argued that individuals who dared to travel abroad in this time of more limited seafaring, accruing exotica and stories from distant lands, would have been accorded exceptional treatment, including the deposition of such items, in death [@Catling1995: 127–8].
 
@@
 Two caches of gold jewellery, found in pits flanking the entrance to the Teke Tholos, were argued by Boardman [-@Boardman1967] to be foundation deposits – a Near Eastern custom – for the tomb of an immigrant Phoenician craftsman.
-Kotsonas has revisited this attribution, deeming the tholos more probably that of an elite individual or family with a monopoly over a metal workshop’s output [-@Kotsonas2006, 155–9].
-While certainty in such individual cases may be moot, it seems unlikely that foreign traders or craftsmen never visited or resided at Knossos during the EIA [@Hoffman1997, 176–85; @Schreiber2003, 293-306].
+Kotsonas has revisited this attribution, deeming the tholos more probably that of an elite individual or family with a monopoly over a metal workshop’s output [-@Kotsonas2006: 155–9].
+While certainty in such individual cases may be moot, it seems unlikely that foreign traders or craftsmen never visited or resided at Knossos during the EIA [@Hoffman1997: 176–85; @Schreiber2003: 293-306].
 
@@
 Generally speaking, the above analyses have tended to rely more on particular finds, tombs, or observable, but limited, patterns in the data; only a few authors have pursued more quantative approaches.
-Cluster analyses have been employed by both Cavanagh [-@Coldstream1996, 653-7] and Antoniadis [-@Antoniadis2012, 193-7], the most sophisticated statistical techniques so far applied to the material.
-Antoniadis [-@Antoniadis2012, 172-6] also divided the EIA tombs of Knossos into three groups, those with fewer than 10 pots, fewer than 50, and more than 50, and found that, in each successive case, the proportions of tombs containing 'oriental' imports and imitations increased, suggesting unequal access to such prestigious goods.
-Yet even here the available data could be further interrogated, such as with the number of imported items *per burial*, something Antoniadis does not explore.
+Cluster analyses have been employed by both Cavanagh [-@Coldstream1996: 653-57] and Antoniadis [-@Antoniadis2012: 193-97], the most sophisticated statistical techniques so far applied to the material.
+Antoniadis [-@Antoniadis2012: 172-6] also divided the EIA tombs of Knossos into three groups, those with fewer than 10 pots, fewer than 50, and more than 50, and found that, in each successive case, the proportions of tombs containing 'oriental' imports and imitations increased, suggesting unequal access to such prestigious goods.
+Yet even here the available data could be further interrogated (with, for instance, the number of imported items *per burial*).
 The present paper thus seeks to extend and nuance a number of observations that have been made in previous studies, as well as revealing patterns thus far unacknowledged, through use of a more systematic approach to quantitative analysis.
@@
 The first includes an entry for each individual ceramic vessel.
-Each entry includes, among other things, the tomb, the vessel shape, its surviving dimensions, its ceramic period and corresponding absolute dating measures.
-In addition, all imports, 'oriental' imitations, and antique (that is, Bronze Age) artefacts were accordingly categorised, using the studies of Antoniadis (2012), Jones [-@Jones2000], and Crowe (2016).
+Variables includes the tomb, the vessel shape, its surviving dimensions, its ceramic period and corresponding absolute dating measures.
+In addition, all imports, 'oriental' imitations, and antique (that is, Bronze Age) artefacts were categorised, using the studies of Antoniadis (2012), Jones [-@Jones2000], and Crowe [-@Crowe2016].
 In total, this database comprises just over 4540 individual artefacts.
@@
 
-Taken together, these databases constitute a detailed record of all the finds from the Fortetsa and KNC publications, and offer the opportunity for carrying out a range of quantitative analyses.
-It must be conceded that, with many of the tombs evidently looted or damaged over the centuries, this cannot be a complete picture of the tombs’ original contents.
+Taken together, these databases make possible a diverse range of quantitative analyses.
+It must be conceded that, with many of the tombs evidently looted or damaged, this cannot be a complete picture of their original contents.
 If we assume, though, that natural or accidental destruction was relatively indiscriminate, and that, even in cases of looting, certain object classes (particularly cinerary urns) were seldom robbed, then the data are by no means rendered useless.
-As Snodgrass [-@Snodgrass1996] notes, the situation means we should be extremely cautious about making negative arguments – that is, conclusions drawn about the lack of features in certain tombs – but may be somewhat more hopeful in the forming of positive ones.
+As Snodgrass [-@Snodgrass1996] cautions, we should be extremely cautious about making negative arguments – that is, conclusions drawn about the *lack* of features in certain tombs – but may be somewhat more hopeful in the forming of positive ones.
 
@@
 
-Evermore archaeologists have begun to advocate the utility, indeed the necessity, of making available not just the data underpinning our academic output, but the full methodologies by which our analyses are conducted, in formats facilitating reproducibilty.
-It is felt that, with the opportunities afforded by current technologies (digital databases, statistical software packages, online repositories etc.), there remain ever fewer obstacles to the full and free sharing of both data and workflows [see @article].
-In Marwick's words [-@Marwick2017, 445], "[t]he technical problems are largely solved; the challenge now is to change the norms of the discipline to make high reproducibility a canonical attribute of high-quality scholarly work".
+Ever more archaeologists have begun to advocate the utility, indeed the necessity, of making available not just the data, but the full methodologies underpinning our academic output.
+With the opportunities afforded by current technologies (digital databases, statistical software packages, online repositories etc.), there remain ever fewer obstacles to the full and free sharing of both data and workflows [see @article].
+In Marwick's words [-@Marwick2017: 445], "[t]he technical problems are largely solved; the challenge now is to change the norms of the discipline to make high reproducibility a canonical attribute of high-quality scholarly work".
 
@@
 
-There are, of course, complications to the sharing of data and methods.
-There are valid reservations around issues of copyright, the protection of sites from looting, the possibility of being scooped with one's own data, or simply the significant time investment of preparing open-access research.
-But I would argue that the potential gains outweigh these concerns, and that research on the EIA would benefit greatly from the greater accessability of datasets and workflows.
-
-In this vein, the present article has been composed making use of Ben Marwick's "rrtools" [see @Marwick2017; @inbook], an open-source package for the statistical software R Studio, which provides the author with the tools to create a self-contained "compendium", which can be freely downloaded, and from which anyone can accurately reproduce the entire workflow of the present analysis.
-Though not visible in the present paper, every chart and table herein was produced directly, through written code, from the raw data using R.
-This code, as well as the raw and derived datasets, complete bibliography (in BibTex format, for export to any major reference manager), figures, and article as originally written are included in the compendium.
+There are, of course, valid reservations regarding issues of copyright, the protection of sites from looting, the possibility of being scooped with one's own data, or simply the significant time investment of preparing open-access research.
+But I would argue that the potential gains outweigh these concerns, and that research on the EIA would benefit hugely from the greater accessability of datasets and workflows.
+
+In this vein, the present article has been composed making use of Ben Marwick's 'rrtools' [see @Marwick2017; @inbook], an open-source package for the statistical software R Studio.
+This provides the author with the tools to create a self-contained 'compendium', which can be freely downloaded, and from which anyone can reproduce the entire workflow of the following analysis.
+Though not visible in the present paper, every chart and table herein was produced directly, through written code, from the raw data using R (and note also that, wherever possible, these are presented in colour-blind friendly palettes).
+This code, as well as the raw and derived datasets, complete bibliography (in BibTex format, for export to any major reference manager), figures, and original article are included in the compendium.
 It can be downloaded as a repository from GitHub, at the following address: https://github.com/DCPollard94/knossoscemeteries, and the raw datasets have been stored with the online repository Figshare, here: (**NB a URL will be included pending acceptance of the article**).
@@
 A first important step in preparing the data for analysis was the estimation of the number of individuals buried in each tomb.
-The method chosen broadly follows that of Cavanagh [-@Cavanagh1996, 659-660], who took the number of pithoi (the most common urn type) and the number of pithoi, amphorae and kraters (which were sometimes used as urns) as middle and upper estimates of the number of interments, with all securely identified interments (in the form of ashes or osteological material) being the lower limit (Figure \@ref(fig:burialestimate)).
-In the present study, the mean of these three estimates was taken, to provide a single aggregate measure for the purposes of analysis.
+The method chosen broadly follows that of Cavanagh [-@Cavanagh1996: 659-60], who took the number of pithoi (the most common urn) and the number of pithoi, amphorae and kraters (which were sometimes used as such) as middle and upper estimates of the number of burials, with all securely identified interments being the lower limit (Figure \@ref(fig:burialestimate)).
+Here, the mean of these three estimates was taken, to provide a single aggregate measure for the purposes of analysis.
 
@@
 
-A central concern of the present study was how to deal with the temporal uncertainty implicit in the assigning of individual vessels to stylistically-defined periods.
-A vessel dated to the PGB period could, on our best estimates, date anywhere between 840 and 810 BC.
-This situation results in major challenges to the visualisation and analysis of such data through means which normally require values and not ranges.
+Another major concern was the temporal uncertainty implicit in the assigning of individual vessels to ceramic periods.
+A Protogeometric B vessel could, on our best estimates, date anywhere between 840 and 810 BC.
+This situation results in major challenges to the visualisation and analysis of such data.
 A simple, if coarse, solution is to use the midpoint of the range assigned to each vessel (i.e.
 825 for our PGB example).
-This is used in a couple of instances in the present paper when the division of vessels into centuries renders the inaccuracies of such a method less significant.
+This is used in a couple of instances in the present paper, for assigning vessels to broad, century-long windows.
 
@@
 
-But the artificial clustering this technique engenders around select absolute dates can prove misleading, as in Figure \@ref(fig:allpithoi), a simple plot of vessel counts based on these average dates; the apparent cessation of depositions for 30 years in the 8th century is simply a fiction generated by the processing of the data.
-
-A alternative way of dealing with temporal uncertainty, and the method employed extensively in this paper, is the use of so-called "aoristic analysis" [for original use in criminology, see @Ratcliffe2000; for archaeological applications, see @Crema2010; @Crema2012; @Johnson2004a].
-Here, rather than assign a pot a single date, the entire time span under investigation is divided into equal chunks (in this case, decades), and probability values summing to 1 are then assigned for each vessel, per chunk of time, based on its accordant date range.
+But the artificial clustering this technique engenders around select absolute dates can prove misleading, as in Figure \@ref(fig:allpithoi); the apparent cessation of depositions for 30 years in the 8th century is simply a fiction generated by the processing of the data.
+
+A alternative approach, and the method employed extensively in this paper, is the use of so-called 'aoristic analysis' [for original use in criminology, see @Ratcliffe2000; for archaeological applications, see @Crema2010; @Crema2012; @Johnson2004a].
+Here, rather than give a pot a single date, the entire time span under investigation is divided into equal chunks (in this case, decades), and probability values summing to one are then assigned for each vessel, per chunk of time, based on its accordant date range.
 So, our PGB vessel, datable between 840 and 810 BC, would register a value of 0.33 for each of those three decades.
 These values for individual vessels can then simply be summed to produce tomb- or cemetery-level signatures.
-The major contribution of this method is that it incorporates temporal uncertainty into the size of the resulting values, that is, "events with tight temporal definition contribute more to the total probability over their range than do loosely defined events" [@Johnson2004a, 450].
+A particular attraction of this method is that it incorporates temporal uncertainty into the size of the resulting values, that is, "events with tight temporal definition contribute more to the total probability over their range than do loosely defined events" [@Johnson2004a, 450].
 
@@
 As seen in Figure \@ref(fig:aoristallcems), the artificial peaks and troughs of the simple averaging method are avoided, and different patterns become visible.
-Though these methods present their own limitations (see @Crema2010, 1123-1124), they are considered a more robust and productive approach to the temporal uncertainty of ceramic dating and, as such, are used for much of the present analysis.
+Though these methods present their own limitations (see @Crema2010, 1123-24), they are considered a more robust and productive approach to the temporal uncertainty of ceramic dating and, as such, are used for much of the present analysis.
 
@@
 ##Introduction
-The great advantage of the large, quantitative data set employed here is that it permits investigation of several nested levels of potentially significant patterning.
-From individual burials, to collective tombs, to the cemeteries as a whole, and between cemeteries in the landscape, we have the opportunity to study and compare both the synchronic and diachronic variation across these various levels.
+The great advantage of the large, quantitative dataset employed here is that it permits investigation of several nested levels of potentially significant patterning.
+From individual burials, to collective tombs, to whole cemeteries, we have the opportunity to study and compare both the synchronic and diachronic variation across these various levels.
 In so doing, I first investigate the broad temporal trends in tomb construction, burial, and pottery deposition across both cemeteries, isolating a notable trend among a select group of tombs.
@@
 Figure \@ref(fig:aoristbycem) presents the aoristic sums of pithoi (as a proxy for burials) and of all vessels, respectively.
-Despite differences in scale and timing, it is immediately apparent that Fortetsa and the KNC experieced similar chronological developments in the volume of burials and, to a slightly lesser degree, grave goods.
-In both cemeteries, the rate of cremation burial began rising around the end of the 9th century, plateaued slightly in the early 8th, and then rose again to a peak in the late 8th and early 7th centuries.
+Despite differences in scale and timing, it is immediately apparent that Fortetsa and the KNC experieced similar chronological developments in burials and, to a slightly lesser degree, grave goods.
+In both cemeteries, the rate of cremation burial began rising around the end of the 9th century, plateaued slightly in the early 8th, and then rose again to a peak in the late 8th and early 7th.
 Across all vessels, the developments are more volatile, but largely similar, especially regarding the late spike in burial activity.
-We might assume that these developments would be mirrored in the construction of new tombs, but this proves not to be the case.
+We might assume that these trends would be mirrored in the construction of new tombs, but this proves not to be the case.
 
@@
 Figure \@ref(fig:tombconstruction) presents two ways of dating the construction of the tombs.
-On the left, counts are based on the ceramic phase considered most likely to correspond to the tomb's first use, including "PG", "G" and "O" where no greater specificity was possible.
-On the right, aoristic sums are presented, spreading the uncertainty associated with those general assignations.
 In both cases, the trend appears clear.
@@
 The Fortetsa, meanwhile, saw next to no new chamber tomb construction from the 8th century onwards.
-The abatement of chamber tomb construction has been noted before (Brock 1957, 4; Cavanagh 1996, 652–3) and somewhat implicitly read as evidence of restructuring of the funerary rite, if not a decline in the fortunes of the wider community.
-But this image of a waigning investment in chamber tombs sits uneasily with the observable growth in mortuary activity in the late 8th century.
+This abatement has been noted before (Brock 1957: 4; Cavanagh 1996: 652–53) and somewhat implicitly read as evidence of restructuring of the funerary rite, if not a decline in the fortunes of the wider community.
+But this image of a waigning investment in chamber tombs sits uneasily with the observable growth in mortuary activity.
 
@@
 These countervalent trends are observable too in the number of tombs receiving depositions in each period (Figure \@ref(fig:tombsinuse)).
-At Forteta, this number remained relatively stable from the MPG to LG periods before declining (and bear in mind, there were only around 20 tombs belonging to this cemetery throughout the period), while at the KNC, the number of tombs in use rose to a preak in the late 8th century before declining.
+At Fortetsa, this number remained relatively stable from the MPG to LG periods before declining (and bear in mind, there were only ever around 20 tombs belonging to this cemetery), while at the KNC, the number of tombs in use rose to a preak in the late 8th century before declining.
 But this is not the whole story.
@@
 
-Figure \@ref(fig:tombsinusebybuilddate) shows the number of tombs in use in each period, divided into the centuries in which those tombs were constructed, all overlaid on the total number of tombs in use across both cemeteries.
-A wave-like pattern can be observed, with each cohort of tombs steadily going out of use as new ones were built.
-This would seem to accord with Cavanagh's suggestion that "[t]he length of time the tombs last seems [...] to reflect the duration of the social group which had access to the sepulchre", and that, if such access was lineally inehrited, frequent fission would result in limited trans-generational use.
+By dividing the tombs in use into the centuries during which they were built (Figure \@ref(fig:tombsinusebybuilddate)), we see a wave-like pattern with each cohort steadily going out of use as new ones were constructed.
+This would seem to accord with Cavanagh's (1996) suggestion that "[t]he length of time the tombs last seems [...] to reflect the duration of the social group which had access to the sepulchre", and that, if such access was lineally inehrited, frequent fission would result in limited trans-generational use.
 We might expect, then, that it was the 8th century tombs coming into use that spurred the dramatic rise in burial rates previoulsy noted.
-But this is not the case.
 
@@
 
-In Figure \@ref(fig:potsbydateandtombdate), all vessels are assigned, firstly, to the century in which they were produced and, secondly, to the century in which the tomb they were found in was built.
-So, though the late 8th and early 7th centuries represented the apex of burial activity across both cemeteries, it was not the tombs built in those centuries which received the lion's share of the burials and associated ceramics.
-This is all the more notable when, returning to Figure \@ref(fig:tombsinusebybuilddate), we see that the number of 9th and 10th century tombs in use by this time was declining.
+However, we see from Figure \@ref(fig:potsbydateandtombdate) that, although the late 8th and early 7th centuries represented the apex of burial activity across both cemeteries, it was tombs built back in the 9th and 10th which received the lion's share of the burials and associated ceramics.
+This is all the more notable when, returning to Figure \@ref(fig:tombsinusebybuilddate), we see that the number of 9th and 10th century tombs in use by this time was, in fact, declining.
 
@@
 Using pithoi as a proxy for burials, we can estimate the changing rates of burial through time.
-And while for the most these remained around three per tomb per century, in the 8th and 7th centuries, tombs founded in the 9th century saw a rapid rise in the rate of burial, peaking at over 11 per tomb per century.
+And while these mostly remained around three per tomb per century, in the 8th and 7th centuries, tombs founded in the 9th century saw a rapid rise in the rate of burial, peaking at over 11 per tomb per century.
 Though these estimates are no doubt very rough, the marked nature of this divergence suggests it is not simply an artefact of imperfect data.
@@
 
-We are yet to draw out which older tombs precisely are responsible for the observible patterning so far discussed.
+We are yet to draw out which older tombs precisely are responsible for the trend so far discussed.
 We might assume the tombs which became the focus of intense later burial activity would be those which yielded the most burials and grave goods when excavated and, happily, this supposition proves true.
@@
 
-tombs_ranked<-tombs[with(tombs, order(-tombs$total_pottery)), ]
+tombs_ranked<-tombs[with(tombs, order(-tombs$total_finds)), ]
 tombs_top_12<-head(tombs_ranked,12)
@@
 
+aorist_all_tomb_names$cemetery<-tombs_top_12$cem_condensed
+
 aorist_all_tomb_melt<-melt(aorist_all_tomb_names)
@@
 
-ggplot(aorist_all_tomb_melt,aes(variable, value))+geom_col()+theme(legend.position = "none")+facet_wrap(~tomb,scales="free_y")+
-  geom_col()+theme_bw()+theme(legend.title=element_blank(),
+ggplot(aorist_all_tomb_melt,aes(variable, value,fill=cemetery))+geom_col()+theme(legend.position = "none")+facet_wrap(~tomb,scales="free_y")+
+  geom_col()+theme_bw()+theme(legend.title=element_blank(),legend.position="bottom",
       text=element_text(family="Garamond", size=12))+
           labs(x="Date (Years BC)", y="Aoristic Sums")+
-              scale_fill_brewer(palette = "Set3")+ 
-                theme(legend.position="none")+
+              scale_fill_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+ 
                   coord_cartesian(xlim=c(1100,610))+scale_y_continuous()+
@@
 
-The 12 tombs with the most associated vessels are all chamber tombs, 10 dating to the 9th century and two to the 10th, and seven of which did exhibit late, intense episodes of deposition (Figure \@ref(fig:toptombs)).
+The 12 tombs with the most associated vessels are all chamber tombs, 10 dating to the 9th century and two to the 10th, and seven of them did exhibit late, intense episodes of deposition (Figure \@ref(fig:toptombs)).
 In fact, Tombs P, 292, 75, 107, 285, II and 218 (henceforth 'Group A') between them account for `r round(calc1,digits=2)`% of the entire ceramic assemblage from both cemeteries.
 All but Tomb 285 date to the 9th century, and together they underpinned the discussed late spate of burials.
-Interestingly, Tombs X, G, Q, 104 and 283 ('Group B'), however, evince a different pattern, with their respective peaks coming around a century earlier, which, as seen in Figure \@ref(fig:aoristbycem) was another period of increased burial activity.
+Interestingly, Tombs X, G, Q, 219 and 283 ('Group B'), however, evince a different pattern, with their respective peaks coming around a century earlier, which, as seen in Figure \@ref(fig:aoristbycem), was another period of increased burial activity.
 So despite receiving the greatest quantities of burials and grave goods, these tombs nonetheless seem to have achieved that distinction via different routes.
-By looking more closely at the periods in which they were built, and in which they were most popular, I believe the differences between these tombs go a long way to explaining the late rise in burial activity here discussed.
+And, indeed, the differences between them may shed light on the late surge in burial activity here discussed.
 
@@
 
-The indices employed so far – the distribution of artefacts between tombs, and the construction and use of the tombs through time – inform us of the temporal or aggregate features of the cemeterie, but they tell us little of the tombs as they were built or first used.
-In attempting more synchronic comparisons, we shall first consider the physical dimensions of the tombs as built - a potentially rich context for the negotiation of broader issues of social power, status and group membership.
+The indices employed so far – the distribution of artefacts between tombs, and the construction and use of the tombs through time – inform us of the temporal or aggregate features of the cemeteries, but tell us little of the tombs as they were built or first used.
+In attempting more synchronic comparisons, we shall first consider the physical dimensions of the tombs as built - a potentially rich context for the negotiation of social power, status and group membership.
 
@@
 
-Furthermore, it seems that no clear relationship existed between the physical size of a tomb and its buried population at EIA Knossos (see Figure \@ref(fig:tombdimensions)).
-True, over half of all tombs do not exceed 2.6 sq m in area, while nine of the ten tombs with the most burials do surpass this value.
+Furthermore, it seems that no clear relationship existed between the physical size of a tomb and its eventual buried population (see Figure \@ref(fig:tombdimensions)).
+True, over half of all tombs do not exceed 2.6 sq m in area, while nine of the ten tombs with the most burials do.
 But many of the very largest tombs are decidedly middling when it comes to burial numbers.
@@
 
-It seems dubious, then, whether the scale of the tombs when built carried overt social meanings but, even if it did, this didn't prove a determining factor in their later reuse.
+It seems dubious, then, whether the scale of the tombs when built carried overt social meanings but, even if it did, this proved irrelevant to their later reuse.
 Several tombs had niches cut into their *dromoi*, or the *stomion* extended, to fit in later burial urns.
-Sometimes, urns from within the chamber were relocated to such niches to make space for new interments.
-There is an impression of competition for inclusion into the sepulchral group, further implying the tombs’ original builders little anticipated their eventual popularity.
+Sometimes, urns from within the chamber were relocated to such niches to make space for new interments (e.g.
+T.
+TFT; Brock 1957: 3–4).
+The impression is of competition for inclusion into the sepulchral group, further implying the tombs’ original builders little anticipated their eventual popularity.
 
@@
 pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A",
-       ifelse(pots$tomb=="X"|pots$tomb=="G"|pots$tomb=="104"|pots$tomb=="Q"|pots$tomb=="283","B","C"))
+       ifelse(pots$tomb=="X"|pots$tomb=="G"|pots$tomb=="219"|pots$tomb=="Q"|pots$tomb=="283","B","C"))
 tombs9th10th<-subset(tombs,century_built=="9th" & total_pottery>50|century_built=="10th" & total_pottery>50)
@@
 
-And in Figure \@ref(fig:tombscompthroughtime) we see that the tombs of Group A were relatively unremarkable in the scale of their depositions during the 9th century, only diverging significantly from the rest in the subsequent two hundred years.
+And in Figure \@ref(fig:tombscompthroughtime), we see that the tombs of Group A were by no means exceptional in the scale of their depositions during the 9th century, only diverging significantly from the rest in the subsequent two hundred years.
 Group B, meanwhile, had their respective peaks in the 9th or 8th centuries, with a near total decline afterwards.
-As a working hypothesis, let us suppose that these two distinct trajectories correspond to changing socio-politcal and economic contexts, made manifest by the burying groups in their mortuary depositions.
+That is, they seem to have followed a more conventional trajectory, going out of use as new tombs were built.
+As a working hypothesis, let us suppose that these two distinct trajecotire correspond to behaviours born of differing socio-politcal and economic contexts.
 How far might this hypothesis take us in explicating these patterns, and how visible might it be?
-
-The late 9th and early 8th centuries are consistently interpreted as periods of artistic development, increased overseas trade, and renewed vigour in the construction of tombs and the visibility of the burial rite.
-The PGB style reflects a uniquely Cretan blend of Attic and Near Eastern forms, while the appearance of BA 'antiques' in the tombs from this period onwards is felt to reflect rejuvinated, even competitive interest in the Minoan past.
-
-```{r imports, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="goes around here"}
-
-imports<-subset(pots,pots$import!="")
-greekimports<-subset(imports,imports$import_region=="Greece")
-NEimports<-subset(imports,imports$import_region=="Near East")
-
-import_aorist<-colSums(imports[27:76])
-import_aorist<-data.frame(import_aorist)
-colnames(import_aorist)<-c("aoristic_sum")
-row.names(import_aorist)<-dates
-setDT(import_aorist, keep.rownames = "decade")
-import_aorist$decade<-as.numeric(import_aorist$decade)
-
-greekimport_aorist<-colSums(greekimports[27:76])
-greekimport_aorist<-data.frame(greekimport_aorist)
-colnames(greekimport_aorist)<-c("aoristic_sum")
-row.names(greekimport_aorist)<-dates
-setDT(greekimport_aorist, keep.rownames = "decade")
-greekimport_aorist$decade<-as.numeric(greekimport_aorist$decade)
-
-NEimport_aorist<-colSums(NEimports[27:76])
-NEimport_aorist<-data.frame(NEimport_aorist)
-colnames(NEimport_aorist)<-c("aoristic_sum")
-row.names(NEimport_aorist)<-dates
-setDT(NEimport_aorist, keep.rownames = "decade")
-NEimport_aorist$decade<-as.numeric(NEimport_aorist$decade)
-
-greekimport_aorist$origin<-"Greece"
-NEimport_aorist$origin<-"Near East"
-allimport_aorist<-rbind(greekimport_aorist,NEimport_aorist)
-
-ggplot()+geom_col(data=import_aorist,aes(decade,aoristic_sum),alpha=0.2)+
-          geom_col(data=allimport_aorist,aes(decade,aoristic_sum, fill=origin),position="dodge")+
-            theme_bw()+theme(text=element_text(family="Garamond", size=12))+
-                labs(x="Date (Years BC)", y="Aoristic Sum")+
-                  scale_fill_brewer(name="Origin",palette = "Set2")+
-                   scale_y_continuous(expand=c(0,0),limits=c(0,20))+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))
-
-```
-
-The rise in imported items is clearly seen in the tomb assemblages, with a succession of increased Greek and Near Eastern imports respectively in the late 9th and early 8th centuries (Figure \@ref(fig:imports)).
-This much has been relatively well understood for some time, at least at the KNC, but we can further examine how these imports were distributed among the various groups of tombs.
-
-```{r importsgroups, fig.dim=c(8,4),dpi=300, fig.align='center',fig.cap="Insert Figure 15 here"}
-
-pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A",
-                   ifelse(pots$tomb=="X"|pots$tomb=="G"|pots$tomb=="104"|pots$tomb=="Q"|pots$tomb=="283","B","C"))
-
-imports<-subset(pots,import!="")
-imports_A<-subset(imports,group=="A")
-imports_B<-subset(imports,group=="B")
-imports_C<-subset(imports,group=="C")
-
-import_A_aorist<-colSums(imports_A[27:76])
-import_A_aorist<-data.frame(import_A_aorist)
-colnames(import_A_aorist)<-c("aoristic_sum")
-row.names(import_A_aorist)<-dates
-setDT(import_A_aorist, keep.rownames = "decade")
-import_A_aorist$decade<-as.numeric(import_A_aorist$decade)
-
-import_A_aorist$group<-"A"
-
-import_B_aorist<-colSums(imports_B[27:76])
-import_B_aorist<-data.frame(import_B_aorist)
-colnames(import_B_aorist)<-c("aoristic_sum")
-row.names(import_B_aorist)<-dates
-setDT(import_B_aorist, keep.rownames = "decade")
-import_B_aorist$decade<-as.numeric(import_B_aorist$decade)
-
-import_B_aorist$group<-"B"
-
-import_C_aorist<-colSums(imports_C[27:76])
-import_C_aorist<-data.frame(import_C_aorist)
-colnames(import_C_aorist)<-c("aoristic_sum")
-row.names(import_C_aorist)<-dates
-setDT(import_C_aorist, keep.rownames = "decade")
-import_C_aorist$decade<-as.numeric(import_C_aorist$decade)
-
-import_C_aorist$group<-"C"
-
-aorist_import_groups<-rbind(import_A_aorist,import_B_aorist,import_C_aorist)
-
-ggplot(aorist_import_groups,aes(decade,aoristic_sum,fill=group))+ geom_col(position="dodge")+
-  theme_bw()+theme(legend.position="bottom",text=element_text(family="Garamond", size=12))+
-  labs(x="Date (Years BC)", y="Aoristic Sum")+
-  guides(fill=guide_legend(title="Group"))+
-  scale_fill_brewer(palette = "Set2")+ 
-  scale_y_continuous(expand=c(0,0),limits=c(0,12))+
-  scale_x_reverse(expand = c(0, 0),limits=c(1000,600))+
-  theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-```
-
-In Figure \@ref(fig:importsgroups), not only is it apparent that a range of tombs were beneficiaries of the increase in imports in the LPG-EG periods, but that there was also potentially significant patterning in where those artefacts ended up.
-The Group A and B tombs closely trace the trends for Greek and Near Eastern imports respectively (and, indeed, tombs Q and G contained only Greek imports), while across all other tombs ('Group C'), there was also no doubt variegation not observable here.
-Imported ceramics are used here as something of a proxy for items of status value, a shorthand that would ideally be bolstered by other artefacts (precious vessels, weapons, jewellery etc.) which, owing to their less-developed chronologies, must unfortunately be excluded.
-But let us suggest, for the time being, that these trends are representative of a larger truth, namely that the late 9th and early 8th centuries witnessed a peak in the variety of depositions, which went unmatched in later periods.
 
@@
 
-To test this, admittedly tentative, proposition, Figure \@ref(fig:bigbox) presents a boxplot of the total aoristic sums across both cemeteries, colour-coded by ceramic phase.
-Note that, for visual clarity, the outliers tomb G in the EG period, and tomb P in the late EO-LO period are excluded, with values ranging from 29-34, and 36-44 respectively.
-The important observation is that the PGB and EG exhibit the largest interquartile ranges (IQRs) of any period, and each possess only a single outlier (beyond 1.5 times the IQR).
+Figure \@ref(fig:bigbox) presents a boxplot of the total aoristic sums across both cemeteries, colour-coded by ceramic phase.
+Looking at the PGB and EG periods, we see that they exhibit the largest interquartile ranges (IQRs) of any period, and each possess only a single outlier (beyond 1.5 times the IQR).
 The Orientalising period, on the other hand, exhibits much reduced IQRs, and a proliferation of outliers far and above the range of the other tombs.
@@
 
-rects <- data.frame(xstart = c(1100,970,920,875,840,810,790,745,70,670,630), 
+rects <- data.frame(xstart = c(1100,970,920,875,840,810,790,745,700,670,630), 
                     xend = c(970,920,875,840,810,790,745,710,670,630,600), 
@@
 
-This possibility is supported further by considering the gini coefficient of the distribution of pottery between the tombs (see Figure \@ref(fig:gini).
-Using the aoristic sums as a probablistically-weighted estimate of the volume of material deposited in each tomb in each decade, use of the gini coefficient can give an (albeit rough) estimate of how unequally distributed that material was.
+This characterisation is supported further by considering the Gini coefficient of the distribution of pottery between the tombs (see Figure \@ref(fig:gini).
+Using aoristic sums as a probablistically-weighted estimate of the volume of material deposited in each tomb in each decade, the Gini coefficient can give an (albeit rough) estimate of how unequally distributed that material was.
 And, though values were always relatively high (i.e.
-unequal) within the cemeteries, the major spike in the gini coefficient seen in the Orientalising period does contrast markedly with the modest rise and plateau of the PGB-EG.
-The gini coefficient is most commonly used in the assessment of income inequality, but I want to stress here that it is not "wealth" in any meaningful sense that we are talking about here.
-Indeed, if anything, the late flourit of burial activity here discussed may be a case of quantity, and not quality.
+unequal) within the cemeteries, the major spike in the Gini coefficient seen in the Orientalising period does contrast markedly with the modest rise and plateau of the PGB-EG.
+
+This is not to suggest that the late 9th to early 8th century was a time of modest, egalitarian burial practices.
+The PGB and EG are consistently interpreted as periods of artistic development, increased overseas trade, and renewed vigour in the construction of tombs and the visibility of the burial rite.
+The PGB style reflects a uniquely Cretan blend of Attic and Near Eastern forms, while the appearance of BA 'antiques' in the tombs from this period onwards is felt to reflect rejuvinated, even competitive interest in the Minoan past.
+
+```{r imports, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="goes around here"}
+
+imports<-subset(pots,pots$import!="")
+greekimports<-subset(imports,imports$import_region=="Greece")
+NEimports<-subset(imports,imports$import_region=="Near East")
+imitations<-subset(pots,pots$imitation!="")
+
+import_aorist<-colSums(imports[27:76])
+import_aorist<-data.frame(import_aorist)
+colnames(import_aorist)<-c("aoristic_sum")
+row.names(import_aorist)<-dates
+setDT(import_aorist, keep.rownames = "decade")
+import_aorist$decade<-as.numeric(import_aorist$decade)
+
+greekimport_aorist<-colSums(greekimports[27:76])
+greekimport_aorist<-data.frame(greekimport_aorist)
+colnames(greekimport_aorist)<-c("aoristic_sum")
+row.names(greekimport_aorist)<-dates
+setDT(greekimport_aorist, keep.rownames = "decade")
+greekimport_aorist$decade<-as.numeric(greekimport_aorist$decade)
+
+NEimport_aorist<-colSums(NEimports[27:76])
+NEimport_aorist<-data.frame(NEimport_aorist)
+colnames(NEimport_aorist)<-c("aoristic_sum")
+row.names(NEimport_aorist)<-dates
+setDT(NEimport_aorist, keep.rownames = "decade")
+NEimport_aorist$decade<-as.numeric(NEimport_aorist$decade)
+
+imitation_aorist<-colSums(imitations[27:76])
+imitation_aorist<-data.frame(imitation_aorist)
+colnames(imitation_aorist)<-c("aoristic_sum")
+row.names(imitation_aorist)<-dates
+setDT(imitation_aorist, keep.rownames = "decade")
+imitation_aorist$decade<-as.numeric(imitation_aorist$decade)
+
+greekimport_aorist$origin<-"Greek Imports"
+NEimport_aorist$origin<-"Near Eastern Imports"
+imitation_aorist$origin<-"'Oriental' Imitations"
+all_import_imitation<-rbind(greekimport_aorist,NEimport_aorist,imitation_aorist)
+
+ggplot()+ geom_col(data=all_import_imitation,aes(decade,aoristic_sum, fill=origin),position="dodge")+
+            theme_bw()+theme(text=element_text(family="Garamond", size=12), legend.position="bottom")+
+                labs(x="Date (Years BC)", y="Aoristic Sum")+
+                  scale_fill_brewer(name="Origin",palette = "Set2")+
+                   scale_y_continuous(expand=c(0,0),limits=c(0,20))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))
+
+```
+
+The rise in imported items is clearly seen in the tomb assemblages, with a succession of increased Greek and Near Eastern imports respectively in the late 9th and early 8th centuries (Figure \@ref(fig:imports)).
+Likewise, the concomitant decline in Near Eastern imports and rise in locally-made imitations thereof, is visible at the end of the 8th century.
+These trends being as a very coarse proxy for its engagement with the wider Mediterranean, the late 9th and early 8th centuries would seem to encompass the most outward-looking phase of the EIA Knossian community.
+This much has been relatively well understood for some time, at least at the KNC, but we can further examine how these imports were distributed among the various groups of tombs.
+
+We have seen that the Group B tombs all experienced their respective peaks in burial activity around the end of the 9th century, before declining in use, while those of Group A went on to receive ever more burials into the late 8th.
+It is notable, then, that among the Group B tombs, imported items constituted `r nrow(subset(pots,pots$group=="B" & pots$import =="Y"))/nrow(subset(pots,pots$group=="B"))*100`% of the total assemblage, compared to `r nrow(subset(pots,pots$group=="C" & pots$import =="Y"))/nrow(subset(pots,pots$group=="C"))*100`% of Group C and only `r nrow(subset(pots,pots$group=="A" & pots$import =="Y"))/nrow(subset(pots,pots$group=="A"))*100`% of Group A.
+This may further endorse the characterisation of the PGB-EG periods, wherein the Group B tombs flourished, and after which they declined, as a time of competitive display between burying parties.
+The contents of the Group A tombs, though probably once also shaped by such strategies, appear to have been diluted, as it were, by their later use which, as is becoming increasingly apparent, was of a different kind.
 
@@
 
-pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A","B")
+pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A",
+       ifelse(pots$tomb=="X"|pots$tomb=="G"|pots$tomb=="219"|pots$tomb=="Q"|pots$tomb=="283","B","C"))
 
@@
                  labs(title="KNC",x="", y="Proportion of Total Assemblage")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" Amphora  "," Jug  "," Oil vessel  ", " Pithos"," Drinking vessel  "," Lid  "," Other"))+ 
+                  scale_fill_brewer(palette = "Set2")+ 
                     scale_y_continuous(expand=c(0,0.0))
@@
                  labs(title="Fortetsa",x="Century BC", y="")+
-                  scale_fill_brewer(palette = "Set2", labels=c(" Amphora  "," Jug  "," Oil vessel  ", " Pithos"," Drinking vessel  "," Lid  "," Other"))+ 
+                  scale_fill_brewer(palette = "Set2")+ 
                     scale_y_continuous(expand=c(0,0.0))
@@
                  labs(title="Group A",x="", y="")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" Amphora  "," Jug  "," Oil vessel  ", " Pithos"," Drinking vessel  "," Lid  "," Other"))+ 
+                  scale_fill_brewer(palette = "Set2")+ 
                     scale_y_continuous(expand=c(0,0.0))
@@
 
-Figure \@ref(fig:comps) shows the composition of the total pottery assemblage from each cemetery, and of the Group A tombs, through the centuries.
-In all cases, there appears to be the steady crystallisation of a basic 'burial kit', comprising a pithos and lid, an oil vessel, a jug, and a drinking vessel.
-The proportion of grave goods comprising 'other' vessels - pyxides, askoi, trays, stands, dinoi, kernoi, kraters, lekanai and more - declines to become only a minor part of the standardised repertoire of the 7th century.
+This is reflected further in the changing composition of the mortuary assemblage through time (Figure \@ref(fig:comps)).
+Across both cemeteries, there appears to have been the steady crystallisation of a basic 'burial kit', comprising a pithos and lid, an oil vessel or two, a jug, and a drinking vessel.
+The proportion of grave goods comprising 'other' vessels - pyxides, askoi, trays, stands, dinoi, kernoi, kraters and more - declines to become only a minor part of the standardised repertoire of the 7th century.
 The tombs of Group A did not break with this mould, and their assemblages seem as conventional as any other during the last century of use in the cemeteries.
 
-Plainly, the main focus of this article has been on ceramics and not other artefacts of metal, stone, glass, and organic materials, due to their less well-developed chronologies, but even the raw distribution of such artefacts may make more sense in the light of the foregoing discussion.
-There is, perhaps unsurprisingly, a relatively strong (Spearman's ρ=`r spearmanobjectspots`) correlation between the number of ceramic and non-ceramic finds across the tombs, and, certainly, many of the most unique and precious objects were recovered from among the larger tombs.
+The main focus of this article has been on ceramics and not other artefacts of metal, stone, glass, and organic materials, due to their less well-developed chronologies, but even the raw distribution of such artefacts may make more sense in the light of the foregoing discussion.
+There is, perhaps unsurprisingly, a relatively strong (Spearman's ρ=`r spearmanobjectspots`) correlation between the number of ceramic and non-ceramic finds across the tombs, and, certainly, many of the most seemingly precious objects were recovered from among the larger tombs.
 But the monopoly those tombs would appear to have over such objects, perhaps tempting one initially to speak of 'elite' or 'high-status' tombs, evaporates when the number of those objects is divided by the number of burials represented.
@@
                     scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-total_metal<-ggplot(subset(tombs,est_burials>1 & weapon>0),aes(weapon, fill=group))+geom_histogram()+
+total_metal<-ggplot(subset(tombs,est_burials>1 & total_metal>0),aes(total_metal, fill=group))+geom_histogram()+
   theme_bw()+theme(legend.title=element_blank(),
@@
                     scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-metalbyburial<-ggplot(subset(tombs,est_burials>1 & weapon>0),aes(weapon/est_burials, fill=group))+geom_histogram()+
+metalbyburial<-ggplot(subset(tombs,est_burials>1 & total_metal>0),aes(total_metal/est_burials, fill=group))+geom_histogram()+
   theme_bw()+theme(legend.title=element_blank(),
@@
 To summarise, towards the end of 8th century BC, a select group of tombs, all of which had been constructed back in the 9th or 10th, became the foci of a major increase in burials, something which did not occur among the tombs built around that time.
-We have seen how this surge in acivity occurred against the backdrop of a declining number of older tombs in use, and I have suggested that the rates of change are too rapid, and unevenly distributed, to be explicable through population growth alone.
 We have seen how little about the physical dimensions or early use of these tombs presaged their later popularity, and that, while a similar, if smaller, increase in burial activity occurred around a century earlier, in the PGB-EG periods, this contrasted in the comparative diversity of its associated assemblages, and their distribution across the tombs.
-The later burials were marked by great uniformity in the associated grave goods, and this apparently limited and consistent repertoire may explain why many of the most-used tombs prove exceptional only in their absolute quantities of artefacts, and not the accoutrements of each individual burial.
+The later burials were marked by great uniformity in the associated grave goods - set against a backdrop of waining tomb construction and declining imports - and this apparently limited and consistent repertoire may explain why many of the most-used tombs prove exceptional only in their absolute quantities of artefacts, and not the accoutrements of each individual burial.
 
@@
 
-The preceding analysis has revealed a number of potentially significant patterns in the use of the Fortetsa and Knossos North Cemeteries; it only remains to consider how we might characterise the burying groups responsible behind them.
-The traditional view is that such groups were immediate families (Brock 1957, 41; @Boardman1967, 63; @Coldstream1996, 291).
+It remains to consider how we might characterise the burying groups responsible behind these various, intriguing trends.
+The traditional view is that such groups were immediate families (Brock 1957, 41; @Boardman1967: 63; @Coldstream1996: 291).
 The relatively short lifespan of many tombs (often only two or three generations) has been taken as evidence for direct, lineal inheritance, with the tomb going out of use as families moved, individuals founded new tombs, or the line came to an end.
@@
 
-The model of lineal inheritance has been recently challenged by Kotsonas [-@Kotsonas2011, 131], however, who considers it ‘probably the greatest factoid in the archaeology of Crete of the given period’.
+The model of lineal inheritance has been recently challenged by Kotsonas [-@Kotsonas2011: 131], however, who considers it ‘probably the greatest factoid in the archaeology of Crete of the given period’.
 He argues that the rise in interments during the 8th century is too great to be explained by population growth, as would be necessary for the possibility of direct lineal inheritance to be preserved.
@@
 In the traditional view, political authority was concentrated in the hands of the *kosmoi*, aristocratic magistrates selected from the leading, hereditary *startoi*, or ‘clans’, networks of extended lineages controlling the ancestral estates on which worked attendent serfs and/or slaves (Willetts 1982, 240).
-But despite elements of this account probably being accurate, others remain equivocal, such as the nature of the *pyla*, or 'tribe', and its relation to the *startos* (which may itself be a military grouping), the distinction between an *astos* and a *poliatas*, both clearly formulations of the rights-bearing civic actor, and the ancestry and spread of the *agelai*, seemingly an adolescent age-grade so far only attested at Eltynia [@Gagarin2016, 36; 78-80].
+But despite elements of this account probably being accurate, others remain equivocal, such as the nature of the *pyla*, or 'tribe', and its relation to the *startos* (which may itself be a military grouping), the distinction between an *astos* and a *poliatas*, both clearly formulations of the rights-bearing civic actor, and the ancestry and spread of the *agelai*, seemingly an adolescent age-grade so far only attested at Eltynia [@Gagarin2016: 36; 78-80].
 Indeed, many terms which might illuminate social divisions and groupings within the Archaic Cretan populace are only attested at one or two sites, and even then sometimes appear to differ in their form or function.
 
-It wouldn't be helpful at present, then, to draw any simple parallel between the Knossian burial groups and, say, powerful *startoi*, a term which is not even attested outside of Gortyn in this period (ibid., 1-2).
+It wouldn't be helpful at present, then, to draw any simple parallel between the Knossian burial groups and, say, powerful *startoi*, a term which is not even attested outside of Gortyn in this period (ibid:, 1-2).
 But archaeological and textual evidence suggests that, in general terms, such supra-familial kinship groups were undergoing a transition during this period, suggesting some such attribution for the burying groups at Knossos is not unreasonable, given the noted changes there also.
-Elsewhere on Crete, we see a shift to unmarked, individual burials, and more austere grave goods, as well as an apparent reorientation of sanctuary deposits away from extra-mural to suburban spaces, and a rise in utilitarian, mass-produced votives (ibid., 34-6).
-At Knossos, a small number of pithos burials in pits do appear in the KNC and Fortetsa volumes (Brock 1957, 98; Coldstream and Catling 1996, 162), and recent re-excavation of the Teke plot identified numerous small pit burials around three chamber tombs [@Evely2012], raising the possibility that others exist, but have gone unrecognised in prior excavations.
+Elsewhere on Crete, we see a shift to unmarked, individual burials, and more austere grave goods, as well as an apparent reorientation of sanctuary deposits away from extra-mural to suburban spaces, and a rise in utilitarian, mass-produced votives (ibid.: 34-6).
+At Knossos, a small number of pithos burials in pits do appear in the KNC and Fortetsa volumes (Brock 1957: 98; Coldstream and Catling 1996: 162), and recent re-excavation of the Teke plot identified numerous small pit burials around three chamber tombs [@Evely2012], raising the possibility that others exist, but have gone unrecognised in prior excavations.
 At Azoria, major urban remodelling in the 7th century saw the earlier, agglutinative house forms of neighbouring Kavousi Vronda and Kastro give way to larger, more formally planned dwellings with little evidence of renovation or remodelling, while larger, apprently civic, buildings were constructed to accommodate major public gatherings [@Haggis2011; @Haggis2011b; @Haggis2014].
 
-Together, these may be indications that with the emergence of the archaic *poleis* on Crete, there was a concomitant restructuring of wider social groups, institutions, and behaviours.
-Was this transformation linked to an emerging emphasis on the citizen, the state, and ideologies of relative egality, within which communal gatherings, intra-mural ritual, and individual graves replaced ostentatious, kin-centred acts of gathering, dedication, and burial?
-It is plausible, though as Perlman and Gagarin (32-37) concede, the decline of ostentatious burial could just as well represent a voluntary abandonment by groups assured of their economic and political influence, finding the symbolic expression of kin-relations in death no longer a necessary expense.
-
-These issues, naturally, recall the conclusions of Morris [-@Morris1987, 171-211], who argued 30 years ago that the emergence of the *polis* across the Aegean led to a tension between elite preferences for conspicuous display, and a downplaying of social difference in death, as promoted by the ‘middling’ ideology that the *polis* engendered and promulgated.
+In the Knossian cemeteries, we have seen a somewhat diverse set of burial practices give way to a more prolific, yet more restricted, enumeration of such rites, focussed on a small number of long-established tombs, in the context of diminishing collective tomb construction, and shortly before a complete cessation of such depositions.
+Together, these varied threads suggest that with the emergence of the archaic *poleis* on Crete, there was a concomitant restructuring of wider social groups, institutions, and behaviours.
+It seems not unreasonable to infer an emergent emphasis on the citizen, the state, and ideologies of relative egality, within which communal gatherings, intra-mural ritual, and individual graves replaced ostentatious, kin-centred acts of gathering, dedication, and burial.
+
+These issues, naturally, recall the conclusions of Morris [-@Morris1987: 171-211], who argued 30 years ago that the emergence of the *polis* across the Aegean led to a tension between elite preferences for conspicuous display, and a downplaying of social difference in death, as promoted by the ‘middling’ ideology that the *polis* engendered and promulgated.
 In his well-known model of the *agathoi* and *kakoi*, Morris argued that, at various times, the relevant spheres of kinship demonstrated in burial could widen and constrict, with more and less distantly related lineages drawn into association through changing social strategies, until the pre-eminence of kinship relations in the structuring of political power declined, along with its manifestiations in mortuary practice.
@@
 But whether this reflected some standard set of accoutrements intentionally shared by the hypothetical *agathoi* of Knossos, or else nascent, egalitarian strictures over elaborate burial is hard to judge.
-The PGB period seems a far better candidate for such an expansion, with its greater variety in chamber dimensions, relatively broad distribution of (local and imported) grave goods, not to mention the innovative decoration and reuse of BA artefacts noted elsewhere.
+The PGB period seems a far better candidate for such an expansion, with its greater variety in chamber dimensions, relatively broad distribution of (local and imported) grave goods, not to mention the innovative ceramic decoration and reuse of BA artefacts noted elsewhere.
 One thing is clear; the social context which produced these signatures contrasted markedly with that of a century later, with its ever greater volume of interments in an ever-smaller cohort of tombs, its more uniform assemblages, and its decline of imported items.
@@
 But, as a more nuanced picture emerges of the island's relationship with the wider Mediterranean, subtler characterisations are possible.
-It is telling that the peak for imported items in the tombs came in the 9th century; the 'Orientalising' as a time period feels increasingly misapplied to Crete, when, at Knossos at least, it witnessed far lower rates of 'oriental' imports, and only limited adoption of the figural motifs which characterise its appearance elsewhere.
-This decline in imports was, however, counterpointed by a rise in imitations of Near Eastern vessels [@Antoniadis2012, 170], something which further accords with our growing awareness of the selective nature of Archaic and Classical Crete's external relations.
+It is telling that the peak for imported items in the tombs came in the 9th century; the Orientalising as a time period feels increasingly misapplied to Crete, when, at Knossos at least, it witnessed far lower rates of 'oriental' imports, and only limited adoption of the figural motifs which characterise its appearance elsewhere.
+This decline in imports was, however, counterpointed by a rise in imitations of Near Eastern vessels [@Antoniadis2012: 170], something which further accords with our growing awareness of the selective nature of Archaic and Classical Crete's external relations.
 Erickson [@Erickson2010] has, likewise, argued that *symposia* were practised on Crete, something often disputed, but that their practitioners merely favoured simpler vessel forms than the figurally-decorated wares popular on the mainland.
@@
 Issues of gender were sadly little considered, though this was largely due to limitations in the data themselves.
-There have also been a number of tombs discovered in the last 20 years that fill in the spaces between the Teke plot and the Medical Faculty Site, and ideally these unpublished tombs should be included in further analyses as they, along with the unconsidered Teke tholos and associated tombs, are almost surely part of a single contiguous cemetery [@Blackman1999, 113; @Catling1983, 51; @French1991, 68-69].
+There have also been a number of tombs discovered in the last 20 years that fill in the spaces between the Teke plot and the Medical Faculty Site, and ideally these unpublished tombs should be included in further analyses as they, along with the unconsidered Teke tholos and associated tombs, are almost surely part of a single contiguous cemetery [@Blackman1999: 113; @Catling1983: 51; @French1991: 68-69].
 Finally, a preliminary study was recently taken by the author of select boxes marked as 'unpublished' from the KNC deposits kept at the British School at Athens at Knossos.
@@
 *Figure 6 Tomb construction through time at both cemeteries.
-On the left are counts based on the earliest dated pottery in each tomb, while on the left are aoristic sums likewise based on the earliest ceramic-based assignations.
+On the left, counts are based on the ceramic phase considered most likely to correspond to the tomb's first use (including simply 'PG', 'G' and 'O' where no greater specificity was possible).
+On the right, aoristic sums are presented, spreading the uncertainty associated with those general assignations.
 In both cases, tombs lacking any securely dateable pottery are excluded.*
+
+*Abbreviations*
+*SM - Subminoan; (E/M/L)PG - (Early/Middle/Late) Protogeometric; PGB - Protogeometric B; (E/M/L)G - (Early/Middle/Late) Geometric; (E/L)O - (Early/Late) Orientalising*
 
@@
 Boxes represent the interquartile range, horizontal lines the median value, whiskers values within 1.5 times the interquartile range, and points outliers beyond this.
-The Y axis has been limited to 20 for ease of interpretation.*
+Note, the Y axis has been limited to 20 to aid readability, meaning the outliers Tomb G in the EG period, and Tomb P in the late EO-LO period are excluded, with values ranging from 29-34, and 36-44 respectively.*
 

### 2019-06-04 analysis/paper/paper2.Rmd
Final rewrite

@@
 library(png)
+library(ggrepel)
 font_import(paths="../paper/font",pattern="GARA")
@@
 Research into this once maligned period has grown apace in recent decades, beginning with the pioneering works of Coldstream [-@Coldstream1977], Desborough [-@Desborough1972], and Snodgrass [-@Snodgrass1971a] and culminating in the rejection by most scholars of the appelation and interpretive bagguage of the so-called ‘Dark Ages’ [see @Kotsonas2016; @Morris1997].
-Nonetheless, our knowledge of many EIA settlements remains often underdeveloped in comparison with those of the second millennium BC.
+Nonetheless, our knowledge of many EIA settlements remains underdeveloped in comparison with those of the second millennium BC.
 
@@
 The cemeteries of Knossos offer an opportunity to introduce such approaches into EIA research on Crete, where they have yet to make much of an impact [see @Kotsonas2011].
-The Fortesta and North Cemetery complexes are both well-published [@Brock1957; @Coldstream1996a] and studied, but a lack of systematic, quantitative comparisons between the cemeteries has left us with many intuited trends, some exceptional tombs, yet a general impression of un-patterned multiplicity.
+The Fortesta and North Cemetery complexes are both well-published [@Brock1957; @Coldstream1996a] and studied, but a lack of systematic, quantitative comparisons between them has left us with many intuited trends, some exceptional tombs, yet a general impression of un-patterned multiplicity.
 
@@
 Settlement evidence at EIA Knossos is, in a word, insubstantial.
-Speaking of the town’s stratigraphic sequence, Coldstream lamented that “the remains of [this] period are like a meagre filling in a very thick sandwich” between Bronze Age and later Greek and Roman levels [@Coldstream1991: 287].
-Most individual finds of Protogeometric (PG) or Geometric date take the form of flimsy foundations, patches of earth floor, wells or pottery scatters, and for a time it was argued that the area in the EIA comprised several dispersed villages, which later coalesced into a single city, as per Aristotle’s model of synoecism [@1950].
+Speaking of the town’s stratigraphic sequence, Coldstream lamented that its “remains [...] are like a meagre filling in a very thick sandwich” between Bronze Age and later Greek and Roman levels [@Coldstream1991: 287].
+Most individual finds of Protogeometric (PG) or Geometric date take the form of flimsy foundations, patches of earth floor, wells or pottery scatters, and for a time it was argued that the area in the EIA comprised several dispersed villages, which later coalesced as per Aristotle’s model of synoecism [@1950].
 
@@
 
-## Heroes and Heirlooms: Object Biographies and Exceptional Burials
-
-A popular approach to the study of Knossos’ EIA tombs has been the investigation of ‘object biography’ [@Appadurai1986; @Kopytoff1986l; @Mauss1954], an anthropological lens emphasising the mutuable nature of an object’s value, the transformations latent within its social existence.
+## Heroes Abroad, Immigrants at Home
+
+A popular approach to the study of Knossos’ EIA tombs has been the investigation of ‘object biography’ [@Appadurai1986; @Kopytoff1986; @Mauss1954], an anthropological lens emphasising the mutuable nature of an object’s value, the transformations latent within its social existence.
 Fertilised by the Homeric motif of gift-giving [@Whitley2002: 220-21], and the remarkable archaeological parallels for such items – such as the boar’s tusk helmet from Tomb 200+ and that given to Odysseus by Meriones (*Iliad* 10.260-271) - such approaches have been brought to bear on certain elaborate burials of the EIA.
@@
 Cluster analyses have been employed by both Cavanagh [-@Coldstream1996: 653-57] and Antoniadis [-@Antoniadis2012: 193-97], the most sophisticated statistical techniques so far applied to the material.
-Antoniadis [-@Antoniadis2012: 172-6] also divided the EIA tombs of Knossos into three groups, those with fewer than 10 pots, fewer than 50, and more than 50, and found that, in each successive case, the proportions of tombs containing 'oriental' imports and imitations increased, suggesting unequal access to such prestigious goods.
+Antoniadis [-@Antoniadis2012: 172-76] also divided the EIA tombs of Knossos into three groups, those with fewer than 10 pots, fewer than 50, and more than 50, and found that, in each successive case, the proportions of tombs containing 'oriental' imports and imitations increased, suggesting unequal access to such prestigious goods.
 Yet even here the available data could be further interrogated (with, for instance, the number of imported items *per burial*).
@@
 The first includes an entry for each individual ceramic vessel.
-Variables includes the tomb, the vessel shape, its surviving dimensions, its ceramic period and corresponding absolute dating measures.
-In addition, all imports, 'oriental' imitations, and antique (that is, Bronze Age) artefacts were categorised, using the studies of Antoniadis (2012), Jones [-@Jones2000], and Crowe [-@Crowe2016].
+Variables include the tomb, the vessel shape, its surviving dimensions, its ceramic period and corresponding absolute dating measures.
+In addition, all imports, 'oriental' imitations, and antique (that is, Bronze Age) artefacts were categorised, using the studies of Antoniadis (2012), Jones [-@Jones2000], and Crowe [-@Crowe2016] respectively.
 In total, this database comprises just over 4540 individual artefacts.
@@
 The final database concerns the tombs themselves.
-Each tomb was accorded its own entry, with variables including dating (both in relative and absolute terms), surviving dimensions, and the known and estimated number of interments, as well as a tally of the objects recorded in the other two databases.
-Finally, the total number of imports, imitations and antiques in each tomb was calculated.
+Each tomb was accorded its own entry, with variables including dating (both relative absolute), surviving dimensions, and the known and estimated number of interments, as well as tallies of the various artefact types recorded in the other two databases.
 
@@
 Ever more archaeologists have begun to advocate the utility, indeed the necessity, of making available not just the data, but the full methodologies underpinning our academic output.
-With the opportunities afforded by current technologies (digital databases, statistical software packages, online repositories etc.), there remain ever fewer obstacles to the full and free sharing of both data and workflows [see @article].
+With the opportunities afforded by current technologies (digital databases, statistical software packages, online repositories etc.), there remain ever fewer obstacles to the full and free sharing of both data and workflows [@article].
 In Marwick's words [-@Marwick2017: 445], "[t]he technical problems are largely solved; the challenge now is to change the norms of the discipline to make high reproducibility a canonical attribute of high-quality scholarly work".
@@
 
-```{r burialestimate, fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="goes around here"}
+```{r burialestimate, fig.dim = c(7, 4), dpi=600, fig.align='center', fig.cap="goes around here"}
 
@@
 For most diachronic analyses, however, pithoi alone were taken as the best proxy for burial numbers.
-In most cases, any tombs lacking identified remains or a single vessel suitable for use as an urn were not considered, thus excluding what are presumably among the most heavily damaged and looted tombs.
+In most cases, any tombs lacking identified remains or a single vessel suitable for use as an urn were not considered, thus excluding the most heavily damaged and looted tombs.
 
@@
 This situation results in major challenges to the visualisation and analysis of such data.
-A simple, if coarse, solution is to use the midpoint of the range assigned to each vessel (i.e.
+A simple, if coarse, solution is to take the midpoint of the range assigned to each vessel (i.e.
 825 for our PGB example).
@@
 
-```{r allpithoi,fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="goes around here"}
+```{r allpithoi,fig.dim = c(7, 4), dpi=600, fig.align='center', fig.cap="goes around here"}
 
@@
 Here, rather than give a pot a single date, the entire time span under investigation is divided into equal chunks (in this case, decades), and probability values summing to one are then assigned for each vessel, per chunk of time, based on its accordant date range.
-So, our PGB vessel, datable between 840 and 810 BC, would register a value of 0.33 for each of those three decades.
+So, our PGB vessel, datable between 840 and 810 BC, would register a value of 0.33 for each of the three decades this encompasses.
 These values for individual vessels can then simply be summed to produce tomb- or cemetery-level signatures.
@@
 
-```{r aoristallcems, fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="goes around here"}
+```{r aoristallcems, fig.dim = c(7, 4), dpi=600, fig.align='center', fig.cap="goes around here"}
 
@@
 As seen in Figure \@ref(fig:aoristallcems), the artificial peaks and troughs of the simple averaging method are avoided, and different patterns become visible.
-Though these methods present their own limitations (see @Crema2010, 1123-24), they are considered a more robust and productive approach to the temporal uncertainty of ceramic dating and, as such, are used for much of the present analysis.
+Though these methods present their own limitations [@Crema2010, 1123-24], they are considered a more robust and productive approach to the temporal uncertainty of ceramic dating and, as such, are used for much of the present analysis.
 
@@
 I then examine these more closely, in an attempt to understand what sets them apart from the rest.
-Tomb dimensions, rates of burial, and changes in the composition of assemblages through time are all considered, to produce an account which speaks to the changing social landscape of EIA Knossos, drawing together issues of group identity, external relations, and the communication of wealth and status through funerary practice.
+Tomb dimensions, artefact distributions, and changes in the composition of assemblages through time are all considered, to produce an account which speaks to the changing social landscape of EIA Knossos, drawing together issues of group identity, external relations, and the communication of wealth and status through funerary practice.
 
@@
 An obvious place to begin is with the chronology of the burials themselves.
-The rate and number of interments being made through time provides the background against which to consider all other developments.
-
-```{r aoristbycem, dpi=300, fig.align="center", fig.cap="goesaround here.", fig.dim=c(8,4)}
+The rate and quantity of interments being made through time provides the background against which to consider all other developments.
+
+```{r aoristbycem, dpi=600, fig.align="center", fig.cap="goesaround here.", fig.dim=c(8,4)}
 KNC<-subset(tombs,tombs$cem_condensed=="KNC")
@@
 Figure \@ref(fig:aoristbycem) presents the aoristic sums of pithoi (as a proxy for burials) and of all vessels, respectively.
-Despite differences in scale and timing, it is immediately apparent that Fortetsa and the KNC experieced similar chronological developments in burials and, to a slightly lesser degree, grave goods.
+Despite differences in scale and timing, it is immediately apparent that Fortetsa and the KNC underwent similar chronological developments with regard to burials and, to a slightly lesser degree, grave goods.
 In both cemeteries, the rate of cremation burial began rising around the end of the 9th century, plateaued slightly in the early 8th, and then rose again to a peak in the late 8th and early 7th.
@@
 
-```{r tombconstruction, fig.dim=c(7.5,4), dpi=300, fig.cap="goes around here"}
+```{r tombconstruction, fig.dim=c(7.5,4), dpi=600, fig.cap="goes around here"}
 
@@
 
-```{r tombsinuse, fig.dim=c(7,4), dpi=300, fig.cap="goes around here"}
+```{r tombsinuse, fig.dim=c(7,4), dpi=600, fig.cap="goes around here"}
 
@@
 
-```{r tombsinusebybuilddate, fig.dim=c(7,4), dpi=300, fig.cap="goes around here"}
+```{r tombsinusebybuilddate, fig.dim=c(7,4), dpi=600, fig.cap="goes around here"}
 tombs_11th<-subset(tombs,tombs$century_built=="11th")
@@
 By dividing the tombs in use into the centuries during which they were built (Figure \@ref(fig:tombsinusebybuilddate)), we see a wave-like pattern with each cohort steadily going out of use as new ones were constructed.
-This would seem to accord with Cavanagh's (1996) suggestion that "[t]he length of time the tombs last seems [...] to reflect the duration of the social group which had access to the sepulchre", and that, if such access was lineally inehrited, frequent fission would result in limited trans-generational use.
+This would seem to accord with Cavanagh's (1996: 659) suggestion that "[t]he length of time the tombs last seems [...] to reflect the duration of the social group which had access to the sepulchre", and that, if such access was lineally inehrited, frequent fission would result in limited trans-generational use.
+But note also that the total increase in active tombs during the 8th century belies the diminishing number of continuing 9th and 10th century tombs.
 We might expect, then, that it was the 8th century tombs coming into use that spurred the dramatic rise in burial rates previoulsy noted.
 
-```{r potsbydateandtombdate,  fig.dim=c(7,4), dpi=300, fig.cap="goes around here"}
+```{r potsbydateandtombdate,  fig.dim=c(7,4), dpi=600, fig.cap="goes around here"}
 
@@
 However, we see from Figure \@ref(fig:potsbydateandtombdate) that, although the late 8th and early 7th centuries represented the apex of burial activity across both cemeteries, it was tombs built back in the 9th and 10th which received the lion's share of the burials and associated ceramics.
-This is all the more notable when, returning to Figure \@ref(fig:tombsinusebybuilddate), we see that the number of 9th and 10th century tombs in use by this time was, in fact, declining.
+This is all the more notable when, as seen in Figure \@ref(fig:tombsinusebybuilddate), they were, at this time, declining in number.
 
@@
 We are yet to draw out which older tombs precisely are responsible for the trend so far discussed.
-We might assume the tombs which became the focus of intense later burial activity would be those which yielded the most burials and grave goods when excavated and, happily, this supposition proves true.
-
-```{r toptombs, fig.dim = c(7, 5), dpi=300, fig.align='center',fig.cap="goes around here"}
+We might assume the tombs which became the focus of such intense later burial activity would be those which yielded the most burials and grave goods when excavated and, happily, this supposition proves true.
+
+```{r toptombs, fig.dim = c(7, 5), dpi=600, fig.align='center',fig.cap="goes around here"}
 
@@
 All but Tomb 285 date to the 9th century, and together they underpinned the discussed late spate of burials.
+
+```{r groups_a_and_b, fig.dim=c(7,5),dpi=600, fig.align='center',fig.cap="goes around here"}
+
+tombs$group<-ifelse(tombs$tomb=="P"|tombs$tomb=="292"|tombs$tomb=="75"|tombs$tomb=="107"|tombs$tomb=="285"|tombs$tomb=="II"|tombs$tomb=="218","A",
+                   ifelse(tombs$tomb=="X"|tombs$tomb=="G"|tombs$tomb=="219"|tombs$tomb=="Q"|tombs$tomb=="283","B","C"))
+
+GroupA<-subset(tombs,group=="A")
+
+aorist_GroupA <- colSums(GroupA[158:207])
+aorist_GroupA<-data.frame(aorist_GroupA)
+
+colnames(aorist_GroupA)<-c("aoristic_sum")
+row.names(aorist_GroupA)<-dates
+aorist_GroupA<-data.frame(aorist_GroupA)
+setDT(aorist_GroupA, keep.rownames = "decade")
+aorist_GroupA$decade<-as.numeric(aorist_GroupA$decade)
+
+aorist_GroupA$group<-"A"
+
+GroupB<-subset(tombs,tombs$group=="B")
+
+aorist_GroupB <-colSums(GroupB[158:207])
+aorist_GroupB<-data.frame(aorist_GroupB)
+
+colnames(aorist_GroupB)<-c("aoristic_sum")
+row.names(aorist_GroupB)<-dates
+aorist_GroupB<-data.frame(aorist_GroupB)
+setDT(aorist_GroupB, keep.rownames = "decade")
+aorist_GroupB$decade<-as.numeric(aorist_GroupB$decade)
+
+aorist_GroupB$group<-"B"
+
+aorist_groups<-rbind(aorist_GroupA,aorist_GroupB)
+
+ggplot(aorist_groups,aes(decade,aoristic_sum,fill=group))+geom_col(position="dodge")+
+  theme_bw()+theme_bw()+theme(text=element_text(family="Garamond", size=12))+
+                labs(x="Date (Years BC)", y="Aoristic Sum")+
+                  scale_fill_brewer(name="Group",palette = "Set2")+
+                   scale_y_continuous(expand=c(0,0),limits=c(0,120))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+```
+
 Interestingly, Tombs X, G, Q, 219 and 283 ('Group B'), however, evince a different pattern, with their respective peaks coming around a century earlier, which, as seen in Figure \@ref(fig:aoristbycem), was another period of increased burial activity.
-So despite receiving the greatest quantities of burials and grave goods, these tombs nonetheless seem to have achieved that distinction via different routes.
-And, indeed, the differences between them may shed light on the late surge in burial activity here discussed.
+So despite receiving the greatest quantities of burials and grave goods, these tombs nonetheless seem to have achieved that distinction via quite different histories of use, a divergence made clear in Figure \@ref(fig:groups_a_and_b).
+It is notable that tombs assinged to each group are seen in both cemeteries, suggesting from the outset that the differences between these may speak to behaviours and social contexts operating across a significant portion of the Knossian community.
 
@@
 
-The indices employed so far – the distribution of artefacts between tombs, and the construction and use of the tombs through time – inform us of the temporal or aggregate features of the cemeteries, but tell us little of the tombs as they were built or first used.
-In attempting more synchronic comparisons, we shall first consider the physical dimensions of the tombs as built - a potentially rich context for the negotiation of social power, status and group membership.
-
-```{r dimensionsboxplots, fig.dim=c(6,5),dpi=300, fig.align='center',fig.cap="goes around here"}
+Having addressed some broad diachronic trends, let us now attempt some more synchronic comparisons,   beginning with the physical dimensions of the tombs as built - a potentially rich context for the negotiation of social power, status and group membership.
+Perhaps the most bountiful tombs were also physically the largest, assertions of status and authority embodied in the man-power invested in their construction.
+
+```{r dimensionsboxplots, fig.dim=c(6,5),dpi=600, fig.align='center',fig.cap="goes around here"}
 
@@
 The 9th-century tombs, though, which include all but two of the 12 discussed above, do not particularly stand out.
-It does seem that chambers were, on average, larger in this period as well as exhibiting greater variety than at any other time.
+It does seem that chambers were, on average, larger in this period, as well as exhibiting greater variety than at any other time.
 Then again, even the largest chambers are not exactly cavernous, and the extra labour expended on construction would have been limited.
@@
 
-```{r tombdimensions, fig.dim = c(7.5, 4), dpi=300, fig.align='center', fig.cap="goes around here"}
-
-tombareas<-ggplot(subset(tombs,tombs$est_burials>1),aes(chamber_area,est_burials,color=cem_condensed))+geom_point()+
-  theme_bw()+theme(legend.title=element_blank(),
-        text=element_text(family="Garamond", size=12))+
-            labs(x=bquote('Chamber Area ('*m^2*')'), y="Estimated Total Burials")+
-               scale_color_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+
+```{r tombgroupdimensions, fig.dim = c(7, 4), dpi=600, fig.align='center', fig.cap="goes around here"}
+
+ggplot(subset(tombs,century_built=="9th" | century_built=="10th"),aes(dromos_length,chamber_area,color=group,label=tomb))+geom_point()+
+  theme_bw()+theme(text=element_text(family="Garamond", size=12))+
+  geom_label_repel(data=subset(tombs,group!="C"),aes(fill=group),color="white",show.legend = FALSE,family="Garamond")+
+            labs(x="Dromos Length (m)", y="Estimated Total Burials")+
+               scale_color_brewer(palette = "Set2",name="Group")+
+               scale_fill_brewer(palette = "Set2")+
+                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,10))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))+coord_fixed()
+
+```
+
+Looking more specifically at the tombs of Groups A and B, against the backdrop of the other 10th and 9th century tombs, there are perhaps hints of differentiation.
+All of them fall within the top 50% based on chamber area, and three have exceptionally long *dromoi*.
+But, later destruction notwithstanding, there appears nothing systematic about the variation visible here.
+If the size of one's tomb was intended to evoke or assert ideas of wealth or status, this appears to have occurred on a largely independent basis.
+
+```{r tombdimensions, fig.dim = c(7.5, 4), dpi=600, fig.align='center', fig.cap="goes around here"}
+
+tombareas<-ggplot(subset(tombs,tombs$est_burials>1),aes(chamber_area,est_burials,color=group))+geom_point()+
+  theme_bw()+theme(text=element_text(family="Garamond", size=12))+
+            labs(x="Chamber Area (sq m)", y="Estimated Total Burials")+
+               scale_color_brewer(palette = "Set2",name="Group")+
                 scale_x_continuous(expand=c(0,0),limits=c(0,10))+
@@
 
-dromoslengths<-ggplot(subset(tombs,tombs$est_burials>1),aes(dromos_length,est_burials,color=cem_condensed))+geom_point()+
-  theme_bw()+theme(legend.title=element_blank(),
-        text=element_text(family="Garamond", size=12))+
+dromoslengths<-ggplot(subset(tombs,tombs$est_burials>1),aes(dromos_length,est_burials,color=group))+geom_point()+
+  theme_bw()+theme(text=element_text(family="Garamond", size=12))+
             labs(x="Dromos Length (m)", y="Estimated Total Burials")+
-               scale_color_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+
+               scale_color_brewer(palette = "Set2",name="Group")+
                 scale_x_continuous(expand=c(0,0),limits=c(0,10))+
@@
 
-Furthermore, it seems that no clear relationship existed between the physical size of a tomb and its eventual buried population (see Figure \@ref(fig:tombdimensions)).
-True, over half of all tombs do not exceed 2.6 sq m in area, while nine of the ten tombs with the most burials do.
-But many of the very largest tombs are decidedly middling when it comes to burial numbers.
-Things are even less clear for the *dromos*; the largest two tombs by burial numbers also had exceptionally long *dromoi*, but the next eight tombs with the longest *dromoi* do not rank highly on interments.
-
-It seems dubious, then, whether the scale of the tombs when built carried overt social meanings but, even if it did, this proved irrelevant to their later reuse.
-Several tombs had niches cut into their *dromoi*, or the *stomion* extended, to fit in later burial urns.
-Sometimes, urns from within the chamber were relocated to such niches to make space for new interments (e.g.
-T.
-TFT; Brock 1957: 3–4).
-The impression is of competition for inclusion into the sepulchral group, further implying the tombs’ original builders little anticipated their eventual popularity.
+What we can say is that no clear relationship existed between the physical size of a tomb and its eventual buried population (see Figure \@ref(fig:tombdimensions)).
+Many of the very largest tombs are decidedly middling when it comes to burial numbers, and vice versa.
+So whatever the scale of the tombs connoted when built, this proved irrelevant to their later reuse.
+Several had niches cut into their *dromoi*, or the *stomion* extended, to fit in later burial urns (including T.
+292, 283 and 107; Cavanagh 1996: 658) and, sometimes, urns from within the chamber were relocated to these to make space for new interments (including T.
+75 and TFT; ibid.: 658; Brock 1957: 3–4).
+The impression is of competition for inclusion into the sepulchral group, further confirming that the tombs’ original builders little anticipated their eventual popularity.
+Whatever led to Group A's curious later use was, it was not presaged by exceptional scale or physical grandeur.
 
@@
 
-Dimensions aside, we can still consider the early use of these tombs through comparisons between the scale and composition of their respective assemblages through time.
-
-```{r tombscompthroughtime, fig.dim=c(7,4),dpi=300,fig.align='center',fig.cap="goes around here"}
+Dimensions aside, we can still investigate the contemporary use of these tombs through comparisons between the scale and composition of their respective assemblages through time.
+
+```{r tombscompthroughtime, fig.dim=c(7,4),dpi=600,fig.align='center',fig.cap="goes around here"}
 
@@
                     scale_y_continuous(expand=c(0,0))+facet_grid(~century,scales="free_x",space="free")
-```
-
-And in Figure \@ref(fig:tombscompthroughtime), we see that the tombs of Group A were by no means exceptional in the scale of their depositions during the 9th century, only diverging significantly from the rest in the subsequent two hundred years.
-Group B, meanwhile, had their respective peaks in the 9th or 8th centuries, with a near total decline afterwards.
-That is, they seem to have followed a more conventional trajectory, going out of use as new tombs were built.
-As a working hypothesis, let us suppose that these two distinct trajecotire correspond to behaviours born of differing socio-politcal and economic contexts.
-How far might this hypothesis take us in explicating these patterns, and how visible might it be?
-
-```{r bigbox, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="goes around here"}
+
+```
+
+And in Figure \@ref(fig:tombscompthroughtime), we see that the tombs of Group A, while not paultry even in their early years, were by no means exceptional during the 9th century.
+They clearly kept pace with the other popular tombs of the time, but their major divergence, their near monopoly, only emerged in the subsequent two hundred years.
+Group B, meanwhile, experienced similar, if not greater, use in the 9th and early 8th centuries, and then declined.
+That is, these tombs seem to have followed a more conventional trajectory viz.
+\@ref(fig:tombsinusebybuilddate).
+What might account for this disparity?
+And if the patterns implied here in \@ref(fig:tombscompthroughtime) - relative diversity, giving way to effective bipolarity in the funerary rite - and their emobdiment in Groups A and B respectively are genuine, how might we further investigate their manifestation across the excavated assemblage?
+
+## A Tale of Two Periods
+
+```{r bigbox, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="goes around here"}
 
@@
                scale_fill_brewer(name="Period",palette = "Set3")+
-                scale_x_reverse(expand=c(0,0))+
+                scale_x_reverse(expand=c(0,0),limits=c(1000,600))+
                   scale_y_continuous(expand=c(0,0),limits=c(0,20))
@@
 Figure \@ref(fig:bigbox) presents a boxplot of the total aoristic sums across both cemeteries, colour-coded by ceramic phase.
-Looking at the PGB and EG periods, we see that they exhibit the largest interquartile ranges (IQRs) of any period, and each possess only a single outlier (beyond 1.5 times the IQR).
-The Orientalising period, on the other hand, exhibits much reduced IQRs, and a proliferation of outliers far and above the range of the other tombs.
-In other words, whilst both the PGB-EG and EO-LO periods witnessed an expanding volume and range of burial activity, in the earlier period this was distributed more evenly across the total burying population.
+Looking at the PGB and EG, we see that they exhibit the largest interquartile ranges (IQRs) of any period, and each possess only a single outlier (beyond 1.5 times the IQR).
+The Orientalising period, on the other hand, exhibits much reduced IQRs, and a proliferation of outliers (that is, Group A) far and above the range of the other tombs.
+In other words, whilst both the PGB-EG and EO-LO periods witnessed an expanding volume and range of burial activity, in the earlier period this was distributed far more evenly across the total burying population.
 One speaks of increasing plurality, the other of divergence.
 
-```{r gini, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="goes around here"}
+```{r gini, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="goes around here"}
 
@@
 
-This characterisation is supported further by considering the Gini coefficient of the distribution of pottery between the tombs (see Figure \@ref(fig:gini).
+This characterisation is supported further by considering the Gini coefficient of the distribution of pottery between the tombs (see Figure \@ref(fig:gini)).
 Using aoristic sums as a probablistically-weighted estimate of the volume of material deposited in each tomb in each decade, the Gini coefficient can give an (albeit rough) estimate of how unequally distributed that material was.
@@
 
-This is not to suggest that the late 9th to early 8th century was a time of modest, egalitarian burial practices.
-The PGB and EG are consistently interpreted as periods of artistic development, increased overseas trade, and renewed vigour in the construction of tombs and the visibility of the burial rite.
+This is not to suggest that the late 9th to early 8th century was a time of understated, egalitarian burial practices.
+The PGB and EG are consistently interpreted as periods of artistic development, increased overseas trade, and renewed vigour in the construction of tombs and the visibility of the burial rite [@Brock1957: 143-44; @Coldstream1994: 110-12].
 The PGB style reflects a uniquely Cretan blend of Attic and Near Eastern forms, while the appearance of BA 'antiques' in the tombs from this period onwards is felt to reflect rejuvinated, even competitive interest in the Minoan past.
 
-```{r imports, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="goes around here"}
+```{r imports, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="goes around here"}
 
@@
 
+groupA7th<-subset(pots,group=="A" & century=="7th")
+groupA7thimports<-(sum(groupA7th$import=="Y")/nrow(groupA7th))*100
 ```
@@
 The rise in imported items is clearly seen in the tomb assemblages, with a succession of increased Greek and Near Eastern imports respectively in the late 9th and early 8th centuries (Figure \@ref(fig:imports)).
-Likewise, the concomitant decline in Near Eastern imports and rise in locally-made imitations thereof, is visible at the end of the 8th century.
-These trends being as a very coarse proxy for its engagement with the wider Mediterranean, the late 9th and early 8th centuries would seem to encompass the most outward-looking phase of the EIA Knossian community.
+Likewise, the decline in Near Eastern imports, and concomitant rise in locally-made imitations thereof, is visible at the end of the 8th century.
+These trends being a very coarse proxy for its engagement with the wider Mediterranean, the late 9th and early 8th centuries would seem to encompass the most outward-looking phase of the EIA Knossian community [see @Whitley2004: 434].
 This much has been relatively well understood for some time, at least at the KNC, but we can further examine how these imports were distributed among the various groups of tombs.
 
-We have seen that the Group B tombs all experienced their respective peaks in burial activity around the end of the 9th century, before declining in use, while those of Group A went on to receive ever more burials into the late 8th.
-It is notable, then, that among the Group B tombs, imported items constituted `r nrow(subset(pots,pots$group=="B" & pots$import =="Y"))/nrow(subset(pots,pots$group=="B"))*100`% of the total assemblage, compared to `r nrow(subset(pots,pots$group=="C" & pots$import =="Y"))/nrow(subset(pots,pots$group=="C"))*100`% of Group C and only `r nrow(subset(pots,pots$group=="A" & pots$import =="Y"))/nrow(subset(pots,pots$group=="A"))*100`% of Group A.
-This may further endorse the characterisation of the PGB-EG periods, wherein the Group B tombs flourished, and after which they declined, as a time of competitive display between burying parties.
-The contents of the Group A tombs, though probably once also shaped by such strategies, appear to have been diluted, as it were, by their later use which, as is becoming increasingly apparent, was of a different kind.
-
-```{r comps, fig.dim=c(7,4),dpi=300, fig.align='center',fig.cap="goes around here"}
+In the Group B tombs, with their more limited flourescence centred on the PGB-EG periods, imported items constitute `r nrow(subset(pots,pots$group=="B" & pots$import =="Y"))/nrow(subset(pots,pots$group=="B"))*100`% of the total excavated assemblage, compared to `r nrow(subset(pots,pots$group=="C" & pots$import =="Y"))/nrow(subset(pots,pots$group=="C"))*100`% in Group C and only `r nrow(subset(pots,pots$group=="A" & pots$import =="Y"))/nrow(subset(pots,pots$group=="A"))*100`% in Group A.
+Indeed, among 7th century vessels, when interments were higher than ever among Group A, the figure is a mere`r groupA7thimports`%.
+Thus, impression grows ever stronger, that the late proliferation of burials in the Group A tombs was altogether different to the comparative diversity of interments a century earlier.
+
+```{r comps, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="goes around here"}
 
@@
 The proportion of grave goods comprising 'other' vessels - pyxides, askoi, trays, stands, dinoi, kernoi, kraters and more - declines to become only a minor part of the standardised repertoire of the 7th century.
-The tombs of Group A did not break with this mould, and their assemblages seem as conventional as any other during the last century of use in the cemeteries.
+The tombs of Group A did not break with this mould, and their assemblages seem largely as conventional as any other during the final century of use in the cemeteries.
 
@@
 There is, perhaps unsurprisingly, a relatively strong (Spearman's ρ=`r spearmanobjectspots`) correlation between the number of ceramic and non-ceramic finds across the tombs, and, certainly, many of the most seemingly precious objects were recovered from among the larger tombs.
-But the monopoly those tombs would appear to have over such objects, perhaps tempting one initially to speak of 'elite' or 'high-status' tombs, evaporates when the number of those objects is divided by the number of burials represented.
+But the near monopoly those tombs would appear to have over such objects, perhaps tempting one initially to speak of 'elite' or 'high-status' tombs, evaporates when the number of those objects is divided by the number of burials represented.
 This fact has seldom, if ever, been appreciated in prior discussions of the cemeteries.
 
-```{r hists, fig.dim=c(6,4),dpi=300, fig.align='center',fig.cap="goes around here"}
+```{r hists, fig.dim=c(6,4),dpi=600, fig.align='center',fig.cap="goes around here"}
 tombs$group<-revalue(tombs$group,c("A"="A","B"="B","C"="B"))
@@
                     scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-total_metal<-ggplot(subset(tombs,est_burials>1 & total_metal>0),aes(total_metal, fill=group))+geom_histogram()+
+total_weapon<-ggplot(subset(tombs,est_burials>1 & weapon>0),aes(weapon, fill=group))+geom_histogram()+
   theme_bw()+theme(legend.title=element_blank(),
                text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Metal Artefacts", y="Number of Tombs")+
+                 labs(x="Total Weapons", y="Number of Tombs")+
                   scale_fill_brewer(palette = "Set2",labels=c(" Group A  ", " Other Tombs"))+
                     scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-total_import<-ggplot(subset(tombs,est_burials>1 & total_import>0),aes(total_import, fill=group))+geom_histogram()+
+total_jewellery<-ggplot(subset(tombs,est_burials>1 & jewellery>0),aes(jewellery, fill=group))+geom_histogram()+
   theme_bw()+theme(legend.title=element_blank(),
                text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Imports", y="Number of Tombs")+
+                 labs(x="Total Jewellery", y="Number of Tombs")+
                   scale_fill_brewer(palette = "Set2",labels=c(" Group A  ", " Other Tombs"))+
@@
                     scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-metalbyburial<-ggplot(subset(tombs,est_burials>1 & total_metal>0),aes(total_metal/est_burials, fill=group))+geom_histogram()+
+weaponbyburial<-ggplot(subset(tombs,est_burials>1 & weapon>0),aes(weapon/est_burials, fill=group))+geom_histogram()+
   theme_bw()+theme(legend.title=element_blank(),
                text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Metal Artefacts/Burials", y="Number of Tombs")+
+                 labs(x="Total Weapons/Burials", y="Number of Tombs")+
                   scale_fill_brewer(palette = "Set2",labels=c(" Group A  ", " Other Tombs"))+
@@
 
-importsbyburial<-ggplot(subset(tombs,est_burials>1 & total_import>0),aes(total_import/est_burials, fill=group))+geom_histogram()+
+jewellerybyburial<-ggplot(subset(tombs,est_burials>1 & jewellery>0),aes(jewellery/est_burials, fill=group))+geom_histogram()+
   theme_bw()+theme(legend.title=element_blank(),
                text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Imports/Burials", y="Number of Tombs")+
+                 labs(x="Total Jewellery/Burials", y="Number of Tombs")+
                   scale_fill_brewer(palette = "Set2",labels=c(" Group A  ", " Other Tombs"))+
@@
 
-ggarrange(total_import,total_metal,total_object,importsbyburial,metalbyburial,objectsbyburial,ncol=3,nrow=2,common.legend = TRUE,legend="bottom")
-```
-
-Figure \@ref(fig:hists) thus supports the view which has emerged from the foregoing analysis, that the major increase in burial activity at both Fortetsa and the KNC was not a phenomenon linked in any straightforward way to simple principles of 'elite' or 'wealthy' burial.
+ggarrange(total_jewellery,total_weapon,total_object,jewellerybyburial,weaponbyburial,objectsbyburial,ncol=3,nrow=2,common.legend = TRUE,legend="bottom")+
+            theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+```
+
+Figure \@ref(fig:hists) thus supports the view which has emerged from the foregoing analysis, that the late surge in burial activity at both Fortetsa and the KNC was not a phenomenon linked in any straightforward way to simple principles of 'elite' or 'wealthy' burial.
 To summarise, towards the end of 8th century BC, a select group of tombs, all of which had been constructed back in the 9th or 10th, became the foci of a major increase in burials, something which did not occur among the tombs built around that time.
@@
 
-It remains to consider how we might characterise the burying groups responsible behind these various, intriguing trends.
+It remains to consider how we might characterise the burying groups behind these various, intriguing trends.
 The traditional view is that such groups were immediate families (Brock 1957, 41; @Boardman1967: 63; @Coldstream1996: 291).
@@
 Certainly, it seems illogical that population increases would lead to the consolidation of burials within only extant tombs.
-The trend instead must reflect reconfigurations of the burying groups themselves, which Kotsonas feels may have, also or instead, comprised age- and sex-grades.
+The trend instead must reflect reconfigurations of the burying groups themselves, which Kotsonas feels may have (also or instead) comprised age- and sex-grades.
 
@@
 In the traditional view, political authority was concentrated in the hands of the *kosmoi*, aristocratic magistrates selected from the leading, hereditary *startoi*, or ‘clans’, networks of extended lineages controlling the ancestral estates on which worked attendent serfs and/or slaves (Willetts 1982, 240).
-But despite elements of this account probably being accurate, others remain equivocal, such as the nature of the *pyla*, or 'tribe', and its relation to the *startos* (which may itself be a military grouping), the distinction between an *astos* and a *poliatas*, both clearly formulations of the rights-bearing civic actor, and the ancestry and spread of the *agelai*, seemingly an adolescent age-grade so far only attested at Eltynia [@Gagarin2016: 36; 78-80].
+But despite elements of this account probably being accurate, others remain equivocal, such as the nature of the *pyla*, or 'tribe', and its relation to the *startos* (which may itself be a military grouping), the distinction between *astos* and *poliatas*, both probably formulations of the rights-bearing civic actor, and the ancestry and spread of the *agelai*, seemingly an adolescent age-grade only attested at Eltynia [@Gagarin2016: 36; 78-80].
 Indeed, many terms which might illuminate social divisions and groupings within the Archaic Cretan populace are only attested at one or two sites, and even then sometimes appear to differ in their form or function.
 
-It wouldn't be helpful at present, then, to draw any simple parallel between the Knossian burial groups and, say, powerful *startoi*, a term which is not even attested outside of Gortyn in this period (ibid:, 1-2).
-But archaeological and textual evidence suggests that, in general terms, such supra-familial kinship groups were undergoing a transition during this period, suggesting some such attribution for the burying groups at Knossos is not unreasonable, given the noted changes there also.
-Elsewhere on Crete, we see a shift to unmarked, individual burials, and more austere grave goods, as well as an apparent reorientation of sanctuary deposits away from extra-mural to suburban spaces, and a rise in utilitarian, mass-produced votives (ibid.: 34-6).
+My aim, however, is not to draw a simple parallel between the Knossian burial groups and, say, powerful *startoi*, a term which is not even attested outside of Gortyn in this period (ibid:, 1-2).
+The fundamental conclusion of the disucussed contrasts between the late 9th and late 8th centuries, between the Group A and B tombs, is not that the burying groups (whatever their composition) changed, as such, it is that the socio-economic and political contexts did, in ways which prompted reformulations of funerary practice, one of which being the intense, iterative burial activity evinced by the patrons of the Group A tombs.
+Such inferences, naturally, recall the conclusions of Morris [-@Morris1987: 171-211], who argued 30 years ago that the emergence of the *polis* across the Aegean led to a tension between elite preferences for ostentatious display, and a downplaying of social difference in death, as promoted by the ‘middling’ ideology that the *polis* engendered and promulgated.
+In his well-known model of the *agathoi* and *kakoi*, Morris argued that, at various times, the relevant spheres of kinship demonstrated in burial could widen and constrict, with more and less distantly related lineages drawn into association through changing social strategies, until the pre-eminence of kinship relations in the structuring of political power declined, along with its manifestiations in mortuary practice.
+
+At Knossos, some aspects of the noted expansion in the burying population seem explicable within a model like Morris'.
+The basic principle, of elites employing access to burial as a strategic act of association and social signalling, is almost certainly in evidence here.
+That the noted spate of burials in the late 8th to early 7th centuries was concentrated solely in older tombs strongly implies that issues of heritage, lineage, and tradition were at play.
+Indeed, the stratigraphy of the niches dug into the *dromos* of tomb 107 imply that depositions continued to be made even after the main chamber collapsed.
+The symbolic, and not the physical, presence these tombs seem to have asserted appears far more instrumental in their extensive reuse, seen also in the lack of correlation between dimensions and ultimate number of burials.
+
+Other features sit less easily with a Morrissian reading of the evidence, though.
+Firstly, the burials at issue are in many ways *not* ostentatious in comparison to the norm.
+We have seen that the 'richest' tombs in terms of finds are far less so when one accounts for the number of burials represented.
+Likewise, Morris claims that periods of expansion in the rite of burial will correspond to increased diversity in the associated assemblages.
+Yet, at Knossos, the range of grave goods diminished and crystallised during the late 8th and early 7th centuries.
+The PGB period seems a far better candidate for such an expansion, with its greater variety in chamber dimensions, relatively broad distribution of (local and imported) grave goods, not to mention the innovative ceramic decoration and reuse of BA artefacts noted elsewhere.
+In both the late 9th and late 8th centuries, then, certain burying groups found it propitious to broaden access to the tombs they patronised, but changing social contexts saw those expansions manifest in quite different associated assemblages.
+But whether the increasingly homogenous iteration of burials seen later reflected some standard set of accoutrements intentionally shared by the hypothetical *agathoi*, or else nascent, egalitarian strictures over elaborate burial remains difficult to judge.
+
+Elsewhere on Crete in the 7th century, we see a shift to unmarked, individual burials, and more austere grave goods, as well as an apparent reorientation of sanctuary deposits away from extra-mural to suburban spaces, and a rise in utilitarian, mass-produced votives [@Gagarin2016: 34-36].
 At Knossos, a small number of pithos burials in pits do appear in the KNC and Fortetsa volumes (Brock 1957: 98; Coldstream and Catling 1996: 162), and recent re-excavation of the Teke plot identified numerous small pit burials around three chamber tombs [@Evely2012], raising the possibility that others exist, but have gone unrecognised in prior excavations.
-At Azoria, major urban remodelling in the 7th century saw the earlier, agglutinative house forms of neighbouring Kavousi Vronda and Kastro give way to larger, more formally planned dwellings with little evidence of renovation or remodelling, while larger, apprently civic, buildings were constructed to accommodate major public gatherings [@Haggis2011; @Haggis2011b; @Haggis2014].
-
-In the Knossian cemeteries, we have seen a somewhat diverse set of burial practices give way to a more prolific, yet more restricted, enumeration of such rites, focussed on a small number of long-established tombs, in the context of diminishing collective tomb construction, and shortly before a complete cessation of such depositions.
-Together, these varied threads suggest that with the emergence of the archaic *poleis* on Crete, there was a concomitant restructuring of wider social groups, institutions, and behaviours.
+At Azoria, major urban remodelling in the 7th century saw the earlier, agglutinative house forms of neighbouring Kavousi Vronda and Kastro give way to larger, more formally planned dwellings with little evidence of renovation or remodelling, while larger, apparently civic, buildings were constructed to accommodate major public gatherings [@Haggis2011; @Haggis2011b; @Haggis2014].
+
+Together, these varied threads suggest that, with the emergence of the archaic *poleis* on Crete, there was a concomitant restructuring of wider social groups, institutions, and behaviours.
 It seems not unreasonable to infer an emergent emphasis on the citizen, the state, and ideologies of relative egality, within which communal gatherings, intra-mural ritual, and individual graves replaced ostentatious, kin-centred acts of gathering, dedication, and burial.
-
-These issues, naturally, recall the conclusions of Morris [-@Morris1987: 171-211], who argued 30 years ago that the emergence of the *polis* across the Aegean led to a tension between elite preferences for conspicuous display, and a downplaying of social difference in death, as promoted by the ‘middling’ ideology that the *polis* engendered and promulgated.
-In his well-known model of the *agathoi* and *kakoi*, Morris argued that, at various times, the relevant spheres of kinship demonstrated in burial could widen and constrict, with more and less distantly related lineages drawn into association through changing social strategies, until the pre-eminence of kinship relations in the structuring of political power declined, along with its manifestiations in mortuary practice.
-
-At Knossos, while some aspects of the noted expansion in the burying population seem explicable within a model like Morris', others sit more uneasily.
-Firstly, the burials at issue here are in many ways *not* ostentatious in comparison to the norm.
-We have seen that the 'richest' tombs in terms of finds are far less so when one accounts for the number of burials represented.
-Likewise, Morris claims that periods of expansion in the rite of burial correspond to greater variety in the associated assemblages, yet, at Knossos, we have seen that the range of grave goods in fact diminished and crystallised during the late 8th and early 7th centuries.
-But whether this reflected some standard set of accoutrements intentionally shared by the hypothetical *agathoi* of Knossos, or else nascent, egalitarian strictures over elaborate burial is hard to judge.
-The PGB period seems a far better candidate for such an expansion, with its greater variety in chamber dimensions, relatively broad distribution of (local and imported) grave goods, not to mention the innovative ceramic decoration and reuse of BA artefacts noted elsewhere.
-One thing is clear; the social context which produced these signatures contrasted markedly with that of a century later, with its ever greater volume of interments in an ever-smaller cohort of tombs, its more uniform assemblages, and its decline of imported items.
-
-It is tempting, though potentially perilous, to link the diminishing range of burial goods to Archaic Crete's purported austerity, its (quite literally) spartan lifestyles.
-But, as a more nuanced picture emerges of the island's relationship with the wider Mediterranean, subtler characterisations are possible.
-It is telling that the peak for imported items in the tombs came in the 9th century; the Orientalising as a time period feels increasingly misapplied to Crete, when, at Knossos at least, it witnessed far lower rates of 'oriental' imports, and only limited adoption of the figural motifs which characterise its appearance elsewhere.
-This decline in imports was, however, counterpointed by a rise in imitations of Near Eastern vessels [@Antoniadis2012: 170], something which further accords with our growing awareness of the selective nature of Archaic and Classical Crete's external relations.
-Erickson [@Erickson2010] has, likewise, argued that *symposia* were practised on Crete, something often disputed, but that their practitioners merely favoured simpler vessel forms than the figurally-decorated wares popular on the mainland.
+And in the Knossian cemeteries, we have seen a somewhat diverse set of burial practices, including the deposition of Minoan antiques and imported items, give way to a more prolific, yet more restricted, suite of funerary rites, focussed on a small number of long-established tombs, in the context of diminishing collective tomb construction, replacement of imports with locally-made imitations, and an imminent cessation of such depositions.
+
+Yet we need not adopt wholesale a model like Morris's "middling" ideology, nor link the diminishing range of burial goods to Archaic Crete's purported austerity, its (quite literally) spartan lifestyles.
+As a more nuanced picture emerges of the island's relationship with the wider Mediterranean, subtler characterisations are possible.
+It is telling that the peak for imported items in the tombs came in the 9th century; the Orientalising as a time period feels increasingly misapplied to Crete [@Whitley2004: 434], when, at Knossos at least, it witnessed far lower rates of 'oriental' imports, and only limited adoption of the figural motifs which characterise its appearance elsewhere.
+The decline in imports was, however, counterpointed by a rise in imitations of Near Eastern vessels [@Antoniadis2012: 170], further according with our growing awareness of the selective nature of Archaic and Classical Crete's external relations.
+Erickson [@Erickson2010], for instance, has argued that *symposia* were practised in Crete, something often disputed, but that their practitioners merely favoured simpler vessel forms than the figurally-decorated wares popular on the mainland.
 Local, selective adoption of, and responses to, wider social and material trends across the contemporary Aegean is perhaps becoming the hallmark of Archaic Crete, and finds further endorsement in the trends identified in the EIA cemeteries at Knossos.
-What seems certain is that this was a period of transition wherein, whatever their exact identity, certain actors invested heavily in a final flurry of burials in chamber tombs, with an emphasis on tradition, an established repertoire of items, rites, and venues for one's passage to the afterlife, before changing social forces effected their abandonment.
+What seems certain is that this was a period of transition wherein, whatever their exact identity, certain actors invested heavily in a final flurry of burials in chamber tombs, with an emphasis on a traditional, established repertoire of items, rites, and venues for one's passage to the afterlife, before changing social forces effected their abandonment.
 
@@
 It is notable how similar the observed trends are at both locations.
-That both underwent similar developments during the EIA suggests, not only that they catered to similar sub-sections of the Knossian population, but that some of the socio-economic and political changes inferred and discussed were real, and community-wide.
+That both underwent comparable developments during the EIA suggests, not only that they catered to similar sub-sections of the Knossian population, but that some of the socio-economic and political changes inferred and discussed were significant and community-wide.
 Further work should seek to extend these comparisons, both into trends unexplored in the present study, and out to the wider funerary landscape of Knossos.
-I hope to produced such a study but, for the time being, the datasets I am making available alongside this article include the other published EIA tombs in the area, so that others may also explore their possible significance.
-
-There were undeniable limitations to the present work.
+I hope to do just that but, for the time being, the datasets I am making available alongside this article include the other published EIA tombs in the area, so that others may also explore their possible significance.
+
+There were undeniable limitations to the present study.
 For one, non-ceramic artefacts were underrepresented due to difficulties with dating, and future work could seek to address this.
 Issues of gender were sadly little considered, though this was largely due to limitations in the data themselves.
-There have also been a number of tombs discovered in the last 20 years that fill in the spaces between the Teke plot and the Medical Faculty Site, and ideally these unpublished tombs should be included in further analyses as they, along with the unconsidered Teke tholos and associated tombs, are almost surely part of a single contiguous cemetery [@Blackman1999: 113; @Catling1983: 51; @French1991: 68-69].
-Finally, a preliminary study was recently taken by the author of select boxes marked as 'unpublished' from the KNC deposits kept at the British School at Athens at Knossos.
-Despite going unmentioned in the publication, this material was not inconsiderable and so, regretably, future study will be necessary to determine whether it threatens the integrity of quantitative analyses such as this.
+There have also been a number of tombs discovered in the last 20 years that fill in the spaces between the Teke plot and the Medical Faculty Site, and ideally these unpublished tombs should be included in further analyses as they, along with the unconsidered Teke tholos and associated tombs, are probably part of a single contiguous cemetery [@Blackman1999: 113; @Catling1983: 51; @French1991: 68-69].
+Finally, a preliminary study was recently undertaken by the author of select boxes marked as 'unpublished' from the KNC deposits kept at the British School at Athens at Knossos.
+Despite going unmentioned in the publication, this material was not inconsiderable and so, regretably, future work will be necessary to determine whether it threatens the integrity of quantitative analyses such as this.
 Suffice to say that, at present, it is another pinch of salt with which to take the present findings.
@@
 These limitations notwithstanding, the foregoing analysis demonstrates the value of quantitative approaches to the archaeological record of the period.
-With large datasets, we are able to locate and visualise covariance and significant correlations between different variables, plot changes in the data through time and, ultimately, draw on our findings to inform discussions of social processes.
 Though the construction of databases can be time consuming, once made they offer the opportunity to subdivide, compare, and plot the data in numerous ways quickly and easily, facilitating investigation of the material from multiple angles.
@@
 In both cases, tombs lacking any securely dateable pottery are excluded.*
-
 *Abbreviations*
@@
 
+*Figure 11 Aoristic sums for the Group A and Group B tombs, as defined above.*
+
 *Figure 11 Boxplots comparing various tomb dimensions, divided, in each case, by the century of tomb construction.*
 
-*Figure 12 Scatterplots comparing tomb dimensions (chamber area and *dromos* length) to the estimated total number of burials by tomb.*
-
-*Figure 13 Counts of vessels dating to the 9th, 8th and 7th centuries respectively, found in tombs built in the 10th or 9th century which yielded upwards of 50 vessels when excavated.
-The seven tombs with late peaks in burial activity (Group A) are shown in green, the remaining five of the largest 12 (Group B) in orange, and all other tombs (Group C) in blue.*
-
-*Figure 14 Aoristic sums of imported ceramics through time across both cemeteries, colour-coded by the broad regional origin of the vessels, and overlain on the grand totals in pale grey.*
-
-*Figure 15 Aoristic sums of imported vessels through time. The sums are colour coded as per Figure 13 (thoug here Group C represents all other tombs, not just those with 50+ vessels).*
-
-*Figure 16 Boxplot showing the aoristic sums of all vessels across both cemeteries through time, colour-coded by ceramic period.
+*Figure 12 Scatterplot comparing the dimensions of* dromos *length and chamber area.
+Groups A and B, defined earlier, are illustrated and labelled in green and orange respectively, while all remaining tombs ('Group C') are in blue.*
+
+*Figure 13 Scatterplots comparing tomb dimensions (chamber area and* dromos *length) to the estimated total number of burials by tomb. The points area colour-coded as per Figure 12.*
+
+*Figure 14 Counts of vessels dating to the 9th, 8th and 7th centuries respectively, found in tombs built in the 10th or 9th century which yielded upwards of 50 vessels when excavated.
+The tombs are colour-coded as per Figure 12.*
+
+*Figure 15 Boxplot showing the aoristic sums of all vessels across both cemeteries through time, colour-coded by ceramic period.
 Boxes represent the interquartile range, horizontal lines the median value, whiskers values within 1.5 times the interquartile range, and points outliers beyond this.
@@
 
-*Figure 17 Gini coefficient values through time based on the aoristic sums for all vessels across both cemeteries. Background colours correspond to ceramic periods.*
+*Figure 16 Gini coefficient values through time based on the aoristic sums for all vessels across both cemeteries. Background colours correspond to ceramic periods.*
+
+*Figure 17 Aoristic sums of imported and 'oriental' imitation vessels through time across both cemeteries, colour-coded by regional origin in the case of imports.*
 
@@
 
-*Figure 19 Histograms corresponding, first, to counts of non-ceramic artefacts, metal objects, and imported items by tomb across both cemeteries and, secondly, to those counts divided by the estimated number of burials in each tomb.*
+*Figure 19 Histograms corresponding, first, to counts of jewellery (beads, necklaces, rings, diadems etc.), metal objects, and imported items by tomb across both cemeteries and, secondly, to those counts divided by the estimated number of burials in each tomb.*
 
@@
 
-*Table 1 The number of pithoi (as a proxy for burials) deposited in tombs dating to the 10th, 9th, 8th and 7th centuries, through each of the centuries they were in use, as well as the number of tombs from which these burials came, allowing calculation of the changing rate of burial through time.*
+*Table 1 The number of pithoi (as a proxy for burials) deposited in tombs dating to the 10th, 9th, 8th and 7th centuries respectively.
+For each of the centuries the tombs were in use (column two), counts of pithoi (column three) are divided by the number of tombs from which these vessels came (column 4), allowing calculation of the changing rate of burial through time (column five).*
 

### 2019-06-06 analysis/paper/paper2.Rmd
Final edits

@@
 For one, though, this is not the Knossos of that illustrous, palatial period, but rather the town as it persisted and evolved during the subsequent Early Iron Age (EIA).
-Research into this once maligned period has grown apace in recent decades, beginning with the pioneering works of Coldstream [-@Coldstream1977], Desborough [-@Desborough1972], and Snodgrass [-@Snodgrass1971a] and culminating in the rejection by most scholars of the appelation and interpretive bagguage of the so-called ‘Dark Ages’ [see @Kotsonas2016; @Morris1997].
+Research into this once maligned period has grown apace in recent decades, beginning with the pioneering works of Coldstream [-@Coldstream1977], Desborough [-@Desborough1972], and Snodgrass [-@Snodgrass1971a] and culminating in the rejection by many scholars of the appelation and interpretive bagguage of the so-called ‘Dark Ages’ [see @Kotsonas2016; @Morris1997].
 Nonetheless, our knowledge of many EIA settlements remains underdeveloped in comparison with those of the second millennium BC.
@@
 
-However, a concentration of activity slightly to the west of the palace presented a good case for continuing, nucleated settlement [@Hood1981], something strongly advocated by Coldstream [-@Coldstream1984a; -@Coldstream2000], and, more recently, findings of the Knossos Urban Landscape Project suggest a PG settlement of up to 40ha [@Kotsonas2011a: 5-8].
+However, a concentration of activity slightly to the west of the palace presented a good case for continuing, nucleated settlement [@Hood1981], something strongly advocated by Coldstream [-@Coldstream1984a; -@Coldstream2000], and, more recently, findings of the Knossos Urban Landscape Project (KULP) suggest a PG settlement of up to 40ha [@Kotsonas2011a: 5-8].
 This estimate remains our best guess in the absence of substantial excavated deposits but, if accurate, suggests that EIA Knossos was one of the largest sites of the contemporary Aegean.
@@
 
-The cemetery on the slope facing the eponymous village of Fortetsa comprises twelve tombs excavated in 1933, and another eight uncovered in 1935 (Brock 1957: xi).
-These cluster in three main groups.
-The largest of these contains nine tombs, arranged on a north-south axis, with the *dromos* of each extending downslope to the west.
-To the north lies a cluster of five tombs and, to the south, another of three.
-Finally, three tombs excavated in 1933 – L, TFT and Π – which lie close to the main road north of the Acropolis hill [@Brock1957: 1-2], are now considered to belong the Knossos North Cemetery (KNC).
+The cemetery on the slope facing the eponymous village of Fortetsa comprises twelve tombs excavated in 1933, and another eight uncovered in 1935 (Brock 1957: xi), grouped into three main clusters running north to south.
+Three of those excavated in 1933 – L, TFT and Π – which lie close to the main road north of the Acropolis hill [@Brock1957: 1-2], are now considered to belong the Knossos North Cemetery (KNC).
 
 To the northeast of the Fortetsa cemetery, a collection of ten badly damaged tombs known as the Fortetsa 1967 tombs were probably the southernmost burials of the KNC.
-Unfortunately, all had been thoroughly looted at the time of excavation, and are thus excluded from present analysis.
+Unfortunately, all had been thoroughly looted at the time of excavation and are thus excluded from present analysis.
 Northwest of these, and just south of the Teke Tholos, the Teke tombs represent the northern reaches of the KNC.
@@
 
-The tombs themselves fall into three main types, the chamber tomb, shaft grave, and pit cave or tomb; the latter two are largely Subminoam phenomena, with the former by far the most abundant.
+The tombs themselves fall largely into three types, the chamber tomb, shaft grave, and pit cave or tomb; the latter two are largely Subminoam phenomena, with the former by far the most abundant.
 It comprises a descending ramp or passage, the *dromos*, leading to the chamber itself, cut into the *kouskouras* bedrock.
@@
 The causes of this lacuna, which persists until the recrudescence of archaeological evidence in the Late Archaic period, remain unknown.
-Although various explanations have been forwarded, [@Coldstream1999: 301-302; @Huxley1994: 126; @Kotsonas2002: 41-44], there are now strong reasons to doubt the settlement was abandoned (Whitelaw, pers comm.).
+Although various explanations have been forwarded [@Coldstream1999: 301-302; @Huxley1994: 126; @Kotsonas2002: 41-44], there are now strong reasons to doubt the settlement was abandoned (Whitelaw, pers.
+comm.).
 
@@
 A popular approach to the study of Knossos’ EIA tombs has been the investigation of ‘object biography’ [@Appadurai1986; @Kopytoff1986; @Mauss1954], an anthropological lens emphasising the mutuable nature of an object’s value, the transformations latent within its social existence.
-Fertilised by the Homeric motif of gift-giving [@Whitley2002: 220-21], and the remarkable archaeological parallels for such items – such as the boar’s tusk helmet from Tomb 200+ and that given to Odysseus by Meriones (*Iliad* 10.260-271) - such approaches have been brought to bear on certain elaborate burials of the EIA.
+Fertilised by the Homeric motif of gift-giving [@Whitley2002: 220-21], and the remarkable archaeological parallels for such items – such as the boar’s tusk helmet from KNC Tomb 201 and that given to Odysseus by Meriones (*Iliad* 10.260-271) - such approaches have been brought to bear on certain elaborate burials of the EIA.
 
@@
 
-Complementing the theme of heroes abroad has been a similar interest in foreigners at home.
+Complementing the theme of heroes abroad has been that of foreigners at home.
 Two caches of gold jewellery, found in pits flanking the entrance to the Teke Tholos, were argued by Boardman [-@Boardman1967] to be foundation deposits – a Near Eastern custom – for the tomb of an immigrant Phoenician craftsman.
-Kotsonas has revisited this attribution, deeming the tholos more probably that of an elite individual or family with a monopoly over a metal workshop’s output [-@Kotsonas2006: 155–9].
+Kotsonas has revisited this attribution, deeming the tholos more probably that of an elite individual or family with a monopoly over a metal workshop’s output [-@Kotsonas2006: 155–59].
 While certainty in such individual cases may be moot, it seems unlikely that foreign traders or craftsmen never visited or resided at Knossos during the EIA [@Hoffman1997: 176–85; @Schreiber2003: 293-306].
@@
 
-Generally speaking, the above analyses have tended to rely more on particular finds, tombs, or observable, but limited, patterns in the data; only a few authors have pursued more quantative approaches.
+Generally speaking, the above analyses have tended to rely on particular finds, tombs, or observable, but limited, patterns in the data; few authors have pursued more quantative approaches.
 Cluster analyses have been employed by both Cavanagh [-@Coldstream1996: 653-57] and Antoniadis [-@Antoniadis2012: 193-97], the most sophisticated statistical techniques so far applied to the material.
 Antoniadis [-@Antoniadis2012: 172-76] also divided the EIA tombs of Knossos into three groups, those with fewer than 10 pots, fewer than 50, and more than 50, and found that, in each successive case, the proportions of tombs containing 'oriental' imports and imitations increased, suggesting unequal access to such prestigious goods.
-Yet even here the available data could be further interrogated (with, for instance, the number of imported items *per burial*).
-The present paper thus seeks to extend and nuance a number of observations that have been made in previous studies, as well as revealing patterns thus far unacknowledged, through use of a more systematic approach to quantitative analysis.
+Yet even here the data could be further interrogated (with, for instance, the number of imported items *per burial*).
+The present paper thus seeks to extend and nuance a number of observations that have been made in previous studies, as well as revealing patterns thus far unacknowledged, through a more systematic quantitative analysis.
 
@@
 The second database is similar, but with each entry corresponding to a find other than a ceramic vessel.
-A ‘material’ variable was added, as of course these finds were not all made of the same substance.
+A ‘material’ variable was added as, of course, these finds were not all made of the same substance.
 This database included some 1620 individual artefacts.
@@
 Taken together, these databases make possible a diverse range of quantitative analyses.
-It must be conceded that, with many of the tombs evidently looted or damaged, this cannot be a complete picture of their original contents.
+Sadly, with many of the tombs evidently looted or damaged, this cannot be a complete picture of their original contents.
 If we assume, though, that natural or accidental destruction was relatively indiscriminate, and that, even in cases of looting, certain object classes (particularly cinerary urns) were seldom robbed, then the data are by no means rendered useless.
@@
 This article is, among other things, an attempt to bring these principles of maximal reproducibility to the archaeology of EIA Crete.
-Within this field, catalogues and excavation reports abound and, though many of these may be forgiven based on their dates of publication, the lack of readily utilisable digital data from a number of recent articles and monographs, indeed, the failure to even present data such that they could be manually digitsed, is less pardonable.
+Within this field, catalogues and excavation reports abound and, though many of these may be forgiven based on their dates of publication, the lack of readily utilisable digital (or even digitisable) data from a number of recent articles and monographs is less pardonable.
 
@@
 
-In this vein, the present article has been composed making use of Ben Marwick's 'rrtools' [see @Marwick2017; @inbook], an open-source package for the statistical software R Studio.
+In this vein, the present article has been composed making use of Ben Marwick's 'rrtools' [@Marwick2017; @inbook], an open-source package for the statistical software R Studio.
 This provides the author with the tools to create a self-contained 'compendium', which can be freely downloaded, and from which anyone can reproduce the entire workflow of the following analysis.
@@
 
-But the artificial clustering this technique engenders around select absolute dates can prove misleading, as in Figure \@ref(fig:allpithoi); the apparent cessation of depositions for 30 years in the 8th century is simply a fiction generated by the processing of the data.
-
-A alternative approach, and the method employed extensively in this paper, is the use of so-called 'aoristic analysis' [for original use in criminology, see @Ratcliffe2000; for archaeological applications, see @Crema2010; @Crema2012; @Johnson2004a].
-Here, rather than give a pot a single date, the entire time span under investigation is divided into equal chunks (in this case, decades), and probability values summing to one are then assigned for each vessel, per chunk of time, based on its accordant date range.
+But the artificial clustering this technique engenders around select absolute dates can prove misleading, as in Figure \@ref(fig:allpithoi); the apparent cessation of depositions in the 8th century is simply a fiction generated by the processing of the data.
+
+A alternative approach, employed extensively in this paper, is the use of so-called 'aoristic analysis' [for original use in criminology, see @Ratcliffe2000; for archaeological applications, see @Crema2010; @Crema2012; @Johnson2004a].
+Here, rather than give a pot a single date, the entire time span under investigation is divided into equal chunks (in this case, decades), and probability values summing to one are then assigned per chunk of time, based on the vessel's date range.
 So, our PGB vessel, datable between 840 and 810 BC, would register a value of 0.33 for each of the three decades this encompasses.
 These values for individual vessels can then simply be summed to produce tomb- or cemetery-level signatures.
-A particular attraction of this method is that it incorporates temporal uncertainty into the size of the resulting values, that is, "events with tight temporal definition contribute more to the total probability over their range than do loosely defined events" [@Johnson2004a, 450].
+A particular attraction of this method is that it incorporates temporal uncertainty into the size of the resulting values, that is, "events with tight temporal definition contribute more to the total probability over their range than do loosely defined events" [@Johnson2004a: 450].
 
@@
 As seen in Figure \@ref(fig:aoristallcems), the artificial peaks and troughs of the simple averaging method are avoided, and different patterns become visible.
-Though these methods present their own limitations [@Crema2010, 1123-24], they are considered a more robust and productive approach to the temporal uncertainty of ceramic dating and, as such, are used for much of the present analysis.
+Though these methods present their own limitations [@Crema2010: 1123-24], they are considered a more robust and productive approach to the temporal uncertainty of ceramic dating.
 
@@
 
-An obvious place to begin is with the chronology of the burials themselves.
-The rate and quantity of interments being made through time provides the background against which to consider all other developments.
+An obvious place to begin is with the chronology of the burials themselves. The rate and quantity of interments through time provides the background against which to consider all other developments.
 
@@
 
-Figure \@ref(fig:aoristbycem) presents the aoristic sums of pithoi (as a proxy for burials) and of all vessels, respectively.
-Despite differences in scale and timing, it is immediately apparent that Fortetsa and the KNC underwent similar chronological developments with regard to burials and, to a slightly lesser degree, grave goods.
+And, despite differences in scale and timing, it is immediately apparent from Figure \@ref(fig:aoristbycem) that Fortetsa and the KNC underwent similar chronological developments with regard to burials and, to a slightly lesser degree, grave goods.
 In both cemeteries, the rate of cremation burial began rising around the end of the 9th century, plateaued slightly in the early 8th, and then rose again to a peak in the late 8th and early 7th.
@@
 The KNC began with a large number of new tombs built during the, admittedly prolongued, SM period, followed by a protracted decline in rates of consruction, offset by small resurgences in the late 9th and 8th centuries.
-It should be noted, too, that many of the later "tombs" are not chamber tombs, but a mix of pit, larnax, and pithos burials.
-The Fortetsa, meanwhile, saw next to no new chamber tomb construction from the 8th century onwards.
+It should be noted, too, that many of the later 'tombs' are not chamber tombs, but a mix of pit, larnax, and pithos burials.
+The Fortetsa, meanwhile, saw almost no new tomb construction from the 8th century onwards.
 This abatement has been noted before (Brock 1957: 4; Cavanagh 1996: 652–53) and somewhat implicitly read as evidence of restructuring of the funerary rite, if not a decline in the fortunes of the wider community.
-But this image of a waigning investment in chamber tombs sits uneasily with the observable growth in mortuary activity.
+But this image of a waining investment in chamber tombs sits uneasily with the observable growth in mortuary activity.
 
@@
 At Fortetsa, this number remained relatively stable from the MPG to LG periods before declining (and bear in mind, there were only ever around 20 tombs belonging to this cemetery), while at the KNC, the number of tombs in use rose to a preak in the late 8th century before declining.
-But this is not the whole story.
 
@@
 
-By dividing the tombs in use into the centuries during which they were built (Figure \@ref(fig:tombsinusebybuilddate)), we see a wave-like pattern with each cohort steadily going out of use as new ones were constructed.
+But the net increase in active tombs in the 8th century belies a more complex pattern (Figure \@ref(fig:tombsinusebybuilddate)).
+By dividing the tombs in use into the centuries during which they were built, we see a wave-like pattern with each cohort steadily going out of use as new ones were constructed.
 This would seem to accord with Cavanagh's (1996: 659) suggestion that "[t]he length of time the tombs last seems [...] to reflect the duration of the social group which had access to the sepulchre", and that, if such access was lineally inehrited, frequent fission would result in limited trans-generational use.
-But note also that the total increase in active tombs during the 8th century belies the diminishing number of continuing 9th and 10th century tombs.
-We might expect, then, that it was the 8th century tombs coming into use that spurred the dramatic rise in burial rates previoulsy noted.
+We might expect, then, that it was the 8th century tombs coming into use that spurred the dramatic rise in burial rates.
 
@@
 And while these mostly remained around three per tomb per century, in the 8th and 7th centuries, tombs founded in the 9th century saw a rapid rise in the rate of burial, peaking at over 11 per tomb per century.
-Though these estimates are no doubt very rough, the marked nature of this divergence suggests it is not simply an artefact of imperfect data.
+Though no doubt rough estimates, the marked nature of this divergence suggests it is not simply an artefact of imperfect data.
 
 But Table 1 also demonstrates the markedly low rates of burial which predominated at both cemeteries; two to four burials per tomb per century amounts to a very selective burial rite.
-It seems unlikely, then, that the rising rates of burial noted above are, for instance, the direct result of population growth; the increase is too rapid, and appears concentrated only in certain (older) tombs, not to mention that, based on KULP’s survey data, the major population increase at EIA Knossos probably occurred back in the Protogeometric period (Whitelaw, per comm.).
+It seems unlikely, then, that the rising rates of burial are, for instance, the direct result of population growth; the increase is too rapid, and concentrated only in older tombs, while, based on KULP’s survey data, the major population increase at EIA Knossos probably occurred back in the Protogeometric period (Whitelaw, pers.
+comm.).
 
@@
 We are yet to draw out which older tombs precisely are responsible for the trend so far discussed.
-We might assume the tombs which became the focus of such intense later burial activity would be those which yielded the most burials and grave goods when excavated and, happily, this supposition proves true.
+We might assume those which became the focus of such intense later burial activity would end up yielding the most burials and grave goods when excavated and, happily, this supposition proves true.
 
@@
 
-```{r groups_a_and_b, fig.dim=c(7,5),dpi=600, fig.align='center',fig.cap="goes around here"}
+```{r groupsab, fig.dim=c(7,5),dpi=600, fig.align='center',fig.cap="goes around here"}
 
@@
 
-Interestingly, Tombs X, G, Q, 219 and 283 ('Group B'), however, evince a different pattern, with their respective peaks coming around a century earlier, which, as seen in Figure \@ref(fig:aoristbycem), was another period of increased burial activity.
-So despite receiving the greatest quantities of burials and grave goods, these tombs nonetheless seem to have achieved that distinction via quite different histories of use, a divergence made clear in Figure \@ref(fig:groups_a_and_b).
-It is notable that tombs assinged to each group are seen in both cemeteries, suggesting from the outset that the differences between these may speak to behaviours and social contexts operating across a significant portion of the Knossian community.
+Interestingly, Tombs X, G, Q, 219 and 283 ('Group B') evince a different pattern, with their respective peaks coming around a century earlier, another period of increased burial activity.
+So despite receiving the greatest quantities of burials and grave goods, these tombs nonetheless seem to have achieved that distinction via quite different histories, a divergence made clear in Figure \@ref(fig:groupsab).
+It is notable that tombs assinged to each group are seen in both cemeteries, suggesting from the outset that the differences between these may speak to social forces operating across the Knossian community.
 
@@
 
-Having addressed some broad diachronic trends, let us now attempt some more synchronic comparisons,   beginning with the physical dimensions of the tombs as built - a potentially rich context for the negotiation of social power, status and group membership.
-Perhaps the most bountiful tombs were also physically the largest, assertions of status and authority embodied in the man-power invested in their construction.
-
-```{r dimensionsboxplots, fig.dim=c(6,5),dpi=600, fig.align='center',fig.cap="goes around here"}
+Having addressed some broad diachronic trends, let us now attempt some more synchronic comparisons, beginning with the physical dimensions of the tombs as built - a potentially rich context for the negotiation of social power and group membership.
+Perhaps the most bountiful tombs were also physically the largest, assertions of status embodied by the man-power invested in their construction.
+
+```{r dimensionsboxplots, fig.dim=c(5,4),dpi=600, fig.align='center',fig.cap="goes around here"}
 
 tombs$century_built<-factor(tombs$century_built,levels=c("11th","10th","9th","8th","7th"))
-dromwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_width,fill=century_built))+geom_boxplot()+
+dromwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_width,fill=century_built))+geom_boxplot(lwd=0.2,outlier.size=0.8,)+
   theme_bw()+theme(legend.position="none",
@@
                     scale_y_continuous(expand=c(0,0.1))
-dromlength<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_length,fill=century_built))+geom_boxplot()+
+dromlength<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_length,fill=century_built))+geom_boxplot(lwd=0.2,outlier.size=0.8)+
   theme_bw()+theme(legend.position="none",
@@
                     scale_y_continuous(expand=c(0,0.5))
-stomwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,stomion_width,fill=century_built))+geom_boxplot()+
+stomwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,stomion_width,fill=century_built))+geom_boxplot(lwd=0.2,outlier.size=0.8)+
   theme_bw()+theme(legend.position="none",
@@
                     scale_y_continuous(expand=c(0,0.1))
-chamarea<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,chamber_area,fill=century_built))+geom_boxplot()+
+chamarea<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,chamber_area,fill=century_built))+geom_boxplot(lwd=0.2,outlier.size=0.8)+
   theme_bw()+theme(legend.position="none",
@@
 The 9th-century tombs, though, which include all but two of the 12 discussed above, do not particularly stand out.
-It does seem that chambers were, on average, larger in this period, as well as exhibiting greater variety than at any other time.
-Then again, even the largest chambers are not exactly cavernous, and the extra labour expended on construction would have been limited.
-The *dromos* is a plausible area for conspicuous demonstrations of scale but, with their frequently poor preservation, we cannot be confident in assessing this.
-It appears at any rate that they were gradually declining in size, albeit with some notable outliers.
+It does seem their chambers were, on average, larger, as well as exhibiting greater variation than at any other time.
+But then again, even the largest chambers are not exactly cavernous, and the extra labour expended would have been limited.
+The *dromos* is a plausible area for conspicuous demonstrations of scale but, with their frequently poor preservation, little confidence is possible.
+It appears at any rate that they were gradually shrinking, albeit with some notable outliers.
 
@@
   geom_label_repel(data=subset(tombs,group!="C"),aes(fill=group),color="white",show.legend = FALSE,family="Garamond")+
-            labs(x="Dromos Length (m)", y="Estimated Total Burials")+
+            labs(x="Dromos Length (m)", y="Chamber Area (sq m)")+
                scale_color_brewer(palette = "Set2",name="Group")+
@@
 
-What we can say is that no clear relationship existed between the physical size of a tomb and its eventual buried population (see Figure \@ref(fig:tombdimensions)).
-Many of the very largest tombs are decidedly middling when it comes to burial numbers, and vice versa.
+What we can say is that no clear relationship existed between the dimensions of a tomb and its eventual buried population (see Figure \@ref(fig:tombdimensions)).
+Many of the very largest tombs are middling when it comes to burial numbers, and vice versa.
 So whatever the scale of the tombs connoted when built, this proved irrelevant to their later reuse.
 Several had niches cut into their *dromoi*, or the *stomion* extended, to fit in later burial urns (including T.
-292, 283 and 107; Cavanagh 1996: 658) and, sometimes, urns from within the chamber were relocated to these to make space for new interments (including T.
+292, 283 and 107; Cavanagh 1996: 658) and, sometimes, urns from within the chamber were relocated to make space for new interments (including T.
 75 and TFT; ibid.: 658; Brock 1957: 3–4).
 The impression is of competition for inclusion into the sepulchral group, further confirming that the tombs’ original builders little anticipated their eventual popularity.
-Whatever led to Group A's curious later use was, it was not presaged by exceptional scale or physical grandeur.
 
@@
 
-Dimensions aside, we can still investigate the contemporary use of these tombs through comparisons between the scale and composition of their respective assemblages through time.
+Dimensions aside, we can investigate the contemporary use of these tombs through comparisons between the scale and composition of their respective assemblages through time.
 
@@
 
-And in Figure \@ref(fig:tombscompthroughtime), we see that the tombs of Group A, while not paultry even in their early years, were by no means exceptional during the 9th century.
-They clearly kept pace with the other popular tombs of the time, but their major divergence, their near monopoly, only emerged in the subsequent two hundred years.
-Group B, meanwhile, experienced similar, if not greater, use in the 9th and early 8th centuries, and then declined.
+And in Figure \@ref(fig:tombscompthroughtime), we see that the tombs of Group A, while not paultry in the 9th century, were by no means exceptional.
+They clearly kept pace with the other popular tombs of the time, but their major divergence only occurred in the subsequent two hundred years.
+Group B, meanwhile, experienced similar, if not greater, use in the 9th century, and then gradually declined.
 That is, these tombs seem to have followed a more conventional trajectory viz.
-\@ref(fig:tombsinusebybuilddate).
+Figure \@ref(fig:tombsinusebybuilddate).
 What might account for this disparity?
-And if the patterns implied here in \@ref(fig:tombscompthroughtime) - relative diversity, giving way to effective bipolarity in the funerary rite - and their emobdiment in Groups A and B respectively are genuine, how might we further investigate their manifestation across the excavated assemblage?
+And if the patterns implied in Figure \@ref(fig:tombscompthroughtime) - relative diversity, giving way to effective bipolarity in depositions - are genuine, how might we further investigate their manifestations across the excavated assemblage?
 
@@
 
-ggplot(aorist_melt,aes(variable,value,fill=period))+geom_boxplot(aes(group = cut_width(variable, 10)),outlier.size=0.8)+theme_bw()+
+ggplot(aorist_melt,aes(variable,value,fill=period))+geom_boxplot(aes(group = cut_width(variable, 10)),outlier.size=0.8,lwd=0.2)+theme_bw()+
       theme(text=element_text(family="Garamond", size=12))+
@@
 
-Figure \@ref(fig:bigbox) presents a boxplot of the total aoristic sums across both cemeteries, colour-coded by ceramic phase.
-Looking at the PGB and EG, we see that they exhibit the largest interquartile ranges (IQRs) of any period, and each possess only a single outlier (beyond 1.5 times the IQR).
-The Orientalising period, on the other hand, exhibits much reduced IQRs, and a proliferation of outliers (that is, Group A) far and above the range of the other tombs.
-In other words, whilst both the PGB-EG and EO-LO periods witnessed an expanding volume and range of burial activity, in the earlier period this was distributed far more evenly across the total burying population.
+Figure \@ref(fig:bigbox) is a boxplot of the total aoristic sums across both cemeteries, colour-coded by ceramic phase.
+Looking at the PGB and EG, we see that they have the largest interquartile ranges (IQRs) of any period, and only a single outlier (beyond 1.5 times the IQR) each.
+The Orientalising period, on the other hand, exhibits much reduced IQRs, and a proliferation of outliers (namely, Group A) far and above the range of the other tombs.
+In other words, whilst both the PGB-EG and EO-LO periods witnessed an expanding volume of burial activity, in the earlier period this was distributed far more evenly across all tombs.
 One speaks of increasing plurality, the other of divergence.
@@
 
-rects <- data.frame(xstart = c(1100,970,920,875,840,810,790,745,700,670,630), 
+rects <- data.frame(xstart = c(1100,970,920,875,840,810,790,745,710,670,630), 
                     xend = c(970,920,875,840,810,790,745,710,670,630,600), 
@@
 
-This characterisation is supported further by considering the Gini coefficient of the distribution of pottery between the tombs (see Figure \@ref(fig:gini)).
+This characterisation is supported further by considering the Gini coefficient for the distribution of pottery between the tombs (see Figure \@ref(fig:gini)).
 Using aoristic sums as a probablistically-weighted estimate of the volume of material deposited in each tomb in each decade, the Gini coefficient can give an (albeit rough) estimate of how unequally distributed that material was.
 And, though values were always relatively high (i.e.
-unequal) within the cemeteries, the major spike in the Gini coefficient seen in the Orientalising period does contrast markedly with the modest rise and plateau of the PGB-EG.
-
-This is not to suggest that the late 9th to early 8th century was a time of understated, egalitarian burial practices.
+unequal) within the cemeteries, the major spike in the Gini coefficient seen in the Orientalising period does contrast markedly with the plateau of the PGB-EG.
+
+This is not to suggest that the late 9th to early 8th century#ies was a time of egalitarian burial practices.
 The PGB and EG are consistently interpreted as periods of artistic development, increased overseas trade, and renewed vigour in the construction of tombs and the visibility of the burial rite [@Brock1957: 143-44; @Coldstream1994: 110-12].
@@
 
+imports1<-nrow(subset(pots,pots$group=="B" & pots$import =="Y"))/nrow(subset(pots,pots$group=="B"))*100
+imports2<-nrow(subset(pots,pots$group=="C" & pots$import =="Y"))/nrow(subset(pots,pots$group=="C"))*100
+imports3<-nrow(subset(pots,pots$group=="A" & pots$import =="Y"))/nrow(subset(pots,pots$group=="A"))*100
 groupA7th<-subset(pots,group=="A" & century=="7th")
 groupA7thimports<-(sum(groupA7th$import=="Y")/nrow(groupA7th))*100
-```
-
-The rise in imported items is clearly seen in the tomb assemblages, with a succession of increased Greek and Near Eastern imports respectively in the late 9th and early 8th centuries (Figure \@ref(fig:imports)).
+
+```
+
+The rise in imported items is readily observable, with a succession of Greek and Near Eastern imports in the late 9th and early 8th centuries (Figure \@ref(fig:imports)).
 Likewise, the decline in Near Eastern imports, and concomitant rise in locally-made imitations thereof, is visible at the end of the 8th century.
-These trends being a very coarse proxy for its engagement with the wider Mediterranean, the late 9th and early 8th centuries would seem to encompass the most outward-looking phase of the EIA Knossian community [see @Whitley2004: 434].
-This much has been relatively well understood for some time, at least at the KNC, but we can further examine how these imports were distributed among the various groups of tombs.
-
-In the Group B tombs, with their more limited flourescence centred on the PGB-EG periods, imported items constitute `r nrow(subset(pots,pots$group=="B" & pots$import =="Y"))/nrow(subset(pots,pots$group=="B"))*100`% of the total excavated assemblage, compared to `r nrow(subset(pots,pots$group=="C" & pots$import =="Y"))/nrow(subset(pots,pots$group=="C"))*100`% in Group C and only `r nrow(subset(pots,pots$group=="A" & pots$import =="Y"))/nrow(subset(pots,pots$group=="A"))*100`% in Group A.
-Indeed, among 7th century vessels, when interments were higher than ever among Group A, the figure is a mere`r groupA7thimports`%.
-Thus, impression grows ever stronger, that the late proliferation of burials in the Group A tombs was altogether different to the comparative diversity of interments a century earlier.
+These trends being a very coarse proxy for its engagement with the wider Mediterranean, the late 9th and early 8th centuries would seem to encompass the most outward- (certainly,eastward-) looking phase of the EIA Knossian community [see @Whitley2004: 434].
+
+In the Group B tombs, with their more limited flourescence centred on the PGB-EG periods, imported items constitute `r round(imports1, digits=2)`% of the total excavated assemblage, compared to `r round(imports2, digits=2)`% in Group C and `r round(imports3, digits=2)`% in Group A.
+Indeed, among 7th century vessels, when interments were higher than ever among Group A, the figure is a mere `r round(groupA7thimports, digits=2)`%.
+The impression grows stronger that the context for the late proliferation of burials in the Group A tombs was altogether different from that underpinning the comparative, competitive diversity of interments a century earlier.
 
@@
                  labs(title="KNC",x="", y="Proportion of Total Assemblage")+
-                  scale_fill_brewer(palette = "Set2")+ 
+                  scale_fill_brewer(palette = "Set2",labels=c(" Amphora  "," Jug  "," Oil vessel  "," Pithos  "," Drinking vessel  "," Lid  "," Other  "))+ 
                     scale_y_continuous(expand=c(0,0.0))
@@
                  labs(title="Fortetsa",x="Century BC", y="")+
-                  scale_fill_brewer(palette = "Set2")+ 
+                  scale_fill_brewer(palette = "Set2",labels=c(" Amphora  "," Jug  "," Oil vessel  "," Pithos  "," Drinking vessel  "," Lid  "," Other  "))+ 
                     scale_y_continuous(expand=c(0,0.0))
@@
                  labs(title="Group A",x="", y="")+
-                  scale_fill_brewer(palette = "Set2")+ 
+                  scale_fill_brewer(palette = "Set2",labels=c(" Amphora  "," Jug  "," Oil vessel  "," Pithos  "," Drinking vessel  "," Lid  "," Other  "))+ 
                     scale_y_continuous(expand=c(0,0.0))
@@
 Across both cemeteries, there appears to have been the steady crystallisation of a basic 'burial kit', comprising a pithos and lid, an oil vessel or two, a jug, and a drinking vessel.
-The proportion of grave goods comprising 'other' vessels - pyxides, askoi, trays, stands, dinoi, kernoi, kraters and more - declines to become only a minor part of the standardised repertoire of the 7th century.
-The tombs of Group A did not break with this mould, and their assemblages seem largely as conventional as any other during the final century of use in the cemeteries.
+The proportion of grave goods comprising 'other' vessels - pyxides, askoi, trays, stands, dinoi, kernoi, kraters and more - declines to become a near insignificant part of the standardised repertoire of the 7th century.
+The tombs of Group A did not break this mould, and their assemblages seem largely as conventional as any other by this time.
 
 The main focus of this article has been on ceramics and not other artefacts of metal, stone, glass, and organic materials, due to their less well-developed chronologies, but even the raw distribution of such artefacts may make more sense in the light of the foregoing discussion.
-There is, perhaps unsurprisingly, a relatively strong (Spearman's ρ=`r spearmanobjectspots`) correlation between the number of ceramic and non-ceramic finds across the tombs, and, certainly, many of the most seemingly precious objects were recovered from among the larger tombs.
-But the near monopoly those tombs would appear to have over such objects, perhaps tempting one initially to speak of 'elite' or 'high-status' tombs, evaporates when the number of those objects is divided by the number of burials represented.
-This fact has seldom, if ever, been appreciated in prior discussions of the cemeteries.
+There is a relatively strong (Spearman's ρ=`r round(spearmanobjectspots,digits=2)`) correlation between the number of ceramic and non-ceramic finds across the tombs, and, certainly, many of the most seemingly precious objects were recovered from among the tombs with the most burials.
 
@@
 
-Figure \@ref(fig:hists) thus supports the view which has emerged from the foregoing analysis, that the late surge in burial activity at both Fortetsa and the KNC was not a phenomenon linked in any straightforward way to simple principles of 'elite' or 'wealthy' burial.
-To summarise, towards the end of 8th century BC, a select group of tombs, all of which had been constructed back in the 9th or 10th, became the foci of a major increase in burials, something which did not occur among the tombs built around that time.
-We have seen how little about the physical dimensions or early use of these tombs presaged their later popularity, and that, while a similar, if smaller, increase in burial activity occurred around a century earlier, in the PGB-EG periods, this contrasted in the comparative diversity of its associated assemblages, and their distribution across the tombs.
-The later burials were marked by great uniformity in the associated grave goods - set against a backdrop of waining tomb construction and declining imports - and this apparently limited and consistent repertoire may explain why many of the most-used tombs prove exceptional only in their absolute quantities of artefacts, and not the accoutrements of each individual burial.
+But this association, perhaps tempting one initially to speak of 'elite' or 'high-status' tombs, evaporates when the number of objects is divided by the number of burials represented.
+This fact has seldom, if ever, been appreciated in prior discussions of the cemeteries (Figure \@ref(fig:hists)).
+Whitley (2004: 437) has pointed out that the proportions of different decorative styles of Orientalising pithoi in Tomb P are replicated across Fortetsa as a whole, but it seems this situation may apply across most artefact classes.
+If so, the late surge in burial activity at both Fortetsa and the KNC was not linked in any straightforward way to simple principles of 'elite' or 'wealthy' burial; many of the most-used tombs prove exceptional only in their absolute quantities of artefacts, and not the accoutrements of each individual burial.
+This fact has seldom been appreciated in prior discussions of the cemeteries.
 
 # Discussion
+
+To summarise, towards the end of 8th century BC, a select group of tombs, all built back in the 9th or 10th, became the foci of a major increase in burials which didn't register among the tombs built during that time.
+Little about the physical dimensions or early use of these tombs presaged their later popularity, while a similar, if smaller, increase in burial activity around a century earlier, in the PGB-EG periods, contrasted in the comparative diversity and broader distribution of its associated assemblages.
 
@@
 The traditional view is that such groups were immediate families (Brock 1957, 41; @Boardman1967: 63; @Coldstream1996: 291).
-The relatively short lifespan of many tombs (often only two or three generations) has been taken as evidence for direct, lineal inheritance, with the tomb going out of use as families moved, individuals founded new tombs, or the line came to an end.
-The idea of lineal groups is supported by the, admittedly circumstantial, evidence of men, women and children being found within single tombs, and one (contentious) instance of an apparently hereditary mandibular deformity in three non-contemporary skulls in the same tomb [@Musgrave1996].
+The relatively short lifespan of many tombs has been taken as evidence for direct, lineal inheritance, with the tomb going out of use as families moved, individuals founded new tombs, or the line fissioned or ended [@Cavanagh1996: 659].
+The idea of family groups is supported by the, admittedly circumstantial, evidence of men, women and children being found within single tombs, and one instance of a (putatively) hereditary mandibular deformity in three skulls in the same tomb [@Musgrave1996].
 
@@
 He argues that the rise in interments during the 8th century is too great to be explained by population growth, as would be necessary for the possibility of direct lineal inheritance to be preserved.
-Certainly, it seems illogical that population increases would lead to the consolidation of burials within only extant tombs.
-The trend instead must reflect reconfigurations of the burying groups themselves, which Kotsonas feels may have (also or instead) comprised age- and sex-grades.
+Rather, the trend must reflect reconfigurations of the burying groups themselves, which Kotsonas feels may have (also or instead) comprised age- and sex-grades.
+We should take this possibility seriously, for one need only look at Tomb P in the 7th century to see that we cannot possibly be dealing with just immediate families.
 
 Yet it is hard to be conclusive, as our current understanding of the social organisation of the early Cretan *poleis* remains patchy.
-Perlman and Gagarin's [-@Gagarin2016] recent study of the Cretan law codes reminds us, importantly, if lamentably, that there is an often underappreciated degree of diversity apparent from the Archaic inscriptions in the offices, institutions, and social groups of the early city-states.
-In the traditional view, political authority was concentrated in the hands of the *kosmoi*, aristocratic magistrates selected from the leading, hereditary *startoi*, or ‘clans’, networks of extended lineages controlling the ancestral estates on which worked attendent serfs and/or slaves (Willetts 1982, 240).
-But despite elements of this account probably being accurate, others remain equivocal, such as the nature of the *pyla*, or 'tribe', and its relation to the *startos* (which may itself be a military grouping), the distinction between *astos* and *poliatas*, both probably formulations of the rights-bearing civic actor, and the ancestry and spread of the *agelai*, seemingly an adolescent age-grade only attested at Eltynia [@Gagarin2016: 36; 78-80].
-Indeed, many terms which might illuminate social divisions and groupings within the Archaic Cretan populace are only attested at one or two sites, and even then sometimes appear to differ in their form or function.
-
-My aim, however, is not to draw a simple parallel between the Knossian burial groups and, say, powerful *startoi*, a term which is not even attested outside of Gortyn in this period (ibid:, 1-2).
-The fundamental conclusion of the disucussed contrasts between the late 9th and late 8th centuries, between the Group A and B tombs, is not that the burying groups (whatever their composition) changed, as such, it is that the socio-economic and political contexts did, in ways which prompted reformulations of funerary practice, one of which being the intense, iterative burial activity evinced by the patrons of the Group A tombs.
-Such inferences, naturally, recall the conclusions of Morris [-@Morris1987: 171-211], who argued 30 years ago that the emergence of the *polis* across the Aegean led to a tension between elite preferences for ostentatious display, and a downplaying of social difference in death, as promoted by the ‘middling’ ideology that the *polis* engendered and promulgated.
-In his well-known model of the *agathoi* and *kakoi*, Morris argued that, at various times, the relevant spheres of kinship demonstrated in burial could widen and constrict, with more and less distantly related lineages drawn into association through changing social strategies, until the pre-eminence of kinship relations in the structuring of political power declined, along with its manifestiations in mortuary practice.
-
-At Knossos, some aspects of the noted expansion in the burying population seem explicable within a model like Morris'.
+Perlman and Gagarin's [-@Gagarin2016] recent study of the Cretan law codes reminds us that there is an often underappreciated degree of diversity in the offices, institutions, and social groups of the early city-states.
+In the traditional view, political authority was vested in the *kosmoi*, aristocratic magistrates selected from the leading, hereditary *startoi*, or ‘clans’, networks of extended lineages controlling the ancestral estates on which worked attendent serfs and/or slaves (Willetts 1982, 240).
+But despite elements of this account probably being accurate, others remain equivocal, such as the nature of the *pyla*, or 'tribe', and its relation to the *startos* (which may itself be a military grouping), the distinction between *astos* and *poliatas*, both probably formulations of the rights-bearing civic actor, and the ancestry and spread of the *agelaos*, an adolescent age-grade currently only attested at Eltynia [@Gagarin2016: 36; 78-80].
+Indeed, many terms which might illuminate social divisions and groupings within the Archaic Cretan populace are only attested at one or two sites, and even then sometimes apparently differ in form or function.
+
+My aim, however, is not to draw a simple parallel between the Knossian burial groups and, say, powerful *startoi*, a term not even attested outside of Gortyn in this period (ibid: 1-2).
+The fundamental conclusion of the disucussed contrasts between the late 9th and late 8th centuries, between the Group A and B tombs, is not that the burying groups changed, as such, it is that wider socio-economic and political contexts did, in ways which prompted reformulations of funerary practice, one of which being the intense, iterative burial activity evinced by the patrons of the Group A tombs.
+Such inferences, naturally, recall the conclusions of Morris [-@Morris1987: 171-211], who argued 30 years ago that the emergence of the *polis* across the Aegean led to a tension between elite preferences for ostentatious display, and a downplaying of social difference in death, as promoted by the ‘middling’ ideology of the city-state.
+In his well-known model of the *agathoi* and *kakoi*, Morris claimed that, at various times, the relevant spheres of kinship demonstrated in burial could widen and constrict, with more and less distantly related lineages drawn into association through changing social strategies, until the pre-eminence of kinship relations in the structuring of political power declined, along with its manifestiations in mortuary practice.
+
+At Knossos, some aspects of the noted expansion in the burying population seem explicable within a model like Morris's.
 The basic principle, of elites employing access to burial as a strategic act of association and social signalling, is almost certainly in evidence here.
-That the noted spate of burials in the late 8th to early 7th centuries was concentrated solely in older tombs strongly implies that issues of heritage, lineage, and tradition were at play.
-Indeed, the stratigraphy of the niches dug into the *dromos* of tomb 107 imply that depositions continued to be made even after the main chamber collapsed.
-The symbolic, and not the physical, presence these tombs seem to have asserted appears far more instrumental in their extensive reuse, seen also in the lack of correlation between dimensions and ultimate number of burials.
+That the noted spate of burials in the late 8th and early 7th centuries was concentrated solely in older tombs strongly implies that issues of heritage, lineage, and tradition were at play.
+In Tomb 107, the stratigraphy of the niches cut into the *dromos* suggest depositions continued to be made even after the main chamber fell into disrepair [@Coldstream1996a: 150]; it is the symbolic and associative, rather than physical qualities of these tombs seem instrumental to their extensive reuse.
 
@@
 Yet, at Knossos, the range of grave goods diminished and crystallised during the late 8th and early 7th centuries.
-The PGB period seems a far better candidate for such an expansion, with its greater variety in chamber dimensions, relatively broad distribution of (local and imported) grave goods, not to mention the innovative ceramic decoration and reuse of BA artefacts noted elsewhere.
-In both the late 9th and late 8th centuries, then, certain burying groups found it propitious to broaden access to the tombs they patronised, but changing social contexts saw those expansions manifest in quite different associated assemblages.
-But whether the increasingly homogenous iteration of burials seen later reflected some standard set of accoutrements intentionally shared by the hypothetical *agathoi*, or else nascent, egalitarian strictures over elaborate burial remains difficult to judge.
+The PGB period better embodies such an expansion, with its greater variety in chamber dimensions and relatively broad distribution of (local and imported) grave goods, not to mention the innovative ceramic decoration and reuse of BA artefacts.
+In both the late 9th and late 8th centuries, then, certain burying groups found it propitious to broaden access to the tombs they patronised, but changing social contexts saw those expansions manifested in quite different ways.
 
@@
 At Knossos, a small number of pithos burials in pits do appear in the KNC and Fortetsa volumes (Brock 1957: 98; Coldstream and Catling 1996: 162), and recent re-excavation of the Teke plot identified numerous small pit burials around three chamber tombs [@Evely2012], raising the possibility that others exist, but have gone unrecognised in prior excavations.
-At Azoria, major urban remodelling in the 7th century saw the earlier, agglutinative house forms of neighbouring Kavousi Vronda and Kastro give way to larger, more formally planned dwellings with little evidence of renovation or remodelling, while larger, apparently civic, buildings were constructed to accommodate major public gatherings [@Haggis2011; @Haggis2011b; @Haggis2014].
+At Azoria, major urban remodelling in the 7th century saw the earlier, agglutinative house forms of neighbouring Kavousi Vronda and Kastro give way to larger, more formally planned dwellings with little evidence of remodelling, while larger, apparently civic, buildings were constructed to accommodate major public gatherings [@Haggis2011; @Haggis2011b; @Haggis2014].
 
@@
 It seems not unreasonable to infer an emergent emphasis on the citizen, the state, and ideologies of relative egality, within which communal gatherings, intra-mural ritual, and individual graves replaced ostentatious, kin-centred acts of gathering, dedication, and burial.
-And in the Knossian cemeteries, we have seen a somewhat diverse set of burial practices, including the deposition of Minoan antiques and imported items, give way to a more prolific, yet more restricted, suite of funerary rites, focussed on a small number of long-established tombs, in the context of diminishing collective tomb construction, replacement of imports with locally-made imitations, and an imminent cessation of such depositions.
-
-Yet we need not adopt wholesale a model like Morris's "middling" ideology, nor link the diminishing range of burial goods to Archaic Crete's purported austerity, its (quite literally) spartan lifestyles.
+And in the Knossian cemeteries, we have seen a somewhat diverse set of burial practices give way to a more prolific, yet more restricted, suite of funerary rites, focussed on a small number of long-established tombs, in the context of diminishing collective tomb construction, replacement of imports with locally-made imitations, and an imminent cessation of such depositions.
+
+Whether the increasingly homogenous burials reflected some standard set of accoutrements intentionally shared by the hypothetical *agathoi*, or else nascent, egalitarian strictures over elaborate burial remains difficult to judge.
+Yet we need not adopt wholesale a model like Morris's "middling" ideology, nor link the diminishing range of burial goods to Archaic Crete's purported austerity.
 As a more nuanced picture emerges of the island's relationship with the wider Mediterranean, subtler characterisations are possible.
-It is telling that the peak for imported items in the tombs came in the 9th century; the Orientalising as a time period feels increasingly misapplied to Crete [@Whitley2004: 434], when, at Knossos at least, it witnessed far lower rates of 'oriental' imports, and only limited adoption of the figural motifs which characterise its appearance elsewhere.
-The decline in imports was, however, counterpointed by a rise in imitations of Near Eastern vessels [@Antoniadis2012: 170], further according with our growing awareness of the selective nature of Archaic and Classical Crete's external relations.
-Erickson [@Erickson2010], for instance, has argued that *symposia* were practised in Crete, something often disputed, but that their practitioners merely favoured simpler vessel forms than the figurally-decorated wares popular on the mainland.
-Local, selective adoption of, and responses to, wider social and material trends across the contemporary Aegean is perhaps becoming the hallmark of Archaic Crete, and finds further endorsement in the trends identified in the EIA cemeteries at Knossos.
+It is telling that the peak for imported items came in the 9th century; the Orientalising as a time period feels increasingly misapplied to Crete [@Whitley2004: 434], when, at Knossos at least, it witnessed far lower rates of 'oriental' imports, and only limited adoption of the figural motifs which characterise its appearance elsewhere.
+
+The decline in imports was, however, counterpointed by a rise in imitations of Near Eastern vessels [@Antoniadis2012: 170], according with our growing awareness of the selective nature of Archaic and Classical Crete's external relations.
+Erickson [@Erickson2010], similarly, has argued that *symposia* were practised in Crete, something often disputed, but that their practitioners merely favoured simpler vessel forms than the figurally-decorated wares popular on the mainland.
+Local, selective adoption of, and responses to, wider social and material trends is perhaps becoming the hallmark of Archaic Crete, and finds further endorsement in the EIA cemeteries at Knossos.
 What seems certain is that this was a period of transition wherein, whatever their exact identity, certain actors invested heavily in a final flurry of burials in chamber tombs, with an emphasis on a traditional, established repertoire of items, rites, and venues for one's passage to the afterlife, before changing social forces effected their abandonment.
@@
 It is important that the present study was a comparative investigation of both the Fortetsa and Knossos North cemeteries.
-It is notable how similar the observed trends are at both locations.
-That both underwent comparable developments during the EIA suggests, not only that they catered to similar sub-sections of the Knossian population, but that some of the socio-economic and political changes inferred and discussed were significant and community-wide.
+That both underwent comparable developments during the EIA suggests, not only that they catered to similar sub-sections of the Knossian population, but that some of the socio-economic and political changes inferred were significant and community-wide.
 Further work should seek to extend these comparisons, both into trends unexplored in the present study, and out to the wider funerary landscape of Knossos.
-I hope to do just that but, for the time being, the datasets I am making available alongside this article include the other published EIA tombs in the area, so that others may also explore their possible significance.
+I hope to do just this but, for the time being, the datasets I am making available alongside this article include the other published EIA tombs in the area, so that others may also explore their possible significance.
 
@@
 For one, non-ceramic artefacts were underrepresented due to difficulties with dating, and future work could seek to address this.
-Issues of gender were sadly little considered, though this was largely due to limitations in the data themselves.
-There have also been a number of tombs discovered in the last 20 years that fill in the spaces between the Teke plot and the Medical Faculty Site, and ideally these unpublished tombs should be included in further analyses as they, along with the unconsidered Teke tholos and associated tombs, are probably part of a single contiguous cemetery [@Blackman1999: 113; @Catling1983: 51; @French1991: 68-69].
-Finally, a preliminary study was recently undertaken by the author of select boxes marked as 'unpublished' from the KNC deposits kept at the British School at Athens at Knossos.
-Despite going unmentioned in the publication, this material was not inconsiderable and so, regretably, future work will be necessary to determine whether it threatens the integrity of quantitative analyses such as this.
+Issues of gender were sadly little considered, though this was largely due to limitations in the data.
+There have also been a number of tombs discovered in the last 20 years that fill in the spaces between the Teke plot and the Medical Faculty Site, and ideally these unpublished tombs should be included in further analyses as they, along with the Teke tholos and associated tombs, are probably part of a single contiguous cemetery [@Blackman1999: 113; @Catling1983: 51; @French1991: 68-69].
+Finally, a preliminary study was recently undertaken by the author of select boxes marked as 'unpublished' from the KNC deposits kept at the British School at Athens, at Knossos.
+Despite little mention in the publication, this material was not inconsiderable and so, regretably, future work will be necessary to determine whether it threatens the integrity of these analyses.
 Suffice to say that, at present, it is another pinch of salt with which to take the present findings.
 
-These limitations notwithstanding, the foregoing analysis demonstrates the value of quantitative approaches to the archaeological record of the period.
-Though the construction of databases can be time consuming, once made they offer the opportunity to subdivide, compare, and plot the data in numerous ways quickly and easily, facilitating investigation of the material from multiple angles.
+These limitations notwithstanding, the foregoing analysis demonstrates the value of applying quantitative approaches to the archaeological record of the period.
+Though the construction of databases can be time consuming, once made they offer the opportunity to subdivide, compare, and plot the data in numerous ways, facilitating investigation of the material from multiple angles.
 By making datasets available for use by others, as with the data and code for the present study, one’s conclusions can also be independently tested, and alternative interpretations can be voiced.

### 2020-10-09 README.Rmd
Edited Readme

@@
 
-This repository contains the data and code for my paper:
+This repository contains the data and code for the paper:
 
@@
 
-You can download the compendium as a zip from from this URL: </archive/master.zip>
+You can download the compendium as a zip from from this URL: <https://github.com/DCPollard94/knossoscemeteries/archive/master.zip>
 
-Or you can install this compendium as an R package, domproject, from GitHub with:
+Or you can install this compendium as an R package from GitHub with:
 ```{r gh-installation, eval = FALSE}

### 2020-10-09 R/Samples.R

@@ -362,11 +362,57 @@ ggplot(top_aorist_imports_melt,aes(variable, value))+geom_col()+theme(legend.pos
   scale_x_reverse(expand = c(0, 0),limits=c(1100,610))+
   theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
 
+import_aorist_tomb<-aggregate(imports[,27:76], by=list(tomb=imports$tomb), FUN=sum)
+
+Opithoi<-subset(pithoi,start_date<711)
+
+tombs$pithoiO<-0
+
+inner_join(Opithoi, tombs, by=c("tomb"="tomb")) %>%
+  group_by(shape,tomb) %>% summarise(pithoiO = sum(shape=="Pithos"))
+
+for(i in colnames(aorist_tomb_values[2:51])) {
+  tombs_gini[[paste0(i)]] <- ineq(aorist_tomb_values[[i]],type="Gini")
+}
+
+ %>%
+  group_by(Col1,Col2) %>%
+  add_count
+
+data1 <- data.frame(table(unlist(Opithoi$tomb))) %>%
+  setNames(c("tomb","total"))
+
+data2 <- data.frame(table(unlist(polychrome$tomb))) %>%
+  setNames(c("tomb","polyrchome"))
+
+pithoicounts<-merge(data1,data2)
+
+pithoicounts<-subset(pithoicounts,total!=0)
+View(pithoicounts)
+pithoicounts$percent_poly<-(pithoicounts$polyrchome/pithoicounts$total)*100
+
+pithoicounts<-pithoicounts[with(pithoicounts, order(-pithoicounts$total)), ]
+pithoitomblevels<-as.vector(pithoicounts$tomb)
+pithoicounts$tomb<-factor(pithoicounts$tomb,levels=pithoitomblevels)
+
+ggplot(pithoicounts,aes(tomb,percent_poly))+geom_col()
+
+pithoi_O<-aggregate(tombs[,126:7], by=list(tomb=tombs$tomb), FUN=sum)
+
+tombs$pithoiO<-count_if(tombs$tomb,Opithoi$tomb)
+
+tombs$pithoi7th<-length(which(pithoi$ave_date == "CA"))
+
+names(pithoi)
 ```
 polychrome<-pithoi[pithoi$additional %like% "Polychrome",]
-View(polychrome)
 
[…]

### 2020-10-09 R/Samples2.R

@@ -68,4 +68,286 @@ row.names(aorist_KNC_pithoibyall)<-dates
 setDT(aorist_KNC_pithoibyall, keep.rownames = "decade")
 aorist_KNC_pithoibyall$decade<-as.numeric(aorist_KNC_pithoibyall$decade)
 View(aorist_KNC_pithoibyall)
-ggplot(aorist_KNC_pithoibyall,aes(decade,aorist_KNC_pithoibyall))+geom_col()
+ggplot(data=subset(pots,pots$height>0 & pots$shape=="Oinochoe"),aes(century,height,fill=century))+geom_boxplot()
+
+ggplot(data=subset(pots,type!="Other"),aes(diameter,height, color=century))+geom_point()+facet_wrap(~type,scales="free_y")
+
+objects<-read.csv("../data/raw_data/KnossosObjects.csv",header=TRUE)
+
+objects$group<-ifelse(objects$tomb=="P"|objects$tomb=="292"|objects$tomb=="75"|objects$tomb=="107"|objects$tomb=="285"|objects$tomb=="II"|objects$tomb=="218","A",
+                     ifelse(objects$tomb=="X"|objects$tomb=="G"|objects$tomb=="219"|objects$tomb=="Q"|objects$tomb=="283","B","C"))
+ggplot(objects,aes(category,fill=group))+geom_bar()
+ggplot(subset(objects,category=="Weapon"),aes(tomb,fill=group))+geom_bar()
+names(objects)
+
+ggplot(subset(imports,century!="NA"))+geom_bar(aes(tomb,fill=group))+facet_grid(~century,scales="free_x")
+
+tomb_rank_names<-as.vector(tombs_ranked$tomb)
+pots$tomb<-factor(pots$tomb,levels=tomb_rank_names)
+
+
+ggplot(subset(tombs,est_burials>1 & cooking>0),aes(cooking/est_burials, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+                   text=element_text(family="Garamond", size=12))+
+  labs(x="Total Non-Ceramics/Burials", y="Number of Tombs")+
+  scale_fill_brewer(palette = "Set2",labels=c(" Group A  ", " Other Tombs"))+
+  scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+
+antiques<-subset(pots,antique!="")
+ggplot(antiques,aes(tomb,fill=group))+geom_bar()
+levels(pots$type)
+
[…]

### 2020-10-12 analysis/paper/Family_Fortunes_Article.Rmd
Article edits

--- 2019-06-06 analysis/paper/paper2.Rmd
@@
 title: "Family Fortunes: A Quantitative Analysis of the Early Iron Age Cemeteries at Knossos, Crete"
-subtitle: "Early Iron Age Cemeteries at Knossos"
 author:
-  - Dominic Pollard, Institute of Archaeology, University College London
+  - Dominic Pollard
+  - Institute of Archaeology, University College London, 31-34 Gordon Square, Bloomsbury, London WC1H 0PY, United Kingdom
   - dominic.pollard.16@ucl.ac.uk
@@
 bibliography: references.bib
-csl: "../templates/world-archaeology.csl" # Insert path for the bib-style
+csl: "../templates/journal-of-anthropological-archaeology.csl" # Insert path for the bib-style
 
 abstract: 
-   "The Early Iron Age cemeteries of Knossos, Crete, represent the richest source of archaeological material at the site for the whole period between the end of the Bronze Age and the emergence of the Archaic *polis*.
-However, despite well-published excavation reports on the two largest burial grounds, the Fortetsa and Knossos North cemeteries, and a number of studies dealing with various artefact classes and individual tombs therein, no substantial comparative, quantitative analysis of these two cemeteries has yet been undertaken.
+   "The Early Iron Age cemeteries of Knossos, Crete, represent the richest source of archaeological material at the site for the whole period between the end of the Bronze Age and the emergence of the Archaic *polis*, or city-state.
+However, despite well-published excavation reports on the two largest burial grounds and a number of studies dealing with various artefacts and individual tombs therein, no substantial comparative, quantitative analysis has yet been undertaken.
 This paper provides such an analysis, examining both the synchronic and diachronic variation among the tombs in these two cemeteries.
-Through the application of a range of quantitative techniques, this study isolates a pattern of increased rates of burial among a few select tombs, and goes on to consider the broader social contexts within which these tombs were built and used, as well as discussing the possible nature of the burying groups.
+Through a range of quantitative techniques, this study isolates a pattern of increased rates of burial among a few select tombs, and considers the broader social contexts within which these tombs were built and used, as well as discussing the possible nature of the burying groups.
 This paper demonstrates the potential for quantitative analyses to contribute to nuanced, context-specific theories of social practice, and advocates their wider application to the archaeological record of the Cretan Early Iron Age.
-It also serves as a demonstration of maximal research reproducibility, with the full datasets and code underpinning the analysis made freely available to download and use"
+It also serves as a demonstration of maximal research reproducibility, with the full datasets and code underpinning the analysis made freely available to download and use."
 
@@
   fig.path = "../figures/",
-  dev="png"
+  dev="jpeg"
 )
@@
 library(data.table)
-library(extrafont)
 library(ggplot2)
@@
 library(ggrepel)
-font_import(paths="../paper/font",pattern="GARA")
-loadfonts(device="win")
 
@@
 
-# Introduction
+# 1. Introduction
 Three principal interests converge in this paper: a reorientation of the focus of Early Iron Age Cretan archaeology; the application of computational, and quantitative methods to supplement traditional modes of archaeological inference; and open-access, reproducible research.
@@
 
-# Early Iron Age Knossos
-## Settlement Evidence: 'A meagre filling in a very thick sandwich'
+# 2. Early Iron Age Knossos
+## 2.1 Settlement Evidence: 'A meagre filling in a very thick sandwich'
 
@@
 
-## The Early Iron Age Cemeteries
-
-```{r map, fig.cap="goes around here"}
-knitr::include_graphics("../figures/Knossos Tombs Map.jpg")
+## 2.2 The Early Iron Age Cemeteries
+
+```{r map, out.width="50%", fig.cap="goes around here"}
+knitr::include_graphics("../figures/Knossos Tombs Map2.jpg")
 ```
@@
 
-## The Fortetsa and Knossos North Cemeteries
+## 2.3 The Fortetsa and Knossos North Cemeteries
 
@@
 
-## Chronological Developments in the Cemeteries
+## 2.4 Chronological Developments in the Cemeteries
 
@@
 
-## Heroes Abroad, Immigrants at Home
+## 2.5 Heroes Abroad, Immigrants at Home
 
@@
 
-## Number Crunching: Quantitative Analyses of the Cemeteries
+## 2.6 Number Crunching: Quantitative Analyses of the Cemeteries
 
@@
 
-# Methods
-
-## Databases
+# 3. Methods
+
+## 3.1 Databases
 
@@
 
-## Reproducibility and Data-sharing
+## 3.2 Reproducibility and Data-sharing
 
@@
 
-## Estimating Interments
+## 3.3 Estimating Interments
 
@@
 
-```{r burialestimate, fig.dim = c(7, 4), dpi=600, fig.align='center', fig.cap="goes around here"}
+```{r burialestimate, fig.dim = c(7, 7), dpi=600, fig.align='center', fig.cap="goes around here"}
 
@@
          theme(legend.title=element_blank(),
-          text=element_text(family="Garamond", size=12))+
+          text=element_text(family="serif", size=12),legend.position="bottom")+
               labs(x="Mean Burial Estimate", y="Interment/Vessel Count")+
@@
 
-##Aoristic Approaches to Dating
+## 3.4 Aoristic Approaches to Dating
 
@@
 
-```{r allpithoi,fig.dim = c(7, 4), dpi=600, fig.align='center', fig.cap="goes around here"}
+```{r allpithoi,fig.dim = c(8, 4), dpi=600, fig.align='center', fig.cap="goes around here"}
 
@@
 
-ggplot(pots, aes(ave_date,fill="Pots"))+geom_histogram()+
+non_aorist_fig<-ggplot(pots, aes(ave_date,fill="Pots"))+geom_histogram(fill="gray40")+
   theme_bw()+theme(legend.title=element_blank(),
-      text=element_text(family="Garamond", size=12))+
+      text=element_text(family="serif", size=12))+
           labs(x="Date (Years BC)", y="Vessel Count")+
-              scale_fill_brewer(palette = "Set3")+ 
                 theme(legend.position="none")+
@@
                       theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+#The aoristic values for every tomb are summed by decade, and then this object is converted to a data frame.
+
+aorist_all_tombs<-colSums(tombs[158:207])
+aorist_all_tombs<-data.frame(aorist_all_tombs)
+
+#The column of values is labelled "aoristic_sum", an vector named "dates" is created with values between 1100 and 610 (in that order) at intervals of 10, and then this used to rename the rows of the data frame.
+Finally, these "row names" are converted into a functioning row in their own right within the data frame (this makes them plottable as a variable) and then are converted into numeric variables, allowing the plotting of them along a continuous scale.
+
+colnames(aorist_all_tombs)<-c("aoristic_sum")
+dates<-seq(from = 1100, to = 610, by = -10)
+row.names(aorist_all_tombs)<-dates
+setDT(aorist_all_tombs, keep.rownames = "decade")
+aorist_all_tombs$decade<-as.numeric(aorist_all_tombs$decade)
+
+#The plot is produced, following the same stylistic adjustments of the previous plot.
+
+aorist_fig<-ggplot(aorist_all_tombs,aes(decade,aoristic_sum, fill="Aoristic Sum"))+
+  geom_col(fill="gray40")+theme_bw()+theme(legend.title=element_blank(),
+      text=element_text(family="serif", size=12))+
+          labs(x="Date (Years BC)", y="Aoristic Sum")+
+               theme(legend.position="none")+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,225))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+ggarrange(non_aorist_fig,aorist_fig,ncol=2,common.legend=FALSE)
 
@@
 
-```{r aoristallcems, fig.dim = c(7, 4), dpi=600, fig.align='center', fig.cap="goes around here"}
-
-#The aoristic values for every tomb are summed by decade, and then this object is converted to a data frame.
-
-aorist_all_tombs<-colSums(tombs[158:207])
-aorist_all_tombs<-data.frame(aorist_all_tombs)
-
-#The column of values is labelled "aoristic_sum", an vector named "dates" is created with values between 1100 and 610 (in that order) at intervals of 10, and then this used to rename the rows of the data frame.
-Finally, these "row names" are converted into a functioning row in their own right within the data frame (this makes them plottable as a variable) and then are converted into numeric variables, allowing the plotting of them along a continuous scale.
-
-colnames(aorist_all_tombs)<-c("aoristic_sum")
-dates<-seq(from = 1100, to = 610, by = -10)
-row.names(aorist_all_tombs)<-dates
-setDT(aorist_all_tombs, keep.rownames = "decade")
-aorist_all_tombs$decade<-as.numeric(aorist_all_tombs$decade)
-
-#The plot is produced, following the same stylistic adjustments of the previous plot.
-
-ggplot(aorist_all_tombs,aes(decade,aoristic_sum, fill="Aoristic Sum"))+
-  geom_col()+theme_bw()+theme(legend.title=element_blank(),
-      text=element_text(family="Garamond", size=12))+
-          labs(x="Date (Years BC)", y="Aoristic Sum")+
-              scale_fill_brewer(palette = "Set3")+ 
-                theme(legend.position="none")+
-                  scale_y_continuous(expand=c(0,0),limits=c(0,225))+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-```
-
 As seen in Figure \@ref(fig:aoristallcems), the artificial peaks and troughs of the simple averaging method are avoided, and different patterns become visible.
@@
 
-#Analysis
-
-##Introduction
+# 4. Analysis
+
 The great advantage of the large, quantitative dataset employed here is that it permits investigation of several nested levels of potentially significant patterning.
@@
 
-##Temporal Developments in the Knossian Cemeteries
+## 4.1 Temporal Developments in the Knossian Cemeteries
 
@@
         theme_bw()+theme(legend.title=element_blank(),
-          text=element_text(family="Garamond", size=12))+
+          text=element_text(family="serif", size=12))+
             labs(x="Date (Years BC)", y="Aoristic Sum (Pithoi)")+
-              scale_fill_brewer(palette = "Set2")+ 
+              scale_fill_manual(values = c("gray50","gray65","gray80"))+ 
                   scale_y_continuous(expand=c(0,0),limits=c(0,45))+
@@
         theme_bw()+theme(legend.title=element_blank(),
-          text=element_text(family="Garamond", size=12))+
+          text=element_text(family="serif", size=12))+
             labs(x="Date (Years BC)", y="Aoristic Sum (All Vessels)")+
-              scale_fill_brewer(palette = "Set2")+ 
+              scale_fill_manual(values = c("gray50","gray65","gray80"))+  
                   scale_y_continuous(expand=c(0,0),limits=c(0,225))+
@@
 
-new_tomb_counts_plot<-ggplot(data=subset(tombs,!is.na(start_period)),aes(start_period, fill=cem_condensed))+geom_bar(position="identity")+theme_classic()+
+new_tomb_counts_plot<-ggplot(data=subset(tombs,!is.na(start_period)),aes(start_period, fill=cem_condensed))+geom_bar(position=position_dodge2(preserve="single"))+theme_classic()+
         theme_bw()+theme(legend.title=element_blank(),
-          text=element_text(family="Garamond", size=12))+
+          text=element_text(family="serif", size=12),
+          axis.text.x=element_text(angle=90, vjust=0.5, hjust=1))+
             labs(x="Period", y="Count of New Tombs")+
-              scale_fill_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+ 
+              scale_fill_manual(values =c("gray50","gray80"),labels=c(" KNC  "," Fortetsa"))+ 
                  scale_y_continuous(expand=c(0,0),limits=c(0,30))
 
-new_tomb_aorist_plot<-ggplot(aorist_tomb_periods_melt,aes(period,aoristic_sum, fill=cemetery))+geom_col(position="identity")+
+new_tomb_aorist_plot<-ggplot(aorist_tomb_periods_melt,aes(period,aoristic_sum, fill=cemetery))+geom_col(position=position_dodge2(preserve="single"))+
   theme_bw()+theme(legend.title=element_blank(),
-          text=element_text(family="Garamond", size=12))+
+          text=element_text(family="serif", size=12),
+          axis.text.x=element_text(angle=90, vjust=0.5, hjust=1))+
             labs(x="Period", y="Aoristic Sum of New Tombs")+
-              scale_fill_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+ 
+               scale_fill_manual(values =c("gray50","gray80"),labels=c(" KNC  "," Fortetsa"))+ 
                  scale_y_continuous(expand=c(0,0),limits=c(0,30))
@@
 
-```{r tombsinuse, fig.dim=c(7,4), dpi=600, fig.cap="goes around here"}
+```{r tombsinuse, fig.dim=c(7,7), dpi=600, fig.cap="goes around here"}
 
@@
        theme_bw()+theme(legend.title=element_blank(),
-          text=element_text(family="Garamond", size=12))+
+          text=element_text(family="serif", size=12),legend.position = "bottom")+
             labs(x="Period", y="Tombs in Use")+
-              scale_fill_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+ 
+              scale_fill_manual(values =c("gray50","gray80"),labels=c(" KNC  "," Fortetsa"))+ 
                  scale_y_continuous(expand=c(0,0),limits=c(0,45))
@@
             geom_col(data=tomb_use,aes(period,no_tombs,fill=tomb_century),position="dodge")+
-              theme_bw()+theme(text=element_text(family="Garamond", size=12))+
+              theme_bw()+theme(text=element_text(family="serif", size=12))+
                  labs(x="Period", y="Estimated No. of Tombs in Use")+
-                  scale_fill_brewer(name="Century 
-Built",palette = "Set2")+ 
+                  scale_fill_brewer(name="Century Built",palette = "Set2")+ 
                     scale_y_continuous(expand=c(0,0),limits=c(0,60))
@@
 
-```{r potsbydateandtombdate,  fig.dim=c(7,4), dpi=600, fig.cap="goes around here"}
+```{r potsbydateandtombdate,  fig.dim=c(7,14), dpi=600, fig.cap="goes around here"}
 
@@
   theme_bw()+theme(legend.position="none",
-               text=element_text(family="Garamond", size=12))+
+               text=element_text(family="serif", size=12))+
                  labs(x="Century BC (Date of Vessel)", y="Vessel Count")+
-                  scale_fill_brewer(palette = "Set2")+ 
+                  scale_fill_manual(values = c("gray50"))+ 
                     scale_y_continuous(expand=c(0,0),limits=c(0,2800))
@@
   theme_bw()+theme(legend.position="none",
-               text=element_text(family="Garamond", size=12))+
+               text=element_text(family="serif", size=12))+
                  labs(x="Century BC (Date of Tomb)", y="Vessel Count")+
-                  scale_fill_brewer(palette = "Set2")+ 
+                  scale_fill_manual(values = c("gray50"))+ 
                     scale_y_continuous(expand=c(0,0),limits=c(0,2800))
 
-ggarrange(date_by_vessel,date_by_tomb,ncol=2)
+ggarrange(date_by_vessel,date_by_tomb,ncol=1)
 
@@
 ```
-
-*Table 1 goes around here*
-
-The significance of these findings can be seen in Table 1.
-Using pithoi as a proxy for burials, we can estimate the changing rates of burial through time.
-And while these mostly remained around three per tomb per century, in the 8th and 7th centuries, tombs founded in the 9th century saw a rapid rise in the rate of burial, peaking at over 11 per tomb per century.
-Though no doubt rough estimates, the marked nature of this divergence suggests it is not simply an artefact of imperfect data.
-
-But Table 1 also demonstrates the markedly low rates of burial which predominated at both cemeteries; two to four burials per tomb per century amounts to a very selective burial rite.
-It seems unlikely, then, that the rising rates of burial are, for instance, the direct result of population growth; the increase is too rapid, and concentrated only in older tombs, while, based on KULP’s survey data, the major population increase at EIA Knossos probably occurred back in the Protogeometric period (Whitelaw, pers.
-comm.).
-
-##Isolating the Phenomenon
-
-We are yet to draw out which older tombs precisely are responsible for the trend so far discussed.
-We might assume those which became the focus of such intense later burial activity would end up yielding the most burials and grave goods when excavated and, happily, this supposition proves true.
-
-```{r toptombs, fig.dim = c(7, 5), dpi=600, fig.align='center',fig.cap="goes around here"}
-
-tombs_ranked<-tombs[with(tombs, order(-tombs$total_finds)), ]
-tombs_top_12<-head(tombs_ranked,12)
-top12<-as.vector(tombs_top_12$tomb)
-pots_top_12<-pots[is.element(pots$tomb, top12),]
-pots_top_12$tomb<-factor(pots_top_12$tomb,levels=top12)
-
-cols<-c(1,158:207)
-aorist_all_tomb_names<-tombs_top_12[,cols]
-aorist_all_tomb_names<-data.frame(aorist_all_tomb_names)
-colnames(aorist_all_tomb_names)<-c("tomb",dates)
-
-aorist_all_tomb_names$cemetery<-tombs_top_12$cem_condensed
-
-aorist_all_tomb_melt<-melt(aorist_all_tomb_names)
-aorist_all_tomb_melt$variable<-as.numeric(as.character(aorist_all_tomb_melt$variable))
-
-top_12<-tombs_top_12$tomb
-aorist_all_tomb_melt$tomb<-factor(aorist_all_tomb_melt$tomb,levels=top_12)
-
-ggplot(aorist_all_tomb_melt,aes(variable, value,fill=cemetery))+geom_col()+theme(legend.position = "none")+facet_wrap(~tomb,scales="free_y")+
-  geom_col()+theme_bw()+theme(legend.title=element_blank(),legend.position="bottom",
-      text=element_text(family="Garamond", size=12))+
-          labs(x="Date (Years BC)", y="Aoristic Sums")+
-              scale_fill_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+ 
-                  coord_cartesian(xlim=c(1100,610))+scale_y_continuous()+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1100,610))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-tombs$group<-ifelse(tombs$tomb=="P"|tombs$tomb=="292"|tombs$tomb=="75"|tombs$tomb=="107"|tombs$tomb=="285"|tombs$tomb=="II"|tombs$tomb=="218","A","B")
-total_pottery_by_groups<-aggregate(tombs$total_pottery, by=list(tombs$group), FUN=sum)
-calc1<-(total_pottery_by_groups[1,2]/sum(total_pottery_by_groups$x))*100
-
-```
-
-The 12 tombs with the most associated vessels are all chamber tombs, 10 dating to the 9th century and two to the 10th, and seven of them did exhibit late, intense episodes of deposition (Figure \@ref(fig:toptombs)).
-In fact, Tombs P, 292, 75, 107, 285, II and 218 (henceforth 'Group A') between them account for `r round(calc1,digits=2)`% of the entire ceramic assemblage from both cemeteries.
-All but Tomb 285 date to the 9th century, and together they underpinned the discussed late spate of burials.
-
-```{r groupsab, fig.dim=c(7,5),dpi=600, fig.align='center',fig.cap="goes around here"}
-
-tombs$group<-ifelse(tombs$tomb=="P"|tombs$tomb=="292"|tombs$tomb=="75"|tombs$tomb=="107"|tombs$tomb=="285"|tombs$tomb=="II"|tombs$tomb=="218","A",
-                   ifelse(tombs$tomb=="X"|tombs$tomb=="G"|tombs$tomb=="219"|tombs$tomb=="Q"|tombs$tomb=="283","B","C"))
-
-GroupA<-subset(tombs,group=="A")
-
-aorist_GroupA <- colSums(GroupA[158:207])
-aorist_GroupA<-data.frame(aorist_GroupA)
-
-colnames(aorist_GroupA)<-c("aoristic_sum")
-row.names(aorist_GroupA)<-dates
-aorist_GroupA<-data.frame(aorist_GroupA)
-setDT(aorist_GroupA, keep.rownames = "decade")
-aorist_GroupA$decade<-as.numeric(aorist_GroupA$decade)
-
-aorist_GroupA$group<-"A"
-
-GroupB<-subset(tombs,tombs$group=="B")
-
-aorist_GroupB <-colSums(GroupB[158:207])
-aorist_GroupB<-data.frame(aorist_GroupB)
-
-colnames(aorist_GroupB)<-c("aoristic_sum")
-row.names(aorist_GroupB)<-dates
-aorist_GroupB<-data.frame(aorist_GroupB)
-setDT(aorist_GroupB, keep.rownames = "decade")
-aorist_GroupB$decade<-as.numeric(aorist_GroupB$decade)
-
-aorist_GroupB$group<-"B"
-
-aorist_groups<-rbind(aorist_GroupA,aorist_GroupB)
-
-ggplot(aorist_groups,aes(decade,aoristic_sum,fill=group))+geom_col(position="dodge")+
-  theme_bw()+theme_bw()+theme(text=element_text(family="Garamond", size=12))+
-                labs(x="Date (Years BC)", y="Aoristic Sum")+
-                  scale_fill_brewer(name="Group",palette = "Set2")+
-                   scale_y_continuous(expand=c(0,0),limits=c(0,120))+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-```
-
-Interestingly, Tombs X, G, Q, 219 and 283 ('Group B') evince a different pattern, with their respective peaks coming around a century earlier, another period of increased burial activity.
-So despite receiving the greatest quantities of burials and grave goods, these tombs nonetheless seem to have achieved that distinction via quite different histories, a divergence made clear in Figure \@ref(fig:groupsab).
-It is notable that tombs assinged to each group are seen in both cemeteries, suggesting from the outset that the differences between these may speak to social forces operating across the Knossian community.
-
-##Not All Tombs are Created Equal
-
-Having addressed some broad diachronic trends, let us now attempt some more synchronic comparisons, beginning with the physical dimensions of the tombs as built - a potentially rich context for the negotiation of social power and group membership.
-Perhaps the most bountiful tombs were also physically the largest, assertions of status embodied by the man-power invested in their construction.
-
-```{r dimensionsboxplots, fig.dim=c(5,4),dpi=600, fig.align='center',fig.cap="goes around here"}
-
-tombs$century_built<-factor(tombs$century_built,levels=c("11th","10th","9th","8th","7th"))
-dromwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_width,fill=century_built))+geom_boxplot(lwd=0.2,outlier.size=0.8,)+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Century of Tomb Construction", y="Dromos Width (m)")+
-                  scale_fill_brewer(palette = "Set2")+ 
-                    scale_y_continuous(expand=c(0,0.1))
-dromlength<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_length,fill=century_built))+geom_boxplot(lwd=0.2,outlier.size=0.8)+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Century of Tomb Construction", y="Dromos Length (m)")+
-                  scale_fill_brewer(palette = "Set2")+ 
-                    scale_y_continuous(expand=c(0,0.5))
-stomwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,stomion_width,fill=century_built))+geom_boxplot(lwd=0.2,outlier.size=0.8)+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Century of Tomb Construction", y="Stomion Width (m)")+
-                  scale_fill_brewer(palette = "Set2")+ 
-                    scale_y_continuous(expand=c(0,0.1))
-chamarea<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,chamber_area,fill=century_built))+geom_boxplot(lwd=0.2,outlier.size=0.8)+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Century of Tomb Construction", y="Chamber Area (sq m)")+
-                  scale_fill_brewer(palette = "Set2")+ 
-                    scale_y_continuous(expand=c(0,0.5))
-
-ggarrange(dromwidth,dromlength,stomwidth,chamarea,nrow=2,ncol=2)
-
-```
-
-In Figure \@ref(fig:dimensionsboxplots), we see that, through the centuries, tomb dimensions did vary, and sometimes consistently, as with the trend toward narrower *dromoi*.
-The 9th-century tombs, though, which include all but two of the 12 discussed above, do not particularly stand out.
-It does seem their chambers were, on average, larger, as well as exhibiting greater variation than at any other time.
-But then again, even the largest chambers are not exactly cavernous, and the extra labour expended would have been limited.
-The *dromos* is a plausible area for conspicuous demonstrations of scale but, with their frequently poor preservation, little confidence is possible.
-It appears at any rate that they were gradually shrinking, albeit with some notable outliers.
-
-```{r tombgroupdimensions, fig.dim = c(7, 4), dpi=600, fig.align='center', fig.cap="goes around here"}
-
-ggplot(subset(tombs,century_built=="9th" | century_built=="10th"),aes(dromos_length,chamber_area,color=group,label=tomb))+geom_point()+
-  theme_bw()+theme(text=element_text(family="Garamond", size=12))+
-  geom_label_repel(data=subset(tombs,group!="C"),aes(fill=group),color="white",show.legend = FALSE,family="Garamond")+
-            labs(x="Dromos Length (m)", y="Chamber Area (sq m)")+
-               scale_color_brewer(palette = "Set2",name="Group")+
-               scale_fill_brewer(palette = "Set2")+
-                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
-                  scale_y_continuous(expand=c(0,0),limits=c(0,10))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))+coord_fixed()
-
-```
-
-Looking more specifically at the tombs of Groups A and B, against the backdrop of the other 10th and 9th century tombs, there are perhaps hints of differentiation.
-All of them fall within the top 50% based on chamber area, and three have exceptionally long *dromoi*.
-But, later destruction notwithstanding, there appears nothing systematic about the variation visible here.
-If the size of one's tomb was intended to evoke or assert ideas of wealth or status, this appears to have occurred on a largely independent basis.
-
-```{r tombdimensions, fig.dim = c(7.5, 4), dpi=600, fig.align='center', fig.cap="goes around here"}
-
-tombareas<-ggplot(subset(tombs,tombs$est_burials>1),aes(chamber_area,est_burials,color=group))+geom_point()+
-  theme_bw()+theme(text=element_text(family="Garamond", size=12))+
-            labs(x="Chamber Area (sq m)", y="Estimated Total Burials")+
-               scale_color_brewer(palette = "Set2",name="Group")+
-                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
-                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-dromoslengths<-ggplot(subset(tombs,tombs$est_burials>1),aes(dromos_length,est_burials,color=group))+geom_point()+
-  theme_bw()+theme(text=element_text(family="Garamond", size=12))+
-            labs(x="Dromos Length (m)", y="Estimated Total Burials")+
-               scale_color_brewer(palette = "Set2",name="Group")+
-                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
-                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-ggarrange(tombareas,dromoslengths,ncol=2,common.legend = TRUE,legend="bottom")
-
-```
-
-What we can say is that no clear relationship existed between the dimensions of a tomb and its eventual buried population (see Figure \@ref(fig:tombdimensions)).
-Many of the very largest tombs are middling when it comes to burial numbers, and vice versa.
-So whatever the scale of the tombs connoted when built, this proved irrelevant to their later reuse.
-Several had niches cut into their *dromoi*, or the *stomion* extended, to fit in later burial urns (including T.
-292, 283 and 107; Cavanagh 1996: 658) and, sometimes, urns from within the chamber were relocated to make space for new interments (including T.
-75 and TFT; ibid.: 658; Brock 1957: 3–4).
-The impression is of competition for inclusion into the sepulchral group, further confirming that the tombs’ original builders little anticipated their eventual popularity.
-
-##Imports, Interments, and Indices of Wealth
-
-Dimensions aside, we can investigate the contemporary use of these tombs through comparisons between the scale and composition of their respective assemblages through time.
-
-```{r tombscompthroughtime, fig.dim=c(7,4),dpi=600,fig.align='center',fig.cap="goes around here"}
-
-pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A",
-       ifelse(pots$tomb=="X"|pots$tomb=="G"|pots$tomb=="219"|pots$tomb=="Q"|pots$tomb=="283","B","C"))
-tombs9th10th<-subset(tombs,century_built=="9th" & total_pottery>50|century_built=="10th" & total_pottery>50)
-pots9th10th<-subset(pots,tomb_century=="9th"|tomb_century=="10th")
-
-tombs9th10th_ranked<-tombs9th10th[with(tombs9th10th, order(-tombs9th10th$total_pottery)), ]
-tombs9th10th_ranked<-as.vector(tombs9th10th_ranked$tomb)
-pots9th10th$tomb<-factor(pots9th10th$tomb,levels=tombs9th10th_ranked)
-pots9th10th<-subset(pots9th10th,century=="9th" & tomb!="NA"|century=="8th" & tomb!="NA"|century=="7th" & tomb!="NA")
-pots9th10th$century<-mapvalues(pots9th10th$century, from = c("9th","8th","7th"), to = c("9th Century", "8th Century","7th Century"))
-
-ggplot(pots9th10th,aes(tomb,fill=group))+geom_bar()+
-  theme_bw()+theme(legend.position="bottom",text=element_text(family="Garamond", size=12),
-               axis.text.x = element_text(angle = 90,hjust=1,vjust=0.4))+
-                 guides(fill=guide_legend(title="Group"))+labs(x="Tomb", y="Vessel Count")+
-                  scale_fill_brewer(palette = "Set2")+ 
-                    scale_y_continuous(expand=c(0,0))+facet_grid(~century,scales="free_x",space="free")
-
-```
-
-And in Figure \@ref(fig:tombscompthroughtime), we see that the tombs of Group A, while not paultry in the 9th century, were by no means exceptional.
-They clearly kept pace with the other popular tombs of the time, but their major divergence only occurred in the subsequent two hundred years.
-Group B, meanwhile, experienced similar, if not greater, use in the 9th century, and then gradually declined.
-That is, these tombs seem to have followed a more conventional trajectory viz.
-Figure \@ref(fig:tombsinusebybuilddate).
-What might account for this disparity?
-And if the patterns implied in Figure \@ref(fig:tombscompthroughtime) - relative diversity, giving way to effective bipolarity in depositions - are genuine, how might we further investigate their manifestations across the excavated assemblage?
-
-## A Tale of Two Periods
-
-```{r bigbox, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="goes around here"}
-
-aorist_tomb_values<-tombs[c(1,158:207)]
-aorist_tomb_values[aorist_tomb_values == 0] <- NA
-
-aorist_tomb_values<-data.frame(aorist_tomb_values)
-colnames(aorist_tomb_values)[2:51]<-dates
-
-aorist_melt<-melt(aorist_tomb_values)
-aorist_melt$variable<-as.numeric(as.character(aorist_melt$variable))
-
-aorist_melt$period<-ifelse(aorist_melt$variable<=1100 & aorist_melt$variable>=980,"SM",
-		ifelse(aorist_melt$variable<=970 & aorist_melt$variable>=930,"EPG",
-		ifelse(aorist_melt$variable<=920 & aorist_melt$variable>=880,"MPG",
-		ifelse(aorist_melt$variable<=870 & aorist_melt$variable>=850,"LPG",
-		ifelse(aorist_melt$variable<=840 & aorist_melt$variable>=820,"PGB",
-	  ifelse(aorist_melt$variable<=810 & aorist_melt$variable>=800,"EG",
-	  ifelse(aorist_melt$variable<=790 & aorist_melt$variable>=750,"MG",
-	  ifelse(aorist_melt$variable<=740 & aorist_melt$variable>=720,"LG",
-	  ifelse(aorist_melt$variable<=710 & aorist_melt$variable>=680,"EO",
-	  ifelse(aorist_melt$variable<=670 & aorist_melt$variable>=640,"LO",
-	  ifelse(aorist_melt$variable<=630 & aorist_melt$variable>=600,"A",""
-		)))))))))))
-
-decade_ranges<-seq(from = 600, to = 1100, by = 10)
-decade_ranges<-rev(decade_ranges)
-decade_ranges<-as.factor(decade_ranges)
-aorist_melt$variable<-as.factor(aorist_melt$variable)
-aorist_melt$variable<-factor(aorist_melt$variable, levels=(decade_ranges))
-aorist_melt$period<-factor(aorist_melt$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
-aorist_melt$variable<-as.numeric(as.character(aorist_melt$variable))
-
-ggplot(aorist_melt,aes(variable,value,fill=period))+geom_boxplot(aes(group = cut_width(variable, 10)),outlier.size=0.8,lwd=0.2)+theme_bw()+
-      theme(text=element_text(family="Garamond", size=12))+
-            labs(x="Date (Years BC)", y="Aoristic Sums per Tomb")+
-               scale_fill_brewer(name="Period",palette = "Set3")+
-                scale_x_reverse(expand=c(0,0),limits=c(1000,600))+
-                  scale_y_continuous(expand=c(0,0),limits=c(0,20))
-
-```
-
-Figure \@ref(fig:bigbox) is a boxplot of the total aoristic sums across both cemeteries, colour-coded by ceramic phase.
-Looking at the PGB and EG, we see that they have the largest interquartile ranges (IQRs) of any period, and only a single outlier (beyond 1.5 times the IQR) each.
-The Orientalising period, on the other hand, exhibits much reduced IQRs, and a proliferation of outliers (namely, Group A) far and above the range of the other tombs.
-In other words, whilst both the PGB-EG and EO-LO periods witnessed an expanding volume of burial activity, in the earlier period this was distributed far more evenly across all tombs.
-One speaks of increasing plurality, the other of divergence.
-
-```{r gini, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="goes around here"}
-
-aorist_tomb_values<-tombs[c(1,158:207)]
-aorist_tomb_values[aorist_tomb_values == 0] <- NA
-
-aorist_tomb_values<-data.frame(aorist_tomb_values)
-colnames(aorist_tomb_values)[2:51]<-dates
-
-tombs_gini<-matrix(0, ncol = 0, nrow = 1)
-tombs_gini<-data.frame(tombs_gini)
-for(i in colnames(aorist_tomb_values[2:51])) {
-  tombs_gini[[paste0(i)]] <- ineq(aorist_tomb_values[[i]],type="Gini")
-}
-
-tombs_gini<-t(tombs_gini)
-tombs_gini<-data.frame(tombs_gini)
-
-colnames(tombs_gini)<-c("gini_coefficient")
-row.names(tombs_gini)<-dates
-tombs_gini<-data.frame(tombs_gini)
-setDT(tombs_gini, keep.rownames = "decade")
-tombs_gini$decade<-as.numeric(tombs_gini$decade)
-
-rects <- data.frame(xstart = c(1100,970,920,875,840,810,790,745,710,670,630), 
-                    xend = c(970,920,875,840,810,790,745,710,670,630,600), 
-                    col = c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
-rects$col<-factor(rects$col,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
-
-ggplot() + 
-  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf, fill = col), alpha = 0.4) +
-     geom_line(data = tombs_gini, aes(decade,gini_coefficient),size=0.5)+
-      theme_bw()+theme(text=element_text(family="Garamond", size=12))+
-              labs(x="Date (Years BC)", y="Gini Coefficient")+
-                 scale_fill_brewer(name="Period",palette = "Set3")+ coord_cartesian(
-                    xlim = c(1100, 600), ylim = c(0.5,0.8))+ scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
-                       scale_y_continuous(expand=c(0,0))
-```
-
-This characterisation is supported further by considering the Gini coefficient for the distribution of pottery between the tombs (see Figure \@ref(fig:gini)).
-Using aoristic sums as a probablistically-weighted estimate of the volume of material deposited in each tomb in each decade, the Gini coefficient can give an (albeit rough) estimate of how unequally distributed that material was.
-And, though values were always relatively high (i.e.
-unequal) within the cemeteries, the major spike in the Gini coefficient seen in the Orientalising period does contrast markedly with the plateau of the PGB-EG.
-
-This is not to suggest that the late 9th to early 8th century#ies was a time of egalitarian burial practices.
-The PGB and EG are consistently interpreted as periods of artistic development, increased overseas trade, and renewed vigour in the construction of tombs and the visibility of the burial rite [@Brock1957: 143-44; @Coldstream1994: 110-12].
-The PGB style reflects a uniquely Cretan blend of Attic and Near Eastern forms, while the appearance of BA 'antiques' in the tombs from this period onwards is felt to reflect rejuvinated, even competitive interest in the Minoan past.
-
-```{r imports, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="goes around here"}
-
-imports<-subset(pots,pots$import!="")
-greekimports<-subset(imports,imports$import_region=="Greece")
-NEimports<-subset(imports,imports$import_region=="Near East")
-imitations<-subset(pots,pots$imitation!="")
-
-import_aorist<-colSums(imports[27:76])
-import_aorist<-data.frame(import_aorist)
-colnames(import_aorist)<-c("aoristic_sum")
-row.names(import_aorist)<-dates
-setDT(import_aorist, keep.rownames = "decade")
-import_aorist$decade<-as.numeric(import_aorist$decade)
-
-greekimport_aorist<-colSums(greekimports[27:76])
-greekimport_aorist<-data.frame(greekimport_aorist)
-colnames(greekimport_aorist)<-c("aoristic_sum")
-row.names(greekimport_aorist)<-dates
-setDT(greekimport_aorist, keep.rownames = "decade")
-greekimport_aorist$decade<-as.numeric(greekimport_aorist$decade)
-
-NEimport_aorist<-colSums(NEimports[27:76])
-NEimport_aorist<-data.frame(NEimport_aorist)
-colnames(NEimport_aorist)<-c("aoristic_sum")
-row.names(NEimport_aorist)<-dates
-setDT(NEimport_aorist, keep.rownames = "decade")
-NEimport_aorist$decade<-as.numeric(NEimport_aorist$decade)
-
-imitation_aorist<-colSums(imitations[27:76])
-imitation_aorist<-data.frame(imitation_aorist)
-colnames(imitation_aorist)<-c("aoristic_sum")
-row.names(imitation_aorist)<-dates
-setDT(imitation_aorist, keep.rownames = "decade")
-imitation_aorist$decade<-as.numeric(imitation_aorist$decade)
-
-greekimport_aorist$origin<-"Greek Imports"
-NEimport_aorist$origin<-"Near Eastern Imports"
-imitation_aorist$origin<-"'Oriental' Imitations"
-all_import_imitation<-rbind(greekimport_aorist,NEimport_aorist,imitation_aorist)
-
-ggplot()+ geom_col(data=all_import_imitation,aes(decade,aoristic_sum, fill=origin),position="dodge")+
-            theme_bw()+theme(text=element_text(family="Garamond", size=12), legend.position="bottom")+
-                labs(x="Date (Years BC)", y="Aoristic Sum")+
-                  scale_fill_brewer(name="Origin",palette = "Set2")+
-                   scale_y_continuous(expand=c(0,0),limits=c(0,20))+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))
-
-imports1<-nrow(subset(pots,pots$group=="B" & pots$import =="Y"))/nrow(subset(pots,pots$group=="B"))*100
-imports2<-nrow(subset(pots,pots$group=="C" & pots$import =="Y"))/nrow(subset(pots,pots$group=="C"))*100
-imports3<-nrow(subset(pots,pots$group=="A" & pots$import =="Y"))/nrow(subset(pots,pots$group=="A"))*100
-groupA7th<-subset(pots,group=="A" & century=="7th")
-groupA7thimports<-(sum(groupA7th$import=="Y")/nrow(groupA7th))*100
-
-```
-
-The rise in imported items is readily observable, with a succession of Greek and Near Eastern imports in the late 9th and early 8th centuries (Figure \@ref(fig:imports)).
-Likewise, the decline in Near Eastern imports, and concomitant rise in locally-made imitations thereof, is visible at the end of the 8th century.
-These trends being a very coarse proxy for its engagement with the wider Mediterranean, the late 9th and early 8th centuries would seem to encompass the most outward- (certainly,eastward-) looking phase of the EIA Knossian community [see @Whitley2004: 434].
-
-In the Group B tombs, with their more limited flourescence centred on the PGB-EG periods, imported items constitute `r round(imports1, digits=2)`% of the total excavated assemblage, compared to `r round(imports2, digits=2)`% in Group C and `r round(imports3, digits=2)`% in Group A.
-Indeed, among 7th century vessels, when interments were higher than ever among Group A, the figure is a mere `r round(groupA7thimports, digits=2)`%.
-The impression grows stronger that the context for the late proliferation of burials in the Group A tombs was altogether different from that underpinning the comparative, competitive diversity of interments a century earlier.
-
-```{r comps, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="goes around here"}
-
-pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A",
-       ifelse(pots$tomb=="X"|pots$tomb=="G"|pots$tomb=="219"|pots$tomb=="Q"|pots$tomb=="283","B","C"))
-
-potcompKNC<-ggplot(subset(pots,century!="" & cem_condensed=="NC"),aes(century,fill=type))+geom_bar(position="fill")+
-  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title = element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(title="KNC",x="", y="Proportion of Total Assemblage")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" Amphora  "," Jug  "," Oil vessel  "," Pithos  "," Drinking vessel  "," Lid  "," Other  "))+ 
-                    scale_y_continuous(expand=c(0,0.0))
-potcompFOR<-ggplot(subset(pots,century!="" & cem_condensed=="FOR"),aes(century,fill=type))+geom_bar(position="fill")+
-  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(title="Fortetsa",x="Century BC", y="")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" Amphora  "," Jug  "," Oil vessel  "," Pithos  "," Drinking vessel  "," Lid  "," Other  "))+ 
-                    scale_y_continuous(expand=c(0,0.0))
-
-potcomptop7<-ggplot(subset(pots,century!="" & century!="11th"& group=="A"),aes(century,fill=type))+geom_bar(position="fill")+
-  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(title="Group A",x="", y="")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" Amphora  "," Jug  "," Oil vessel  "," Pithos  "," Drinking vessel  "," Lid  "," Other  "))+ 
-                    scale_y_continuous(expand=c(0,0.0))
-
-ggarrange(potcompKNC,potcompFOR,potcomptop7,ncol=3,common.legend =TRUE,legend="bottom")
-
-spearmanobjectspots<-cor(tombs$total_object,tombs$total_pottery,method="spearman")
-```
-
-This is reflected further in the changing composition of the mortuary assemblage through time (Figure \@ref(fig:comps)).
-Across both cemeteries, there appears to have been the steady crystallisation of a basic 'burial kit', comprising a pithos and lid, an oil vessel or two, a jug, and a drinking vessel.
-The proportion of grave goods comprising 'other' vessels - pyxides, askoi, trays, stands, dinoi, kernoi, kraters and more - declines to become a near insignificant part of the standardised repertoire of the 7th century.
-The tombs of Group A did not break this mould, and their assemblages seem largely as conventional as any other by this time.
-
-The main focus of this article has been on ceramics and not other artefacts of metal, stone, glass, and organic materials, due to their less well-developed chronologies, but even the raw distribution of such artefacts may make more sense in the light of the foregoing discussion.
-There is a relatively strong (Spearman's ρ=`r round(spearmanobjectspots,digits=2)`) correlation between the number of ceramic and non-ceramic finds across the tombs, and, certainly, many of the most seemingly precious objects were recovered from among the tombs with the most burials.
-
-```{r hists, fig.dim=c(6,4),dpi=600, fig.align='center',fig.cap="goes around here"}
-tombs$group<-revalue(tombs$group,c("A"="A","B"="B","C"="B"))
-total_object<-ggplot(subset(tombs,est_burials>1 & total_object>0),aes(total_object, fill=group))+geom_histogram()+
-  theme_bw()+theme(legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Non-Ceramic Finds", y="Number of Tombs")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" Group A  ", " Other Tombs"))+
-                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-total_weapon<-ggplot(subset(tombs,est_burials>1 & weapon>0),aes(weapon, fill=group))+geom_histogram()+
-  theme_bw()+theme(legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Weapons", y="Number of Tombs")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" Group A  ", " Other Tombs"))+
-                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-total_jewellery<-ggplot(subset(tombs,est_burials>1 & jewellery>0),aes(jewellery, fill=group))+geom_histogram()+
-  theme_bw()+theme(legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Jewellery", y="Number of Tombs")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" Group A  ", " Other Tombs"))+
-                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-
-objectsbyburial<-ggplot(subset(tombs,est_burials>1 & total_object>0),aes(total_object/est_burials, fill=group))+geom_histogram()+
-  theme_bw()+theme(legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Non-Ceramics/Burials", y="Number of Tombs")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" Group A  ", " Other Tombs"))+
-                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-weaponbyburial<-ggplot(subset(tombs,est_burials>1 & weapon>0),aes(weapon/est_burials, fill=group))+geom_histogram()+
-  theme_bw()+theme(legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Weapons/Burials", y="Number of Tombs")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" Group A  ", " Other Tombs"))+
-                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-
-jewellerybyburial<-ggplot(subset(tombs,est_burials>1 & jewellery>0),aes(jewellery/est_burials, fill=group))+geom_histogram()+
-  theme_bw()+theme(legend.title=element_blank(),
-               text=element_text(family="Garamond", size=12))+
-                 labs(x="Total Jewellery/Burials", y="Number of Tombs")+
-                  scale_fill_brewer(palette = "Set2",labels=c(" Group A  ", " Other Tombs"))+
-                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
-
-ggarrange(total_jewellery,total_weapon,total_object,jewellerybyburial,weaponbyburial,objectsbyburial,ncol=3,nrow=2,common.legend = TRUE,legend="bottom")+
-            theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-```
-
-But this association, perhaps tempting one initially to speak of 'elite' or 'high-status' tombs, evaporates when the number of objects is divided by the number of burials represented.
-This fact has seldom, if ever, been appreciated in prior discussions of the cemeteries (Figure \@ref(fig:hists)).
-Whitley (2004: 437) has pointed out that the proportions of different decorative styles of Orientalising pithoi in Tomb P are replicated across Fortetsa as a whole, but it seems this situation may apply across most artefact classes.
-If so, the late surge in burial activity at both Fortetsa and the KNC was not linked in any straightforward way to simple principles of 'elite' or 'wealthy' burial; many of the most-used tombs prove exceptional only in their absolute quantities of artefacts, and not the accoutrements of each individual burial.
-This fact has seldom been appreciated in prior discussions of the cemeteries.
-
-# Discussion
-
-To summarise, towards the end of 8th century BC, a select group of tombs, all built back in the 9th or 10th, became the foci of a major increase in burials which didn't register among the tombs built during that time.
-Little about the physical dimensions or early use of these tombs presaged their later popularity, while a similar, if smaller, increase in burial activity around a century earlier, in the PGB-EG periods, contrasted in the comparative diversity and broader distribution of its associated assemblages.
-
-It remains to consider how we might characterise the burying groups behind these various, intriguing trends.
-The traditional view is that such groups were immediate families (Brock 1957, 41; @Boardman1967: 63; @Coldstream1996: 291).
-The relatively short lifespan of many tombs has been taken as evidence for direct, lineal inheritance, with the tomb going out of use as families moved, individuals founded new tombs, or the line fissioned or ended [@Cavanagh1996: 659].
-The idea of family groups is supported by the, admittedly circumstantial, evidence of men, women and children being found within single tombs, and one instance of a (putatively) hereditary mandibular deformity in three skulls in the same tomb [@Musgrave1996].
-
-The model of lineal inheritance has been recently challenged by Kotsonas [-@Kotsonas2011: 131], however, who considers it ‘probably the greatest factoid in the archaeology of Crete of the given period’.
-He argues that the rise in interments during the 8th century is too great to be explained by population growth, as would be necessary for the possibility of direct lineal inheritance to be preserved.
-Rather, the trend must reflect reconfigurations of the burying groups themselves, which Kotsonas feels may have (also or instead) comprised age- and sex-grades.
-We should take this possibility seriously, for one need only look at Tomb P in the 7th century to see that we cannot possibly be dealing with just immediate families.
-
-Yet it is hard to be conclusive, as our current understanding of the social organisation of the early Cretan *poleis* remains patchy.
-Perlman and Gagarin's [-@Gagarin2016] recent study of the Cretan law codes reminds us that there is an often underappreciated degree of diversity in the offices, institutions, and social groups of the early city-states.
-In the traditional view, political authority was vested in the *kosmoi*, aristocratic magistrates selected from the leading, hereditary *startoi*, or ‘clans’, networks of extended lineages controlling the ancestral estates on which worked attendent serfs and/or slaves (Willetts 1982, 240).
-But despite elements of this account probably being accurate, others remain equivocal, such as the nature of the *pyla*, or 'tribe', and its relation to the *startos* (which may itself be a military grouping), the distinction between *astos* and *poliatas*, both probably formulations of the rights-bearing civic actor, and the ancestry and spread of the *agelaos*, an adolescent age-grade currently only attested at Eltynia [@Gagarin2016: 36; 78-80].
-Indeed, many terms which might illuminate social divisions and groupings within the Archaic Cretan populace are only attested at one or two sites, and even then sometimes apparently differ in form or function.
-
-My aim, however, is not to draw a simple parallel between the Knossian burial groups and, say, powerful *startoi*, a term not even attested outside of Gortyn in this period (ibid: 1-2).
-The fundamental conclusion of the disucussed contrasts between the late 9th and late 8th centuries, between the Group A and B tombs, is not that the burying groups changed, as such, it is that wider socio-economic and political contexts did, in ways which prompted reformulations of funerary practice, one of which being the intense, iterative burial activity evinced by the patrons of the Group A tombs.
-Such inferences, naturally, recall the conclusions of Morris [-@Morris1987: 171-211], who argued 30 years ago that the emergence of the *polis* across the Aegean led to a tension between elite preferences for ostentatious display, and a downplaying of social difference in death, as promoted by the ‘middling’ ideology of the city-state.
-In his well-known model of the *agathoi* and *kakoi*, Morris claimed that, at various times, the relevant spheres of kinship demonstrated in burial could widen and constrict, with more and less distantly related lineages drawn into association through changing social strategies, until the pre-eminence of kinship relations in the structuring of political power declined, along with its manifestiations in mortuary practice.
-
-At Knossos, some aspects of the noted expansion in the burying population seem explicable within a model like Morris's.
-The basic principle, of elites employing access to burial as a strategic act of association and social signalling, is almost certainly in evidence here.
-That the noted spate of burials in the late 8th and early 7th centuries was concentrated solely in older tombs strongly implies that issues of heritage, lineage, and tradition were at play.
-In Tomb 107, the stratigraphy of the niches cut into the *dromos* suggest depositions continued to be made even after the main chamber fell into disrepair [@Coldstream1996a: 150]; it is the symbolic and associative, rather than physical qualities of these tombs seem instrumental to their extensive reuse.
-
-Other features sit less easily with a Morrissian reading of the evidence, though.
-Firstly, the burials at issue are in many ways *not* ostentatious in comparison to the norm.
-We have seen that the 'richest' tombs in terms of finds are far less so when one accounts for the number of burials represented.
-Likewise, Morris claims that periods of expansion in the rite of burial will correspond to increased diversity in the associated assemblages.
-Yet, at Knossos, the range of grave goods diminished and crystallised during the late 8th and early 7th centuries.
-The PGB period better embodies such an expansion, with its greater variety in chamber dimensions and relatively broad distribution of (local and imported) grave goods, not to mention the innovative ceramic decoration and reuse of BA artefacts.
-In both the late 9th and late 8th centuries, then, certain burying groups found it propitious to broaden access to the tombs they patronised, but changing social contexts saw those expansions manifested in quite different ways.
-
-Elsewhere on Crete in the 7th century, we see a shift to unmarked, individual burials, and more austere grave goods, as well as an apparent reorientation of sanctuary deposits away from extra-mural to suburban spaces, and a rise in utilitarian, mass-produced votives [@Gagarin2016: 34-36].
-At Knossos, a small number of pithos burials in pits do appear in the KNC and Fortetsa volumes (Brock 1957: 98; Coldstream and Catling 1996: 162), and recent re-excavation of the Teke plot identified numerous small pit burials around three chamber tombs [@Evely2012], raising the possibility that others exist, but have gone unrecognised in prior excavations.
-At Azoria, major urban remodelling in the 7th century saw the earlier, agglutinative house forms of neighbouring Kavousi Vronda and Kastro give way to larger, more formally planned dwellings with little evidence of remodelling, while larger, apparently civic, buildings were constructed to accommodate major public gatherings [@Haggis2011; @Haggis2011b; @Haggis2014].
-
-Together, these varied threads suggest that, with the emergence of the archaic *poleis* on Crete, there was a concomitant restructuring of wider social groups, institutions, and behaviours.
-It seems not unreasonable to infer an emergent emphasis on the citizen, the state, and ideologies of relative egality, within which communal gatherings, intra-mural ritual, and individual graves replaced ostentatious, kin-centred acts of gathering, dedication, and burial.
-And in the Knossian cemeteries, we have seen a somewhat diverse set of burial practices give way to a more prolific, yet more restricted, suite of funerary rites, focussed on a small number of long-established tombs, in the context of diminishing collective tomb construction, replacement of imports with locally-made imitations, and an imminent cessation of such depositions.
-
-Whether the increasingly homogenous burials reflected some standard set of accoutrements intentionally shared by the hypothetical *agathoi*, or else nascent, egalitarian strictures over elaborate burial remains difficult to judge.
-Yet we need not adopt wholesale a model like Morris's "middling" ideology, nor link the diminishing range of burial goods to Archaic Crete's purported austerity.
-As a more nuanced picture emerges of the island's relationship with the wider Mediterranean, subtler characterisations are possible.
-It is telling that the peak for imported items came in the 9th century; the Orientalising as a time period feels increasingly misapplied to Crete [@Whitley2004: 434], when, at Knossos at least, it witnessed far lower rates of 'oriental' imports, and only limited adoption of the figural motifs which characterise its appearance elsewhere.
-
-The decline in imports was, however, counterpointed by a rise in imitations of Near Eastern vessels [@Antoniadis2012: 170], according with our growing awareness of the selective nature of Archaic and Classical Crete's external relations.
-Erickson [@Erickson2010], similarly, has argued that *symposia* were practised in Crete, something often disputed, but that their practitioners merely favoured simpler vessel forms than the figurally-decorated wares popular on the mainland.
-Local, selective adoption of, and responses to, wider social and material trends is perhaps becoming the hallmark of Archaic Crete, and finds further endorsement in the EIA cemeteries at Knossos.
-What seems certain is that this was a period of transition wherein, whatever their exact identity, certain actors invested heavily in a final flurry of burials in chamber tombs, with an emphasis on a traditional, established repertoire of items, rites, and venues for one's passage to the afterlife, before changing social forces effected their abandonment.
-
-# Conclusion
-
-It is important that the present study was a comparative investigation of both the Fortetsa and Knossos North cemeteries.
-That both underwent comparable developments during the EIA suggests, not only that they catered to similar sub-sections of the Knossian population, but that some of the socio-economic and political changes inferred were significant and community-wide.
-Further work should seek to extend these comparisons, both into trends unexplored in the present study, and out to the wider funerary landscape of Knossos.
-I hope to do just this but, for the time being, the datasets I am making available alongside this article include the other published EIA tombs in the area, so that others may also explore their possible significance.
-
-There were undeniable limitations to the present study.
-For one, non-ceramic artefacts were underrepresented due to difficulties with dating, and future work could seek to address this.
-Issues of gender were sadly little considered, though this was largely due to limitations in the data.
-There have also been a number of tombs discovered in the last 20 years that fill in the spaces between the Teke plot and the Medical Faculty Site, and ideally these unpublished tombs should be included in further analyses as they, along with the Teke tholos and associated tombs, are probably part of a single contiguous cemetery [@Blackman1999: 113; @Catling1983: 51; @French1991: 68-69].
-Finally, a preliminary study was recently undertaken by the author of select boxes marked as 'unpublished' from the KNC deposits kept at the British School at Athens, at Knossos.
-Despite little mention in the publication, this material was not inconsiderable and so, regretably, future work will be necessary to determine whether it threatens the integrity of these analyses.
-Suffice to say that, at present, it is another pinch of salt with which to take the present findings.
-
-These limitations notwithstanding, the foregoing analysis demonstrates the value of applying quantitative approaches to the archaeological record of the period.
-Though the construction of databases can be time consuming, once made they offer the opportunity to subdivide, compare, and plot the data in numerous ways, facilitating investigation of the material from multiple angles.
-By making datasets available for use by others, as with the data and code for the present study, one’s conclusions can also be independently tested, and alternative interpretations can be voiced.
-The databases produced for the above analysis contain a large of amount of unutilised data that could facilitate a range of future investigations, and all are welcome to it.
-With more such analyses, we could begin to a construct a fuller account of the regional diversity in mortuary, and wider social, practices that characterised the communities of EIA and archaic Crete.
-
-##### pagebreak
-# Acknowledgements
-
-I would like to thank Borja Legarra Herrero and Corinna Riva for their help and advice from the earliest stages of the present research, as well as the BSA at Knossos for permitting me to examine some of the material from the Medical Faculty excavations.
-I am indebted most of all to Professor Todd Whitelaw, for his unstinting guidance, feedback and facilitation of my ongoing research.
-
-# About the Author
-
-Dominic Pollard is a PhD candidate at University College London.
-His research focusses on the Early Iron Age of Crete, with particular interest in the use of computational methods to supplement traditional modes of archaeological inference.
-
-##### pagebreak
-# References 
-<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
-<div id="refs"></div>
-
-##### pagebreak
-# Captions
-
-*Figure 1 Map of the Knossos area, showing the locations of the various EIA tomb clusters and cemeteries, with EIA pottery distributions from the Knossos Urban Landscape Project marked in orange, giving an impression of the contemporary urban extent (underlying image credit: T.
-Whitelaw).*
-
-*Figure 2 Relationship between known interments, counts of pithoi, and counts of pithoi, amphorae and kraters across both cemeteries.
-The straight line represents the mean estimate based on these three values.*
-
-*Figure 3 Counts of pithoi across both cemeteries through time. Vessel dates have here been based on the midpoint (or mean) of the absolute date range accompanying their respective ceramic periods.*
-
-*Figure 4 Aoristic sums of all vessels across both cemeteries. This represents a probabilistically weighted visualisation of depositional activity through time.*
-
-*Figure 5 Aoristic sums of pithoi (left), as a proxy for cremation burials, and of all vessels (right). The sums are colour coded to show the values of the two cemeteries, and the total values*
-
-*Figure 6 Tomb construction through time at both cemeteries.
-On the left, counts are based on the ceramic phase considered most likely to correspond to the tomb's first use (including simply 'PG', 'G' and 'O' where no greater specificity was possible).
-On the right, aoristic sums are presented, spreading the uncertainty associated with those general assignations.
-In both cases, tombs lacking any securely dateable pottery are excluded.*
-*Abbreviations*
-*SM - Subminoan; (E/M/L)PG - (Early/Middle/Late) Protogeometric; PGB - Protogeometric B; (E/M/L)G - (Early/Middle/Late) Geometric; (E/L)O - (Early/Late) Orientalising*
-
-*Figure 7 Estimates of the number of tombs in use during each of the ceramic phases attested in both cemetaries, based on counts of tombs yielding vessels of each period.*
-
-*Figure 8 Estimates of the number of tombs in use through time across both cemeteries.
-The grey background represents the total number of tombs in use, while the smaller coloured bars correspond to tombs built in each of the five centuries of EIA activity at the cemeteries.*
-
-*Figure 9 Counts of vessels from all tombs in both cemeteries, arranged, on the left, by the century to which they are assigned based on their ceramic period and, on the right, by the centuries in which the tombs they come from were built.*
-
-*Figure 10 Aoristic sums for the 12 tombs with the largest ceramic assemblages, arranged in decreasing order.
-Note, for the sake of better illustrating the relative temporal patterns, rather than absolute values, the scales on each plot are distinct.*
-
-*Figure 11 Aoristic sums for the Group A and Group B tombs, as defined above.*
-
-*Figure 11 Boxplots comparing various tomb dimensions, divided, in each case, by the century of tomb construction.*
-
-*Figure 12 Scatterplot comparing the dimensions of* dromos *length and chamber area.
-Groups A and B, defined earlier, are illustrated and labelled in green and orange respectively, while all remaining tombs ('Group C') are in blue.*
-
-*Figure 13 Scatterplots comparing tomb dimensions (chamber area and* dromos *length) to the estimated total number of burials by tomb. The points area colour-coded as per Figure 12.*
-
-*Figure 14 Counts of vessels dating to the 9th, 8th and 7th centuries respectively, found in tombs built in the 10th or 9th century which yielded upwards of 50 vessels when excavated.
-The tombs are colour-coded as per Figure 12.*
-
-*Figure 15 Boxplot showing the aoristic sums of all vessels across both cemeteries through time, colour-coded by ceramic period.
-Boxes represent the interquartile range, horizontal lines the median value, whiskers values within 1.5 times the interquartile range, and points outliers beyond this.
-Note, the Y axis has been limited to 20 to aid readability, meaning the outliers Tomb G in the EG period, and Tomb P in the late EO-LO period are excluded, with values ranging from 29-34, and 36-44 respectively.*
-
-*Figure 16 Gini coefficient values through time based on the aoristic sums for all vessels across both cemeteries. Background colours correspond to ceramic periods.*
-
-*Figure 17 Aoristic sums of imported and 'oriental' imitation vessels through time across both cemeteries, colour-coded by regional origin in the case of imports.*
-
-*Figure 18 The composition of the total pottery assemblage for each century across both cemeteries, and among the 7 notable tombs (Group A) drawn out earlier in the analysis.*
-
-*Figure 19 Histograms corresponding, first, to counts of jewellery (beads, necklaces, rings, diadems etc.), metal objects, and imported items by tomb across both cemeteries and, secondly, to those counts divided by the estimated number of burials in each tomb.*
-
-##### pagebreak
-# Figures
-
-##### pagebreak
-# Tables
 
@@
 
+The significance of these findings can be seen in Table 1.
+Using pithoi as a proxy for burials, we can estimate the changing rates of burial through time.
+And while these mostly remained around three per tomb per century, in the 8th and 7th centuries, tombs founded in the 9th century saw a rapid rise in the rate of burial, peaking at over 11 per tomb per century.
+Though no doubt rough estimates, the marked nature of this divergence suggests it is not simply an artefact of imperfect data.
+
+But Table 1 also demonstrates the markedly low rates of burial which predominated at both cemeteries; two to four burials per tomb per century amounts to a very selective burial rite.
+It seems unlikely, then, that the rising rates of burial are, for instance, the direct result of population growth; the increase is too rapid, and concentrated only in older tombs, while, based on KULP’s survey data, the major population increase at EIA Knossos probably occurred back in the Protogeometric period (Whitelaw, pers.
+comm.).
+
+## 4.2 Isolating the Phenomenon
+
+We are yet to draw out which older tombs precisely are responsible for the trend so far discussed.
+We might assume those which became the focus of such intense later burial activity would end up yielding the most burials and grave goods when excavated and, happily, this supposition proves true.
+
+```{r toptombs, fig.dim = c(7, 5), dpi=600, fig.align='center',fig.cap="goes around here"}
+
+tombs_ranked<-tombs[with(tombs, order(-tombs$total_finds)), ]
+tombs_top_12<-head(tombs_ranked,12)
+top12<-as.vector(tombs_top_12$tomb)
+pots_top_12<-pots[is.element(pots$tomb, top12),]
+pots_top_12$tomb<-factor(pots_top_12$tomb,levels=top12)
+
+cols<-c(1,158:207)
+aorist_all_tomb_names<-tombs_top_12[,cols]
+aorist_all_tomb_names<-data.frame(aorist_all_tomb_names)
+colnames(aorist_all_tomb_names)<-c("tomb",dates)
+
+aorist_all_tomb_names$cemetery<-tombs_top_12$cem_condensed
+
+aorist_all_tomb_melt<-melt(aorist_all_tomb_names)
+aorist_all_tomb_melt$variable<-as.numeric(as.character(aorist_all_tomb_melt$variable))
+
+top_12<-tombs_top_12$tomb
+aorist_all_tomb_melt$tomb<-factor(aorist_all_tomb_melt$tomb,levels=top_12)
+
+ggplot(aorist_all_tomb_melt,aes(variable, value,fill=cemetery))+geom_col()+theme(legend.position = "none")+facet_wrap(~tomb,scales="free_y")+
+  geom_col()+theme_bw()+theme(legend.title=element_blank(),legend.position="bottom",
+      text=element_text(family="serif", size=12),
+      axis.text.x=element_text(angle=45, vjust=1.2, hjust=1))+
+          labs(x="Date (Years BC)", y="Aoristic Sums")+
+              scale_fill_manual(values = c("gray50","gray80"),labels=c(" KNC  "," Fortetsa"))+ 
+                  coord_cartesian(xlim=c(1100,610))+scale_y_continuous()+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1100,610))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+tombs$group<-ifelse(tombs$tomb=="P"|tombs$tomb=="292"|tombs$tomb=="75"|tombs$tomb=="107"|tombs$tomb=="285"|tombs$tomb=="II"|tombs$tomb=="218","A","B")
+total_pottery_by_groups<-aggregate(tombs$total_pottery, by=list(tombs$group), FUN=sum)
+calc1<-(total_pottery_by_groups[1,2]/sum(total_pottery_by_groups$x))*100
+
+```
+
+The 12 tombs with the most associated vessels are all chamber tombs, 10 dating to the 9th century and two to the 10th, and seven of them did exhibit late, intense episodes of deposition (Figure \@ref(fig:toptombs)).
+In fact, Tombs P, 292, 75, 107, 285, II and 218 (henceforth 'Group A') between them account for `r round(calc1,digits=2)`% of the entire ceramic assemblage from both cemeteries.
+All but Tomb 285 date to the 9th century, and together they underpinned the discussed late spate of burials.
+
+```{r groupsab, fig.dim=c(8,4),dpi=600, fig.align='center',fig.cap="goes around here"}
+
+tombs$group<-ifelse(tombs$tomb=="P"|tombs$tomb=="292"|tombs$tomb=="75"|tombs$tomb=="107"|tombs$tomb=="285"|tombs$tomb=="II"|tombs$tomb=="218","A",
+                   ifelse(tombs$tomb=="X"|tombs$tomb=="G"|tombs$tomb=="219"|tombs$tomb=="Q"|tombs$tomb=="283","B","C"))
+
+GroupA<-subset(tombs,group=="A")
+
+aorist_GroupA <- colSums(GroupA[158:207])
+aorist_GroupA<-data.frame(aorist_GroupA)
+
+colnames(aorist_GroupA)<-c("aoristic_sum")
+row.names(aorist_GroupA)<-dates
+aorist_GroupA<-data.frame(aorist_GroupA)
+setDT(aorist_GroupA, keep.rownames = "decade")
+aorist_GroupA$decade<-as.numeric(aorist_GroupA$decade)
+
+aorist_GroupA$group<-"A"
+
+GroupB<-subset(tombs,tombs$group=="B")
+
+aorist_GroupB <-colSums(GroupB[158:207])
+aorist_GroupB<-data.frame(aorist_GroupB)
+
+colnames(aorist_GroupB)<-c("aoristic_sum")
+row.names(aorist_GroupB)<-dates
+aorist_GroupB<-data.frame(aorist_GroupB)
+setDT(aorist_GroupB, keep.rownames = "decade")
+aorist_GroupB$decade<-as.numeric(aorist_GroupB$decade)
+
+aorist_GroupB$group<-"B"
+
+aorist_groups<-rbind(aorist_GroupA,aorist_GroupB)
+
+ggplot(aorist_groups,aes(decade,aoristic_sum,fill=group))+geom_col(position="dodge")+
+  theme_bw()+theme_bw()+theme(text=element_text(family="serif", size=12))+
+                labs(x="Date (Years BC)", y="Aoristic Sum")+
+                  scale_fill_manual(values=c("gray50","gray80"),name="Group")+
+                   scale_y_continuous(expand=c(0,0),limits=c(0,120))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+```
+
+Interestingly, Tombs X, G, Q, 219 and 283 ('Group B') evince a different pattern, with their respective peaks coming around a century earlier, another period of increased burial activity.
+So despite receiving the greatest quantities of burials and grave goods, these tombs nonetheless seem to have achieved that distinction via quite different histories, a divergence made clear in Figure \@ref(fig:groupsab).
+It is notable that tombs assinged to each group are seen in both cemeteries, suggesting from the outset that the differences between these may speak to social forces operating across the Knossian community.
+
+## 4.3 Not All Tombs are Created Equal
+
+Having addressed some broad diachronic trends, let us now attempt some more synchronic comparisons, beginning with the physical dimensions of the tombs as built - a potentially rich context for the negotiation of social power and group membership.
+Perhaps the most bountiful tombs were also physically the largest, assertions of status embodied by the man-power invested in their construction.
+
+```{r dimensionsboxplots, fig.dim=c(7,6),dpi=600, fig.align='center',fig.cap="goes around here"}
+
+tombs$century_built<-factor(tombs$century_built,levels=c("11th","10th","9th","8th","7th"))
+dromwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_width,fill="gray80"))+geom_boxplot(lwd=0.2,outlier.size=0.8)+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="serif", size=10))+
+                 scale_fill_manual(values=c("gray80"))+
+                   labs(x="Century of Tomb Construction", y="Dromos Width (m)")+
+                    scale_y_continuous(expand=c(0,0.1))
+dromlength<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_length,fill="gray80"))+geom_boxplot(lwd=0.2,outlier.size=0.8)+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="serif", size=10))+
+  scale_fill_manual(values=c("gray80"))+
+                 labs(x="Century of Tomb Construction", y="Dromos Length (m)")+
+                    scale_y_continuous(expand=c(0,0.5))
+stomwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,stomion_width,fill="gray80"))+geom_boxplot(lwd=0.2,outlier.size=0.8)+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="serif", size=10))+
+  scale_fill_manual(values=c("gray80"))+
+                 labs(x="Century of Tomb Construction", y="Stomion Width (m)")+
+                    scale_y_continuous(expand=c(0,0.1))
+chamarea<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,chamber_area,fill="gray80"))+geom_boxplot(lwd=0.2,outlier.size=0.8)+
+  theme_bw()+theme(legend.position="none",
+               text=element_text(family="serif", size=10))+
+  scale_fill_manual(values=c("gray80"))+
+                 labs(x="Century of Tomb Construction", y="Chamber Area (sq. m)")+
+                    scale_y_continuous(expand=c(0,0.5))
+
+ggarrange(dromwidth,dromlength,stomwidth,chamarea,nrow=2,ncol=2)
+
+```
+
+In Figure \@ref(fig:dimensionsboxplots), we see that, through the centuries, tomb dimensions did vary, and sometimes consistently, as with the trend toward narrower *dromoi*.
+The 9th-century tombs, though, which include all but two of the 12 discussed above, do not particularly stand out.
+It does seem their chambers were, on average, larger, as well as exhibiting greater variation than at any other time.
+But then again, even the largest chambers are not exactly cavernous, and the extra labour expended would have been limited.
+The *dromos* is a plausible area for conspicuous demonstrations of scale but, with their frequently poor preservation, little confidence is possible.
+It appears at any rate that they were gradually shrinking, albeit with some notable outliers.
+
+```{r tombgroupdimensions, fig.dim = c(7, 6), dpi=600, fig.align='center', fig.cap="goes around here"}
+
+ggplot(subset(tombs,century_built=="9th" | century_built=="10th"),aes(dromos_length,chamber_area,color=group,label=tomb))+geom_point()+
+  theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom")+
+  geom_label_repel(data=subset(tombs,group!="C"),aes(fill=group),color="white",show.legend = FALSE,family="serif")+
+            labs(x="Dromos Length (m)", y="Chamber Area (sq m)")+
+               scale_color_brewer(palette = "Set2",name="Group")+
+               scale_fill_brewer(palette = "Set2")+
+                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,10))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))+coord_fixed()
+
+```
+
+Looking more specifically at the tombs of Groups A and B, against the backdrop of the other 10th and 9th century tombs, there are perhaps hints of differentiation.
+All of them fall within the top 50% based on chamber area, and three have exceptionally long *dromoi*.
+But, later destruction notwithstanding, there appears nothing systematic about the variation visible here.
+If the size of one's tomb was intended to evoke or assert ideas of wealth or status, this appears to have occurred on a largely independent basis.
+
+```{r tombdimensions, fig.dim = c(7.5, 4), dpi=600, fig.align='center', fig.cap="goes around here"}
+
+tombareas<-ggplot(subset(tombs,tombs$est_burials>1),aes(chamber_area,est_burials,color=group))+geom_point()+
+  theme_bw()+theme(text=element_text(family="serif", size=10))+
+            labs(x="Chamber Area (sq m)", y="Estimated Total Burials")+
+               scale_color_brewer(palette = "Set2",name="Group")+
+                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+dromoslengths<-ggplot(subset(tombs,tombs$est_burials>1),aes(dromos_length,est_burials,color=group))+geom_point()+
+  theme_bw()+theme(text=element_text(family="serif", size=10))+
+            labs(x="Dromos Length (m)", y="Estimated Total Burials")+
+               scale_color_brewer(palette = "Set2",name="Group")+
+                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+
+ggarrange(tombareas,dromoslengths,ncol=2,common.legend = TRUE,legend="bottom")
+
+```
+
+What we can say is that no clear relationship existed between the dimensions of a tomb and its eventual buried population (see Figure \@ref(fig:tombdimensions)).
+Many of the very largest tombs are middling when it comes to burial numbers, and vice versa.
+So whatever the scale of the tombs connoted when built, this proved irrelevant to their later reuse.
+Several had niches cut into their *dromoi*, or the *stomion* extended, to fit in later burial urns (including T.
+292, 283 and 107; Cavanagh 1996: 658) and, sometimes, urns from within the chamber were relocated to make space for new interments (including T.
+75 and TFT; ibid.: 658; Brock 1957: 3–4).
+The impression is of competition for inclusion into the sepulchral group, further confirming that the tombs’ original builders little anticipated their eventual popularity.
+
+## 4.4 Imports, Interments, and Indices of Wealth
+
+Dimensions aside, we can investigate the contemporary use of these tombs through comparisons between the scale and composition of their respective assemblages through time.
+
+```{r tombscompthroughtime, fig.dim=c(7,4),dpi=600,fig.align='center',fig.cap="goes around here"}
+
+pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A",
+       ifelse(pots$tomb=="X"|pots$tomb=="G"|pots$tomb=="219"|pots$tomb=="Q"|pots$tomb=="283","B","C"))
+tombs9th10th<-subset(tombs,century_built=="9th" & total_pottery>50|century_built=="10th" & total_pottery>50)
+pots9th10th<-subset(pots,tomb_century=="9th"|tomb_century=="10th")
+
+tombs9th10th_ranked<-tombs9th10th[with(tombs9th10th, order(-tombs9th10th$total_pottery)), ]
+tombs9th10th_ranked<-as.vector(tombs9th10th_ranked$tomb)
+pots9th10th$tomb<-factor(pots9th10th$tomb,levels=tombs9th10th_ranked)
+pots9th10th<-subset(pots9th10th,century=="9th" & tomb!="NA"|century=="8th" & tomb!="NA"|century=="7th" & tomb!="NA")
+pots9th10th$century<-mapvalues(pots9th10th$century, from = c("9th","8th","7th"), to = c("9th Century", "8th Century","7th Century"))
+
+ggplot(pots9th10th,aes(tomb,fill=group))+geom_bar()+
+  theme_bw()+theme(legend.position="bottom",text=element_text(family="serif", size=10),
+               axis.text.x = element_text(angle = 90,hjust=1,vjust=0.4))+
+                 guides(fill=guide_legend(title="Group"))+labs(x="Tomb", y="Vessel Count")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0))+facet_grid(~century,scales="free_x",space="free")
+
+```
+
+And in Figure \@ref(fig:tombscompthroughtime), we see that the tombs of Group A, while not paultry in the 9th century, were by no means exceptional.
+They clearly kept pace with the other popular tombs of the time, but their major divergence only occurred in the subsequent two hundred years.
+Group B, meanwhile, experienced similar, if not greater, use in the 9th century, and then gradually declined.
+That is, these tombs seem to have followed a more conventional trajectory viz.
+Figure \@ref(fig:tombsinusebybuilddate).
+What might account for this disparity?
+And if the patterns implied in Figure \@ref(fig:tombscompthroughtime) - relative diversity, giving way to effective bipolarity in depositions - are genuine, how might we further investigate their manifestations across the excavated assemblage?
+
+## 4.5 A Tale of Two Periods
+
+```{r bigbox, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="goes around here"}
+
+aorist_tomb_values<-tombs[c(1,158:207)]
+aorist_tomb_values[aorist_tomb_values == 0] <- NA
+
+aorist_tomb_values<-data.frame(aorist_tomb_values)
+colnames(aorist_tomb_values)[2:51]<-dates
+
+aorist_melt<-melt(aorist_tomb_values)
+aorist_melt$variable<-as.numeric(as.character(aorist_melt$variable))
+
+aorist_melt$period<-ifelse(aorist_melt$variable<=1100 & aorist_melt$variable>=980,"SM",
+		ifelse(aorist_melt$variable<=970 & aorist_melt$variable>=930,"EPG",
+		ifelse(aorist_melt$variable<=920 & aorist_melt$variable>=880,"MPG",
+		ifelse(aorist_melt$variable<=870 & aorist_melt$variable>=850,"LPG",
+		ifelse(aorist_melt$variable<=840 & aorist_melt$variable>=820,"PGB",
+	  ifelse(aorist_melt$variable<=810 & aorist_melt$variable>=800,"EG",
+	  ifelse(aorist_melt$variable<=790 & aorist_melt$variable>=750,"MG",
+	  ifelse(aorist_melt$variable<=740 & aorist_melt$variable>=720,"LG",
+	  ifelse(aorist_melt$variable<=710 & aorist_melt$variable>=680,"EO",
+	  ifelse(aorist_melt$variable<=670 & aorist_melt$variable>=640,"LO",
+	  ifelse(aorist_melt$variable<=630 & aorist_melt$variable>=600,"A",""
+		)))))))))))
+
+decade_ranges<-seq(from = 600, to = 1100, by = 10)
+decade_ranges<-rev(decade_ranges)
+decade_ranges<-as.factor(decade_ranges)
+aorist_melt$variable<-as.factor(aorist_melt$variable)
+aorist_melt$variable<-factor(aorist_melt$variable, levels=(decade_ranges))
+aorist_melt$period<-factor(aorist_melt$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
+aorist_melt$variable<-as.numeric(as.character(aorist_melt$variable))
+
+ggplot(aorist_melt,aes(variable,value,fill=period))+geom_boxplot(aes(group = cut_width(variable, 10)),outlier.size=1,lwd=0.3)+theme_bw()+
+      theme(text=element_text(family="serif", size=10))+
+            labs(x="Date (Years BC)", y="Aoristic Sums per Tomb")+
+               scale_fill_brewer(name="Period",palette = "Set3")+
+                scale_x_reverse(expand=c(0,0),limits=c(1000,600))+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,20))
+
+```
+
+Figure \@ref(fig:bigbox) is a boxplot of the total aoristic sums across both cemeteries, colour-coded by ceramic phase.
+Looking at the PGB and EG, we see that they have the largest interquartile ranges (IQRs) of any period, and only a single outlier (beyond 1.5 times the IQR) each.
+The Orientalising period, on the other hand, exhibits much reduced IQRs, and a proliferation of outliers (namely, Group A) far and above the range of the other tombs.
+In other words, whilst both the PGB-EG and EO-LO periods witnessed an expanding volume of burial activity, in the earlier period this was distributed far more evenly across all tombs.
+One speaks of increasing plurality, the other of divergence.
+
+```{r gini, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="goes around here"}
+
+aorist_tomb_values<-tombs[c(1,158:207)]
+aorist_tomb_values[aorist_tomb_values == 0] <- NA
+
+aorist_tomb_values<-data.frame(aorist_tomb_values)
+colnames(aorist_tomb_values)[2:51]<-dates
+
+tombs_gini<-matrix(0, ncol = 0, nrow = 1)
+tombs_gini<-data.frame(tombs_gini)
+for(i in colnames(aorist_tomb_values[2:51])) {
+  tombs_gini[[paste0(i)]] <- ineq(aorist_tomb_values[[i]],type="Gini")
+}
+
+tombs_gini<-t(tombs_gini)
+tombs_gini<-data.frame(tombs_gini)
+
+colnames(tombs_gini)<-c("gini_coefficient")
+row.names(tombs_gini)<-dates
+tombs_gini<-data.frame(tombs_gini)
+setDT(tombs_gini, keep.rownames = "decade")
+tombs_gini$decade<-as.numeric(tombs_gini$decade)
+
+rects <- data.frame(xstart = c(1100,970,920,875,840,810,790,745,710,670,630), 
+                    xend = c(970,920,875,840,810,790,745,710,670,630,600), 
+                    col = c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
+rects$col<-factor(rects$col,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
+
+ggplot() + 
+  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf, fill = col), alpha = 0.4) +
+     geom_line(data = tombs_gini, aes(decade,gini_coefficient),size=0.5)+
+      theme_bw()+theme(text=element_text(family="serif", size=12))+
+              labs(x="Date (Years BC)", y="Gini Coefficient")+
+                 scale_fill_brewer(name="Period",palette = "Set3")+ coord_cartesian(
+                    xlim = c(1100, 600), ylim = c(0.5,0.8))+ scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
+                       scale_y_continuous(expand=c(0,0))
+```
+
+This characterisation is supported further by considering the Gini coefficient for the distribution of pottery between the tombs (see Figure \@ref(fig:gini)).
+Using aoristic sums as a probablistically-weighted estimate of the volume of material deposited in each tomb in each decade, the Gini coefficient can give an (albeit rough) estimate of how unequally distributed that material was.
+And, though values were always relatively high (i.e.
+unequal) within the cemeteries, the major spike in the Gini coefficient seen in the Orientalising period does contrast markedly with the plateau of the PGB-EG.
+
+This is not to suggest that the late 9th to early 8th century#ies was a time of egalitarian burial practices.
+The PGB and EG are consistently interpreted as periods of artistic development, increased overseas trade, and renewed vigour in the construction of tombs and the visibility of the burial rite [@Brock1957: 143-44; @Coldstream1994: 110-12].
+The PGB style reflects a uniquely Cretan blend of Attic and Near Eastern forms, while the appearance of BA 'antiques' in the tombs from this period onwards is felt to reflect rejuvinated, even competitive interest in the Minoan past.
+
+```{r imports, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="goes around here"}
+
+imports<-subset(pots,pots$import!="")
+greekimports<-subset(imports,imports$import_region=="Greece")
+NEimports<-subset(imports,imports$import_region=="Near East")
+imitations<-subset(pots,pots$imitation!="")
+
+import_aorist<-colSums(imports[27:76])
+import_aorist<-data.frame(import_aorist)
+colnames(import_aorist)<-c("aoristic_sum")
+row.names(import_aorist)<-dates
+setDT(import_aorist, keep.rownames = "decade")
+import_aorist$decade<-as.numeric(import_aorist$decade)
+
+greekimport_aorist<-colSums(greekimports[27:76])
+greekimport_aorist<-data.frame(greekimport_aorist)
+colnames(greekimport_aorist)<-c("aoristic_sum")
+row.names(greekimport_aorist)<-dates
+setDT(greekimport_aorist, keep.rownames = "decade")
+greekimport_aorist$decade<-as.numeric(greekimport_aorist$decade)
+
+NEimport_aorist<-colSums(NEimports[27:76])
+NEimport_aorist<-data.frame(NEimport_aorist)
+colnames(NEimport_aorist)<-c("aoristic_sum")
+row.names(NEimport_aorist)<-dates
+setDT(NEimport_aorist, keep.rownames = "decade")
+NEimport_aorist$decade<-as.numeric(NEimport_aorist$decade)
+
+imitation_aorist<-colSums(imitations[27:76])
+imitation_aorist<-data.frame(imitation_aorist)
+colnames(imitation_aorist)<-c("aoristic_sum")
+row.names(imitation_aorist)<-dates
+setDT(imitation_aorist, keep.rownames = "decade")
+imitation_aorist$decade<-as.numeric(imitation_aorist$decade)
+
+greekimport_aorist$origin<-"Greek Imports"
+NEimport_aorist$origin<-"Near Eastern Imports"
+imitation_aorist$origin<-"'Oriental' Imitations"
+all_import_imitation<-rbind(greekimport_aorist,NEimport_aorist,imitation_aorist)
+
+ggplot()+ geom_col(data=all_import_imitation,aes(decade,aoristic_sum, fill=origin),position="dodge")+
+            theme_bw()+theme(text=element_text(family="serif", size=10), legend.position="bottom")+
+                labs(x="Date (Years BC)", y="Aoristic Sum")+
+                  scale_fill_brewer(name="Origin",palette = "Set2")+
+                   scale_y_continuous(expand=c(0,0),limits=c(0,20))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))
+
+imports1<-nrow(subset(pots,pots$group=="B" & pots$import =="Y"))/nrow(subset(pots,pots$group=="B"))*100
+imports2<-nrow(subset(pots,pots$group=="C" & pots$import =="Y"))/nrow(subset(pots,pots$group=="C"))*100
+imports3<-nrow(subset(pots,pots$group=="A" & pots$import =="Y"))/nrow(subset(pots,pots$group=="A"))*100
+groupA7th<-subset(pots,group=="A" & century=="7th")
+groupA7thimports<-(sum(groupA7th$import=="Y")/nrow(groupA7th))*100
+
+```
+
+The rise in imported items is readily observable, with a succession of Greek and Near Eastern imports in the late 9th and early 8th centuries (Figure \@ref(fig:imports)).
+Likewise, the decline in Near Eastern imports, and concomitant rise in locally-made imitations thereof, is visible at the end of the 8th century.
+These trends being a very coarse proxy for its engagement with the wider Mediterranean, the late 9th and early 8th centuries would seem to encompass the most outward- (certainly,eastward-) looking phase of the EIA Knossian community [see @Whitley2004: 434].
+
+In the Group B tombs, with their more limited flourescence centred on the PGB-EG periods, imported items constitute `r round(imports1, digits=2)`% of the total excavated assemblage, compared to `r round(imports2, digits=2)`% in Group C and `r round(imports3, digits=2)`% in Group A.
+Indeed, among 7th century vessels, when interments were higher than ever among Group A, the figure is a mere `r round(groupA7thimports, digits=2)`%.
+The impression grows stronger that the context for the late proliferation of burials in the Group A tombs was altogether different from that underpinning the comparative, competitive diversity of interments a century earlier.
+
+```{r comps, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="goes around here"}
+
+pots$group<-ifelse(pots$tomb=="P"|pots$tomb=="292"|pots$tomb=="75"|pots$tomb=="107"|pots$tomb=="285"|pots$tomb=="II"|pots$tomb=="218","A",
+       ifelse(pots$tomb=="X"|pots$tomb=="G"|pots$tomb=="219"|pots$tomb=="Q"|pots$tomb=="283","B","C"))
+
+potcompKNC<-ggplot(subset(pots,century!="" & cem_condensed=="NC"),aes(century,fill=type))+geom_bar(position="fill")+
+  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title = element_blank(),
+               text=element_text(family="serif", size=12))+
+                 labs(title="KNC",x="", y="Proportion of Total Assemblage")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0.0))
+potcompFOR<-ggplot(subset(pots,century!="" & cem_condensed=="FOR"),aes(century,fill=type))+geom_bar(position="fill")+
+  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title=element_blank(),
+               text=element_text(family="serif", size=12))+
+                 labs(title="Fortetsa",x="Century BC", y="")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0.0))
+
+potcomptop7<-ggplot(subset(pots,century!="" & century!="11th"& group=="A"),aes(century,fill=type))+geom_bar(position="fill")+
+  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title=element_blank(),
+               text=element_text(family="serif", size=12))+
+                 labs(title="Group A",x="", y="")+
+                  scale_fill_brewer(palette = "Set2")+ 
+                    scale_y_continuous(expand=c(0,0.0))
+
+ggarrange(potcompKNC,potcompFOR,potcomptop7,ncol=3,common.legend =TRUE,legend="bottom")
+
+spearmanobjectspots<-cor(tombs$total_object,tombs$total_pottery,method="spearman")
+```
+
+This is reflected further in the changing composition of the mortuary assemblage through time (Figure \@ref(fig:comps)).
+Across both cemeteries, there appears to have been the steady crystallisation of a basic 'burial kit', comprising a pithos and lid, an oil vessel or two, a jug, and a drinking vessel.
+The proportion of grave goods comprising 'other' vessels - pyxides, askoi, trays, stands, dinoi, kernoi, kraters and more - declines to become a near insignificant part of the standardised repertoire of the 7th century.
+The tombs of Group A did not break this mould, and their assemblages seem largely as conventional as any other by this time.
+
+The main focus of this article has been on ceramics and not other artefacts of metal, stone, glass, and organic materials, due to their less well-developed chronologies, but even the raw distribution of such artefacts may make more sense in the light of the foregoing discussion.
+There is a relatively strong (Spearman's ρ=`r round(spearmanobjectspots,digits=2)`) correlation between the number of ceramic and non-ceramic finds across the tombs, and, certainly, many of the most seemingly precious objects were recovered from among the tombs with the most burials.
+
+```{r hists, fig.dim=c(6,4),dpi=600, fig.align='center',fig.cap="goes around here"}
+tombs$group<-revalue(tombs$group,c("A"="A","B"="B","C"="B"))
+total_object<-ggplot(subset(tombs,est_burials>1 & total_object>0),aes(total_object, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="serif", size=10))+
+                 labs(x="Total Non-Ceramic Finds", y="Number of Tombs")+
+                  scale_fill_manual(values = c("gray80","gray50"),labels=c(" Group A  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+total_weapon<-ggplot(subset(tombs,est_burials>1 & weapon>0),aes(weapon, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="serif", size=10))+
+                 labs(x="Total Weapons", y="Number of Tombs")+
+                 scale_fill_manual(values = c("gray80","gray50"),labels=c(" Group A  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+total_jewellery<-ggplot(subset(tombs,est_burials>1 & jewellery>0),aes(jewellery, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="serif", size=10))+
+                 labs(x="Total Jewellery", y="Number of Tombs")+
+                  scale_fill_manual(values = c("gray80","gray50"),labels=c(" Group A  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+
+objectsbyburial<-ggplot(subset(tombs,est_burials>1 & total_object>0),aes(total_object/est_burials, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="serif", size=10))+
+                 labs(x="Total Non-Ceramics/Burials", y="Number of Tombs")+
+                  scale_fill_manual(values = c("gray80","gray50"),labels=c(" Group A  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+weaponbyburial<-ggplot(subset(tombs,est_burials>1 & weapon>0),aes(weapon/est_burials, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="serif", size=10))+
+                 labs(x="Total Weapons/Burials", y="Number of Tombs")+
+                  scale_fill_manual(values = c("gray80","gray50"),labels=c(" Group A  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+
+jewellerybyburial<-ggplot(subset(tombs,est_burials>1 & jewellery>0),aes(jewellery/est_burials, fill=group))+geom_histogram()+
+  theme_bw()+theme(legend.title=element_blank(),
+               text=element_text(family="serif", size=10))+
+                 labs(x="Total Jewellery/Burials", y="Number of Tombs")+
+                  scale_fill_manual(values = c("gray80","gray50"),labels=c(" Group A  ", " Other Tombs"))+
+                    scale_y_continuous(expand=c(0,0))+scale_x_continuous(expand=c(0,0))
+
+ggarrange(total_jewellery,total_weapon,total_object,jewellerybyburial,weaponbyburial,objectsbyburial,ncol=3,nrow=2,common.legend = TRUE,legend="bottom")+
+            theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+```
+
+But this association, perhaps tempting one initially to speak of 'elite' or 'high-status' tombs, evaporates when the number of objects is divided by the number of burials represented.
+This fact has seldom, if ever, been appreciated in prior discussions of the cemeteries (Figure \@ref(fig:hists)).
+Whitley (2004: 437) has pointed out that the proportions of different decorative styles of Orientalising pithoi in Tomb P are replicated across Fortetsa as a whole, but it seems this situation may apply across most artefact classes.
+If so, the late surge in burial activity at both Fortetsa and the KNC was not linked in any straightforward way to simple principles of 'elite' or 'wealthy' burial; many of the most-used tombs prove exceptional only in their absolute quantities of artefacts, and not the accoutrements of each individual burial.
+This fact has seldom been appreciated in prior discussions of the cemeteries.
+
+# 5. Discussion
+
+To summarise, towards the end of 8th century BC, a select group of tombs, all built back in the 9th or 10th, became the foci of a major increase in burials which didn't register among the tombs built during that time.
+Little about the physical dimensions or early use of these tombs presaged their later popularity, while a similar, if smaller, increase in burial activity around a century earlier, in the PGB-EG periods, contrasted in the comparative diversity and broader distribution of its associated assemblages.
+
+It remains to consider how we might characterise the burying groups behind these various, intriguing trends.
+The traditional view is that such groups were immediate families (Brock 1957, 41; @Boardman1967: 63; @Coldstream1996: 291).
+The relatively short lifespan of many tombs has been taken as evidence for direct, lineal inheritance, with the tomb going out of use as families moved, individuals founded new tombs, or the line fissioned or ended [@Cavanagh1996: 659].
+The idea of family groups is supported by the, admittedly circumstantial, evidence of men, women and children being found within single tombs, and one instance of a (putatively) hereditary mandibular deformity in three skulls in the same tomb [@Musgrave1996].
+
+The model of lineal inheritance has been recently challenged by Kotsonas [-@Kotsonas2011: 131], however, who considers it ‘probably the greatest factoid in the archaeology of Crete of the given period’.
+He argues that the rise in interments during the 8th century is too great to be explained by population growth, as would be necessary for the possibility of direct lineal inheritance to be preserved.
+Rather, the trend must reflect reconfigurations of the burying groups themselves, which Kotsonas feels may have (also or instead) comprised age- and sex-grades.
+We should take this possibility seriously, for one need only look at Tomb P in the 7th century to see that we cannot possibly be dealing with just immediate families.
+
+Yet it is hard to be conclusive, as our current understanding of the social organisation of the early Cretan *poleis* remains patchy.
+Perlman and Gagarin's [-@Gagarin2016] recent study of the Cretan law codes reminds us that there is an often underappreciated degree of diversity in the offices, institutions, and social groups of the early city-states.
+In the traditional view, political authority was vested in the *kosmoi*, aristocratic magistrates selected from the leading, hereditary *startoi*, or ‘clans’, networks of extended lineages controlling the ancestral estates on which worked attendent serfs and/or slaves (Willetts 1982, 240).
+But despite elements of this account probably being accurate, others remain equivocal, such as the nature of the *pyla*, or 'tribe', and its relation to the *startos* (which may itself be a military grouping), the distinction between *astos* and *poliatas*, both probably formulations of the rights-bearing civic actor, and the ancestry and spread of the *agelaos*, an adolescent age-grade currently only attested at Eltynia [@Gagarin2016: 36; 78-80].
+Indeed, many terms which might illuminate social divisions and groupings within the Archaic Cretan populace are only attested at one or two sites, and even then sometimes apparently differ in form or function.
+
+My aim, however, is not to draw a simple parallel between the Knossian burial groups and, say, powerful *startoi*, a term not even attested outside of Gortyn in this period (ibid: 1-2).
+The fundamental conclusion of the disucussed contrasts between the late 9th and late 8th centuries, between the Group A and B tombs, is not that the burying groups changed, as such, it is that wider socio-economic and political contexts did, in ways which prompted reformulations of funerary practice, one of which being the intense, iterative burial activity evinced by the patrons of the Group A tombs.
+Such inferences, naturally, recall the conclusions of Morris [-@Morris1987: 171-211], who argued 30 years ago that the emergence of the *polis* across the Aegean led to a tension between elite preferences for ostentatious display, and a downplaying of social difference in death, as promoted by the ‘middling’ ideology of the city-state.
+In his well-known model of the *agathoi* and *kakoi*, Morris claimed that, at various times, the relevant spheres of kinship demonstrated in burial could widen and constrict, with more and less distantly related lineages drawn into association through changing social strategies, until the pre-eminence of kinship relations in the structuring of political power declined, along with its manifestiations in mortuary practice.
+
+At Knossos, some aspects of the noted expansion in the burying population seem explicable within a model like Morris's.
+The basic principle, of elites employing access to burial as a strategic act of association and social signalling, is almost certainly in evidence here.
+That the noted spate of burials in the late 8th and early 7th centuries was concentrated solely in older tombs strongly implies that issues of heritage, lineage, and tradition were at play.
+In Tomb 107, the stratigraphy of the niches cut into the *dromos* suggest depositions continued to be made even after the main chamber fell into disrepair [@Coldstream1996a: 150]; it is the symbolic and associative, rather than physical qualities of these tombs seem instrumental to their extensive reuse.
+
+Other features sit less easily with a Morrissian reading of the evidence, though.
+Firstly, the burials at issue are in many ways *not* ostentatious in comparison to the norm.
+We have seen that the 'richest' tombs in terms of finds are far less so when one accounts for the number of burials represented.
+Likewise, Morris claims that periods of expansion in the rite of burial will correspond to increased diversity in the associated assemblages.
+Yet, at Knossos, the range of grave goods diminished and crystallised during the late 8th and early 7th centuries.
+The PGB period better embodies such an expansion, with its greater variety in chamber dimensions and relatively broad distribution of (local and imported) grave goods, not to mention the innovative ceramic decoration and reuse of BA artefacts.
+In both the late 9th and late 8th centuries, then, certain burying groups found it propitious to broaden access to the tombs they patronised, but changing social contexts saw those expansions manifested in quite different ways.
+
+Elsewhere on Crete in the 7th century, we see a shift to unmarked, individual burials, and more austere grave goods, as well as an apparent reorientation of sanctuary deposits away from extra-mural to suburban spaces, and a rise in utilitarian, mass-produced votives [@Gagarin2016: 34-36].
+At Knossos, a small number of pithos burials in pits do appear in the KNC and Fortetsa volumes (Brock 1957: 98; Coldstream and Catling 1996: 162), and recent re-excavation of the Teke plot identified numerous small pit burials around three chamber tombs [@Evely2012], raising the possibility that others exist, but have gone unrecognised in prior excavations.
+At Azoria, major urban remodelling in the 7th century saw the earlier, agglutinative house forms of neighbouring Kavousi Vronda and Kastro give way to larger, more formally planned dwellings with little evidence of remodelling, while larger, apparently civic, buildings were constructed to accommodate major public gatherings [@Haggis2011; @Haggis2011b; @Haggis2014].
+
+Together, these varied threads suggest that, with the emergence of the archaic *poleis* on Crete, there was a concomitant restructuring of wider social groups, institutions, and behaviours.
+It seems not unreasonable to infer an emergent emphasis on the citizen, the state, and ideologies of relative egality, within which communal gatherings, intra-mural ritual, and individual graves replaced ostentatious, kin-centred acts of gathering, dedication, and burial.
+And in the Knossian cemeteries, we have seen a somewhat diverse set of burial practices give way to a more prolific, yet more restricted, suite of funerary rites, focussed on a small number of long-established tombs, in the context of diminishing collective tomb construction, replacement of imports with locally-made imitations, and an imminent cessation of such depositions.
+
+Whether the increasingly homogenous burials reflected some standard set of accoutrements intentionally shared by the hypothetical *agathoi*, or else nascent, egalitarian strictures over elaborate burial remains difficult to judge.
+Yet we need not adopt wholesale a model like Morris's "middling" ideology, nor link the diminishing range of burial goods to Archaic Crete's purported austerity.
+As a more nuanced picture emerges of the island's relationship with the wider Mediterranean, subtler characterisations are possible.
+It is telling that the peak for imported items came in the 9th century; the Orientalising as a time period feels increasingly misapplied to Crete [@Whitley2004: 434], when, at Knossos at least, it witnessed far lower rates of 'oriental' imports, and only limited adoption of the figural motifs which characterise its appearance elsewhere.
+
+The decline in imports was, however, counterpointed by a rise in imitations of Near Eastern vessels [@Antoniadis2012: 170], according with our growing awareness of the selective nature of Archaic and Classical Crete's external relations.
+Erickson [@Erickson2010], similarly, has argued that *symposia* were practised in Crete, something often disputed, but that their practitioners merely favoured simpler vessel forms than the figurally-decorated wares popular on the mainland.
+Local, selective adoption of, and responses to, wider social and material trends is perhaps becoming the hallmark of Archaic Crete, and finds further endorsement in the EIA cemeteries at Knossos.
+What seems certain is that this was a period of transition wherein, whatever their exact identity, certain actors invested heavily in a final flurry of burials in chamber tombs, with an emphasis on a traditional, established repertoire of items, rites, and venues for one's passage to the afterlife, before changing social forces effected their abandonment.
+
+# 6. Conclusion
+
+It is important that the present study was a comparative investigation of both the Fortetsa and Knossos North cemeteries.
+That both underwent comparable developments during the EIA suggests, not only that they catered to similar sub-sections of the Knossian population, but that some of the socio-economic and political changes inferred were significant and community-wide.
+Further work should seek to extend these comparisons, both into trends unexplored in the present study, and out to the wider funerary landscape of Knossos.
+I hope to do just this but, for the time being, the datasets I am making available alongside this article include the other published EIA tombs in the area, so that others may also explore their possible significance.
+
+There were undeniable limitations to the present study.
+For one, non-ceramic artefacts were underrepresented due to difficulties with dating, and future work could seek to address this.
+Issues of gender were sadly little considered, though this was largely due to limitations in the data.
+There have also been a number of tombs discovered in the last 20 years that fill in the spaces between the Teke plot and the Medical Faculty Site, and ideally these unpublished tombs should be included in further analyses as they, along with the Teke tholos and associated tombs, are probably part of a single contiguous cemetery [@Blackman1999: 113; @Catling1983: 51; @French1991: 68-69].
+Finally, a preliminary study was recently undertaken by the author of select boxes marked as 'unpublished' from the KNC deposits kept at the British School at Athens, at Knossos.
+Despite little mention in the publication, this material was not inconsiderable and so, regretably, future work will be necessary to determine whether it threatens the integrity of these analyses.
+Suffice to say that, at present, it is another pinch of salt with which to take the present findings.
+
+These limitations notwithstanding, the foregoing analysis demonstrates the value of applying quantitative approaches to the archaeological record of the period.
+Though the construction of databases can be time consuming, once made they offer the opportunity to subdivide, compare, and plot the data in numerous ways, facilitating investigation of the material from multiple angles.
+By making datasets available for use by others, as with the data and code for the present study, one’s conclusions can also be independently tested, and alternative interpretations can be voiced.
+The databases produced for the above analysis contain a large of amount of unutilised data that could facilitate a range of future investigations, and all are welcome to it.
+With more such analyses, we could begin to a construct a fuller account of the regional diversity in mortuary, and wider social, practices that characterised the communities of EIA and archaic Crete.
+
+##### pagebreak
+# Acknowledgements
+
+I would like to thank Borja Legarra Herrero and Corinna Riva for their help and advice from the earliest stages of the present research, as well as the BSA at Knossos for permitting me to examine some of the material from the Medical Faculty excavations.
+Many thanks to Anna Bloxam for pointing me towards aoristic analysis.
+I am indebted most of all to Professor Todd Whitelaw, for his unstinting guidance, feedback and facilitation of my ongoing research.
+
+# Funding
+
+This research did not receive any specific grant from funding agencies in the public, commercial, or not-for-profit sectors.
+
+##### pagebreak
+# References 
+<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
+<div id="refs"></div>
+
+##### pagebreak
+# Captions
+
+*Figure 1 Map of the Knossos area, showing the locations of the various EIA tomb clusters and cemeteries, with EIA pottery distributions from the Knossos Urban Landscape Project marked in orange, giving an impression of the contemporary urban extent (underlying image credit: T.
+Whitelaw).*
+
+*Figure 2 Relationship between known interments, counts of pithoi, and counts of pithoi, amphorae and kraters across both cemeteries.
+The straight line represents the mean estimate based on these three values.*
+
+*Figure 3 Counts of pithoi across both cemeteries through time. Vessel dates have here been based on the midpoint (or mean) of the absolute date range accompanying their respective ceramic periods.*
+
+*Figure 4 Aoristic sums of all vessels across both cemeteries. This represents a probabilistically weighted visualisation of depositional activity through time.*
+
+*Figure 5 Aoristic sums of pithoi (left), as a proxy for cremation burials, and of all vessels (right). The sums are colour coded to show the values of the two cemeteries, and the total values*
+
+*Figure 6 Tomb construction through time at both cemeteries.
+On the left, counts are based on the ceramic phase considered most likely to correspond to the tomb's first use (including simply 'PG', 'G' and 'O' where no greater specificity was possible).
+On the right, aoristic sums are presented, spreading the uncertainty associated with those general assignations.
+In both cases, tombs lacking any securely dateable pottery are excluded.*
+*Abbreviations*
+*SM - Subminoan; (E/M/L)PG - (Early/Middle/Late) Protogeometric; PGB - Protogeometric B; (E/M/L)G - (Early/Middle/Late) Geometric; (E/L)O - (Early/Late) Orientalising*
+
+*Figure 7 Estimates of the number of tombs in use during each of the ceramic phases attested in both cemetaries, based on counts of tombs yielding vessels of each period.*
+
+*Figure 8 Estimates of the number of tombs in use through time across both cemeteries.
+The grey background represents the total number of tombs in use, while the smaller coloured bars correspond to tombs built in each of the five centuries of EIA activity at the cemeteries.*
+
+*Figure 9 Counts of vessels from all tombs in both cemeteries, arranged, on the left, by the century to which they are assigned based on their ceramic period and, on the right, by the centuries in which the tombs they come from were built.*
+
+*Figure 10 Aoristic sums for the 12 tombs with the largest ceramic assemblages, arranged in decreasing order.
+Note, for the sake of better illustrating the relative temporal patterns, rather than absolute values, the scales on each plot are distinct.*
+
+*Figure 11 Aoristic sums for the Group A and Group B tombs, as defined above.*
+
+*Figure 11 Boxplots comparing various tomb dimensions, divided, in each case, by the century of tomb construction.*
+
+*Figure 12 Scatterplot comparing the dimensions of* dromos *length and chamber area.
+Groups A and B, defined earlier, are illustrated and labelled in green and orange respectively, while all remaining tombs ('Group C') are in blue.*
+
+*Figure 13 Scatterplots comparing tomb dimensions (chamber area and* dromos *length) to the estimated total number of burials by tomb. The points area colour-coded as per Figure 12.*
+
+*Figure 14 Counts of vessels dating to the 9th, 8th and 7th centuries respectively, found in tombs built in the 10th or 9th century which yielded upwards of 50 vessels when excavated.
+The tombs are colour-coded as per Figure 12.*
+
+*Figure 15 Boxplot showing the aoristic sums of all vessels across both cemeteries through time, colour-coded by ceramic period.
+Boxes represent the interquartile range, horizontal lines the median value, whiskers values within 1.5 times the interquartile range, and points outliers beyond this.
+Note, the Y axis has been limited to 20 to aid readability, meaning the outliers Tomb G in the EG period, and Tomb P in the late EO-LO period are excluded, with values ranging from 29-34, and 36-44 respectively.*
+
+*Figure 16 Gini coefficient values through time based on the aoristic sums for all vessels across both cemeteries. Background colours correspond to ceramic periods.*
+
+*Figure 17 Aoristic sums of imported and 'oriental' imitation vessels through time across both cemeteries, colour-coded by regional origin in the case of imports.*
+
+*Figure 18 The composition of the total pottery assemblage for each century across both cemeteries, and among the 7 notable tombs (Group A) drawn out earlier in the analysis.*
+
+*Figure 19 Histograms corresponding, first, to counts of jewellery (beads, necklaces, rings, diadems etc.), metal objects, and imported items by tomb across both cemeteries and, secondly, to those counts divided by the estimated number of burials in each tomb.*
+
+##### pagebreak
+# Figures
+
+##### pagebreak
+# Tables
+
 #####pagebreak

### 2020-10-12 analysis/paper/paper2.Rmd
Article edits

@@
 author:
-  - Dominic Pollard, Institute of Archaeology, University College London
+  - Dominic Pollard
+  - Institute of Archaeology, University College London, 31-34 Gordon Square, Bloomsbury, London WC1H 0PY, United Kingdom
   - dominic.pollard.16@ucl.ac.uk
@@
 bibliography: references.bib
-csl: "../templates/world-archaeology.csl" # Insert path for the bib-style
+csl: "../templates/journal-of-anthropological-archaeology.csl" # Insert path for the bib-style
 

### 2020-10-12 analysis/templates/template.Rmd
Beginning writing
Article edits

@@
 output: word_document
+highlights: "My highlights \n"
 abstract: "My abstract \n"
@@
   word_document:
-    reference_docx: templates/template1.docx
+    reference_docx: templates/articletemplate.docx
 ```

### 2020-10-13 analysis/paper/Family_Fortunes_Article.Rmd
Figure edits

@@
 The colour scheme in this paper is chosen from those that are suitable for colour-blind readers.
-The font Garmond is installed and stored.
 library(ggpubr)
@@
 library(ggforce)
-library(png)
+library(jpeg)
 library(ggrepel)
@@
 
-```{r map, out.width="50%", fig.cap="goes around here"}
+```{r map, fig.align= "centre", fig.cap="goes around here"}
+
 knitr::include_graphics("../figures/Knossos Tombs Map2.jpg")
@@
 
-```{r burialestimate, fig.dim = c(7, 7), dpi=600, fig.align='center', fig.cap="goes around here"}
-
-#The tombs database is imported and, for the purposes of this article, the entire database is subsetted to include only those tombs from the Knossos North and Fortetsa cemeteries (excluding the Tekke Tombs).
-
-tombs<-read.csv("../data/raw_data/KnossosTombs.csv",header=TRUE)
-tombs<-subset(tombs,tombs$cem_condensed=="KNC"|tombs$cem_condensed=="FOR")
-
-#Figure 1 is plotted, with axis limits set (excluding tomb P, resulting in the warning message delivered) and the style of the plot adjusted (including font and text size, the appearance of the gridlines, selecting a colour palette etc.)
-
-p<-ggplot(tombs)+geom_point(aes(est_burials,all_urns,color="All urns"))+
-  geom_point(aes(est_burials,total_burials,color="Known interments"))+
-    geom_point(aes(est_burials,pithoi,colour="Pithoi"))+
-      geom_line(aes(est_burials,est_burials),alpha=0.4)+
-       scale_x_continuous(expand=c(0,0),limits=c(0,55))+
-        scale_y_continuous(expand=c(0,0),limits=c(0,55))+theme_bw()+
-         theme(legend.title=element_blank(),
-          text=element_text(family="serif", size=12),legend.position="bottom")+
-              labs(x="Mean Burial Estimate", y="Interment/Vessel Count")+
-                 scale_color_brewer(palette = "Set2")
-
-#The coordinates of the plot are adjusted to render the axes equal, and the corresponding gridlines as squares, and then the figure is plotted.
-
-p<-p+coord_fixed()
-plot(p)
-
-```
-
 For most diachronic analyses, however, pithoi alone were taken as the best proxy for burial numbers.
@@
 
-```{r allpithoi,fig.dim = c(8, 4), dpi=600, fig.align='center', fig.cap="goes around here"}
+```{r allpithoi,fig.dim = c(8, 3), dpi=600, fig.align='center', fig.cap="goes around here"}
+
+#The tombs database is imported and, for the purposes of this article, the entire database is subsetted to include only those tombs from the Knossos North and Fortetsa cemeteries (excluding the Tekke Tombs).
+
+tombs<-read.csv("../data/raw_data/KnossosTombs.csv",header=TRUE)
+tombs<-subset(tombs,tombs$cem_condensed=="KNC"|tombs$cem_condensed=="FOR")
 
@@
   theme_bw()+theme(legend.title=element_blank(),
-      text=element_text(family="serif", size=12))+
+      text=element_text(family="serif", size=11))+
           labs(x="Date (Years BC)", y="Vessel Count")+
                 theme(legend.position="none")+
-                  scale_y_continuous(expand=c(0,0),limits=c(0,600))+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,610))+
                     scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
@@
   geom_col(fill="gray40")+theme_bw()+theme(legend.title=element_blank(),
-      text=element_text(family="serif", size=12))+
+      text=element_text(family="serif", size=11))+
           labs(x="Date (Years BC)", y="Aoristic Sum")+
@@
 
-```{r aoristbycem, dpi=600, fig.align="center", fig.cap="goesaround here.", fig.dim=c(8,4)}
+```{r aoristbycem, dpi=600, fig.align="center", fig.cap="goesaround here.", fig.dim=c(8,3.5)}
 KNC<-subset(tombs,tombs$cem_condensed=="KNC")
@@
         theme_bw()+theme(legend.title=element_blank(),
-          text=element_text(family="serif", size=12))+
+          text=element_text(family="serif", size=11))+
             labs(x="Date (Years BC)", y="Aoristic Sum (Pithoi)")+
@@
         theme_bw()+theme(legend.title=element_blank(),
-          text=element_text(family="serif", size=12))+
+          text=element_text(family="serif", size=11))+
             labs(x="Date (Years BC)", y="Aoristic Sum (All Vessels)")+
@@
 
-```{r tombsinuse, fig.dim=c(7,7), dpi=600, fig.cap="goes around here"}
+```{r tombsinuse, fig.dim=c(3,3),out.width="25%", dpi=600, fig.cap="goes around here"}
 
@@
        theme_bw()+theme(legend.title=element_blank(),
-          text=element_text(family="serif", size=12),legend.position = "bottom")+
+          text=element_text(family="serif", size=10),legend.position = "bottom")+
             labs(x="Period", y="Tombs in Use")+
@@
 
-```{r potsbydateandtombdate,  fig.dim=c(7,14), dpi=600, fig.cap="goes around here"}
+```{r potsbydateandtombdate,  fig.dim=c(3,7),out.width="0.5%", dpi=600, fig.cap="goes around here"}
 
@@
   theme_bw()+theme(legend.position="none",
-               text=element_text(family="serif", size=12))+
+               text=element_text(family="serif", size=10),axis.text.y=element_text(angle=90, vjust=0, hjust=0.5))+
                  labs(x="Century BC (Date of Vessel)", y="Vessel Count")+
@@
   theme_bw()+theme(legend.position="none",
-               text=element_text(family="serif", size=12))+
+               text=element_text(family="serif", size=10),axis.text.y=element_text(angle=90, vjust=0, hjust=0.5))+
                  labs(x="Century BC (Date of Tomb)", y="Vessel Count")+
@@
 
-```{r dimensionsboxplots, fig.dim=c(7,6),dpi=600, fig.align='center',fig.cap="goes around here"}
+```{r dimensionsboxplots, fig.dim=c(7,7),dpi=600, fig.align='center',fig.cap="goes around here"}
 
@@
 
-```{r tombgroupdimensions, fig.dim = c(7, 6), dpi=600, fig.align='center', fig.cap="goes around here"}
-
-ggplot(subset(tombs,century_built=="9th" | century_built=="10th"),aes(dromos_length,chamber_area,color=group,label=tomb))+geom_point()+
-  theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom")+
+```{r tombgroupdimensions, fig.dim = c(3.5, 3.5),out.width="10%", dpi=600, fig.align='center', fig.cap="goes around here"}
+
+ggplot(subset(tombs,century_built=="9th" | century_built=="10th"),aes(dromos_length,chamber_area,color=group,label=tomb))+geom_point(size=1)+
+  theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom", legend.margin=margin(0,0,0,0),legend.box.margin=margin(-10,-10,-10,-10))+
   geom_label_repel(data=subset(tombs,group!="C"),aes(fill=group),color="white",show.legend = FALSE,family="serif")+
@@
                scale_fill_brewer(palette = "Set2")+
-                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
+                scale_x_continuous(expand=c(0,0),limits=c(0,8))+
                   scale_y_continuous(expand=c(0,0),limits=c(0,10))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))+coord_fixed()
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
 

### 2020-10-13 analysis/paper/Family_Fortunes_Article.Rmd
Repo name changes

@@
 
-```{r map, fig.align= "centre", fig.cap="goes around here"}
+```{r map, fig.align= "centre", fig.cap="Map of the Knossos area, showing the locations of the various EIA tomb clusters and cemeteries, with EIA pottery distributions from the Knossos Urban Landscape Project marked in orange, giving an impression of the contemporary urban extent (underlying image credit: T.
+Whitelaw)."}
 
@@
 
-```{r allpithoi,fig.dim = c(8, 3), dpi=600, fig.align='center', fig.cap="goes around here"}
+```{r allpithoi,fig.dim = c(8, 3), dpi=600, fig.align='center', fig.cap="(Left) Counts of pithoi across both cemeteries through time.
+Vessel dates have here been based on the midpoint (or mean) of the absolute date range accompanying their respective ceramic periods.
+(Right) Aoristic sums of all vessels across both cemeteries.
+This represents a probabilistically weighted visualisation of depositional activity through time."}
 
@@
 
-```{r aoristbycem, dpi=600, fig.align="center", fig.cap="goesaround here.", fig.dim=c(8,3.5)}
+```{r aoristbycem, dpi=600, fig.align="center", fig.cap="Aoristic sums of pithoi (left), as a proxy for cremation burials, and of all vessels (right).
+The sums are colour coded to show the values of the two cemeteries, and the total values.", fig.dim=c(8,3.5)}
 KNC<-subset(tombs,tombs$cem_condensed=="KNC")
@@
 
-```{r tombconstruction, fig.dim=c(7.5,4), dpi=600, fig.cap="goes around here"}
+```{r tombconstruction, fig.dim=c(7.5,4), dpi=600, fig.cap="Tomb construction through time at both cemeteries.
+On the left, counts are based on the ceramic phase considered most likely to correspond to the tomb's first use (including simply 'PG', 'G' and 'O' where no greater specificity was possible).
+On the right, aoristic sums are presented, spreading the uncertainty associated with those general assignations.
+In both cases, tombs lacking any securely dateable pottery are excluded.
+Abbreviations: SM - Subminoan; (E/M/L)PG - (Early/Middle/Late) Protogeometric; PGB - Protogeometric B; (E/M/L)G - (Early/Middle/Late) Geometric; (E/L)O - (Early/Late) Orientalising"}
 
@@
 
-```{r tombsinuse, fig.dim=c(3,3),out.width="25%", dpi=600, fig.cap="goes around here"}
+```{r tombsinuse, fig.dim=c(3,3),out.width="25%", dpi=600, fig.cap="Estimates of the number of tombs in use during each of the ceramic phases attested in both cemetaries, based on counts of tombs yielding vessels of each period."}
 
@@
        theme_bw()+theme(legend.title=element_blank(),
-          text=element_text(family="serif", size=10),legend.position = "bottom")+
+          text=element_text(family="serif", size=10),legend.position = "bottom", legend.margin=margin(0,0,0,0),legend.box.margin=margin(-10,-10,10,-10))+
             labs(x="Period", y="Tombs in Use")+
@@
 
-```{r tombsinusebybuilddate, fig.dim=c(7,4), dpi=600, fig.cap="goes around here"}
+```{r tombsinusebybuilddate, fig.dim=c(7,4), dpi=600, fig.cap="Estimates of the number of tombs in use through time across both cemeteries.
+The grey background represents the total number of tombs in use, while the smaller coloured bars correspond to tombs built in each of the five centuries of EIA activity at the cemeteries."}
 tombs_11th<-subset(tombs,tombs$century_built=="11th")
@@
 
-```{r potsbydateandtombdate,  fig.dim=c(3,7),out.width="0.5%", dpi=600, fig.cap="goes around here"}
+```{r potsbydateandtombdate,  fig.dim=c(3,7),out.width="0.5%", dpi=600, fig.cap="Counts of vessels from all tombs in both cemeteries, arranged, above, by the century to which they are assigned based on their ceramic period and, below, by the centuries in which the tombs they come from were built."}
 
@@
 
-```{r toptombs, fig.dim = c(7, 5), dpi=600, fig.align='center',fig.cap="goes around here"}
+```{r toptombs, fig.dim = c(7, 5), dpi=600, fig.align='center',fig.cap="Aoristic sums for the 12 tombs with the largest ceramic assemblages, arranged in decreasing order.
+Note, for the sake of better illustrating the relative temporal patterns, rather than absolute values, the scales on each plot are distinct."}
 
@@
       text=element_text(family="serif", size=12),
-      axis.text.x=element_text(angle=45, vjust=1.2, hjust=1))+
+      axis.text.x=element_text(angle=45, vjust=1.2, hjust=1), legend.margin=margin(0,0,0,0),legend.box.margin=margin(-10,-10,10,-10))+
           labs(x="Date (Years BC)", y="Aoristic Sums")+
@@
 
-```{r groupsab, fig.dim=c(8,4),dpi=600, fig.align='center',fig.cap="goes around here"}
+```{r groupsab, fig.dim=c(8,4),dpi=600, fig.align='center',fig.cap="Aoristic sums for the Group A and Group B tombs, as defined above."}
 
@@
 
-```{r dimensionsboxplots, fig.dim=c(7,7),dpi=600, fig.align='center',fig.cap="goes around here"}
+```{r dimensionsboxplots, fig.dim=c(7,7),dpi=600, fig.align='center',fig.cap="Boxplots comparing various tomb dimensions, divided, in each case, by the century of tomb construction."}
 
@@
 
-```{r tombgroupdimensions, fig.dim = c(3.5, 3.5),out.width="10%", dpi=600, fig.align='center', fig.cap="goes around here"}
+```{r tombgroupdimensions, fig.dim = c(3.5, 3.5),out.width="10%", dpi=600, fig.align='center', fig.cap="Scatterplot comparing the dimensions of* dromos *length and chamber area.
+Groups A and B, defined earlier, are illustrated and labelled in green and orange respectively, while all remaining tombs ('Group C') are in blue."}
 
 ggplot(subset(tombs,century_built=="9th" | century_built=="10th"),aes(dromos_length,chamber_area,color=group,label=tomb))+geom_point(size=1)+
-  theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom", legend.margin=margin(0,0,0,0),legend.box.margin=margin(-10,-10,-10,-10))+
-  geom_label_repel(data=subset(tombs,group!="C"),aes(fill=group),color="white",show.legend = FALSE,family="serif")+
+  theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom", legend.margin=margin(0,0,0,0),legend.box.margin=margin(-10,-10,10,-10))+
+  geom_label_repel(data=subset(tombs,group!="C"),aes(fill=group),color="white",show.legend = FALSE,family="serif", size=3)+
             labs(x="Dromos Length (m)", y="Chamber Area (sq m)")+
@@
 
-```{r tombdimensions, fig.dim = c(7.5, 4), dpi=600, fig.align='center', fig.cap="goes around here"}
+```{r tombdimensions, fig.dim = c(7.5, 4), dpi=600, fig.align='center', fig.cap="Scatterplots comparing tomb dimensions (chamber area and* dromos *length) to the estimated total number of burials by tomb.
+The points area colour-coded as per Figure 11."}
 
@@
 
-```{r tombscompthroughtime, fig.dim=c(7,4),dpi=600,fig.align='center',fig.cap="goes around here"}
+```{r tombscompthroughtime, fig.dim=c(7,4),dpi=600,fig.align='center',fig.cap="Counts of vessels dating to the 9th, 8th and 7th centuries respectively, found in tombs built in the 10th or 9th century which yielded upwards of 50 vessels when excavated."}
 
@@
   theme_bw()+theme(legend.position="bottom",text=element_text(family="serif", size=10),
-               axis.text.x = element_text(angle = 90,hjust=1,vjust=0.4))+
+               axis.text.x = element_text(angle = 90,hjust=1,vjust=0.4), legend.margin=margin(0,0,0,0),legend.box.margin=margin(-10,-10,10,-10))+
                  guides(fill=guide_legend(title="Group"))+labs(x="Tomb", y="Vessel Count")+
@@
 
-```{r bigbox, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="goes around here"}
+```{r bigbox, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="Boxplot showing the aoristic sums of all vessels across both cemeteries through time, colour-coded by ceramic period.
+Boxes represent the interquartile range, horizontal lines the median value, whiskers values within 1.5 times the interquartile range, and points outliers beyond this.
+Note, the Y axis has been limited to 20 to aid readability, meaning the outliers Tomb G in the EG period, and Tomb P in the late EO-LO period are excluded, with values ranging from 29-34, and 36-44 respectively."}
 
@@
 
-ggplot(aorist_melt,aes(variable,value,fill=period))+geom_boxplot(aes(group = cut_width(variable, 10)),outlier.size=1,lwd=0.3)+theme_bw()+
+ggplot(aorist_melt,aes(variable,value,fill=period))+geom_boxplot(aes(group = cut_width(variable, 10)),outlier.size=0.9,lwd=0.3)+theme_bw()+
       theme(text=element_text(family="serif", size=10))+
@@
 
-```{r gini, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="goes around here"}
+```{r gini, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="Gini coefficient values through time based on the aoristic sums for all vessels across both cemeteries.
+Background colours correspond to ceramic periods."}
 
@@
 
-```{r imports, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="goes around here"}
+```{r imports, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="Aoristic sums of imported and 'oriental' imitation vessels through time across both cemeteries, colour-coded by regional origin in the case of imports."}
 
@@
 ggplot()+ geom_col(data=all_import_imitation,aes(decade,aoristic_sum, fill=origin),position="dodge")+
-            theme_bw()+theme(text=element_text(family="serif", size=10), legend.position="bottom")+
+            theme_bw()+theme(text=element_text(family="serif", size=10), legend.position="bottom", legend.margin=margin(0,0,0,0),legend.box.margin=margin(2,-10,10,-10))+
                 labs(x="Date (Years BC)", y="Aoristic Sum")+
@@
 
-```{r comps, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="goes around here"}
+```{r comps, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="The composition of the total pottery assemblage for each century across both cemeteries, and among the 7 notable tombs (Group A) drawn out earlier in the analysis."}
 
@@
   theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title = element_blank(),
-               text=element_text(family="serif", size=12))+
+               text=element_text(family="serif", size=11))+
                  labs(title="KNC",x="", y="Proportion of Total Assemblage")+
@@
   theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title=element_blank(),
-               text=element_text(family="serif", size=12))+
+               text=element_text(family="serif", size=11))+
                  labs(title="Fortetsa",x="Century BC", y="")+
@@
   theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title=element_blank(),
-               text=element_text(family="serif", size=12))+
+               text=element_text(family="serif", size=11))+
                  labs(title="Group A",x="", y="")+
@@
 
-```{r hists, fig.dim=c(6,4),dpi=600, fig.align='center',fig.cap="goes around here"}
+```{r hists, fig.dim=c(6,4),dpi=600, fig.align='center',fig.cap="Histograms corresponding, first, to counts of jewellery (beads, necklaces, rings, diadems etc.), metal objects, and imported items by tomb across both cemeteries and, secondly, to those counts divided by the estimated number of burials in each tomb."}
+
 tombs$group<-revalue(tombs$group,c("A"="A","B"="B","C"="B"))
@@
 It remains to consider how we might characterise the burying groups behind these various, intriguing trends.
-The traditional view is that such groups were immediate families (Brock 1957, 41; @Boardman1967: 63; @Coldstream1996: 291).
+The traditional view is that such groups were immediate families [@Brock1957: 41; @Boardman1967: 63; @Coldstream1996: 291].
 The relatively short lifespan of many tombs has been taken as evidence for direct, lineal inheritance, with the tomb going out of use as families moved, individuals founded new tombs, or the line fissioned or ended [@Cavanagh1996: 659].
@@
 ##### pagebreak
-# Captions
-
-*Figure 1 Map of the Knossos area, showing the locations of the various EIA tomb clusters and cemeteries, with EIA pottery distributions from the Knossos Urban Landscape Project marked in orange, giving an impression of the contemporary urban extent (underlying image credit: T.
-Whitelaw).*
-
-*Figure 2 Relationship between known interments, counts of pithoi, and counts of pithoi, amphorae and kraters across both cemeteries.
-The straight line represents the mean estimate based on these three values.*
-
-*Figure 3 Counts of pithoi across both cemeteries through time. Vessel dates have here been based on the midpoint (or mean) of the absolute date range accompanying their respective ceramic periods.*
-
-*Figure 4 Aoristic sums of all vessels across both cemeteries. This represents a probabilistically weighted visualisation of depositional activity through time.*
-
-*Figure 5 Aoristic sums of pithoi (left), as a proxy for cremation burials, and of all vessels (right). The sums are colour coded to show the values of the two cemeteries, and the total values*
-
-*Figure 6 Tomb construction through time at both cemeteries.
-On the left, counts are based on the ceramic phase considered most likely to correspond to the tomb's first use (including simply 'PG', 'G' and 'O' where no greater specificity was possible).
-On the right, aoristic sums are presented, spreading the uncertainty associated with those general assignations.
-In both cases, tombs lacking any securely dateable pottery are excluded.*
-*Abbreviations*
-*SM - Subminoan; (E/M/L)PG - (Early/Middle/Late) Protogeometric; PGB - Protogeometric B; (E/M/L)G - (Early/Middle/Late) Geometric; (E/L)O - (Early/Late) Orientalising*
-
-*Figure 7 Estimates of the number of tombs in use during each of the ceramic phases attested in both cemetaries, based on counts of tombs yielding vessels of each period.*
-
-*Figure 8 Estimates of the number of tombs in use through time across both cemeteries.
-The grey background represents the total number of tombs in use, while the smaller coloured bars correspond to tombs built in each of the five centuries of EIA activity at the cemeteries.*
-
-*Figure 9 Counts of vessels from all tombs in both cemeteries, arranged, on the left, by the century to which they are assigned based on their ceramic period and, on the right, by the centuries in which the tombs they come from were built.*
-
-*Figure 10 Aoristic sums for the 12 tombs with the largest ceramic assemblages, arranged in decreasing order.
-Note, for the sake of better illustrating the relative temporal patterns, rather than absolute values, the scales on each plot are distinct.*
-
-*Figure 11 Aoristic sums for the Group A and Group B tombs, as defined above.*
-
-*Figure 11 Boxplots comparing various tomb dimensions, divided, in each case, by the century of tomb construction.*
-
-*Figure 12 Scatterplot comparing the dimensions of* dromos *length and chamber area.
-Groups A and B, defined earlier, are illustrated and labelled in green and orange respectively, while all remaining tombs ('Group C') are in blue.*
-
-*Figure 13 Scatterplots comparing tomb dimensions (chamber area and* dromos *length) to the estimated total number of burials by tomb. The points area colour-coded as per Figure 12.*
-
-*Figure 14 Counts of vessels dating to the 9th, 8th and 7th centuries respectively, found in tombs built in the 10th or 9th century which yielded upwards of 50 vessels when excavated.
-The tombs are colour-coded as per Figure 12.*
-
-*Figure 15 Boxplot showing the aoristic sums of all vessels across both cemeteries through time, colour-coded by ceramic period.
-Boxes represent the interquartile range, horizontal lines the median value, whiskers values within 1.5 times the interquartile range, and points outliers beyond this.
-Note, the Y axis has been limited to 20 to aid readability, meaning the outliers Tomb G in the EG period, and Tomb P in the late EO-LO period are excluded, with values ranging from 29-34, and 36-44 respectively.*
-
-*Figure 16 Gini coefficient values through time based on the aoristic sums for all vessels across both cemeteries. Background colours correspond to ceramic periods.*
-
-*Figure 17 Aoristic sums of imported and 'oriental' imitation vessels through time across both cemeteries, colour-coded by regional origin in the case of imports.*
-
-*Figure 18 The composition of the total pottery assemblage for each century across both cemeteries, and among the 7 notable tombs (Group A) drawn out earlier in the analysis.*
-
-*Figure 19 Histograms corresponding, first, to counts of jewellery (beads, necklaces, rings, diadems etc.), metal objects, and imported items by tomb across both cemeteries and, secondly, to those counts divided by the estimated number of burials in each tomb.*
-
-##### pagebreak
-# Figures
-
-##### pagebreak
-# Tables
 

### 2020-10-13 README.Rmd
Repo name changes
Final edits

@@
 
-> Pollard, Dominic (2019). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
+> Pollard, Dominic (2020). _`r Title`_. Name of journal/book, <https://doi.org/xxx/xxx>
 
@@
 
-> Pollard, Dominic (2019). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Pollard, Dominic (2020). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+
+### Repository contents
+
+All data, text and code to recreate the paper and analysis therein are found in the 'analysis' folder.
+Within this, 'analysis/data' contains all raw and derived data, in .csv format.
+All charts and images from the paper can be found in 'analysis/figures', while the paper itself, and the BibTex file containing all cited works are found in 'analysis/paper'.
+The folder 'analysis/templates' contains the citation style file, and a template file used to set the style for the Word document produced by 'Family_Fortunes_Article.Rmd'.
 
@@
 
-> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Pollard, Dominic (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
 

### 2020-10-13 analysis/paper/Family_Fortunes_Article.Rmd
Final edits

@@
 However, despite well-published excavation reports on the two largest burial grounds and a number of studies dealing with various artefacts and individual tombs therein, no substantial comparative, quantitative analysis has yet been undertaken.
-This paper provides such an analysis, examining both the synchronic and diachronic variation among the tombs in these two cemeteries.
+Such an analysis is provided here, examining both the synchronic and diachronic variation among the tombs of these two cemeteries.
 Through a range of quantitative techniques, this study isolates a pattern of increased rates of burial among a few select tombs, and considers the broader social contexts within which these tombs were built and used, as well as discussing the possible nature of the burying groups.
@@
 To begin with the first, it may be felt that, of all ancient sites on Crete, Knossos is the last in need of further archaeological attention.
-The city and palace of Knossos, first excavated well over a century ago [@Evans1921; -@Evans1928; -@Evans1930], retain a pre-eminent position in popular and academic accounts the island’s Bronze Age (BA), and few, if any, other Cretan sites have been so thoroughly dug, reconstructed, and contested.
+The city and palace of Knossos, first excavated well over a century ago [@Evans1921; @Evans1928; @Evans1930], retain a pre-eminent position in popular and academic accounts the island’s Bronze Age (BA), and few, if any, other Cretan sites have been so thoroughly dug, reconstructed, and contested.
 
@@
 Speaking of the town’s stratigraphic sequence, Coldstream lamented that its “remains [...] are like a meagre filling in a very thick sandwich” between Bronze Age and later Greek and Roman levels [@Coldstream1991: 287].
-Most individual finds of Protogeometric (PG) or Geometric date take the form of flimsy foundations, patches of earth floor, wells or pottery scatters, and for a time it was argued that the area in the EIA comprised several dispersed villages, which later coalesced as per Aristotle’s model of synoecism [@1950].
+Most individual finds of Protogeometric (PG) or Geometric date take the form of flimsy foundations, patches of earth floor, wells or pottery scatters, and for a time it was argued that the area in the EIA comprised several dispersed villages, which later coalesced as per Aristotle’s model of synoecism [@Alexiou1950].
 
 However, a concentration of activity slightly to the west of the palace presented a good case for continuing, nucleated settlement [@Hood1981], something strongly advocated by Coldstream [-@Coldstream1984a; -@Coldstream2000], and, more recently, findings of the Knossos Urban Landscape Project (KULP) suggest a PG settlement of up to 40ha [@Kotsonas2011a: 5-8].
-This estimate remains our best guess in the absence of substantial excavated deposits but, if accurate, suggests that EIA Knossos was one of the largest sites of the contemporary Aegean.
+If accurate, this suggests that EIA Knossos was one of the largest sites of the contemporary Aegean.
 
@@
 Fortunately, far larger burial accumulations are to be found in the Fortetsa and the Knossos North cemeteries, which together comprise some 134 tombs, ranging in date from the Subminoan (SM) to Late Orienatalising (LO) periods.
-Though many were robbed in antiquity, and others destroyed by the digging of later burials, building work, or just the ravages of time, they still represent the most abundant source of evidence for not just mortuary, but social practice of any kind at EIA Knossos.
+TAken together, they represent the most abundant source of evidence for not just mortuary, but social practice of any kind at EIA Knossos.
 
@@
 
-The cemetery on the slope facing the eponymous village of Fortetsa comprises twelve tombs excavated in 1933, and another eight uncovered in 1935 (Brock 1957: xi), grouped into three main clusters running north to south.
-Three of those excavated in 1933 – L, TFT and Π – which lie close to the main road north of the Acropolis hill [@Brock1957: 1-2], are now considered to belong the Knossos North Cemetery (KNC).
+The cemetery on the slope facing the eponymous village of Fortetsa comprises twelve tombs excavated in 1933, and another eight uncovered in 1935 [@Brock1957: xi, 1-2], grouped into three main clusters running north to south.
+Three of those excavated in 1933 – L, TFT and Π – which lie close to the main road north of the Acropolis hill , are now considered to belong the Knossos North Cemetery (KNC).
 
@@
 comm.).
-
-During the use of the cemeteries, correspondences have been highlighted between changing pottery styles and episodes of tomb construction and abandonment.
-The Protogeometric B (PGB) style, first defined by Brock [-@Brock1957: 143], who hailed it as “the most remarkable phase in Cretan vase-painting”, is apparently a Cretan anomaly, evincing a melding of Attic geometric motifs, patterns possibly deriving from Near Eastern metalwork, and iconography drawn from the Minoan repertoire [@Coldstream1984b: 93-94].
-Coldstream has argued that the concurrence of this style with a rash of newly founded tombs points to a reformulation of elite mortuary display, noting how, in this period also, redeposited Minoan larnakes began appearing in and around some tombs, in one case even inspiring the decoration of a PGB vessel [in T.
-107; see @Coldstream1984b; -@Coldstream1994: 112–13].
-Whether simply nostalgia [@Coldstream1998: 60], or a more tactical strategy by newly emerging elite groups [@Coldstream1994: 114-15], the Bronze Age past clearly possessed a significant and enduring cultural cachet for the EIA Knossians.
 
@@
 Fertilised by the Homeric motif of gift-giving [@Whitley2002: 220-21], and the remarkable archaeological parallels for such items – such as the boar’s tusk helmet from KNC Tomb 201 and that given to Odysseus by Meriones (*Iliad* 10.260-271) - such approaches have been brought to bear on certain elaborate burials of the EIA.
-
-For instance, Tomb 201 in the KNC contained a bronze sword, spearhead, and arrowhead, an iron dirk or knife, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned helmet [@Catling1995: 123].
-Catling identifies similarities between this tomb and burials at Tiryns, Kaloriziki, and the famous *hērōön* of Lefkandi; all contained imported items, were of ‘warrior grave’ type, and formed loci for later burials.
-He has argued that individuals who dared to travel abroad in this time of more limited seafaring, accruing exotica and stories from distant lands, would have been accorded exceptional treatment, including the deposition of such items, in death [@Catling1995: 127–8].
+For instance, Tomb 201 in the KNC contained various weapons, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned helmet [@Catling1995: 123].
+Based similarities between this tomb and burials at Tiryns, Kaloriziki, and the famous *hērōön* of Lefkandi, Catling has argued that these burials represent individuals who dared to travel abroad in this time of more limited seafaring, accruing exotica and stories from distant lands, and accorded exceptional treatment in death [@Catling1995: 127–8].
 
@@
 Variables include the tomb, the vessel shape, its surviving dimensions, its ceramic period and corresponding absolute dating measures.
-In addition, all imports, 'oriental' imitations, and antique (that is, Bronze Age) artefacts were categorised, using the studies of Antoniadis (2012), Jones [-@Jones2000], and Crowe [-@Crowe2016] respectively.
+In addition, all imports, 'Oriental' (read, Cypriot or Near Eastern) imitations, and antique (that is, Bronze Age) artefacts were categorised, using the studies of Antoniadis (2012), Jones [-@Jones2000], and Crowe [-@Crowe2016] respectively.
 In total, this database comprises just over 4540 individual artefacts.
@@
 Within this field, catalogues and excavation reports abound and, though many of these may be forgiven based on their dates of publication, the lack of readily utilisable digital (or even digitisable) data from a number of recent articles and monographs is less pardonable.
-
-There are, of course, valid reservations regarding issues of copyright, the protection of sites from looting, the possibility of being scooped with one's own data, or simply the significant time investment of preparing open-access research.
+There are, of course, valid reservations regarding the risks involved in sharing  archaeological data, not to mention the significant time investment of preparing open-access research.
 But I would argue that the potential gains outweigh these concerns, and that research on the EIA would benefit hugely from the greater accessability of datasets and workflows.
@@
 In this vein, the present article has been composed making use of Ben Marwick's 'rrtools' [@Marwick2017; @inbook], an open-source package for the statistical software R Studio.
-This provides the author with the tools to create a self-contained 'compendium', which can be freely downloaded, and from which anyone can reproduce the entire workflow of the following analysis.
-Though not visible in the present paper, every chart and table herein was produced directly, through written code, from the raw data using R (and note also that, wherever possible, these are presented in colour-blind friendly palettes).
+This provides the author with the tools to create a self-contained 'compendium', which can be freely downloaded, and from which anyone can reproduce the entire workflow of the published analysis.
+Though not visible in the present paper, every graph and table herein was produced directly, through written code, from the raw data using R (and note also that, wherever possible, these are presented in colour-blind friendly palettes).
 This code, as well as the raw and derived datasets, complete bibliography (in BibTex format, for export to any major reference manager), figures, and original article are included in the compendium.
 It can be downloaded as a repository from GitHub, at the following address: https://github.com/DCPollard94/knossoscemeteries, and the raw datasets have been stored with the online repository Figshare, here: (**NB a URL will be included pending acceptance of the article**).
-The data are made available through an MIT license, which permits unlimited use, ammendation, and dissemination thereof, yet frees the present author from any responsibilty for that use, or perceived errors or inaccuracies in the data.
 It is sincerely hoped that others may further explore and expand upon the research presented here by downloading this compendium.
@@
 A first important step in preparing the data for analysis was the estimation of the number of individuals buried in each tomb.
-The method chosen broadly follows that of Cavanagh [-@Cavanagh1996: 659-60], who took the number of pithoi (the most common urn) and the number of pithoi, amphorae and kraters (which were sometimes used as such) as middle and upper estimates of the number of burials, with all securely identified interments being the lower limit (Figure \@ref(fig:burialestimate)).
+The method chosen broadly follows that of Cavanagh [-@Cavanagh1996: 659-60], who took the number of pithoi (the most common urn) and the number of pithoi, amphorae and kraters (which were sometimes used as such) as middle and upper estimates of the number of burials, with all securely identified interments being the lower limit.
 Here, the mean of these three estimates was taken, to provide a single aggregate measure for the purposes of analysis.
@@
 
-But the artificial clustering this technique engenders around select absolute dates can prove misleading, as in Figure \@ref(fig:allpithoi); the apparent cessation of depositions in the 8th century is simply a fiction generated by the processing of the data.
+But the artificial clustering this technique engenders around select absolute dates can prove misleading, as in Figure \@ref(fig:allpithoi) (left); the apparent cessation of depositions in the 8th century is simply a fiction generated by the processing of the data.
 
@@
 
-As seen in Figure \@ref(fig:aoristallcems), the artificial peaks and troughs of the simple averaging method are avoided, and different patterns become visible.
+As seen in Figure \@ref(fig:allpithoi) (right), the artificial peaks and troughs of the simple averaging method are avoided, and different patterns become visible.
 Though these methods present their own limitations [@Crema2010: 1123-24], they are considered a more robust and productive approach to the temporal uncertainty of ceramic dating.
@@
         theme_bw()+theme(legend.title=element_blank(),
-          text=element_text(family="serif", size=11))+
+          text=element_text(family="serif", size=12))+
             labs(x="Date (Years BC)", y="Aoristic Sum (Pithoi)")+
@@
         theme_bw()+theme(legend.title=element_blank(),
-          text=element_text(family="serif", size=11))+
+          text=element_text(family="serif", size=12))+
             labs(x="Date (Years BC)", y="Aoristic Sum (All Vessels)")+
@@
 In both cases, tombs lacking any securely dateable pottery are excluded.
-Abbreviations: SM - Subminoan; (E/M/L)PG - (Early/Middle/Late) Protogeometric; PGB - Protogeometric B; (E/M/L)G - (Early/Middle/Late) Geometric; (E/L)O - (Early/Late) Orientalising"}
+Abbreviations: SM - Subminoan; (E/M/L)PG - (Early/Middle/Late) Protogeometric; PGB - Protogeometric B; (E/M/L)G - (Early/Middle/Late) Geometric; (E/L)O - (Early/Late) Orientalising."}
 
@@
                text=element_text(family="serif", size=10),axis.text.y=element_text(angle=90, vjust=0, hjust=0.5))+
-                 labs(x="Century BC (Date of Vessel)", y="Vessel Count")+
+                 labs(x="Date of Vessel (Century BC)", y="Vessel Count")+
                   scale_fill_manual(values = c("gray50"))+ 
@@
                text=element_text(family="serif", size=10),axis.text.y=element_text(angle=90, vjust=0, hjust=0.5))+
-                 labs(x="Century BC (Date of Tomb)", y="Vessel Count")+
+                 labs(x="Date of Tomb (Century BC)", y="Vessel Count")+
                   scale_fill_manual(values = c("gray50"))+ 
@@
 *Table 1 The number of pithoi (as a proxy for burials) deposited in tombs dating to the 10th, 9th, 8th and 7th centuries respectively.
-For each of the centuries the tombs were in use (column two), counts of pithoi (column three) are divided by the number of tombs from which these vessels came (column 4), allowing calculation of the changing rate of burial through time (column five).*
+For each of the centuries the tombs were in use (column two), counts of pithoi (column three) are divided by the number of tombs from which these vessels came (column four), allowing calculation of the changing rate of burial through time (column five).*
 
@@
 Using pithoi as a proxy for burials, we can estimate the changing rates of burial through time.
-And while these mostly remained around three per tomb per century, in the 8th and 7th centuries, tombs founded in the 9th century saw a rapid rise in the rate of burial, peaking at over 11 per tomb per century.
+And while these mostly remained around three per tomb per century, in the 8th and 7th centuries tombs founded in the 9th saw a rapid rise in the rate of burial, peaking at over 11 per tomb per century.
 Though no doubt rough estimates, the marked nature of this divergence suggests it is not simply an artefact of imperfect data.
@@
 The 12 tombs with the most associated vessels are all chamber tombs, 10 dating to the 9th century and two to the 10th, and seven of them did exhibit late, intense episodes of deposition (Figure \@ref(fig:toptombs)).
-In fact, Tombs P, 292, 75, 107, 285, II and 218 (henceforth 'Group A') between them account for `r round(calc1,digits=2)`% of the entire ceramic assemblage from both cemeteries.
+In fact, Tombs P, 292, 75, 107, 285, II and 218 (henceforth, 'Group A') between them account for `r round(calc1,digits=2)`% of the entire ceramic assemblage from both cemeteries.
 All but Tomb 285 date to the 9th century, and together they underpinned the discussed late spate of burials.
@@
 Interestingly, Tombs X, G, Q, 219 and 283 ('Group B') evince a different pattern, with their respective peaks coming around a century earlier, another period of increased burial activity.
-So despite receiving the greatest quantities of burials and grave goods, these tombs nonetheless seem to have achieved that distinction via quite different histories, a divergence made clear in Figure \@ref(fig:groupsab).
+So despite receiving the most burials and grave goods across both cemeteries, these 12 tombs nonetheless seem to have achieved that distinction via quite different histories, a divergence made clear in Figure \@ref(fig:groupsab).
 It is notable that tombs assinged to each group are seen in both cemeteries, suggesting from the outset that the differences between these may speak to social forces operating across the Knossian community.
@@
 
-```{r tombgroupdimensions, fig.dim = c(3.5, 3.5),out.width="10%", dpi=600, fig.align='center', fig.cap="Scatterplot comparing the dimensions of* dromos *length and chamber area.
+```{r tombgroupdimensions, fig.dim = c(3.5, 3.5),out.width="10%", dpi=600, fig.align='center', fig.cap="Scatterplot comparing the dimensions of *dromos* length and chamber area.
 Groups A and B, defined earlier, are illustrated and labelled in green and orange respectively, while all remaining tombs ('Group C') are in blue."}
@@
 
-Looking more specifically at the tombs of Groups A and B, against the backdrop of the other 10th and 9th century tombs, there are perhaps hints of differentiation.
+Looking more specifically at the tombs of Groups A and B, against the backdrop of the other 10th and 9th century tombs, there are perhaps hints of differentiation (Figure \@ref(fig:tombgroupdimensions)).
 All of them fall within the top 50% based on chamber area, and three have exceptionally long *dromoi*.
@@
 
-```{r tombdimensions, fig.dim = c(7.5, 4), dpi=600, fig.align='center', fig.cap="Scatterplots comparing tomb dimensions (chamber area and* dromos *length) to the estimated total number of burials by tomb.
-The points area colour-coded as per Figure 11."}
+```{r tombdimensions, fig.dim = c(7.5, 4), dpi=600, fig.align='center', fig.cap="Scatterplots comparing tomb dimensions (chamber area and *dromos* length) to the estimated total number of burials by tomb.
+The points are colour-coded as per Figure 11."}
 
 tombareas<-ggplot(subset(tombs,tombs$est_burials>1),aes(chamber_area,est_burials,color=group))+geom_point()+
-  theme_bw()+theme(text=element_text(family="serif", size=10))+
+  theme_bw()+theme(text=element_text(family="serif", size=11))+
             labs(x="Chamber Area (sq m)", y="Estimated Total Burials")+
@@
 dromoslengths<-ggplot(subset(tombs,tombs$est_burials>1),aes(dromos_length,est_burials,color=group))+geom_point()+
-  theme_bw()+theme(text=element_text(family="serif", size=10))+
+  theme_bw()+theme(text=element_text(family="serif", size=11))+
             labs(x="Dromos Length (m)", y="Estimated Total Burials")+
@@
 
-And in Figure \@ref(fig:tombscompthroughtime), we see that the tombs of Group A, while not paultry in the 9th century, were by no means exceptional.
-They clearly kept pace with the other popular tombs of the time, but their major divergence only occurred in the subsequent two hundred years.
+And in Figure \@ref(fig:tombscompthroughtime), we see that the tombs of Group A, while not paultry in the 9th century, were by no means exceptional in the quantity of burial goods they received.
+They  kept pace with the other popular tombs of the time, but their major divergence only occurred in the subsequent two centuries.
 Group B, meanwhile, experienced similar, if not greater, use in the 9th century, and then gradually declined.
@@
 
-```{r bigbox, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="Boxplot showing the aoristic sums of all vessels across both cemeteries through time, colour-coded by ceramic period.
-Boxes represent the interquartile range, horizontal lines the median value, whiskers values within 1.5 times the interquartile range, and points outliers beyond this.
+```{r bigbox, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="Boxplot showing variation in aoristic sums of all vessels across both cemeteries through time, colour-coded by ceramic period.
 Note, the Y axis has been limited to 20 to aid readability, meaning the outliers Tomb G in the EG period, and Tomb P in the late EO-LO period are excluded, with values ranging from 29-34, and 36-44 respectively."}
@@
 
-Figure \@ref(fig:bigbox) is a boxplot of the total aoristic sums across both cemeteries, colour-coded by ceramic phase.
+Figure \@ref(fig:bigbox) is a boxplot of the total aoristic sums per tomb across both cemeteries, colour-coded by ceramic phase.
 Looking at the PGB and EG, we see that they have the largest interquartile ranges (IQRs) of any period, and only a single outlier (beyond 1.5 times the IQR) each.
@@
 
-This is not to suggest that the late 9th to early 8th century#ies was a time of egalitarian burial practices.
+This is not to suggest that the late 9th to early 8th centuries was a time of egalitarian burial practices.
 The PGB and EG are consistently interpreted as periods of artistic development, increased overseas trade, and renewed vigour in the construction of tombs and the visibility of the burial rite [@Brock1957: 143-44; @Coldstream1994: 110-12].
-The PGB style reflects a uniquely Cretan blend of Attic and Near Eastern forms, while the appearance of BA 'antiques' in the tombs from this period onwards is felt to reflect rejuvinated, even competitive interest in the Minoan past.
+The PGB style, first defined by Brock [-@Brock1957: 143], who hailed it as “the most remarkable phase in Cretan vase-painting”, is apparently a Cretan anomaly, evincing a melding of Attic geometric motifs, patterns possibly deriving from Near Eastern metalwork, and iconography drawn from the Minoan repertoire [@Coldstream1984b: 93-94].
+Coldstream has tied the emergence of this style to a reformulation of elite mortuary display, noting how, in this period also, redeposited Minoan larnakes began appearing in and around some tombs, in one case even inspiring the decoration of a PGB vessel [in T.
+107; see @Coldstream1984b; -@Coldstream1994: 112–13].
+Whether simply nostalgia [@Coldstream1998: 60], or a more tactical strategy by newly emerging elite groups [@Coldstream1994: 114-15], the Bronze Age past appears to have possessed a significant cultural cachet for the EIA Knossians, as may the new, elaborately decorated pottery of the PGB, and items imported from overseas.
 
@@
 
-The rise in imported items is readily observable, with a succession of Greek and Near Eastern imports in the late 9th and early 8th centuries (Figure \@ref(fig:imports)).
+The rise in imported items around this time is readily observable, with a succession of Greek and Near Eastern imports in the late 9th and early 8th centuries (Figure \@ref(fig:imports)).
 Likewise, the decline in Near Eastern imports, and concomitant rise in locally-made imitations thereof, is visible at the end of the 8th century.
@@
 Indeed, among 7th century vessels, when interments were higher than ever among Group A, the figure is a mere `r round(groupA7thimports, digits=2)`%.
-The impression grows stronger that the context for the late proliferation of burials in the Group A tombs was altogether different from that underpinning the comparative, competitive diversity of interments a century earlier.
+The impression grows stronger that the context for the late proliferation of burials in the Group A tombs was altogether different from that underpinning the comparative - if competitive - diversity of interments a century earlier.
 
@@
 
-But this association, perhaps tempting one initially to speak of 'elite' or 'high-status' tombs, evaporates when the number of objects is divided by the number of burials represented.
-This fact has seldom, if ever, been appreciated in prior discussions of the cemeteries (Figure \@ref(fig:hists)).
+But this association, perhaps tempting one initially to speak of 'elite' or 'high-status' tombs, evaporates when the number of objects is divided by the number of burials represented (Figure \@ref(fig:hists)).
 Whitley (2004: 437) has pointed out that the proportions of different decorative styles of Orientalising pithoi in Tomb P are replicated across Fortetsa as a whole, but it seems this situation may apply across most artefact classes.
@@
 In the traditional view, political authority was vested in the *kosmoi*, aristocratic magistrates selected from the leading, hereditary *startoi*, or ‘clans’, networks of extended lineages controlling the ancestral estates on which worked attendent serfs and/or slaves (Willetts 1982, 240).
-But despite elements of this account probably being accurate, others remain equivocal, such as the nature of the *pyla*, or 'tribe', and its relation to the *startos* (which may itself be a military grouping), the distinction between *astos* and *poliatas*, both probably formulations of the rights-bearing civic actor, and the ancestry and spread of the *agelaos*, an adolescent age-grade currently only attested at Eltynia [@Gagarin2016: 36; 78-80].
+But despite elements of this account probably being accurate, others remain equivocal, such as the nature of the *pyla*, or 'tribe', and its relation to the *startos* (which may itself be a military grouping), the distinction between *astos* and *poliatas*, both probably formulations of the rights-bearing civic actor, and the ancestry and spread of the *agelaos*, an adolescent age-grade currently only attested at Eltynia [@Gagarin2016: 36, 78-80].
 Indeed, many terms which might illuminate social divisions and groupings within the Archaic Cretan populace are only attested at one or two sites, and even then sometimes apparently differ in form or function.
@@
 That the noted spate of burials in the late 8th and early 7th centuries was concentrated solely in older tombs strongly implies that issues of heritage, lineage, and tradition were at play.
-In Tomb 107, the stratigraphy of the niches cut into the *dromos* suggest depositions continued to be made even after the main chamber fell into disrepair [@Coldstream1996a: 150]; it is the symbolic and associative, rather than physical qualities of these tombs seem instrumental to their extensive reuse.
+In Tomb 107, the stratigraphy of the niches cut into the *dromos* suggest depositions continued to be made even after the main chamber fell into disrepair [@Coldstream1996a: 150].
+It is the symbolic and associative, rather than physical qualities of these tombs which seem instrumental to their extensive reuse, and the emphasising of group membership, of belonging - be it to a powerful family, tribe or age-grade - is surely crucial to understanding these patterns.
 
@@
 
-Together, these varied threads suggest that, with the emergence of the archaic *poleis* on Crete, there was a concomitant restructuring of wider social groups, institutions, and behaviours.
+These varied lines of evidence, and others, suggest that, with the emergence of the archaic *poleis* on Crete, there was a concomitant restructuring of wider social groups, institutions, and behaviours.
 It seems not unreasonable to infer an emergent emphasis on the citizen, the state, and ideologies of relative egality, within which communal gatherings, intra-mural ritual, and individual graves replaced ostentatious, kin-centred acts of gathering, dedication, and burial.
-And in the Knossian cemeteries, we have seen a somewhat diverse set of burial practices give way to a more prolific, yet more restricted, suite of funerary rites, focussed on a small number of long-established tombs, in the context of diminishing collective tomb construction, replacement of imports with locally-made imitations, and an imminent cessation of such depositions.
+And at Knossos, we have seen a somewhat diverse set of burial practices give way to a more prolific, yet more restricted, suite of funerary rites, focussed on a small number of long-established tombs, in the context of diminishing collective tomb construction, replacement of imports with locally-made imitations, not to mention the imminent cessation of such burials in both cemeteries.
 
@@
 As a more nuanced picture emerges of the island's relationship with the wider Mediterranean, subtler characterisations are possible.
-It is telling that the peak for imported items came in the 9th century; the Orientalising as a time period feels increasingly misapplied to Crete [@Whitley2004: 434], when, at Knossos at least, it witnessed far lower rates of 'oriental' imports, and only limited adoption of the figural motifs which characterise its appearance elsewhere.
+It is telling that the peak for imported items came in the 9th century; the Orientalising as a time period feels increasingly misapplied to Crete [@Whitley2004: 434], when, at Knossos at least, it witnessed far lower rates of 'Oriental' imports, and only limited adoption of the figural motifs which characterise the style's appearance elsewhere.
 
@@
 Further work should seek to extend these comparisons, both into trends unexplored in the present study, and out to the wider funerary landscape of Knossos.
-I hope to do just this but, for the time being, the datasets I am making available alongside this article include the other published EIA tombs in the area, so that others may also explore their possible significance.
+I hope to do just this but, for the time being, the datasets I am making available alongside this article include the other published EIA tombs in the Knossos area, so that others may also explore their possible significance.
 
@@
 Despite little mention in the publication, this material was not inconsiderable and so, regretably, future work will be necessary to determine whether it threatens the integrity of these analyses.
-Suffice to say that, at present, it is another pinch of salt with which to take the present findings.
+At present, it is a pinch of salt with which to take the present findings.
 
@@
 The databases produced for the above analysis contain a large of amount of unutilised data that could facilitate a range of future investigations, and all are welcome to it.
-With more such analyses, we could begin to a construct a fuller account of the regional diversity in mortuary, and wider social, practices that characterised the communities of EIA and archaic Crete.
+With more such analyses, we could begin to a construct a fuller account of the regional diversity in mortuary, and wider social, practices among the communities of EIA and Archaic Crete.
 
@@
 ##### pagebreak
-
-#####pagebreak
-
 ### Colophon

### 2021-02-19 analysis/paper/Family_Fortunes_Article.Rmd
Edits

@@
 ---
-title: "Family Fortunes: A Quantitative Analysis of the Early Iron Age Cemeteries at Knossos, Crete"
+title: "Family Fortunes: A Quantitative Analysis of the Early Iron Age Cemeteries of Knossos, Crete"
 author:
@@
 library(ineq)
-library(reshape)
+library(reshape2)
 library(data.table)
@@
 library(ggrepel)
+library(forcats)
+library(factoextra)
+library(cluster)
 
@@
 Speaking of the town’s stratigraphic sequence, Coldstream lamented that its “remains [...] are like a meagre filling in a very thick sandwich” between Bronze Age and later Greek and Roman levels [@Coldstream1991: 287].
-Most individual finds of Protogeometric (PG) or Geometric date take the form of flimsy foundations, patches of earth floor, wells or pottery scatters, and for a time it was argued that the area in the EIA comprised several dispersed villages, which later coalesced as per Aristotle’s model of synoecism [@Alexiou1950].
+Most individual finds of Protogeometric (PG) or Geometric (G) date take the form of flimsy foundations, patches of earth floor, wells or pottery scatters, and for a time it was argued that the area in the EIA comprised several dispersed villages, which later coalesced as per Aristotle’s model of synoecism [@Alexiou1950].
 
@@
 
-```{r map, fig.align= "centre", fig.cap="Map of the Knossos area, showing the locations of the various EIA tomb clusters and cemeteries, with EIA pottery distributions from the Knossos Urban Landscape Project marked in orange, giving an impression of the contemporary urban extent (underlying image credit: T.
+```{r map, fig.align= "centre", fig.cap="Map of the Knossos area, showing the locations of the various EIA tombs (crosses) and cemeteries, with EIA pottery distributions from the Knossos Urban Landscape Project (circles), giving an impression of the contemporary urban extent (underlying image credit: T.
 Whitelaw)."}
@@
 The transitional period between the Late Bronze Age (LBA) and EIA yields few if any securely datable interments, but later evidence is more plentiful, with burials known from the Kephala ridge [@Coldstream1963: 38; @Hogarth1899: 82-5], the modern suburbs of Ayios Ioannis and Atsalenio [@Boardman1960a; @Davaras1968: 133–46], the cave of Mavro Spelio [@Antoniadis2012: 58-9], and the Khaniale Teke site with its well-known tholos [@Boardman1967;@Hutchinson1954; cf.
-@Hoffman1997 @Kotsonas2006].
+@Hoffman1997;@Kotsonas2006].
 Taken together, these scattered tombs point to a richly variegated set of funerary practices but, despite their reasonable number, are so dispersed that they provide little scope for systematic quantitative analysis.
 
-Fortunately, far larger burial accumulations are to be found in the Fortetsa and the Knossos North cemeteries, which together comprise some 134 tombs, ranging in date from the Subminoan (SM) to Late Orienatalising (LO) periods.
-TAken together, they represent the most abundant source of evidence for not just mortuary, but social practice of any kind at EIA Knossos.
+Fortunately, far larger burial accumulations are to be found in the Fortetsa and the Knossos North cemeteries (Figure \@ref(fig:map2)), which together comprise some 134 tombs, ranging in date from the Subminoan (SM) to Late Orienatalising (LO) periods.
+Taken together, they represent the most abundant source of evidence for not just mortuary, but social practice of any kind at EIA Knossos.
 
@@
 Three of those excavated in 1933 – L, TFT and Π – which lie close to the main road north of the Acropolis hill , are now considered to belong the Knossos North Cemetery (KNC).
+
+```{r map2, fig.align= "centre", fig.cap="Closer view of the Knossos North and Fortetsa cemeteries.
+Tombs were digitised and georeferenced from the original publications, Coldstream and Catling (1996) and Brock (1957)."}
+
+knitr::include_graphics("../figures/Knossos Tombs.jpg")
+```
 
@@
 Northwest of these, and just south of the Teke Tholos, the Teke tombs represent the northern reaches of the KNC.
-The central section of the KNC comprises the ‘Medical Faculty’ site, named for the extension of the University of Crete that prompted major rescue excavations there in 1978 [@Coldstream1996a: 53-55].
+The central section of the KNC comprises the ‘Medical Faculty’ site (KNC Main and NW in Figure \@ref(fig:map2)), named for the extension of the University of Crete that prompted major rescue excavations there in 1978 [@Coldstream1996a: 53-55].
 
@@
 The *dromos* and chamber are separated by the *stomion*, an opening often set a step or two lower than the *dromos*, which is covered by a large stone slab, smaller stacked stones, or combination of both [@Antoniadis2012: 47-48].
-The majority of burials are cremations, a rite that appears at Knossos in the Subminoan period, in the Tomb 200+ complex in the KNC [@Coldstream1994: 109].
+The vast majority of burials are cremations deposited in pithoi or other urns, a rite that appears at Knossos in the Subminoan period, in the Tomb 200+ complex in the KNC [@Coldstream1994: 109].
 Inhumation does not completely disappear in the EIA, but by the Orientalising period it appears restricted solely to child burials [@Antoniadis2012: 69].
@@
 For instance, Tomb 201 in the KNC contained various weapons, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned helmet [@Catling1995: 123].
-Based similarities between this tomb and burials at Tiryns, Kaloriziki, and the famous *hērōön* of Lefkandi, Catling has argued that these burials represent individuals who dared to travel abroad in this time of more limited seafaring, accruing exotica and stories from distant lands, and accorded exceptional treatment in death [@Catling1995: 127–8].
+Based similarities between this tomb and burials at Tiryns, Kaloriziki, and the famous *hērōön* of Lefkandi, Catling has argued that these burials represent individuals who dared to travel abroad in this time of more limited seafaring, accruing exotica and stories from distant lands, and so were accorded exceptional treatment in death [@Catling1995: 127–8].
 
@@
 
-## 2.6 Number Crunching: Quantitative Analyses of the Cemeteries
+## 2.6 Quantitative Analyses of the Cemeteries
 
@@
 
-```{r allpithoi,fig.dim = c(8, 3), dpi=600, fig.align='center', fig.cap="(Left) Counts of pithoi across both cemeteries through time.
-Vessel dates have here been based on the midpoint (or mean) of the absolute date range accompanying their respective ceramic periods.
-(Right) Aoristic sums of all vessels across both cemeteries.
+```{r allpithoi,fig.dim = c(7, 7), dpi=600, fig.align='center', fig.cap="(A) Counts of all vessels per decade across both cemeteries.
+Vessel dates have here been based on the mean of the absolute date range corresponding to their respective ceramic periods.
+(B) The same but with aoristic sums replacing raw counts.
 This represents a probabilistically weighted visualisation of depositional activity through time."}
@@
 
-non_aorist_fig<-ggplot(pots, aes(ave_date,fill="Pots"))+geom_histogram(fill="gray40")+
+rects <- data.frame(xstart = c(1100,970,920,875,840,810,790,745,710,670,630), 
+                    xend =    c(970,920,875,840,810,790,745,710,670,630,600), 
+                    xmid =   c(1035,945,897.5,857.5,825,800,767.5,727.5,690,650,615),
+                    col = c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
+rects$col<-factor(rects$col,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
+
+non_aorist_fig<-ggplot()+geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf), fill=NA, colour="grey90",alpha = 0.4) +
+  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = 590, ymax = 620), fill=NA, colour="black",alpha = 0.4) +
+  geom_histogram(data=pots, aes(ave_date,fill="Pots"),fill="gray40",binwidth = 10)+
+  geom_text(data=rects,aes(xmid, 607,label=col),family="serif",size=3)+
   theme_bw()+theme(legend.title=element_blank(),
@@
           labs(x="Date (Years BC)", y="Vessel Count")+
-                theme(legend.position="none")+
-                  scale_y_continuous(expand=c(0,0),limits=c(0,610))+
+                theme(legend.position="none",panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,620))+
                     scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
@@
 
-aorist_all_tombs<-colSums(tombs[158:207])
+aorist_all_tombs<-colSums(tombs[160:209])
 aorist_all_tombs<-data.frame(aorist_all_tombs)
@@
 
-aorist_fig<-ggplot(aorist_all_tombs,aes(decade,aoristic_sum, fill="Aoristic Sum"))+
-  geom_col(fill="gray40")+theme_bw()+theme(legend.title=element_blank(),
+aorist_fig<-ggplot()+
+   geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf), fill=NA, colour="grey90",alpha = 0.4) +
+  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = 214, ymax = 225), fill=NA, colour="black",alpha = 0.4) +
+  geom_text(data=rects,aes(xmid, 220,label=col),family="serif",size=3)+
+  geom_col(data=aorist_all_tombs,aes(decade,aoristic_sum, fill="Aoristic
+  Sum"),fill="gray40")+theme_bw()+theme(legend.title=element_blank(),
       text=element_text(family="serif", size=11))+
           labs(x="Date (Years BC)", y="Aoristic Sum")+
-               theme(legend.position="none")+
+              theme(legend.position="none",panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
                   scale_y_continuous(expand=c(0,0),limits=c(0,225))+
@@
 
-ggarrange(non_aorist_fig,aorist_fig,ncol=2,common.legend=FALSE)
+ggarrange(non_aorist_fig,aorist_fig,nrow=2,labels=c("A","B"),align="v",common.legend=FALSE)
 
@@
 
-```{r aoristbycem, dpi=600, fig.align="center", fig.cap="Aoristic sums of pithoi (left), as a proxy for cremation burials, and of all vessels (right).
-The sums are colour coded to show the values of the two cemeteries, and the total values.", fig.dim=c(8,3.5)}
+```{r aoristbycem, dpi=600, fig.align="center", fig.cap="(A) Aoristic sums of pithoi, as a proxy for cremation burials, through time. (B) Aoristic sums of all vessels through time.", fig.dim=c(7,7.5)}
 KNC<-subset(tombs,tombs$cem_condensed=="KNC")
 
-aorist_KNC_pithoi <- colSums(KNC[208:257])
+aorist_KNC_pithoi <- colSums(KNC[210:259])
 aorist_KNC_pithoi<-data.frame(aorist_KNC_pithoi)
@@
 
-aorist_FOR_pithoi <-colSums(FOR[208:257])
+aorist_FOR_pithoi <-colSums(FOR[210:259])
 aorist_FOR_pithoi<-data.frame(aorist_FOR_pithoi)
@@
 
-aorist_all_pithoi <- colSums(tombs[208:257])
+aorist_all_pithoi <- colSums(tombs[210:259])
 aorist_all_pithoi<-data.frame(aorist_all_pithoi)
@@
 
+aorist_FOR_pithoi$cemetery<-"FC"
+aorist_KNC_pithoi$cemetery<-"KNC"
+aorist_all_pithoi$cemetery<-"All tombs"
+aorist_pithoi<-rbind(aorist_all_pithoi,aorist_KNC_pithoi,aorist_FOR_pithoi)
+aorist_pithoi$cemetery<-factor(aorist_pithoi$cemetery,levels=c("All tombs", "KNC", "FC"))
+
 aorist_pithoi_plot<-ggplot(NULL)+
-  geom_col(data=aorist_all_pithoi,aes(decade,aoristic_sum,fill=" All tombs  "))+
-    geom_col(data=aorist_KNC_pithoi,aes(decade,aoristic_sum,fill=" KNC  "))+
-      geom_col(data=aorist_FOR_pithoi,aes(decade,aoristic_sum, fill=" Fortetsa  "))+
+  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf), fill=NA, colour="grey90",alpha = 0.4) +
+  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = 43, ymax = 45), fill=NA, colour="black",alpha = 0.4) +
+  geom_text(data=rects,aes(xmid, 44,label=col),family="serif",size=3)+
+  geom_line(data=aorist_pithoi,aes(decade,aoristic_sum,linetype=cemetery))+ 
        theme_bw()+theme(legend.title=element_blank(),
-          text=element_text(family="serif", size=12))+
+          text=element_text(family="serif", size=12),panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+          plot.margin=unit(c(0.1,1,0.5,1),"cm"))+
             labs(x="Date (Years BC)", y="Aoristic Sum (Pithoi)")+
-              scale_fill_manual(values = c("gray50","gray65","gray80"))+ 
                 scale_y_continuous(expand=c(0,0),limits=c(0,45))+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-aorist_KNC <- colSums(KNC[158:207])
+                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))
+
+aorist_KNC <- colSums(KNC[160:209])
 aorist_KNC<-data.frame(aorist_KNC)
@@
 
-aorist_FOR <-colSums(FOR[158:207])
+aorist_FOR <-colSums(FOR[160:209])
 aorist_FOR<-data.frame(aorist_FOR)
@@
 
+aorist_FOR$cemetery<-"FC"
+aorist_KNC$cemetery<-"KNC"
+aorist_all_tombs$cemetery<-"All tombs"
+aorist_tombs<-rbind(aorist_all_tombs,aorist_KNC,aorist_FOR)
+aorist_tombs$cemetery<-factor(aorist_pithoi$cemetery,levels=c("All tombs", "KNC", "FC"))
+
 aorist_pots_plot<-ggplot(NULL)+
-  geom_col(data=aorist_all_tombs,aes(decade,aoristic_sum,fill=" All tombs  "))+
-    geom_col(data=aorist_KNC,aes(decade,aoristic_sum,fill=" KNC  "))+
-      geom_col(data=aorist_FOR,aes(decade,aoristic_sum, fill=" Fortetsa  "))+
+   geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf), fill=NA, colour="grey90",alpha = 0.4) +
+  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = 220, ymax = 230), fill=NA, colour="black",alpha = 0.4) +
+  geom_text(data=rects,aes(xmid, 225,label=col),family="serif",size=3)+
+  geom_line(data=aorist_tombs,aes(decade,aoristic_sum,linetype=cemetery))+
         theme_bw()+theme(legend.title=element_blank(),
-          text=element_text(family="serif", size=12))+
+          text=element_text(family="serif", size=12),panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+          plot.margin=unit(c(0.1,1,0.1,1),"cm"))+
             labs(x="Date (Years BC)", y="Aoristic Sum (All Vessels)")+
               scale_fill_manual(values = c("gray50","gray65","gray80"))+  
-                  scale_y_continuous(expand=c(0,0),limits=c(0,225))+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-ggarrange(aorist_pithoi_plot,aorist_pots_plot,ncol=2,common.legend=TRUE, legend="bottom")
+                  scale_y_continuous(expand=c(0,0),limits=c(0,230))+
+                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))
+
+ggarrange(aorist_pithoi_plot,aorist_pots_plot,nrow=2,labels=c("A","B"),align="v",common.legend=TRUE, legend="bottom")
 ```
@@
 And, despite differences in scale and timing, it is immediately apparent from Figure \@ref(fig:aoristbycem) that Fortetsa and the KNC underwent similar chronological developments with regard to burials and, to a slightly lesser degree, grave goods.
-In both cemeteries, the rate of cremation burial began rising around the end of the 9th century, plateaued slightly in the early 8th, and then rose again to a peak in the late 8th and early 7th.
+In both cemeteries, the rate of cremation burial began rising around the end of the 9th century BC, plateaued slightly in the early 8th, and then rose again to a peak in the late 8th and early 7th.
 Across all vessels, the developments are more volatile, but largely similar, especially regarding the late spike in burial activity.
@@
 
-```{r tombconstruction, fig.dim=c(7.5,4), dpi=600, fig.cap="Tomb construction through time at both cemeteries.
+```{r tombconstruction, fig.dim=c(7,4), dpi=600, fig.cap="Tomb construction through time at both cemeteries.
 On the left, counts are based on the ceramic phase considered most likely to correspond to the tomb's first use (including simply 'PG', 'G' and 'O' where no greater specificity was possible).
 On the right, aoristic sums are presented, spreading the uncertainty associated with those general assignations.
-In both cases, tombs lacking any securely dateable pottery are excluded.
-Abbreviations: SM - Subminoan; (E/M/L)PG - (Early/Middle/Late) Protogeometric; PGB - Protogeometric B; (E/M/L)G - (Early/Middle/Late) Geometric; (E/L)O - (Early/Late) Orientalising."}
+In both cases, tombs lacking any securely dateable pottery are excluded."}
 
@@
 
-new_tomb_counts_plot<-ggplot(data=subset(tombs,!is.na(start_period)),aes(start_period, fill=cem_condensed))+geom_bar(position=position_dodge2(preserve="single"))+theme_classic()+
+cbPalette <- c("#999999","#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+
+new_tomb_counts_plot<-ggplot(data=subset(tombs,!is.na(start_period)),aes(start_period, fill=forcats::fct_rev(tomb_type_condensed)))+geom_bar()+
+        theme_bw()+theme(legend.title=element_blank(),legend.spacing.x = unit(0.1,"cm"), text=element_text(family="serif", size=12),
+          axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),
+          panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
+            labs(x="Period", y="Count of New Tombs")+scale_fill_manual(values=cbPalette,
+                                                                                        labels=c("Unknown","Shaft Grave","Pit Cave",
+                                                                                        "Pithos Burial","Pit","Other","Larnax Burial","Chamber Tomb"))+
+             scale_y_continuous(expand=c(0,0),limits=c(0,25))+facet_wrap(~cem_condensed)+
+               theme(strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA))
+
+new_tomb_aorist_plot<-ggplot()+geom_line(data=aorist_tomb_periods_melt,aes(period,aoristic_sum,group=cemetery,linetype=cemetery))+
   theme_bw()+theme(legend.title=element_blank(),
           text=element_text(family="serif", size=12),
-          axis.text.x=element_text(angle=90, vjust=0.5, hjust=1))+
-            labs(x="Period", y="Count of New Tombs")+
-              scale_fill_manual(values =c("gray50","gray80"),labels=c(" KNC  "," Fortetsa"))+ 
-                 scale_y_continuous(expand=c(0,0),limits=c(0,30))
-
-new_tomb_aorist_plot<-ggplot(aorist_tomb_periods_melt,aes(period,aoristic_sum, fill=cemetery))+geom_col(position=position_dodge2(preserve="single"))+
-  theme_bw()+theme(legend.title=element_blank(),
-          text=element_text(family="serif", size=12),
-          axis.text.x=element_text(angle=90, vjust=0.5, hjust=1))+
+          axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),
+          panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
             labs(x="Period", y="Aoristic Sum of New Tombs")+
-               scale_fill_manual(values =c("gray50","gray80"),labels=c(" KNC  "," Fortetsa"))+ 
-                 scale_y_continuous(expand=c(0,0),limits=c(0,30))
-
-ggarrange(new_tomb_counts_plot,new_tomb_aorist_plot,ncol=2,common.legend=TRUE, legend="bottom")
+               scale_fill_manual(values =c("gray50","gray80"),labels=c(" KNC  "," FC"))+ 
+                 scale_y_continuous(expand=c(0,0),limits=c(0,28))+scale_x_discrete(expand=c(0,0))
+
+ggarrange(new_tomb_counts_plot,new_tomb_aorist_plot,ncol=2,align="h",labels=c("A","B"), widths=c(1.6,1),common.legend=FALSE, legend="bottom")
 
@@
 
-KNC_tombs_in_use<-colSums(KNC[6:15]!="")
-KNC_tombs_in_use<-data.frame(KNC_tombs_in_use)
+KNC_tombs_in_use<-data.frame(colSums(KNC[6:15]!=""))
 colnames(KNC_tombs_in_use)<-c("no_tombs")
@@
 
-FOR_tombs_in_use<-colSums(FOR[6:15]!="")
-FOR_tombs_in_use<-data.frame(FOR_tombs_in_use)
+FOR_tombs_in_use<-data.frame(colSums(FOR[6:15]!=""))
 colnames(FOR_tombs_in_use)<-c("no_tombs")
@@
 
-ggplot(tombs_in_use_cem,aes(period,no_tombs, fill=cemetery))+geom_col(position="dodge")+theme_classic()+
+tombs_in_use<-data.frame(colSums(tombs[6:15]!=""))
+colnames(tombs_in_use)<-c("no_tombs")
+setDT(tombs_in_use, keep.rownames = "period")
+tombs_in_use$period<-factor(tombs_in_use$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
+
+ggplot(tombs_in_use_cem,aes(period,no_tombs, linetype=cemetery,group=cemetery))+geom_line()+theme_classic()+
        theme_bw()+theme(legend.title=element_blank(),
-          text=element_text(family="serif", size=10),legend.position = "bottom", legend.margin=margin(0,0,0,0),legend.box.margin=margin(-10,-10,10,-10))+
+          text=element_text(family="serif", size=10),legend.position = "bottom", legend.margin=margin(0,0,0,0),legend.box.margin=margin(-10,-10,10,-10), axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),
+          panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
             labs(x="Period", y="Tombs in Use")+
-              scale_fill_manual(values =c("gray50","gray80"),labels=c(" KNC  "," Fortetsa"))+ 
-                 scale_y_continuous(expand=c(0,0),limits=c(0,45))
+              scale_y_continuous(expand=c(0,0),limits=c(0,45))+
+              scale_x_discrete(expand=c(0,0))
 ```
@@
 
-```{r tombsinusebybuilddate, fig.dim=c(7,4), dpi=600, fig.cap="Estimates of the number of tombs in use through time across both cemeteries.
+```{r tombsinusebybuilddate, fig.dim=c(7,5), dpi=600, fig.cap="Estimates of the number of tombs in use through time across both cemeteries.
 The grey background represents the total number of tombs in use, while the smaller coloured bars correspond to tombs built in each of the five centuries of EIA activity at the cemeteries."}
@@
 
-ggplot(data=tomb_use,aes(period,no_tombs))+geom_col(alpha=0.2)+
-            geom_col(data=tomb_use,aes(period,no_tombs,fill=tomb_century),position="dodge")+
-              theme_bw()+theme(text=element_text(family="serif", size=12))+
-                 labs(x="Period", y="Estimated No. of Tombs in Use")+
-                  scale_fill_brewer(name="Century Built",palette = "Set2")+ 
-                    scale_y_continuous(expand=c(0,0),limits=c(0,60))
+ggplot()+geom_area(data=tombs_in_use,aes(period,no_tombs,fill="All tombs",group=1),alpha=0.2)+
+            geom_line(data=tomb_use,aes(period,no_tombs,linetype=tomb_century,group=tomb_century))+
+              theme_bw()+theme(text=element_text(family="serif", size=12),
+          plot.margin=unit(c(0.5,1,0.5,1),"cm"),
+          panel.grid.major = element_blank(), panel.grid.minor = element_blank(),legend.position = "bottom")+
+                 labs(x="Period", y="Tombs in use",linetype="Century of tomb construction",fill="")+
+                  scale_fill_manual(values="#999999")+
+                    scale_y_continuous(expand=c(0,0),limits=c(0,60))+
+                      scale_x_discrete(expand=c(0,0))
 
@@
 
-```{r potsbydateandtombdate,  fig.dim=c(3,7),out.width="0.5%", dpi=600, fig.cap="Counts of vessels from all tombs in both cemeteries, arranged, above, by the century to which they are assigned based on their ceramic period and, below, by the centuries in which the tombs they come from were built."}
+```{r potsbydateandtombdate,  fig.dim=c(6,4),out.width="0.5%", dpi=600, fig.cap="Counts of vessels from all tombs in both cemeteries, arranged, above, by the century to which they are assigned based on their ceramic period and, below, by the centuries in which the tombs they come from were built."}
 
@@
 date_by_vessel<-ggplot(data=subset(pots,pots$century!=""),aes(century,fill="Date"))+geom_bar()+
-  theme_bw()+theme(legend.position="none",
+  theme_bw()+theme(legend.position="none", panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                text=element_text(family="serif", size=10),axis.text.y=element_text(angle=90, vjust=0, hjust=0.5))+
@@
 date_by_tomb<-ggplot(data=subset(pots,pots$century!=""),aes(tomb_century,fill="Date"))+geom_bar()+
-  theme_bw()+theme(legend.position="none",
+  theme_bw()+theme(legend.position="none", panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                text=element_text(family="serif", size=10),axis.text.y=element_text(angle=90, vjust=0, hjust=0.5))+
@@
 
-ggarrange(date_by_vessel,date_by_tomb,ncol=1)
+ggarrange(date_by_vessel,date_by_tomb,ncol=2,align="h",labels=c("A","B"))
 
@@
 And while these mostly remained around three per tomb per century, in the 8th and 7th centuries tombs founded in the 9th saw a rapid rise in the rate of burial, peaking at over 11 per tomb per century.
+A given that the cemeteries went out of use around 630 BC, this in reality amounts to a burial every six to seven years, compared to once a generation or so among most other tombs.
 Though no doubt rough estimates, the marked nature of this divergence suggests it is not simply an artefact of imperfect data.
@@
 But Table 1 also demonstrates the markedly low rates of burial which predominated at both cemeteries; two to four burials per tomb per century amounts to a very selective burial rite.
-It seems unlikely, then, that the rising rates of burial are, for instance, the direct result of population growth; the increase is too rapid, and concentrated only in older tombs, while, based on KULP’s survey data, the major population increase at EIA Knossos probably occurred back in the Protogeometric period (Whitelaw, pers.
+It seems unlikely, then, that the rising rates of burial are, for instance, the direct result of population growth; the increase is too rapid, and concentrated only in particular tombs, while, based on KULP’s survey data, the major population increase at EIA Knossos probably occurred back in the Protogeometric period (Whitelaw, pers.
 comm.).
@@
 
+```{r cluster, fig.dim=c(3.4,3.9),dpi=600,fig.align='center',out.width="0.5%", fig.cap="Results of cluster analysis based on aoristic sums per decade for each tomb.
+Ward's method was used and tombs that yielded no pots at all were excluded." }
+
+df_prep<-subset(tombs,total_pottery>0)
+df<-df_prep[-c(2:159,208:260)]
+row.names(df)<-df[,1]
+df[,1]<-NULL
+df<-scale(df)
+hc3 <- agnes(df, method = "ward")
+sub_grp <- cutree(hc3, k = 4)
+
+df_prep$cluster<-sub_grp
+df_prep2<-df_prep
+df_prep2 <- within(df_prep2, tomb[cluster<2] <- NA)
+
+cbPalette <- c("#E69F00","#999999", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+
+fviz_cluster(list(data = df, cluster = sub_grp),ellipse.alpha=0.1,pointsize=1,show.clust.cent=FALSE,main=NULL,xlab="Dimension 1",ylab="Dimension 2",
+             ggtheme=theme_bw()+theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank(),
+                                      text=element_text(family="serif", size=9),legend.position = "bottom"),
+             geom="point")+ geom_text_repel(label=df_prep2$tomb,color="Black",family="serif",size=2.5)+
+                scale_fill_manual('Cluster',values = cbPalette)+scale_color_manual('Cluster',values = cbPalette)+
+  scale_shape_manual('Cluster', values=c(15,16,17,18))
+
+```
+
 ```{r toptombs, fig.dim = c(7, 5), dpi=600, fig.align='center',fig.cap="Aoristic sums for the 12 tombs with the largest ceramic assemblages, arranged in decreasing order.
@@
 
-tombs_ranked<-tombs[with(tombs, order(-tombs$total_finds)), ]
-tombs_top_12<-head(tombs_ranked,12)
-top12<-as.vector(tombs_top_12$tomb)
-pots_top_12<-pots[is.element(pots$tomb, top12),]
-pots_top_12$tomb<-factor(pots_top_12$tomb,levels=top12)
-
-cols<-c(1,158:207)
-aorist_all_tomb_names<-tombs_top_12[,cols]
+tombs_ranked<-tombs[with(tombs, order(-tombs$total_pottery)), ]
+tombs_top_20<-head(tombs_ranked,20)
+top20<-as.vector(tombs_top_20$tomb)
+pots_top_20<-pots[is.element(pots$tomb, top20),]
+pots_top_20$tomb<-factor(pots_top_20$tomb,levels=top20)
+
+cols<-c(1,160:209)
+aorist_all_tomb_names<-tombs_top_20[,cols]
 aorist_all_tomb_names<-data.frame(aorist_all_tomb_names)
@@
 
-aorist_all_tomb_names$cemetery<-tombs_top_12$cem_condensed
+aorist_all_tomb_names$cemetery<-tombs_top_20$cem_condensed
 
@@
 
-top_12<-tombs_top_12$tomb
-aorist_all_tomb_melt$tomb<-factor(aorist_all_tomb_melt$tomb,levels=top_12)
-
-ggplot(aorist_all_tomb_melt,aes(variable, value,fill=cemetery))+geom_col()+theme(legend.position = "none")+facet_wrap(~tomb,scales="free_y")+
+top_20<-tombs_top_20$tomb
+aorist_all_tomb_melt$tomb<-factor(aorist_all_tomb_melt$tomb,levels=top_20)
+
+ggplot(aorist_all_tomb_melt,aes(variable, value,fill=cemetery))+geom_col()+theme(legend.position = "none")+
   geom_col()+theme_bw()+theme(legend.title=element_blank(),legend.position="bottom",
-      text=element_text(family="serif", size=12),
-      axis.text.x=element_text(angle=45, vjust=1.2, hjust=1), legend.margin=margin(0,0,0,0),legend.box.margin=margin(-10,-10,10,-10))+
+      text=element_text(family="serif", size=11), panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+      axis.text.x=element_text(angle=45, vjust=1.2, hjust=1), legend.margin=margin(0.5,0,0,0),legend.box.margin=margin(1,-10,1,-10))+
           labs(x="Date (Years BC)", y="Aoristic Sums")+
-              scale_fill_manual(values = c("gray50","gray80"),labels=c(" KNC  "," Fortetsa"))+ 
-                  coord_cartesian(xlim=c(1100,610))+scale_y_continuous()+
+              scale_fill_manual(values = c("gray80","gray50"),labels=c(" KNC  "," Fortetsa"))+ 
+                  coord_cartesian(xlim=c(1100,610))+scale_y_continuous(expand = expansion(mult = c(0, 0.05)))+
                     scale_x_reverse(expand = c(0, 0),limits=c(1100,610))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))+facet_wrap(~tomb,scales="free_y")+
+                       theme(strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA),
+                             strip.text.x = element_text(size=8,margin = margin(0.05,0,0.05,0, "cm")))
 
@@
 
-```{r groupsab, fig.dim=c(8,4),dpi=600, fig.align='center',fig.cap="Aoristic sums for the Group A and Group B tombs, as defined above."}
-
-tombs$group<-ifelse(tombs$tomb=="P"|tombs$tomb=="292"|tombs$tomb=="75"|tombs$tomb=="107"|tombs$tomb=="285"|tombs$tomb=="II"|tombs$tomb=="218","A",
-                   ifelse(tombs$tomb=="X"|tombs$tomb=="G"|tombs$tomb=="219"|tombs$tomb=="Q"|tombs$tomb=="283","B","C"))
-
-GroupA<-subset(tombs,group=="A")
-
-aorist_GroupA <- colSums(GroupA[158:207])
+```{r groupsab, fig.dim=c(6,6.5),dpi=600, fig.align='center', fig.cap="Aoristic sums for the Group A and Group B tombs, as defined above."}
+
+GroupA<-subset(df_prep,cluster==3)
+
+aorist_GroupA <- colSums(GroupA[160:209])
 aorist_GroupA<-data.frame(aorist_GroupA)
@@
 aorist_GroupA$decade<-as.numeric(aorist_GroupA$decade)
-
-aorist_GroupA$group<-"A"
-
-GroupB<-subset(tombs,tombs$group=="B")
-
-aorist_GroupB <-colSums(GroupB[158:207])
+aorist_GroupA$no_tombs<-colSums(GroupA[160:209]!=0)
+
+aorist_GroupA$sum_over_tombs<-ifelse(is.na(aorist_GroupA$aoristic_sum/aorist_GroupA$no_tombs),0,
+                                     (aorist_GroupA$aoristic_sum/aorist_GroupA$no_tombs))
+
+aorist_GroupA$group<-"Group A"
+
+GroupB<-subset(df_prep,cluster==4)
+
+aorist_GroupB <-colSums(GroupB[160:209])
 aorist_GroupB<-data.frame(aorist_GroupB)
@@
 aorist_GroupB$decade<-as.numeric(aorist_GroupB$decade)
-
-aorist_GroupB$group<-"B"
-
-aorist_groups<-rbind(aorist_GroupA,aorist_GroupB)
-
-ggplot(aorist_groups,aes(decade,aoristic_sum,fill=group))+geom_col(position="dodge")+
-  theme_bw()+theme_bw()+theme(text=element_text(family="serif", size=12))+
-                labs(x="Date (Years BC)", y="Aoristic Sum")+
-                  scale_fill_manual(values=c("gray50","gray80"),name="Group")+
-                   scale_y_continuous(expand=c(0,0),limits=c(0,120))+
-                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-```
+aorist_GroupB$no_tombs<-colSums(GroupB[160:209]!=0)
+
+aorist_GroupB$sum_over_tombs<-ifelse(is.na(aorist_GroupB$aoristic_sum/aorist_GroupB$no_tombs),0,
+                                     (aorist_GroupB$aoristic_sum/aorist_GroupB$no_tombs))
+
+aorist_GroupB$group<-"Group B"
+
+Other<-subset(df_prep,cluster==1|cluster==2)
+
+aorist_Other <-colSums(Other[160:209])
+aorist_Other<-data.frame(aorist_Other)
+
+colnames(aorist_Other)<-c("aoristic_sum")
+row.names(aorist_Other)<-dates
+aorist_Other<-data.frame(aorist_Other)
+setDT(aorist_Other, keep.rownames = "decade")
+aorist_Other$decade<-as.numeric(aorist_Other$decade)
+aorist_Other$no_tombs<-colSums(Other[160:209]!=0)
+
+aorist_Other$sum_over_tombs<-ifelse(is.na(aorist_Other$aoristic_sum/aorist_Other$no_tombs),0,
+                                     (aorist_Other$aoristic_sum/aorist_Other$no_tombs))
+
+aorist_Other$group<-"Other"
+
+aorist_groups<-rbind(aorist_GroupA,aorist_GroupB,aorist_Other)
+
+unscaled<-ggplot()+  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf), fill=NA, colour="grey90",alpha = 0.4) +
+  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = 118, ymax = 126), fill=NA, colour="black",alpha = 0.4) +
+  geom_text(data=rects,aes(xmid, 122,label=col),family="serif",size=2.5)+
+  geom_line(data=aorist_groups,aes(decade,aoristic_sum,linetype=group,group=group))+
+  theme_bw()+theme_bw()+theme(text=element_text(family="serif", size=10),
+                              panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+                              plot.margin=unit(c(0.1,1,0.5,1),"cm"),
+                              legend.title = element_blank())+
+    labs(x="Date (Years BC)", y="Aoristic sum/No. of tombs")+
+  scale_y_continuous(expand=c(0,0),limits=c(0,126))+
+  scale_x_reverse(expand = c(0, 0),limits=c(1100,600))
+
+scaled<-ggplot()+  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf), fill=NA, colour="grey90",alpha = 0.4) +
+  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = 13.3, ymax = 14), fill=NA, colour="black",alpha = 0.4) +
+  geom_text(data=rects,aes(xmid, 13.65,label=col),family="serif",size=2.5)+
+  geom_line(data=aorist_groups,aes(decade,sum_over_tombs,linetype=group,group=group))+
+  theme_bw()+theme_bw()+theme(text=element_text(family="serif", size=10),
+                              panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+                              plot.margin=unit(c(0.1,1,0.5,1),"cm"),
+                              legend.title = element_blank())+
+    labs(x="Date (Years BC)", y="Aoristic sum/No. of tombs")+
+  scale_y_continuous(expand=c(0,0),limits=c(0,14))+
+  scale_x_reverse(expand = c(0, 0),limits=c(1100,600))
+
+ggarrange(unscaled,scaled,nrow=2,labels=c("A","B"),align="v",common.legend=TRUE, legend="bottom")
+
+```
+
+```{r table2}
+
+pithoi$cluster <- df_prep$cluster[match(pithoi$tomb,df_prep$tomb)]
+pithoi$cluster[is.na(pithoi$cluster)]<- "Other"
+pithoi$cluster[pithoi$cluster<3]<- "Other"
+pithoi$cluster[pithoi$cluster==3]<- "Group A"
+pithoi$cluster[pithoi$cluster==4]<- "Group B"
+
+tombs$cluster <- df_prep$cluster[match(tombs$tomb,df_prep$tomb)]
+tombs$cluster[is.na(tombs$cluster)]<- "Other"
+tombs$cluster[tombs$cluster<3]<- "Other"
+tombs$cluster[tombs$cluster==3]<- "Group A"
+tombs$cluster[tombs$cluster==4]<- "Group B"
+
+potsA10th<-sum(pithoi$cluster=="Group A" & pithoi$century=="10th",na.rm=TRUE)
+potsA9th<-sum(pithoi$cluster=="Group A" & pithoi$century=="9th",na.rm=TRUE)
+potsA8th<-sum(pithoi$cluster=="Group A" & pithoi$century=="8th",na.rm=TRUE)
+potsA7th<-sum(pithoi$cluster=="Group A" & pithoi$century=="7th",na.rm=TRUE)
+
+notombsA10th<-nrow(unique(subset(pithoi,cluster=="Group A" & century=="10th",select=tomb)))
+notombsA9th<-nrow(unique(subset(pithoi,cluster=="Group A" & century=="9th",select=tomb)))
+notombsA8th<-nrow(unique(subset(pithoi,cluster=="Group A" & century=="8th",select=tomb)))
+notombsA7th<-nrow(unique(subset(pithoi,cluster=="Group A" & century=="7th",select=tomb)))
+
+potsB10th<-sum(pithoi$cluster=="Group B" & pithoi$century=="10th",na.rm=TRUE)
+potsB9th<-sum(pithoi$cluster=="Group B" & pithoi$century=="9th",na.rm=TRUE)
+potsB8th<-sum(pithoi$cluster=="Group B" & pithoi$century=="8th",na.rm=TRUE)
+potsB7th<-sum(pithoi$cluster=="Group B" & pithoi$century=="7th",na.rm=TRUE)
+
+notombsB10th<-nrow(unique(subset(pithoi,cluster=="Group B" & century=="10th",select=tomb)))
+notombsB9th<-nrow(unique(subset(pithoi,cluster=="Group B" & century=="9th",select=tomb)))
+notombsB8th<-nrow(unique(subset(pithoi,cluster=="Group B" & century=="8th",select=tomb)))
+notombsB7th<-nrow(unique(subset(pithoi,cluster=="Group B" & century=="7th",select=tomb)))
+
+potsother10th<-sum(pithoi$cluster=="Other" & pithoi$century=="10th",na.rm=TRUE)
+potsother9th<-sum(pithoi$cluster=="Other" & pithoi$century=="9th",na.rm=TRUE)
+potsother8th<-sum(pithoi$cluster=="Other" & pithoi$century=="8th",na.rm=TRUE)
+potsother7th<-sum(pithoi$cluster=="Other" & pithoi$century=="7th",na.rm=TRUE)
+
+notombsother10th<-nrow(unique(subset(pithoi,cluster=="Other" & century=="10th",select=tomb)))
+notombsother9th<-nrow(unique(subset(pithoi,cluster=="Other" & century=="9th",select=tomb)))
+notombsother8th<-nrow(unique(subset(pithoi,cluster=="Other" & century=="8th",select=tomb)))
+notombsother7th<-nrow(unique(subset(pithoi,cluster=="Other" & century=="7th",select=tomb)))
+
+```
+
+|**Tombs**          | **Date of pithoi (burials)**| **No. of pithoi (burials)**| **Number of tombs represented** | **Burials per tomb**                   |
+|-------------------|--------------------------|-------------------------|-----------------------------|--------------------------------------------------|
+|*Group A*|                         |                         |                             |                                                  |
+|                   |10th century              | `r potsA10th`        | `r notombsA10th`         | `r 0` |
+|                   |9th century               | `r potsA9th`         | `r notombsA9th`          | `r round(potsA9th/notombsA9th,digits=2)`   |
+|                   |8th century               | `r potsA8th`         | `r notombsA8th`          | `r round(potsA8th/notombsA8th,digits=2)`   |
+|                   |7th century               | `r potsA7th`         | `r notombsA7th`          | `r round(potsA7th/notombsA7th,digits=2)`   |
+|*Group B*|                          |                         |                             |                                                  |
+|                   |10th century              | `r potsB10th`        | `r notombsB10th`         | `r round(potsB10th/notombsB10th,digits=2)` |
+|                   |9th century               | `r potsB9th`          | `r notombsB9th`           | `r round(potsB9th/notombsB9th,digits=2)`     |
+|                   |8th century               | `r potsB8th`          | `r notombsB8th`           | `r round(potsB8th/notombsB8th,digits=2)`     |
+|                   |7th century               | `r potsB7th`          | `r notombsB7th`           | `r round(potsB7th/notombsB7th,digits=2)`     |
+|*Other tombs*|                          |                         |                             |                                                  |
+|                   |10th century              | `r potsother10th`        | `r notombsother10th`         | `r round(potsother10th/notombsother10th,digits=2)` |
+|                   |9th century               | `r potsother9th`          | `r notombsother9th`           | `r round(potsother9th/notombsother9th,digits=2)`     |
+|                   |8th century               | `r potsother8th`          | `r notombsother8th`           | `r round(potsother8th/notombsother8th,digits=2)`     |
+|                   |7th century               | `r potsother7th`          | `r notombsother7th`           | `r round(potsother7th/notombsother7th,digits=2)`     |
 
@@
 
-```{r dimensionsboxplots, fig.dim=c(7,7),dpi=600, fig.align='center',fig.cap="Boxplots comparing various tomb dimensions, divided, in each case, by the century of tomb construction."}
+```{r dimensionsboxplots, fig.dim=c(6,5.9),dpi=600, fig.align='center',fig.cap="Boxplots comparing various tomb dimensions, divided, in each case, by the century of tomb construction."}
 
 tombs$century_built<-factor(tombs$century_built,levels=c("11th","10th","9th","8th","7th"))
-dromwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_width,fill="gray80"))+geom_boxplot(lwd=0.2,outlier.size=0.8)+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="serif", size=10))+
+dromwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_width_max,fill="gray80"))+geom_boxplot(lwd=0.2,outlier.size=0.8)+
+  theme_bw()+theme(legend.position="none", panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+               text=element_text(family="serif", size=9))+
                  scale_fill_manual(values=c("gray80"))+
@@
 dromlength<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,dromos_length,fill="gray80"))+geom_boxplot(lwd=0.2,outlier.size=0.8)+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="serif", size=10))+
+  theme_bw()+theme(legend.position="none", panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+               text=element_text(family="serif", size=9))+
   scale_fill_manual(values=c("gray80"))+
@@
 stomwidth<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,stomion_width,fill="gray80"))+geom_boxplot(lwd=0.2,outlier.size=0.8)+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="serif", size=10))+
+  theme_bw()+theme(legend.position="none", panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+               text=element_text(family="serif", size=9))+
   scale_fill_manual(values=c("gray80"))+
@@
 chamarea<-ggplot(data=subset(tombs, !is.na(century_built)),aes(century_built,chamber_area,fill="gray80"))+geom_boxplot(lwd=0.2,outlier.size=0.8)+
-  theme_bw()+theme(legend.position="none",
-               text=element_text(family="serif", size=10))+
+  theme_bw()+theme(legend.position="none", panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+               text=element_text(family="serif", size=9))+
   scale_fill_manual(values=c("gray80"))+
@@
 
-```{r tombgroupdimensions, fig.dim = c(3.5, 3.5),out.width="10%", dpi=600, fig.align='center', fig.cap="Scatterplot comparing the dimensions of *dromos* length and chamber area.
+```{r tombgroupdimensions, fig.dim = c(7, 4),out.width="10%", dpi=600, fig.align='center', fig.cap="Scatterplot comparing the dimensions of *dromos* length and chamber area.
 Groups A and B, defined earlier, are illustrated and labelled in green and orange respectively, while all remaining tombs ('Group C') are in blue."}
 
-ggplot(subset(tombs,century_built=="9th" | century_built=="10th"),aes(dromos_length,chamber_area,color=group,label=tomb))+geom_point(size=1)+
-  theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom", legend.margin=margin(0,0,0,0),legend.box.margin=margin(-10,-10,10,-10))+
-  geom_label_repel(data=subset(tombs,group!="C"),aes(fill=group),color="white",show.legend = FALSE,family="serif", size=3)+
-            labs(x="Dromos Length (m)", y="Chamber Area (sq m)")+
-               scale_color_brewer(palette = "Set2",name="Group")+
-               scale_fill_brewer(palette = "Set2")+
-                scale_x_continuous(expand=c(0,0),limits=c(0,8))+
+cbPalette <- c("#56B4E9", "#009E73","#E69F00")
+
+dim_by_cent<-ggplot(subset(tombs,century_built!="7th" & !is.na(tombs$century_built)), aes(dromos_area,chamber_area))+geom_point(size=1)+
+    theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",
+                     panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+                     legend.margin=margin(0,0,0,0),legend.box.margin=margin(-10,-10,10,-10))+
+            labs(x="Dromos area (sq. m)", y="Chamber Area (sq. m)")+
+                             scale_x_continuous(expand=c(0,0),limits=c(0,12))+
                   scale_y_continuous(expand=c(0,0),limits=c(0,10))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
+                      theme(plot.margin=unit(c(0,0.1,0.2,0.5),"cm"))+
+  stat_cor(method="pearson",label.x=6,label.y=8,family="serif",size=3,label.sep="\n")+facet_wrap(~century_built)+
+                       theme(strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA),
+                             strip.text.x = element_text(size=8))
+
+dim_all<-ggplot(tombs,aes(dromos_area,chamber_area,color=cluster))+geom_point(size=1)+
+    theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
+                     panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+                     legend.margin=margin(0.8,0,0,0),legend.box.margin=margin(-10,-10,10,-10))+
+            labs(x="Dromos area (sq. m)", y="Chamber Area (sq. m)")+
+                        scale_color_manual(values=cbPalette)+
+                             scale_x_continuous(expand=c(0,0),limits=c(0,12))+
+                  scale_y_continuous(expand=c(0,0),limits=c(0,10))+
+                      theme(plot.margin=unit(c(0.55,0.3,0.2,0.2),"cm"))
+
+ggarrange(dim_all,dim_by_cent,ncol=2,labels=c("A","B"))
 
@@
 ```{r tombdimensions, fig.dim = c(7.5, 4), dpi=600, fig.align='center', fig.cap="Scatterplots comparing tomb dimensions (chamber area and *dromos* length) to the estimated total number of burials by tomb.
-The points are colour-coded as per Figure 11."}
-
-tombareas<-ggplot(subset(tombs,tombs$est_burials>1),aes(chamber_area,est_burials,color=group))+geom_point()+
-  theme_bw()+theme(text=element_text(family="serif", size=11))+
-            labs(x="Chamber Area (sq m)", y="Estimated Total Burials")+
-               scale_color_brewer(palette = "Set2",name="Group")+
+The points are colour-coded as per Figure 12."}
+
+tombareas<-ggplot(subset(tombs,tombs$est_burials>1),aes(chamber_area,est_burials,color=cluster))+geom_point()+
+  theme_bw()+theme(text=element_text(family="serif", size=11),
+                   panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+                   legend.title = element_blank())+
+            labs(x="Chamber Area (sq. m)", y="Estimated Total Burials")+
+               scale_color_manual(values=cbPalette)+
                 scale_x_continuous(expand=c(0,0),limits=c(0,10))+
                   scale_y_continuous(expand=c(0,0),limits=c(0,50))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-dromoslengths<-ggplot(subset(tombs,tombs$est_burials>1),aes(dromos_length,est_burials,color=group))+geom_point()+
-  theme_bw()+theme(text=element_text(family="serif", size=11))+
-            labs(x="Dromos Length (m)", y="Estimated Total Burials")+
-               scale_color_brewer(palette = "Set2",name="Group")+
+                      theme(plot.margin=unit(c(0.1,0.2,0.1,0.1),"cm"))
+
+dromosareas<-ggplot(subset(tombs,tombs$est_burials>1),aes(dromos_area,est_burials,color=cluster))+geom_point()+
+  theme_bw()+theme(text=element_text(family="serif", size=11),
+                   panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+                   legend.title = element_blank())+
+            labs(x="Dromos Area (sq. m)", y="Estimated Total Burials")+
+              scale_color_manual(values=cbPalette)+
                 scale_x_continuous(expand=c(0,0),limits=c(0,10))+
                   scale_y_continuous(expand=c(0,0),limits=c(0,50))+
-                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))
-
-ggarrange(tombareas,dromoslengths,ncol=2,common.legend = TRUE,legend="bottom")
+                      theme(plot.margin=unit(c(0.1,0.2,0.1,0.1),"cm"))
+
+ggarrange(tombareas,dromosareas,ncol=2,common.legend = TRUE,legend="bottom")
 
@@
 Many of the very largest tombs are middling when it comes to burial numbers, and vice versa.
-So whatever the scale of the tombs connoted when built, this proved irrelevant to their later reuse.
+So whatever the scale of the tombs connoted when built, this proved largely irrelevant to their later reuse.
 Several had niches cut into their *dromoi*, or the *stomion* extended, to fit in later burial urns (including T.
@@
 
-```{r tombscompthroughtime, fig.dim=c(7,4),dpi=600,fig.align='center',fig.cap="Counts of vessels dating to the 9th, 8th and 7th centuries respectively, found in tombs built in the 10th or 9th century which yielded upwards of 50 vessels when excavated."}
+```{r tombscompthroughtime, fig.dim=c(7,4),dpi=600,fig.align='center',fig.cap="Counts of vessels dating to the 9th, 8th and 7th centuries respectively, found in tombs built in the 10th or 9th century which yielded upwards of 50 vessels when excavated.
+The tombs are ordered left to right by total excavated vessels, and the diminishing number of tombs in each plot reflects those falling into disuse in the 8th and 7th centuries."}
 
@@
   theme_bw()+theme(legend.position="bottom",text=element_text(family="serif", size=10),
+                   panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                axis.text.x = element_text(angle = 90,hjust=1,vjust=0.4), legend.margin=margin(0,0,0,0),legend.box.margin=margin(-10,-10,10,-10))+
@@
 Group B, meanwhile, experienced similar, if not greater, use in the 9th century, and then gradually declined.
-That is, these tombs seem to have followed a more conventional trajectory viz.
-Figure \@ref(fig:tombsinusebybuilddate).
+That is, these tombs seem to have followed a more conventional trajectory vis-à-vis Figure \@ref(fig:tombsinusebybuilddate).
 What might account for this disparity?
@@
 
-aorist_tomb_values<-tombs[c(1,158:207)]
+aorist_tomb_values<-tombs[c(1,160:209)]
 aorist_tomb_values[aorist_tomb_values == 0] <- NA
@@
 
-ggplot(aorist_melt,aes(variable,value,fill=period))+geom_boxplot(aes(group = cut_width(variable, 10)),outlier.size=0.9,lwd=0.3)+theme_bw()+
+ggplot(aorist_melt,aes(variable,value,fill=period))+geom_boxplot(aes(group = cut_width(variable, 10)),outlier.size=0.8,lwd=0.3)+theme_bw()+
       theme(text=element_text(family="serif", size=10))+
@@
 
-This is reflected further in the changing composition of the mortuary assemblage through time (Figure \@ref(fig:comps)).
+This is reflected further in the changing composition of the mortuary assemblages through time (Figure \@ref(fig:comps)).
 Across both cemeteries, there appears to have been the steady crystallisation of a basic 'burial kit', comprising a pithos and lid, an oil vessel or two, a jug, and a drinking vessel.
@@
 
-Whether the increasingly homogenous burials reflected some standard set of accoutrements intentionally shared by the hypothetical *agathoi*, or else nascent, egalitarian strictures over elaborate burial remains difficult to judge.
+Whether the increasingly homogenous rites reflected some standard set of accoutrements intentionally shared by the hypothetical *agathoi*, or else nascent, egalitarian strictures over elaborate burial remains difficult to judge.
 Yet we need not adopt wholesale a model like Morris's "middling" ideology, nor link the diminishing range of burial goods to Archaic Crete's purported austerity.
 As a more nuanced picture emerges of the island's relationship with the wider Mediterranean, subtler characterisations are possible.
-It is telling that the peak for imported items came in the 9th century; the Orientalising as a time period feels increasingly misapplied to Crete [@Whitley2004: 434], when, at Knossos at least, it witnessed far lower rates of 'Oriental' imports, and only limited adoption of the figural motifs which characterise the style's appearance elsewhere.
-
-The decline in imports was, however, counterpointed by a rise in imitations of Near Eastern vessels [@Antoniadis2012: 170], according with our growing awareness of the selective nature of Archaic and Classical Crete's external relations.
-Erickson [@Erickson2010], similarly, has argued that *symposia* were practised in Crete, something often disputed, but that their practitioners merely favoured simpler vessel forms than the figurally-decorated wares popular on the mainland.
+It is telling that the peak for imported items came in the 9th century; the Orientalising as a period label feels increasingly misapplied to Crete [@Whitley2004: 434], when, at Knossos at least, it witnessed a diminution in 'Oriental' imports, and only limited adoption of the figural motifs which characterise the style's appearance elsewhere.
+
+The decline in imports was, however, counterpointed by a rise in imitations of Near Eastern vessels [@Antoniadis2012: 170], something which accords with our growing awareness of the selective nature of Archaic and Classical Crete's external relations.
+Erickson [@Erickson2010], similarly, has argued that *symposia* were practised in Crete - something often disputed - but that their practitioners merely favoured simpler vessel forms than the figurally-decorated wares popular on the mainland.
 Local, selective adoption of, and responses to, wider social and material trends is perhaps becoming the hallmark of Archaic Crete, and finds further endorsement in the EIA cemeteries at Knossos.
@@
 There have also been a number of tombs discovered in the last 20 years that fill in the spaces between the Teke plot and the Medical Faculty Site, and ideally these unpublished tombs should be included in further analyses as they, along with the Teke tholos and associated tombs, are probably part of a single contiguous cemetery [@Blackman1999: 113; @Catling1983: 51; @French1991: 68-69].
-Finally, a preliminary study was recently undertaken by the author of select boxes marked as 'unpublished' from the KNC deposits kept at the British School at Athens, at Knossos.
-Despite little mention in the publication, this material was not inconsiderable and so, regretably, future work will be necessary to determine whether it threatens the integrity of these analyses.
-At present, it is a pinch of salt with which to take the present findings.
 

### 2021-05-21 README.Rmd
Last edits

@@
 
-> Pollard, Dominic (2020). _`r Title`_. Name of journal/book, <https://doi.org/xxx/xxx>
+> Pollard, Dominic (2021). _`r Title`_. Name of journal/book, <https://doi.org/xxx/xxx>
 
@@
 
-> Pollard, Dominic (2020). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Pollard, Dominic (2021). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
 
@@
 ```
+
+The compendium is also available from figshare at: https://doi.org/10.6084/m9.figshare.13019588.v1
+
 ### Licenses

### 2021-05-21 analysis/paper/Knossos_Cemeteries_Article.Rmd
Last edits

---
title: "All Equal in the Presence of Death? A Quantitative Analysis of the Early Iron Age Cemeteries of Knossos, Crete"
author:
  - Dominic Pollard
  - Institute of Archaeology, University College London, 31-34 Gordon Square, Bloomsbury, London WC1H 0PY, United Kingdom
  - dominic.pollard.16@ucl.ac.uk
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/articletemplate.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-anthropological-archaeology.csl" # Insert path for the bib-style

abstract: 
   "The Early Iron Age cemeteries of Knossos, Crete, represent the richest source of archaeological material at the site for the period between the end of the Bronze Age and the emergence of the Archaic *polis*, or city-state. However, despite well-published excavation reports on the two largest burial grounds and studies dealing with various artefacts and individual tombs therein, no substantial comparative, quantitative analysis has yet been published. Such an analysis is provided here, examining both the synchronic and diachronic variation among the tombs of these two cemeteries. Through a range of quantitative techniques, this study isolates two significant patterns of burial among the largest tombs at both cemeteries and considers the broader social contexts within which these were constructed and used. The relationship between the patrons of these tombs and the development of the city-state of Knossos is also discussed. This paper demonstrates the potential for quantitative analyses to contribute to nuanced, context-specific theories of social practice, and advocates their wider application to the archaeological record of the Cretan Early Iron Age. It also serves as a demonstration of maximal research reproducibility, with the full datasets and code underpinning the analysis made freely available to download and use."

---
# Keywords: 
Early Iron Age; Knossos; Burial; Crete; Quantitative Methods; Open-Access

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dev="jpeg"
)

#The appropriate packages are installed. Extrafont allows the use of non-standard fonts in R plots, ggplot2 is a powerful figure and graph plotting tool, and RColorBrewer includes a number of colour pallettes which can be applied to plots. The colour scheme in this paper is chosen from those that are suitable for colour-blind readers.
library(ggpubr)
library(ineq)
library(reshape2)
library(data.table)
library(ggplot2)
library(RColorBrewer)
library(grid)
library(dplyr)
library(plyr)
library(ggforce)
library(jpeg)
library(ggrepel)
library(forcats)
library(factoextra)
library(cluster)

```

##### pagebreak

# 1. Introduction
This paper presents a novel quantitative analysis of the Early Iron Age (EIA) cemeteries of Knossos, Crete. Within this period, these cemeteries represent not only the richest source of archaeological material at the site, but also the largest published mortuary assemblage across the entire island. However, despite well-published excavation reports on the two largest Knossian burial grounds, and a number of studies dealing with various artefacts and individual tombs therein, no substantial comparative, quantitative analysis of these cemeteries has yet been published.

## 1.1 Early Iron Age Research in Crete
The EIA of Greece has witnessed a dramatic growth in archaeological interest in recent decades. This era is bookended by the collapse of the Late Bronze Age (LBA) palace societies of the Aegean – the so-called Minoan and Mycenaean cultures of Crete and mainland Greece respectively – and the emergence of the nascent Greek city-states or *poleis* [@Dickinson2006; @Lemos2020; for Crete, see @Wallace2010a]. This corresponds broadly to the 11th through 8th centuries BCE, though use of the term varies, and here it is applied to all periods from the Subminoan to the Orientalising (see Table 1 for chronology and abbreviations used throughout).

The EIA was once often termed the ‘Greek Dark Ages’ [@Bintliff2012 209-213; @Morris1997], a time of assumed instability and hardship following the loss of literate, state administration at the end of the Bronze Age (BA), with accompanying declines in population, technology and craftsmanship, and inter-regional connectivity. In turn, signs of increasing settlement size, flourishing seaborne trade, and the growing breadth and refinement of artistic and ceramic styles in the 8th century were taken to herald a ‘Greek Renaissance’, presaging the development of the numerous and diverse city-states of the Classical period [@Kotsonas2020, 79; @Ure1921].

Though not challenging this characterisation as such, the pioneering works of Coldstream [-@Coldstream1977], Desborough [-@Desborough1972], and Snodgrass [-@Snodgrass1971a] nonetheless marked the start of a shift in focus for research on the EIA. For one, these scholars prioritised archaeological assemblages at a time when reasoning back from later historical and Homeric sources was the norm, but more importantly they asserted, particularly in the case of Snodgrass, that Greece's EIA communities were descended directly from their LBA predecessors and that, in turn, understanding these groups was essential for theorising the development of the later *poleis*.

Interest in the period grew in subsequent decades, as excavation at sites like the Toumba cemetery at Lefkandi in Euboea [@Desborough1970], and the North Cemetery at Knossos - discussed in this paper - yielded evidence for the maintenance of long-distance networks of exchange and interaction at the dawn of the EIA, as well as artefacts and rituals redolent of the Homeric epics [@Catling1995]. The growth of archaeological survey, with its broad chronological purview, began revealing long-term settlement dynamics in many regions, often transcending the traditional divides of the LBA, EIA and Classical era. Meanwhile a new generation of scholars in the 1980s and 1990s, particularly under the supervision of Snodgrass, pursued studies in burial practice, social organisation, the emergence of sanctuaries, and civic identities in the EIA [@Morgan1990; @Morgan2003; @Morris1987; @Whitley1991a]. All this work has culminated in the active rejection by many scholars of the appellation and interpretive baggage of the so-called ‘Dark Ages’ [see @Kotsonas2016; @Morris1997], and excavation and survey work pertaining to the EIA continue to grow apace [e.g. on Crete: @Haggis2016; @Zographaki2013].

Crete has had a complex relationship with these broader trends. Here, the EIA was long eclipsed not so much by the Classical era as the Minoan Bronze Age, from which it was seen to mark a severe decline in political complexity and cultural achievement. In turn, the representation of the island in ancient Greek texts as archaising, idiosyncratic and parochial – not to mention, piratical – has seen it generally side-lined from discussions of the political development of Greece in the first millennium [@Prent2005 69-70; @Whitley2009 273-74]. Indeed, in Lemos’ [-@Lemos2002 1] overview of Protogeometric Greece, the island is excluded altogether as 'its character . . . is different from most of the rest of the Aegean’.

These differences include an early and notable shift in settlement at the end of the BA towards inland, defensible locations [@Nowicki2000]; mixed populations which may at various points in the EIA have included native Cretans, mainlanders of various origins, Phoenicians, and even groups from the Central Mediterranean [@DAgata2012; @Kanta2003; @Stampolidis2006]; a precocious engagement with Near Eastern material culture, artistic styles and perhaps even modes of mortuary and religious practice [@Morris1992 110; @Stampolidis1990]; the worship of local deities into the Classical period, as well as specifically Cretan forms of several of the Olympians [@Prent2005]; and a number of city-states codifying and inscribing law codes at an early date, which detail constitutions in some ways akin, but in others strikingly different, to accounts of Spartan law [@Gagarin2016]. Rather than being dismissed as atypical, such features have much to offer to the development of alternative, decentred accounts of early Greek state-formation, approaches to which have been historically Atheno-centric.

## 1.2 The Present Study
There are, given the history of research outlined above, limitations in the data available and the methods so far employed in the study of the Cretan EIA, and this article aims to confront certain of these head-on. Excavated mortuary contexts far outstrip those of settlements for the period, and yet, despite the volume of material, these have seldom been submitted to statistical analyses or even proper quantification [see @Kotsonas2011]. The Fortetsa and North Cemetery complexes, the two largest EIA cemeteries at Knossos, are a case in point. Both are well-published [@Brock1957; @Coldstream1996a] and have been the subject of multiple studies concerning particular tombs, object classes, and even single artefacts, but a lack of systematic, quantitative comparisons between them has left us with many intuited trends, some exceptional tombs, and yet a general impression of an un-patterned multiplicity in burial practices.

The city and palace of Knossos, first excavated well over a century ago [@Evans1921; @Evans1928; @Evans1930], retain a pre-eminent position in popular and academic accounts of the island’s Bronze Age, but the city remains of great importance in the subsequent EIA, being one of only two sites to have boasted a Minoan palace and later to have become a *polis*, with an uninterrupted sequence of occupation in between [@Cucuzza2013 31]. It can therefore offer significant insights into not only the processes of resilience, consolidation, and redevelopment in the aftermath of state collapse, but also into what has been termed the role of ‘the past in the present’, that is, the encounters between the later Knossian community and its BA heritage. The ruined palace remained visible for centuries after its destruction, while BA artefacts have repeatedly been found in EIA contexts – particularly tombs [@Cucuzza2013; @Crowe2018]. And of course, though the myths of King Minos are now afforded no place in the real history of the site, those myths were part of the identity of the Classical Greek community, as evidenced, for instance, in the minting of coins depicting the labyrinth [@Mielczarek2013]. Finally, the present analysis has implications for understanding the relationship between burial rites and wider socio-political developments, particularly in the context of state-formation, something doubtless pertinent to other periods and regions.

Utilising a dataset including information on every find and tomb recorded in the Fortetsa and Knossos North Cemetery publications [@Brock1957; @Coldstream1996a], I here aim to demonstrate the potential for quantitative methods, firstly, to permit the identification and nuanced characterisation of significant patterning in the archaeological material; secondly, to facilitate a more systematic appraisal of the cemeteries’ largest tombs and their distinctive histories; and, finally, to offer tangible and context-specific evidence as to the behaviours and identities of the burying groups, with a view to eventually comparing such results with mortuary assemblages across Crete.

# 2. Early Iron Age Knossos
## 2.1 Settlement Evidence: 'A meagre filling in a very thick sandwich'

Survey and excavation at Knossos suggest that, despite the final destruction of the palace and a contraction in size at the end of the BA, occupation into the EIA was continuous [for an overview, see @Hatzaki2020]. But compared to the preceding era, archaeological contexts from the EIA town are, in a word, insubstantial. Coldstream lamented that the period's 'remains [...] are like a meagre filling in a very thick sandwich' between BA and later Greaco-Roman levels [@Coldstream1991 287]. Most individual finds of PG or G date take the form of flimsy foundations, patches of earth floor, wells or pottery scatters, and for a time it was argued that the area in the EIA comprised several dispersed villages, which later coalesced as per Aristotle’s model of synoecism [@Alexiou1950].

However, a concentration of activity slightly to the west of the palace has long presented a good case for continuing, nucleated settlement [@Hood1981], something strongly advocated by Coldstream [-@Coldstream1984a; -@Coldstream2000]. More recently, findings of the Knossos Urban Landscape Project (KULP) suggest a PG settlement of up to 40ha in size [@Kotsonas2018]. If accurate, this suggests that EIA Knossos was one of the largest sites of the contemporary Aegean.

```{r map, fig.align= "centre", fig.cap="Map of the Knossos area, showing the locations of the various EIA cemeteries and clusters of tombs. The area of the EIA settlement is based upon Kotsonas et al. [-@Kotsonas2018, Fig. 2]."}

knitr::include_graphics("../figures/Knossos_Area.jpg")
```

## 2.2 The Early Iron Age Cemeteries

Survey aside, our most abundant evidence for the period comes from the mortuary record (see Fig. \@ref(fig:map)). The final BA - Late Minoan IIIC - yields few if any securely datable interments, but from the start of the EIA evidence becomes more plentiful, with burials known from the Kephala ridge [@Coldstream1963 38; @Hogarth1899 82-85], the modern suburbs of Ayios Ioannis and Atsalenio [@Boardman1960a; @Davaras1968 133–46], the cave of Mavro Spelio [@Coldstream2000 291-94], and the Khaniale Teke site with its well-known *tholos* [@Boardman1967;@Hutchinson1954; cf. @Hoffman1997;@Kotsonas2006]. Together these tombs point to a richly variegated set of funerary practices but, despite their reasonable number, their scattered locations present difficulties for analysing the behaviours of distinct burying groups within shared mortuary contexts.

Fortunately, larger and more concentrated areas of burial activity are to be found in the form of the Fortetsa and Knossos North cemeteries (Fig. \@ref(fig:map2)), which together comprise over 130 tombs, ranging in date from the SM to LO periods. Collectively, they represent the most abundant source of evidence for not just mortuary, but social practices of any kind at EIA Knossos.

## 2.3 The Fortetsa and Knossos North Cemeteries

The cemetery on the slope facing the eponymous village of Fortetsa comprises twelve tombs excavated in 1933, and another eight uncovered in 1935 [@Brock1957 xi, 1-2], grouped into three main clusters running north to south. Three of those excavated in 1933 – L, TFT and Π – which lie close to the main road north of the Acropolis hill and southwest of the university hospital, are now considered to belong to the Knossos North Cemetery (KNC).

```{r map2, fig.align= "centre", fig.cap="Closer view of the Knossos North and Fortetsa cemeteries. Tombs were digitised and georeferenced from the original publications, Coldstream and Catling (1996) and Brock (1957)."}

knitr::include_graphics("../figures/Knossos Tombs.jpg")

```

To the northeast of the Fortetsa cemetery, a collection of ten badly damaged tombs known as the Fortetsa 1967 tombs were probably a south-western extension of the KNC. Unfortunately, all had been thoroughly looted at the time of excavation and are thus excluded from the present analysis. Northwest of these, and just south of the Teke Tholos, the Teke tombs belong to the northern part of KNC. The central section of the KNC comprises the ‘Medical Faculty’ site (KNC Main and NW in Fig. \@ref(fig:map2)), named for the extension of the University of Crete that prompted major rescue excavations there in 1978 [@Coldstream1996a 53-55].

As considered here, the Fortetsa and North Cemeteries comprise 20 and 103 tombs respectively. It is possible the Fortetsa tombs form part of a larger cemetery but clear evidence of this has yet to be identified. At both cemeteries, the rock-cut chamber tomb is the most common type; indeed, it is the only type at Fortetsa, while at the KNC it accounts for 63% of all tombs. The chamber tomb comprises a descending ramp or passage, the *dromos*, leading to the chamber itself, cut into the *kouskouras* bedrock. The *dromos* and chamber are separated by the *stomion*, an opening often set a step or two lower than the *dromos*, which is blocked by a large stone slab, smaller stacked stones, or combination of both. The larger size of the KNC is matched by a greater diversity of burial types, which vary by period. In the earliest, SM phase, shaft graves and pit caves are the second and third most common type. Later, in the Geometric period, various simpler pits, sometimes containing *pithoi* or *larnakes* as burial containers, are also encountered. The vast majority of burials are cremations deposited in *pithoi* or other urns, a rite that appears in the SM period, in the Tomb 200+ complex in the KNC [@Coldstream1994 109]. Inhumation, which began as the majority rite, declined from the 10th century. Though it did not completely disappear in the EIA, by the Orientalising period inhumation appears restricted solely to child burials [@Cavanagh1996 652].

## 2.4 Chronological Developments in the Cemeteries

The foundation of the cemeteries has received much attention. It has been claimed that no tomb at Knossos was used continuously from LM IIIC to SM [@Coldstream1998 58; @Coldstream2006 582], though there is a possibility that this is a classificatory artefact [@Hallager2010]. Instances of BA tombs reused in the EIA have been suggested (e.g. the Khaniale Teke tholos), and Cavanagh [-@Cavanagh1996], based on a cluster analysis of tomb dimensions, could not reject the possibility of some KNC tombs being of BA date, though ongoing study of the retained sherds from the KNC excavation suggests very little Late Minoan III activity in the area of the later cemetery (Whitelaw, pers. comm.).

As contentious as their founding is the seemingly abrupt abandonment of both cemeteries around 630 BCE. Admittedly, few new tombs had been built in the preceding century, though interments in old tombs had certainly continued [@Brock1957 4; @Cavanagh1996 651-53]. The ensuing period, accompanied by equally scanty settlement evidence at Knossos, has been dubbed by some the ‘Archaic Gap'. The causes of this lacuna, which persists until the resurgence of archaeological evidence in the Late Archaic period, remain unknown. Although various explanations have been forwarded [@Coldstream1999 301-302; @Huxley1994 126; @Kotsonas2002 41-44], there are now strong reasons to doubt the settlement was abandoned [@Kotsonas2018], and we should always be open to the possibility of revisions in the Archaic ceramic chronology, which remains less secure than that of earlier periods [@Whitley2009 276-77].

## 2.5 Heroes Abroad, Immigrants at Home

A popular approach to the study of Knossos’ EIA tombs has been the investigation of ‘object biography’ [@Appadurai1986; @Kopytoff1986; @Mauss1954], an anthropological lens emphasising the mutable nature of an object’s value, the transformations latent within its social existence. Inspired by the Homeric motif of gift-giving [@Whitley2002 220-21], and the remarkable archaeological parallels for such items – such as between the boar’s tusk helmet from KNC Tomb 201 and that given to Odysseus by Meriones (*Iliad* 10.260-271) - such approaches have been brought to bear on certain elaborate burials of the EIA [most recently @Kotsonas2018]. For instance, Tomb 201 in the KNC contained various weapons, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned helmet [@Catling1995 123]. Based on similarities between this tomb and burials at Tiryns, Kaloriziki, and the Toumba building at Lefkandi, Catling has argued that these burials represent individuals who dared to travel abroad in a time of more limited seafaring, accruing exotica and stories from distant lands, and so were accorded exceptional treatment in death [@Catling1995 127–8].

Complementing the theme of heroes abroad has been that of foreigners at home. Two caches of gold jewellery, found in pits flanking the entrance to the Teke Tholos, were argued by Boardman [-@Boardman1967] to be foundation deposits – a Near Eastern custom – for the tomb of an immigrant Phoenician craftsman. Kotsonas has revisited this attribution, arguing the tomb was rather that of an elite individual or family with a monopoly over a metal workshop’s output [-@Kotsonas2006 155–59]. Either way, it seems unlikely that foreign traders or craftsmen never visited or resided at Knossos during the EIA [@Hoffman1997 176–85; @Schreiber2003 293-306; @Stampolidis2006]. Recent re-examinations of the mortuary material from Lefkandi [@Arrington2016], and Perati in Attica [@Murray2018] have further raised the possibility of heterogenous communities existing in the EIA which, by way of travel or exchange, acquired and expressed a diversity of burial habits within individual sites, in a manner which undermines traditional associations of 'exotic imports' with an elite class. Similar complications to the notion of 'elite' or 'wealthy' tombs will be explored below.

## 2.6 Quantitative Analyses of the Cemeteries

To practitioners in other regions, the application of quantitative methods to analysis of mortuary assemblages may not seem particularly novel. However, in archaeological work on the EIA of Greece more generally, and Crete in particular, such approaches remain far from common. Snodgrass' [e.g. -@Snodgrass1980 22-24] attempts to extrapolate demographic trends for 8th century Attica from estimated burial rates represents a pioneering case, but his assumptions were in time challenged, most notably by Morris [-@Morris1987]. Morris examined the relationship between mortuary visibility and processes of inclusivity and exclusion in access to the burial rites, arguing that ideological, and not demographic, forces were behind the patterns Snodgrass had observed. Meanwhile, Whitley (1991) considered the relationship between pottery style and mortuary practice, which he similarly took to vary according to periods of more open and restricted access to certain burial rites. The work of Whitley and Morris, both students of Snodgrass, remains after 30 years perhaps the most systematic quantitative work yet applied to EIA Greek burial. While not outwardly rejected, such quantitative methods have been little utilised in EIA archaeology since, often from a sense that the available data are insufficient [e.g. @Eaby2007 20-21; though note @Kotsonas2011; @Murray2015].

This situation is reflected at Knossos, where analyses have tended to rely on particular finds [@Coldstream1984b], tombs [@Catling1995], or observable, but limited, patterns in the data [@Cavanagh1996]. Cluster analyses, the most sophisticated statistical techniques so far applied to the material from Knossos, have been employed by Cavanagh [-@Coldstream1996 653-57], Antoniadis [-@Antoniadis2017 131-34] and Whitley, though the state of publication of the KNC at the time of Whitley's doctoral study precluded him from publishing the analyses in full in the eventual monograph. Cavanagh undertook some analysis of the use of the tombs through time, the nature of individual burial assemblages, and changes in burial rites, though as part of the KNC publication, his report is only preliminary. Meanwhile, Antoniadis [-@Antoniadis2017 116-119]  divided the EIA tombs of Knossos into three groups, those with fewer than 10 pots, fewer than 50, and more than 50, and found that, in each successive case, the proportions of tombs containing Near Eastern imports and imitations increased, suggesting unequal access to such prestigious goods. Yet even here the data could be further interrogated (considering, for example, the number of imported items *per burial*). Through a more systematic quantitative analysis, and a range of statistical methods, the present paper seeks to extend and nuance a number of observations that have been made in previous studies of the Knossos cemeteries, as well as revealing patterns thus far unrecognised.

# 3. Methods

## 3.1 Databases

The cornerstone of the present study is a database created from the excavation reports for the Fortetsa and Knossos North cemeteries [@Brock1957; @Coldstream1996a], supplemented by the studies of Jones [-@Jones2000], Antoniadis (2017) and Crowe [-@Crowe2016] to identify all imported vessels, vessels imitating Near Eastern prototypes, and BA antiques respectively. It includes information on every published ceramic vessel (n=4481), non-ceramic object (n=1606), and tomb (n=123), with this last including tallies and summary data based on the first two.

Taken together, these databases make possible a diverse range of quantitative analyses. Sadly, with many of the tombs evidently looted or damaged, this cannot be a complete picture of their original contents. If we assume, though, that natural or accidental destruction was relatively indiscriminate, and that, even in cases of looting, certain object classes (particularly cinerary urns) were seldom robbed, then the data are by no means rendered useless. As Snodgrass [-@Snodgrass1996] cautions, we should be chary of making negative arguments – that is, conclusions drawn about the *lack* of features in certain tombs – but may be somewhat more hopeful in the forming of positive ones. The great advantage of the large, quantitative dataset employed here is that it permits investigation of several nested levels of potentially significant patterning. From individual burials, to collective tombs, to whole cemeteries, we have the opportunity to study and compare both synchronic and diachronic variation across these various levels.

## 3.2 Reproducibility and Data-sharing

Ever more archaeologists have begun to advocate the utility, indeed the necessity, of making available not just the data, but the full methodologies underpinning their published work. With the opportunities afforded by current technologies (digital databases, statistical software packages, online repositories etc.), there remain ever fewer obstacles to the full and free sharing of both data and workflows [@article]. In Marwick's words [-@Marwick2017 445], "[t]he technical problems are largely solved; the challenge now is to change the norms of the discipline to make high reproducibility a canonical attribute of high-quality scholarly work."

In this vein, the present article was written using Marwick's [-@Marwick2017; @inbook] 'rrtools', an open-source package for the statistical software R Studio. This provides the author with the tools to create a self-contained 'compendium', which can be freely downloaded, and from which anyone can reproduce the entire workflow of the published analysis. Every graph presented below was produced directly, through written code, from the raw data using R (and note also that these are presented in colour-blind friendly palettes). This code, as well as the raw and derived datasets, complete bibliography (in BibTex format, for export to any major reference manager), figures, and original article are included in the compendium. It can be downloaded as a repository from GitHub, at the following address: https://github.com/DCPollard94/knossoscemeteries, and the raw datasets have been stored with the online repository Figshare, here: https://doi.org/10.6084/m9.figshare.13019588.v1.

## 3.3 Estimating Interments

A first important step in preparing the data for analysis was the estimation of the number of individuals buried in each tomb. The method employed here broadly follows that of Cavanagh [-@Cavanagh1996 659-60], who took the number of *pithoi* (the most common cremation urn) and the number of *pithoi*, amphorae and kraters (which were sometimes used as such) as middle and upper estimates of the number of burials, with all securely identified interments being the lower limit. Here, the mean of these three estimates was taken, to provide a single aggregate measure for the purposes of analysis.

For most diachronic analyses, however, *pithoi* alone were used as the best proxy for burial numbers. As such, in the following discussion, '*pithoi*' should be taken to indicate urns, while 'vessels' embraces the total ceramic assemblage in each tomb. In most cases, any tombs lacking identified remains or a single vessel suitable for use as an urn were not considered, thus excluding the most heavily damaged and looted tombs.

Two points of caution should be noted. Firstly, and as already mentioned, cremation began as a minority rite in the 11th century, and only came to systematically replace inhumation during the 10th century. However, it appears to be only from around the mid-9th century that *pithoi* became the *de facto* cremation urn. Earlier, ashes may have been placed in other vessels, or even loose in the tombs [@Cavanagh1996 651-52, 659-60]. This change can be seen in Fig. \@ref(fig:aoristbycem)A below, with the sharp uptick in deposited *pithoi* around 850 BCE. Thus we must be careful in making quantitative comparisons between the rates of burial pre- and post-850, because of the differential preservation of inhumations or loose cremations versus ceramic urns. Secondly, *pithoi* were not always used as urns; occasionally they served simply as grave goods like any other vessel, while from the end of the 8th century a class of polychrome *pithoi* could occur in pairs, but accompanying only a single cremation. These caveats should be borne in mind, but needn't undermine the present analysis. The main trends examined here post-date the 9th century shift towards *pithos*-based cremation, while even halving the counts of all polychrome *pithoi* would not greatly affect the ranking of tombs by their estimated number of burials, and not all such vessels were deposited in twos.

## 3.4 Aoristic Approaches to Dating

Another major concern is the temporal uncertainty implicit in the assigning of individual vessels to ceramic periods. A PGB vessel could, on our current best estimates, date anywhere between 840 and 810 BCE. This situation poses challenges to visualisation and analysis. A simple, if coarse, solution is to take the mean date (i.e. 825 for our PGB example). This is used in a couple of instances in the present paper, for assigning vessels to broad, century-long windows.

```{r allpots,fig.dim = c(5.4, 6), dpi=600, fig.align='center', fig.cap="A) Counts of all vessels per decade across both cemeteries. Vessels' dates have here been based on the mean of the absolute date range corresponding to their respective ceramic periods. B) The same but with aoristic sums replacing raw counts. This represents a probabilistically weighted visualisation of depositional activity through time."}

#The tombs database is imported and, for the purposes of this article, the entire database is subsetted to include only those tombs from the Knossos North and Fortetsa cemeteries (excluding the Tekke Tombs).

tombs<-read.csv("../data/raw_data/KnossosTombs.csv",header=TRUE)
tombs<-subset(tombs,tombs$cem_condensed=="KNC"|tombs$cem_condensed=="FOR")

#The database of all vessels is imported and, again only those tombs from the KNC and Fortetsa cemetery are selected. A new object "pithoi" is made by subsetting only those vessels labelled "Pithos", and a similar object "all_urns" by subsetting all pithoi, amphorae and kraters.

pots<-read.csv("../data/raw_data/KnossosPots.csv",header=TRUE)
pots<-subset(pots,pots$cem_condensed=="FOR"|pots$cem_condensed=="KNC")
pithoi<-subset(pots,pots$shape=="Pithos")
all_urns<-subset(pots,pots$shape=="Pithos"|pots$shape=="Amphora"|pots$shape=="Krater")

assemblages<-read.csv("../data/derived_data/KnossosAssemblages.csv",header=TRUE)
objects<-read.csv("../data/raw_data/KnossosObjects.csv")

#The figure is plotted by creating a histogram of all vessels based on their average dates ("ave_date"), which is simply the midpoint of the absolute chronological range associated with their respective ceramic phase. Similar stylistic adjustments are made (font, labels, colour palette, removing a legend, setting the x and y scales, and expanding the borders of the plot slightly to prevent the "600" label being cut off, which would occur without the final line of code.)

rects <- data.frame(xstart = c(1100,970,920,875,840,810,790,745,710,670,630), 
                    xend =    c(970,920,875,840,810,790,745,710,670,630,600), 
                    xmid =   c(1035,945,897.5,857.5,825,800,767.5,727.5,690,650,615),
                    col = c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
rects$col<-factor(rects$col,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))

non_aorist_fig<-ggplot()+
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf), fill=NA, colour="grey90",alpha = 0.4) +
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = 585, ymax = 620), fill=NA, colour="black",size=0.5) +
  geom_histogram(data=pots, aes(ave_date,fill="Pots",color="Pots"),fill="grey90",color="black",binwidth = 10,size=0.4)+
  geom_text(data=rects,aes(xmid, 602,label=col),family="serif",size=2.8)+
  theme_bw()+theme(legend.title=element_blank(),
      text=element_text(family="serif", size=11),
          plot.margin=unit(c(0.1,0.5,0.1,0.2),"cm"))+
          labs(x="Date (Years BCE)", y="Vessel count per decade")+
                theme(legend.position="none",panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
                  scale_y_continuous(expand=c(0,0),limits=c(0,620))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))

#The aoristic values for every tomb are summed by decade, and then this object is converted to a data frame.

aorist_all_tombs<-colSums(tombs[160:209])
aorist_all_tombs<-data.frame(aorist_all_tombs)

#The column of values is labelled "aoristic_sum", an vector named "dates" is created with values between 1100 and 610 (in that order) at intervals of 10, and then this used to rename the rows of the data frame. Finally, these "row names" are converted into a functioning row in their own right within the data frame (this makes them plottable as a variable) and then are converted into numeric variables, allowing the plotting of them along a continuous scale.

colnames(aorist_all_tombs)<-c("aoristic_sum")
dates<-seq(from = 1100, to = 610, by = -10)
row.names(aorist_all_tombs)<-dates
setDT(aorist_all_tombs, keep.rownames = "decade")
aorist_all_tombs$decade<-as.numeric(aorist_all_tombs$decade)

#The plot is produced, following the same stylistic adjustments of the previous plot.

aorist_fig<-ggplot()+
   geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf), fill=NA, colour="grey90",alpha = 0.4) +
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = 212, ymax = 226), fill=NA, colour="black",size=0.4) +
  geom_text(data=rects,aes(xmid, 219,label=col),family="serif",size=2.8)+
  geom_col(data=aorist_all_tombs,aes(decade,aoristic_sum, fill="Aoristic Sum",color="Aoristic Sum"),
           fill="grey90",color="black",width=10,size=0.4)+
  theme_bw()+theme(legend.title=element_blank(),
      text=element_text(family="serif", size=11),
          plot.margin=unit(c(0.1,0.5,0.1,0.2),"cm"))+
          labs(x="Date (Years BCE)", y="Aoristic sum per decade")+
              theme(legend.position="none",panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
                  scale_y_continuous(expand=c(0,0),limits=c(0,226))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))

ggarrange(non_aorist_fig,aorist_fig,nrow=2,labels=c("A","B"),align="v",common.legend=FALSE)

```

But the artificial clustering this technique engenders around select absolute dates can prove misleading, as in the many apparent oscillations and interruptions in depositional activity visible in Fig. \@ref(fig:allpots)A.

An alternative approach, employed extensively in this paper, is the use of so-called 'aoristic analysis' [for original use in criminology, see @Ratcliffe2000; for archaeological applications, see @Crema2010; @Crema2012; @Johnson2004a]. Here, rather than give a pot a single date, the entire time span under investigation is divided into equal chunks (in this case, decades), and probability values summing to one are then assigned per chunk of time, based on the vessel's date range. So, our PGB vessel, datable between 840 and 810 BCE, would register a value of 0.33 for each of the three decades this encompasses. These values for individual vessels can then be summed to produce tomb- or cemetery-level signatures. A particular attraction of this method is that it incorporates temporal uncertainty into the size of the resulting values, that is, 'events with tight temporal definition contribute more to the total probability over their range than do loosely defined events' [@Johnson2004a 450].

As seen in Fig. \@ref(fig:allpots)B, the artificial peaks and troughs of the simple averaging method are avoided, and different patterns become visible. Though these methods present their own limitations [@Crema2010 1123-24], they are considered a more robust and productive approach to the temporal uncertainties implicit in ceramic dating.

# 4. Analysis

The present analysis proceeds by first identifying two distinctive trends in burial activity among certain tombs at Fortetsa and the KNC, before advancing to questions of what their use can reveal about broader changes in the social context and significance of burial within the Knossian community.

## 4.1 Temporal Developments in the Knossian Cemeteries

An obvious place to begin is with the construction of the tombs themselves. Sadly, the establishment of a secure chronology in this regard at Fortetsa and the KNC is quite difficult, given the frequency of later reuse, looting or other disturbances.

```{r tombconstruction, fig.dim=c(7,4), dpi=600, fig.cap="Tomb construction through time at both cemeteries. A) Counts based on the ceramic phase considered most likely to correspond to the tomb's first use (including simply 'PG', 'G' and 'O' where no greater specificity was possible). B) Aoristic sums of the same, where the uncertainty associated with general period assignations is spread across the corresponding sub-periods. In both cases, tombs lacking any securely dateable pottery are excluded."}

tomb_dates_aorist<-read.csv("../data/derived_data/tomb_dates_aorist.csv",header=TRUE)
tomb_dates_aorist_KNC<-subset(tomb_dates_aorist,tomb_dates_aorist$cem_condensed=="NC")
tomb_dates_aorist_FOR<-subset(tomb_dates_aorist,tomb_dates_aorist$cem_condensed=="FOR")

KNC<-subset(tombs,tombs$cem_condensed=="KNC")
FOR<-subset(tombs,tombs$cem_condensed=="FOR")

aorist_tomb_periods<-colSums(tomb_dates_aorist_KNC[5:14])
aorist_tomb_periods<-data.frame(aorist_tomb_periods)
aorist_tomb_periods$FOR<-colSums(tomb_dates_aorist_FOR[5:14])
colnames(aorist_tomb_periods)<-c("KNC","FOR")
setDT(aorist_tomb_periods, keep.rownames = "period")
aorist_tomb_periods$period<-factor(aorist_tomb_periods$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
aorist_tomb_periods_melt<-melt(aorist_tomb_periods)
colnames(aorist_tomb_periods_melt)<-c("period","cemetery","aoristic_sum")

tombs$start_period<-factor(tombs$start_period,levels=c("SM", "EPG", "MPG", "PG","LPG", "PGB", "EG", "MG","G", "LG", "EO","O","LO"))

tombs$cem_condensed<-factor(tombs$cem_condensed,levels=c("KNC","FOR"))

cbPalette <- c("#999999","#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

new_tomb_counts_plot<-ggplot(data=subset(tombs,!is.na(start_period)),aes(start_period, fill=forcats::fct_rev(tomb_type_condensed),color=forcats::fct_rev(tomb_type_condensed)))+geom_bar(width=1,size=0.4)+
  scale_color_manual(values=c("black","black","black","black","black","black","black","black"),labels=c("Unknown","Shaft Grave","Pit Cave","Pithos Burial","Pit","Other","Larnax Burial","Chamber Tomb"))+
        theme_bw()+theme(legend.title=element_blank(),legend.spacing.x = unit(0.1,"cm"), text=element_text(family="serif", size=12),
          axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),
          panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
            labs(x="Period", y="Count of new tombs")+scale_fill_manual(values=cbPalette,
                                                                                        labels=c("Unknown","Shaft Grave","Pit Cave",
                                                                                        "Pithos Burial","Pit","Other","Larnax Burial","Chamber Tomb"))+
             scale_y_continuous(expand=c(0,0),limits=c(0,25))+scale_x_discrete(expand=c(0,0))+
facet_wrap(~cem_condensed)+
               theme(strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA))

new_tomb_aorist_plot<-ggplot()+geom_line(data=aorist_tomb_periods_melt,aes(period,aoristic_sum,group=cemetery,linetype=cemetery))+
  theme_bw()+theme(legend.title=element_blank(),
          text=element_text(family="serif", size=12),
          axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),
          panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
            labs(x="Period", y="Aoristic sum of new tombs")+
               scale_fill_manual(values =c("gray50","gray80"),labels=c(" KNC  "," FC"))+ 
                 scale_y_continuous(expand=c(0,0))+scale_x_discrete(expand=c(0,0))

ggarrange(new_tomb_counts_plot,new_tomb_aorist_plot,ncol=2,align="h",labels=c("A","B"), widths=c(1.6,1),common.legend=FALSE, legend="bottom")

```

The periods of construction shown in Fig. \@ref(fig:tombconstruction) correspond, therefore, largely with the earliest surviving pottery from each tomb. On this basis, it would appear that the KNC began with many tombs built during the - admittedly prolonged - SM period, followed by a protracted decline in construction, offset by resurgences in the LPG and LG periods. The Fortetsa tombs, which were all of chamber type, mostly date to PG, with almost no new tomb construction post-PGB. Indeed, it has been suggested that chamber tombs actually ceased to be created into the 8th century, with those appearing as MG or later in date being heavily looted and thus, in reality, older (Brock 1957, p.4; Cavanagh 1996, pp.652–53). What is clear is that many later tombs in the KNC are not chambers, but a mix of pit, *larnax*, and *pithos* burials, as well as unidentified forms. It is tempting to hypothesise similar burials having been overlooked at Fortetsa, but as a smaller cemetery it may have simply served a more restricted burying population and, thus, hosted more limited rites. The patterns at both cemeteries are by no means identical, then, even if there is some similarity in the declining investment in chamber tomb construction in the latter part of the EIA.

```{r tombsinusebybuilddate, fig.dim=c(7,3), dpi=600, fig.cap=" A) Estimates of the number of tombs in use through time at the KNC and B) at Fortetsa. The grey background represents the total number of tombs in use, while the lines correspond to tombs built in each of the five centuries of EIA activity at the cemeteries. C) The KNC again, but only showing chamber tombs definitely built before the 8th century."}

KNC_tombs_in_use<-data.frame(colSums(KNC[6:15]!=""))
colnames(KNC_tombs_in_use)<-c("no_tombs")
setDT(KNC_tombs_in_use, keep.rownames = "period")
KNC_tombs_in_use$period<-factor(KNC_tombs_in_use$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
KNC_tombs_in_use$cemetery<-"KNC"

FOR_tombs_in_use<-data.frame(colSums(FOR[6:15]!=""))
colnames(FOR_tombs_in_use)<-c("no_tombs")
setDT(FOR_tombs_in_use, keep.rownames = "period")
FOR_tombs_in_use$period<-factor(FOR_tombs_in_use$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
FOR_tombs_in_use$cemetery<-"FOR"

tombs_in_use_cem<-rbind(KNC_tombs_in_use,FOR_tombs_in_use)
tombs_in_use_cem$cemetery<-factor(tombs_in_use_cem$cemetery,levels=c("KNC","FOR"))

tombs_in_use<-data.frame(colSums(tombs[6:15]!=""))
colnames(tombs_in_use)<-c("no_tombs")
setDT(tombs_in_use, keep.rownames = "period")
tombs_in_use$period<-factor(tombs_in_use$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))

tombs_11th<-subset(KNC,century_built=="11th")
tombs_10th<-subset(KNC,century_built=="10th")
tombs_9th<-subset(KNC,century_built=="9th")
tombs_8th<-subset(KNC,century_built=="8th")
tombs_7th<-subset(KNC,century_built=="7th")

tombs_in_use_11th<-colSums(tombs_11th[6:15]!="")
tombs_in_use_11th<-data.frame(tombs_in_use_11th)
colnames(tombs_in_use_11th)<-c("no_tombs")
setDT(tombs_in_use_11th, keep.rownames = "period")
tombs_in_use_11th$period<-factor(tombs_in_use_11th$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
tombs_in_use_11th$tomb_century<-"11th"

tombs_in_use_10th<-colSums(tombs_10th[6:15]!="")
tombs_in_use_10th<-data.frame(tombs_in_use_10th)
colnames(tombs_in_use_10th)<-c("no_tombs")
setDT(tombs_in_use_10th, keep.rownames = "period")
tombs_in_use_10th$period<-factor(tombs_in_use_10th$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
tombs_in_use_10th$tomb_century<-"10th"

tombs_in_use_9th<-colSums(tombs_9th[6:15]!="")
tombs_in_use_9th<-data.frame(tombs_in_use_9th)
colnames(tombs_in_use_9th)<-c("no_tombs")
setDT(tombs_in_use_9th, keep.rownames = "period")
tombs_in_use_9th$period<-factor(tombs_in_use_9th$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
tombs_in_use_9th$tomb_century<-"9th"

tombs_in_use_8th<-colSums(tombs_8th[6:15]!="")
tombs_in_use_8th<-data.frame(tombs_in_use_8th)
colnames(tombs_in_use_8th)<-c("no_tombs")
setDT(tombs_in_use_8th, keep.rownames = "period")
tombs_in_use_8th$period<-factor(tombs_in_use_8th$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
tombs_in_use_8th$tomb_century<-"8th"

tombs_in_use_7th<-colSums(tombs_7th[6:15]!="")
tombs_in_use_7th<-data.frame(tombs_in_use_7th)
colnames(tombs_in_use_7th)<-c("no_tombs")
setDT(tombs_in_use_7th, keep.rownames = "period")
tombs_in_use_7th$period<-factor(tombs_in_use_7th$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
tombs_in_use_7th$tomb_century<-"7th"

tomb_use<-rbind(tombs_in_use_11th,tombs_in_use_10th,tombs_in_use_9th,tombs_in_use_8th,tombs_in_use_7th)
tomb_use$tomb_century<-factor(tomb_use$tomb_century,levels=c("11th","10th","9th","8th","7th"))

KNC_use<-ggplot()+geom_area(data=subset(tombs_in_use_cem,cemetery=="KNC"),aes(period,no_tombs,fill="All tombs",group=cemetery),alpha=0.2)+
            geom_line(data=tomb_use,aes(period,no_tombs,linetype=tomb_century,group=tomb_century))+
              theme_bw()+theme(text=element_text(family="serif", size=11),axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),
          panel.grid.major = element_blank(), panel.grid.minor = element_blank(),legend.position = "bottom",
          legend.text=element_text(size=10))+
                 labs(x="Period", y="Tombs in use",linetype="Century of tomb construction",fill="")+
                  scale_fill_manual(values="#999999")+
                    scale_y_continuous(expand=c(0,0),limits=c(0,60))+
                      scale_x_discrete(expand=c(0,0))

Ftombs_11th<-subset(FOR,century_built=="11th")
Ftombs_10th<-subset(FOR,century_built=="10th")
Ftombs_9th<-subset(FOR,century_built=="9th")
Ftombs_8th<-subset(FOR,century_built=="8th")

Ftombs_in_use_11th<-colSums(Ftombs_11th[6:15]!="")
Ftombs_in_use_11th<-data.frame(Ftombs_in_use_11th)
colnames(Ftombs_in_use_11th)<-c("no_tombs")
setDT(Ftombs_in_use_11th, keep.rownames = "period")
Ftombs_in_use_11th$period<-factor(Ftombs_in_use_11th$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
Ftombs_in_use_11th$tomb_century<-"11th"

Ftombs_in_use_10th<-colSums(Ftombs_10th[6:15]!="")
Ftombs_in_use_10th<-data.frame(Ftombs_in_use_10th)
colnames(Ftombs_in_use_10th)<-c("no_tombs")
setDT(Ftombs_in_use_10th, keep.rownames = "period")
Ftombs_in_use_10th$period<-factor(Ftombs_in_use_10th$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
Ftombs_in_use_10th$tomb_century<-"10th"

Ftombs_in_use_9th<-colSums(Ftombs_9th[6:15]!="")
Ftombs_in_use_9th<-data.frame(Ftombs_in_use_9th)
colnames(Ftombs_in_use_9th)<-c("no_tombs")
setDT(Ftombs_in_use_9th, keep.rownames = "period")
Ftombs_in_use_9th$period<-factor(Ftombs_in_use_9th$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
Ftombs_in_use_9th$tomb_century<-"9th"

Ftombs_in_use_8th<-colSums(Ftombs_8th[6:15]!="")
Ftombs_in_use_8th<-data.frame(Ftombs_in_use_8th)
colnames(Ftombs_in_use_8th)<-c("no_tombs")
setDT(Ftombs_in_use_8th, keep.rownames = "period")
Ftombs_in_use_8th$period<-factor(Ftombs_in_use_8th$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
Ftombs_in_use_8th$tomb_century<-"8th"

Ftomb_use<-rbind(Ftombs_in_use_11th,Ftombs_in_use_10th,Ftombs_in_use_9th,Ftombs_in_use_8th)
Ftomb_use$tomb_century<-factor(Ftomb_use$tomb_century,levels=c("11th","10th","9th","8th"))

FOR_use<-ggplot()+geom_area(data=subset(tombs_in_use_cem,cemetery=="FOR"),aes(period,no_tombs,fill="All tombs",group=cemetery),alpha=0.2)+
            geom_line(data=Ftomb_use,aes(period,no_tombs,linetype=tomb_century,group=tomb_century))+
              theme_bw()+theme(text=element_text(family="serif", size=11),axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),
          panel.grid.major = element_blank(), panel.grid.minor = element_blank(),legend.position = "bottom",
          legend.text=element_text(size=10))+
                 labs(x="Period", y="Tombs in use",linetype="Century of tomb construction",fill="")+
                  scale_fill_manual(values="#999999")+
                    scale_y_continuous(expand=c(0,0),limits=c(0,20))+
                      scale_x_discrete(expand=c(0,0))

KNC_CT<-subset(KNC,tomb_type_condensed=="CT" & KNC$start_date>799)
KNC_tombs_in_use_CT<-data.frame(colSums(KNC_CT[6:15]!=""))
colnames(KNC_tombs_in_use_CT)<-c("no_tombs")
setDT(KNC_tombs_in_use_CT, keep.rownames = "period")
KNC_tombs_in_use_CT$period<-factor(KNC_tombs_in_use_CT$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))

tombs_11th_CT<-subset(KNC_CT,century_built=="11th")
tombs_10th_CT<-subset(KNC_CT,century_built=="10th")
tombs_9th_CT<-subset(KNC_CT,century_built=="9th")

tombs_in_use_11th_CT<-colSums(tombs_11th_CT[6:15]!="")
tombs_in_use_11th_CT<-data.frame(tombs_in_use_11th_CT)
colnames(tombs_in_use_11th_CT)<-c("no_tombs")
setDT(tombs_in_use_11th_CT, keep.rownames = "period")
tombs_in_use_11th_CT$period<-factor(tombs_in_use_11th_CT$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
tombs_in_use_11th_CT$tomb_century<-"11th"

tombs_in_use_10th_CT<-colSums(tombs_10th_CT[6:15]!="")
tombs_in_use_10th_CT<-data.frame(tombs_in_use_10th_CT)
colnames(tombs_in_use_10th_CT)<-c("no_tombs")
setDT(tombs_in_use_10th_CT, keep.rownames = "period")
tombs_in_use_10th_CT$period<-factor(tombs_in_use_10th_CT$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
tombs_in_use_10th_CT$tomb_century<-"10th"

tombs_in_use_9th_CT<-colSums(tombs_9th_CT[6:15]!="")
tombs_in_use_9th_CT<-data.frame(tombs_in_use_9th_CT)
colnames(tombs_in_use_9th_CT)<-c("no_tombs")
setDT(tombs_in_use_9th_CT, keep.rownames = "period")
tombs_in_use_9th_CT$period<-factor(tombs_in_use_9th_CT$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))
tombs_in_use_9th_CT$tomb_century<-"9th"

tomb_use_CT<-rbind(tombs_in_use_11th_CT,tombs_in_use_10th_CT,tombs_in_use_9th_CT)
tomb_use_CT$tomb_century<-factor(tomb_use_CT$tomb_century,levels=c("11th","10th","9th"))

KNC_CT_use<-ggplot()+geom_area(data=(KNC_tombs_in_use_CT),aes(period,no_tombs,fill="All tombs",group=1),alpha=0.2)+
  geom_line(data=tomb_use_CT,aes(period,no_tombs,linetype=tomb_century,group=tomb_century))+
  theme_bw()+theme(text=element_text(family="serif", size=11),axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),
                   panel.grid.major = element_blank(), panel.grid.minor = element_blank(),legend.position = "bottom",
                   legend.text=element_text(size=10))+
  labs(x="Period", y="Tombs in use",linetype="Century of tomb construction",fill="")+
  scale_fill_manual(values="#999999")+
  scale_y_continuous(expand=c(0,0),limits=c(0,60))+
  scale_x_discrete(expand=c(0,0))

ggarrange(KNC_use,FOR_use,KNC_CT_use,ncol=3,common.legend = TRUE,legend="bottom", labels=c("A","B","C"))
```

A further contrast is seen in the number of tombs in use in each period (Fig. \@ref(fig:tombsinusebybuilddate)). While at Fortetsa the pattern skews early, with several graves apparently active solely during the PG period, at the KNC the number of tombs in use rose to a peak in the late 8th century before declining. This increase belies a more complex reality, however, with older tombs being steadily abandoned as new ones were constructed (Fig. \@ref(fig:tombsinusebybuilddate)A). This is discernible too at Fortetsa, though in a more sporadic manner largely attributable to the smaller number of tombs (Fig. \@ref(fig:tombsinusebybuilddate)B). The rise in tombs in use at the KNC, and lack thereof at Fortetsa, is thus primarily due to the construction of later tombs, as shown in Fig. \@ref(fig:tombsinusebybuilddate)C, where only pre-8th century chamber tombs at the KNC are presented, and the pattern is more similar to that at Fortetsa. However, an additional phenomenon at the KNC (Figs \@ref(fig:tombsinusebybuilddate)A and C) is the reuse of some SM tombs at much later dates, with an abandonment in between. This reuse is unlikely to have been knowingly undertaken by distant descendants of the original occupants and so raises questions of how the right of access to a sepulchre was maintained, or re-established.

```{r aoristbycem, fig.dim=c(5.5,6.2), dpi=600, fig.align="center", fig.cap="A) Aoristic sums of pithoi, as a proxy for cremation burials, through time for Fortetsa, the KNC, and all tombs together. B) Aoristic sums of all vessels through time."}

aorist_KNC_pithoi <- colSums(KNC[210:259])
aorist_KNC_pithoi<-data.frame(aorist_KNC_pithoi)

colnames(aorist_KNC_pithoi)<-c("aoristic_sum")
row.names(aorist_KNC_pithoi)<-dates
aorist_KNC_pithoi<-data.frame(aorist_KNC_pithoi)
setDT(aorist_KNC_pithoi, keep.rownames = "decade")
aorist_KNC_pithoi$decade<-as.numeric(aorist_KNC_pithoi$decade)

FOR<-subset(tombs,tombs$cem_condensed=="FOR")

aorist_FOR_pithoi <-colSums(FOR[210:259])
aorist_FOR_pithoi<-data.frame(aorist_FOR_pithoi)

colnames(aorist_FOR_pithoi)<-c("aoristic_sum")
row.names(aorist_FOR_pithoi)<-dates
aorist_FOR_pithoi<-data.frame(aorist_FOR_pithoi)
setDT(aorist_FOR_pithoi, keep.rownames = "decade")
aorist_FOR_pithoi$decade<-as.numeric(aorist_FOR_pithoi$decade)

aorist_all_pithoi <- colSums(tombs[210:259])
aorist_all_pithoi<-data.frame(aorist_all_pithoi)

colnames(aorist_all_pithoi)<-c("aoristic_sum")

row.names(aorist_all_pithoi)<-dates
aorist_all_pithoi<-data.frame(aorist_all_pithoi)
setDT(aorist_all_pithoi, keep.rownames = "decade")
aorist_all_pithoi$decade<-as.numeric(aorist_all_pithoi$decade)

aorist_FOR_pithoi$cemetery<-"FOR"
aorist_KNC_pithoi$cemetery<-"KNC"
aorist_all_pithoi$cemetery<-"All tombs"
aorist_pithoi<-rbind(aorist_all_pithoi,aorist_KNC_pithoi,aorist_FOR_pithoi)
aorist_pithoi$cemetery<-factor(aorist_pithoi$cemetery,levels=c("All tombs", "KNC", "FOR"))

aorist_pithoi_plot<-ggplot(NULL)+
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf), fill=NA, colour="grey90",alpha = 0.4) +
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = 42.6, ymax = 45), fill=NA, colour="black",size = 0.5) +
  geom_text(data=rects,aes(xmid, 43.8,label=col),family="serif",size=2.9)+
  geom_line(data=aorist_pithoi,aes(decade,aoristic_sum,linetype=cemetery))+ 
       theme_bw()+theme(legend.title=element_blank(),
          text=element_text(family="serif", size=12),panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          plot.margin=unit(c(0.1,0.2,0.1,0.2),"cm"))+
            labs(x="Date (Years BCE)", y="Aoristic sum (Pithoi)")+
                scale_y_continuous(expand=c(0,0),limits=c(0,45))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))

aorist_KNC <- colSums(KNC[160:209])
aorist_KNC<-data.frame(aorist_KNC)

colnames(aorist_KNC)<-c("aoristic_sum")
row.names(aorist_KNC)<-dates
aorist_KNC<-data.frame(aorist_KNC)
setDT(aorist_KNC, keep.rownames = "decade")
aorist_KNC$decade<-as.numeric(aorist_KNC$decade)

FOR<-subset(tombs,tombs$cem_condensed=="FOR")

aorist_FOR <-colSums(FOR[160:209])
aorist_FOR<-data.frame(aorist_FOR)

colnames(aorist_FOR)<-c("aoristic_sum")
row.names(aorist_FOR)<-dates
aorist_FOR<-data.frame(aorist_FOR)
setDT(aorist_FOR, keep.rownames = "decade")
aorist_FOR$decade<-as.numeric(aorist_FOR$decade)

aorist_FOR$cemetery<-"FOR"
aorist_KNC$cemetery<-"KNC"
aorist_all_tombs$cemetery<-"All tombs"
aorist_tombs<-rbind(aorist_all_tombs,aorist_KNC,aorist_FOR)
aorist_tombs$cemetery<-factor(aorist_pithoi$cemetery,levels=c("All tombs", "KNC", "FOR"))

aorist_pots_plot<-ggplot(NULL)+
   geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf), fill=NA, colour="grey90",alpha = 0.4) +
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = 218, ymax = 230), fill=NA, colour="black",size = 0.5) +
  geom_text(data=rects,aes(xmid, 224,label=col),family="serif",size=2.9)+
  geom_line(data=aorist_tombs,aes(decade,aoristic_sum,linetype=cemetery))+
        theme_bw()+theme(legend.title=element_blank(),
          text=element_text(family="serif", size=12),panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          plot.margin=unit(c(0.1,0.2,0.1,0.2),"cm"))+
            labs(x="Date (Years BCE)", y="Aoristic sum (All vessels)")+
              scale_fill_manual(values = c("gray50","gray65","gray80"))+  
                  scale_y_continuous(expand=c(0,0),limits=c(0,230))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))

ggarrange(aorist_pithoi_plot,aorist_pots_plot,nrow=2,labels=c("A","B"),align="v",common.legend=TRUE, legend="bottom")
```

Turning to the burials themselves, and despite differences in scale and timing, Fortetsa and the KNC underwent similar chronological developments with regard to cremations (*pithoi*) and, to a slightly lesser degree, associated vessels (Fig. \@ref(fig:aoristbycem)). In both cemeteries, the rate of cremation burial began rising in the mid-9th century, plateaued slightly in the early 8th, and then rose again to a peak in the late 8th (KNC) and early 7th (Fortetsa). Regarding all ceramic vessels, the developments align less well; Fortetsa lacks the peak seen in the KNC in PGB-EG, depositions instead remaining relatively steady until the second half of the 8th century. What both cemeteries share, then, is a notable increase in both cremation burials and associated vessels in the LG-EO periods, with a subsequent decline. While in the case of the KNC, these developments seemingly correspond with the number of tombs in use, at Fortetsa there is a disparity, with burial activity increasing despite little change (and ultimately a decline) in tomb-use.

```{r potsbydateandtombdate,  fig.dim=c(6.5,4.6), dpi=600, fig.cap="Counts of pithoi from all tombs arranged A) by the century to which they are assigned based on their ceramic period and B) by the centuries in which the tombs they come from were built. C) and D) represent the same but for all ceramic vessels."}

pots$century<-factor(pots$century,levels=c("11th","10th","9th","8th","7th"))
pots$tomb_century<-factor(pots$tomb_century,levels=c("11th","10th","9th","8th","7th"))
pots$tomb_type <- tombs$tomb_type[match(pots$tomb,tombs$tomb)]
pots$tomb_type<-as.character(pots$tomb_type)
pots$tomb_type[pots$tomb_type!="CT"]<-"Other"
pots$tomb_type<-as.factor(pots$tomb_type)
pots$tomb_type<-factor(pots$tomb_type,levels=c("Other","CT"))

date_by_vessel<-ggplot(data=subset(pots,pots$century!="" & !is.na(tomb_type)),
                       aes(century,fill=tomb_type,color=tomb_type))+geom_bar(size=0.5)+
  theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
               text=element_text(family="serif", size=10),axis.text.y=element_text(angle=90, vjust=0, hjust=0.5),
               legend.title = element_blank(),legend.text=element_text(size=10))+
                 labs(x="Date of vessel (century BCE)", y="Count")+
  scale_fill_manual(values=c("white","grey90"),labels=c("Other","Chamber tomb"),guide = guide_legend(reverse = TRUE))+
  scale_color_manual(values=c("black","black"),labels=c("Other","Chamber tomb"),guide = guide_legend(reverse = TRUE))+
                    scale_y_continuous(expand=c(0,0),limits=c(0,1800))+facet_wrap(~cem_condensed)+
                       theme(strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA),
                             strip.text.x = element_text(size=9,margin = margin(0.05,0,0.1,0, "cm")))

date_by_tomb<-ggplot(data=subset(pots,pots$century!="" & !is.na(tomb_type)),
                     aes(tomb_century,fill=tomb_type,color=tomb_type))+geom_bar(size=0.5)+
  theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
               text=element_text(family="serif", size=10),axis.text.y=element_text(angle=90, vjust=0, hjust=0.5),
               legend.title = element_blank(),legend.text=element_text(size=10))+
                 labs(x="Date of tomb (century BCE)", y="Count")+
  scale_fill_manual(values=c("white","grey90"),labels=c("Other","Chamber tomb"),guide = guide_legend(reverse = TRUE))+
    scale_color_manual(values=c("black","black"),labels=c("Other","Chamber tomb"),guide = guide_legend(reverse = TRUE))+
                    scale_y_continuous(expand=c(0,0),limits=c(0,1800))+facet_wrap(~cem_condensed)+
                       theme(strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA),
                             strip.text.x = element_text(size=9,margin = margin(0.05,0,0.1,0, "cm")))

date_by_pithos<-ggplot(data=subset(pots,century!="" & shape=="Pithos" & !is.na(tomb_type)),
                       aes(century,fill=tomb_type,color=tomb_type))+geom_bar(size=0.5)+
  theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
               text=element_text(family="serif", size=10),axis.text.y=element_text(angle=90, vjust=0, hjust=0.5),
               legend.title = element_blank(),legend.text=element_text(size=10))+
                 labs(x="Date of pithos (century BCE)", y="Count")+
  scale_fill_manual(values=c("white","grey90"),labels=c("Other","Chamber tomb"),guide = guide_legend(reverse = TRUE))+
    scale_color_manual(values=c("black","black"),labels=c("Other","Chamber tomb"),guide = guide_legend(reverse = TRUE))+
                    scale_y_continuous(expand=c(0,0),limits=c(0,310))+facet_wrap(~cem_condensed)+
                       theme(strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA),
                             strip.text.x = element_text(size=9,margin = margin(0.05,0,0.1,0, "cm")))

date_by_tomb_pithos<-ggplot(data=subset(pots,pots$century!="" & shape=="Pithos" & !is.na(tomb_type)),
                            aes(tomb_century,fill=tomb_type,color=tomb_type))+geom_bar(size=0.5)+
  theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
               text=element_text(family="serif", size=10),axis.text.y=element_text(angle=90, vjust=0, hjust=0.5),
               legend.title = element_blank(),legend.text=element_text(size=10))+
                 labs(x="Date of tomb (century BCE)", y="Count")+
  scale_fill_manual(values=c("white","grey90"),labels=c("Other","Chamber tomb"),guide = guide_legend(reverse = TRUE))+
    scale_color_manual(values=c("black","black"),labels=c("Other","Chamber tomb"),guide = guide_legend(reverse = TRUE))+
                    scale_y_continuous(expand=c(0,0),limits=c(0,310))+facet_wrap(~cem_condensed)+
                       theme(strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA),
                             strip.text.x = element_text(size=9,margin = margin(0.05,0,0.1,0, "cm")))

ggarrange(date_by_pithos, date_by_tomb_pithos,date_by_vessel,date_by_tomb,nrow=2,ncol=2,labels=c("A","B","C","D"),
          common.legend = TRUE,legend="bottom")

```

This shared rise in burial activity was not, however, evenly distributed amongst all tombs (Fig. \@ref(fig:potsbydateandtombdate)). That is, though largely deposited during the late 8th and early 7th centuries, the greatest proportion of cremation *pithoi* and associated vessels ended up in chamber tombs built back in the 9th and 10th. The similarities in this pattern at Fortetsa and the KNC may prove significant, but an obvious objection is that chamber tombs - which dominate the total assemblage at the KNC, and are the only type at Fortetsa - may have ceased to be constructed in the 8th century, and so any later rise in interments would naturally be concentrated in older tombs. Therefore, it is necessary to establish whether all or merely some tombs of this type received large quantities of later burials.

## 4.2 Isolating the Phenomenon

A first step might be to reason that those tombs which became the focus of increased later burial activity would end up yielding the most *pithoi* and grave goods when excavated.

```{r centsize,fig.dim=c(5,5), dpi=600, fig.cap="Scatterplots showing the relationship between the number of vessels dated to the A) 10th, B) 9th, C) 8th, and D) 7th centuries in tombs and the total number of vessels found therein. In each case, only tombs built by the century in question are plotted."}

tombs$vessels_10th<-rowSums(tombs[,170:179])
tombs$vessels_9th<-rowSums(tombs[,180:189])
tombs$vessels_8th<-rowSums(tombs[,190:199])
tombs$vessels_7th<-rowSums(tombs[,200:209])

cbPalette <- c("#E69F00", "#56B4E9")

cor10th<-ggplot()+geom_point(data=subset(tombs,start_date>900),aes(vessels_10th,total_pottery,color=cem_condensed),size=0.95)+
  stat_cor(data=subset(tombs,start_date>900),aes(vessels_10th,total_pottery),method="pearson",
           label.x.npc=0.4,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(y="Total vessels", x="10th century vessels")+
                               scale_x_continuous(expand = expansion(mult = c(0.001, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0.001, 0.05)),limits=c(0,470))+
  scale_color_manual(values=cbPalette)

cor9th<-ggplot()+geom_point(data=subset(tombs,start_date>800),aes(vessels_9th,total_pottery,color=cem_condensed),size=0.95)+
  stat_cor(data=subset(tombs,start_date>800),aes(vessels_9th,total_pottery),method="pearson",
           label.x.npc=0.7,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(y="Total vessels", x="9th century vessels")+
                               scale_x_continuous(expand = expansion(mult = c(0.001, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0.001, 0.05)),limits=c(0,470))+
  scale_color_manual(values=cbPalette)

cor8th<-ggplot()+geom_point(data=subset(tombs,start_date>700),aes(vessels_8th,total_pottery,color=cem_condensed),size=0.95)+
  stat_cor(data=subset(tombs,start_date>700),aes(vessels_8th,total_pottery),method="pearson",
           label.x.npc=0.2,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(y="Total vessels", x="8th century vessels")+
                               scale_x_continuous(expand = expansion(mult = c(0.001, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0.001, 0.05)),limits=c(0,470))+
  scale_color_manual(values=cbPalette)

cor7th<-ggplot()+geom_point(data=tombs,aes(vessels_7th,total_pottery,color=cem_condensed),size=0.95)+
  stat_cor(data=tombs,aes(vessels_7th,total_pottery),method="pearson",
           label.x.npc=0.6,label.y.npc=0.3,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(y="Total vessels", x="7th century vessels")+
                               scale_x_continuous(expand = expansion(mult = c(0.001, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0.001, 0.05)),limits=c(0,470))+
  scale_color_manual(values=cbPalette)

ggarrange(cor10th,cor9th,cor8th,cor7th,nrow=2,ncol=2,labels=c("A","B","C","D"),common.legend = TRUE,legend="bottom")
```

This suspicion is well supported by Fig. \@ref(fig:centsize), where the strongest correlations are seen between tombs which received more 8th and 7th century vessels, and which had the largest assemblages overall. Therefore, the cemeteries' largest tombs seem a sensible starting point for investigating the dynamics of this late rise in burial activity.

```{r toptombs, fig.dim = c(7, 5), dpi=600, fig.align='center',fig.cap="Aoristic sums through time for the 20 tombs with the largest ceramic assemblages when excavated, arranged in decreasing order. Tomb names/numbers are shown above each plot. Note, for the sake of better illustrating the relative temporal patterns, rather than absolute values, the scales on each plot are distinct. The total number of vessels in each case is shown."}

tombs_ranked<-tombs[with(tombs, order(-tombs$total_pottery)), ]
tombs_top_20<-head(tombs_ranked,20)
top20<-as.vector(tombs_top_20$tomb)
pots_top_20<-pots[is.element(pots$tomb, top20),]
pots_top_20$tomb<-factor(pots_top_20$tomb,levels=top20)

cols<-c(1,160:209)
aorist_all_tomb_names<-tombs_top_20[,cols]
aorist_all_tomb_names<-data.frame(aorist_all_tomb_names)
colnames(aorist_all_tomb_names)<-c("tomb",dates)

aorist_all_tomb_names$cemetery<-tombs_top_20$cem_condensed

aorist_all_tomb_melt<-melt(aorist_all_tomb_names)
aorist_all_tomb_melt$variable<-as.numeric(as.character(aorist_all_tomb_melt$variable))

top_20<-tombs_top_20$tomb
aorist_all_tomb_melt$tomb<-factor(aorist_all_tomb_melt$tomb,levels=top_20)

tombs_top_20[, "max"] <- apply(tombs_top_20[, 160:209], 1, max)
tombs_top_20$total_pottery <- paste0("n=", as.factor(tombs_top_20$total_pottery))

ggplot()+
  geom_area(data=aorist_all_tomb_melt,aes(variable, value,fill=cemetery))+theme(legend.position = "none")+
  geom_text(data=tombs_top_20,aes(x=1000,y=max*0.9,label=total_pottery),family="serif",size=3)+
  theme_bw()+theme(legend.title=element_blank(),legend.position="bottom", legend.text = element_text(family="serif",size=9),
      text=element_text(family="serif", size=11), panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
      axis.text.x=element_text(angle=45, vjust=1.2, hjust=1), legend.margin=margin(0.5,0,0,0),legend.box.margin=margin(1,-10,1,-10))+
          labs(x="Date (Years BCE)", y="Aoristic sum")+
              scale_fill_manual(values = c("gray80","gray50"),labels=c(" KNC  "," Fortetsa"))+ 
                  coord_cartesian(xlim=c(1100,610))+scale_y_continuous(expand = expansion(mult = c(0, 0.05)))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,610))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))+facet_wrap(~tomb,scales="free_y")+
                       theme(strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA),
                             strip.text.x = element_text(size=8,margin = margin(0.05,0,0.05,0, "cm")))

```

The 20 tombs which yielded the most ceramic vessels are all chamber tombs, except for the unidentified T.134, and all but 294, 207 and P2 date to the 10th or 9th centuries (Fig. \@ref(fig:toptombs)). Several likewise evince late episodes of substantial use: Tombs P, 292, 75, 107, II and 218 for instance. Meanwhile, tombs including X, G, Q, 283 and 100 show more marked activity about a century earlier, in the 9th century. These patterns are suggestive, but a more robust method of sorting the tombs presents itself in the use of cluster analysis. As discussed above, this technique been employed by Cavanagh (1996) and Antoniadis (2017), in the first instance using tomb dimensions to discriminate EIA from LBA tombs, and in the second using a combination of artefact counts and dimensions to identify tombs of similar character and use. The aoristic sums calculated for the period under investigation amount to a new class of quantitative data with which to perform such an analysis, one which should ideally group tombs based on their chronological patterns of use.

```{r cluster, fig.dim=c(3.5,4.1),dpi=600,fig.align='center',out.width="0.5%", fig.cap="Results of cluster analysis based on aoristic sums per decade for each tomb. Ward's method was used and tombs that yielded no pots at all were excluded. Four groups derived using a k-means clustering. Names/numbers of all tombs aside from those of Cluster 1 are shown." }

df_prep<-subset(tombs,total_pottery>0)
df<-df_prep[-c(2:159,210:260)]
row.names(df)<-df[,1]
df[,1]<-NULL
df<-scale(df)
hc3 <- agnes(df, method = "ward")
sub_grp <- cutree(hc3, k = 4)

df_prep$cluster<-sub_grp
df_prep2<-df_prep
df_prep2 <- within(df_prep2, tomb[cluster<2] <- NA)

cbPalette <- c("#E69F00","#999999", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

fviz_cluster(list(data = df, cluster = sub_grp),ellipse.alpha=0.1,pointsize=1,
             show.clust.cent=FALSE,main=NULL,xlab="Dimension 1",ylab="Dimension 2",
             ggtheme=theme_bw()+theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank(),
                                      text=element_text(family="serif", size=10),legend.position = "bottom"),
             geom="point",legend.text=element_text(size=10))+
  geom_text_repel(label=df_prep2$tomb,color="Black",family="serif",size=2.5)+
                scale_fill_manual('Cluster',values = cbPalette)+scale_color_manual('Cluster',values = cbPalette)+
  scale_shape_manual('Cluster', values=c(15,16,17,18))+scale_y_reverse()

```

This analysis was done using Ward's method, and Fig. \@ref(fig:cluster) presents the results as a scatterplot, with the four groups separated by way of a k-means clustering. Cluster 1 comprises the vast majority of tombs in both cemeteries, while tombs Π (Pi) and 40 appear to be outliers. However, Clusters 3 and 4 reflect remarkably closely the preliminary observations above; Cluster 3 includes all those tombs which in Fig. \@ref(fig:toptombs) exhibit a late period of increased burial activity, while Cluster 4 largely comprises those with an earlier peak in depositions. All tombs in both clusters are chamber tombs, dating mostly to the 10th century in Cluster 4, and all to the 9th century in Cluster 3. Given that just the aoristic sums were submitted to analysis, it is encouraging that the results follow so closely an intuitive assessment of the data, and that tombs from both cemeteries appear in each group, suggesting from the outset that the differences between these may speak to social forces operating across the Knossian community. Tomb II is perhaps a notable absence, being the top-left-most point of Cluster 1, and we might choose to consider it a close relative of Cluster 3, but for the time being the results of the analysis will be adhered to. From now on, Clusters 3 and 4 shall be referred to as Groups A and B respectively.

```{r groupsab, fig.dim=c(5,6),dpi=600, fig.align='center', fig.cap="A) Aoristic sums through time for the Group A and Group B tombs, as identified through the cluster analysis discussed above, and all remaining tombs. B) Aoristic sums divided in each decade by the number of tombs from which they were calculated (i.e. excluding all those with values of 0)."}

GroupA<-subset(df_prep,cluster==3)

aorist_GroupA <- colSums(GroupA[160:209])
aorist_GroupA<-data.frame(aorist_GroupA)

colnames(aorist_GroupA)<-c("aoristic_sum")
row.names(aorist_GroupA)<-dates
aorist_GroupA<-data.frame(aorist_GroupA)
setDT(aorist_GroupA, keep.rownames = "decade")
aorist_GroupA$decade<-as.numeric(aorist_GroupA$decade)
aorist_GroupA$no_tombs<-colSums(GroupA[160:209]!=0)

aorist_GroupA$sum_over_tombs<-ifelse(is.na(aorist_GroupA$aoristic_sum/aorist_GroupA$no_tombs),0,
                                     (aorist_GroupA$aoristic_sum/aorist_GroupA$no_tombs))

aorist_GroupA$group<-"Group A"

GroupB<-subset(df_prep,cluster==4)

aorist_GroupB <-colSums(GroupB[160:209])
aorist_GroupB<-data.frame(aorist_GroupB)

colnames(aorist_GroupB)<-c("aoristic_sum")
row.names(aorist_GroupB)<-dates
aorist_GroupB<-data.frame(aorist_GroupB)
setDT(aorist_GroupB, keep.rownames = "decade")
aorist_GroupB$decade<-as.numeric(aorist_GroupB$decade)
aorist_GroupB$no_tombs<-colSums(GroupB[160:209]!=0)

aorist_GroupB$sum_over_tombs<-ifelse(is.na(aorist_GroupB$aoristic_sum/aorist_GroupB$no_tombs),0,
                                     (aorist_GroupB$aoristic_sum/aorist_GroupB$no_tombs))

aorist_GroupB$group<-"Group B"

Other<-subset(df_prep,cluster==1|cluster==2)

aorist_Other <-colSums(Other[160:209])
aorist_Other<-data.frame(aorist_Other)

colnames(aorist_Other)<-c("aoristic_sum")
row.names(aorist_Other)<-dates
aorist_Other<-data.frame(aorist_Other)
setDT(aorist_Other, keep.rownames = "decade")
aorist_Other$decade<-as.numeric(aorist_Other$decade)
aorist_Other$no_tombs<-colSums(Other[160:209]!=0)

aorist_Other$sum_over_tombs<-ifelse(is.na(aorist_Other$aoristic_sum/aorist_Other$no_tombs),0,
                                     (aorist_Other$aoristic_sum/aorist_Other$no_tombs))

aorist_Other$group<-"Other"

aorist_groups<-rbind(aorist_GroupA,aorist_GroupB,aorist_Other)

unscaled<-ggplot()+geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf), fill=NA, colour="grey90",alpha = 0.4) +
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = 118, ymax = 126), fill=NA, colour="black",size = 0.5) +
  geom_text(data=rects,aes(xmid, 122,label=col),family="serif",size=2.5)+
  geom_line(data=aorist_groups,aes(decade,aoristic_sum,linetype=group,group=group))+
  theme_bw()+theme_bw()+theme(text=element_text(family="serif", size=10),
                              panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                              legend.title = element_blank(),
                              plot.margin=unit(c(0.1,0.3,0.1,0.1),"cm"))+
    labs(x="Date (Years BCE)", y="Aoristic sum/No. of tombs")+
  scale_y_continuous(expand=c(0,0),limits=c(0,126))+
  scale_x_reverse(expand = c(0, 0),limits=c(1100,600))

scaled<-ggplot()+  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf), fill=NA, colour="grey90",alpha = 0.4) +
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = 13.1, ymax = 14), fill=NA, colour="black",size = 0.5) +
  geom_text(data=rects,aes(xmid, 13.55,label=col),family="serif",size=2.5)+
  geom_line(data=aorist_groups,aes(decade,sum_over_tombs,linetype=group,group=group),
            plot.margin=unit(c(0.1,0.3,0.1,0.1),"cm"))+
  theme_bw()+theme_bw()+theme(text=element_text(family="serif", size=10),
                              panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                              legend.title = element_blank())+
    labs(x="Date (Years BCE)", y="Aoristic sum/No. of tombs")+
  scale_y_continuous(expand=c(0,0),limits=c(0,14))+
  scale_x_reverse(expand = c(0, 0),limits=c(1100,600))

ggarrange(unscaled,scaled,nrow=2,labels=c("A","B"),align="v",common.legend=TRUE, legend="bottom")

```

The divergence in use between Groups A and B can be seen in Fig. \@ref(fig:groupsab), which shows the aoristic sums through time of these two groups, as well as of the remaining tombs. Where Group A embodies the late rise in burial activity already noted, Group B peaked instead in the PGB-EG periods, and then declined. The final group, as an aggregate of all remaining tombs, understandably presents an intermediate pattern. The different scale of the depositions made in the tombs of Group A and B can be seen clearly in Fig. \@ref(fig:groupsab)B, where the aoristic sums in each decade are divided by the number of tombs from which they came. Groups A and B roughly retain their relative scale, while the signal from the remaining tombs appears much reduced. That is, these other tombs received far fewer burials or offerings per tomb.

```{r table2}

pots$cluster <- df_prep$cluster[match(pots$tomb,df_prep$tomb)]
pots$cluster[is.na(pots$cluster)]<- "Other"
pots$cluster[pots$cluster<3]<- "Other"
pots$cluster[pots$cluster==3]<- "Group A"
pots$cluster[pots$cluster==4]<- "Group B"

pithoi$cluster <- df_prep$cluster[match(pithoi$tomb,df_prep$tomb)]
pithoi$cluster[is.na(pithoi$cluster)]<- "Other"
pithoi$cluster[pithoi$cluster<3]<- "Other"
pithoi$cluster[pithoi$cluster==3]<- "Group A"
pithoi$cluster[pithoi$cluster==4]<- "Group B"

tombs$cluster <- df_prep$cluster[match(tombs$tomb,df_prep$tomb)]
tombs$cluster[is.na(tombs$cluster)]<- "Other"
tombs$cluster[tombs$cluster<3]<- "Other"
tombs$cluster[tombs$cluster==3]<- "Group A"
tombs$cluster[tombs$cluster==4]<- "Group B"

potsA10th<-sum(pithoi$cluster=="Group A" & pithoi$century=="10th",na.rm=TRUE)
potsA9th<-sum(pithoi$cluster=="Group A" & pithoi$century=="9th",na.rm=TRUE)
potsA8th<-sum(pithoi$cluster=="Group A" & pithoi$century=="8th",na.rm=TRUE)
potsA7th<-sum(pithoi$cluster=="Group A" & pithoi$century=="7th",na.rm=TRUE)

notombsA10th<-nrow(unique(subset(pithoi,cluster=="Group A" & century=="10th",select=tomb)))
notombsA9th<-nrow(unique(subset(pithoi,cluster=="Group A" & century=="9th",select=tomb)))
notombsA8th<-nrow(unique(subset(pithoi,cluster=="Group A" & century=="8th",select=tomb)))
notombsA7th<-nrow(unique(subset(pithoi,cluster=="Group A" & century=="7th",select=tomb)))

potsB10th<-sum(pithoi$cluster=="Group B" & pithoi$century=="10th",na.rm=TRUE)
potsB9th<-sum(pithoi$cluster=="Group B" & pithoi$century=="9th",na.rm=TRUE)
potsB8th<-sum(pithoi$cluster=="Group B" & pithoi$century=="8th",na.rm=TRUE)
potsB7th<-sum(pithoi$cluster=="Group B" & pithoi$century=="7th",na.rm=TRUE)

notombsB10th<-nrow(unique(subset(pithoi,cluster=="Group B" & century=="10th",select=tomb)))
notombsB9th<-nrow(unique(subset(pithoi,cluster=="Group B" & century=="9th",select=tomb)))
notombsB8th<-nrow(unique(subset(pithoi,cluster=="Group B" & century=="8th",select=tomb)))
notombsB7th<-nrow(unique(subset(pithoi,cluster=="Group B" & century=="7th",select=tomb)))

potsother10th<-sum(pithoi$cluster=="Other" & pithoi$century=="10th",na.rm=TRUE)
potsother9th<-sum(pithoi$cluster=="Other" & pithoi$century=="9th",na.rm=TRUE)
potsother8th<-sum(pithoi$cluster=="Other" & pithoi$century=="8th",na.rm=TRUE)
potsother7th<-sum(pithoi$cluster=="Other" & pithoi$century=="7th",na.rm=TRUE)

notombsother10th<-nrow(unique(subset(pithoi,cluster=="Other" & century=="10th",select=tomb)))
notombsother9th<-nrow(unique(subset(pithoi,cluster=="Other" & century=="9th",select=tomb)))
notombsother8th<-nrow(unique(subset(pithoi,cluster=="Other" & century=="8th",select=tomb)))
notombsother7th<-nrow(unique(subset(pithoi,cluster=="Other" & century=="7th",select=tomb)))

```

|**Tombs**          | **Date of *pithoi* (burials)**| **No. of *pithoi* (burials)**| **Number of tombs represented** | **Burials per tomb**                   |
|-------------------|--------------------------|-------------------------|-----------------------------|--------------------------------------------------|
|*Group A*|                         |                         |                             |                                                  |
|                   |9th century               | `r potsA9th`         | `r notombsA9th`          | `r round(potsA9th/notombsA9th,digits=2)`   |
|                   |8th century               | `r potsA8th`         | `r notombsA8th`          | `r round(potsA8th/notombsA8th,digits=2)`   |
|                   |7th century               | `r potsA7th`         | `r notombsA7th`          | `r round(potsA7th/notombsA7th,digits=2)`   |
|*Group B*|                          |                         |                             |                                                  |
|                   |10th century              | `r potsB10th`        | `r notombsB10th`         | `r round(potsB10th/notombsB10th,digits=2)` |
|                   |9th century               | `r potsB9th`          | `r notombsB9th`           | `r round(potsB9th/notombsB9th,digits=2)`     |
|                   |8th century               | `r potsB8th`          | `r notombsB8th`           | `r round(potsB8th/notombsB8th,digits=2)`     |
|                   |7th century               | `r potsB7th`          | `r notombsB7th`           | `r round(potsB7th/notombsB7th,digits=2)`     |
|*Other tombs*|                          |                         |                             |                                                  |
|                   |10th century              | `r potsother10th`        | `r notombsother10th`         | `r round(potsother10th/notombsother10th,digits=2)` |
|                   |9th century               | `r potsother9th`          | `r notombsother9th`           | `r round(potsother9th/notombsother9th,digits=2)`     |
|                   |8th century               | `r potsother8th`          | `r notombsother8th`           | `r round(potsother8th/notombsother8th,digits=2)`     |
|                   |7th century               | `r potsother7th`          | `r notombsother7th`           | `r round(potsother7th/notombsother7th,digits=2)`     |

*Table 2. The number of pithoi (as a proxy for burials) deposited in the tombs of Group A, Group B, and all other tombs. For each of the centuries the tombs were in use (column two), counts of pithoi (column three) are divided by the number of tombs from which these vessels came (column four), allowing calculation of the changing rate of cremation burial through time (column five).*

The singular pattern of use among the Group A tombs is apparent in Table 2. We can be quite certain that the burying populations already represented a small sub-section of the Knossian community, and for most tombs in most periods the burial rite itself was even more selective. Rates of cremation burial never top 3.5 per tomb per century among the vast majority of the tombs - barely one per generation - while among those of Group B they peak at 6.5. In the Group A tombs, however, the large numbers of *pithoi* deposited in the 8th and 7th centuries represent up to 18 burials per tomb per century, or one every five to six years. A general increase in cremation burials in the 8th century has been noted before (Cavanagh 1996), and more recently argued to be too great and rapid to be explained by population growth alone [@Kotsonas2011]. Based on Table 2, it would be more accurate to say that it is the disproportionate concentration of this increase among the Group A tombs that makes purely demographic explanations unlikely.

## 4.3 Not All Tombs are Created Equal

Having thus sketched out these differing patterns of use, particularly the contrasting trajectories of the Group A and B tombs, it is necessary to examine in what other ways, if any, these tombs might have differed from those around them. Though many are not large, the physical dimensions of the tombs as built represent a plausible context for the negotiation and display of social power and group membership, both in terms of the labour expended in construction, and the greater impression larger or more elaborate tombs may have made.

```{r tombgroupdimensions, fig.dim = c(7, 4),dpi=600, fig.align='center', fig.cap="A) Histograms showing the distribution of chamber tomb chamber areas, in square meters. B) The same but for dromos areas. In each case, these measurements are planimetric rather than volumetric, as accurate heights and depths for chambers and dromoi are so rarely recorded."}

cbPalette <- c("#56B4E9", "#009E73","#E69F00")

chamber_area<-ggplot(data=subset(tombs, chamber_area>0),aes(chamber_area,fill=cluster,color=cluster))+
  geom_histogram(binwidth = 0.5,size=0.4)+
    theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                     panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                     legend.text=element_text(size=10))+
            labs(x="Chamber area ("~m^2~")", y="Count of tombs per 0.5"~m^2~"")+
                        scale_fill_manual(values=cbPalette)+scale_color_manual(values=c("black","black","black"))+
                             scale_x_continuous(expand=expansion(mult=c(0,0.05)),breaks = seq(0, 10, by = 2))+
                  scale_y_continuous(expand=expansion(mult=c(0,0.05)),breaks = seq(0, 14, by = 2),limits=c(0,14))

dromos_area<-ggplot(data=subset(tombs, dromos_area>0),aes(dromos_area,fill=cluster,color=cluster))+
  geom_histogram(binwidth = 0.5,size=0.4)+
    theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                     panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                     legend.text=element_text(size=10))+
            labs(x="Dromos area ("~m^2~")", y="Count of tombs per 0.5"~m^2~"")+
                        scale_fill_manual(values=cbPalette)+scale_color_manual(values=c("black","black","black"))+
                             scale_x_continuous(expand=expansion(mult=c(0,0.05)),breaks = seq(0, 12, by = 2))+
                  scale_y_continuous(expand=expansion(mult=c(0,0.05)),breaks = seq(0, 14, by = 2),limits=c(0,14))

ggarrange(chamber_area,dromos_area,ncol=2, labels=c("A","B"),align="hv",common.legend = TRUE, legend="bottom",widths=c(10,12))

```

However, in considering those tombs where chamber and *dromos* area are calculable (Fig. \@ref(fig:tombgroupdimensions)), Groups A and B do not seem systematically marked out by their scale. The *dromos*, as the most visible area with regard to the funerary rite, might be considered the more indicative measure, and indeed three Group A and B tombs (75, 219 and 100) had very substantial *dromoi.* These features were also commonly disturbed by later activity, meaning other long *dromoi* may have gone unrecognised, but on the present evidence there is no general demarcation between the tombs based on size.

```{r tombdimensions, fig.dim = c(7, 4), dpi=600, fig.align='center', fig.cap="Scatterplots showing A) chamber areas of 10th century tombs, plotted against the number of 10th and 9th century vessels they yielded respectively. B) The same but using combined chamber and *dromos* areas. C) As A, but with tombs of the 9th century and vessels of the 9th and 8th centuries. D) As B, but with tombs of the 9th century and vessels of the 9th and 8th centuries."}

cbPalette <- c("#56B4E9", "#009E73","#E69F00")

  cham_9th_pots_9th<-ggplot()+geom_point(data=subset(tombs,century_built=="9th"),aes(chamber_area,vessels_9th,color=cluster),size=1)+  stat_cor(data=subset(tombs,century_built=="9th"),aes(chamber_area,vessels_9th),method="pearson",label.x.npc=0.1,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(x="Chamber area ("~m^2~")", y="9th century vessels")+
                          scale_color_manual(values=cbPalette)+
                               scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

  cham_9th_pots_8th<-ggplot()+geom_point(data=subset(tombs,century_built=="9th"),aes(chamber_area,vessels_8th,color=cluster),size=1)+  stat_cor(data=subset(tombs,century_built=="9th"),aes(chamber_area,vessels_8th),method="pearson",label.x.npc=0.6,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(x="Chamber area ("~m^2~")", y="8th century vessels")+
                          scale_color_manual(values=cbPalette)+
                               scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

    dim_9th_pots_9th<-ggplot()+geom_point(data=subset(tombs,century_built=="9th"),aes(chamber_area+dromos_area,vessels_9th,color=cluster),size=1)+  stat_cor(data=subset(tombs,century_built=="9th"),aes(chamber_area+dromos_area,vessels_9th),method="pearson",label.x.npc=0.1,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10),axis.title.x = element_text(hjust=1))+
              labs(x="Dromos+chamber area ("~m^2~")", y="9th century vessels")+
                          scale_color_manual(values=cbPalette)+
                               scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

  dim_9th_pots_8th<-ggplot()+geom_point(data=subset(tombs,century_built=="9th"),aes(chamber_area+dromos_area,vessels_8th,color=cluster),size=1)+  stat_cor(data=subset(tombs,century_built=="9th"),aes(chamber_area+dromos_area,vessels_8th),method="pearson",label.x.npc=0.1,label.y.npc=0.8,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10),axis.title.x = element_text(hjust=1))+
              labs(x="Dromos+chamber area ("~m^2~")", y="8th century vessels")+
                          scale_color_manual(values=cbPalette)+
                               scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

  cbPalette <- c("#009E73","#E69F00")

  cham_10th_pots_10th<-ggplot()+geom_point(data=subset(tombs,century_built=="10th"),aes(chamber_area,vessels_10th,color=cluster),size=1)+  stat_cor(data=subset(tombs,century_built=="10th"),aes(chamber_area,vessels_10th),method="pearson",label.x.npc=0.6,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(x="Chamber area ("~m^2~")", y="10th century vessels")+
                          scale_color_manual(values=cbPalette)+
                               scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

  cham_10th_pots_9th<-ggplot()+geom_point(data=subset(tombs,century_built=="10th"),aes(chamber_area,vessels_9th,color=cluster),size=1)+  stat_cor(data=subset(tombs,century_built=="10th"),aes(chamber_area,vessels_9th),method="pearson",label.x.npc=0.1,label.y.npc=0.85,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10))+
              labs(x="Chamber area ("~m^2~")", y="9th century vessels")+
                          scale_color_manual(values=cbPalette)+
                               scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

    dim_10th_pots_10th<-ggplot()+geom_point(data=subset(tombs,century_built=="10th"),aes(chamber_area+dromos_area,vessels_10th,color=cluster),size=1)+  stat_cor(data=subset(tombs,century_built=="10th"),aes(chamber_area+dromos_area,vessels_10th),method="pearson",label.x.npc=0.6,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10),axis.title.x = element_text(hjust=1))+
              labs(x="Dromos+chamber area ("~m^2~")", y="10th century vessels")+
                          scale_color_manual(values=cbPalette)+
                               scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

  dim_10th_pots_9th<-ggplot()+geom_point(data=subset(tombs,century_built=="10th"),aes(chamber_area+dromos_area,vessels_9th,color=cluster),size=1)+  stat_cor(data=subset(tombs,century_built=="10th"),aes(chamber_area+dromos_area,vessels_9th),method="pearson",label.x.npc=0.1,label.y.npc=0.8,family="serif",size=3,label.sep="\n")+
        theme_bw()+theme(text=element_text(family="serif", size=10),legend.position = "bottom",legend.title = element_blank(),
                       panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                       legend.text=element_text(size=10),axis.title.x = element_text(hjust=1))+
              labs(x="Dromos+chamber area ("~m^2~")", y="9th century vessels")+
                          scale_color_manual(values=cbPalette)+
                               scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
                        scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

 scatters<-ggarrange(ggarrange(cham_10th_pots_10th,cham_10th_pots_9th,dim_10th_pots_10th,dim_10th_pots_9th,
                    ncol=2,nrow=2,align="hv",common.legend = TRUE,legend="bottom",labels=c("A","","B",""),label.y=1.1),
ggarrange(cham_9th_pots_9th,cham_9th_pots_8th,dim_9th_pots_9th,dim_9th_pots_8th,
                    ncol=2,nrow=2,align="hv",common.legend = TRUE,legend="bottom",labels=c("C","","D",""),label.y=1.1),align="hv",widths=c(9.5,10))

 annotate_figure(scatters, top=text_grob("10th Century Tombs                                                            9th Century Tombs",
                 x=0.51,family="serif"))

```

There are hints of differentiation among contemporary tombs, however. Fig. \@ref(fig:tombdimensions)A shows that the Group B tombs of the 10th century were among the largest of the time, and went on to receive the greatest quantities of vessels in the subsequent 9th century. It is difficult to say if this correlation suggests the tombs' builders pre-empted their later use, or if they were motivated by some kind of competitive display (or a combination of these things); however, a similar situation is not immediately clear in the case of the Group A tombs. They were not marked out by their size, even against contemporary tombs, and their later use thus appears unforeseen. This fact is supported by the presence of niches cut into the *dromoi* of some Group A tombs, to fit in later burial urns (including T. 292, 283 and 107; Cavanagh 1996, p.658), while sometimes urns from within the chamber itself were removed to make space for new interments (including T. 75 and TFT; ibid., p.658; Brock 1957, pp.3–4). The impression is of competition for inclusion into the sepulchral group, demand outstripping supply as it were, further confirming that, whatever accounted for the 8th-7th century growth in burial activity, the tombs' original builders little anticipated it.

## 4.4 Imports, Interments, and Indices of Wealth

In the past it has been conventional to talk about the tombs with the most finds (i.e. those of Groups A and B) as the 'richest' or possessing the most 'wealth items' [e.g. @Coldstream1996d 718; @Wallace2010a 304-311]. They certainly yielded more vessels, more imports, more weapons and so on than other tombs. However, it will be suggested here that such statements fail to capture the complex realities behind the use of these tombs through time, and their relationship to wider trends across both cemeteries. To begin on a fundamental level, raw quantities of grave goods, in which Groups A and B are doubtless unmatched, have seldom been divided by the estimated number of burials within each tomb.

```{r histpots, fig.dim=c(5,5),dpi=600, fig.align='center',fig.cap="Empirical cumulative distribution function (ECDF) plots showing the distribution of A) total pottery counts across tombs of Group A, Group B, and Others and B) those counts divided by the estimated number of burials in each tomb. C) and D) show the same, but with all grave goods aside from ceramic vessels. These plots are akin to cumulative frequency graphs, with all tombs plotted left to right in ascending order. However, unlike cumulative frequency plots, the totals are not summed; rather the 'tread' of each step corresponds to the actual value registered by one or more tombs. Furthermore, the x-axis does not present counts of tombs, but rather the proportion of all tombs registering less than or equal to the value at that point on the axis. This permits samples of different sizes to be plotted side-by-side, as with the three groups of tombs presented here."}

tombs$cluster <- df_prep$cluster[match(tombs$tomb,df_prep$tomb)]
tombs$cluster[is.na(tombs$cluster)]<- "Other"
tombs$cluster[tombs$cluster<3]<- "Other"
tombs$cluster[tombs$cluster==3]<- "Group A"
tombs$cluster[tombs$cluster==4]<- "Group B"

cbPalette <- c("#56B4E9", "#009E73","#E69F00")

no_pith<-aggregate(pots$shape!="Pithos", by=list(tomb=pots$tomb), FUN=sum)
tombs$pots_no_pith<-no_pith$x[match(tombs$tomb,no_pith$tomb)]

total_pottery<-ggplot(subset(tombs,pots_no_pith>0),aes(pots_no_pith,color=cluster))+stat_ecdf(cex=0.8,alpha=0.6)+coord_flip()+
  theme_bw()+theme(legend.title=element_blank(),
                   text=element_text(family="serif", size=10),panel.grid = element_blank(),
                   legend.text=element_text(size=10))+
  labs(x="Total pottery", y="Proportion of tombs")+
  scale_color_manual(values = cbPalette)+
  scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))

pottery_by_burial<-ggplot(subset(tombs,pots_no_pith>0),aes(pots_no_pith/est_burials,color=cluster))+
  stat_ecdf(cex=0.8,alpha=0.6)+coord_flip()+
  theme_bw()+theme(legend.title=element_blank(),
                   text=element_text(family="serif", size=10),panel.grid = element_blank(),
                   legend.text=element_text(size=10))+
  labs(x="Total pottery/burials", y="Proportion of tombs")+
  scale_color_manual(values = cbPalette)+
  scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))

total_object<-ggplot(subset(tombs,total_object>0),aes(total_object,color=cluster))+stat_ecdf(cex=0.8,alpha=0.6)+coord_flip()+
  theme_bw()+theme(legend.title=element_blank(),
                   text=element_text(family="serif", size=10),panel.grid = element_blank(),
                   legend.text=element_text(size=10))+
  labs(x="Total objects", y="Proportion of tombs")+
  scale_color_manual(values = cbPalette)+
  scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))

object_by_burial<-ggplot(subset(tombs,total_object>0),aes(total_object/est_burials,color=cluster))+
  stat_ecdf(cex=0.8,alpha=0.6)+coord_flip()+
  theme_bw()+theme(legend.title=element_blank(),
                   text=element_text(family="serif", size=10),panel.grid = element_blank(),
                   legend.text=element_text(size=10))+
  labs(x="Total objects/burials", y="Proportion of tombs")+
  scale_color_manual(values = cbPalette)+
  scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))

ggarrange(total_pottery,pottery_by_burial,total_object,object_by_burial,labels=c("A","B","C","D"),common.legend = TRUE,legend="bottom",ncol=2,nrow = 2,widths=c(10,9.6))
```

As seen in Fig. \@ref(fig:histpots), even this simple step quite significantly alters the gap between Groups A and B and the remaining tombs. In the case of pottery, the disparity is much reduced; in that of all grave goods besides pottery, it is almost eradicated. Furthermore, in both instances, the relationship between the Group A and B tombs is reversed, with the latter tending to have yielded more finds *per burial*, if only slightly.

```{r secureassems, fig.dim=c(3,3.2),dpi=600,fig.align='center',fig.cap="ECDF plot showing the size of securely identified single burial assemblages, divided into the tombs of Group A, Group B and all other tombs."}

assemblages$cluster <- df_prep$cluster[match(assemblages$tomb,df_prep$tomb)]
assemblages$cluster[is.na(assemblages$cluster)]<- "Other"
assemblages$cluster[assemblages$cluster<3]<- "Other"
assemblages$cluster[assemblages$cluster==3]<- "Group A"
assemblages$cluster[assemblages$cluster==4]<- "Group B"

assem2<-unique(assemblages$assemblage)

assem2<-data.frame(assem2)
colnames(assem2)<-"assemblage"
assem2$cluster <- assemblages$cluster[match(assem2$assemblage,assemblages$assemblage)]
counts<-count(assemblages,"assemblage")
assem_counts<-merge(assem2,counts)

ggplot(assem_counts,aes(freq,color=cluster))+stat_ecdf(cex=0.8,alpha=0.6)+coord_flip()+
  theme_bw()+theme(legend.title=element_blank(),legend.position = "bottom",
                   text=element_text(family="serif", size=10),panel.grid = element_blank(),
        legend.box.margin=margin(-10,-10,-10,-10),
                   legend.text=element_text(size=10))+
  labs(x="Assemblage size", y="Proportion of tombs")+
  scale_color_manual(values = cbPalette)+
  scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))

```

It might be argued these figures are based on tomb-level counts, and thus average out the differences between individual burials. This is certainly true, and there are a number of more securely identified individual burial assemblages across Fortetsa and the KNC. However, if anything these cases urge greater caution. As seen in Fig. \@ref(fig:secureassems), the grave goods accompanying individual burials could in some cases be numerous, but nothing about Group A or B suggests they were uncommonly characterised by such offerings. They adhere closely to the more general trend. In all cases, between around 75% and 82% of these secure assemblages numbered five items or fewer.

```{r hists, fig.dim=c(7,5),dpi=600, fig.align='center',fig.cap="ECDF plots showing the distribution of A) items of adornment (beads, necklaces, rings, pins, fibulae etc.), weapons, and imported items by tomb across both cemeteries, and B) those counts divided by the estimated number of burials in each tomb."}

total_jewellery<-ggplot(subset(tombs,adornment>0 & jewellery>0),aes(adornment+jewellery,color=cluster))+
  stat_ecdf(cex=0.6,alpha=0.6)+coord_flip()+
  theme_bw()+theme(legend.title=element_blank(),
                   text=element_text(family="serif", size=10),panel.grid = element_blank(),
                   legend.text=element_text(size=10))+
  labs(x="Total adornment", y="Proportion of tombs")+
  scale_color_manual(values = cbPalette)+
  scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))

total_weapon<-ggplot(subset(tombs,weapon>0),aes(weapon,color=cluster))+stat_ecdf(cex=0.6,alpha=0.6)+coord_flip()+
  theme_bw()+theme(legend.title=element_blank(),
                   text=element_text(family="serif", size=10),panel.grid = element_blank(),
                   legend.text=element_text(size=10))+
  labs(x="Total weapons", y="Proportion of tombs")+
  scale_color_manual(values = cbPalette)+
  scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))

total_import<-ggplot(subset(tombs,total_import>0),aes(total_import,color=cluster))+stat_ecdf(cex=0.6,alpha=0.6)+coord_flip()+
  theme_bw()+theme(legend.title=element_blank(),
                   text=element_text(family="serif", size=10),panel.grid = element_blank(),
                   legend.text=element_text(size=10))+
  labs(x="Total imports", y="Proportion of tombs")+
  scale_color_manual(values = cbPalette)+
  scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))

jewellerybyburial<-ggplot(subset(tombs,adornment>0 & jewellery>0),aes((adornment+jewellery)/est_burials,color=cluster))+
  stat_ecdf(cex=0.6,alpha=0.6)+coord_flip()+
  theme_bw()+theme(legend.title=element_blank(),
                   text=element_text(family="serif", size=10),panel.grid = element_blank(),
                   legend.text=element_text(size=10))+
  labs(x="Total adornment/burials", y="Proportion of tombs")+
  scale_color_manual(values = cbPalette)+
  scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))

weaponbyburial<-ggplot(subset(tombs,weapon>0),aes(weapon/est_burials,color=cluster))+
  stat_ecdf(cex=0.6,alpha=0.6)+coord_flip()+
  theme_bw()+theme(legend.title=element_blank(),
                   text=element_text(family="serif", size=10),panel.grid = element_blank(),
                   legend.text=element_text(size=10))+
  labs(x="Total weapons/burials", y="Proportion of tombs")+
  scale_color_manual(values = cbPalette)+
  scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))

importbyburial<-ggplot(subset(tombs,total_import>0),aes(total_import/est_burials,color=cluster))+
  stat_ecdf(cex=0.6,alpha=0.6)+coord_flip()+
  theme_bw()+theme(legend.title=element_blank(),
                   text=element_text(family="serif", size=10),panel.grid = element_blank(),
                   legend.text=element_text(size=10))+
  labs(x="Total imports/burials", y="Proportion of tombs")+
  scale_color_manual(values = cbPalette)+
  scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))

ggarrange(total_jewellery,total_weapon,total_import,jewellerybyburial,weaponbyburial,importbyburial,ncol=3,nrow=2,common.legend = TRUE,legend="bottom",align="hv",labels=c("A","","","B","",""))+
            theme(plot.margin=unit(c(0.1,0.1,0.1,0.1),"cm"))

```

Similar patterns are found among many classes of grave good (Fig. \@ref(fig:histpots)). In both Groups A and B, items of adornment (pins, fibulae, rings, beads etc.) and weapons, when divided by the estimated number of burials in each tomb, fall in line with, or even below, the distribution across all other tombs. As above, the tombs of Group B are less diminished and, indeed, in the case of imports, their pre-eminence is retained even when adjusted by estimated burials.

This last fact is notable for, as discussed, the Group B tombs were at their peak during the PGB and EG periods. These decades are consistently interpreted as a time of artistic development, increased overseas trade, and renewed vigour in the  visibility of the burial rite [@Brock1957 143-44; @Coldstream1994 110-12]. The PGB style, first defined by Brock [-@Brock1957 143], who hailed it as 'the most remarkable phase in Cretan vase-painting', is apparently unique to the island, evincing a melding of Attic Geometric motifs, patterns possibly derived from Near Eastern metalwork, and iconography drawn from the Minoan repertoire [@Coldstream1984b 93-94]. Coldstream has tied the emergence of this style to a reformulation of elite mortuary display, noting how, in this period also, redeposited Minoan *larnakes* began appearing in and around some tombs, in one case even inspiring the decoration of a PGB vessel [in T. 107; see @Coldstream1984b; -@Coldstream1994 112–13]. Whether simply nostalgia [@Coldstream1998 60], or a more tactical strategy by newly emerging elite groups [@Coldstream1994 114-15], the Bronze Age past appears to have possessed a significant cultural cachet for the EIA Knossians.

Crowe [-@Crowe2018] has recently argued that these antiques cannot be linked in any obvious way to the wealth of the burying groups *per se*, with LBA *larnakes*, for instance, being found across some 21 tombs of varying richness (in terms of their quantities of grave goods). Rather, she suggests, BA artefacts found in or looted from earlier tombs formed a relatively unrestricted type of mortuary accessory, often incorporated into the burial rite in ways in-keeping with contemporary traditions. A plurality of burial accoutrements, then, seems a fair characterisation of the late 9th to early 8th centuries, within which BA antiques, vibrantly decorated pottery, and items imported from overseas were variously deployed in the elaboration of funerary rites and offerings.

```{r importaorist,fig.dim=c(5,4),dpi=600, fig.align='center',fig.cap="Aoristic sums through time of imports, divided by the number of tombs from which they came."}

GroupA_imports<-subset(pots,cluster=="Group A" & import=="Y")
GroupA_all<-subset(pots,cluster=="Group A" & shape=="Pithos")

imports_GroupA <- colSums(GroupA_imports[27:76])
imports_GroupA<-data.frame(imports_GroupA)

colnames(imports_GroupA)<-c("aoristic_sum")
row.names(imports_GroupA)<-dates
imports_GroupA<-data.frame(imports_GroupA)
setDT(imports_GroupA, keep.rownames = "decade")
imports_GroupA$decade<-as.numeric(imports_GroupA$decade)
imports_GroupA$no_tombs<-colSums(GroupA[160:209]!=0)

imports_GroupA$group<-"Group A"

GroupB_imports<-subset(pots,cluster=="Group B" & import=="Y")
GroupB_all<-subset(pots,cluster=="Group B" & shape=="Pithos")

imports_GroupB <-colSums(GroupB_imports[27:76])
imports_GroupB<-data.frame(imports_GroupB)

colnames(imports_GroupB)<-c("aoristic_sum")
row.names(imports_GroupB)<-dates
imports_GroupB<-data.frame(imports_GroupB)
setDT(imports_GroupB, keep.rownames = "decade")
imports_GroupB$decade<-as.numeric(imports_GroupB$decade)
imports_GroupB$no_tombs<-colSums(GroupB[160:209]!=0)

imports_GroupB$group<-"Group B"

Other_imports<-subset(pots,cluster=="Other" & import=="Y")
Other_all<-subset(pots,cluster=="Other" & shape=="Pithos")

imports_Other <-colSums(Other_imports[27:76])
imports_Other<-data.frame(imports_Other)

colnames(imports_Other)<-c("aoristic_sum")
row.names(imports_Other)<-dates
imports_Other<-data.frame(imports_Other)
setDT(imports_Other, keep.rownames = "decade")
imports_Other$decade<-as.numeric(imports_Other$decade)
imports_Other$no_tombs<-colSums(Other[160:209]!=0)

imports_Other$group<-"Other"

imports_groups<-rbind(imports_GroupA,imports_GroupB,imports_Other)

ggplot()+geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf), fill=NA, colour="grey90",alpha = 0.4) +
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = 1.15, ymax = 1.2), fill=NA, colour="black",alpha = 0.4) +
  geom_text(data=rects,aes(xmid, 1.175,label=col),family="serif",size=2.5)+
  geom_line(data=imports_groups,aes(decade,aoristic_sum/no_tombs,linetype=group))+
  theme_bw()+theme_bw()+theme(text=element_text(family="serif", size=10), legend.position = "bottom",
                              panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                              legend.title = element_blank(),
        legend.box.margin=margin(-10,-10,-10,-10),
                   legend.text=element_text(size=10))+
    labs(x="Date (Years BCE)", y="Aoristic sum of imports/No. of tombs")+
  scale_y_continuous(expand=c(0,0),limits=c(0,1.2),breaks = seq(0, 1.1, by = 0.2))+
  scale_x_reverse(expand = c(0, 0),limits=c(970,630))

```

It must be stressed that the Group A tombs were not dislocated from these trends. As shown in Fig. \@ref(fig:importaorist), by the middle of EG they were receiving more imports per tomb than even the Group B tombs, and both Groups' trajectories through the late 9th and early 8th centuries are similar with regard to these items. Following this period, imports declined, more muted burial forms proliferated as chamber tomb construction probably ceased, and the rate of burial among the Group A tombs grew substantially. All this suggests that, by the end of the 8th century, quite different social conditions had ensued.

```{r gini, fig.dim=c(5,3.2),dpi=600, fig.align='center',fig.cap="Gini coefficient values through time based on aoristic sums for all vessels across both cemeteries."}

cbPalette <- c("#56B4E9", "#009E73","#E69F00")

pots$cluster <- df_prep$cluster[match(pots$tomb,df_prep$tomb)]
pots$cluster[is.na(pots$cluster)]<- "Other"
pots$cluster[pots$cluster<3]<- "Other"
pots$cluster[pots$cluster==3]<- "Group A"
pots$cluster[pots$cluster==4]<- "Group B"

aorist_tomb_values<-tombs[c(1,160:209)]
aorist_tomb_values[aorist_tomb_values == 0] <- NA

aorist_tomb_values<-data.frame(aorist_tomb_values)
colnames(aorist_tomb_values)[2:51]<-dates

tombs_gini<-matrix(0, ncol = 0, nrow = 1)
tombs_gini<-data.frame(tombs_gini)
for(i in colnames(aorist_tomb_values[2:51])) {
  tombs_gini[[paste0(i)]] <- ineq(aorist_tomb_values[[i]],type="Gini")
}

tombs_gini<-t(tombs_gini)
tombs_gini<-data.frame(tombs_gini)

colnames(tombs_gini)<-c("gini_coefficient")
row.names(tombs_gini)<-dates
tombs_gini<-data.frame(tombs_gini)
setDT(tombs_gini, keep.rownames = "decade")
tombs_gini$decade<-as.numeric(tombs_gini$decade)

rects <- data.frame(xstart = c(1100,970,920,875,840,810,790,745,710,670,630), 
                    xend =    c(970,920,875,840,810,790,745,710,670,630,600), 
                    xmid =   c(1035,945,897.5,857.5,825,800,767.5,727.5,690,650,615),
                    col = c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))
rects$col<-factor(rects$col,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO","A"))

ggplot() + geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf), fill=NA, colour="grey90",alpha = 0.4) +
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = 0.786, ymax = 0.8), fill=NA, colour="black",alpha = 0.4) +
    geom_text(data=rects,aes(xmid, 0.793,label=col),family="serif",size=2.5)+
    geom_line(data = tombs_gini, aes(decade,gini_coefficient),size=0.5)+
      theme_bw()+theme(text=element_text(family="serif", size=10),
                       panel.grid = element_blank())+
              labs(x="Date (Years BCE)", y="Gini Coefficient")+ coord_cartesian(
                    xlim = c(1100, 600), ylim = c(0.5,0.8))+ scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
                       scale_y_continuous(expand=c(0,0))
```

The contrasts between these two periods are further evidenced in the Gini coefficient for the distribution of pottery between the tombs (Fig. \@ref(fig:gini)). Using aoristic sums as a probabilistically weighted estimate of the quantity of vessels deposited in each tomb in each decade, the Gini coefficient can give an (albeit rough) estimate of how unequally distributed those offerings were. Although values were always relatively high (i.e. unequal), the major spike in the Gini coefficient seen at the end of LG does contrast markedly with the plateau of the PGB-MG periods. This is not to suggest that the late 9th to early 8th centuries were egalitarian by any means - they do, after all, represent a slightly *more* unequal period than those preceding and succeeding them. But the general trend does intuitively accord with the image sketched above, of more plural modes of burial giving way to practices increasingly polarised between the overflowing tombs of Group A, and the remaining chamber tombs and more ephemeral types constructed in later years.

```{r comps, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="The composition of the total pottery assemblage for each century across each cemetery (excluding Groups A and B), among the Group A and B tombs, and across all tombs in both cemeteries. The number of vessels in each case is indicated at the top of each bar."}

pots_no_pith<-subset(pots,type!="Pithos"&type!="Lid")

cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

potcompKNC<-ggplot(subset(pots_no_pith,century!="" & cem_condensed=="KNC" & cluster=="Other"),aes(century,fill=type))+geom_bar(position="fill")+
  theme_bw()+theme(plot.title = element_text(hjust=0.5), axis.title.x = element_blank(),
               text=element_text(family="serif", size=11),
               plot.margin=margin(0.1,0.1,0.1,0.1),legend.title.align=0.5)+
                 labs(y="Proportion of total assemblage")+ scale_fill_manual(values=cbPalette)+ 
                    scale_y_continuous(expand=c(0,0.0))+scale_x_discrete(expand=c(0,0.0))+
  guides(fill=guide_legend(title="Century BCE",title.position = "top",nrow=1))

potcompFOR<-ggplot(subset(pots_no_pith,century!="" & cem_condensed=="FOR" & cluster=="Other"),aes(century,fill=type))+geom_bar(position="fill")+
  theme_bw()+theme(plot.title = element_text(hjust=0.5),axis.title.x = element_blank(),axis.text.y = element_blank(),
               text=element_text(family="serif", size=11),
               plot.margin=margin(0.1,0.1,0.1,0.1),legend.title.align=0.5)+
                 labs(y="")+
                  scale_fill_manual(values=cbPalette)+ 
                    scale_y_continuous(expand=c(0,0.0))+scale_x_discrete(expand=c(0,0.0))+
  guides(fill=guide_legend(title="Century BCE",title.position = "top",nrow=1))

potcompA<-ggplot(subset(pots_no_pith,century!="" & cluster=="Group A"),aes(century,fill=type))+geom_bar(position="fill")+
  theme_bw()+theme(plot.title = element_text(hjust=0.5),axis.title.x = element_blank(),axis.text.y = element_blank(),
               text=element_text(family="serif", size=11),
               plot.margin=margin(0.1,0.1,0.1,0.1),legend.title.align=0.5)+
                 labs(y="")+
                  scale_fill_manual(values=cbPalette)+ scale_x_discrete(expand=c(0,0.0))+
                    scale_y_continuous(expand=c(0,0.0))+guides(fill=guide_legend(title="Century BCE",title.position = "top",nrow=1))

potcompB<-ggplot(subset(pots_no_pith,century!="" & cluster=="Group B"),aes(century,fill=type))+geom_bar(position="fill")+
  theme_bw()+theme(plot.title = element_text(hjust=0.5),axis.title.x = element_blank(),axis.text.y = element_blank(),
               text=element_text(family="serif", size=11),
               plot.margin=margin(0.1,0.1,0.1,0.1),legend.title.align=0.5)+
                 labs(y="")+
                  scale_fill_manual(values=cbPalette)+ scale_x_discrete(expand=c(0,0.0))+
                    scale_y_continuous(expand=c(0,0.0))+guides(fill=guide_legend(title="Century BCE",title.position = "top",nrow=1))

potcomp_all<-ggplot(subset(pots_no_pith,!is.na(century)),aes(century,fill=type))+geom_bar(position="fill")+
  theme_bw()+theme(plot.title = element_text(hjust=0.5),axis.title.x = element_blank(),axis.text.y = element_blank(),
               text=element_text(family="serif", size=11),
               plot.margin=margin(0.1,0.1,0.1,0.1),legend.title.align=0.5)+
                 labs(y="")+
                  scale_fill_manual(values=cbPalette)+ scale_x_discrete(expand=c(0,0.0))+
                    scale_y_continuous(expand=c(0,0.0))+guides(fill=guide_legend(title="Century BCE",title.position = "top",nrow=1))
###

counts <- ddply(pots_no_pith, .(pots_no_pith$cluster, pots_no_pith$cem_condensed, pots_no_pith$century), nrow)
names(counts) <- c("group","cem", "century", "Freq")
counts<-counts[complete.cases(counts), ]
counts$level<-10

counts2 <- ddply(pots_no_pith, .(pots_no_pith$cluster, pots_no_pith$century), nrow)
names(counts2) <- c("group", "century", "Freq")
counts2<-counts2[complete.cases(counts2), ]
counts2$level<-10

countsall <- ddply(pots_no_pith, .(pots_no_pith$century), nrow)
names(countsall) <- c("century", "Freq")
countsall$level<-10

potcompKNC_total<-ggplot(subset(counts,group=="Other" & cem=="KNC"),aes(century,level,label=Freq))+geom_text(family="serif",size=3)+
  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title = element_blank(),
               plot.margin=margin(0.1,0.1,0.1,30.5), text=element_text(family="serif", size=9),
               axis.title.x=element_blank(),axis.text.x=element_blank(),axis.ticks.x=element_blank(),
               axis.title.y=element_blank(),axis.text.y=element_blank(),axis.ticks.y=element_blank(),
               panel.grid = element_blank())+geom_vline(xintercept = seq(0.5, length(counts$century),
                                                                         by = 1), color="black", size=.5)+
                scale_y_continuous(expand=c(0,0),limits=c(0,20))+scale_x_discrete(expand=c(0,0.5))+labs(title="KNC")

potcompFOR_total<-ggplot(subset(counts,group=="Other" & cem=="FOR"),aes(century,level,label=Freq))+geom_text(family="serif",size=3)+
  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title = element_blank(),
               plot.margin=margin(0.1,0.1,0.1,12.5),text=element_text(family="serif", size=9),
               axis.title.x=element_blank(),axis.text.x=element_blank(),axis.ticks.x=element_blank(),
               axis.title.y=element_blank(),axis.text.y=element_blank(),axis.ticks.y=element_blank(),
               panel.grid = element_blank())+geom_vline(xintercept = seq(0.5, length(counts$century),
                                                                         by = 1), color="black", size=.5)+
                scale_y_continuous(expand=c(0,0),limits=c(0,20))+scale_x_discrete(expand=c(0,0.5))+labs(title="FOR")

potcompA_total<-
ggplot(subset(counts2,group=="Group A"),aes(century,level,label=Freq))+geom_text(family="serif",size=3)+
  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title = element_blank(),
               plot.margin=margin(0.1,0.1,0.1,12.5),text=element_text(family="serif", size=9),
               axis.title.x=element_blank(),axis.text.x=element_blank(),axis.ticks.x=element_blank(),
               axis.title.y=element_blank(),axis.text.y=element_blank(),axis.ticks.y=element_blank(),
               panel.grid = element_blank())+geom_vline(xintercept = seq(0.5, length(counts2$century),
                                                                         by = 1), color="black", size=.5)+
                scale_y_continuous(expand=c(0,0),limits=c(0,20))+scale_x_discrete(expand=c(0,0.5))+labs(title="Group A")

potcompB_total<-ggplot(subset(counts2,group=="Group B"),aes(century,level,label=Freq))+geom_text(family="serif",size=3)+
  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title = element_blank(),
               plot.margin=margin(0.1,0.1,0.1,12.5),text=element_text(family="serif", size=9),
               axis.title.x=element_blank(),axis.text.x=element_blank(),axis.ticks.x=element_blank(),
               axis.title.y=element_blank(),axis.text.y=element_blank(),axis.ticks.y=element_blank(),
               panel.grid = element_blank())+geom_vline(xintercept = seq(0.5, length(counts2$century),
                                                                         by = 1), color="black", size=.5)+
                scale_y_continuous(expand=c(0,0),limits=c(0,20))+scale_x_discrete(expand=c(0,0.5))+labs(title="Group B")

potcompall_total<-ggplot(subset(countsall,!is.na(countsall$century)),aes(century,level,label=Freq))+geom_text(family="serif",size=3)+
  theme_bw()+theme(plot.title = element_text(hjust=0.5),legend.title = element_blank(),
               plot.margin=margin(0.1,0.1,0.1,12.5),text=element_text(family="serif", size=9),
               axis.title.x=element_blank(),axis.text.x=element_blank(),axis.ticks.x=element_blank(),
               axis.title.y=element_blank(),axis.text.y=element_blank(),axis.ticks.y=element_blank(),
               panel.grid = element_blank())+geom_vline(xintercept = seq(0.5, length(counts2$century),
                                                                         by = 1), color="black", size=.5)+
                scale_y_continuous(expand=c(0,0),limits=c(0,20))+scale_x_discrete(expand=c(0,0.5))+labs(title="All Tombs")

ggarrange(ggarrange(potcompKNC_total,potcompFOR_total,potcompA_total,potcompB_total,potcompall_total, ncol=5,widths=c(2.35,2,2,2,2)),ggarrange(potcompKNC,potcompFOR,potcompA,potcompB,potcomp_all, ncol=5,common.legend =TRUE,legend="bottom",widths=c(2.36,2,2,2,2)),nrow=2,heights=c(1,9),align="hv")

```

The changing composition of the tomb assemblages through time contributes further to our understanding of these developments (Fig. \@ref(fig:comps)). It appears that something of a standardised 'burial kit' crystallised by the 8th or 7th century, comprising roughly an oil vessel or two, a jug, and a drinking vessel. The proportion of 'other' vessels - *pyxides*, *askoi*, trays, stands, *kernoi* and more - declined to become a very minor part of the repertoire of the 7th century.

The 10th and 9th centuries were characterised by a more varied assemblage with, besides these 'other' shapes, mixing vessels (mostly kraters) being present in larger numbers. Though sometimes used as urns, finds of drinking sets inside kraters [e.g. the Group B T. 285: @Cavanagh1996 670-72] suggest these vessels occasionally served their more traditional function in acts of communal drinking at the graveside. Such vessels become far rarer at later dates, and are completely absent from the nearly 450 vessels datable to the 7th century from Group A. Indeed, in this regard, the assemblages of Group A in the 7th century appear, if anything, more restricted than those of the remaining KNC tombs, where kraters and other vessels were deposited in similar quantities to the preceding century. If modes of funerary conviviality persisted among the patrons of Group A, they evidently took on a somewhat different character.

```{r assemsize, fig.dim=c(3.5,3.9),dpi=600, fig.align='center',fig.cap="Aoristic sums of all vessels - except pithoi - through time, divided in each decade by the aoristic sums of pithoi from each group."}

pithoi_A<-subset(pithoi,cluster=="Group A")
pithoi_aorist_A<-colSums(pithoi_A[27:76])
pithoi_aorist_A<-data.frame(pithoi_aorist_A)
colnames(pithoi_aorist_A)<-c("aoristic_sum")
row.names(pithoi_aorist_A)<-dates
setDT(pithoi_aorist_A, keep.rownames = "decade")
pithoi_aorist_A$decade<-as.numeric(pithoi_aorist_A$decade)

pots_A<-subset(pots,cluster=="Group A")
pots_aorist_A<-colSums(pots_A[27:76])
pots_aorist_A<-data.frame(pots_aorist_A)
colnames(pots_aorist_A)<-c("aoristic_sum")
row.names(pots_aorist_A)<-dates
setDT(pots_aorist_A, keep.rownames = "decade")
pots_aorist_A$decade<-as.numeric(pots_aorist_A$decade)

pots_aorist_A$pots_over_pithoi<-ifelse(is.na(pots_aorist_A$aoristic_sum/pithoi_aorist_A$aoristic_sum),0,
                                       (pots_aorist_A$aoristic_sum/pithoi_aorist_A$aoristic_sum))

pithoi_B<-subset(pithoi,cluster=="Group B")
pithoi_aorist_B<-colSums(pithoi_B[27:76])
pithoi_aorist_B<-data.frame(pithoi_aorist_B)
colnames(pithoi_aorist_B)<-c("aoristic_sum")
row.names(pithoi_aorist_B)<-dates
setDT(pithoi_aorist_B, keep.rownames = "decade")
pithoi_aorist_B$decade<-as.numeric(pithoi_aorist_B$decade)

pots_B<-subset(pots,cluster=="Group B")
pots_aorist_B<-colSums(pots_B[27:76])
pots_aorist_B<-data.frame(pots_aorist_B)
colnames(pots_aorist_B)<-c("aoristic_sum")
row.names(pots_aorist_B)<-dates
setDT(pots_aorist_B, keep.rownames = "decade")
pots_aorist_B$decade<-as.numeric(pots_aorist_B$decade)

pots_aorist_B$pots_over_pithoi<-ifelse(is.na(pots_aorist_B$aoristic_sum/pithoi_aorist_B$aoristic_sum),0,
                                       (pots_aorist_B$aoristic_sum/pithoi_aorist_B$aoristic_sum))

pithoi_Other<-subset(pithoi,cluster=="Other")
pithoi_aorist_Other<-colSums(pithoi_Other[27:76])
pithoi_aorist_Other<-data.frame(pithoi_aorist_Other)
colnames(pithoi_aorist_Other)<-c("aoristic_sum")
row.names(pithoi_aorist_Other)<-dates
setDT(pithoi_aorist_Other, keep.rownames = "decade")
pithoi_aorist_Other$decade<-as.numeric(pithoi_aorist_Other$decade)

pots_Other<-subset(pots,cluster=="Other")
pots_aorist_Other<-colSums(pots_Other[27:76])
pots_aorist_Other<-data.frame(pots_aorist_Other)
colnames(pots_aorist_Other)<-c("aoristic_sum")
row.names(pots_aorist_Other)<-dates
setDT(pots_aorist_Other, keep.rownames = "decade")
pots_aorist_Other$decade<-as.numeric(pots_aorist_Other$decade)

pots_aorist_Other$pots_over_pithoi<-ifelse(is.na(pots_aorist_Other$aoristic_sum/pithoi_aorist_Other$aoristic_sum),0,
                                       (pots_aorist_Other$aoristic_sum/pithoi_aorist_Other$aoristic_sum))

pots_aorist_Other$pots_over_pithoi<-ifelse(pots_aorist_Other$pots_over_pithoi>50,0,pots_aorist_Other$pots_over_pithoi)

pots_aorist_A$cluster<-"Group A"
pots_aorist_B$cluster<-"Group B"
pots_aorist_Other$cluster<-"Other"
pots_aorist_all<-rbind(pots_aorist_A,pots_aorist_B,pots_aorist_Other)

ggplot()+geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf), fill=NA, colour="grey90",alpha = 0.4) +
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = 23.8, ymax = 25), fill=NA, colour="black",alpha = 0.4) +
  geom_text(data=rects,aes(xmid, 24.4,label=col),family="serif",size=2.8)+
  geom_line(data=pots_aorist_all,aes(decade,pots_over_pithoi,linetype=cluster))+
  theme_bw()+theme_bw()+theme(text=element_text(family="serif", size=10),
                              panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                              legend.title = element_blank(),legend.position = "bottom",
        legend.box.margin=margin(-10,-10,-10,-10),
                   legend.text=element_text(size=10))+
    labs(x="Date (Years BCE)", y="Aoristic sum of all vessels/pithoi")+
  scale_y_continuous(expand=c(0,0),limits=c(0,25))+
  scale_x_reverse(expand = c(0, 0),limits=c(840,630))

```

Not only did tomb assemblages become more homogenous in composition; they apparently also grew more limited and consistent in size. In Fig. \@ref(fig:histpots) the aoristic sums of all vessels - excluding *pithoi* - are divided in each decade by those of *pithoi* and show a steady convergence around approximately five vessels per *pithos* (remember, these are not counts of vessels *per se*), a figure consistent with the four-five vessel 'burial kit' suggested by Fig. \@ref(fig:comps). Again, the Group A and B tombs, which are marked in the PGB period by more plentiful assemblages, followed this general trend. Setting this against the significant rise in cremation burials made in the Group A tombs in the 8th and 7th centuries, it is clear this increase in no way equated to more substantial offerings.

```{r objects,fig.dim=c(6,5),dpi=600, fig.align='center',fig.cap="Total counts of different kinds of non-ceramic grave goods, divided into Group A, Group B, and all other tombs."}

objects$cluster <- df_prep$cluster[match(objects$tomb,df_prep$tomb)]
objects$cluster[is.na(objects$cluster)]<- "Other"
objects$cluster[objects$cluster<3]<- "Other"
objects$cluster[objects$cluster==3]<- "Group A"
objects$cluster[objects$cluster==4]<- "Group B"

ggplot(subset(objects,material!="Clay"),aes(category,color="black",fill="grey"))+geom_bar(size=0.4)+
  theme_bw()+theme(legend.title=element_blank(),legend.position="none",
      text=element_text(family="serif", size=11), panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
      axis.text.x=element_text(angle=45, vjust=1, hjust=1))+
          labs(x="Object type", y="Count")+ scale_color_manual(values="black")+scale_fill_manual(values="grey90")+
         scale_y_continuous(expand = expansion(mult = c(0, 0.05)))+
                    scale_x_discrete(expand = c(0, 0))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))+facet_col(~cluster)+
                       theme(strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA),
                             strip.text.x = element_text(size=10,margin = margin(0.05,0,0.05,0, "cm")))

```

Finally, it is not only in the case of pottery that grave goods show remarkable consistencies across tombs. Though lack of dates for most non-ceramic finds precludes analysis of any changes through time, even the raw counts of different object classes reveal clear parallels between Groups A and B, and the remaining tombs (Fig. \@ref(fig:objects)). The one exception may be the case of 'cooking' items. This class comprises so-called *obeloi* and firedogs, interpreted as cooking spits and frames on which to rest them [@Snodgrass1996]. These are essentially absent in all tombs aside from those of Groups A and B, and there is a possibility that certain kinds of funerary commensality were characteristic of these burying groups, perhaps in the form of lavish feasts of roasted meat held for the deceased.

Considering the evidence outlined above, it seems the growth in burial activity among the Group A tombs cannot be linked in any straightforward way to simple principles of 'elite' or 'wealthy' burial. In fact, distinguishing higher and lower status tombs (especially among those of chamber type) is fraught with difficulty in any period. Many of the most-used tombs prove exceptional only in their absolute quantities of artefacts, and not the accoutrements of each individual interment. This is not to say there were no instances of burials accompanied by lavish displays of grave goods; Tomb 285, for instance, may have received many of its iron artefacts (of which it contained 61), as a single deposit (ibid., p591). But even where such cases occurred, the general consistency in assemblages across tombs and through time suggest they cannot have been the norm, and for every 'rich' burial in a Group A tomb, there must have been many more that differed little from the average.

# 5. Discussion

The excavators of the KNC have lamented that, across the cemetery, '[w]ith very few exceptions, the types of offering essentially repeat themselves - almost monotonously...Within such heavily used graves as T. 75 or T. 285 we look in vain for any urn treated with particular respect. The overwhelming impression of the North Cemetery's Dark Age population is of individuals who had much in common, not least their status, not all of whom were either able or willing to advertise their success in life by the same degree of consumption in death as some of their peers' [@Coldstream1996d 720]. The foregoing analysis suggests that, while in many ways this is true, there remain patterns in the data which can nuance our understanding of the chronological developments and wider social contexts of this cemetery and that at Fortetsa.

To summarise, it has been suggested that the tombs with the most substantial assemblages can broadly be divided into two groups. Firstly, a number of tombs constructed mostly in the 10th century went on to receive large quantities of burials and associated grave goods particularly in the PGB-EG periods, a time of comparative diversity in mortuary assemblages and practices, and then largely declined. Secondly, a group of later tombs built in the 9th century witnessed a growth in cremation burial in the 8th and early 7th centuries quite unmatched by the other tombs still in use at the time. This growth, however, did not entail larger individual sets of grave goods, and the composition of the assemblages in these tombs broadly reflected those of the cemeteries as a whole.

My argument, however, is not for the existence of two distinct populations within the Knossian community. These two 'groups' are really two prevailing trajectories in the use of certain chamber tombs. There was, after all, significant overlap between them. The Group A tombs in their early years also received many imports, while the Group B tombs which did persist late into the 8th and 7th centuries exhibited a retraction in assemblage size and diversity comparable to Group A. The similarities in their assemblages - not only in total, but as they changed through the centuries, and in specific artefact classes like *obeloi* and firedogs - suggest similar social collectives, with closely aligned mortuary rites. Their major divergence came in the chronological offset of their foundation, use, and abandonment. Cavanagh (1996: 659) has shown how most chamber tombs at the KNC lasted no more than three-four generations, with only around 15% used for over two centuries. The dwindling of the Group B tombs in the 8th century, then, would be nothing extraordinary. Furthermore, it may account for features like their greater numbers of imports and other grave goods per burial. Lacking the rise in repetitive, more homogenous offerings which characterised the Group A tombs, remnants of a period of more plural funerary rites may have left traces in the assemblages of Group B. Groups A and B reflect, in other words, two episodes in the life of the cemeteries, wherein wider socio-economic and political changes prompted reformulations of funerary practice, away from the more diverse modes of PGB-EG and toward the increased, iterative burial activity exemplified by Group A.

This latter trend likely corresponded to a broadening of access to these tombs and their associated rites, a conclusion which naturally recalls the work of Morris [-@Morris1987 171-211], who argued 30 years ago that the emergence of the *polis* across the Aegean led to a tension between elite preferences for ostentatious display, and a downplaying of social difference in death, as promoted by the ‘middling’ ideology of the city-state. In his well-known model of the *agathoi* and *kakoi*, Morris claimed that, at various times, the relevant spheres of kinship demonstrated in burial could widen and constrict, with more and less distantly related lineages drawn into association through changing social strategies, until the pre-eminence of kinship relations in the structuring of political power declined, along with its manifestations in mortuary practice.

At Knossos, some aspects of the noted expansion in the burying population seem explicable via a model like Morris's. The basic principle, of elites employing access to burial as a strategic act of association and social signalling, is almost certainly in evidence here. That the noted spate of burials in the late 8th and early 7th centuries was concentrated solely in established, chamber tombs strongly implies that issues of tradition and lineage were at play. That is not to say that the Group A tombs were reused simply because they were old, however. The Group B tombs, after all, were older, but only saw more limited use throughout the 8th century. Rather, the Group A tombs seem to be a case of good timing, old enough to embody established rites of burial vis-à-vis new grave and tomb types, but also patronised by groups still evidently holding firm and active claims over their use, and no doubt looming large in the political community of the contemporary settlement. The wave of the Group B tombs had crested already, and their remaining patrons showed little capacity or will to engage in the concerted rise in burial seen in the Group A tombs. This, in turn, took on a life of its own, until some graves literally overflowed. In Tomb 107, the stratigraphy of the niches cut into the *dromos* suggests depositions continued to be made even after the main chamber fell into disrepair [@Coldstream1996a 150]. It was the symbolic and associative, rather than physical qualities of these tombs which became instrumental to their extensive reuse, and such qualities must have been vested in the patrons of the tombs themselves. The emphasising of group membership - of belonging - is surely crucial to understanding these trends.

Other features sit less easily with a model like Morris's, however. Firstly, the burials at issue are in many ways *not* ostentatious in comparison to the norm. We have seen that the ostensibly 'richest' tombs are far less so when one accounts for the number of burials represented. Similarly, Whitley [-@Whitley2004 437], while suggesting links between decorative styles of Orientalising *pithoi* and the number of offerings accompanying individual burials therein, conceded that the proportions of these different styles in Tomb P - the most-used tomb of all - were also replicated across Fortetsa as a whole. In other words, and as suggested in Fig. \@ref(fig:secureassems) above, though burials themselves could range in 'richness', such variation operated largely at the level of the individual, not the tomb or burial group. In addition, Morris claims that periods of expansion in the rite of burial will correspond to increased diversity in the associated assemblages. Yet, at Knossos, the range of grave goods diminished and crystallised during the late 8th and early 7th centuries. The PGB period better embodies such an expansion, with its greater variety and broader distribution of (local and imported) grave goods, not to mention the innovative ceramic decoration and reuse of BA artefacts.

The limited applicability of Morris's model to the Knossian case raises the recurrent spectre of Cretan 'exceptionalism' [@Whitley2004], the oft-noted idiosyncrasies of the island both in the structure and constitution of its later *poleis*, and the historical developments which underpinned their emergence. Crete has long been recognised as precocious in the emergence of its stratified and apparently quite stable political institutions. In the traditional view, political authority was vested in the *kosmoi*, aristocratic magistrates selected from the leading, hereditary *startoi* or ‘clans’, networks of extended lineages controlling the ancestral estates on which worked attendant serfs and/or slaves [@Willetts1955; @Willetts1982]. Aspects of this account have been maintained, but many others have been challenged, from the nature of Cretan economy [@Perlman2004a] and slavery [@Lewis2018], to the role of symposia [@Erickson2011] and the reliability of the later historical sources [@Perlman1992; cf.  @Seelentag2015]. Gagarin and Perlman's [-@Gagarin2016] recent study of the Archaic Cretan law codes reminds us that there is an often-underappreciated degree of diversity in the offices, institutions, and social groups of the early city-states. The precise nature of many of these remains equivocal, such as the *pyla* or 'tribe' and its relation to the *startos* (which may itself be a military grouping), the distinction between *astos* and *poliatas*, both probably formulations of the rights-bearing civic actor, and the ancestry and spread of the *agelaos*, an adolescent age-grade currently only attested at Eltynia [@Gagarin2016 36, 78-80]. Indeed, many terms which might illuminate social divisions and groupings within the Archaic Cretan populace are only attested at one or two sites, and even then sometimes apparently differ in form or function.

In other words, the composition of the Archaic communities of Crete remains somewhat elusive, as do the socio-political, economic and religious developments which brought them into being. The foregoing analysis, while not resolving these issues, does suggest some possible aspects of the process at Knossos. A key consideration is how we might characterise the burying groups behind the observed trends. The traditional view is that the patrons of each tomb were immediate families [@Brock1957 41; @Boardman1967 63; @Cavanagh1996 291]. The relatively short lifespan of many tombs has been taken as evidence for direct, lineal inheritance, with the tomb going out of use as families moved, individuals founded new tombs, or the line fissioned or ended [@Cavanagh1996 659]. The idea of family groups is supported by the, admittedly circumstantial, evidence of men, women and children being found within single tombs, and one instance of a (putatively) hereditary mandibular deformity in three skulls in the same tomb [@Musgrave1996].

The model of lineal inheritance has been recently challenged by Kotsonas [-@Kotsonas2011 131], however, who considers it 'probably the greatest factoid in the archaeology of Crete of the given period'. He argues that the rise in interments during the 8th century is too great to be explained by population growth, as would be necessary for the possibility of direct lineal inheritance to be preserved. Rather, the trend must reflect reconfigurations of the burying groups themselves, which Kotsonas feels may have (also or instead) comprised age- and sex-grades. While, as noted earlier, Kotsonas' suggestion is likely correct, that we are witnessing changes in access to visible burial rather than population growth, this need not negate the idea that the burying groups comprised some kind of familial unit. If the figure of around four burials per century is taken as a rough average, this would correspond to but a single family member per generation (it could scarcely be less), reflecting a very restricted rite. The Group A tombs achieved something in the region of 17 burials per century, based on the estimates presented here. While a fourfold increase is no doubt significant, a simple extension of the funerary rite to all members of a nuclear family would roughly account for the change, while the inclusion of only slightly more distant kin could well do likewise. The possibility of familial groups cannot be ruled out on numbers alone then. However, assuming that the broadening of access to burial was a socially strategic act, it does seem more plausible that it would be done to mobilise and bind together more politically meaningful - and formerly more distant - entities, be they segmentary lineages, clans, or age- and sex-grades.

Evidence from elsewhere on the island does certainly suggest that, with the emergence of the archaic *poleis* on Crete, there was a concomitant restructuring of wider social groups, institutions, and behaviours, particularly among the elite classes. These trends include a shift to unmarked, individual burials and more austere grave goods, as well as an apparent reorientation of sanctuary deposits away from extra-mural to suburban spaces, and a rise in utilitarian, mass-produced votives [@Prent2005]. At Azoria, major urban remodelling in the 7th century saw the earlier, agglutinative house forms of neighbouring Kavousi Vronda and Kastro give way to larger, more formally planned dwellings with little evidence of remodelling, while larger, apparently civic, buildings were constructed to accommodate major public gatherings [@Haggis2011; @Haggis2011b; @Haggis2014]. These architectural changes have been suggested to accompany the codification of an elite, urban class engaged in modes of civic commensality which transcended traditional ties of kinship.

It would be wrong, however, to assume that these changes occurred in a uniform or uncontested manner across the island. Recent scholarship has drawn attention to the precarious and performative aspects of aristocratic lifestyles in Greece's emerging city-states [@Duplouy2006], as well as the modes of tactical cooperation which developed to minimise competition between elites, whilst maintaining their material and social pre-eminence within wider society [@Seelentag2020]. The Knossos cemeteries provide evidence of both these countervalent forces. On the one hand, as discussed, the imports, BA antiques, kraters and drinking sets of the PGB-EG period gave way during the 8th and 7th centuries to more limited, repetitive rites, and it is tempting to read into this the gradual restriction of more ostentatious modes of funerary practice and display. At the same time, the significant expansion in access to the Group A tombs, set against the spread of simpler grave types and more modest use of other chamber tombs, suggests a period of tension and contestation in the relevant spheres of familial, corporate, and civic affiliation advertised in death. Though these other grave types are not apparent at Fortetsa as they are at the KNC, the distinct rise in later burial is just as visible - indeed, it is exemplified by Tomb P - and thus these developments would appear to have affected the Knossian community at large.

Two interpretations of the Group A tombs present themselves, then, although at present it remains difficult to decide between them. On the one hand, they may have been the sites of retrenchment by established lineage groups, who broadened access to more distant family in the face of burial rites - and a political community - moving slowly away from such kin-centred modes of public ritual, of which the traditional chamber tomb burial was archetypal. Or else, their patrons were active adopters of more extended, supra-familial modes of political affiliation, with the retention of traditional rites and venues for burial serving as a means of legitimation, as rights of access changed. In either case, it is surely salient that these developments proved ultimately abortive, and there is an irony in that the most visible burials at both cemeteries in their final decades were also those most decidedly halted by their abandonment. That these trends are traceable throughout the 8th and 7th centuries reflects, I would suggest, the fitful and protracted birth of the political community of Archaic Knossos, and the modes of public ritual deemed appropriate therein. While the later burials themselves seem to profess relative equality in death, and thus in life - with occasional ostentatious exceptions - the growing disparities in who was permitted to be buried where, make clear that the reality was far more complicated.

# 6. Conclusion
Tombs have been discovered in the last 20 years which fill in some of the gaps between the several plots assigned to the KNC, and ideally these unpublished tombs should be included in further analyses as they, along with the Teke tholos and associated tombs, are probably part of a single contiguous cemetery [@Blackman1999 113; @Catling1983 51; @French1991 68-69]. Further work should likewise extend to the previously published tombs from the wider funerary landscape of Knossos, for which the KNC and Fortetsa may serve as points of reference and comparison. Such an extension of the current analysis will form a part of my doctoral thesis, which places mortuary data alongside settlement patterning, and archaeological and textual sources to investigate socio-political changes across Crete in the EIA. In the meantime, the datasets made available alongside this article include the other published EIA tombs in the Knossos area, so that others may also explore their possible significance.

The foregoing analysis demonstrates the clear value of applying quantitative approaches to the archaeological record of the period, and to mortuary contexts more generally. Aoristic methods enabled not only a clearer sense of chronological developments in the cemeteries, they provided new quantitative data through which a cluster analysis helped identify distinctive patterns of tomb use. Though some of the developments discussed here have been identified before, they have not until now been so thoroughly isolated and explored. Furthermore, far from reinforcing simple associations between the number of finds from a tomb and the supposed 'wealth' of its patrons, the application of quantitative methods has here challenged the relationship between grave goods and social status, suggesting a more complex picture wherein the modalities and meanings of different burial rites varied through time. As more is learnt about the Archaic towns and cities of Crete, it is hoped these patterns will form part of a narrative around the creation and consolidation of new civic identities, and their impact on forms of behaviour, including burial practice. Finally, the databases produced for the above analysis contain a large of amount of unutilised data that could facilitate a range of future investigations, and all are welcome to it. With more such analyses, we could begin to a construct a fuller account of the regional diversity in mortuary, and wider social, practices among the communities of EIA and Archaic Crete.

##### pagebreak
# Acknowledgements

I would like to thank Borja Legarra Herrero and Corinna Riva for their help and advice from the earliest stages of the present research, as well as the BSA at Knossos for permitting me to examine some of the material from the Medical Faculty excavations. Many thanks to Anna Bloxam for pointing me towards aoristic analysis. My thanks also to the Editors and anonymous reviewers of this paper, whose suggestions have improved it enormously. I am indebted most of all to Professor Todd Whitelaw, for his unstinting guidance, feedback and facilitation of my ongoing research.

# Funding

This research did not receive any specific grant from funding agencies in the public, commercial, or not-for-profit sectors.

##### pagebreak
# References 
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

##### pagebreak
### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
git2r::repository(here::here())
```

### 2021-05-21 R/Samples3.R

@@ -1,0 +1,781 @@
+```{r violin, fig.dim=c(10,20), dpi=600, fig.cap="Tomb construction through time at both cemeteries. On the left, counts are based on the ceramic phase considered most likely to correspond to the tomb's first use (including simply 'PG', 'G' and 'O' where no greater specificity was possible). On the right, aoristic sums are presented, spreading the uncertainty associated with those general assignations. In both cases, tombs lacking any securely dateable pottery are excluded."}
+
+
+names(df_prep)
+aorist_tomb_dates<-subset(df_prep,total_pottery>10)
+aorist_tomb_dates<-aorist_tomb_dates[-c(2,4:15,17,19:111,113:157,208:258)]
+aorist_tomb_dates<-subset(aorist_tomb_dates,start_period!="")
+
+aorist_tomb_dates_melt<-melt(as.data.table(aorist_tomb_dates), id.vars = c("tomb","cem_condensed","start_period","start_date","total_pottery","cluster"), measure.vars = c("X1100","X1090","X1080","X1070","X1060","X1050","X1040","X1030","X1020","X1010","X1000","X990","X980","X970","X960","X950","X940","X930","X920","X910","X900","X890","X880","X870","X860","X850","X840","X830","X820","X810","X800","X790","X780","X770","X760","X750","X740","X730","X720","X710","X700","X690","X680","X670","X660","X650","X640","X630","X620","X610"))
+
+aorist_tomb_dates_melt$variable<-gsub("X","",aorist_tomb_dates_melt$variable)
+aorist_tomb_dates_melt$variable<-as.numeric(aorist_tomb_dates_melt$variable)
+
+aorist_tomb_dates_melt$value_100<-aorist_tomb_dates_melt$value*100
+aorist_tomb_dates_melt<-data.frame(aorist_tomb_dates_melt)
+df.expanded <- aorist_tomb_dates_melt[rep(row.names(aorist_tomb_dates_melt), aorist_tomb_dates_melt$value_100), 1:9]
+names(aorist_tomb_dates_melt)
+
[…]

### 2021-06-22 README.Rmd
Readme changes

@@
 # Please put your title here to include it in the file below. 
-Title <- "Family Fortunes: A Quantitative Analysis of the Early Iron Age Cemeteries of Knossos"
+Title <- "All Equal in the Presence of Death? A Quantitative Analysis of the Early Iron Age Cemeteries of Knossos"
 ```
@@
 
-This repository contains the data and code for the paper:
+This compendium contains the data and code for the paper:
 
-> Pollard, Dominic (2021). _`r Title`_. Name of journal/book, <https://doi.org/xxx/xxx>
+> Pollard, Dominic (2021). _`r Title`_. Journal of Anthropological Archaeology, 63, 101320. <https://doi.org/10.1016/j.jaa.2021.101320>
 
-The pre-print available is online here:
+### Compendium contents
 
-> Pollard, Dominic (2021). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+This compendium consists of the raw data, full R code and text, and bibliographic information used in the creation of the article 'All Equal in the Presence of Death?
+A Quantitative Analysis of the Early Iron Age Cemeteries at Knossos, Crete'.
+This article provides a novel study of two large mortuary assemblages of the Cretan Early Iron Age, which have never been systematically compared through quantitative analysis.
+As part of the author's commitment to data sharing and maximal reproducibility, this compendium makes available all associated data and materials utilised in the production of the article.
 
-### Repository contents
+The data pertain to the Knossos North and Fortetsa cemeteries, as presented in their respective publications (Brock 1957; Coldstream and Catling 1996).
+This compendium includes three related databases.
+The first contains entries for each individual ceramic vessel.
+Variables include the tomb, the vessel shape, its surviving dimensions, its ceramic period and corresponding absolute dating measures.
+In addition, all imports, ‘Oriental’ imitations, and antique (that is, Bronze Age) artefacts are identified.
+In total, this database comprises just over 4540 individual artefacts.
+
+The second is similar, but with each entry corresponding to a find other than a ceramic vessel.
+A ‘material’ variable was added as these finds were not all made of the same substance.
+This database includes some 1620 individual artefacts.
+
+The final database concerns the tombs themselves.
+Each tomb was accorded its own entry, with variables including dating (both relative and absolute), surviving dimensions, and the known and estimated number of interments, as well as tallies of the various artefact types recorded in the other two databases.
+
+These data are utilised in the article, itself written in R, with all figures and tables generated on-the-fly through code.
+The full text and code, as well as the produced images, are stored in the compendium, as is all bibliographic information, in the form of a BibTex file which can be added to any reference manager.
+
+This compendium allows the user to fully reproduce the analysis presented in the article, as well as to freely download and use the raw data for their own research.
+Further tombs in the wider Knossos area have since been added and are likewise made available in the compendium.
 
@@
 All charts and images from the paper can be found in 'analysis/figures', while the paper itself, and the BibTex file containing all cited works are found in 'analysis/paper'.
-The folder 'analysis/templates' contains the citation style file, and a template file used to set the style for the Word document produced by 'Family_Fortunes_Article.Rmd'.
+The folder 'analysis/templates' contains the citation style file, and a template file used to set the style for the Word document produced by 'Knossos_Cemeteries_Article.Rmd'.
+
+**References**:
+
+>Brock, J.K. 1957. *Fortetsa: Early Greek Tombs Near Knossos*. Cambridge: Cambridge University Press.
+
+>Coldstream, J. N., and H. W. Catling (eds.) 1996. *Knossos North Cemetery: Early Greek Tombs*. London: The British School at Athens Supplement 28. Vol. I-IV.
 
@@
 
-> Pollard, Dominic (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Pollard, Dominic (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for '`r Title`'_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <http://doi.org/10.6084/m9.figshare.13019588>
 
@@
 
-The compendium is also available from figshare at: https://doi.org/10.6084/m9.figshare.13019588.v1
+The compendium is also available from figshare at:
+
+[![DOI](https://img.shields.io/badge/doi-10.6084/m9.figshare.13019588-blue.svg?style=flat&labelColor=whitesmoke&logo=data%3Aimage%2Fpng%3Bbase64%2CiVBORw0KGgoAAAANSUhEUgAAAB8AAAAfCAYAAAAfrhY5AAAJsklEQVR42qWXd1DTaRrHf%2BiB2Hdt5zhrAUKz4IKEYu9IGiGFFJJQ0gkJCAKiWFDWBRdFhCQUF3UVdeVcRQEBxUI3yY9iEnQHb3bdW1fPubnyz%2F11M7lvEHfOQee2ZOYzPyDv%2B3yf9%2Fk95YX4fx%2BltfUt08GcFEuPR4U9hDDZ%2FVngIlhb%2FSiI6InkTgLzgDcgfvtnovhH4BzoVlrbwr55QnhCtBW4QHXnFrZbPBaQoBh4%2FSYH2EnpBEtqcDMVzB93wA%2F8AFwa23XFGcc8CkT3mxz%2BfXWtq9T9IQlLIXYEuHojudb%2BCM7Hgdq8ydi%2FAHiBXyY%2BLjwFlAEnS6Jnar%2FvnQVhvdzasad0eKvWZKe8hvDB2ofLZ%2FZEcWsh%2BhyIuyO5Bxs2iZIE4nRv7NWAb0EO8AC%2FWPxjYAWuOEX2MSXZVgPxzmRL3xKz3ScGpx6p6QnOx4mDIFqO0w6Q4fEhO5IzwxlSwyD2FYHzwAW%2BAZ4fEsf74gCumykwNHskLM7taQxLYjjIyy8MUtraGhTWdkfhkFJqtvuVl%2F9l2ZquDfEyrH8B0W06nnpH3JtIyRGpH1iJ6SfxDIHjRXHJmdQjLpfHeN54gnfFx4W9QRnovx%2FN20aXZeTD2J84hn3%2BqoF2Tqr14VqTPUCIcP%2B5%2Fly4qC%2BUL3sYxSvNj1NwsVYPsWdMUfomsdkYm3Tj0nbV0N1wRKwFe1MgKACDIBdMAhPE%2FwicwNWxll8Ag40w%2BFfhibJkGHmutjYeQ8gVlaN%2BjO51nDysa9TwNUFMqaGbKdRJZFfOJSp6mkRKsv0rRIpEVWjAvyFkxNOEpwvcAVPfEe%2Bl8ojeNTx3nXLBcWRrYGxSRjDEk0VlpxYrbe1ZmaQ5xuT0u3r%2B2qe5j0J5uytiZPGsRL2Jm32AldpxPUNJ3jmmsN4x62z1cXrbedXBQf2yvIFCeZrtyicZZG2U2nrrBJzYorI2EXLrvTfCSB43s41PKEvbZDEfQby6L4JTj%2FfIwam%2B4%2BwucBu%2BDgNK05Nle1rSt9HvR%2FKPC4U6LTfvUIaip1mjIa8fPzykii23h2eanT57zQ7fsyYH5QjywwlooAUcAdOh5QumgTHx6aAO7%2FL52eaQNEShrxfhL6albEDmfhGflrsT4tps8gTHNOJbeDeBlt0WJWDHSgxs6cW6lQqyg1FpD5ZVDfhn1HYFF1y4Eiaqa18pQf3zzYMBhcanlBjYfgWNayAf%2FASOgklu8bmgD7hADrk4cRlOL7NSOewEcbqSmaivT33QuFdHXj5sdvjlN5yMDrAECmdgDWG2L8P%2BAKLs9ZLZ7dJda%2BB4Xl84t7QvnKfvpXJv9obz2KgK8dXyqISyV0sXGZ0U47hOA%2FAiigbEMECJxC9aoKp86re5O5prxOlHkcksutSQJzxZRlPZmrOKhsQBF5zEZKybUC0vVjG8PqOnhOq46qyDTDnj5gZBriWCk4DvXrudQnXQmnXblebhAC2cCB6zIbM4PYgGl0elPSgIf3iFEA21aLdHYLHUQuVkpgi02SxFdrG862Y8ymYGMvXDzUmiX8DS5vKZyZlGmsSgQqfLub5RyLNS4zfDiZc9Edzh%2FtCE%2BX8j9k%2FqWB071rcZyMImne1SLkL4GRw4UPHMV3jjwEYpPG5uW5fAEot0aTSJnsGAwHJi2nvF1Y5OIqWziVCQd5NT7t6Q8guOSpgS%2Fa1dSRn8JGGaCD3BPXDyQRG4Bqhu8XrgAp0yy8DMSvvyVXDgJcJTcr1wQ2BvFKf65jqhvmxXUuDpGBlRvV36XvGjQzLi8KAKT2lYOnmxQPGorURSV0NhyTIuIyqOmKTMhQ%2BieEsgOgpc4KBbfDM4B3SIgFljvfHF6cef7qpyLBXAiQcXvg5l3Iunp%2FWv4dH6qFziO%2BL9PbrimQ9RY6MQphEfGUpOmma7KkGzuS8sPUFnCtIYcKCaI9EXo4HlQLgGrBjbiK5EqMj2AKWt9QWcIFMtnVvQVDQV9lXJJqdPVtUQpbh6gCI2Ov1nvZts7yYdsnvRgxiWFOtNJcOMVLn1vgptVi6qrNiFOfEjHCDB3J%2BHDLqUB77YgQGwX%2Fb1eYna3hGKdlqJKIyiE4nSbV8VFgxmxR4b5mVkkeUhMgs5YTi4ja2XZ009xJRHdkfwMi%2BfocaancuO7h%2FMlcLOa0V%2FSw6Dq47CumRQAKhgbOP8t%2BMTjuxjJGhXCY6XpmDDFqWlVYbQ1aDJ5Cptdw4oLbf3Ck%2BdWkVP0LpH7s9XLPXI%2FQX8ws%2Bj2In63IcRvOOo%2BTTjiN%2BlssfRsanW%2B3REVKoavBOAPTXABW4AL7e4NygHdpAKBscmlDh9Jysp4wxbnUNna3L3xBvyE1jyrGIkUHaqQMuxhHElV6oj1picvgL1QEuS5PyZTEaivqh5vUCKJqOuIgPFGESns8kyFk7%2FDxyima3cYxi%2FYOQCj%2F%2B9Ms2Ll%2Bhn4FmKnl7JkGXQGDKDAz9rUGL1TIlBpuJr9Be2JjK6qPzyDg495UxXYF7JY1qKimw9jWjF0iV6DRIqE%2B%2FeWG0J2ofmZTk0mLYVd4GLiFCOoKR0Cg727tWq981InYynvCuKW43aXgEjofVbxIqrm0VL76zlH3gQzWP3R3Bv9oXxclrlO7VVtgBRpSP4hMFWJ8BrUSBCJXC07l40X4jWuvtc42ofNCxtlX2JH6bdeojXgTh5TxOBKEyY5wvBE%2BACh8BtOPNPkApjoxi5h%2B%2FFMQQNpWvZaMH7MKFu5Ax8HoCQdmGkJrtnOiLHwD3uS5y8%2F2xTSDrE%2F4PT1yqtt6vGe8ldMBVMEPd6KwqiYECHDlfbvzphcWP%2BJiZuL5swoWQYlS%2Br7Yu5mNUiGD2retxBi9fl6RDGn4Ti9B1oyYy%2BMP5G87D%2FCpRlvdnuy0PY6RC8BzTA40NXqckQ9TaOUDywkYsudxJzPgyDoAWn%2BB6nEFbaVxxC6UXjJiuDkW9TWq7uRBOJocky9iMfUhGpv%2FdQuVVIuGjYqACbXf8aa%2BPeYNIHZsM7l4s5gAQuUAzRUoT51hnH3EWofXf2vkD5HJJ33vwE%2FaEWp36GHr6GpMaH4AAPuqM5eabH%2FhfG9zcCz4nN6cPinuAw6IHwtvyB%2FdO1toZciBaPh25U0ducR2PI3Zl7mokyLWKkSnEDOg1x5fCsJE9EKhH7HwFNhWMGMS7%2BqxyYsbHHRUDUH4I%2FAheQY7wujJNnFUH4KdCju83riuQeHU9WEqNzjsJFuF%2FdTDAZ%2FK7%2F1WaAU%2BAWymT59pVMT4g2AxcwNa0XEBDdBDpAPvgDIH73R25teeuAF5ime2Ul0OUIiG4GpSAEJeYW9wDTf43wfwHgHLKJoPznkwAAAABJRU5ErkJggg%3D%3D)](http://doi.org/10.6084/m9.figshare.13019588)
 

### 2021-06-22 analysis/paper/Knossos_Cemeteries_Article.Rmd
Readme changes

@@
 
-Interest in the period grew in subsequent decades, as excavation at sites like the Toumba cemetery at Lefkandi in Euboea [@Desborough1970], and the North Cemetery at Knossos - discussed in this paper - yielded evidence for the maintenance of long-distance networks of exchange and interaction at the dawn of the EIA, as well as artefacts and rituals redolent of the Homeric epics [@Catling1995].
+Interest in the period grew in subsequent decades, as excavation at sites like the Toumba cemetery at Lefkandi in Euboea [e.g.
+@Desborough1970], and the North Cemetery at Knossos - discussed in this paper - yielded evidence for the maintenance of long-distance networks of exchange and interaction at the dawn of the EIA, as well as artefacts and rituals redolent of the Homeric epics [@Catling1995].
 The growth of archaeological survey, with its broad chronological purview, began revealing long-term settlement dynamics in many regions, often transcending the traditional divides of the LBA, EIA and Classical era.

### 2021-06-22 analysis/paper/Knossos_Cemeteries_Article_B&W.Rmd
Readme changes

--- 2021-06-22 analysis/paper/Knossos_Cemeteries_Article.Rmd
@@
 library(cluster)
+library(ggpattern)
 
@@
 
-```{r tombconstruction, fig.dim=c(7,4), dpi=600, fig.cap="Tomb construction through time at both cemeteries.
+```{r tombconstructionbW, fig.dim=c(7,4), dpi=600, fig.cap="Tomb construction through time at both cemeteries.
 A) Counts based on the ceramic phase considered most likely to correspond to the tomb's first use (including simply 'PG', 'G' and 'O' where no greater specificity was possible).
@@
 
-new_tomb_counts_plot<-ggplot(data=subset(tombs,!is.na(start_period)),aes(start_period, fill=forcats::fct_rev(tomb_type_condensed),color=forcats::fct_rev(tomb_type_condensed)))+geom_bar(width=1,size=0.4)+
-  scale_color_manual(values=c("black","black","black","black","black","black","black","black"),labels=c("Unknown","Shaft Grave","Pit Cave","Pithos Burial","Pit","Other","Larnax Burial","Chamber Tomb"))+
+new_tomb_counts_plot<-ggplot(data=subset(tombs,!is.na(start_period)),aes(start_period,pattern_type=forcats::fct_rev(tomb_type_condensed)))+geom_bar_pattern(pattern              = 'magick',
+      pattern_fill         = 'black',
+      pattern_aspect_ratio = 1.75,
+      fill                 = 'white',
+      colour               = 'black',width=1)+
         theme_bw()+theme(legend.title=element_blank(),legend.spacing.x = unit(0.1,"cm"), text=element_text(family="serif", size=12),
@@
           panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
-            labs(x="Period", y="Count of new tombs")+scale_fill_manual(values=cbPalette,
-                                                                                        labels=c("Unknown","Shaft Grave","Pit Cave",
-                                                                                        "Pithos Burial","Pit","Other","Larnax Burial","Chamber Tomb"))+
+            labs(x="Period", y="Count of new tombs")+scale_pattern_type_discrete(choices = ggpattern::magick_pattern_names,labels=c("Unknown","Shaft Grave","Pit Cave","Pithos Burial","Pit","Other","Larnax Burial","Chamber Tomb"))+
              scale_y_continuous(expand=c(0,0),limits=c(0,25))+scale_x_discrete(expand=c(0,0))+
@@
 The periods of construction shown in Fig.
-\@ref(fig:tombconstruction) correspond, therefore, largely with the earliest surviving pottery from each tomb.
+\@ref(fig:tombconstructionbw) correspond, therefore, largely with the earliest surviving pottery from each tomb.
 On this basis, it would appear that the KNC began with many tombs built during the - admittedly prolonged - SM period, followed by a protracted decline in construction, offset by resurgences in the LPG and LG periods.
@@
 
-```{r centsize,fig.dim=c(5,5), dpi=600, fig.cap="Scatterplots showing the relationship between the number of vessels dated to the A) 10th, B) 9th, C) 8th, and D) 7th centuries in tombs and the total number of vessels found therein.
+```{r centsizebw,fig.dim=c(5,5), dpi=600, fig.cap="Scatterplots showing the relationship between the number of vessels dated to the A) 10th, B) 9th, C) 8th, and D) 7th centuries in tombs and the total number of vessels found therein.
 In each case, only tombs built by the century in question are plotted."}
@@
 tombs$vessels_7th<-rowSums(tombs[,200:209])
-
-cbPalette <- c("#E69F00", "#56B4E9")
 
@@
                         scale_y_continuous(expand = expansion(mult = c(0.001, 0.05)),limits=c(0,470))+
-  scale_color_manual(values=cbPalette)
+  scale_color_grey(start=0.2,end=0.7)
 
@@
 This suspicion is well supported by Fig.
-\@ref(fig:centsize), where the strongest correlations are seen between tombs which received more 8th and 7th century vessels, and which had the largest assemblages overall.
+\@ref(fig:centsizebw), where the strongest correlations are seen between tombs which received more 8th and 7th century vessels, and which had the largest assemblages overall.
 Therefore, the cemeteries' largest tombs seem a sensible starting point for investigating the dynamics of this late rise in burial activity.
@@
 
-```{r cluster, fig.dim=c(3.5,4.1),dpi=600,fig.align='center',out.width="0.5%", fig.cap="Results of cluster analysis based on aoristic sums per decade for each tomb.
+```{r clusterbw, fig.dim=c(3.5,4.1),dpi=600,fig.align='center',out.width="0.5%", fig.cap="Results of cluster analysis based on aoristic sums per decade for each tomb.
 Ward's method was used and tombs that yielded no pots at all were excluded.
@@
   geom_text_repel(label=df_prep2$tomb,color="Black",family="serif",size=2.5)+
-                scale_fill_manual('Cluster',values = cbPalette)+scale_color_manual('Cluster',values = cbPalette)+
+                scale_fill_grey('Cluster')+scale_color_grey('Cluster')+
   scale_shape_manual('Cluster', values=c(15,16,17,18))+scale_y_reverse()
@@
 This analysis was done using Ward's method, and Fig.
-\@ref(fig:cluster) presents the results as a scatterplot, with the four groups separated by way of a k-means clustering.
+\@ref(fig:clusterbw) presents the results as a scatterplot, with the four groups separated by way of a k-means clustering.
 Cluster 1 comprises the vast majority of tombs in both cemeteries, while tombs Π (Pi) and 40 appear to be outliers.
@@
 
-```{r tombgroupdimensions, fig.dim = c(7, 4),dpi=600, fig.align='center', fig.cap="A) Histograms showing the distribution of chamber tomb chamber areas, in square meters.
+```{r tombgroupdimensionsbw, fig.dim = c(7, 4),dpi=600, fig.align='center', fig.cap="A) Histograms showing the distribution of chamber tomb chamber areas, in square meters.
 B) The same but for dromos areas.
 In each case, these measurements are planimetric rather than volumetric, as accurate heights and depths for chambers and dromoi are so rarely recorded."}
-
-cbPalette <- c("#56B4E9", "#009E73","#E69F00")
 
@@
             labs(x="Chamber area ("~m^2~")", y="Count of tombs per 0.5"~m^2~"")+
-                        scale_fill_manual(values=cbPalette)+scale_color_manual(values=c("black","black","black"))+
+                        scale_fill_grey(start=0.2,end=0.9)+scale_color_manual(values=c("black","black","black"))+
                              scale_x_continuous(expand=expansion(mult=c(0,0.05)),breaks = seq(0, 10, by = 2))+
@@
             labs(x="Dromos area ("~m^2~")", y="Count of tombs per 0.5"~m^2~"")+
-                        scale_fill_manual(values=cbPalette)+scale_color_manual(values=c("black","black","black"))+
+                        scale_fill_grey(start=0.2,end=0.9)+scale_color_manual(values=c("black","black","black"))+
                              scale_x_continuous(expand=expansion(mult=c(0,0.05)),breaks = seq(0, 12, by = 2))+
@@
 However, in considering those tombs where chamber and *dromos* area are calculable (Fig.
-\@ref(fig:tombgroupdimensions)), Groups A and B do not seem systematically marked out by their scale.
+\@ref(fig:tombgroupdimensionsbw)), Groups A and B do not seem systematically marked out by their scale.
 The *dromos*, as the most visible area with regard to the funerary rite, might be considered the more indicative measure, and indeed three Group A and B tombs (75, 219 and 100) had very substantial *dromoi.* These features were also commonly disturbed by later activity, meaning other long *dromoi* may have gone unrecognised, but on the present evidence there is no general demarcation between the tombs based on size.
 
-```{r tombdimensions, fig.dim = c(7, 4), dpi=600, fig.align='center', fig.cap="Scatterplots showing A) chamber areas of 10th century tombs, plotted against the number of 10th and 9th century vessels they yielded respectively.
+```{r tombdimensionsbw, fig.dim = c(7, 4), dpi=600, fig.align='center', fig.cap="Scatterplots showing A) chamber areas of 10th century tombs, plotted against the number of 10th and 9th century vessels they yielded respectively.
 B) The same but using combined chamber and *dromos* areas.
@@
 
-cbPalette <- c("#56B4E9", "#009E73","#E69F00")
-
   cham_9th_pots_9th<-ggplot()+geom_point(data=subset(tombs,century_built=="9th"),aes(chamber_area,vessels_9th,color=cluster),size=1)+  stat_cor(data=subset(tombs,century_built=="9th"),aes(chamber_area,vessels_9th),method="pearson",label.x.npc=0.1,label.y.npc=0.9,family="serif",size=3,label.sep="\n")+
@@
               labs(x="Chamber area ("~m^2~")", y="9th century vessels")+
-                          scale_color_manual(values=cbPalette)+
+                          scale_color_grey(start=0.2,end=0.8)+
                                scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
@@
               labs(x="Chamber area ("~m^2~")", y="8th century vessels")+
-                          scale_color_manual(values=cbPalette)+
+                          scale_color_grey(start=0.2,end=0.8)+
                                scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
@@
               labs(x="Dromos+chamber area ("~m^2~")", y="9th century vessels")+
-                          scale_color_manual(values=cbPalette)+
+                          scale_color_grey(start=0.2,end=0.8)+
                                scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
@@
               labs(x="Dromos+chamber area ("~m^2~")", y="8th century vessels")+
-                          scale_color_manual(values=cbPalette)+
+                          scale_color_grey(start=0.2,end=0.8)+
                                scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
                         scale_y_continuous(expand = expansion(mult = c(0, 0.05)))
-
-  cbPalette <- c("#009E73","#E69F00")
 
@@
               labs(x="Chamber area ("~m^2~")", y="10th century vessels")+
-                          scale_color_manual(values=cbPalette)+
+                          scale_color_grey(start=0.5,end=0.8)+
                                scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
@@
               labs(x="Chamber area ("~m^2~")", y="9th century vessels")+
-                          scale_color_manual(values=cbPalette)+
+                          scale_color_grey(start=0.5,end=0.8)+
                                scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
@@
               labs(x="Dromos+chamber area ("~m^2~")", y="10th century vessels")+
-                          scale_color_manual(values=cbPalette)+
+                          scale_color_grey(start=0.5,end=0.8)+
                                scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
@@
               labs(x="Dromos+chamber area ("~m^2~")", y="9th century vessels")+
-                          scale_color_manual(values=cbPalette)+
+                          scale_color_grey(start=0.5,end=0.8)+
                                scale_x_continuous(expand = expansion(mult = c(0, 0.05)))+
@@
 Fig.
-\@ref(fig:tombdimensions)A shows that the Group B tombs of the 10th century were among the largest of the time, and went on to receive the greatest quantities of vessels in the subsequent 9th century.
+\@ref(fig:tombdimensionsbw)A shows that the Group B tombs of the 10th century were among the largest of the time, and went on to receive the greatest quantities of vessels in the subsequent 9th century.
 It is difficult to say if this correlation suggests the tombs' builders pre-empted their later use, or if they were motivated by some kind of competitive display (or a combination of these things); however, a similar situation is not immediately clear in the case of the Group A tombs.
@@
 
-```{r histpots, fig.dim=c(5,5),dpi=600, fig.align='center',fig.cap="Empirical cumulative distribution function (ECDF) plots showing the distribution of A) total pottery counts across tombs of Group A, Group B, and Others and B) those counts divided by the estimated number of burials in each tomb.
+```{r histpotsbw, fig.dim=c(5,5),dpi=600, fig.align='center',fig.cap="Empirical cumulative distribution function (ECDF) plots showing the distribution of A) total pottery counts across tombs of Group A, Group B, and Others and B) those counts divided by the estimated number of burials in each tomb.
 C) and D) show the same, but with all grave goods aside from ceramic vessels.
@@
 
-cbPalette <- c("#56B4E9", "#009E73","#E69F00")
-
 no_pith<-aggregate(pots$shape!="Pithos", by=list(tomb=pots$tomb), FUN=sum)
@@
   labs(x="Total pottery", y="Proportion of tombs")+
-  scale_color_manual(values = cbPalette)+
+  scale_color_grey(start=0.1,end=0.8)+
   scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))
@@
   labs(x="Total pottery/burials", y="Proportion of tombs")+
-  scale_color_manual(values = cbPalette)+
+  scale_color_grey(start=0.1,end=0.8)+
   scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))
@@
   labs(x="Total objects", y="Proportion of tombs")+
-  scale_color_manual(values = cbPalette)+
+  scale_color_grey(start=0.1,end=0.8)+
   scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))
@@
   labs(x="Total objects/burials", y="Proportion of tombs")+
-  scale_color_manual(values = cbPalette)+
+  scale_color_grey(start=0.1,end=0.8)+
   scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))
@@
 As seen in Fig.
-\@ref(fig:histpots), even this simple step quite significantly alters the gap between Groups A and B and the remaining tombs.
+\@ref(fig:histpotsbw), even this simple step quite significantly alters the gap between Groups A and B and the remaining tombs.
 In the case of pottery, the disparity is much reduced; in that of all grave goods besides pottery, it is almost eradicated.
@@
   labs(x="Assemblage size", y="Proportion of tombs")+
-  scale_color_manual(values = cbPalette)+
+  scale_color_grey(start=0.1,end=0.8)+
   scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))
@@
 
-```{r hists, fig.dim=c(7,5),dpi=600, fig.align='center',fig.cap="ECDF plots showing the distribution of A) items of adornment (beads, necklaces, rings, pins, fibulae etc.), weapons, and imported items by tomb across both cemeteries, and B) those counts divided by the estimated number of burials in each tomb."}
+```{r histsbw, fig.dim=c(7,5),dpi=600, fig.align='center',fig.cap="ECDF plots showing the distribution of A) items of adornment (beads, necklaces, rings, pins, fibulae etc.), weapons, and imported items by tomb across both cemeteries, and B) those counts divided by the estimated number of burials in each tomb."}
 
@@
   labs(x="Total adornment", y="Proportion of tombs")+
-  scale_color_manual(values = cbPalette)+
+  scale_color_grey(start=0.1,end=0.8)+
   scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))
@@
   labs(x="Total weapons", y="Proportion of tombs")+
-  scale_color_manual(values = cbPalette)+
+  scale_color_grey(start=0.1,end=0.8)+
   scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))
@@
   labs(x="Total imports", y="Proportion of tombs")+
-  scale_color_manual(values = cbPalette)+
+  scale_color_grey(start=0.1,end=0.8)+
   scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))
@@
   labs(x="Total adornment/burials", y="Proportion of tombs")+
-  scale_color_manual(values = cbPalette)+
+  scale_color_grey(start=0.2,end=0.8)+
   scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))
@@
   labs(x="Total weapons/burials", y="Proportion of tombs")+
-  scale_color_manual(values = cbPalette)+
+  scale_color_grey(start=0.1,end=0.8)+
   scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))
@@
   labs(x="Total imports/burials", y="Proportion of tombs")+
-  scale_color_manual(values = cbPalette)+
+  scale_color_grey(start=0.1,end=0.8)+
   scale_y_continuous(expand=expansion(mult=c(0,0.05)))+scale_x_continuous(expand=c(0,0))
@@
 Similar patterns are found among many classes of grave good (Fig.
-\@ref(fig:histpots)).
+\@ref(fig:histsbw)).
 In both Groups A and B, items of adornment (pins, fibulae, rings, beads etc.) and weapons, when divided by the estimated number of burials in each tomb, fall in line with, or even below, the distribution across all other tombs.
@@
 
-potcompKNC<-ggplot(subset(pots_no_pith,century!="" & cem_condensed=="KNC" & cluster=="Other"),aes(century,fill=type))+geom_bar(position="fill")+
+potcompKNC<-ggplot(subset(pots_no_pith,century!="" & cem_condensed=="KNC" & cluster=="Other"),aes(century,fill=type,color=type))+geom_bar(position="fill",size=0.5)+
   theme_bw()+theme(plot.title = element_text(hjust=0.5), axis.title.x = element_blank(),
@@
                plot.margin=margin(0.1,0.1,0.1,0.1),legend.title.align=0.5)+
-                 labs(y="Proportion of total assemblage")+ scale_fill_manual(values=cbPalette)+ 
+                 labs(y="Proportion of total assemblage")+ scale_fill_grey(start=0.1,end=0.9)+
+  scale_color_manual(values=c("Black","Black","Black","Black","Black","Black"))+
                     scale_y_continuous(expand=c(0,0.0))+scale_x_discrete(expand=c(0,0.0))+
-  guides(fill=guide_legend(title="Century BCE",title.position = "top",nrow=1))
-
-potcompFOR<-ggplot(subset(pots_no_pith,century!="" & cem_condensed=="FOR" & cluster=="Other"),aes(century,fill=type))+geom_bar(position="fill")+
+  guides(fill=guide_legend(title="Century BCE",title.position = "top",nrow=1),
+         color=guide_legend(title="Century BCE",title.position = "top",nrow=1))
+
+potcompFOR<-ggplot(subset(pots_no_pith,century!="" & cem_condensed=="FOR" & cluster=="Other"),aes(century,fill=type,color=type))+geom_bar(position="fill",size=0.5)+
   theme_bw()+theme(plot.title = element_text(hjust=0.5),axis.title.x = element_blank(),axis.text.y = element_blank(),
@@
                plot.margin=margin(0.1,0.1,0.1,0.1),legend.title.align=0.5)+
-                 labs(y="")+
-                  scale_fill_manual(values=cbPalette)+ 
+                 labs(y="")+scale_fill_grey(start=0.1,end=0.9)+ 
+  scale_color_manual(values=c("Black","Black","Black","Black","Black","Black"))+
                     scale_y_continuous(expand=c(0,0.0))+scale_x_discrete(expand=c(0,0.0))+
-  guides(fill=guide_legend(title="Century BCE",title.position = "top",nrow=1))
-
-potcompA<-ggplot(subset(pots_no_pith,century!="" & cluster=="Group A"),aes(century,fill=type))+geom_bar(position="fill")+
+  guides(fill=guide_legend(title="Century BCE",title.position = "top",nrow=1),
+         color=guide_legend(title="Century BCE",title.position = "top",nrow=1))
+
+potcompA<-ggplot(subset(pots_no_pith,century!="" & cluster=="Group A"),aes(century,fill=type,color=type))+geom_bar(position="fill",size=0.5)+
   theme_bw()+theme(plot.title = element_text(hjust=0.5),axis.title.x = element_blank(),axis.text.y = element_blank(),
@@
                plot.margin=margin(0.1,0.1,0.1,0.1),legend.title.align=0.5)+
-                 labs(y="")+
-                  scale_fill_manual(values=cbPalette)+ scale_x_discrete(expand=c(0,0.0))+
-                    scale_y_continuous(expand=c(0,0.0))+guides(fill=guide_legend(title="Century BCE",title.position = "top",nrow=1))
-
-potcompB<-ggplot(subset(pots_no_pith,century!="" & cluster=="Group B"),aes(century,fill=type))+geom_bar(position="fill")+
+                 labs(y="")+scale_fill_grey(start=0.1,end=0.9)+
+  scale_color_manual(values=c("Black","Black","Black","Black","Black","Black"))+
+  scale_x_discrete(expand=c(0,0.0))+scale_y_continuous(expand=c(0,0.0))+guides(fill=guide_legend(title="Century BCE",title.position = "top",nrow=1),
+         color=guide_legend(title="Century BCE",title.position = "top",nrow=1))
+
+potcompB<-ggplot(subset(pots_no_pith,century!="" & cluster=="Group B"),aes(century,fill=type,color=type))+geom_bar(position="fill",size=0.5)+
   theme_bw()+theme(plot.title = element_text(hjust=0.5),axis.title.x = element_blank(),axis.text.y = element_blank(),
@@
                plot.margin=margin(0.1,0.1,0.1,0.1),legend.title.align=0.5)+
-                 labs(y="")+
-                  scale_fill_manual(values=cbPalette)+ scale_x_discrete(expand=c(0,0.0))+
-                    scale_y_continuous(expand=c(0,0.0))+guides(fill=guide_legend(title="Century BCE",title.position = "top",nrow=1))
-
-potcomp_all<-ggplot(subset(pots_no_pith,!is.na(century)),aes(century,fill=type))+geom_bar(position="fill")+
+                 labs(y="")+scale_color_manual(values=c("Black","Black","Black","Black","Black","Black"))+
+                  scale_fill_grey(start=0.1,end=0.9)+ scale_x_discrete(expand=c(0,0.0))+
+                    scale_y_continuous(expand=c(0,0.0))+
+  guides(fill=guide_legend(title="Century BCE",title.position = "top",nrow=1),
+         color=guide_legend(title="Century BCE",title.position = "top",nrow=1))
+
+potcomp_all<-ggplot(subset(pots_no_pith,!is.na(century)),aes(century,fill=type,color=type))+geom_bar(position="fill",size=0.5)+
   theme_bw()+theme(plot.title = element_text(hjust=0.5),axis.title.x = element_blank(),axis.text.y = element_blank(),
@@
                plot.margin=margin(0.1,0.1,0.1,0.1),legend.title.align=0.5)+
-                 labs(y="")+
-                  scale_fill_manual(values=cbPalette)+ scale_x_discrete(expand=c(0,0.0))+
-                    scale_y_continuous(expand=c(0,0.0))+guides(fill=guide_legend(title="Century BCE",title.position = "top",nrow=1))
+                 labs(y="")+scale_color_manual(values=c("Black","Black","Black","Black","Black","Black"))+
+                  scale_fill_grey(start=0.1,end=0.9)+ scale_x_discrete(expand=c(0,0.0))+
+                    scale_y_continuous(expand=c(0,0.0))+
+  guides(fill=guide_legend(title="Century BCE",title.position = "top",nrow=1),
+         color=guide_legend(title="Century BCE",title.position = "top",nrow=1))
 ###
@@
 In Fig.
-\@ref(fig:histpots) the aoristic sums of all vessels - excluding *pithoi* - are divided in each decade by those of *pithoi* and show a steady convergence around approximately five vessels per *pithos* (remember, these are not counts of vessels *per se*), a figure consistent with the four-five vessel 'burial kit' suggested by Fig.
+\@ref(fig:assemsize) the aoristic sums of all vessels - excluding *pithoi* - are divided in each decade by those of *pithoi* and show a steady convergence around approximately five vessels per *pithos* (remember, these are not counts of vessels *per se*), a figure consistent with the four-five vessel 'burial kit' suggested by Fig.
 \@ref(fig:comps).
